### 2019-12-01 aap/Revision/vr-mcmc-revision.tex
Revision

\documentclass[bj]{imsart}
\makeatletter
\def\set@curr@file#1{%
  \begingroup
    \escapechar\m@ne
    \xdef\@curr@file{\expandafter\string\csname #1\endcsname}%
  \endgroup
}
\def\quote@name#1{"\quote@@name#1\@gobble""}
\def\quote@@name#1"{#1\quote@@name}
\def\unquote@name#1{\quote@@name#1\@gobble"}
\makeatother
\usepackage{graphics}

\usepackage{lineno,hyperref}
\usepackage{amsfonts,amssymb}
\usepackage{mathtools}
%\usepackage{mathbbol}
\usepackage{bbm}
\usepackage{multirow,algorithm2e}%
\usepackage{xargs}
%\usepackage{showlabels}

\newcommandx{\CPE}[3][1=]{{\mathsf E}_{#1}\left[\left. #2 \, \right| #3 \right]}
\usepackage[textsize=footnotesize]{todonotes} % pour voir les commentaires

\newcommand{\eric}[1]{\todo[color=purple!10, author=Eric]{#1}}
\newcommand{\erici}[1]{\todo[inline,color=purple!10, author=Eric]{#1}}
\newcommand{\denis}[1]{\todo[color=green!10, author=Eric]{#1}}
\newcommand{\denisi}[1]{\todo[inline,color=green!10, author=Eric]{#1}}

\def\rmd{\mathrm{d}}
\def\rme{\mathrm{e}}
\def\rset{\mathbb{R}}
\def\NtrainPath{T}
\def\TrainSet{\mathcal{T}}
\modulolinenumbers[5]
\newcommand{\proofendsign}{$\Box$}
\newtheorem{thm}{Theorem}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{prop}[thm]{Proposition}
\newtheorem{remark}{Remark}
\newtheorem{cor}{Corollary}
\newtheorem{example}{Example}
\newenvironment{proof}{{\noindent \bf Proof }}
 {{\hspace*{\fill}\proofendsign\par\bigskip}}
\def\TV{\operatorname{TV}}
\newcommand{\jac}[1]{\operatorname{J} _{#1}}

\newcommand*{\argmin}{\operatornamewithlimits{arg\,min}}
\newcommand*{\const}{\mathrm{const}}
\usepackage[lofdepth,lotdepth]{subfig}
\usepackage{graphicx}
\newcommand*{\ol}{\overline}

\begin{document}

\begin{frontmatter}
\title{Variance reduction for MCMC methods via martingale representations}
\runtitle{Variance reduction for MCMC}
%\thankstext{T1}{This research was }

\begin{aug}
\author{\snm{D. Belomestny}\thanksref{a,b}\ead[label=e1]{denis.belomestny@uni-due.de}}
\and
\author{\snm{E. Moulines}\thanksref{c, b}\ead[label=e2]{eric.moulines@polytechnique.edu}}
\and
\author{\snm{S. Samsonov}\thanksref{b}\ead[label=e3]{svsamsonov@hse.ru}}
\and
\author{\snm{N. Shagadatov}\thanksref{b}\ead[label=e4]{nshagadatov@edu.hse.ru}}

\runauthor{D. Belomestny et al}

\affiliation{Duisburg-Essen University, National Research University Higher School of Economics,  Ecole Polytechnique  and Duisburg-Essen University}

\address[a]{
Duisburg-Essen University,
Faculty of Mathematics,
D-45127 Essen
Germany }

\address[c]{
Centre de Math\'ematiques Appliqu\'ees, UMR 7641, Ecole Polytechnique,
France
}

\address[b]{
National University Higher School of Economics, Moscow, Russia
}

\end{aug}

\begin{abstract}
In this paper we propose an efficient variance reduction approach for  MCMC algorithms relying on a novel discrete time martingale representation for Markov chains. Our approach is fully non-asymptotic and does not require any type of ergodicity or special product structure of the underlying density.  By rigorously analyzing the  convergence properties of the proposed algorithm, we show that it's complexity is indeed asymptotically smaller than one of the original MCMC algorithm. The numerical performance of the new method is illustrated  in the case of Gaussian mixtures and Bayesian regression models.
\end{abstract}

\begin{keyword}[class=MSC]
\kwd[Primary ]{60G40}
\kwd{60G40}
\kwd[; secondary ]{91G80}
\end{keyword}

\begin{keyword}
\kwd{MCMC}
\kwd{variance reduction}
\kwd{martingale representation}
\end{keyword}

\end{frontmatter}

\section{Introduction}

Monte Carlo integration typically has an error variance of the form
$\sigma^{2}/n,$ where $n$ is a sample size and \(\sigma^2\) is the variance of  integrand. We can make the variance smaller
by using a larger value of $n$.  Alternatively,
 we can  reduce $\sigma^2$ instead of increasing the sample size \(n\). To this end, one can try to construct
a new Monte Carlo experiment with the same expectation as the original one
but with a lower variance $\sigma^2$. Methods to achieve this are known as variance
reduction techniques. Variance reduction plays an important role in
Monte Carlo and Markov Chain Monte Carlo methods. Introduction to many of the variance reduction techniques can be found in \cite{christian1999monte}, \cite{rubinstein2016simulation}, \cite{GobetBook} and \cite{glasserman2013monte}. Recently one witnessed a revival of interest in efficient variance reduction methods  for MCMC algorithms, see for example \cite{dellaportas2012control}, \cite{mira2013zero}, \cite{brosse2018diffusion} and references therein.
\par
Suppose that we wish to compute $\pi(f):=\mathsf{E}\left[f(X)\right]$, where $X$
is a random vector-valued in $\mathcal{X}\subseteq\mathbb{R}^{d}$ with a density $\pi$ and $f:$
$\mathcal{X}\to\mathbb{R}$ with $f\in L^2(\pi)$.
The idea of the so-called control variates variance reduction method
 is to find a cheaply computable random variable $\zeta$ with $\mathsf{E}[\zeta]=0$ and \(\mathsf{E}[\zeta^2]<\infty,\)
such that the variance of the r.v. $f(X)-\zeta$ is small.  The complexity of the problem of constructing classes $Z$ of control variates \(\zeta\) satisfying   $\mathsf{E}[\zeta]=0$ essentially depends on the degree of our knowledge on \(\pi.\)
For example, if \(\pi\) is analytically known and satisfies some regularity conditions, one can apply the well-known technique of  polynomial interpolation to construct control variates enjoying  some optimality properties, see, for example, Section~3.2 in \cite{dimov2008monte}. Alternatively, if an orthonormal system in \(L^2(\pi)\) is analytically available, one can build control variates \(\zeta\) as a linear combination of the corresponding basis functions, see \cite{GobetCV}. Furthermore, if \(\pi\) is known only up to a normalizing constant (which is often the case in Bayesian statistics), one can apply the recent approach of control variates  depending only on the gradient \(\nabla \log \pi\)  using Schr\(\ddot{\text{o}}\)dinger-type Hamiltonian operator in  \cite{mira2013zero} and so-called Stein operator in \cite{brosse2018diffusion}.
In some situations \(\pi\) is not known analytically, but \(X\) can be represented as a function of  simple random variables with known distribution.
Such  situation arises, for example, in the case of functionals of  discretized diffusion processes. In this case a Wiener chaos-type decomposition can be used to construct control variates with nice theoretical properties, see \cite{belomestny2018stratified}.
Note that in order to compare different  variance reduction approaches, one has to analyze their complexity, that is, the number of numerical operations required to achieve a prescribed magnitude of the resulting variance.

The situation becomes much more difficult in the case of MCMC algorithms, where one  has to work with a
Markov chain \(X_p,\) \(p=0,1,2,\ldots,\) whose marginal distribution  converges  to \(\pi\) as time grows. One important class of the variance reduction methods in this case  is based on the so-called Poisson equation for the corresponding  Markov chain. It was observed in Henderson~\cite{henderson1997variance}  that if a time-homogeneous Markov chain \((X_p)\) is stationary with stationary distribution \(\pi,\) then for any real-valued function \(G \in L^1(\pi) \) defined on the state space of the Markov chain \((X_p),\)  the function \(U(x) := G(x)-\mathsf{E}[G(X_{1})|X_0 = x]\) has zero mean with respect to \(\pi\).  The best choice for the function \(G\) corresponds to a solution of the so-called Poisson equation  \(\mathsf{E}[G(X_{1})|X_0 = x]-G(x)=-f(x)+\pi(f)\).  Moreover, it is also related to the minimal asymptotic variance in the corresponding central limit theorem, see \cite{duncan2016variance} and \cite{mira2013zero}.   Although the Poisson equation involves the quantity of interest \(\pi(f)\)  and can not be  solved explicitly in most cases, this idea still can be used to construct some  approximations for the optimal zero-variance control variates. For example,  Henderson~\cite{henderson1997variance} proposed to compute approximations for the solution of the Poisson equation for specific Markov chains with particular emphasis on models arising in stochastic network theory. In \cite{dellaportas2012control} and \cite{brosse2018diffusion}  series-type control variates are introduced and studied for reversible Markov chains. It is assumed in \cite{dellaportas2012control}  that the one-step conditional expectations  can be computed explicitly  for a set of basis functions. The authors in \cite{brosse2018diffusion} proposed another approach tailored to diffusion setting which doesn't require the computation of integrals of basis functions and only involves  applications of the underlying generator.
\par
In this paper we focus on the  Langevin type algorithms which got much attention recently, see \cite{dalalyan2017theoretical,durmus:moulines:2017, MR2353037, MR3861816, MR2977986} and references therein. We propose  a generic variance reduction method for these and other types algorithms, which is purely non-asymptotic and does not require that    the conditional expectations of the corresponding Markov chain can be computed  or that the generator is known analytically. Moreover, we do not need to assume stationarity or/and sampling under the invariant distribution \(\pi.\) We rigorously analyse the convergence of the method and study its complexity. It is shown that  our variance-reduced Langevin algorithm outperforms the standard Langevin algorithms in terms of complexity.
\par
The paper is organized as follows.  In Section~\ref{sec:setup} we set up the problem and introduce some notations. Section~\ref{seq:mart_repr} contains a novel martingale representation and shows how this representation can be used for variance reduction. In Section~\ref{sec:ula_analysis} we analyze the performance of the proposed variance reduction algorithm in the case of Unadjusted Langevin Algorithm (ULA). Section~\ref{sec:coeff} studies the complexity of the variance reduced ULA. Finally, numerical examples are presented in Section~\ref{sec:num}.

\section{Setup}\label{sec:setup}
Let  \(\mathcal{X}\) be a domain in \( \mathbb{R}^d.\)  Our aim is to numerically compute  expectations of the form
\[
\pi(f)=\int_{\mathcal{X}} f(x)\pi(\rmd x),
\]
where \(f:\) \(\mathcal{X}\longrightarrow \mathbb{R}\) and \(\pi\) is a probability measure supported on \(\mathcal{X}.\)
If  the dimension of the space \(\mathcal{X}\) is large and \(\pi(f)\) can not be computed analytically, one can apply Monte Carlo methods. However, in many practical situations  direct sampling from \(\pi\) is impossible and this precludes the use of plain Monte Carlo methods in this case. One popular alternative to Monte Carlo  is Markov Chain Monte Carlo (MCMC), where one is looking for a discrete time  (possibly non-homogeneous) Markov chain   \((X_p)_{p\geq 0}\) such that \(\pi\) is its unique invariant measure. In this paper we study a class of MCMC algorithms with \((X_p)_{p\geq 0}\) satisfying the  the following recurrence relation:
\begin{equation}
\label{eq:chain_gen}
X_{p}=\Phi_{p}(X_{p-1},\xi_{p}),\quad p=1,2,\ldots ,\quad X_{0}=x_0,
\end{equation}
for some i.i.d.  random vectors \(\xi_p\in \mathbb{R}^m\) with distribution \(P_{\xi}\)
and some Borel-measurable
functions $\Phi_{p}\colon\mathcal{X}\times\mathbb{R}^{m}\to\mathcal{X}.$
In fact, this is quite general class of Markov chains (see Theorem~1.3.6 in ~\cite{moulines2018})
and many well-known MCMC algorithms can be represented in form \eqref{eq:chain_gen}.
Let us consider two popular examples.
\begin{example}[Unadjusted Langevin Algorithm]
\label{exam:langevin-algorithm}
Fix a sequence of positive time steps \((\gamma_p)_{p\geq 1}.\) Given a Borel function $\mu\colon\mathbb{R}^{d}\to\mathbb{R}^{d}$,
consider a non-homogeneous
discrete-time Markov chain $(X_{p})_{p\geq0}$ defined by
\begin{equation}\label{eq:chain}
X_{p+1}=X_{p}-\gamma_{p+1}\mu(X_{p})+\sqrt{\gamma_{p+1}}Z_{p+1},\end{equation}
where $\left(Z_{p}\right)_{p\geq1}$ is an i.i.d.\ sequence of $d$-dimensional
standard Gaussian random vectors. If
${\mu=\frac{\nabla U}{2}}$
for some continuously differentiable function $U,$ then Markov chain~\eqref{eq:chain} can be used to approximately sample from the density
\begin{equation}\label{eq:stationary_distr}
\pi(x)= \mathrm{const}\,\rme^{-U(x)}, \quad \mathrm{const}= \left. 1 \middle /{\int_{\mathbb{R}^{d}} \rme^{-U(x)}\, \rmd x,} \right.
\end{equation}
provided that \(\int_{\mathbb{R}^{d}} \rme^{-U(x)}\,\rmd x\) is finite. This method is usually referred to as
Unadjusted Langevin Algorithm (ULA).
In fact the Markov chain~\eqref{eq:chain}
arises as the Euler-Maruyama discretization
of the Langevin diffusion
\[
\rmd Y_t=-\mu(Y_t)\,\rmd t+ \rmd W_t
\]
with nonnegative time steps $(\gamma_p)_{p\ge1}$,
and,  under mild technical conditions, the latter Langevin diffusion admits $\pi$
of~\eqref{eq:stationary_distr}
as its unique invariant distribution; see \cite{dalalyan2017theoretical} and \cite{durmus:moulines:2017}.
\end{example}
\begin{example}[Metropolis-Adjusted Langevin Algorithm]
The Metropolis-Hastings algorithm
associated with a target density \(\pi\) requires the choice of a sequence of conditional densities  \((q_p)_{p\geq 1}\) also called proposal or candidate kernels. The transition from the value  of the Markov chain \(X_p\)  at time \(p\)
and its value at time \(p + 1\) proceeds via the following transition step:

\begin{algorithm}[H]
Given \(X_p=x\)\;
\begin{enumerate}
\item Generate \(Y_p\sim q_p(\cdot|x)\)\;
\item Put
\begin{eqnarray*}
X_{p+1}=
\begin{cases}
Y_p, & \texttt{ with probability } \alpha(x,Y_p),
\\
x, &  \texttt{ with probability } 1-\alpha(x,Y_p),
\end{cases}
\end{eqnarray*}
where
\begin{eqnarray*}
\alpha(x,y)=\min\left\{1,\frac{\pi(y)}{\pi(x)}\frac{q_p(x|y)}{q_p(y|x)}\right\}.
\end{eqnarray*}
\end{enumerate}
\end{algorithm}
Then, as shown in Metropolis et al.~\cite{metropolis1953equation}, this transition is reversible with respect to \(\pi\) and therefore preserves the stationary density \(\pi\). If \(q_p\) have a wide enough
support to eventually reach any region
of the state space \(\mathcal{X}\) with positive mass
under \(\pi\), then this transition is irreducible and $\pi$ is a maximal irreducibility measure \cite{mengersen:tweedie:1996}. The  Metropolis-Adjusted Langevin algorithm (MALA) takes  \eqref{eq:chain} as proposal, that is,
\begin{eqnarray*}
q_p(y|x)=(\gamma_{p+1})^{-d/2}\boldsymbol{\varphi}\Bigl([y-x+\gamma_{p+1}\mu(x)]/\sqrt{\gamma_{p+1}}\Bigr),
\end{eqnarray*}
where
$\boldsymbol{\varphi}(z)=(2\pi)^{-d/2} \exp\{-|z|^2/2\}$,
$z\in\mathbb R^d$, denotes the density of a $d$-dimensional
standard Gaussian random vector.  The MALA algorithms usually provide noticeable speed-ups in convergence for most problems. It is not difficult to see that the MALA chain can be compactly represented in the form
\begin{align*}
X_{p+1} &=X_p+\mathbbm{1}\bigl(U_{p+1}\leq \alpha(X_{p},Y_{p})\bigr)(Y_{p}-X_p),  \,
Y_{p}&=X_p-\gamma_{p+1}\mu(X_p)+\sqrt{\gamma_{p+1}}Z_{p+1},
\end{align*}
where \((U_{p})_{p\geq 1}\) is an i.i.d. sequence of uniformly distributed on \([0,1]\) random variables independent of \((Z_p)_{p\geq 1}.\) Thus we recover \eqref{eq:chain_gen} with  \(\xi_p=(U_p,Z_p)\in \mathbb{R}^{d+1}\) and
\begin{eqnarray*}
\Phi_p(x,(u,z)^\top)=x+\mathbbm{1}\bigl(u\leq \alpha(x,x-\gamma_{p}\mu(x)+\sqrt{\gamma_{p}}z)\bigr)(-\gamma_{p}\mu(x)+\sqrt{\gamma_{p}}z).
\end{eqnarray*}
\end{example}
\begin{example}
  Let \((X_t)_{t\in [0,T]}\) be the unique strong solution to a  SDE of the form:
\begin{eqnarray}
\label{eq:sde-inv}
dX_t=b(X_t)\,dt+\sigma(X_t)\,dW_t,\quad t\geq 0,	
\end{eqnarray}
where \(W\) is a standard \(\mathbb{R}^m\)-valued Brownian motion, \(b:\) \(\mathbb{R}^d\to \mathbb{R}^d\) and \(\sigma:\) \(\mathbb{R}^d\times \mathbb{R}^m \to \mathbb{R}^d\) are locally Lipschitz continuous functions with at most linear growth.
The process \((X_t)_{t\geq 0}\) is a Markov
process and let \(L\) denote its infinitesimal generator defined by
\begin{eqnarray*}
Lg=b^\top \nabla g+\frac{1}{2}\text{Tr}(\sigma^\top D^2g\sigma)
\end{eqnarray*}
for any \(g\in C^2(\mathbb{R}^d).\)
If there exists a continuously twice differentiable Lyapunov function \(V:\) \(\mathbb{R}^d\to \mathbb{R}_{+}\) such that
\begin{eqnarray*}
\sup_{x\in \mathbb{R}^d} LV(x) <\infty,\quad \limsup_{|x|\to \infty} LV(x)<0,
\end{eqnarray*}
then there is an invariant probability measure \(\pi\)  for \(X,\) that is, \(X_t\sim \pi\) for all \(t> 0\) if \(X_0\sim \pi.\) Invariant measures are crucial in the study of the long term behaviour of stochastic differential
systems \eqref{eq:sde-inv}.    Under some additional assumptions,  the invariant
measure \(\pi\) is  ergodic and this property
can be exploited  to compute  the integrals \(\pi(f)\) for \(f\in L^2(\pi)\) by means of ergodic averages. The idea is to replace the diffusion \(X\) by a (simulable) discretization scheme of the form (see e.g. \cite{MR3861816})
\begin{eqnarray*}
\bar X_{n+1}=\bar X_n+\gamma_{n+1} b(\bar X_n)+\sigma(\bar X_n)(W_{\Gamma_{n+1}}-W_{\Gamma_n}), \quad n\geq 0,\quad \bar X_0=X_0,
\end{eqnarray*}
where \(\Gamma_n=\gamma_1+\ldots+\gamma_n\) and \((\gamma_n)_{n\geq 1}\) is a non-increasing sequence of time steps. Then for a function \(f\in L^2(\pi)\) we can approximate \(\pi(f)\) via
\begin{eqnarray*}
\pi_n^\gamma(f)=\frac{1}{\Gamma_n}\sum_{i=1}^n \gamma_{i}f(\bar X_{i-1}).
\end{eqnarray*}
Due to typically high correlation between \(X_0,X_1,\ldots\)  variance reduction is of crucial importance here.
As a matter of fact, in many cases there is no explicit formula for the invariant measure and this makes the use of gradient  based  variance reduction techniques (see e.g. \cite{mira2013zero})   impossible in this case.
On the contrary, our method can be directly used to reduce the variance of the ergodic estimator \(\pi_n^\gamma\) without explicit knowledge of \(\pi.\)
\end{example}
\section{Martingale representation and variance reduction}
\label{seq:mart_repr}
In this section we give a general discrete-time martingale representation for the Markov chain \eqref{eq:chain_gen} which is  used below to construct an efficient variance reduction algorithm. Let \((\phi_k)_{k\in \mathbb{Z}_+}\) be a complete orthonormal system in \(L^2(\mathbb{R}^m, P_{\xi})\) with \(\phi_0\equiv 1\).  In particular, we have
\begin{eqnarray*}
\mathsf{E}[\phi_i(\xi)\phi_j(\xi)]=\delta_{ij},\quad i,j\in  \mathbb{Z}_{+}
\end{eqnarray*}
with \(\xi \sim P_{\xi}.\)
Notice that this implies that the random variables
$\phi_k(\xi)$, $k\ge1$, are centered. As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of Shifted Legendre polynomials for "uniform part" and Hermite polynomials for "gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the Shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\) In the sequel, we denote by $(\mathcal{G}_p)_{p\in \mathbb{Z}_+}$  the filtration
generated by generated by $(\xi_p)_{p \in \mathbb{N}^*}$ with the convention $\mathcal{G}_0=\mathrm{triv}$.

\begin{thm}\label{prop:29032018a1}
 Let $f$ be a Borel function $\mathbb{R}^{d}\to\mathbb R$ such that
$\mathsf{E}\left[\left|f(X_{p})\right|^{2}\right]<\infty$, with a Markov chain $(X_{p})_{p\geq 0}$ defined in \eqref{eq:chain_gen}.
Then,
for $p>j$, the following representation holds in \(L^2(P)\)
\begin{eqnarray}
\label{eq:mart_repr}
f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}a_{p,l,k}(X_{l-1})\phi_k\left(\xi_{l}\right),
\end{eqnarray}
where
\begin{eqnarray}
\label{eq:coeff_mart}
a_{p,l,k}(x)=\mathsf{E}\left[\left.f(X_{p})\phi_k\left(\xi_{l}\right)\right|X_{l-1}=x\right], \quad p\geq l, \quad k\in \mathbb{N}.
\end{eqnarray}
\end{thm}
\begin{proof}
The expansion obviously holds for $p=1$ and $j=0$.
Indeed, due to the orthonormality and completeness
of the system $\left(\phi_{k}\right)$, we have
\[
f(X_{1})=\mathsf{E}\left[f(X_{1})\right]+\sum_{k\geq1}a_{1,1,k}(X_0)\phi_{k}(\xi_{1})
\]
with
\[
a_{1,1,k}(x_0)=\CPE{f(X_{1})\phi_{k}\left(\xi_{1}\right)}{X_0=x_0},
\]
provided $\mathsf{E}\left[\left|f(X_{1})\right|^{2}\right]<\infty.$
Recall that $\mathcal{G}_l=\sigma(\xi_{1},\ldots,\xi_{l}),$
$l=1,2,\ldots, $
and $\mathcal{G}_0=\mathrm{triv}$.
Suppose that (\ref{eq:mart_repr})
holds for $p=q$, all $j<q$, and all Borel-measurable functions
$f$ with $\mathsf{E}\left[|f(X_{q})|^2\right]<\infty$.
Let us prove it for $p=q+1$.
Given $f$ with $\mathsf{E}\left[|f(X_{p})|^2\right]<\infty$,
due to the orthonormality and completeness
of the system $\left(\phi_{k}\right)$, we get by
conditioning on $\mathcal{G}_{q}$,
\[
f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|{\mathcal{G}}_q\right]+\sum_{k\geq1}\alpha_{p,q+1,k}\phi_{k}(\xi_{q+1}),
\]
where
\begin{equation*}
\alpha_{p,q+1,k}
=\mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{q+1})\right|\mathcal G_q\right].
\end{equation*}
By the Markov property of $(X_{l})$,
we have
$\mathsf{E}[f(X_{p})|\mathcal{G}_q]
=\mathsf{E}[f(X_{p})|X_{q}]$.
Furthermore, a calculation involving
intermediate conditioning on $\mathcal G_{q+1}$
and the recurrence relation
$X_{q+1}=\Phi_{q+1}(X_{q},\xi_{q+1})$
verifies that
\[
\alpha_{p,q+1,k}=\mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{q+1})\right|X_{q}\right]
=a_{p,q+1,k}(X_{q})
\]
for suitably chosen
Borel-measurable functions $a_{p,q+1,k}$.
We thus arrive at
\begin{align}
\label{berm:sig_X}
f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|X_{q}\right]+\sum_{k\geq1}a_{p,q+1,k}(X_{q})\phi_{k}(\xi_{q+1}),
\end{align}
which is the required statement in the case $j=q$.
Now assume $j<q$.
The random variable
$\mathsf{E}\left[\left.f(X_{p})\right|X_{q}\right]$
is square integrable and has the form
$g(X_{q})$,
hence the induction hypothesis applies, and we get
\begin{equation}\label{eq:28082017a1}
\mathsf{E}\left[\left.f(X_{p})\right|X_{q}\right]=\mathsf{E}\left[\left.f(X_{p})\right|X_{j}\right]+\sum_{k\geq1}\sum_{l=j+1}^{q}a_{p,l,k}(X_{l-1})\phi_{k}(\xi_{l})
\end{equation}
with
\begin{align*}
a_{p,l,k}(X_{l-1}) &= \mathsf{E}\left[\left.\mathsf{E}\left[\left.f(X_{p})\right|\mathcal{G}_{q}\right]\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]
= \mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]\\
&=\mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{l})\right|X_{l-1}\right].
\end{align*}
Formulas \eqref{berm:sig_X}
and~\eqref{eq:28082017a1} conclude the proof.
\end{proof}
From  numerical point of view another representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
\begin{prop}
The coefficients \(a_{p,l,k}\) in \eqref{eq:coeff_mart}  can be alternatively represented as
\begin{eqnarray*}
a_{p,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{p,l}\left(\Phi_l(x,\xi)\right)\right]
\end{eqnarray*}
with \(Q_{p,l}(x)=\mathsf{E}\left[\left.f(X_{p})\right|X_{l}=x\right],\) \(p\geq l.\)
%where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
The functions \((Q_{p,l})_{l=0}^p\)  can be computed by the backward recurrence: $Q_{p,p}(x)=f(x)$ and for $l \in \{0,\cdots,p-1\}$
\begin{eqnarray}
\label{eq:qpl}
Q_{p,l}(x)=\mathsf{E}\left[\left.Q_{p,l+1}(X_{l+1})\right|X_{l}=x\right].
%=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))]
\end{eqnarray}
\end{prop}
Next we show how  the representation \eqref{eq:mart_repr} can be used to efficiently reduce the variance of MCMC algorithms.  Let $(\gamma_{p})_{p\in\mathbb N}$ be a sequence of positive and non-increasing
step sizes
with $\sum_{p=1}^\infty \gamma_p=\infty$
and, for $n,l\in\mathbb{N}$, $n\le l$, we set
\[
\Gamma_{n,l}=\sum_{p=n}^{l}\gamma_{p}.
\]
 Consider a weighted average estimator $\pi_{n}^{N}(f)$ of the form
\begin{equation}\label{eq:29032018a2}
\pi_{n}^{N}(f)=\sum_{p=N+1}^{N+n}\omega_{p,n}^{N}f(X_{p}),\quad\omega_{p,n}^{N}=\gamma_{p+1}\Gamma_{N+2,N+n+1}^{-1},
\end{equation}
where $N\in\mathbb N_0$ is the length of the burn-in period and $n\in\mathbb N$
the number of effective samples.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{equation}
M_{K,n}^N(f) =\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\bar a_{l,k}(X_{l-1})\phi_k(\xi_{l}),
\label{eq:29032018a5}
\end{equation}
where
\begin{align}
\label{eq:first-expression-bar-a-l-k}
\bar{a}_{l,k}(x)
=\sum_{p=l}^{N+n}\omega_{p,n}^{N}a_{p,l,k}(x)=\mathsf{E}\left[\left.\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right)\phi_k(\xi_{l})\right|X_{l-1}=x\right].
\end{align}
Since \(X_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\geq 1,\)  the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.

\section{Analysis of variance reduced ULA}
\label{sec:ula_analysis}
The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
\begin{equation}
\label{eq:29032018a3}
\pi_{K,n}^N(f)=\pi_n^N(f)-M_{K,n}^N(f)
\end{equation}
should be small for \(K\) large enough. In this section   we provide  an analysis of the variance-reduced ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
We shall use the notations $\mathbb N=\{1,2,\ldots\}$ and $\mathbb N_0=\mathbb N\cup\{0\}$.
By $H_k$, $k\in\mathbb N_0$,
we denote the normalized Hermite polynomial on $\mathbb R$, that is,
$$
H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
\quad x\in\mathbb R.
$$
For a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
$\mathbf{H}_\mathbf{k}$ denotes the normalized Hermite polynomial on $\mathbb R^d$, that is,
$$
\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
$$
In what follows, we also use the notation
$|\mathbf{k}|=\sum_{i=1}^d k_i$ for $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
and we set $\mathcal G_p=\sigma(Z_1,\ldots,Z_p)$, $p\in\mathbb N$, and $\mathcal G_0=\mathrm{triv}$.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{align}
M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{p=N+1}^{N+n}\omega_{p,n}^{N}\sum_{l=N+1}^{p}a_{p,l,\mathbf{k}}(X_{l-1})\mathbf{H}_\mathbf{k}(Z_{l})
\label{eq:29032018a5}\\
&=\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{l=N+1}^{N+n}\bar{a}_{l,\mathbf{k}}(X_{l-1})\mathbf{H}_\mathbf{k}(Z_{l})
\notag
\end{align}
with \(\|\mathbf{k}\|=\max_{i} k_i\) and
\begin{align*}
\bar{a}_{l,\mathbf{k}}(x)
& =\sum_{p=l}^{N+n}\omega_{p,n}^{N}a_{p,l,\mathbf{k}}(x) =\mathsf{E}\left[\left.\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right)\mathbf{H}_\mathbf{k}\left(Z_{l}\right)\right|X_{l-1}=x\right].
\end{align*}
For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we are interested in its conditional
Mean Squared Error (MSE),
which can be decomposed as the sum of the squared conditional bias
and the conditional variance:
\begin{eqnarray}\label{eq:29032018a4}
\mathrm{MSE}\left[\rho(f)|\mathcal G_N\right]
&=&\mathsf{E}\left[(\rho(f)-\pi(f))^2|\mathcal G_N\right]
\\
\nonumber
&=&\left(\mathsf{E}[\rho(f)|\mathcal G_N]-\pi(f)\right)^2
+\mathsf{Var}\left[\rho(f)|\mathcal G_N\right].
\end{eqnarray}
The quantities in~\eqref{eq:29032018a4}
are conditioned on $\mathcal G_N$,
as it reflects the way the estimators are used for MCMC:
the path of the Markov chain is simulated only once,
and we start to use the realized values
of the Markov chain to construct
our estimate only after the burn-in period of length~$N$.
We also notice that, due to the Markovian structure,
the conditioning on $\mathcal G_N$
in~\eqref{eq:29032018a4}
is equivalent to conditioning on $X_N$ only
(this is particularly clear in the case $\rho(f)=\pi_n^N(f)$
but requires some calculation
in the remaining case $\rho(f)=\pi_{K,n}^N(f)$).

\subsection{Squared conditional bias}
Due to the martingale transform structure
of $M_{K,n}^N(f)$,
we have
\[
\mathsf E\left[\left.M_{K,n}^N(f)\right|\mathcal G_N\right]=0,
\]
Hence both estimators
$\pi_n^N(f)$ and $\pi_{K,n}^N(f)$ have the same conditional bias.
Notice that this remains true also when we substitute
the coefficients $a_{p,l,\mathbf{k}}$ in~\eqref{eq:29032018a5}
with some independent approximations $\widehat a_{p,l,\mathbf{k}}.$ For a bounded Borel function~$f$,
we can estimate the conditional bias
similarly to the beginning of ~\cite[Section~4]{durmus:moulines:2017}:
\begin{align}
\left(\mathsf{E}[\pi_{K,n}^N(f)|\mathcal G_N]-\pi(f)\right)^2
&=\left(\mathsf{E}[\pi_{n}^N(f)|\mathcal G_N]-\pi(f)\right)^2
\label{eq:06042018a2}\\
&\le
\mathrm{osc}(f)^2
\sum_{p=N+1}^{N+n}
\omega_{p,n}^N\,
\|Q_\gamma^{N+1,p}(X_N,\cdot)-\pi(\cdot)\|_{TV}^2,
\notag
\end{align}
where
$\mathrm{osc}(f):=\sup_{x\in\mathbb R^d}f(x)-\inf_{x\in\mathbb R^d}f(x)$,
$\|\mu-\nu\|_{\TV}$ denotes the total variation distance
between probability measures $\mu$ and $\nu$, that is,
$$
\|\mu-\nu\|_{TV}=\sup_{A\in\mathcal B(\mathbb R^d)}
|\mu(A)-\nu(A)|,
$$
$\pi(\cdot)$ denotes the probability measure on $\mathbb R^d$
with density $\pi$ of~\eqref{eq:stationary_distr};
for $\gamma>0$, the Markov kernel $R_\gamma$
from $(\mathbb R^d,\mathcal B(\mathbb R^d))$
into $(\mathbb R^d,\mathcal B(\mathbb R^d))$
is defined by
$$
R_\gamma(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\}dy,
$$
while, for $k,l\in\mathbb N$, $k\le l$,
the kernel $Q_\gamma^{k,l}$ is
$
Q_\gamma^{k,l}=R_{\gamma_l}\cdots R_{\gamma_k},
$
which, finally, provides the (random) measure
$Q_\gamma^{N+1,p}(X_N,\cdot)$
used in the right-hand side of~\eqref{eq:06042018a2}.
Different specific upper bounds for the squared bias
can be deduced from~\eqref{eq:06042018a2}
using results of Section~3 in~\cite{durmus:moulines:2017}
on bounds in the total variation distance.

\subsection{Conditional variance}
An upper bound for the variance of the classical estimator~\eqref{eq:29032018a2}
is provided in  in~\cite[Theorem~17]{durmus:moulines:2017}.
As for the estimator~\eqref{eq:29032018a3}, we introduce lemma of conditional variance decomposition.
It follows from~\eqref{eq:29032018a5}, Theorem~\ref{prop:29032018a1} and independence condition for the sequence $(Z_p)_{p \geq 1}$.
\begin{lem}
\label{cond_var_dec}
	For variance-reduced estimator~\eqref{eq:29032018a3}, defined for the Markov Chain~\eqref{eq:chain_gen} the following representation holds
\begin{equation}\label{eq:29032018a6}
\mathsf{Var}\left[\left.\pi_{K,n}^{N}(f)\right|\mathcal G_N\right]
%=\sum_{\min_i (k_i)\ge K+1}
=\sum_{\mathbf k\colon \|\mathbf k\|\ge K+1}
\sum_{l=N+1}^{N+n}
\mathsf{E}\left[\left.\left|\bar a_{l,\mathbf{k}}(X_{l-1})\right|^{2}
\right|\mathcal G_N\right].
\end{equation}
\end{lem}
Now we present an upper bound for the right-hand sides of~\eqref{eq:29032018a6} in case of ULA algorithm.
\begin{thm}\label{th:mr}
Fix $K\in\mathbb N$.
Suppose that $f$  and \(\mu\) are \(K+1\) times continuously differentiable
and it holds
\begin{eqnarray*}
|\partial^{\mathbf{k}} f(x)|\le  B_f,\quad |\partial^{\mathbf{k}}\mu(x)|\le  B_\mu, \quad
x\in\mathbb R^d
\end{eqnarray*}
for all \(\mathbf{k}\) with \(0<\|\mathbf{k}\|\leq  K+1\),
\[
\jac{\mu}(x)\ge b_\mu I ,\quad x\in\mathbb R^d,
\]
for some positive number $b_\mu\in(0,B_\mu]$  where  \(\jac{\mu}\) stands for Jacobian of \(\mu.\)
Let $(\gamma_k)_{k\in\mathbb N}$ be a sequence of positive
and non-increasing step sizes with $\sum_{k=1}^\infty \gamma_k=\infty$.
We  assume that $\gamma_1<\frac1{B_\mu}$ and that
\begin{equation}\label{eq:17042018a1}
\sum_{r=j}^\infty \gamma_r\prod_{k=j}^{r}\left[1-\gamma_{k}b_\mu\right]\leq C,\quad\text{for all }j\in\mathbb N,
\end{equation}
with some constant \(C\). Then it holds
\begin{eqnarray}
\label{eq:var-bound}
\mathsf{Var}\left[\left.\pi_{K,n}^{N}(f)\right|\mathcal G_N\right]\lesssim \frac{1}{\Gamma^2_{N+2,N+n+1}}\sum_{l=N+1}^{N+n}\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
\left(\frac{\gamma_{l}}{2}\right)^{|I|K},
\end{eqnarray}
where the sum in \eqref{eq:var-bound} is taken over all subsets \(I\) of \(\{1,\ldots,d\}\) and \(\lesssim\) stands for inequality up to a constant not depending  on \(n\) and \(N.\)
\end{thm}
\begin{remark}
Note that in  Theorem~\eqref{th:mr} we don't require that  the function \(\mu\) itself is bounded.
Assumption~\eqref{eq:17042018a1} is not restrictive.
For instance, a straightforward calculation shows
that \eqref{eq:17042018a1} is satisfied in most interesting case
$\gamma_k=\const/k^\alpha$ with $\alpha\in (0,1)$.
\end{remark}

\section{Nonparametric regression}
The functions \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. This amounts to solve a nonparametric regression problem.
We present a generic regression algorithm. Algorithm starts with estimating the functions $\bar{Q}_{l}$ for $l=N+1,\ldots,N+n,$ where
\begin{equation*}
\bar{Q}_{l}(x)=\sum_{p=l}^{N+n} \omega_{p,n}^{N} Q_{p,l}\left(x\right)=\omega_{l,n}^{N}f(x)+\mathsf{E}\left[\left.\sum_{p=l+1}^{N+n}\omega_{p,n}^{N}f(X_{p})\right|X_{l}=x\right] .
\end{equation*}
We first generate \(\NtrainPath\)  paths conditionally independent of the $\sigma$-algebra generated by the burn-in sequence $X_1,\dots,X_N$:
\[
\TrainSet=\Bigl\{(X^{(s)}_{N+1},\ldots,X^{(s)}_{N+n}),\quad s=1,\ldots,\NtrainPath\Bigr\}
\]
of the chain \(X\)
(the so-called ``training paths'').
The regression algorithm proceeds with estimating the  functions \((\bar Q_{l})\) by solving  the least squares optimization problems
\begin{equation}\label{eq:06042018a1}
{\widehat  Q}_{l}=\argmin_{\psi\in \Psi}\sum_{s=1}^{\NtrainPath} \left | \sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X^{(s)}_{p})-\psi(X^{(s)}_{l})\right|^2
\end{equation}
for \(l=N+1,\ldots, N+n-1.\)
where \(\Psi\) is a  class of  functions on \(\mathbb{R}^d.\)
Next  we estimate the coefficients \( \bar a_{l,k}\) using
the formula
\begin{equation}
\label{eq:a_est_int}
\widehat a_{l,k}(x)=\mathsf{E}\left[\left.\phi_k\left(\xi\right)\widehat Q_{l}\left(\Phi_l(x,\xi)\right)\right | \TrainSet\right] \,,
\end{equation}
where $\xi$ is a random variable independent from $\mathcal{T}$ with law $P_\xi,$
$\Phi_l$ is defined in \eqref{eq:chain_gen} and the expectation is taken according to known distribution \(P_\xi\). Conditioning on $\TrainSet$ underlines that estimation of $\widehat Q_{l}$ derived from training trajectories. In many cases integration can be done in closed analytical form.

 For example, in the case of ULA algorithm we have \(P_\xi$ is the multivariate normal distribution $ \mathcal{N}(0,I_d)\) and \(\Phi_l(x,\xi)=x-\gamma_{l}\mu(x)+\sqrt{\gamma_{l}}\xi.\) Therefore if we use polynomials to approximate \((\bar Q_{l}),\) then \((\widehat a_{l,k})\) can be even computed in closed form; implementations details are provided in  Section~\ref{sec:num}.

Upon estimating  the coefficients \((\widehat a_{l,k})\), one can construct an empirical estimate  of \(M_{K,n}^N(f)\) in the form
\begin{eqnarray*}
\widehat M_{K,n}^N(f) =\sum_{1\leq  k \le K}\sum_{l=N+1}^{N+n}\widehat a_{l,k}(X_{l-1})\phi_k(\xi_{l}).
\end{eqnarray*}
Obviously \(\mathsf{E}[\widehat M_{K,n}^N(f)|\TrainSet]=0\) and \(\widehat M_{K,n}^N(f)\) is indeed a valid control variate in that it
does not introduce any bias.

\section{Complexity analysis for ULA}
\label{sec:coeff}

The following result quantifies the error of estimating the functions \((\bar Q_l)\)
via the algorithm and its proof follows from Theorem~2.2 in \cite{audibert2011robust}.
\begin{thm}\label{berm:theorem:regression_cv}
 Suppose that for any \(l\in \{N+1,\ldots,N+n+1\},\)
\[
\mathsf{E}\left[\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})-\bar Q_l(X_l)\right)^4\right]\leq \sigma_l^4,
\]
for some finite positive numbers \(\sigma_{N+1},\ldots, \sigma_{N+n}. \)
Furthermore, assume that \(\Psi=\mathrm{span}\{\psi_1,\ldots,\psi_D\},\) where the functions  \(\psi_1,\ldots,\psi_D\) are uniformly bounded and satisfy
\[
\max_{l}\sup_{g\in \Psi\setminus \{0\}} \|g\|^2_\infty/\mathsf{E}[g^2(X_l) | \mathcal{G}_N]\leq B<\infty.
\]
Then for any values of \(\varepsilon\) and \(\NtrainPath\) such that \(2/\NtrainPath\leq \varepsilon\leq 1\) and
\begin{eqnarray*}
\NtrainPath\gtrsim B^2\left[BD+\log(2/\varepsilon)+\frac{B^2D^2}{\NtrainPath}\right]
\end{eqnarray*}
it holds with probability at least \(1-\varepsilon,\)
\begin{multline}
\mathsf{E}\left[ \left.\Bigl|\bar Q_l(X_l)
-\widehat Q_l(X_l)\Bigr|^2\right | \TrainSet \vee \mathcal{G}_N\right]
\lesssim  \sigma^2_l B
\left(\frac{BD+\log(2/\varepsilon)}{\NtrainPath}+\frac{B^2D^2}{\NtrainPath^2}\right)
\\
 +
\inf_{\psi\in \Psi}
\mathsf{E}\left[\left.\left|\bar Q_l(X_{l})-\psi(X_{l})\right|^{2}\right | \mathcal{G}_N \right], \label{eq:regr_error}
\end{multline}
for \(l=N+1,\ldots,N+n,
\) with \(\lesssim\) standing for inequality up to a universal multiplicative constant.
\end{thm}
Now we are able to give  a bound for the difference between \( M_{K,n}^N\) and \(\widehat M_{K,n}^N.\)
\begin{prop}
\label{cor:dif_m}
Under  conditions of Theorem~\ref{berm:theorem:regression_cv}, we have with probability at least \(1-\varepsilon,\)
\begin{align}
\label{eq:main_bound}
&\mathsf{E}\left[\left.\left|\widehat{M}_{K,n}^{N}(f)-M_{K,n}^{N}(f)\right|^{2}\right | \TrainSet,\mathcal{G}_N\right] \\
\nonumber
&\phantom{\mathsf{E}\widehat{M}_{K,n}^{N}}  \lesssim K\left[\sum_{l=N+1}^{N+n} \sigma^2_l \right]
B
\left(\frac{BD+\log(2/\varepsilon)}{\NtrainPath}+\frac{B^2D^2}{\NtrainPath^2}\right)
\\ \nonumber
& \phantom{\mathsf{E}|\widehat{M}_{K,n}^{N}} + K \sum_{l=n+1}^{N+n}\inf_{\psi\in \Psi}
\mathsf{E}\left[\left.\left|\bar Q_l(X_{l})-\psi(X_{l})\right|^{2}\right | \mathcal{G}_N \right].
\end{align}
\end{prop}
\begin{proof}
Using the conditional Cauchy-Schwarz inequality and  orthonormality of \((\phi_k)_{k\geq 0}\), we derive
\begin{eqnarray*}
\mathsf{E}\left[\left.\left|\widehat{a}_{l,k}(X_{l-1})-\bar a_{l,k}(X_{l-1})\right|^{2}\right | \TrainSet \vee \mathcal{G}_N\right]\leq\mathsf{E}\left[\left.\left|\widehat{Q}_{l}\left(X_{l}\right)-\bar Q_{l}\left(X_{l}\right)\right|^{2}\right | \TrainSet \vee \mathcal{G}_N\right]
\end{eqnarray*}
By the Jensen inequality and orthonormality of \((\phi_k)_{k\geq 0},\)
\begin{multline*}
\mathsf{E}\left[\left.\left|\widehat{M}_{K,n}^{N}(f)-M_{K,n}^{N}(f)\right|^{2}\right | \TrainSet \vee \mathcal{G}_N\right]
\\
\leq \sum_{1\leq k\le K}\sum_{l=N+1}^{N+n}\mathsf{E}\left[\left.\left|\widehat{a}_{l,k}(X_{l-1})-\bar a_{l,k}(X_{l-1})\right|^{2}\right | \TrainSet \vee \mathcal{G}_N\right].
\end{multline*}
\end{proof}
\subsection{Complexity analysis of variance reduction for the ULA algorithm}
In the case of ULA one can bound the quantities \(\sigma_l^2\) using  \(L^p\)-type Sobolev inequalities  (see \cite{Adamczak}), see Remark~\ref{rem:p-norm} in Section~\ref{sec:proofs}.
In particular, we can derive that  under conditions of Theorem~\ref{th:mr} with \(K=1,\)
\begin{eqnarray*}
\sigma_l^2\lesssim \frac{1}{\Gamma^2_{N+2,N+n+1}},\quad l=N+1,\ldots,N+n+1,
\end{eqnarray*}
where \(\lesssim \) stands for the inequality up to a multiplicative constant not depending on \(n\) and \(N.\) Using this inequality and combining \eqref{eq:main_bound} with Theorem~\ref{th:mr}, we get for the variance of \(\widehat \pi_{K,n}^N(f)=\pi_n^N(f)-\widehat M_{K,n}^N(f),\) with probability at least \(1-\varepsilon,\)
\begin{eqnarray}
\nonumber
\mathsf{Var}\left[\left.\widehat \pi_{K,n}^N(f)\right | \TrainSet, \mathcal{G}_N\right]&\lesssim & \frac{n KB}{\Gamma^2_{N+2,N+n+1}}
\left(\frac{BD+\log(2/\varepsilon)}{\NtrainPath}+\frac{B^2D^2}{\NtrainPath^2}\right)
\\
\nonumber
&& + \frac{1}{\Gamma^2_{N+2,N+n+1}}\sum_{l=N+1}^{N+n}\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
\left(\frac{\gamma_{l}}{2}\right)^{|I|K}
\\
\label{eq:ula_red_var}
&& +
K \sum_{l=n+1}^{N+n}\inf_{\psi\in \Psi}
\mathsf{E}\left[\left.\left|\bar Q_l(X_{l})-\psi(X_{l})\right|^{2}\right | \TrainSet \vee \mathcal{G}_N \right].
\end{eqnarray}
In order to assess the complexity of the proposed algorithm,  we first prove that under some conditions the coefficients \(a_{p,l,\mathbf{k}}\) decay exponentially fast as \(|p-l|\to \infty.\)
\begin{lem}
\label{a_decay}
Suppose that \(f,\mu\in C^1(\mathbb{R}^d),\) \(0<b_\mu I\leq\jac{\mu}(x)\leq B_\mu I\) and \(B_\mu\gamma_{l}\leq 1,\) then
\begin{eqnarray*}
\|a_{p,l,\mathbf{k}}\|_{\infty}\leq \sqrt{\gamma_l}\, \|\nabla f\|_\infty\exp\left(-b_{\mu}\sum_{r=l+1}^p \gamma_{r}  \right),\quad \mathbf{k}\in \mathbb{N}^d\setminus \{\mathbf{0}\}.
\end{eqnarray*}
\end{lem}
\begin{cor}
Assume that
\begin{eqnarray}
\label{eq:gamk-decay}
\gamma_k=\gamma_1 k^{-\alpha},\quad \alpha\in (0,1),
\end{eqnarray}
 then
\begin{eqnarray*}
\|a_{p,l,\mathbf{k}}\|_{\infty}\leq \sqrt{\gamma_1} l^{-\alpha/2}\, \|\nabla f\|_\infty\exp\left(-c b_{\mu}\gamma_1 (p^{1-\alpha}-l^{1-\alpha}) \right)
\end{eqnarray*}
for some constant \(c>0.\)
\end{cor}
Suppose now that the chain is close to stationarity, then \(Q_{p,l}\) are functions of \(p-l\) only.
Furthermore,  according to Lemma~\ref{a_decay}, \(a_{p,l,\mathbf{k}}\) are exponentially small for \(p-l\) large.
As a result we only need to compute a logarithmic (in \(n\)) number of functions \(Q_{p,l}.\)
Hence the cost of computing the estimates  \(\widehat {a}_{p,l,\mathbf{k}}(x)\) for \(l=N+1,\ldots,N+n,\) and \(\|\mathbf{k}\|\leq K\) using regression on \(\Psi,\) is of order
\[
\log^{1/(1-\alpha)} (n) K^d  \NtrainPath D^2.
\]
Suppose for simplicity that all functions \(\bar{Q}_l\) are in \(\Psi\) for some \(D>0,\) that is, the third term in \eqref{eq:ula_red_var} is zero.  Then under \eqref{eq:gamk-decay} it is enough to take \(K=\lceil 1/\alpha \rceil+1\) to get
\begin{eqnarray*}
\sum_{l=N+1}^{N+n}\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
\left(\frac{\gamma_{l}}{2}\right)^{|I|K}\leq C
\end{eqnarray*}
for some constant \(C\) not depending on \(d.\) Then we have with high probability
\begin{eqnarray*}
\mathsf{Var}\left[\left.\widehat \pi_{K,n}^N(f)\right | \TrainSet \vee \mathcal{G}_N\right]\lesssim \frac{1}{n^{2(1-\alpha)}}\left[1+\frac{n}{\NtrainPath}\right],
\end{eqnarray*}
with corresponding cost proportional to \(\NtrainPath\log^{1/(1-\alpha)} (n) ,\) provided \(N/n=o(1).\) We should compare this to the standard weighted estimator \(\pi_{n}^N(f)\) with variance
\begin{eqnarray*}
\mathsf{Var}\left[\left. \pi_{n}^N(f)\right |  \mathcal{G}_N\right]\lesssim \frac{1}{n^{1-\alpha}}
\end{eqnarray*}
and cost of order \(n.\) Thus while the cost of achieving
\begin{eqnarray*}
\mathsf{Var}\left[\left. \pi_{n}^N(f)\right | \TrainSet \vee \mathcal{G}_N\right]\leq \varepsilon^2
\end{eqnarray*}
is of order  \(\varepsilon^{-2/(1-\alpha)},\) we get the same bound for \(\pi_{K,n}^N(f)\) at a cost of order
\[
\varepsilon^{-1/(1-\alpha)}\log^{1/(1-\alpha)} (\varepsilon).
\]
In the presence of approximation errors we have to increase \(D\) with \(n\) to balance  the statistical and approximation errors resulting in a smaller complexity reduction. Note that our approach does not require the computation of \(\nabla \log(\pi)\) for the construction of control variates. This leads to an additional complexity reduction especially in statistical application where the computation of \(\nabla \log(\pi)\) can be very costly.
\section{Numerical results}
\label{sec:num}
If each function  \(Q_{p,l}(x) \) can be well approximated by polynomials, then we can use polynomial basis functions in   \eqref{eq:06042018a1} to approximate \(\bar Q_l\). Suppose that we constructed a polynomial approximation  for each function \(\bar Q_{l}\)  in the form:
\begin{equation*}
\widehat{Q}_{l}(x) = \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} x^{\mathbf{s}},\quad s=(s_1,\ldots,s_d)
\end{equation*}
with some coefficients \(\beta_{\mathbf{s}}\in \mathbb{R}.\)
Then using the identity
\[
\xi^j = j! \sum_{r = 0}^{\lfloor j/2 \rfloor} \frac{1}{2^r  r! \sqrt{(j-2r)!}} H_{j-2r}(\xi),\quad \xi \in \mathbb{R},
\]
we derive for all $x \in \mathbb{R}^d$,
\begin{eqnarray*}
\widehat a_{l,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(x) \widehat{Q}_{l}(x - \gamma_l \mu(x) + \sqrt{\gamma_l}\xi)}{\TrainSet \vee \mathcal{G}_N}
= \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{i,k_i,s_i}(x),
\end{eqnarray*}
where for all integers $i,k,s_i$ and $x \in \rset^d$,
\begin{align*}
P_{l,k,s}(x) &= \mathsf{E} \left[ H_{k}(\xi_l) (x_l - \gamma_l \mu_l(x) + \sqrt{\gamma_l} \xi_l)^{s} \right]
\end{align*}
is a one-dimensional polynomial (in \(x\)) of degree at most \(s\) with analytically known coefficients. Alternatively, on practice we can work directly with functions \(Q_{p,l}.\) Due to (almost) time-homogenous Markov chain generated by ULA, each function $Q_{p,l}$ can be approximated by function, which depends only on the  $p-l$, that is,
\[
 Q_{p,l}(x) = \mathsf{E} \left[f(X_p) | X_l = x\right] \approx Q^\circ_{p-l}(x)
 \]
Thus we have
\[
  Q^\circ_{r}(x) = \mathsf{E} \left[f(X_{l+r}) | X_l = x \right].
 \]
Consequently, the functions $Q^\circ_r$ can be estimated using a modified least squares criteria
\begin{eqnarray}
\label{eq:Qcircr}
\widehat{Q}^\circ_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
\end{eqnarray}
for \(1 \leq r \leq n-1,\) where  $\widehat{Q}^\circ_0(x) = f(x)$ by definition.
Due to Lemma~\ref{a_decay}, it is enough to estimate $Q^\circ_{r}$ for $|r| < n_{\mathrm{trunc}}$ for some truncation level \(n_{\mathrm{trunc}}\) depending on \(d\) and \(\gamma.\) It allows us to use a smaller amount of training trajectories to approximate conditional expectations $Q^\circ_{r}$.
Finally  we construct a truncated version of our estimator:
\[
\pi_{K,n,n_{\mathrm{trunc}}}^N(f) = \pi_n^N(f) -  M_{K,n, n_{\mathrm{trunc}}}^{N}(f),
\]
where
\begin{eqnarray*}
{\widehat  M}_{K,n, n_{\mathrm{trunc}}}^N &=& \sum_{0<\left\| \mathbf{k}\right\| \leq K} \sum_{p=N+1}^{N+n} \omega_{p,n}^N
\\
&&\times
\sum_{l=N+1}^p {\widehat  a}_{p-l,\mathbf{k}}(X_{l-1})\mathbf{H}_{\mathbf{k}}(Z_l) \mathbbm{1}\{ |p-l| < n_{\mathrm{trunc}} \}
\end{eqnarray*}
and
\begin{eqnarray*}
\widehat a_{p-l,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(x) \widehat{Q}^\circ_{p-l}(x - \gamma_l \mu(x) + \sqrt{\gamma_l}\xi)}{\TrainSet \vee \mathcal{G}_N}.
\end{eqnarray*}
\subsection{Gaussian mixtures}
 We consider a sample generated by ULA with $\pi$  given by the mixture of two Gaussian distributions  with equal weights:
\[
\pi (x) = \frac{1}{2(2\pi)^{d/2}} \left( \rme^{\frac{-\| x-a\|_2^2}{2}}  + \rme^{\frac{-\| x+a\|_2^2}{2}} \right), \quad x \in\mathbb R^d
\]
where $a \in \mathbb{R}^d$ is a given vector. The function $U(x)$ and its gradient are given by
\[
U(x) = \frac{1}{2} \|x - a\|_2^2 - \text{log}(1 + \rme^{-2x^\top a}), \,\,
\nabla U(x) = x-a +2a(1 + \rme^{2 x^\top a})^{-1},
\]
respectively.
In our experiments we considered dimensions $d = 2$ and $d = 8$ and take $a=( (2d)^{-1/2}, \dots, (2d)^{-1/2})^\top$.
In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i,\)
 we have used constant step sizes $\gamma_i = \gamma=0.2$ and sampled $n = 10^4$ samples for $d = 2$ and $n = 2 \times 10^3$ for $d = 8$.  We generated $\NtrainPath = 10$ independent "training" trajectories and solved the least squares  problems \eqref{eq:Qcircr} using the first order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section. The truncation level \(n_{\mathrm{trunc}}\) is chosen to be \(50.\) To test our variance reduction algorithm, we generated 100 independent trajectories.
 In Figure ~\ref{fig:1} we compare our approach to variance reduction methods of \cite{mira2013zero} and \cite{brosse2018diffusion}.

 \begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][d = 2]{
\includegraphics[width=0.45\textwidth]{pictures/GM_2d_f1.png}
\label{fig:subfig1}}
\qquad
\subfloat[Subfigure 2 list of figures text][d = 8]{
\includegraphics[width=0.45\textwidth]{pictures/GM_8d_f1.png}
\label{fig:subfig2}}
\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Gaussian mixture model. The compared estimators are the ordinary empirical average  (O), our estimator of control variates (CV-B), zero variance estimator (ZV) and the control variates using diffusion approximation (CV), along with the region obtained by the ordinary ULA of the length $10^2 \times n$ (green regions). \label{fig:1}}
\end{figure}

\subsection{Binary Logistic Regression}

Second experiment considers the problem of logistic regression. Suppose we have i.i.d. sample $\left\{ (\mathbf{X}_i, Y_i)\right\}$ for $ i =1, \dots , m$ with features $\mathbf{X}_i \in \mathbb{R}^d $ and binary labels $Y_i \in \left\{0,1 \right\}$. The binary logistic regression model defines the conditional distribution of $Y$ given $X$ by a logistic function
\begin{eqnarray*}
r(\theta, x) = \frac{e^{\theta^T x}}{1 + e^{\theta^T x}},
\end{eqnarray*}
where $\theta$ is a parameter of model. In order to estimate $\theta$ according to given data, the Bayessian approach introduces prior distribution $\pi_0(\theta)$ and inferences the posterior density $\pi(\theta)$ using Bayes' rule.

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\textwidth]{pictures/BLR_banknotes_4d.png}
\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Logistic Regression. The compared estimators are the ordinary empirical average  (O), our estimator of control variates (CV-B), zero variance estimator (ZV) and control variates using diffusion approximation (CV).
 \label{fig:2}}
\end{center}
\end{figure}
In the case of  Gaussian prior $\pi_0$ with zero mean and covariance matrix $\sigma^2 I_d$, the posterior density takes the form:
\begin{eqnarray*}
\pi(\theta) \propto \exp \left\{ -  \mathbf{Y}^T \mathbf{X \theta} - \sum_{i=1}^m \text{log} (1 + e^{-\mathbf{\theta}^T \mathbf{X}_i}) - \frac{1}{2\sigma^2} \left\| \mathbf{\theta}\right\|_2^2 \right\},
\end{eqnarray*}
where $\mathbf{Y}$ is defined as $(Y_1, \dots, Y_m)^T \in \left\{ 0,1\right\}$ and $\sigma^2 $ is an additional parameter. We have
\begin{eqnarray*}
U( \mathbf{\theta}) = \mathbf{Y}^T \mathbf{X \theta} + \sum_{i=1}^m \text{log} (1 + e^{-\mathbf{\theta}^T \mathbf{X}_i}) + \frac{1}{2\sigma^2} \left\|\mathbf{\theta}\right\|_2^2, \,\,
\nabla U(\mathbf{\theta}) = \mathbf{X}^T \mathbf{Y} - \sum_{i=1}^m \frac{\mathbf{X}_i}{1 + e^{\mathbf{\theta}^T \mathbf{X}_i}} + \frac{1}{\sigma^2} \mathbf{\theta}.
\end{eqnarray*}
To demonstrate the performance of the proposed control variates approach in the above Bayesian logistic regression model, we take a simple dataset from \cite{mira2013zero}, which contains the measurements of four variables on $m=200$ Swiss banknotes. Prior distribution of the regression parameter $\mathbf{\theta} = \left( \theta_1, \theta_2, \theta_3, \theta_4 \right)$ assumed to be normal with the covariance matrix $\sigma^2 I_4$, where $\sigma^2 = 100$. To construct trajectories of length $n = 5 \times 10^3, $ we take  constant step sizes $\gamma_i = 0.1$ for the ULA scheme with $N = 10^3$ burn-in steps. As in the previous experiment we use the first order polynomials approximations to analytically compute the coefficients $\hat{a}_{p-l,\mathbf{k}},$ where $T = 10$ and $n_{trunc} = 50$. The target function is taken to be $f(\theta) = \sum_{i=1}^{i=d}\theta_i$.  In order to test our variance reduction algorithm, we generates \(100\) independent test trajectories.
In Figure ~\ref{fig:2} we compare our approach to variance reduction methods of \cite{mira2013zero} and \cite{brosse2018diffusion}.

\subsection{Binary Probit Regression}

Finally we consider a Bayesian probit regression model in its ordinary form, following \cite{mira2013zero}. More precisely, the log-likelihood of the model looks as follows
\begin{equation*}
    \mathbf{L}(\mathbf{Y}| \theta, \mathbf{X}) = \sum_{i=1}^{m} \left[ \mathbf{Y}_i \text{log} (\Phi(\theta^T \mathbf{X}_i)) + (1-\mathbf{Y}_i) \text{log}(\Phi(-\theta^T \mathbf{X}_i))\right],
\end{equation*}
where $\theta \in \mathbb{R}^d$ is a parameter of the model and $\Phi$ is a cumulative distribution function of the standard normal distribution $\mathcal{N}(0,1)$. Then, if the prior $\pi(\mathbf{\theta}) = \mathcal{N}(0, \sigma^2 I_d)$, then the posterior distribution expresses as
\begin{equation*}
    \pi(\theta | \mathbf{Y}, \mathbf{X})  \propto \text{exp} \left\{\sum_{i=1}^{m} \left[\mathbf{Y}_i \text{log} (\Phi(\theta^T \mathbf{X}_i)) + (1-\mathbf{Y}_i) \text{log}(\Phi(-\theta^T \mathbf{X}_i))\right]- \frac{1}{2\sigma^2} \left\| \mathbf{\theta}\right\|_2^2  \right\}
\end{equation*}
Consequently,
\begin{equation*}
    U(\theta) =  - \sum_{i=1}^{m} \left[\mathbf{Y}_i \text{log} (\Phi(\theta^T \mathbf{X}_i)) + (1-\mathbf{Y}_i) \text{log}(\Phi(-\theta^T \mathbf{X}_i))\right] + \frac{1}{2\sigma^2} \left\| \mathbf{\theta}\right\|_2^2
\end{equation*}

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\textwidth]{pictures/Probit_banknotes_4d.png}
\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Probit Regression. The compared estimators are the ordinary empirical average  (O), our modified estimator (CV-B), zero variance estimator (ZV) and the estimator with control variates using asymptotic variance minimization (CV).
 \label{fig:3}}
\end{center}
\end{figure}

\begin{equation*}
    \nabla U(\theta) = - \mathbf{X}^T s + \frac{1}{\sigma^2} \theta,
\end{equation*}
where the elements of the \(m\)-dimensional column vector $s$ are given by
\begin{equation*}
    s_i = \mathbf{Y}_i \Omega (\theta^T \mathbf{X}_i) - (1 - \mathbf{Y}_i) \Omega(- \theta^T \mathbf{X}_i), \quad i = 1, 2, \dots, m,
\end{equation*}
and $\Omega(\theta^T \mathbf{X}_i) : = \phi(\theta^T \mathbf{X}_i) / \Phi(\theta^T \mathbf{X}_i)$, where $\phi$ stands for the density of the standard normal distribution.
As in the previous examples, to show the performance  of the proposed variance reduced estimator for the Bayesian probit regression model, banknotes dataset is used and the regularization parameter is taken to be $\sigma^2=100$.
To generate trajectories of the length $n = 5 \times 10^3, $ we take a constant step size $\gamma_i = 0.1$ for the ULA scheme after $N = 10^3$ burn-in steps.  We use again the first order polynomials approximations to  compute the coefficients $\hat{a}_{p-l,\mathbf{k}}$ and fixed $T = 10$, $n_{trunc} = 50$ and $K=1$. The target function is now $f(\theta) = \sum_{i=1}^{i=d}\theta_i$.  In order to test our variance reduction algorithm, we generate \(100\) independent test trajectories. In Figure ~\ref{fig:3} we compare our approach to the variance reduction methods of \cite{mira2013zero} and \cite{brosse2018diffusion}.

\section{Proofs}
\label{sec:proofs}
\subsection{Proof of Theorem~\ref{th:mr}}
We start with introducing some notations.
For $m\in\mathbb N$, a smooth function
$h\colon\mathbb R^{d\times m}\to\mathbb R$
with arguments being denoted
$(y_1,\ldots,y_m)$, $y_i\in\mathbb R^d$, $i=1,\ldots,m$,
a multi-index $\mathbf k=(k_i)\in\mathbb N_0^d$,
and $j\in\{1,\ldots,m\}$,
we use the notation $\partial^{\mathbf k}_{y_j} h$ for the multiple derivative of $h$
with respect to the components of~$y_j$:
$$
\partial^{\mathbf k}_{y_j} h(y_1,\ldots,y_m)
:=\partial^{k_d}_{y_j^d}
\ldots
\partial^{k_1}_{y_j^1}
h(y_1,\ldots,z_m),
\quad y_j=(y_j^1,\ldots,y_j^d).
$$
In the particular case $m=1$ we can drop
the subscript $y_1$ in that notation.
For $l\le p$, we have the representation
\[
X_{p}=G_{p,l}(X_{l-1},\sqrt{\gamma_{l}}Z_{l},\ldots,\sqrt{\gamma_{p}}Z_{p}),
\]
where the function \(G_{p,l}:\) \(\mathbb{R}^{d\times(p-l+2)}\to \mathbb{R}^{d}\) is defined as
\begin{equation}
\label{eq:definition-G-p-l}
G_{p,l}(x,y_l,\ldots,y_p):=\Phi_{p}(\cdot,y_{p})\circ\Phi_{p-1}(\cdot,y_{p-1})\circ\ldots\circ\Phi_{l}(x,y_{l})
\end{equation}
with
\[
\Phi_{j}(x,y)=x-\gamma_{j}\mu(x)+y,
\quad x,y\in\mathbb R^d.
\]
As a consequence,
for a function $f\colon\mathbb R^d\to\mathbb R$
as in Section~\ref{sec:setup}, we have
$$
f\left(X_{p}\right) =f\circ G_{p,l}(X_{l-1},\sqrt{\gamma_{l}}Z_{l},\ldots,\sqrt{\gamma_{p}}Z_{p}).
$$
In what follows, for $\mathbf{k}\in\mathbb N_0^d$,
we use the shorthand notation
\begin{equation}
\label{eq:definition-differential-f-p}
\partial_{y_l}^{\mathbf k} f\left(X_{p}\right)
:=\partial_{y_l}^{\mathbf k} [f\circ G_{p,l}](X_{l-1},\sqrt{\gamma_{l}}Z_{l},\ldots,\sqrt{\gamma_{p}}Z_{p})
\end{equation}
whenever the function $f\circ G_{p,l}$ is smooth enough
(that is, $f$ and $\mu$ need to be smooth enough).
Finally, for a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$, we use the notation
\(\mathbf{k} ! :=k_1!\cdot\ldots\cdot k_d !\)

\begin{lem}\label{eq:a_repr}
Fix $l\le p$ and some \(\mathbf{k},\mathbf{k}'\in \mathbb{N}_0^d\) with \(\mathbf{k}'\le \mathbf{k}\) componentwise.  Then the following representation holds
$$
a_{p,l,\mathbf{k}}(X_{l-1})=\left(\gamma_{l}^{|\mathbf{k}'|}\frac{(\mathbf{k}-\mathbf{k}') !}{\mathbf{k}!}\right)^{1/2}
\mathsf E\left[\left.
\partial_{y_l}^{\mathbf{k}'}f(X_p)\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_l)\right|X_{l-1}\right].
$$
\end{lem}

\begin{proof}
Let $\boldsymbol{\varphi}(z)=\frac{1}{(2\pi)^{d/2}} \exp\{-|z|^2/2\}$,
$z\in\mathbb R^d$, denote the density of a $d$-dimensional
standard Gaussian random vector.
We first remark that, for the normalized Hermite polynomial $\mathbf H_{\mathbf k}$ on $\mathbb R^d$,
$\mathbf k\in\mathbb N_0^d$, it holds
$$
\mathbf{H}_{\mathbf{k}}(z)\boldsymbol{\varphi}(z)
=\frac{(-1)^{|\mathbf{k}|}}{\sqrt{\mathbf{k} !}} \partial^{\mathbf{k}} \boldsymbol{\varphi}(z).
$$
This enables to use the integration by parts in vector form as follows
(below $\prod_{j=l+1}^p:=1$ whenever $l=p$)
\begin{align*}
&a_{p,l,\mathbf k}(x)
 =
\int_{\mathbb R^d}\ldots\int_{\mathbb R^d}
f\circ G_{p,l}(x,\sqrt{\gamma_{l}}z_{l},\ldots,\sqrt{\gamma_{p}}z_{p})
\mathbf{H}_{\mathbf{k}}(z_{l})\boldsymbol{\varphi}(z_l)
\prod_{j=l+1}^p\boldsymbol{\varphi}(z_j)\, dz_{l}\ldots dz_{p}
\\
& =
\frac{1}{\sqrt{\mathbf k!}}
\int_{\mathbb R^d}\ldots\int_{\mathbb R^d}
f\circ G_{p,l}(x,\sqrt{\gamma_{l}}z_{l},\ldots,\sqrt{\gamma_{p}}z_{p})
(-1)^{|\mathbf{k}|}\partial^{\mathbf{k}} \boldsymbol{\varphi}(z_l)
\prod_{j=l+1}^p\boldsymbol{\varphi}(z_j)\, dz_{l}\ldots dz_{p}
\\
& =
\gamma_l^{|\mathbf k'|/2}\frac{1}{\sqrt{\mathbf k!}}
\int_{\mathbb R^d}\ldots\int_{\mathbb R^d}
\partial_{y_l}^{\mathbf k'}[f\circ G_{p,l}](x,\sqrt{\gamma_{l}}z_{l},\ldots,\sqrt{\gamma_{p}}z_{p})
\\
&\hspace{1em}\times
(-1)^{|\mathbf{k}-\mathbf k'|}\partial^{\mathbf{k}-\mathbf k'} \boldsymbol{\varphi}(z_l)
\prod_{j=l+1}^p\boldsymbol{\varphi}(z_j)\, dz_{l}\ldots dz_{p}
\\
%& =
%\gamma_l^{|\mathbf k'|/2}\frac{\sqrt{(\mathbf k-\mathbf k')!}}{\sqrt{\mathbf k!}}
%\int_{\mathbb R^d}\ldots\int_{\mathbb R^d}
%\partial_{y_l}^{\mathbf k'}[f\circ G_{p,l}](x,\sqrt{\gamma_{l}}z_{l},\ldots,\sqrt{\gamma_{p}}z_{p})
%\\
&\hspace{1em}\times
\mathbf{H}_{\mathbf{k}-\mathbf k'}(z_{l})\boldsymbol{\varphi}(z_l)
\prod_{j=l+1}^p\boldsymbol{\varphi}(z_j)\, dz_{l}\ldots dz_{p}
\\
 & =\gamma_{l}^{|\mathbf{k}'|/2}\frac{\sqrt{(\mathbf{k}-\mathbf{k}')!}}{\sqrt{\mathbf{k}!}}\mathsf{E}\left[\partial_{y_{l}}^{\mathbf{k}'}[f\circ G_{p,l}](x,\sqrt{\gamma_{l}}Z_{l},\ldots,\sqrt{\gamma_{p}}Z_{p})\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_{l})\right].
\end{align*}
The last expression yields the result.
\end{proof}

For multi-indices $\mathbf k,\mathbf k'\in\mathbb N_0^d$
with $\mathbf k'\le\mathbf k$ componentwise
and $\mathbf k'\ne\mathbf k$,
we get applying first Lemma~\ref{eq:a_repr}
\begin{equation*}
\begin{split}
\bar a_{l,\mathbf{k}}(X_{l-1})
&=\left(\gamma_{l}^{|\mathbf{k}'|}\frac{(\mathbf{k}-\mathbf{k}')!}{\mathbf{k}!}\right)^{1/2}
\,\sum_{p=l}^{N+n}\omega_{p,n}^N
\mathsf E\left[\left.
\partial_{y_l}^{\mathbf{k}'}f(X_p)\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_l)\right|X_{l-1}\right]\\
&=\left(\gamma_{l}^{|\mathbf{k}'|}\frac{(\mathbf{k}-\mathbf{k}')!}{\mathbf{k}!}\right)^{1/2}
\\
&\times \sum_{p=l}^{N+n}\omega_{p,n}^N
\mathsf E\left[\left.
\left(
\partial_{y_l}^{\mathbf{k}'}f(X_p)-\mathsf E\left[\partial_{y_l}^{\mathbf{k}'}f(X_p)|X_{l-1}\right]
\right)
\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_l)\right|X_{l-1}\right].
\end{split}
\end{equation*}

Assume that $\mu$ and $f$ are $K\times d$ times continuously differentiable.
Then, given $\mathbf k\in\mathbb N_0^d$,
by taking $\mathbf k'= \mathbf k'(\mathbf k)
=(K1_{\{k_1>K\}},\ldots,K1_{\{k_d>K\}})$,
for each $l\in\{N+1,\ldots,N+n\}$, we get
\begin{align}
\label{eq:sum_abar}
\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}\left[\bar{a}_{l,\mathbf{k}}^{2}(X_{l-1})|\mathcal{G}_{N}\right] & =\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}\left(\gamma_{l}^{|\mathbf{k}'|}\frac{(\mathbf{k}-\mathbf{k}')!}{\mathbf{k}!}\right)Q(\mathbf{k}',\mathbf{k}-\mathbf{k}')\\
\nonumber
 & =\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}\gamma_{l}^{|I|K}\sum_{\mathbf{m}_{I}\in\mathbb{N}_{I}^{d}}\frac{\mathbf{m}_{I}!}{\left(\mathbf{m}_{I}+\mathbf{K}_{I}\right)!}
 \\
 \nonumber
 & \hspace{1em}\times\sum_{\mathbf{m}_{I^c}\in \mathbb{N}^d_{0,I^c},\,\|\mathbf{m}_{I^c}\|\leq K}Q(\mathbf{\mathbf{K}}_{I},\mathbf{m}_{I}+\mathbf{m}_{I^c}),
\end{align}
where for any two multi-indices \(\mathbf{r},\) \(\mathbf{q}\) from \(\mathbb{N}_0^d\)
\begin{multline*}
Q(\mathbf{r},\mathbf{q})
\\
=\mathsf{E}\left\{
\left(\mathsf{E}\left[\sum_{p=l}^{N+n}\omega_{p,n}^{N}\left(\partial_{y_{l}}^{\mathbf r}f\left(X_{p}\right)-\mathsf{E}\left[\partial_{y_{l}}^{\mathbf r}f\left(X_{p}\right)|X_{l-1}\right]\right)\mathbf{H}_{\mathbf{q}}(Z_{l})\Big|X_{l-1}\right]\right)^{2}
\Bigg|\mathcal G_N \right\}.
\end{multline*}
In \eqref{eq:sum_abar} the first sum runs over all nonempty subsets $I$ of the set $\{1,\ldots,d\}.$
For any subset $I,$ $\mathbb{N}_{I}^{d}$ stands for a set
of multi-indices $\mathbf{m}_{I}$ with elements $m_{i}=0,$ $i\not\in I,$
and $m_{i}\in\mathbb{N},$  $i\in I.$ Moreover, \(I^c=\{1,\ldots,d\}\setminus I\) and \(\mathbb{N}^d_{0,I^c}\) stands for a set
of multi-indices $\mathbf{m}_{I^c}$ with elements $m_{i}=0,$ $i\in I,$
and $m_{i}\in\mathbb{N}_0,$  $i\not\in I.$ Finally, the multi-index \(\mathbf{K}_I\) is defined as $\mathbf{\mathbf{K}}_{I}=(K1_{\{1\in I\}},\ldots,K1_{\{d\in I\}}).$
Applying the estimate
\begin{eqnarray*}
\frac{\mathbf{m}_{I}!}{\left(\mathbf{m}_{I}+\mathbf{K}_{I}\right)!}\leq (1/2)^{|I| K},
\end{eqnarray*}
we get
\begin{eqnarray*}
\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}\left[\bar{a}_{l,\mathbf{k}}^{2}(X_{l-1})|\mathcal{G}_{N}\right]&\leq &
\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset} (\gamma_{l}/2)^{|I|K}
\\
&& \times\sum_{\mathbf{m}_{I}\in\mathbb{N}_{I}^{d}} \sum_{\mathbf{m}_{I^c}\in \mathbb{N}^d_{0,I^c},\,\|\mathbf{m}_{I^c}\|\leq K} Q(\mathbf{\mathbf{K}}_{I},\mathbf{m}_{I}+\mathbf{m}_{I^c})
\\
&\leq &
\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset} (\gamma_{l}/2)^{|I|K} \sum_{\mathbf{m}\in\mathbb{N}_0^{d}} Q(\mathbf{\mathbf{K}}_{I},\mathbf{m}).
\end{eqnarray*}
Now using the consequence
$\sum_{\mathbf{m}\in \mathbb{N}^d_0} \langle\xi,\mathbf{H}_\mathbf{m}(Z_l)\rangle^2\le\langle\xi,\xi\rangle$
of Parseval's identity
(the latter is used conditionally on $X_{l-1}$
%equivalently, on $\mathcal G_{l-1}$,
which is possible because the system
$\{\mathbf{H}_\mathbf{m}(Z_l)\}_{m\in \mathbb{N}^d_0}$
is orthonormal conditionally on $X_{l-1}$),
%and the tower property $\mathsf E[\mathsf E(\cdot|\mathcal G_{l-1})|\mathcal G_N]=\mathsf E[\cdot|\mathcal G_N]$
we derive
\begin{eqnarray*}
\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}\left[\bar{a}_{l,\mathbf{k}}^{2}(X_{l-1})|\mathcal{G}_{N}\right]
&\leq & \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
\left(\frac{\gamma_{l}}{2}\right)^{|I|K}
\\
&& \times\mathsf E\left[
\mathsf{Var}\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}\partial_{y_{l}}^{\mathbf{K}_I}f\left(X_{p}\right)
\Big|X_{l-1}\right)
\Bigg|\mathcal G_N\right]
\\
&=& \frac{1}{\Gamma^2_{N+2,N+n+1}}\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
\left(\frac{\gamma_{l}}{2}\right)^{|I|K}
R_{l,n,N}^{I,K}
\end{eqnarray*}
with
\begin{eqnarray*}
R_{l,n,N}^{I,K}=\mathsf E\left[
\mathsf{Var}\left(\sum_{p=l}^{N+n}\gamma_{p+1}\partial_{y_{l}}^{\mathbf{K}_I}f\left(X_{p}\right)
\Big|X_{l-1}\right)
\Bigg|\mathcal G_N\right].
\end{eqnarray*}
As a result
\begin{eqnarray*}
\mathsf{Var}\left[\left.\pi_{K,n}^{N}(f)\right|\mathcal G_N\right]\leq \frac{1}{\Gamma^2_{N+2,N+n+1}}\sum_{l=N+1}^{N+n}\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
\left(\frac{\gamma_{l}}{2}\right)^{|I|K}
R_{l,n,N}^{I,K}.
\end{eqnarray*}
Next we show that under the conditions  of Theorem~\ref{th:mr}, the quantity \(R_{l,n,N}^{I,K}\) is uniformly bounded in \(l,n,N,I.\) For the sake of simplicity we present the proof only in one-dimensional case. First we need to prove several auxiliary results.
\begin{lem}\label{lem:06062018a1}
Let $(x_p)_{p\in\mathbb N_0}$
and $(\epsilon_p)_{p\in\mathbb N}$
be sequences of nonnegative real numbers
satisfying $x_0=\ol C_0$ and
\begin{align}
0&\le x_p\le\alpha_p x_{p-1}+\gamma_p \epsilon_p,\quad p\in\mathbb N,
\label{eq:06062018a1}\\
0&\le\epsilon_p\le\ol C_1\prod_{k=1}^p \alpha_k^2,\quad p\in\mathbb N,
\label{eq:06062018a2}
\end{align}
where $\alpha_p,\gamma_p\in(0,1)$, $p\in\mathbb N$,
and $\ol C_0,\ol C_1$ are some nonnegative constants. Assume
\begin{equation}\label{eq:06062018a3}
\sum_{r=1}^\infty\gamma_r \prod_{k=1}^r \alpha_k\le\ol C_2
\end{equation}
for some constant $\ol C_2$. Then
$$
x_p\le(\ol C_0+\ol C_1\ol C_2)\prod_{k=1}^p \alpha_k,\quad p\in\mathbb N.
$$
\end{lem}

\begin{proof}
Applying~\eqref{eq:06062018a1} recursively, we get
$$
x_p\le\ol C_0\prod_{k=1}^p \alpha_k
+\sum_{r=1}^p \gamma_r\epsilon_r
\prod_{k=r+1}^p \alpha_k,
$$
where we use the convention $\prod_{k=p+1}^p:=1$.
Substituting estimate~\eqref{eq:06062018a2}
into the right-hand side, we obtain
$$
x_p\le\left(\ol C_0+\ol C_1
\sum_{r=1}^p \gamma_r \prod_{k=1}^r \alpha_k
\right)
\prod_{k=1}^p \alpha_k,
$$
which, together with~\eqref{eq:06062018a3}, completes the proof.
\end{proof}

In what follows, we use the notation
\begin{equation}
\label{eq:definition-alpha}
\alpha_k=1-\gamma_k b_\mu,\quad k\in\mathbb N.
\end{equation}

\begin{remark}\label{rem:10062018a1}
Notice that, under~\eqref{eq:17042018a1},
not only~\eqref{eq:06062018a3} but also
\begin{equation}\label{eq:10062018a1}
\sum_{r=j}^\infty\gamma_r \prod_{k=j}^r \alpha_k\le\ol C_2
\end{equation}
is satisfied with the same constant~$\ol C_2$
(which is $C$ of~\eqref{eq:17042018a1})
simultaneously for all $j\in\mathbb N$.
Below this will allow us to apply
Lemma~\ref{lem:06062018a1}
to bound double indexed sequences
$(x_{j,p})_{j\ge1,p\ge j}$
satisfying
$$
0\le x_{j,p}\le\alpha_p x_{j,p-1}+\gamma_p \epsilon_{j,p},\quad p\ge j+1,
$$
with suitable $(\epsilon_{j,p})_{j\ge1,p\ge j+1}$
and the constant $\ol C_2$ in~\eqref{eq:10062018a1}
being independent of~$j$.
\end{remark}

\begin{lem}\label{lem:06062018a2}
Under assumptions of Theorem~\ref{th:mr},
for all natural $r\le K$ and $l\le p$, there exist constants $C_r$ (not depending on $l$ and $p$) such that
\begin{equation}
\label{eq:08062018b2}
\left|\partial_{y_l}^r X_p \right| \le C_r\prod_{k=l+1}^p \alpha_k \,,
\end{equation}
where $\partial_{y_l}^r X_p$ is defined in \eqref{eq:definition-differential-f-p}. Moreover, we can choose $C_1=1$.
\end{lem}
\begin{proof}
The proof is along the same lines as Lemma~\ref{lem:06062018a3}.
\end{proof}
\begin{lem}\label{lem:06062018a3}
Under assumptions of Theorem~\ref{th:mr},
for all natural $r\le K$, $j\ge l$ and $p>j$, we have
\begin{equation}\label{eq:08062018b3}
\left|\partial_{y_j} \partial_{y_l}^r X_p\right|
\le c_r\prod_{k=l+1}^p \alpha_k
\end{equation}
with some constants $c_r$
not depending on $j$, $l$ and $p$,
while, for $p\le j$, it holds
$\partial_{y_{j}}\partial_{y_{l}}^{r}X_{p}=0$.
\end{lem}

\begin{proof}
The last statement is straightforward.
We fix natural numbers $j\ge l$ and prove~\eqref{eq:08062018b3}
for all $p>j$ by induction in~$r$.
First, for $p>j$, we write
$$
\partial_{y_{l}}X_{p}
=\left[1-\gamma_{p}\mu'(X_{p-1})\right]\partial_{y_{l}}X_{p-1}
$$
and differentiate this identity with respect to~$y_j$
$$
\partial_{y_{j}}\partial_{y_{l}}X_{p}
=\left[1-\gamma_{p}\mu'(X_{p-1})\right]\partial_{y_{j}}\partial_{y_{l}}X_{p-1}-\gamma_{p}\mu''(X_{p-1})\partial_{y_{j}}X_{p-1}\partial_{y_{l}}X_{p-1}.
$$
By Lemma~\ref{lem:06062018a2}, we have
\begin{align*}
|\partial_{y_{j}}\partial_{y_{l}}X_{p}|
&\le\alpha_p|\partial_{y_{j}}\partial_{y_{l}}X_{p-1}|
+\gamma_{p}
B_\mu
\prod_{k=l+1}^{p-1}\alpha_k
\prod_{k=j+1}^{p-1}\alpha_k\\
&\le\alpha_p|\partial_{y_{j}}\partial_{y_{l}}X_{p-1}|
+\gamma_{p}
\const
\prod_{k=l+1}^{j}\alpha_k
\prod_{k=j+1}^{p}\alpha_k^2,
\quad p\ge j+1,
\end{align*}
with a suitable constant
(we can take, e.g., $\const=\frac{B_\mu}{(1-\gamma_1 b_\mu)^2}$).
By Lemma~\ref{lem:06062018a1}
together with Remark~\ref{rem:10062018a1} applied
to bound $|\partial_{y_{j}}\partial_{y_{l}}X_{p}|$
for $p\ge j+1$
(notice that $\partial_{y_j}\partial_{y_l}X_j=0$, that is,
$\ol C_0$ in Lemma~\ref{lem:06062018a1} is zero,
while $\ol C_1$ in Lemma~\ref{lem:06062018a1}
has the form $\const\prod_{k=l+1}^j \alpha_k$),
we obtain~\eqref{eq:08062018b3} for $r=1$.
The induction hypothesis is now that the inequality
\begin{equation}\label{eq:induc-derive}
\left|\partial_{y_{j}}\partial_{y_{l}}^{k}X_{p}\right|
\leq c_{k}\prod_{s=l+1}^{p}\alpha_{s}
\end{equation}
holds for all natural $k<r\;(\le K)$ and $p>j$.
We need to show~\eqref{eq:induc-derive} for $k=r$.
Fa\`a di Bruno's formula implies for $2\le r\le K$ and $p>l$
\begin{align}
\partial_{y_{l}}^{r}X_{p}
&=\left[1-\gamma_{p}\mu'(X_{p-1})\right]\partial_{y_{l}}^{r}X_{p-1}
\label{eq:08062018a1}\\
&\hspace{1em}-\gamma_{p}\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\mu^{(m_{1}+\ldots+m_{r-1})}(X_{p-1})\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X_{p-1}}{k!}\right)^{m_{k}},
\notag
\end{align}
where the sum is taken over all $(r-1)$-tuples of nonnegative integers
$(m_{1},\ldots,m_{r-1})$ satisfying the constraint
\begin{equation}\label{eq:09062018a1}
1\cdot m_{1}+2\cdot m_{2}+\ldots+(r-1)\cdot m_{r-1}=r.
\end{equation}
Notice that we work with $(r-1)$-tuples
rather than with $r$-tuples
because the term containing
$\partial_{y_l}^r X_{p-1}$
on the right-hand side of~\eqref{eq:08062018a1}
is listed separately.
For $p>j$, we then have
\begin{align}
&\partial_{y_{j}}\partial_{y_{l}}^{r}X_{p}
=\left[1-\gamma_{p}\mu'(X_{p-1})\right]\partial_{y_{j}}\partial_{y_{l}}^{r}X_{p-1}-\gamma_{p}\mu''(X_{p-1})\partial_{y_{l}}^{r}X_{p-1}\partial_{y_{j}}X_{p-1}
\label{eq:09062018a2}\\
&\hspace{1em}-\gamma_{p}\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\mu^{(m_{1}+\ldots+m_{r-1}+1)}(X_{p-1})\partial_{y_{j}}X_{p-1}\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X_{p-1}}{k!}\right)^{m_{k}}
\notag\\
&\hspace{1em}-\gamma_{p}\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\mu^{(m_{1}+\ldots+m_{r-1})}(X_{p-1})\partial_{y_{j}}\left[\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X_{p-1}}{k!}\right)^{m_{k}}\right]
\notag\\
&\hspace{1em} =\left[1-\gamma_{p}\mu'(X_{p-1})\right]\partial_{y_{j}}\partial_{y_{l}}^{r}X_{p-1}+\gamma_{p}\epsilon_{l,j,p},
\notag
\end{align}
where the last equality defines the quantity $\epsilon_{l,j,p}$.
Furthermore,
\begin{eqnarray*}
\partial_{y_{j}}\left[\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X_{p-1}}{k!}\right)^{m_{k}}\right]&=&\sum_{s=1}^{r-1}\frac{m_{s}}{s!}\left(\frac{\partial_{y_{l}}^{s}X_{p-1}}{s!}\right)^{m_{s}-1}\partial_{y_{j}}\partial_{y_{l}}^{s}X_{p-1}
\\
&& \times \prod_{k\le r-1,\,k\neq s}\left(\frac{\partial_{y_{l}}^{k}X_{p-1}}{k!}\right)^{m_{k}}.
\end{eqnarray*}
Using Lemma~\ref{lem:06062018a2},
induction hypothesis~\eqref{eq:induc-derive}
and the fact that $m_{1}+\ldots+m_{r-1}\ge2$
for $(r-1)$-tuples of nonnegative integers
satisfying~\eqref{eq:09062018a1},
we can bound $|\epsilon_{l,j,p}|$ as follows
\begin{align*}
&\left|\epsilon_{l,j,p}\right|  \leq  B_{\mu}C_{r}\prod_{k=l+1}^{p-1}\alpha_{k}\prod_{k=j+1}^{p-1}\alpha_{k}+B_{\mu}\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\left[\prod_{k=j+1}^{p-1}\alpha_{k}\right]
\\
&\times \prod_{s=1}^{r-1}\left(\frac{C_{s}\prod_{k=l+1}^{p-1}\alpha_{k}}{s!}\right)^{m_{s}}\\
&+B_{\mu}\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\sum_{t=1}^{r-1}\frac{m_t}{t!}\left(\frac{C_{t}\prod_{k=l+1}^{p-1}\alpha_{k}}{t!}\right)^{m_{t}-1}c_t\left[\prod_{k=l+1}^{p-1}\alpha_{k}\right]
\\
&\times \prod_{s\le r-1,\,s\neq t}\left(\frac{C_{s}\prod_{k=l+1}^{p-1}\alpha_{k}}{s!}\right)^{m_{s}}\leq \const\prod_{k=l+1}^{j}\alpha_{k}\prod_{k=j+1}^{p}\alpha_{k}^{2}
\end{align*}
with some constant ``$\const$'',
which is, in fact, $\frac1{(1-\gamma_1 b_\mu)^2}$ times
the expression involving
$B_\mu,r,C_1,\ldots,C_r,c_1,\ldots,c_{r-1}$.
Thus, \eqref{eq:09062018a2} now implies
$$
|\partial_{y_{j}}\partial_{y_{l}}^{r}X_{p}|
\le\alpha_p|\partial_{y_{j}}\partial_{y_{l}}^{r}X_{p-1}|
+\gamma_p \const\prod_{k=l+1}^{j}\alpha_{k}\prod_{k=j+1}^{p}\alpha_{k}^{2},
\quad p\ge j+1.
$$
We can again apply Lemma~\ref{lem:06062018a1}
and Remark~\ref{rem:10062018a1}
to bound $|\partial_{y_{j}}\partial_{y_{l}}^r X_{p}|$
for $p\ge j+1$
(notice that $\partial_{y_j}\partial_{y_l}^r X_j=0$, that is,
$\ol C_0$ in Lemma~\ref{lem:06062018a1} is zero,
while $\ol C_1$ in Lemma~\ref{lem:06062018a1}
has the form $\const\prod_{k=l+1}^j \alpha_k$),
and we obtain~\eqref{eq:induc-derive} for $k=r$.
This concludes the proof.
\end{proof}

\begin{lem}
\label{lem:var_poincare}
Under assumptions of Theorem~\ref{th:mr},
for all natural $l\le q$, it holds
$$
\mathsf{Var}\left[\sum_{p=l}^{q}\gamma_{p+1}\partial_{y_{l}}^{K}f\left(X_{p}\right)\Big|X_{l-1}\right]\le B_K\quad\text{a.s.},
$$
where $B_K$ is a deterministic bound that does not depend on $l$ and~$q$.
\end{lem}

\begin{proof}
The expression
$\sum_{p=l}^q \gamma_{p+1} \partial_{y_l}^K f(X_p)$
can be viewed as a deterministic function of
$X_{l-1},Z_l,Z_{l+1},\ldots,Z_q$
$$
\sum_{p=l}^q \gamma_{p+1} \partial_{y_l}^K f(X_p)
=F(X_{l-1},Z_l,Z_{l+1},\ldots,Z_q).
$$
By the (conditional) Gaussian Poincar\'e inequality,
we have
$$
\mathsf{Var}\left[\sum_{p=l}^{q}\gamma_{p+1}\partial_{y_{l}}^{K}f\left(X_{p}\right)\Big|X_{l-1}\right]
\le\mathsf E\left[\left.
\|\nabla_Z F(X_{l-1},Z_l,Z_{l+1},\ldots,Z_q)\|^2
\right|X_{l-1}\right],
$$
where $\nabla_Z F=(\partial_{Z_l} F,\ldots,\partial_{Z_q} F)$,
and $\|\cdot\|$ denotes the Euclidean norm.
Notice that \(\partial_{Z_j} F=\sqrt{\gamma_j}\,\partial_{y_j} F\) and
hence
$$
\mathsf{Var}\left[\sum_{p=l}^{q}\gamma_{p+1}\partial_{y_{l}}^{K}f\left(X_{p}\right)\Big|X_{l-1}\right]\leq\sum_{j=l}^{q}\gamma_{j}\mathsf{E}\left[\left(\sum_{p=l}^{q}\gamma_{p+1}\partial_{y_{j}}\partial_{y_{l}}^{K}f\left(X_{p}\right)\right)^{2}\Big|X_{l-1}\right].
$$
It is straightforward to check that
$\partial_{y_j}\partial_{y_l}^K f(X_p)=0$
whenever $p<j$. Therefore, we get
\begin{equation}\label{eq:10062018a2}
\mathsf{Var}\left[\sum_{p=l}^{q}\gamma_{p+1}\partial_{y_{l}}^{K}f\left(X_{p}\right)\Big|X_{l-1}\right]\leq\sum_{j=l}^{q}\gamma_{j}\mathsf{E}\left[\left(\sum_{p=j}^{q}\gamma_{p+1}\partial_{y_{j}}\partial_{y_{l}}^{K}f\left(X_{p}\right)\right)^{2}\Big|X_{l-1}\right].
\end{equation}
Now fix $p$ and $j$, $p\ge j$, in $\{l,\ldots,q\}$.
By Fa\`a di Bruno's formula
\[
\partial_{y_{l}}^{K}f\left(X_{p}\right)=\sum\frac{K!}{m_{1}!\ldots m_{K}!}f^{(m_{1}+\ldots+m_{K})}(X_{p})\prod_{k=1}^{K}\left(\frac{\partial_{y_{l}}^{k}X_{p}}{k!}\right)^{m_{k}},
\]
where the sum is taken over all $K$-tuples of nonnegative integers
$(m_1,\ldots,m_K)$ satisfying
$$
1\cdot m_{1}+2\cdot m_{2}+\ldots+K\cdot m_{K}=K.
$$
Then
\begin{eqnarray*}
\partial_{y_{j}}\partial_{y_{l}}^{K}f\left(X_{p}\right)
&=&\sum\frac{K!}{m_{1}!\ldots m_{K}!}f^{(m_{1}+\ldots+m_{K}+1)}(X_{p})\left[\partial_{y_{j}}X_{p}\right]\prod_{k=1}^{K}\left(\frac{\partial_{y_{l}}^{k}X_{p}}{k!}\right)^{m_{k}} \\
&&+\sum\frac{K!}{m_{1}!\ldots m_{K}!}f^{(m_{1}+\ldots+m_{K})}(X_{p})
\sum_{s=1}^{K}\frac{m_{s}}{s!}
\left(\frac{\partial_{y_{l}}^{s}X_{p}}{s!}\right)^{m_{s}-1}
\\
&& \times\left[\partial_{y_{j}}\partial_{y_{l}}^{s}X_{p}\right]
\prod_{k\le K,\,k\neq s}\left(\frac{\partial_{y_{l}}^{k}X_{p}}{k!}\right)^{m_{k}}.
\end{eqnarray*}
Using the bounds of
Lemmas \ref{lem:06062018a2} and~\ref{lem:06062018a3},
we obtain
\begin{eqnarray}
\label{eq:part-part}
\left|\partial_{y_{j}}\partial_{y_{l}}^{K}f\left(X_{p}\right)\right|
\leq A_{K}\prod_{k=l+1}^{p}\alpha_{k}
\end{eqnarray}
with a suitable constant $A_{K}$.
Substituting this in~\eqref{eq:10062018a2},
we proceed as follows
\begin{align*}
\mathsf{Var}\left[\sum_{p=l}^{q}\gamma_{p+1}\partial_{y_{l}}^{K}f\left(X_{p}\right)\Big|X_{l-1}\right]
&\le
A_{K}^{2}\sum_{j=l}^{q}\gamma_{j}
\left(\sum_{p=j}^{q}\gamma_{p+1}\prod_{k=l+1}^{p}\alpha_{k}\right)^{2}
\\
&\le
\frac{A_{K}^{2}}{(1-\gamma_1 b_\mu)^2}
\sum_{j=l}^{q} \gamma_{j}
\left(
\sum_{p=j+1}^{q+1} \gamma_{p} \prod_{k=l+1}^{p} \alpha_{k}
\right)^{2}
\\
%&=
%\frac{A_{K}^{2}}{(1-\gamma_1 b_\mu)^2}
%\sum_{j=l}^{q} \gamma_{j}
%\prod_{k=l+1}^{j} \alpha_{k}^2
%\left(
%\sum_{p=j+1}^{q+1} \gamma_{p} \prod_{k=j+1}^{p} \alpha_{k}
%\right)^{2}
%\\
&\le
\frac{A_{K}^{2}}{(1-\gamma_1 b_\mu)^3}
\sum_{j=l}^{q} \gamma_{j}
\prod_{k=l}^{j} \alpha_{k}
\left(
\sum_{p=j+1}^{q+1} \gamma_{p} \prod_{k=j+1}^{p} \alpha_{k}
\right)^{2}
\\
&\le
\frac{A_{K}^{2}}{(1-\gamma_1 b_\mu)^3}C^3=B_K,
\end{align*}
where $C$ is the bound from~\eqref{eq:17042018a1}.
The proof is completed.
\end{proof}
\begin{remark}\label{rem:p-norm}
In fact in the proof of Lemma~\ref{lem:var_poincare} using \eqref{eq:part-part} we have shown that the expectations
\begin{eqnarray*}
\mathsf E\left[\left.
\|\nabla_Z F(X_{l-1},Z_l,Z_{l+1},\ldots,Z_q)\|^r
\right|X_{l-1}\right]
\end{eqnarray*}
are uniformly bounded for all natural \(l\leq q\) and any \(r\geq 2.\) Indeed it follows from \eqref{eq:17042018a1} that  the quantities
\begin{eqnarray*}
\left(\sum_{j=l}^{q} \gamma_{j}
\prod_{k=l}^{j} \alpha_{k}\right)^{r/2}
\left(
\sum_{p=j+1}^{q+1} \gamma_{p} \prod_{k=j+1}^{p} \alpha_{k}
\right)^{r}
\end{eqnarray*}
are bounded. This stronger result in combination with a \(L^r\)-Sobolev-type inequalities (see e.g. \cite{Adamczak}) implies that also higher order conditional (on \(X_{l-1}\)) central moments of the r.v.
\begin{eqnarray*}
\sum_{p=l}^{q}\gamma_{p+1}\partial_{y_{l}}^{K}f\left(X_{p}\right)
\end{eqnarray*}
are uniformly bounded in \(l\leq q\) and \(p.\)

\end{remark}

\subsection{Proof of Lemma~\ref{a_decay}}

We have for any \(\mathbf{k}\neq \mathbf{0}\)
\begin{eqnarray*}
a_{p,l,\mathbf{k}}(x)=\mathsf{E}\left[\mathbf{H}_\mathbf{k}\left(Z_l\right)[Q_{p,l}\left(\Phi_l(x,Z_l)\right)-Q_{p,l}\left(\Phi_l(x,0)\right)]\right].
\end{eqnarray*}
Hence
\begin{eqnarray*}
\|a_{p,l,\mathbf{k}}\|_{\infty}\leq \sqrt{\gamma_l}\|\nabla Q_{p,l}\|_{\infty}.
\end{eqnarray*}
Since \(f,\mu\in C^1(\mathbb{R}^d),\)  we have
\begin{eqnarray*}
\nabla Q_{p,l}(x)=\mathsf{E}\left[(I-\gamma_{l+1}\nabla\mu(x))\nabla Q_{p,l+1}(x-\gamma_{l+1}\mu(x)+\sqrt{\gamma_{l+1}}\xi)\right].
\end{eqnarray*}
Hence
\begin{eqnarray*}
\|\nabla Q_{p,l}\|_\infty\leq (1-b_{\mu}\gamma_{l+1})\|\nabla Q_{p,l+1}\|_\infty
\end{eqnarray*}
and
\begin{eqnarray*}
\|\nabla Q_{p,l}\|_\infty&\leq & \|\nabla f\|_\infty\prod_{r=l+1}^p (1-b_{\mu}\gamma_{r})
\\
&\leq & \|\nabla f\|_\infty\exp\left(-b_{\mu}\sum_{r=l+1}^p \gamma_{r}  \right).
\end{eqnarray*}
As a result
\begin{eqnarray*}
\|a_{p,l,\mathbf{k}}\|_{\infty}\leq \sqrt{\gamma_l}\, \|\nabla f\|_\infty\exp\left(-b_{\mu}\sum_{r=l+1}^p \gamma_{r}  \right).
\end{eqnarray*}

\bibliographystyle{plain}
\bibliography{refs-1}
\end{document}

### 2019-12-01 aap/Revision/vr-mcmc-revision.tex
Revision

@@
 \subfloat[Subfigure 1 list of figures text][d = 2]{
-\includegraphics[width=0.45\textwidth]{pictures/GM_2d_f1.png}
+\includegraphics[width=0.45\textwidth]{../pictures/GM_2d_f1.png}
 \label{fig:subfig1}}
@@
 \subfloat[Subfigure 2 list of figures text][d = 8]{
-\includegraphics[width=0.45\textwidth]{pictures/GM_8d_f1.png}
+\includegraphics[width=0.45\textwidth]{../pictures/GM_8d_f1.png}
 \label{fig:subfig2}}
@@
 \begin{center}
-\includegraphics[width=0.6\textwidth]{pictures/BLR_banknotes_4d.png}
+\includegraphics[width=0.6\textwidth]{../pictures/BLR_banknotes_4d.png}
 \caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Logistic Regression.
@@
 \begin{center}
-\includegraphics[width=0.6\textwidth]{pictures/Probit_banknotes_4d.png}
+\includegraphics[width=0.6\textwidth]{../pictures/Probit_banknotes_4d.png}
 \caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Probit Regression.
@@
 
+\appendix
+
+\section{Concentration bounds for  ULA}
+For $\gamma>0$, the Markov kernel $R_\gamma$
+from $(\mathbb R^d,\mathcal B(\mathbb R^d))$
+into $(\mathbb R^d,\mathcal B(\mathbb R^d))$
+is defined by
+$$
+R_\gamma(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\}dy,\quad A\in \mathcal B(\mathbb R^d),
+$$
+while, for $k,l\in\mathbb N$, $k\le l$,
+the kernel $Q_{\boldsymbol{\gamma}}^{k,l}$ is
+$
+Q_{\boldsymbol{\gamma}}^{k,l}=R_{\gamma_l}\cdots R_{\gamma_k}.
+$
+ We make the following assumptions.
+
+\begin{description}
+\item[(H1)] Assume that $\nabla U$ is Lipschitz, that is, there exists $L\geq 0$ such that 
+\begin{eqnarray*}
+\|\nabla U(x)-\nabla U(y)\|\leq L\|x-y\|, \quad x,y\in \mathbb{R}^d
+\end{eqnarray*}
+
+\item[(H2)] There exist $K>0,$ $M>0$ and $m>0$ such that for any $x\not\in B_K(0)	$ and any $y\in \mathbb{R}^d$ it holds 
+\begin{eqnarray*}
+\langle D^2U(x)y,y\rangle \geq m\|y\|^2, \quad \|D^3 U(y)\| \leq M. 
+\end{eqnarray*}
+
+\end{description}
+
 \bibliographystyle{plain}

### 2019-12-01 aap/Revision/vr-mcmc-revision.tex
Revision

@@
 
-\section{Concentration bounds for  ULA}
+\section{Bounds for moments of  ULA}
 For $\gamma>0$, the Markov kernel $R_\gamma$
-from $(\mathbb R^d,\mathcal B(\mathbb R^d))$
-into $(\mathbb R^d,\mathcal B(\mathbb R^d))$
+on $(\mathbb R^d,\mathcal B(\mathbb R^d))$
 is defined by
 $$
-R_\gamma(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\}dy,\quad A\in \mathcal B(\mathbb R^d),
+R_\gamma(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \nabla U(x)\right\|^2\right\}dy,\quad A\in \mathcal B(\mathbb R^d),
 $$
@@
 \begin{eqnarray*}
-\|\nabla U(x)-\nabla U(y)\|\leq L\|x-y\|, \quad x,y\in \mathbb{R}^d
+\|\nabla U(x)-\nabla U(y)\|\leq L\|x-y\|, \quad x,y\in \mathbb{R}^d.
 \end{eqnarray*}
@@
 \end{description}
+
+\begin{thm}
+Suppose that
+\begin{eqnarray*}
+\sum_{p=1}^{\infty} \gamma^2_p<\infty.
+\end{eqnarray*}
+Under assumptions (H1) and (H2), then for any bounded function $f$ on $\mathbb{R}^d$ 
+we have
+\begin{eqnarray*}
+\mathsf{E}\left[\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})-\bar Q_l(X_l)\right)^{4}\right] \lesssim \frac{1}{\Gamma^4_{N+2,N+n+1}},
+\end{eqnarray*}
+ and
+\begin{eqnarray*}
+\bar Q_l(X_l)=\mathsf{E}\left[\left.\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right|X_{l}\right].
+\end{eqnarray*}
+Moreover 
+\begin{eqnarray*}
+\mathsf{E}\left[\exp\left(\rho\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k})-\left(\sum_{k=l}^{p}\gamma_{k+1}\right)\pi(g)\right)\right)\right]\leq C
+\end{eqnarray*}
+for some small enough $\rho>0,$  all natural $0<l\leq p<\infty$ and constant $C$ not depending on $l,p.$
+\end{thm}
 

### 2019-12-11 aap/Revision/vr-mcmc-revision.tex
Revision

@@
 \end{prop}
+\par
 Next we show how  the representation \eqref{eq:mart_repr} can be used to efficiently reduce the variance of MCMC algorithms.
@@
 Since \(X_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\geq 1,\)  the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
-
-\section{Analysis of variance reduced ULA}
-\label{sec:ula_analysis}
 The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
@@
 should be small for \(K\) large enough. In this section   we provide  an analysis of the variance-reduced ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
+\section{Analysis of variance reduced ULA}
+\label{sec:ula_analysis}
+
 We shall use the notations $\mathbb N=\{1,2,\ldots\}$ and $\mathbb N_0=\mathbb N\cup\{0\}$.
@@
 For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
-of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we are interested in its conditional
-Mean Squared Error (MSE),
-which can be decomposed as the sum of the squared conditional bias
-and the conditional variance:
+of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we shall be interested in the Mean Squared Error (MSE),
+which can be decomposed as the sum of the squared  bias
+and the  variance:
 \begin{eqnarray}\label{eq:29032018a4}
-\mathrm{MSE}\left[\rho(f)|\mathcal G_N\right]
-&=&\mathsf{E}\left[(\rho(f)-\pi(f))^2|\mathcal G_N\right]
+\mathrm{MSE}\left[\rho(f)\right]
+&=&\mathsf{E}\left[(\rho(f)-\pi(f))^2\right]
 \\
 \nonumber
-&=&\left(\mathsf{E}[\rho(f)|\mathcal G_N]-\pi(f)\right)^2
-+\mathsf{Var}\left[\rho(f)|\mathcal G_N\right].
+&=&\left(\mathsf{E}[\rho(f)]-\pi(f)\right)^2
++\mathsf{Var}[\rho(f)].
 \end{eqnarray}
-The quantities in~\eqref{eq:29032018a4}
-are conditioned on $\mathcal G_N$,
-as it reflects the way the estimators are used for MCMC:
-the path of the Markov chain is simulated only once,
-and we start to use the realized values
-of the Markov chain to construct
-our estimate only after the burn-in period of length~$N$.
-We also notice that, due to the Markovian structure,
-the conditioning on $\mathcal G_N$
-in~\eqref{eq:29032018a4}
-is equivalent to conditioning on $X_N$ only
-(this is particularly clear in the case $\rho(f)=\pi_n^N(f)$
-but requires some calculation
-in the remaining case $\rho(f)=\pi_{K,n}^N(f)$).
-
-\subsection{Squared conditional bias}
+Our analysis will be done under the following two assumptions.
+
+\begin{description}
+\item[(H1)][\textbf{Lipschitz continuity}] Assume that $\nabla U$ is Lipschitz, that is, there exists $L\geq 0$ such that 
+\begin{eqnarray*}
+\|\nabla U(x)-\nabla U(y)\|\leq L\|x-y\|, \quad x,y\in \mathbb{R}^d.
+\end{eqnarray*}
+
+\item[(H2)][\textbf{Convexity outside a ball}] There exist $K>0,$ $M>0$ and $m>0$ such that for any $x\not\in B_K(0)	$ and any $y\in \mathbb{R}^d$ it holds 
+\begin{eqnarray*}
+\langle D^2U(x)y,y\rangle \geq m\|y\|^2, \quad \|D^3 U(y)\| \leq M. 
+\end{eqnarray*}
+\end{description}
+First we analyse the biases of the estimators
+$\pi_n^N(f)$ and $\pi_{K,n}^N(f).$
+\subsection{Squared bias}
 Due to the martingale transform structure
@@
 \[
-\mathsf E\left[\left.M_{K,n}^N(f)\right|\mathcal G_N\right]=0,
+\mathsf E\left[M_{K,n}^N(f)\right]=0,
 \]
@@
 \begin{align}
-\left(\mathsf{E}[\pi_{K,n}^N(f)|\mathcal G_N]-\pi(f)\right)^2
-&=\left(\mathsf{E}[\pi_{n}^N(f)|\mathcal G_N]-\pi(f)\right)^2
+\left(\mathsf{E}[\pi_{K,n}^N(f)]-\pi(f)\right)^2
+&=\left(\mathsf{E}[\pi_{n}^N(f)]-\pi(f)\right)^2
 \label{eq:06042018a2}\\
@@
 \omega_{p,n}^N\,
-\|Q_\gamma^{N+1,p}(X_N,\cdot)-\pi(\cdot)\|_{TV}^2,
+\|Q_\gamma^{p}(x_0,\cdot)-\pi(\cdot)\|_{TV}^2,
 \notag
@@
 using results of Section~3 in~\cite{durmus:moulines:2017}
-on bounds in the total variation distance.
-
+on bounds in the total variation distance. For example, the following proposition holds.
+\begin{prop}
+Suppose that the assumption (H1) holds, then it holds
+\begin{eqnarray*}
+\|Q_\gamma^{p}(x,\cdot)-\pi(\cdot)\|^2_{TV}\lesssim \sum_{k=r+1}^{p} \gamma_k^2+\exp\left(-\varkappa\sum_{k=r+1}^{p} \gamma_k\right),
+\end{eqnarray*}
+for some $\varkappa>0,$ where $\lesssim$ stands for  inequality up to a constant depending on $L$ and $\mathcal G_N.$ In particular if $\gamma_k=k^{-\alpha}$ for some $\alpha\in (0.5,1], $ then 
+\begin{eqnarray*}
+\|Q_\gamma^{N+1,N+n}(x,\cdot)-\pi(\cdot)\|^2_{TV}\lesssim \frac{1}{(N+1)^{2\alpha-1}}
+\end{eqnarray*}
+provided that $N/n\to 0$ as $N,n\to \infty.$
+\end{prop}
 \subsection{Conditional variance}
@@
 \begin{thm}\label{th:mr}
-Fix $K\in\mathbb N$.
-Suppose that $f$  and \(\mu\) are \(K+1\) times continuously differentiable
+Suppose that $f$  is two times continuously differentiable
 and it holds
 \begin{eqnarray*}
-|\partial^{\mathbf{k}} f(x)|\le  B_f,\quad |\partial^{\mathbf{k}}\mu(x)|\le  B_\mu, \quad
+|\partial^{\mathbf{k}} f(x)|\le  B_f, \quad
 x\in\mathbb R^d
 \end{eqnarray*}
-for all \(\mathbf{k}\) with \(0<\|\mathbf{k}\|\leq  K+1\),
-\[
-\jac{\mu}(x)\ge b_\mu I ,\quad x\in\mathbb R^d,
-\]
-for some positive number $b_\mu\in(0,B_\mu]$  where  \(\jac{\mu}\) stands for Jacobian of \(\mu.\)
+for all \(\mathbf{k}\) with \(0<\|\mathbf{k}\|\leq  2.\) Moreover assume that there exist $K>0,$ $M>0$ and $m>0$ such that for any $x\not\in B_K(0)$ and any $y\in \mathbb{R}^d$ it holds
+\[
+\langle \nabla \mu(x)y,y\rangle \geq m\|y\|^2.
+\]
 Let $(\gamma_k)_{k\in\mathbb N}$ be a sequence of positive
-and non-increasing step sizes with $\sum_{k=1}^\infty \gamma_k=\infty$.
-We  assume that $\gamma_1<\frac1{B_\mu}$ and that
-\begin{equation}\label{eq:17042018a1}
-\sum_{r=j}^\infty \gamma_r\prod_{k=j}^{r}\left[1-\gamma_{k}b_\mu\right]\leq C,\quad\text{for all }j\in\mathbb N,
-\end{equation}
-with some constant \(C\). Then it holds
+and non-increasing step sizes with $\sum_{k=1}^\infty \gamma_k=\infty$ and $\sum_{k=1}^\infty \gamma^2_k<\infty.$
+Then it holds under 
 \begin{eqnarray}
 \label{eq:var-bound}
-\mathsf{Var}\left[\left.\pi_{K,n}^{N}(f)\right|\mathcal G_N\right]\lesssim \frac{1}{\Gamma^2_{N+2,N+n+1}}\sum_{l=N+1}^{N+n}\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
-\left(\frac{\gamma_{l}}{2}\right)^{|I|K},
+\mathsf{Var}\left[\left.\pi_{K,n}^{N}(f)\right|\mathcal G_N\right]\lesssim \frac{1}{\Gamma^2_{N+2,N+n+1}},
 \end{eqnarray}
-where the sum in \eqref{eq:var-bound} is taken over all subsets \(I\) of \(\{1,\ldots,d\}\) and \(\lesssim\) stands for inequality up to a constant not depending  on \(n\) and \(N.\)
+where \(\lesssim\) stands for inequality up to a constant not depending  on \(n\) and \(N.\)
 \end{thm}
-\begin{remark}
-Note that in  Theorem~\eqref{th:mr} we don't require that  the function \(\mu\) itself is bounded.
-Assumption~\eqref{eq:17042018a1} is not restrictive.
-For instance, a straightforward calculation shows
-that \eqref{eq:17042018a1} is satisfied in most interesting case
-$\gamma_k=\const/k^\alpha$ with $\alpha\in (0,1)$.
-\end{remark}
+\begin{cor}
+
+\end{cor}
 
@@
 \begin{eqnarray*}
-\sum_{p=1}^{\infty} \gamma^2_p<\infty.
-\end{eqnarray*}
-Under assumptions (H1) and (H2), then for any bounded function $f$ on $\mathbb{R}^d$ 
+\sum_{p=1}^{\infty} \gamma_p=\infty, \quad \sum_{p=1}^{\infty} \gamma^2_p<\infty.
+\end{eqnarray*}
+Under assumptions (H1) and (H2), then for any bounded function $g$ on $\mathbb{R}^d$ 
 we have
 \begin{eqnarray*}
-\mathsf{E}\left[\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})-\bar Q_l(X_l)\right)^{4}\right] \lesssim \frac{1}{\Gamma^4_{N+2,N+n+1}},
+\mathsf{E}\left[\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}g(X_{p})-\bar Q_l(X_l)\right)^{4}\right] \lesssim \frac{1}{\Gamma^4_{N+2,N+n+1}}
 \end{eqnarray*}
@@
 \end{eqnarray*}
-Moreover 
+Moreover, 
 \begin{eqnarray*}

### 2019-12-11 Note_exponential_moment/MarkovCon.tex

\documentclass[leqno,11pt,a4paper]{article}
\usepackage{amsmath,amsthm}

\textheight=9.0truein
\hoffset=-0.4truein
\textwidth=6.1truein
\voffset=-0.5truein

\input{header.tex}
\input{defs_AHM.tex}

%\usepackage{mathabx} %package utilisé pour définir "\lll" comme le symbole d'absolue continuité de 2 mesures
\usepackage[toc,page]{appendix} %package utilisé pour l'appendice voir : https://texfaq.org/FAQ-appendix
\usepackage{tikz}
%\usetikzlibrary{shapes}
%"Ce document a été généré avec la version de TikZ \pgfversion ." : à mettre après "begin{document}" si affichage souhaité...
\newtheorem{rem}{Remark}
\setcounter{secnumdepth}{3} %permet de numéroter les sous-sous-sections

\begin{document}
\title{A quantitative Mac Diarmid's inequality for geometrically ergodic Markov chains}
%\title{Density estimation for RWRE : a Bayesian non-parametric approach}
%\author{A. Havet, M. Lerasle, E. Moulines and E. Vernet}
\author{D. Belomestny, E. Moulines and S. Samsonov}
\maketitle

\begin{abstract}
 We state and prove a quantitative version of the bounded difference inequality for geometrically ergodic Markov chains.
 Our proof uses the same martingale decomposition as \cite{MR3407208} but, compared to this paper, the exact coupling argument is modified to fill a gap between the strongly aperiodic case and the general aperiodic case.
\end{abstract}

\noindent
{\small {\bf Keywords:} Concentration inequalities ; Markov chains ; Geometric ergodicity ; Coupling.}

\noindent
{\small {\bf AMS MSC 2010:} 60J05; 60E15.}

\section{Introduction}
The purpose of this note is to establish a quantitative version of Mc Diarmid's inequality for geometrically ergodic Markov chains.
Let $X_0,\ldots,X_{n-1}$ denote independent random variables taking values in a measurable space $(\sX,\cX)$ and $c=(c_0,\ldots,c_{n-1})$ denote a vector of non-negative real numbers. A function  $f:\sX^n\to\Rbb$ satisfies the bounded difference inequality if for all $x=(x_0,\ldots,x_{n-1})$ and
$y=(y_0,\ldots,y_{n-1})\in \sX^{n}$, we have
\begin{equation}\label{def:BDP}
|f(x)-f(y)|\leqslant \sum_{i=0}^{n-1}c_i\indiacc{x_i\ne y_i}\eqsp.
\end{equation}
The bounded difference inequality, first established in \cite{MR1036755}, shows that for all $t > 0$,
\[
\Pbb\big(f(X_0,\ldots,X_{n-1})-\Ebb[f(X_0,\ldots,X_{n-1})]>t\big)\leqslant \rme^{-2t^2/\|c\|^2}\eqsp,
\]
where $\|c\|^2=\sum_{i=0}^{n-1}c_i^2$.
Several attempts have been made to extend this result to Markov chains.
In \cite{MR2424985}, the concentration of particular functionals of the form $f(x_0,\ldots,x_{n-1})=\sup_{g\in \cF}\sum_{i=0}^{n-1}g(x_i)$, for centered functions $g$ in a class $\cF$ is established.
The concentration of general functionals (satisfying \eqref{def:BDP}) of geometrically ergodic Markov chains was established in \cite{MR3407208}, where it is also proved that geometric ergodicity is a necessary assumption. However, the result in \cite{MR3407208} is not quantitative.
It states that for all geometrically recurrent set $C$, there exists a constant $\beta$, depending on $C$ such that for all $x \in C$ and $t > 0$,
\begin{equation}
\label{eq:mcdiarmid-markov}
\Pbb_x\big(f(X_0,\ldots,X_{n-1})-\Ebb_x[f(X_0,\ldots,X_{n-1})]>t\big)\leqslant \rme^{-\beta t^2/\|c\|^2}\eqsp,
\end{equation}
where for any $x\in\Xsigma$, $\Pbb_x$ is the distribution of the Markov chain $\{X_k\}_{k=0}^\infty$ starting from $x$ (see the precise definition below).
In many  applications, it is necessary to get the explicit dependence of the constant $\beta$ as a function of the set $C$.
In particular, this problem arises when establishing posterior concentration rates of Bayesian non-parametric estimators; see for example \cite{rousseau:2016,ghosal:vandervaart:2017} for recent accounts on this theory.
To extend these results to Markovian settings, the result of \cite{MR3407208} cannot be applied directly and a quantitative version of  \eqref{eq:mcdiarmid-markov} is required, where the dependence of $\beta$ on constants characterizing the mixing of the Markov chain is needed; see for example \cite{vernet:2015,lecorff:lerasle:vernet:2018}.

A quantitative version of Mc Diarmid's inequality for Markov chains was established in \cite{paulin:2015}, where the constant $\beta$ depends here explicitly on the mixing time of the chain.
The existence of finite mixing times requires \emph{uniform} ergodicity of the chain, see for example \cite[Section 3.3]{MR2095565}, an  assumption that typically fails when the chain takes value in general state spaces.
In this note, we prove an extension of Mc Diarmid's inequality to geometrically ergodic Markov chains.
Our proof is based on  \cite{MR3407208},
but avoids the use of \cite[Lemma~6]{MR3407208} which requires the construction of an exact coupling.
Exact coupling can actually be built in the strongly aperiodic case but there is a gap in the general aperiodic case.

The remaining of the paper is decomposed as follows, Section~\ref{sec:Setting} introduces formally the notations and the assumptions of the main result, which is stated and proved in Section~\ref{sec:Main}.

\section{Notations and assumptions}\label{sec:Setting}
Let $(\sX,\cX)$ be a measurable space.
We denote by $\tvdistsym$  the total variation distance between probability measures.
For any sequence $x=\sequence{x}[n][\Nbb]$ and any non-negative integers $a$ and $b$, with $a\leqslant b$, let $\chunk[a]{x}[b]=(x_a,x_{a+1},\ldots,x_b)$.
For any $n\geqslant 0$ and any vector $c=\chunk[0]{c}[n-1]\in \R^n$, let $\|c\|$ denote the Euclidean norm of $c$ and $\|c\|_\infty=\max_{0\leqslant i\leqslant n-1}|c_i|$ denote its sup-norm.

We denote by $(\sX^{\Zbb_+},\cX^{\otimes\Zbb_+}, (\cF_k)_{k\geqslant 0})$ the canonical filtered space,
$\{X_n\}_{n=0}^\infty$ the canonical process and $\theta: \sX^{\Zbb_+}\to\sX^{\Zbb_+}$  the shift operator on the canonical space defined, for any $x=(x_{n})_{n\geqslant 0}\in \sX^{\Zbb_+}$ by $\theta(x)\in \sX^{\Zbb_+}$, where, for any $n\geqslant 0$, $\theta(x)_n=x_{n+1}$.
Set $\theta_1=\theta$ and for $n\in \Nbb^*$, define inductively, $\theta_n=\theta_{n-1} \circ \theta$. We also need to define $\theta_\infty$. To this aim, fix an arbitrary $x^*\in \Xset$, we define $\theta_\infty:\Xset^\Nbb \to \Xset^\Nbb$ such that for $z=\sequence{z}[k][\Nbb] \in \Xset^\Nbb$, $\theta_\infty z\in \Xset^\Nbb$ is the constant sequence  $(\theta_\infty z)_k=x^*$ for all $k \in \Nbb$.

Let $P$ be a Markov kernel on $\sX \times \cX$.
For any probability measure $\xi$ on $(\Xset,\Xsigma)$, denote by $\Pbb_{\xi}$  the unique probability under which $(X_n)_{n\geqslant 0}$ is a Markov chain with Markov kernel $P$ and initial distribution $\xi$ and let $\Ebb_{\xi}$ denote the expectation under the distribution $\Pbb_{\xi}$.
Recall that $\cF_n$ denotes the $\sigma$-algebra generated by $X_0, \ldots,X_n$.
For any $x\in \sX$, let $\delta_x$ denote the Dirac mass at point $x$.
With some abuse of notation, we also denote $\Pbb_x$ (resp. $\Ebb_x$) instead of $\Pbb_{\delta_x}$ (resp. $\Ebb_{\delta_x}$).

For any $\cB\in \cX$ and any integer $i\geqslant 0$, let
\[
\tau_{\cB}^i=\inf\{n\geqslant i: X_n\in \cB\}=i+\tau_{\cB}^0\circ \theta^i
\qquad \text{and} \qquad
\sigma_{\cB}=\tau_{\cB}^1=1+\tau_{\cB}^0\circ \theta\eqsp.
\]
For $c = \chunk[0]{c}[n-1] \in \R_+^n$, we denote by  $\functionboundeddiff{\sX^n}{c}$  the set of measurable functions $f:\sX^n\to\R$ such that for all $x= (x_0,\dots,x_{n-1})$ and $y= (y_0,\dots,y_{n-1})$, $|f(x)-f(y)|\leqslant \sum_{i=0}^{n-1} c_i\indiacc{x_i\ne y_i}$
The main result is established under the following conditions.
\begin{itemize}
\item[{\bf H1}] The Markov kernel $P$ is irreducible and aperiodic, with unique invariant probability $\pi$.
\item[{\bf H2}] There exist a non-empty set $\cC \in \cX$ and two real numbers $u>1$ and $M>0$ such that
  \[\sup_{x\in \cC}\Ebb_x[u^{\sigma_{\cC}}]\leqslant M\eqsp.\]
\item[{\bf H3}] There exist $r\in (0,1)$ and $L\geqslant 1$ such that, for any $x$ in the set $\cC$ of {\bf H2} and any $n\geqslant 0$,
\[
\tvdist{\delta_xP^n}{\pi} \leqslant Lr^n\eqsp,
\]
where $\pi$ is the unique invariant measure granted in {\bf H1}.
\end{itemize}
When the Markov kernel $P$ is uniformly ergodic, then {\bf H3} holds with $\cC=\sX$. The following Lemma is a coupling result that replaces \cite[Lemma~6]{MR3407208}.
It is instrumental in the sequel.
\begin{lemma}\label{lem:DMPS18}
 For any probability measures $\xi$ and $\xi'$ on $(\sX,\cX)$, any $n\geqslant 1$, any $c\in \R_+^n$ and any $h\in \bB\Dbb(\sX^n,c)$,
\begin{equation*}\label{eq:Step1}
 |\Ebb_\xi[h(\chunk[0]{X}[n-1])]-\Ebb_{\xi'}[h(\chunk[0]{X}[n-1])]|\leqslant 2\sum_{i=0}^{n-1}c_i \tvdist{\xi P^i}{\xi'P^i}\eqsp.
\end{equation*}
\end{lemma}
\begin{remark}
 It is possible to avoid the factor $2$ in \eqref{eq:Step1} under additional technical conditions, for example, when there exists a maximal coupling for $(\Pbb_\xi,\Pbb_{\xi'})$, see \cite[Lemma 23.2.1]{douc:moulines:priouret:soulier:2018}.
\end{remark}
\begin{proof}
Fix an arbitrary
$x^* \in \Xset$. For $i \in \{1,\ldots,n-1\}$, we set $\bar{h}_i(\chunk[i]{x}[n-1])=h(x^*,\dots,x^*,\chunk[i]{x}[n-1])$. By convention,
we set $\bar{h}_n$ the constant function $\bar{h}_n=h(x^*,\ldots,x^*)$ and $\bar{h}_0=h$. With these
notations, we have the decomposition
\[
h(\chunk{x})=\sum_{i=0}^{n-1} \{\bar{h}_i(\chunk[i]{x}[n-1]) - \bar{h}_{i+1}(\chunk[i+1]{x}[n-1]) \} +\bar{h}_n \eqsp.
\]
For all $i \in \{0,\ldots,n-1\}$ and all $x_i \in \Xset$, let
\begin{align}
\bar{w}_i(x_i)&=\int \lrcb{\bar{h}_i(\chunk[i]{x}[n-1])-\bar{h}_{i+1}(\chunk[i+1]{x}[n-1])}  \prod_{\ell=i+1}^{n-1} P(x_{\ell-1}, \rmd x_{\ell})\eqsp, \nonumber \\
&=\int \lrcb{h(x^*,\ldots,x^*,\chunk[i]{x})-h(x^*,\ldots,x^*,\chunk[i+1]{x})}  \prod_{\ell=i+1}^{n-1} P(x_{\ell-1}, \rmd x_{\ell}) \eqsp. \label{eq:delta:diarmid:fond}
\end{align}
It is easily seen that $\CPE{\{\bar{h}_i(\chunk[i]{X}[n-1])-\bar{h}_{i+1}(\chunk[i+1]{X}[n-1])\}}{\cF_{i}}=\bar{w}_i(X_i)$, $\Pbb_\xi-\as$,  which implies that
\[
\Ebb_\xi \lrb{h(\chunk{X})}=\sum_{i=0}^{n-1} \xi P^i \bar{w}_i + \bar{h}_n \eqsp.
\]
Since  $h \in\functionboundeddiff{\Xset^n}{c}$, \eqref{eq:delta:diarmid:fond} shows that  $\supnorm{\bar{w}_i} \leq c_i$.
Therefore,
\[
|\Ebb_\xi \lrb{h(X^{n-1})}-\Ebb_{\xi'} \lrb{h(X^{n-1})} |\\
\leq \sum_{i=0}^{n-1} |\xi P^i \bar{w}_i-\xi' P^i \bar{w}_i| \leq 2 \sum_{i=0}^{n-1} c_{i}\tvdist{\xi P^{i}}{\xi' P^{i}} \eqsp.
\]
\end{proof}

\section{Main result}\label{sec:Main}
The main result of this paper  is the following quantitative version of Mac Diarmid's inequality for geometrically ergodic Markov chains.
\begin{theorem}\label{thm:ConcMarkQuant}
 Assume {\bf H1}, {\bf H2}, {\bf H3}. Let $n\geqslant 1$, $c\in \R^n$ and  $f\in \functionboundeddiff{\sX^n}{c}$. Then, for all $x\in \cC$ and $t > 0$,
 \[
\Pbb_x\big(f( \chunk[0]{X}[n-1])-\Ebb_x[f(\chunk[0]{X}[n-1])]>t\big)\leqslant \exp\bigg(-\frac{\beta t^2}{\|c\|^2}\bigg)\eqsp,
 \]
where $\beta$ is given by
 \[
 \beta=\frac{(1-r\vee u^{-1/4})^2}{16L}\bigg(\frac5{\log u}+4ML\bigg)^{-1}\eqsp.
 \]
\end{theorem}
\begin{proof}[Proof of \Cref{thm:ConcMarkQuant}]
Fix $c\in \Rbb^n$, $x\in\sX$ and $f\in \functionboundeddiff{\sX^n}{c}$.
Following \cite{MR3407208}, we decompose $f(\chunk[0]{X}[n-1])-\Ebb_x[f(\chunk[0]{X}[n-1])]$ into martingale increments by conditioning to the stopping times $\tau_{\cC}^i$, $i=0,\ldots, n-1$.
For any integer $i\in [0,n-1]$, define
\[
G_i=\Ebb_x\big[f(\chunk[0]{X}[n-1])|\cF_{\tau_{\cC}^i}\big]\eqsp.
\]
As $\tau_{\cC}^0=0$ $\Pbb_x$-a.s., it holds $\Ebb_x[f(\chunk[0]{X}[n-1])]=\Ebb_x[f(\chunk[0]{X}[n-1])|\cF_{\tau_{\cC}^0}]=G_0$.
Moreover, as $\tau_{\cC}^{n-1}\geqslant n-1$, it also holds $G_{n-1}=\Ebb_x[f(\chunk[0]{X}[n-1])|\cF_{\tau_{\cC}^{n-1}}]=f(\chunk[0]{X}[n-1])$.
Therefore, the difference $f(\chunk[0]{X}[n-1])-\Ebb_x[f(\chunk[0]{X}[n-1])]$ is decomposed into a sum of the martingale increments $G_{i+1}-G_i$ as follows
\begin{equation}\label{eq:LinkWithG}
 f(\chunk[0]{X}[n-1])-\Ebb_x[f(\chunk[0]{X}[n-1])]=G_{n-1}-G_0=\sum_{i=0}^{n-2}(G_{i+1}-G_i)\eqsp.
\end{equation}
The proof is now decomposed into three facts that aim at bounding the Laplace transform of $f(\chunk[0]{X}[n-1])-\Ebb_x[f(\chunk[0]{X}[n-1])]$.

\noindent
{\bf Fact 1.} \emph{For any $i\in \{1,\ldots,n-1\}$,}
\begin{equation}\label{eq:Fact1}
 G_i-G_{i-1}=(G_i-G_{i-1})\indiacc{\tau_{\cC}^{i-1}=i-1}\eqsp.
\end{equation}
\begin{proof}[Proof of Fact 1.] By definition $\tau_{\cC}^{i-1}\geqslant i-1$ and $\tau_{\cC}^{i-1}> i-1$ if and only if $\tau_{\cC}^{i-1}=\tau_{\cC}^{i}$.
 Therefore,
 \[
 G_i-G_{i-1}=(G_i-G_{i-1})\left(\indiacc{\tau_{\cC}^{i-1}=i-1}+\indiacc{\tau_{\cC}^{i-1}=\tau_{\cC}^{i}}\right)\eqsp.
 \]
To prove that $(G_i-G_{i-1})\indiacc{\tau_{\cC}^{i-1}=\tau_{\cC}^{i}}=0$, we decompose according to the values of $\tau_{\cC}^{i}$:
 \[
(G_i-G_{i-1})\indiacc{\tau_{\cC}^{i-1}=\tau_{\cC}^{i}}=\sum_{j\geqslant i} (G_i-G_{i-1})\indiacc{\tau_{\cC}^{i-1}=\tau_{\cC}^{i}=j}\eqsp.
 \]
Now, remark that, for any $i\geqslant 0$,
\begin{align}
 G_i\indiacc{\tau_{\cC}^{i}=j}=
\begin{cases}\label{eq:Gitaui=j}
 \Ebb_x\big[f(\chunk[0]{X}[n-1])|\cF_j\big]&\text{ if } j\leqslant n-2\eqsp,\\
 f(\chunk[0]{X}[n-1])&\text{ if } j\geqslant n-1\eqsp.
\end{cases}
\end{align}
Then, for any $j\geqslant i$,
\[
 G_i\indiacc{\tau_{\cC}^{i}=j}\indiacc{\tau_{\cC}^{i-1}=\tau_{\cC}^{i}}= G_{i-1}\indiacc{\tau_{\cC}^{i-1}=j}\indiacc{\tau_{\cC}^{i-1}=\tau_{\cC}^{i}}=G_{i-1}\indiacc{\tau_{\cC}^{i}=j}\indiacc{\tau_{\cC}^{i-1}=\tau_{\cC}^{i}}\eqsp.
\]
This proves Fact 1.
\end{proof}

Fact 2. bounds the increments $G_i-G_{i-1}$. The proof relies on the following lemma which is a consequence of the coupling result Lemma~\ref{lem:DMPS18}.
Define $g_{n-1}=g_{n-1,\pi}=f$ and, for any $i\in [0,n-2]$, let $g_i$ and $g_{i,\pi}$ denote the functions defined for any $\chunk[0]{x}[i]\in\sX^{i+1}$ by 
\begin{equation}
\label{eq:definition-g-i}
g_i(\chunk[0]{x}[i])=\Ebb_{x_i}[f(\chunk[0]{x}[i],\chunk[1]{X}[n-1-i])],\qquad g_{i,\pi}(\chunk[0]{x}[i])=\Ebb_{\pi}[f(\chunk[0]{x}[i],\chunk[1]{X}[n-1-i])]\eqsp.
\end{equation}

\begin{lemma}
\label{lem:gi-gipi}
 Assume {\bf H1}, {\bf H2}, {\bf H3}. 
 For any $i\in \{0,\ldots,n-1\}$ and $(\chunk[0]{x}[i-1],x_i)$ in $\sX^i\times\cC$,
\begin{equation}\label{eq:gi-gipi}
|g_i(\chunk[0]{x}[i])-g_{i,\pi}(\chunk[0]{x}[i])|\leqslant 2L\sum_{j=i+1}^{n-1}c_jr^{j-i}\eqsp.
\end{equation}
\end{lemma}
\begin{proof}
 Fix $i\in\{0,\ldots,n-1\}$ and $\chunk[0]{x}[i]\in \sX^{i+1}$.
As $f\in\functionboundeddiff{\sX^n}{c}$, the function $\tilde{f}_i:\chunk[1]{y}[n-1-i]\in \sX^{n-1-i}\mapsto f(\chunk[0]{x}[i],\chunk[1]{y}[n-1-i])\in \R$ satisfies
\[
|\tilde{f}_i(\chunk[1]{y}[n-1-i])-\tilde{f}_i(\chunk[1]{z}[n-1-i])|\leqslant \sum_{k=1}^{n-1-i}c_{i+k}\indiacc{y_k\ne z_k}\eqsp.
\]
Hence, $\tilde{f}_i\in\bB\Dbb(\sX^{n-1-i},c_{i+1:n-1})$. Applying Lemma~\ref{lem:DMPS18} to the function $h=\tilde{f}_i$ yields
\begin{align*}
 |g_i(\chunk[0]{x}[i])-g_{i,\pi}(\chunk[0]{x}[i])|&=|\Ebb_{x_i}[f(\chunk[0]{x}[i],\chunk[1]{X}[n-1-i])]-\Ebb_{\pi}[f(\chunk[0]{x}[i],\chunk[1]{X}[n-1-i])]|\\
 &=|\Ebb_{x_i}[\tilde{f}_i(\chunk[1]{X}[n-1-i])]-\Ebb_{\pi}[\tilde{f}_i(\chunk[1]{X}[n-1-i])]| \leqslant 2\sum_{j=i+1}^{n-1}c_j \tvdist{\delta_{x_i}P^j}{\pi}\eqsp.
\end{align*}
Inequality \eqref{eq:gi-gipi} follows from {\bf H3}.
\end{proof}

\noindent
{\bf Fact 2.} \emph{Let $\rho$ such that $r\leqslant \rho<1$ and $i\in\{1,\ldots,n-1\}$.
Then,}
\begin{align}
\label{eq:Gi-Gi-1} |G_i-G_{i-1}|&\leqslant C_1\|c\|_{\infty}\indiacc{\tau_{\cC}^{i-1}=i-1}\sigma_{\cC}\circ\theta^{i-1}\eqsp,\\
\label{eq:Gi-Gi-1Sq}  |G_i-G_{i-1}|^2&\leqslant C_2\indiacc{\tau_{\cC}^{i-1}=i-1}\frac1{\rho^{2\sigma_{\cC}\circ\theta^{i-1}}}\sum_{k=i}^{n-1}c_k^2\rho^{k-i}\eqsp.
\end{align}
\emph{where, $C_1=5L/(1-r)$ and $C_2=16L^2/(1-\rho)$.}
\begin{proof}[Proof of Fact 2.]
For any integer $i\in\{1,\ldots,n\}$, let
\[
G_{i,1}=\Ebb_x[f(\chunk[0]{X}[n-1])|\cF_{\tau_{\cC}^{i-1}}]\indiacc{\tau_{\cC}^{i-1}=i-1},\qquad G_{i,2}=\Ebb_x[f(\chunk[0]{X}[n-1])|\cF_{\tau_{\cC}^{i}}]\indiacc{\tau_{\cC}^{i-1}=i-1}\eqsp.
\]
From Fact 1., $G_i-G_{i-1}=G_{i,2}-G_{i,1}$.
By Markov's property, for any $i\in \{0,\ldots,n-1\}$ and  $x\in \sX$,
\[
\Ebb_x[f(\chunk[0]{X}[n-1])|\cF_i]=g_i(X_{0:i}),\qquad \Pbb_{x}-\text{a.s.}\eqsp.
\]
Now, let $R_{i,1}=g_{i-1}(\chunk[0]{X}[i-1])\indiacc{\tau_{\cC}^{i-1}=i-1}-g_{i-1,\pi}(\chunk[0]{X}[i-1])\indiacc{\tau_{\cC}^{i-1}=i-1}$.
We have
\begin{align}
\notag G_{i,1}&=\Ebb_x[f(\chunk[0]{X}[n-1])|\cF_{\tau_{\cC}^{i-1}}]\indiacc{\tau_{\cC}^{i-1}=i-1}=\Ebb_x[f(\chunk[0]{X}[n-1])|\cF_{i-1}]\indiacc{\tau_{\cC}^{i-1}=i-1}\\
\label{eq:Gi1} &=g_{i-1}(\chunk[0]{X}[i-1])\indiacc{\tau_{\cC}^{i-1}=i-1}=g_{i-1,\pi}(\chunk[0]{X}[i-1])\indiacc{\tau_{\cC}^{i-1}=i-1}+R_{i,1}\eqsp.
\end{align}
Moreover, as $\tau_{\cC}^i\geqslant i$, by \eqref{eq:Gitaui=j},
\begin{align}
\notag G_{i,2}&=\sum_{j\geqslant i}\Ebb_x[f(\chunk[0]{X}[n-1])|\cF_{\tau_{\cC}^{i}}]\indiacc{\tau_{\cC}^{i-1}=i-1}\indiacc{\tau_{\cC}^i=j}\\
%\notag &=\sum_{j=i}^{n-2}\Ebb_x[f(\chunk[0]{X}[n-1])|\cF_{j}]\indiacc{\tau_{\cC}^{i-1}=i-1}\indiacc{\tau_{\cC}^i=j}+f(\chunk[0]{X}[n-1])\indiacc{\tau_{\cC}^{i-1}=i-1}\indiacc{\tau_{\cC}^i\geqslant n-1}\\
\label{eq:Gi2} &=\sum_{j=i}^{n-2}g_j(\chunk[0]{X}[j])\indiacc{\tau_{\cC}^{i-1}=i-1,\tau_{\cC}^i=j}+f(\chunk[0]{X}[n-1])\indiacc{\tau_{\cC}^{i-1}=i-1,\tau_{\cC}^i\geqslant n-1}\eqsp.
\end{align}
Let $R_{i,2}=\sum_{j=i}^{n-2}(g_j(\chunk[0]{X}[j])-g_{j,\pi}(\chunk[0]{X}[j]))\indiacc{\tau_{\cC}^{i-1}=i-1,\tau_{\cC}^i=j}$.
From \eqref{eq:Gi1} and \eqref{eq:Gi2},
\begin{align}
\label{eq:Gi1-Gi2} |G_{i,2}-G_{i,1}|=|&R_{i,2}-R_{i,1}+\sum_{j=i}^{n-2}(g_{j,\pi}(\chunk[0]{X}[j])-g_{i-1,\pi}(\chunk[0]{X}[i-1]))\indiacc{\tau_{\cC}^{i-1}=i-1,\tau_{\cC}^i=j}\\
\notag &+(f(\chunk[0]{X}[n-1])-g_{i-1,\pi}(\chunk[0]{X}[i-1]))\indiacc{\tau_{\cC}^{i-1}=i-1,\tau_{\cC}^i\geqslant n-1}|
% \\
% \leqslant 4&\indiacc{\tau_{\cC}^{i-1}=i-1}L\sum_{j=i+1}^{n-1}c_jr^{j-i}+\sum_{j=i}^{n-2}|(g_{j,\pi}(\chunk[0]{X}[j])-g_{i-1,\pi}(\chunk[0]{X}[i-1]))\indiacc{\tau_{\cC}^{i-1}=i-1,\tau_{\cC}^i=j}|\\
% &+|(f(\chunk[0]{X}[n-1])-g_{i-1,\pi}(\chunk[0]{X}[i-1]))\indiacc{\tau_{\cC}^{i-1}=i-1,\tau_{\cC}^i\geqslant n-1}|
\eqsp.
\end{align}
We bound separately all the terms in this decomposition.
First, as $\pi$ is invariant and $f\in \functionboundeddiff{\sX^n}{c}$, for any $j\in \{i+1,\ldots,n-1\}$ and any $\chunk[0]{x}[j]\in \sX^{j+1}$,
\[
|g_{j,\pi}(\chunk[0]{x}[j])-g_{i-1,\pi}(\chunk[0]{x}[i-1])|=\Ebb_{\pi}[f(\chunk[0]{x}[j],\chunk[j+1]{X}[n-1])-f(\chunk[0]{x}[i-1],\chunk[i]{X}[n-1])]\leqslant \sum_{k=i}^jc_k\eqsp.
\]
Hence,
\begin{equation}
\label{eq:Bound:Easyterms}
\begin{aligned} \sum_{j=i}^{n-2}|(g_{j,\pi}(\chunk[0]{X}[j])-g_{i-1,\pi}(\chunk[0]{X}[i-1]))|\indiacc{\tau_{\cC}^i=j}&\leqslant \sum_{j=i}^{n-2}\indiacc{\tau_{\cC}^i=j}\sum_{k=i}^jc_k=\indiacc{\tau_{\cC}^i\leqslant n-2}\sum_{k=i}^{\tau_{\cC}^i}c_k\eqsp,\\
|f(\chunk[0]{X}[n-1])-g_{i-1,\pi}(\chunk[0]{X}[i-1])|\indiacc{\tau_{\cC}^i\geqslant n-1}&\leqslant \indiacc{\tau_{\cC}^i\geqslant n-1}\sum_{k=i}^{n-1}c_k\eqsp.
\end{aligned}
\end{equation}

 %where $|R_{i,1}|$ and $|R_{i,2}|$ are respectively bounded in \eqref{eq:Bound:R1} and \eqref{eq:Bound:R2}.
%\anici{Selon moi, ta majoration de $|R_{i,1}|$ et $|R_{i,2}|$ par le même terme n'était pas juste : j'ai commenté dans le fichier source (au dessus de cette anotation) et propose donc}
%\begin{align*}
% |G_{i,2}-G_{i,1}|=|&R_{i,2}-R_{i,1}+\sum_{j=i}^{n-2}(g_{j,\pi}(\chunk[0]{X}[j])-g_{i-1,\pi}(\chunk[0]{X}[i-1]))\indiacc{\tau_{\cC}^{i-1}=i-1,\tau_{\cC}^i=j}\\
% &+(f(\chunk[0]{X}[n-1])-g_{i-1,\pi}(\chunk[0]{X}[i-1]))\indiacc{\tau_{\cC}^{i-1}=i-1,\tau_{\cC}^i\geqslant n-1}|\\
% \leqslant  |& R_{i,1}| + |R_{i,2}|+\sum_{j=i}^{n-2}|(g_{j,\pi}(\chunk[0]{X}[j])-g_{i-1,\pi}(\chunk[0]{X}[i-1]))\indiacc{\tau_{\cC}^{i-1}=i-1,\tau_{\cC}^i=j}|\\
% &+|(f(\chunk[0]{X}[n-1])-g_{i-1,\pi}(\chunk[0]{X}[i-1]))\indiacc{\tau_{\cC}^{i-1}=i-1,\tau_{\cC}^i\geqslant n-1}|\\
% \leqslant  2&\indiacc{\tau_{\cC}^{i-1}=i-1}L \left( \sum_{j=i+1}^{n-1}c_jr^{j-i} + \sum_{k=\tau_{\cC}^i+1}^{n-1}c_kr^{k-\tau_{\cC}^i} \right)\\
% & +  \sum_{j=i}^{n-2}|(g_{j,\pi}(\chunk[0]{X}[j])-g_{i-1,\pi}(\chunk[0]{X}[i-1]))\indiacc{\tau_{\cC}^{i-1}=i-1,\tau_{\cC}^i=j}|\\
% &+|(f(\chunk[0]{X}[n-1])-g_{i-1,\pi}(\chunk[0]{X}[i-1]))\indiacc{\tau_{\cC}^{i-1}=i-1,\tau_{\cC}^i\geqslant n-1}|\eqsp.
%\end{align*}
To bound $|R_{i,1}|$ and $|R_{i,2}|$ in \eqref{eq:Gi1-Gi2}, we use Lemma~\ref{lem:gi-gipi}.
First, \eqref{eq:gi-gipi} directly yields
\begin{equation}\label{eq:Bound:R1}
|R_{i,1}|\leqslant 2\indiacc{\tau_{\cC}^{i-1}=i-1}L\sum_{j=i+1}^{n-1}c_jr^{j-i}\eqsp.
\end{equation}
Moreover, as $\{\tau_{\cC}^i=j\}\subset \{X_j\in \cC\}$, \eqref{eq:gi-gipi} also yields
\[
(g_j(\chunk[0]{X}[j])-g_{j,\pi}(\chunk[0]{X}[j]))\indiacc{\tau_{\cC}^i=j}\leqslant 2L\sum_{k=j+1}^{n-1}c_kr^{k-j}\indiacc{\tau_{\cC}^i=j}\leqslant 2L\indiacc{\tau_{\cC}^i=j}\sum_{k=\tau_{\cC}^i+1}^{n-1}c_kr^{k-\tau_{\cC}^i}\eqsp.
\]
Therefore,
\begin{equation}\label{eq:Bound:R2}
|R_{i,2}|\leqslant 2L\indiacc{\tau_{\cC}^{i-1}=i-1}\sum_{k=\tau_{\cC}^i+1}^{n-1}c_kr^{k-\tau_{\cC}^i}\eqsp.
\end{equation}
Plugging \eqref{eq:Bound:Easyterms}, \eqref{eq:Bound:R1} and \eqref{eq:Bound:R2} in \eqref{eq:Gi1-Gi2} yields
\begin{align}
\label{eq:Gi1-Gi22} |G_{i,2}-G_{i,1}|\leqslant &2L\bigg(\sum_{j=i+1}^{n-1}c_jr^{j-i}+\sum_{k=\tau_{\cC}^i+1}^{n-1}c_kr^{k-\tau_{\cC}^i}+\frac1{2L}\sum_{k=i}^{\tau_{\cC}^i\wedge (n-1)}c_k\bigg)\indiacc{\tau_{\cC}^{i-1}=i-1}\eqsp.
\end{align}
Both \eqref{eq:Gi-Gi-1} and \eqref{eq:Gi-Gi-1Sq} follow from \eqref{eq:Gi1-Gi22} by bounding separately the $3$ terms in the right-hand side of this inequality.
Let us first establish \eqref{eq:Gi-Gi-1}.
Since $r<1$,
\[
\sum_{j=i+1}^{n-1}c_jr^{j-i}\leqslant \frac{\|c\|_\infty r}{1-r},\qquad \sum_{k=\tau_{\cC}^i+1}^{n-1}c_kr^{k-\tau_{\cC}^i}\leqslant \frac{\|c\|_\infty r}{1-r}\eqsp.
\]
%\anici{Dans toutes les équations suivantes, je pense qu'il y avait une erreur dans l'indice du $\theta$ car par définition $\tau_{\cC}^i = i + \tau_{\cC}^0 \circ \theta^i$ et $\sigma_{\cC}=1+\tau_{\cC}^0 \circ \theta$ d'où $1-i+\tau_{\cC}^i\wedge (n-1) \leqslant 1-i+\tau_{\cC}^i = 1 - i + i +\tau_{\cC}^0 \circ\theta \circ \theta^{i-1}=\sigma_{\cC}\circ\theta^{i-1}$. Cela est par ailleurs cohérent avec \eqref{eq:Gi-Gi-1} que l'on souhaite montrer. J'ai donc modifié en conséquence.}
Moreover,
\[
\sum_{k=i}^{\tau_{\cC}^i\wedge (n-1)}c_k\leqslant \|c\|_{\infty}[1-i+\tau_{\cC}^i\wedge (n-1) ]\leqslant \|c\|_{\infty}[1+\tau_{\cC}^0\circ \theta^i]=\|c\|_{\infty}\sigma_{\cC}\circ\theta^{i-1}\eqsp.
\]
As $r<1\leqslant \sigma_{\cC}\circ\theta^{i-1}$, plugging these upper bounds in \eqref{eq:Gi1-Gi22} shows
\[
|G_{i}-G_{i-1}|=|G_{i,2}-G_{i,1}|\leqslant \frac{5L\|c\|_{\infty}}{1-r}\sigma_{\cC}\circ\theta^{i-1}\indiacc{\tau_{\cC}^{i-1}=i-1}\eqsp.
\]
This proves \eqref{eq:Gi-Gi-1}.
We use slightly different controls to prove \eqref{eq:Gi-Gi-1Sq} from \eqref{eq:Gi1-Gi22}.
As $r\leqslant \rho<1$, $\rho^{-\sigma_{\cC}\circ\theta^{i-1}}\geqslant 1$, and
\begin{align}
\label{eq:Step1Var} \sum_{j=i+1}^{n-1}c_jr^{j-i}\leqslant \sum_{j=i}^{n-1}c_j\rho^{j-i}\leqslant \rho^{-\sigma_{\cC}\circ\theta^{i-1}}\sum_{j=i}^{n-1}c_j\rho^{j-i}\eqsp.
\end{align}
%\anici{Ci après, je pense qu'il y avait également la même erreur d'indice pour $\theta$ et j'ai modifié en conséquence.}
Moreover,
\begin{align*}
 \sum_{k=\tau_{\cC}^i+1}^{n-1}c_kr^{k-\tau_{\cC}^i}\leqslant\rho^{i-\tau_{\cC}^i}\sum_{k=\tau_{\cC}^i+1}^{n-1}c_k\rho^{k-i}\eqsp.
\end{align*}
As $\tau_{\cC}^i\geqslant i$ and $i-\tau_{\cC}^i=1-\sigma_{\cC}\circ\theta^{i-1}$,
\begin{equation}\label{eq:Step2Var}
\sum_{k=\tau_{\cC}^i+1}^{n-1}c_kr^{k-\tau_{\cC}^i}\leqslant \rho^{1-\sigma_{\cC}\circ\theta^{i-1}} \sum_{j=\tau^i_{\cC}+1}^{n-1}c_j\rho^{j-i}\leqslant \rho^{-\sigma_{\cC}\circ\theta^{i-1}}\sum_{j=\tau^i_{\cC}+1}^{n-1}c_j\rho^{j-i}\eqsp.
\end{equation}
In addition,
\begin{align}\label{eq:Step3Var}
 \sum_{k=i}^{\tau_{\cC}^i\wedge (n-1)}c_k\leqslant &\sum_{k=i}^{\tau_{\cC}^i\wedge (n-1)}c_k\rho^{k-\tau_{\cC}^i}=\sum_{k=i}^{\tau_{\cC}^i\wedge (n-1)}c_k\rho^{k-i-\sigma_{\cC}\circ\theta^{i-1}+1}\leqslant \rho^{-\sigma_{\cC}\circ\theta^{i-1}}\sum_{k=i}^{\tau_{\cC}^i\wedge (n-1)}c_k\rho^{k-i}\eqsp.
\end{align}
%\anici{Je ne sais pas si c'est important quantitativement mais dans les deux majorations précédentes, on se débarasses du $\tau_{\cC}^i$ assez "brutalement" alors que si on fait comme dans mon manuscrit de thèse, on parvient dans l'inégalité suivante non pas à $25L^2$ mais $16 L^2$ (en supposant bien évidemment comme ici que $L\geq 1$).}
Plugging \eqref{eq:Step1Var}, \eqref{eq:Step2Var} and \eqref{eq:Step3Var} in \eqref{eq:Gi1-Gi22} and applying Cauchy-Schwarz inequality shows
\begin{align*}
 |G_{i}-G_{i-1}|^2=&|G_{i,2}-G_{i,1}|^2\leqslant16L^2\rho^{-2\sigma_{\cC}\circ\theta^{i-1}}\bigg(\sum_{k=i}^{n-1}c_k\rho^{k-i}\bigg)^2\indiacc{\tau_{\cC}^{i-1}=i-1}\\
 \leqslant &\frac{16L^2}{1-\rho}\rho^{-2\sigma_{\cC}\circ\theta^{i-1}}\sum_{k=i}^{n-1}c_k^2\rho^{k-i}\indiacc{\tau_{\cC}^{i-1}=i-1}\eqsp.
\end{align*}
This proves \eqref{eq:Gi-Gi-1Sq} and thus {\bf Fact 2.}
\end{proof}

\noindent
{\bf Fact 3.} \emph{ Assume {\bf H1}, {\bf H2}, {\bf H3}. For any $x\in \cC$,}
\begin{align}
\Ebb_x\bigg[\rme^{f(\chunk[0]{X}[n-1])-\Ebb_x[f(\chunk[0]{X}[n-1])]}\bigg]\leqslant \rme^{C_3\|c\|^2}\eqsp.
\end{align}
\emph{where $C_3=\fracaa{4L\left(\fracaa{5}{\log u}+4ML\right)}{(1-r\vee u^{-1/4})^2}$.}
\begin{proof}[Proof of Fact 3.]
For any $t\in \Rbb$, $\rme^t\leqslant 1+t+t^2\rme^{|t|}$. Hence, as $\Ebb_{x}[G_{i+1}-G_i|\cF_{\tau_{\cC}^{i}}]=0$, for any $i\geqslant 0$, we have
\begin{align*}
 \Ebb_x[\rme^{G_{i+1}-G_i}|\cF_{\tau_{\cC}^{i}}]\leqslant 1+\Ebb_x[(G_{i+1}-G_i)^2\rme^{|G_{i+1}-G_i|}|\cF_{\tau_{\cC}^{i}}]\eqsp.
\end{align*}
By Fact 2.,
\begin{align*}
 \Ebb_x[\rme^{G_{i+1}-G_i}|\cF_{\tau_{\cC}^{i}}]&\leqslant 1+C_2\sum_{k=i+1}^{n-1}c_k^2\rho^{k-i-1}\indiacc{\tau_{\cC}^{i}=i}\Ebb_x[\rho^{-2\sigma_{\cC}\circ\theta^{i}}\rme^{C_1\|c\|_{\infty}\sigma_{\cC}\circ\theta^{i}}|\cF_{\tau_{\cC}^{i}}]\eqsp.
 \end{align*}
 Now by Markov's property,
 %
\begin{align*}
\indiacc{\tau_{\cC}^{i}=i} \Ebb_x[\rho^{-2\sigma_{\cC}\circ\theta^{i}}\rme^{C_1\|c\|_{\infty}\sigma_{\cC}\circ\theta^{i}}|\cF_{\tau_{\cC}^{i}}]&=\indiacc{\tau_{\cC}^{i}=i} \Ebb_x[\rho^{-2\sigma_{\cC}\circ\theta^{i}}\rme^{C_1\|c\|_{\infty}\sigma_{\cC}\circ\theta^{i}}|\cF_{i}]\\
&=\indiacc{\tau_{\cC}^{i}=i}\Ebb_{X_i}[\rho^{-2\sigma_{\cC}}\rme^{C_1\|c\|_{\infty}\sigma_{\cC}}]\eqsp.
\end{align*}
Hence,
 \begin{align*}
 \Ebb_x[\rme^{G_{i+1}-G_i}|\cF_{\tau_{\cC}^{i}}]
 &=1+C_2\sum_{k=i+1}^{n-1}c_k^2\rho^{k-i-1}\indiacc{\tau_{\cC}^{i}=i}\Ebb_{X_i}[\rho^{-2\sigma_{\cC}}\rme^{C_1\|c\|_{\infty}\sigma_{\cC}}]\eqsp.
 \end{align*}
Let $\rho=r\vee u^{-1/4}$, $\varepsilon=\log u/(2C_1)$ and assume first that $\|c\|_{\infty}\leqslant \varepsilon$. By {\bf H$2$},
\[
\indiacc{\tau_{\cC}^{i}=i}\Ebb_{X_i}[\rho^{-2\sigma_{\cC}}\rme^{C_1\|c\|_{\infty}\sigma_{\cC}}]\leqslant \indiacc{\tau_{\cC}^{i}=i}\sup_{x\in\cC}\Ebb_x[\rho^{-2\sigma_{\cC}}\rme^{C_1\|c\|_{\infty}\sigma_{\cC}}]\leqslant\sup_{x\in\cC}\Ebb_x[u^{\sigma_{\cC}}]\leqslant M\enspace.
\]
Hence,
\[
 \Ebb_x[\rme^{G_{i+1}-G_i}|\cF_{\tau_{\cC}^{i}}]\leqslant 1+C_2M\sum_{k=i+1}^{n-1}c_k^2\rho^{k-i-1}\leqslant \rme^{C_2M\sum_{k=i+1}^{n-1}c_k^2\rho^{k-i-1}}\eqsp.
\]
By recurrence, it follows that
\begin{align*}
\Ebb_x\bigg[\rme^{f(\chunk[0]{X}[n-1])-\Ebb_x[f(\chunk[0]{X}[n-1])]}\bigg]&\leqslant \rme^{C_2M\sum_{i=0}^{n-2}\sum_{k=i+1}^{n-1}c_k^2\rho^{k-i-1}}\\
&=\rme^{C_2M\sum_{k=1}^{n-1}c_k^2\sum_{i=0}^{k-1}\rho^{k-i-1}}\leqslant \rme^{\frac{C_2M}{1-\rho}\|c\|^2} \eqsp.
\end{align*}
Fix $\tilde{x}$ in $\sX$ and let $\tilde{f} : \sX^n \rightarrow \Rbb$ be defined, for any $x_{0:n-1}$ in $\sX^n$, by
\[
\tilde{f}(\chunk[0]{X}[n-1])
=
f(x_0\mathbbm{1}_{\{c_{0}\leq\varepsilon\}} + \tilde{x} \mathbbm{1}_{\{c_{0}>\varepsilon\}},\dots, x_{n-1}\mathbbm{1}_{\{c_{n-1}\leq\varepsilon\}} + \tilde{x} \mathbbm{1}_{\{c_{n-1}>\varepsilon\}})
\eqsp .
\]
As $f$ belongs to $\mathbb{BD}\left(\sX^n,c\right)$, $\tilde{f}$ belongs to $ \mathbb{BD}\left(\sX^n,\tilde{c}\right)$, where
\[
\tilde{c}
=
\left( c_0\mathbbm{1}_{\{c_{0}\leq\varepsilon\}},\dots,c_{n-1}\mathbbm{1}_{\{c_{n-1}\leq\varepsilon\}}\right)
\eqsp .
\]
Since $\|\tilde{c}\|_{\infty} < \varepsilon$ and $\|\tilde{c}\|\leqslant \|c\|$, $\tilde{f}$ satisfies
\begin{align}\label{ineq:max-gamma-leq-varepsilon-ftilde}
\Ebb_{x}\left[\rme^{\tilde{f}(\chunk[0]{X}[n-1])-\Ebb_{x}[\tilde{f}(\chunk[0]{X}[n-1])]}\right]
& \leqslant
\rme^{\frac{MC_2}{1-\rho}
\| \tilde{c}\|^2} \leqslant
\rme^{\frac{MC_2}{1-\rho}
\| c\|^2}
\eqsp .
\end{align}
Furthermore, by definition of $\tilde{f}$ and since $f$ is in $\mathbb{BD}(\sX^n,c)$, for any $x\in\sX^n$,
\begin{align}\label{ineq:difference-f-tildef}
|f(x)-\tilde{f}(x)|
& =
%|f(x_0,\dots,x_{n-1})-f(x_0\mathbbm{1}_{\{\gamma_{0}\leq\varepsilon\}} + \tilde{x} \mathbbm{1}_{\{\gamma_{0}>\varepsilon\}},\dots, x_{n-1}\mathbbm{1}_{\{\gamma_{n-1}\leq\varepsilon\}} + \tilde{x} \mathbbm{1}_{\{\gamma_{n-1}>\varepsilon\}})|\\
%& \leq
\sum_{i=0}^{n-1} c_i \mathbbm{1}_{\{c_i>\varepsilon\}}
\leq
\sum_{i=0}^{n-1} c_i \frac{c_i}{\varepsilon}
\leq
\frac{\|c\|^2}{\varepsilon}.
\end{align}
This implies
\[
\Ebb_x\bigg[\rme^{f(\chunk[0]{X}[n-1])-\Ebb_x[f(\chunk[0]{X}[n-1])]}\bigg]\leqslant \rme^{\frac{2\|c\|^2}{\varepsilon}}\Ebb_{x}\left[\rme^{\tilde{f}(\chunk[0]{X}[n-1])-\Ebb_{x}[\tilde{f}(\chunk[0]{X}[n-1])]}\right]\leqslant \rme^{\bigg(\frac2{\varepsilon}+\frac{MC_2}{1-\rho}\bigg)
\| c\|^2}
\eqsp .
\]
This shows {\bf Fact 3} since
\[
\frac2{\varepsilon}+\frac{MC_2}{1-\rho}\leqslant \frac{4L}{(1-r\vee u^{-1/4})^2}\bigg(\frac5{\log u}+4ML\bigg)\eqsp.
\]
\end{proof}

\noindent
{\bf Fact 3} proves that there exists a constant $C=2C_3$ such that, for any $c\in \Rbb^n$, $f\in \functionboundeddiff{\sX^n}{c}$ and $x\in \cC$,
\begin{equation}\label{ControlLaplace}
 \Ebb_x\bigg[\rme^{f(\chunk[0]{X}[n-1])-\Ebb_x[f(\chunk[0]{X}[n-1])]}\bigg]\leqslant \rme^{C\| c\|^2/2}
\eqsp .
\end{equation}
Let $f\in \functionboundeddiff{\sX^n}{c}$ and $x\in \cC$. 
For any $s>0$, $sf\in \functionboundeddiff{\sX^n}{c}$.
Hence, from \eqref{ControlLaplace}, for any $s,t>0$,
\begin{align*}
 \Pbb\big(f(\chunk[0]{X}[n-1])-\Ebb_x[f(\chunk[0]{X}[n-1])]>t\big)&\leqslant \rme^{-st+\log\Ebb_x\big[\rme^{sf(\chunk[0]{X}[n-1])-\Ebb_x[sf(\chunk[0]{X}[n-1])]}\big]}\\
 &\leqslant \rme^{-st+s^2C\| c\|^2/2}\enspace.
\end{align*}
Choosing $s=t/(C\|c\|^2)$ proves Theorem~\ref{thm:ConcMarkQuant} with
\[
\beta=\frac1{2C}=\frac1{4C_3}=\frac{(1-r\vee u^{-1/4})^2}{16L}\bigg(\frac5{\log u}+4ML\bigg)^{-1}\eqsp.
\]
\end{proof}

\bibliographystyle{plain}
\bibliography{BiblioSource}

\end{document}

### 2019-12-12 aap/Revision/vr-mcmc-revision.tex
Revision

@@
 \begin{description}
-\item[(H1)][\textbf{Lipschitz continuity}] Assume that $\nabla U$ is Lipschitz, that is, there exists $L\geq 0$ such that 
-\begin{eqnarray*}
-\|\nabla U(x)-\nabla U(y)\|\leq L\|x-y\|, \quad x,y\in \mathbb{R}^d.
-\end{eqnarray*}
-
-\item[(H2)][\textbf{Convexity outside a ball}] There exist $K>0,$ $M>0$ and $m>0$ such that for any $x\not\in B_K(0)	$ and any $y\in \mathbb{R}^d$ it holds 
-\begin{eqnarray*}
-\langle D^2U(x)y,y\rangle \geq m\|y\|^2, \quad \|D^3 U(y)\| \leq M. 
+\item[(H1)][\textbf{Lipschitz continuity}] Assume that $\nabla U$ is Lipschitz, that is, there exists $L_U\geq 0$ such that 
+\begin{eqnarray*}
+|\nabla U(x)-\nabla U(y)|\leq L_U|x-y|, \quad x,y\in \mathbb{R}^d.
+\end{eqnarray*}
+\item[(H2)][\textbf{Convexity outside a ball}] There exist $K_U>0,$ $M_U>0$ and $m_U>0$ such that for any $x\not\in B_{K_U}(0)	$ and any $y\in \mathbb{R}^d$ it holds 
+\begin{eqnarray*}
+\langle D^2U(x)y,y\rangle \geq m_U\|y\|^2, \quad \|D^3 U(y)\| \leq M_U. 
 \end{eqnarray*}
@@
 for $\gamma>0$, the Markov kernel $R_\gamma$
-from $(\mathbb R^d,\mathcal B(\mathbb R^d))$
-into $(\mathbb R^d,\mathcal B(\mathbb R^d))$
 is defined by
@@
 $
-which, finally, provides the (random) measure
-$Q_\gamma^{N+1,p}(X_N,\cdot)$
+which, finally, provides the  measure
+$Q_\gamma^{p}(x_0,\cdot)=Q_\gamma^{p,1}$
 used in the right-hand side of~\eqref{eq:06042018a2}.
@@
 is provided in  in~\cite[Theorem~17]{durmus:moulines:2017}.
-As for the estimator~\eqref{eq:29032018a3}, we introduce lemma of conditional variance decomposition.
-It follows from~\eqref{eq:29032018a5}, Theorem~\ref{prop:29032018a1} and independence condition for the sequence $(Z_p)_{p \geq 1}$.
-\begin{lem}
+As for the estimator~\eqref{eq:29032018a3}, 
+it follows from~\eqref{eq:29032018a5}, Theorem~\ref{prop:29032018a1} and independence condition for the sequence $(Z_p)_{p \geq 1}$.
+\begin{prop}
 \label{cond_var_dec}
@@
 \begin{equation}\label{eq:29032018a6}
-\mathsf{Var}\left[\left.\pi_{K,n}^{N}(f)\right|\mathcal G_N\right]
+\mathsf{Var}\left[\pi_{K,n}^{N}(f)\right]
 %=\sum_{\min_i (k_i)\ge K+1}
@@
 \sum_{l=N+1}^{N+n}
-\mathsf{E}\left[\left.\left|\bar a_{l,\mathbf{k}}(X_{l-1})\right|^{2}
-\right|\mathcal G_N\right].
-\end{equation}
-\end{lem}
+\mathsf{E}\left[\left|\bar a_{l,\mathbf{k}}(X_{l-1})\right|^{2}\right].
+\end{equation}
+\end{prop}
 Now we present an upper bound for the right-hand sides of~\eqref{eq:29032018a6} in case of ULA algorithm.
 \begin{thm}\label{th:mr}
+Assume (H1) and (H2).
 Suppose that $f$  is two times continuously differentiable
@@
 \end{eqnarray*}
-for all \(\mathbf{k}\) with \(0<\|\mathbf{k}\|\leq  2.\) Moreover assume that there exist $K>0,$ $M>0$ and $m>0$ such that for any $x\not\in B_K(0)$ and any $y\in \mathbb{R}^d$ it holds
-\[
-\langle \nabla \mu(x)y,y\rangle \geq m\|y\|^2.
-\]
-Let $(\gamma_k)_{k\in\mathbb N}$ be a sequence of positive
+for all \(\mathbf{k}\) with \(0<\|\mathbf{k}\|\leq  2.\) Let $(\gamma_k)_{k\in\mathbb N}$ be a sequence of positive
 and non-increasing step sizes with $\sum_{k=1}^\infty \gamma_k=\infty$ and $\sum_{k=1}^\infty \gamma^2_k<\infty.$ 
@@
 \label{eq:var-bound}
-\mathsf{Var}\left[\left.\pi_{K,n}^{N}(f)\right|\mathcal G_N\right]\lesssim \frac{1}{\Gamma^2_{N+2,N+n+1}},
+\mathsf{Var}\left[\pi_{K,n}^{N}(f)\right]\lesssim \frac{1}{\Gamma^2_{N+2,N+n+1}},
 \end{eqnarray}
@@
 \left(\mathsf{E}\left[\sum_{p=l}^{N+n}\omega_{p,n}^{N}\left(\partial_{y_{l}}^{\mathbf r}f\left(X_{p}\right)-\mathsf{E}\left[\partial_{y_{l}}^{\mathbf r}f\left(X_{p}\right)|X_{l-1}\right]\right)\mathbf{H}_{\mathbf{q}}(Z_{l})\Big|X_{l-1}\right]\right)^{2}
-\Bigg|\mathcal G_N \right\}.
+\right\}.
 \end{multline*}
@@
 \mathsf{Var}\left(\sum_{p=l}^{N+n}\gamma_{p+1}\partial_{y_{l}}^{\mathbf{K}_I}f\left(X_{p}\right)
-\Big|X_{l-1}\right)
-\Bigg|\mathcal G_N\right].
+\Big|X_{l-1}\right)\right].
 \end{eqnarray*}
@@
 \label{eq:definition-alpha}
-\alpha_k=1-\gamma_k b_\mu,\quad k\in\mathbb N.
+\alpha_k=1-\gamma_k \mu'(X_{k-1}),\quad k\in\mathbb N.
 \end{equation}
@@
 \begin{remark}\label{rem:10062018a1}
-Notice that, under~\eqref{eq:17042018a1},
-not only~\eqref{eq:06062018a3} but also
-\begin{equation}\label{eq:10062018a1}
-\sum_{r=j}^\infty\gamma_r \prod_{k=j}^r \alpha_k\le\ol C_2
-\end{equation}
-is satisfied with the same constant~$\ol C_2$
-(which is $C$ of~\eqref{eq:17042018a1})
-simultaneously for all $j\in\mathbb N$.
+
 Below this will allow us to apply
@@
 $$
-with suitable $(\epsilon_{j,p})_{j\ge1,p\ge j+1}$
-and the constant $\ol C_2$ in~\eqref{eq:10062018a1}
-being independent of~$j$.
+with suitable $(\epsilon_{j,p})_{j\ge1,p\ge j+1}$.
 \end{remark}
-
+The assumption (H1) implies that  $|\mu'(x)|\leq B_\mu$ for some constant $B_\mu>0$ and all $x\in \mathbb{R}^d$. Without loss of generality we suppose that $\gamma_N B_\mu<1.$
 \begin{lem}\label{lem:06062018a2}
@@
 \label{eq:08062018b2}
-\left|\partial_{y_l}^r X_p \right| \le C_r\prod_{k=l+1}^p \alpha_k \,,
+\left|\partial_{y_l}^r X_p \right| \le C_r\prod_{k=l+1}^p (1-\gamma_k\mu'(X_{k-1}))\quad \text{a.s.} 
 \end{equation}
@@
 \end{lem}
-\begin{proof}
-The proof is along the same lines as Lemma~\ref{lem:06062018a3}.
-\end{proof}
 \begin{lem}\label{lem:06062018a3}
@@
 \left|\partial_{y_j} \partial_{y_l}^r X_p\right|
-\le c_r\prod_{k=l+1}^p \alpha_k
+\le c_r\prod_{k=l+1}^p (1-\gamma_k\mu'(X_{k-1})), \quad \text{a.s.}
 \end{equation}
@@
 \end{align*}
-with a suitable constant
-(we can take, e.g., $\const=\frac{B_\mu}{(1-\gamma_1 b_\mu)^2}$).
+with a suitable constant.
 By Lemma~\ref{lem:06062018a1}
@@
 \end{align*}
-with some constant ``$\const$'',
-which is, in fact, $\frac1{(1-\gamma_1 b_\mu)^2}$ times
-the expression involving
+with some constant ``$\const$'' depending on
 $B_\mu,r,C_1,\ldots,C_r,c_1,\ldots,c_{r-1}$.
@@
 A_{K}^{2}\sum_{j=l}^{q}\gamma_{j}
-\left(\sum_{p=j}^{q}\gamma_{p+1}\prod_{k=l+1}^{p}\alpha_{k}\right)^{2}
+\mathsf{E}\left(\sum_{p=j}^{q}\gamma_{p+1}\prod_{k=l+1}^{p}\alpha_{k}\right)^{2}
 \\
 &\le
-\frac{A_{K}^{2}}{(1-\gamma_1 b_\mu)^2}
+\frac{A_{K}^{2}}{(1-\gamma_1 B_\mu)^2}
 \sum_{j=l}^{q} \gamma_{j}
@@
 &\le
-\frac{A_{K}^{2}}{(1-\gamma_1 b_\mu)^3}
+\frac{A_{K}^{2}}{(1-\gamma_1 B_\mu)^3}
 \sum_{j=l}^{q} \gamma_{j}
@@
 &\le
-\frac{A_{K}^{2}}{(1-\gamma_1 b_\mu)^3}C^3=B_K,
+\frac{A_{K}^{2}}{(1-\gamma_1 B_\mu)^3}C^3=B_K,
 \end{align*}

### 2019-12-12 Note_exponential_moment/MarkovCon.tex
update

@@
 
+\section{Applications to Unadjusted Langevin Algorithm}\label{sec:ULA}
+We illustrate the applicability of the results by the example on one of the MCMC methods.
+Suppose that we aim at sampling from the distribution with density $\pi(x) = \frac{e^{-U(x)}}{\int\limits_{\mathbb{R}^d} e^{-U(y)}\,dy}$.
+One of the popular algorithms for solving this problem is the Unadjusted Langevin Algorithm, which suggests constructing the Markov Chain
+$$
+X_{k+1} = h_{k+1}(X_k) + \sqrt{2\gamma_{k+1}}Z_{k+1}, h_{k+1}(X_k) := X_{k} - \gamma_{k+1}\nabla U(X_{k}) 
+$$
+with some non-increasing sequence $\gamma_k \geq 0$, where $Z_k \sim \mathcal{N}(0,I)$ are i.i.d.
+standard normal $d$-dimensional random variables.
+Standard assumptions on the sequence $\{\gamma_k\}$ are $\sum\limits_{k=0}^{\infty}\gamma_k = \infty$ and $\sum\limits_{k=0}^{\infty}\gamma_k^2 < \infty$.
+Thus it is impossible to analyze properties of additive functionals of $\{X_k\}$ by standard tools, since the assumptions {\bf H2} and {\bf H3} fails in case $\gamma_k \rightarrow 0$.
+Yet we might establish some alternative bounds on the speed of convergence to stationary distribution.
+We need to impose condition
+\par
+[{\bf C1}] Potential $U(x)$ satisfies for all $x,y \in \mathbb{R}^d$ the condition
+$$
+\|U(x) - U(y)\| \leq L\|x-y\|
+$$
+This condition obviously implies
+$$
+\|h_{k}(X) - h_k(Y)\| \leq (1+L\gamma_k)\|X-Y\|
+$$
+Hence, by \cite[Theorem~19]{durmus:moulines:2018}, we might obtain the following minorization condition
+\begin{lemma} Assume that the condition {\bf C1} holds. Then for all $x,y \in \mathbb{R}^d, n\geq 1$ it holds
+$$
+\|\delta_xQ^n - \delta_yQ^n\|_{TV} \leq \mathbb{I}(x \neq y)\left(1 - 2\Phi\left(-\frac{\|x-y\|}{2\sqrt{b_n}}\right)\right)
+$$
+where $b_n := \sum\nolimits_{i=1}^{k}\frac{\gamma_i}{\prod_{j=1}^{i}(1+L\gamma_j)^2}$
+\end{lemma}
+Now we aim at establishing Foster-Lyapunov drift conditions for the kernel $P_{\gamma_n}$.We will say that function $V(x): \mathbb{R}^d \rightarrow [1,+\infty)$ satisfies a Foster-Lyapunov drift condition for the Markov kernel $P_\gamma$ if.
+there are such $\overline{\gamma} > 0, \lambda \in [0,1), C > 0$ such that for any $\gamma \in (0, \overline{\gamma}]$ it holds
+\begin{equation}
+\label{eq:drift}
+P_{\gamma}V \leq \lambda^{\gamma}V + \gamma C
+\end{equation}
+Given that this condition holds, we may prove drift condition for. the kernel $Q^n$.
+\begin{lemma} Assume that $P_{\gamma}$ satisfies condition \ref{eq:drift} for any $0 < \gamma < \overline{gamma}$. Then it holds
+$$
+Q^{n,p}V \leq \lambda^{\Gamma_{p,n}}V + C\frac{\lambda^{\Gamma_{1,p}} - \lambda^{\Gamma_{1,n}}}{\log{\frac{1}{\lambda}}}
+$$ 
+\end{lemma}
+\begin{proof} For notation simplicity let us consider only $p=1$. Then by induction
+$$
+Q^{n}V \leq \lambda^{\Gamma_{1,n}}V + C \sum\limits_{k=1}^{n}\gamma_k \lambda^{\Gamma_{1,k-1}} \leq \lambda^{\Gamma_{1,n}}V + C \int\limits_{0}^{\Gamma_{1,n}}\lambda^{x}\,dx \leq
+$$
+$$
+\leq \lambda^{\Gamma_{1,n}}V + C\frac{1-\lambda^{\Gamma_{1,n}}}{\log{\frac{1}{\lambda}}}
+$$
+\end{proof}
+
 \bibliographystyle{plain}

### 2019-12-13 Note_exponential_moment/MarkovCon.tex
update

@@
 standard normal $d$-dimensional random variables.
+Let us denote ULA kernel with parameter $\gamma$ as $P_{\gamma}$, that is,
+\begin{equation}
+\label{eq:ula_kernel}
+P_{\gamma}(x,dy) = \frac{1}{(4\pi\gamma)^{\frac{d}{2}}}\exp{\left(-\frac{\|y-x+\gamma\nabla U(x)\|^2}{4\gamma}\right)}\,dy
+\end{equation}
+Let us also introduce the multiple steps kernel
+$$
+Q^{p,n} = P_{\gamma_p}\ldots P_{\gamma_n}, Q^{n} := Q^{1,n}
+$$
 Standard assumptions on the sequence $\{\gamma_k\}$ are $\sum\limits_{k=0}^{\infty}\gamma_k = \infty$ and $\sum\limits_{k=0}^{\infty}\gamma_k^2 < \infty$.
@@
 Given that this condition holds, we may prove drift condition for. the kernel $Q^n$.
-\begin{lemma} Assume that $P_{\gamma}$ satisfies condition \ref{eq:drift} for any $0 < \gamma < \overline{gamma}$. Then it holds
+
+\begin{lemma}
+\label{lem:drift_prod}
+ Assume that $P_{\gamma}$ satisfies condition \ref{eq:drift} for any $0 < \gamma < \overline{gamma}$. Then it holds
 $$
@@
 \end{proof}
+Let us check that ULA kernel \ref{} satisfies the drift condition \label{eq:drift}. We will need additional assumption
+\par
+[{\bf C2}] There exists such constants $K_1 > 0$ and $m > 0$, such that for any $x \notin B(0,K_1)$ it holds
+$$
+\langle \nabla U(x), x \rangle \geq \frac{m}{2}\|x\|^2
+$$
+Another popular condition on the potential is strong convexity outside some euclidean ball:
+\par
+[{\bf C3}] There exist constants $K_1 > 0$ and $m > 0$ such that for any $x \notin B(0,K_1)$ and $y \in \mathbb{R}^d$, it holds $\langle D^2U(x)y,y\rangle \geq m\|y\|^2$.
+Moreover, there exists $M \geq 0$ such that for any $x \in \mathbb{R}^d$, $\|D^3U(x)\| \leq M$.
+\par
+Yet it is possible to show that assumption {\bf C3} and {\bf C1} imply {\bf C2} (see \cite[Lemma~14]{brosse:2018:diffusion}), so in the sequel we are going to work with the condition {\bf C2}.
+It allows us to prove the following result
+
+\begin{lemma} Assume that the potential $U(x)$ satisfies conditions {\bf C1} and {\bf C2}.
+Then the kernel \ref{} satisfies drift condition \ref{eq:drift} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L^2}$ with drift function $V(x) = 1 + \|x\|^2$, constants $\lambda = \exp{-\left(\frac{m}{2}\right)}$, $C = 3K_1^2 + 2d + C_1$ for some absolute constant $C_1$.
+\end{lemma}
+
+\begin{proof} Consider $V(x) = 1 + \|x\|^2$, then
+$$
+P_{\gamma}V(x) = \int\limits_{\mathbb{R}^d}V(y)P_{\gamma}(x,dy) = \int\limits_{\mathbb{R}^d}(1+\|y\|^2)\frac{1}{(4\pi\gamma)^{\frac{d}{2}}}\exp{\left(-\frac{\|y-x+\gamma\nabla U(x)\|^2}{4\gamma}\right)}\,dy =
+$$
+
+$$
+=1 + \frac{1}{(4\pi\gamma)^{\frac{d}{2}}}\int\limits_{\mathbb{R}^d}\|z + x - \gamma \nabla U(x)\|^2 \exp{\left(-\frac{\|z\|^2}{4\gamma}\right)}\,dz
+$$
+Let us first consider the case $x \notin B(0,K_1)$.
+Note that $\|z+x-\gamma \nabla U(x)\|^2 = \|z\|^2 + 2 \langle z, x - \gamma \nabla U(x) \rangle + \gamma^2 \|x - \gamma \nabla U(x)\|^2$ and the linear term vanishes after integration, moreover, for $Z \sim \mathcal{N}(0,2\gamma I_d$, it holds $\mathbb{E}\|Z\|^2 = 2\gamma d$.
+It remains to notice that due to {\bf C1} and {\bf C2}
+$$
+\|x - \gamma \nabla U(x) \|^2 = \|x\|^2 - 2\gamma \langle \nabla U(x), x\rangle + \gamma^2 \|\nabla U(x)\|^2 \leq
+$$
+$$
+ \leq (1-\gamma m)\|x\|^2 + 2\gamma^2 L^2 \|x\|^2 + 2\|\nabla U(0)\|^2
+$$
+Thus, plugging everything into expression for $P_{\gamma}V$ and using $\gamma < \frac{m}{4L}$, we obtain
+$$
+P_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + 2\gamma d + (\gamma m - 2\gamma^2 L^2) + 2\gamma^2\|\nabla U(0)\|^2 \leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma(d + 2\gamma \|\nabla U(0)\|^2)
+$$
+Now let $x \in B(0,K_1)$. Then simply using $\|x-\gamma \nabla U(x)\|^2 \leq 2(1+L\gamma)^2\|x\|^2 + 2\gamma^2\|\nabla U(0)\|^2$, we obtain
+$$
+P_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma\left((m - 2\gamma L^2)(1+\|x\|^2) + 2d + 2(1+L\gamma)^2\|x\|^d + 2\gamma \|\nabla U(0)\|^2\right) \leq 
+$$
+
+$$
+\leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma(3K_1^2 + 2d + C_1)
+$$
+for some absolute constant $C_1$, which does not depend on $\gamma, d, x$.
+\end{proof}
+
+It is possible to use \ref{lem:drift_prod} in order to obtain convergence rates in $V$-norm, following the ideas of \cite[Theorem~18.4.3]{douc:moulines:priouret:soulier:2018}.
+Namely, the following lemma holds:
 

### 2019-12-30 Note_exponential_moment/MarkovCon.tex
update

@@
 \item[{\bf H1}] The Markov kernel $P$ is irreducible and aperiodic, with unique invariant probability $\pi$.
-\item[{\bf H2}] There exist a non-empty set $\cC \in \cX$ and two real numbers $u>1$ and $M>0$ such that
-  \[\sup_{x\in \cC}\Ebb_x[u^{\sigma_{\cC}}]\leqslant M\eqsp.\]
-\item[{\bf H3}] There exist $r\in (0,1)$ and $L\geqslant 1$ such that, for any $x$ in the set $\cC$ of {\bf H2} and any $n\geqslant 0$,
+\item[{\bf H2 (drift condition)}] There exist a measurable function $V: X \rightarrow [1; +\infty)$, real numbers $\lambda \in (0,1)$, $b > 0$ and a non-empty set $\cC \in \cX$ such that for any $x \in X$,
 \[
-\tvdist{\delta_xP^n}{\pi} \leqslant Lr^n\eqsp,
+PV(x) \leq \lambda V(x) + b\indiacc{x \in \cC}
 \]
-where $\pi$ is the unique invariant measure granted in {\bf H1}.
+%\item[{\bf H2}] There exist a non-empty set $\cC \in \cX$ and two real numbers $u>1$ and $M>0$ such that
+%  \[\sup_{x\in \cC}\Ebb_x[u^{\sigma_{\cC}}]\leqslant M\eqsp.\]
+%\item[{\bf H3}] There exist $r\in (0,1)$ and $L\geqslant 1$ such that, for any $x$ in the set $\cC$ of {\bf H2} and any $n\geqslant 0$,
+%\[
+%\tvdist{\delta_xP^n}{\pi} \leqslant Lr^n\eqsp,
+%\]
+%where $\pi$ is the unique invariant measure granted in {\bf H1}.
+\item[{\bf H3 (V-ergodicity)}] There exist $r \in (0,1)$, $L \geqslant 1$ and function $V: X \rightarrow \mathbb{R}$ such that for any $x  \in  X$ and any $n \geqslant 0$, it holds
+\[
+d_V(\delta_xP^n,\pi) \leqslant L V(x) r^n\eqsp,
+\]
 \end{itemize}
-When the Markov kernel $P$ is uniformly ergodic, then {\bf H3} holds with $\cC=\sX$. The following Lemma is a coupling result that replaces \cite[Lemma~6]{MR3407208}.
-It is instrumental in the sequel.
+where $\pi$ is the unique invariant measure from {\bf H1}.
+\par
+When the Markov kernel $P$ is uniformly ergodic, then {\bf H3} holds with $V(x) = 1$.   
+\par
+Note that \cite[Proposition 14.3.1]{douc:moulines:priouret:soulier:2018} implies that if Markov kernel $P$ satisfies the drift condition {\bf H2} with the set $\cC \in \cX$, $\cC = \{x: V(x) < d\}$, then it holds for any $x \in \cC$ that
+\[
+\sup_{x\in \cC}\Ebb_x[\lambda^{-\sigma_{\cC}}]\leqslant M\eqsp
+\]
+with some $M$ depending on $V(x)$ and $d$.
+\par
+The following Lemma is a coupling result that replaces \cite[Lemma~6]{MR3407208}. It is instrumental in the sequel.
 \begin{lemma}\label{lem:DMPS18}
@@
 \begin{equation*}\label{eq:Step1}
- |\Ebb_\xi[h(\chunk[0]{X}[n-1])]-\Ebb_{\xi'}[h(\chunk[0]{X}[n-1])]|\leqslant 2\sum_{i=0}^{n-1}c_i \tvdist{\xi P^i}{\xi'P^i}\eqsp.
+ |\Ebb_\xi[h(\chunk[0]{X}[n-1])]-\Ebb_{\xi'}[h(\chunk[0]{X}[n-1])]|\leqslant 2\sum_{i=0}^{n-1}c_i \tvdist{\xi P^i}{\xi'P^i} \leqslant 2\sum_{i=0}^{n-1}c_i \vfuncdist{\xi P^i}{\xi'P^i} \eqsp.
 \end{equation*}
@@
 \]
+The result now follows from the trivial bound $\tvdist{\xi}{\eta} \leq \vfuncdist{\xi}{\eta}$ for arbitrary measures $\xi,\eta$ given that $V(x) \geq 1$ for any $x \in X$.
 \end{proof}
@@
 \section{Main result}\label{sec:Main}
-The main result of this paper  is the following quantitative version of Mac Diarmid's inequality for geometrically ergodic Markov chains.
-\begin{theorem}\label{thm:ConcMarkQuant}
- Assume {\bf H1}, {\bf H2}, {\bf H3}. Let $n\geqslant 1$, $c\in \R^n$ and  $f\in \functionboundeddiff{\sX^n}{c}$. Then, for all $x\in \cC$ and $t > 0$,
+The main result of this paper  is the following quantitative version of Mac Diarmid's inequality for $V-$ergodic Markov chains under certain drift condition.
+%\begin{theorem}\label{thm:ConcMarkQuant}
+ %Assume {\bf H1}, {\bf H2}, {\bf H3}. Let $n\geqslant 1$, $c\in \R^n$ and  $f\in \functionboundeddiff{\sX^n}{c}$. Then, for all $x\in \cC$ and $t > 0$,
+ %\[
+%\Pbb_x\big(f( \chunk[0]{X}[n-1])-\Ebb_x[f(\chunk[0]{X}[n-1])]>t\big)\leqslant \exp\bigg(-\frac{\beta t^2}{\|c\|^2}\bigg)\eqsp,
+% \]
+%where $\beta$ is given by
+ %\[
+ %\beta=\frac{(1-r\vee u^{-1/4})^2}{16L}\bigg(\frac5{\log u}+4ML\bigg)^{-1}\eqsp.
+ %\]
+%\end{theorem}
+\begin{theorem}\label{thm:ConcMarkVErgodic}
+ Assume {\bf H1}, {\bf H2} and {\bf H3} holds with function $V(x)$.
+Let $\cC = \{V(x) < d\}$, $n\geqslant 1$, $c\in \R^n$ and  $f\in \functionboundeddiff{\sX^n}{c}$.
+Then, for all $x\in \cC$ and $t > 0$,
  \[
@@
  \[
- \beta=\frac{(1-r\vee u^{-1/4})^2}{16L}\bigg(\frac5{\log u}+4ML\bigg)^{-1}\eqsp.
+ \beta=\frac{(1-r\vee \lambda^{1/4})^2}{16Ld}\bigg(-\frac5{\log \lambda}+4MLd\bigg)^{-1}\eqsp.
  \]
 \end{theorem}
-\begin{proof}[Proof of \Cref{thm:ConcMarkQuant}]
-Fix $c\in \Rbb^n$, $x\in\sX$ and $f\in \functionboundeddiff{\sX^n}{c}$.
+
+\begin{proof}[Proof of \Cref{thm:ConcMarkVErgodic}]
+Fix $c\in \Rbb^n$, $x\in\sX$ and $f\in \functionboundeddiff{\sX^n}{c}$. Note that drift condition {\bf H3} implies that
+\[
+\sup_{x\in \cC}\Ebb_x[\lambda^{-\sigma_{\cC}}]\leqslant M\eqsp
+\]
 Following \cite{MR3407208}, we decompose $f(\chunk[0]{X}[n-1])-\Ebb_x[f(\chunk[0]{X}[n-1])]$ into martingale increments by conditioning to the stopping times $\tau_{\cC}^i$, $i=0,\ldots, n-1$.
@@
 \label{lem:gi-gipi}
- Assume {\bf H1}, {\bf H2}, {\bf H3}. 
+ Assume {\bf H1}, {\bf H2}, {\bf H3} and let $\cC = \{x:V(x) < d\}$. 
  For any $i\in \{0,\ldots,n-1\}$ and $(\chunk[0]{x}[i-1],x_i)$ in $\sX^i\times\cC$,
 \begin{equation}\label{eq:gi-gipi}
-|g_i(\chunk[0]{x}[i])-g_{i,\pi}(\chunk[0]{x}[i])|\leqslant 2L\sum_{j=i+1}^{n-1}c_jr^{j-i}\eqsp.
+|g_i(\chunk[0]{x}[i])-g_{i,\pi}(\chunk[0]{x}[i])|\leqslant 2LV(x_i)\sum_{j=i+1}^{n-1}c_jr^{j-i} \leqslant 2Ld \sum_{j=i+1}^{n-1}c_jr^{j-i}\eqsp.
 \end{equation}
@@
  |g_i(\chunk[0]{x}[i])-g_{i,\pi}(\chunk[0]{x}[i])|&=|\Ebb_{x_i}[f(\chunk[0]{x}[i],\chunk[1]{X}[n-1-i])]-\Ebb_{\pi}[f(\chunk[0]{x}[i],\chunk[1]{X}[n-1-i])]|\\
- &=|\Ebb_{x_i}[\tilde{f}_i(\chunk[1]{X}[n-1-i])]-\Ebb_{\pi}[\tilde{f}_i(\chunk[1]{X}[n-1-i])]| \leqslant 2\sum_{j=i+1}^{n-1}c_j \tvdist{\delta_{x_i}P^j}{\pi}\eqsp.
+ &=|\Ebb_{x_i}[\tilde{f}_i(\chunk[1]{X}[n-1-i])]-\Ebb_{\pi}[\tilde{f}_i(\chunk[1]{X}[n-1-i])]| \leqslant 2\sum_{j=i+1}^{n-1}c_j \vfuncdist{\delta_{x_i}P^{j-i}}{\pi} \eqsp.
 \end{align*}
-Inequality \eqref{eq:gi-gipi} follows from {\bf H3}.
+Now inequality \eqref{eq:gi-gipi} follows from {\bf H3}.
 \end{proof}
@@
 \end{align}
-\emph{where, $C_1=5L/(1-r)$ and $C_2=16L^2/(1-\rho)$.}
+\emph{where, $C_1=5Ld/(1-r)$ and $C_2=16L^2d^2/(1-\rho)$.}
 \begin{proof}[Proof of Fact 2.]
@@
 \begin{equation}\label{eq:Bound:R1}
-|R_{i,1}|\leqslant 2\indiacc{\tau_{\cC}^{i-1}=i-1}L\sum_{j=i+1}^{n-1}c_jr^{j-i}\eqsp.
+|R_{i,1}|\leqslant 2\indiacc{\tau_{\cC}^{i-1}=i-1}Ld\sum_{j=i+1}^{n-1}c_jr^{j-i}\eqsp.
 \end{equation}
@@
 \[
-(g_j(\chunk[0]{X}[j])-g_{j,\pi}(\chunk[0]{X}[j]))\indiacc{\tau_{\cC}^i=j}\leqslant 2L\sum_{k=j+1}^{n-1}c_kr^{k-j}\indiacc{\tau_{\cC}^i=j}\leqslant 2L\indiacc{\tau_{\cC}^i=j}\sum_{k=\tau_{\cC}^i+1}^{n-1}c_kr^{k-\tau_{\cC}^i}\eqsp.
+(g_j(\chunk[0]{X}[j])-g_{j,\pi}(\chunk[0]{X}[j]))\indiacc{\tau_{\cC}^i=j}\leqslant 2L\sum_{k=j+1}^{n-1}c_kr^{k-j}\indiacc{\tau_{\cC}^i=j}\leqslant 2Ld\indiacc{\tau_{\cC}^i=j}\sum_{k=\tau_{\cC}^i+1}^{n-1}c_kr^{k-\tau_{\cC}^i}\eqsp.
 \]
@@
 \begin{equation}\label{eq:Bound:R2}
-|R_{i,2}|\leqslant 2L\indiacc{\tau_{\cC}^{i-1}=i-1}\sum_{k=\tau_{\cC}^i+1}^{n-1}c_kr^{k-\tau_{\cC}^i}\eqsp.
+|R_{i,2}|\leqslant 2Ld\indiacc{\tau_{\cC}^{i-1}=i-1}\sum_{k=\tau_{\cC}^i+1}^{n-1}c_kr^{k-\tau_{\cC}^i}\eqsp.
 \end{equation}
@@
 \begin{align}
-\label{eq:Gi1-Gi22} |G_{i,2}-G_{i,1}|\leqslant &2L\bigg(\sum_{j=i+1}^{n-1}c_jr^{j-i}+\sum_{k=\tau_{\cC}^i+1}^{n-1}c_kr^{k-\tau_{\cC}^i}+\frac1{2L}\sum_{k=i}^{\tau_{\cC}^i\wedge (n-1)}c_k\bigg)\indiacc{\tau_{\cC}^{i-1}=i-1}\eqsp.
+\label{eq:Gi1-Gi22} |G_{i,2}-G_{i,1}|\leqslant &2Ld\bigg(\sum_{j=i+1}^{n-1}c_jr^{j-i}+\sum_{k=\tau_{\cC}^i+1}^{n-1}c_kr^{k-\tau_{\cC}^i}+\frac1{2Ld}\sum_{k=i}^{\tau_{\cC}^i\wedge (n-1)}c_k\bigg)\indiacc{\tau_{\cC}^{i-1}=i-1}\eqsp.
 \end{align}
@@
 \[
-|G_{i}-G_{i-1}|=|G_{i,2}-G_{i,1}|\leqslant \frac{5L\|c\|_{\infty}}{1-r}\sigma_{\cC}\circ\theta^{i-1}\indiacc{\tau_{\cC}^{i-1}=i-1}\eqsp.
+|G_{i}-G_{i-1}|=|G_{i,2}-G_{i,1}|\leqslant \frac{5Ld\|c\|_{\infty}}{1-r}\sigma_{\cC}\circ\theta^{i-1}\indiacc{\tau_{\cC}^{i-1}=i-1}\eqsp.
 \]
@@
 \begin{align*}
- |G_{i}-G_{i-1}|^2=&|G_{i,2}-G_{i,1}|^2\leqslant16L^2\rho^{-2\sigma_{\cC}\circ\theta^{i-1}}\bigg(\sum_{k=i}^{n-1}c_k\rho^{k-i}\bigg)^2\indiacc{\tau_{\cC}^{i-1}=i-1}\\
- \leqslant &\frac{16L^2}{1-\rho}\rho^{-2\sigma_{\cC}\circ\theta^{i-1}}\sum_{k=i}^{n-1}c_k^2\rho^{k-i}\indiacc{\tau_{\cC}^{i-1}=i-1}\eqsp.
+ |G_{i}-G_{i-1}|^2=&|G_{i,2}-G_{i,1}|^2\leqslant16L^2d^2\rho^{-2\sigma_{\cC}\circ\theta^{i-1}}\bigg(\sum_{k=i}^{n-1}c_k\rho^{k-i}\bigg)^2\indiacc{\tau_{\cC}^{i-1}=i-1}\\
+ \leqslant &\frac{16L^2d^2}{1-\rho}\rho^{-2\sigma_{\cC}\circ\theta^{i-1}}\sum_{k=i}^{n-1}c_k^2\rho^{k-i}\indiacc{\tau_{\cC}^{i-1}=i-1}\eqsp.
 \end{align*}
@@
 \noindent
-{\bf Fact 3.} \emph{ Assume {\bf H1}, {\bf H2}, {\bf H3}. For any $x\in \cC$,}
+{\bf Fact 3.} \emph{ Assume {\bf H1}, {\bf H2}, {\bf H3}. Let $\cC = \{x: V(x) < d\}$. Then for any $x\in \cC$,}
 \begin{align}
@@
 \end{align}
-\emph{where $C_3=\fracaa{4L\left(\fracaa{5}{\log u}+4ML\right)}{(1-r\vee u^{-1/4})^2}$.}
+\emph{where $C_3=\fracaa{4Ld\left(-\fracaa{5}{\log \lambda}+4MLd\right)}{(1-r\vee \lambda^{1/4})^2}$.}
 \begin{proof}[Proof of Fact 3.]
@@
 \[
-\frac2{\varepsilon}+\frac{MC_2}{1-\rho}\leqslant \frac{4L}{(1-r\vee u^{-1/4})^2}\bigg(\frac5{\log u}+4ML\bigg)\eqsp.
+\frac2{\varepsilon}+\frac{MC_2}{1-\rho}\leqslant \frac{4Ld}{(1-r\vee \lambda^{1/4})^2}\bigg(-\frac5{\log \lambda}+4MLd\bigg)\eqsp.
 \]
@@
 \[
-\beta=\frac1{2C}=\frac1{4C_3}=\frac{(1-r\vee u^{-1/4})^2}{16L}\bigg(\frac5{\log u}+4ML\bigg)^{-1}\eqsp.
+\beta=\frac1{2C}=\frac1{4C_3}=\frac{(1-r\vee \lambda^{1/4})^2}{16Ld}\bigg(-\frac5{\log \lambda}+4MLd\bigg)^{-1}\eqsp.
 \]
@@
 
-It is possible to use \ref{lem:drift_prod} in order to obtain convergence rates in $V$-norm, following the ideas of \cite[Theorem~18.4.3]{douc:moulines:priouret:soulier:2018}.
-Namely, the following lemma holds:
-
 \bibliographystyle{plain}

### 2020-01-07 Note_exponential_moment/MarkovCon.tex
updat

@@
 Note that \cite[Proposition 14.3.1]{douc:moulines:priouret:soulier:2018} implies that if Markov kernel $P$ satisfies the drift condition {\bf H2} with the set $\cC \in \cX$, $\cC = \{x: V(x) < d\}$, then it holds for any $x \in \cC$ that
-\[
+\begin{equation}
+\label{eq:exp_moment_hit}
 \sup_{x\in \cC}\Ebb_x[\lambda^{-\sigma_{\cC}}]\leqslant M\eqsp
-\]
+\end{equation}
 with some $M$ depending on $V(x)$ and $d$.
@@
 \begin{theorem}\label{thm:ConcMarkVErgodic}
- Assume {\bf H1}, {\bf H2} and {\bf H3} holds with function $V(x)$.
-Let $\cC = \{V(x) < d\}$, $n\geqslant 1$, $c\in \R^n$ and  $f\in \functionboundeddiff{\sX^n}{c}$.
+ Assume {\bf H1}, {\bf H2} and {\bf H3} holds with drift function $V(x)$.
+Let $M$ be a constant from \ref{eq:exp_moment_hit}, $\cC = \{V(x) < d\}$, $n\geqslant 1$, $c\in \R^n$ and  $f\in \functionboundeddiff{\sX^n}{c}$.
 Then, for all $x\in \cC$ and $t > 0$,
@@
 \label{lem:drift_prod}
- Assume that $P_{\gamma}$ satisfies condition \ref{eq:drift} for any $0 < \gamma < \overline{gamma}$. Then it holds
+ Assume that $P_{\gamma}$ satisfies condition \ref{eq:drift} for any $0 < \gamma < \overline{\gamma}$. Then it holds
 $$
@@
 
-\begin{lemma} Assume that the potential $U(x)$ satisfies conditions {\bf C1} and {\bf C2}.
-Then the kernel \ref{} satisfies drift condition \ref{eq:drift} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L^2}$ with drift function $V(x) = 1 + \|x\|^2$, constants $\lambda = \exp{-\left(\frac{m}{2}\right)}$, $C = 3K_1^2 + 2d + C_1$ for some absolute constant $C_1$.
+\begin{lemma} Assume that the potential $U(x)$ satisfies conditions {\bf C1} and {\bf C2} and without loss of generality consider $\nabla U(0) = 0$.
+Then the kernel \ref{} satisfies drift condition \ref{eq:drift} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L^2}$ with drift function $V(x) = 1 + \|x\|^2$, constants $\lambda = \exp{-\left(\frac{m}{2}\right)}$, $C = 3K_1^2 + 2d + C_1$ for some absolute constant $C_1 > 0$.
 \end{lemma}
@@
 $$
- \leq (1-\gamma m)\|x\|^2 + 2\gamma^2 L^2 \|x\|^2 + 2\|\nabla U(0)\|^2
+ \leq (1-\gamma m)\|x\|^2 + 2\gamma^2 L^2 \|x\|^2
 $$
@@
 $$
-P_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + 2\gamma d + (\gamma m - 2\gamma^2 L^2) + 2\gamma^2\|\nabla U(0)\|^2 \leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma(d + 2\gamma \|\nabla U(0)\|^2)
+P_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + 2\gamma d + (\gamma m - 2\gamma^2 L^2) \leq \exp^{-\frac{\gamma m}{2}}V(x) + 2\gamma d
 $$
-Now let $x \in B(0,K_1)$. Then simply using $\|x-\gamma \nabla U(x)\|^2 \leq 2(1+L\gamma)^2\|x\|^2 + 2\gamma^2\|\nabla U(0)\|^2$, we obtain
+Now let $x \in B(0,K_1)$. Then simply using $\|x-\gamma \nabla U(x)\|^2 \leq 2(1+L\gamma)^2\|x\|^2$, we obtain
 $$
-P_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma\left((m - 2\gamma L^2)(1+\|x\|^2) + 2d + 2(1+L\gamma)^2\|x\|^d + 2\gamma \|\nabla U(0)\|^2\right) \leq 
+P_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma\left((m - 2\gamma L^2)(1+\|x\|^2) + 2d + 2(1+L\gamma)^2\|x\|^d\right) \leq 
 $$
@@
 
+Now the geometric ergodicity result readily follows from \cite[Theorem~19.4.1]{douc:moulines:priouret:soulier:2018}, namely, the following lemma holds:
+
+\begin{lemma} Let the potential $U(x), x \in \mathbb{R}^D$ satisfy conditions {\bf C1} and {\bf C2}. Then for $0 < \gamma < \overline{\gamma}=\frac{m}{4L^2}$, for any $x \in X$ it holds
+$$
+d_V(\delta_xQ^n_\gamma,\pi_\gamma) \leq C\rho^n\left(V(x) + \pi_\gamma(V)\right)
+$$ 
+with $V(x) = 1 + \|x\|^2$ and constants
+$$
+b = \gamma(3K_1^2 + 2D + C_1); \quad \overline{b} = b\exp^{-\frac{m\gamma}{2}} + d; \quad \varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right); \quad
+$$
+
+$$
+C = \left(1 + \exp^{-\frac{m\gamma}{2}}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp^{-\frac{m\gamma}{2}} - \frac{2b}{1+d})}\right);
+$$
+
+$$
+\rho = \exp^{-\gamma\left(\frac{m}{2} - 2\frac{3K_1^2 + 2D + C_1}{d}\right)\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp^{-\frac{m\gamma}{2}} + \frac{2b}{d+1})}}}
+$$
+for some $C_1 > 0$, which does not depend on $d, \gamma$ and $n$.
+\end{lemma}
+
+\begin{proof} Note that the condition {\bf C1} implies that the Markov kernel $Q^n_\gamma$ satisfies $(1,\varepsilon)$-Doeblin condition with $\varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right)$.
+Together with drift condition {\bf C2} it allows to apply \cite[Theorem~19.4.1]{douc:moulines:priouret:soulier:2018} with appropriate constants.
+\end{proof}
+
 \bibliographystyle{plain}

### 2020-01-19 aap/Revision/vr-mcmc-revision.tex
update

@@
 \frac{A_{K}^{2}}{(1-\gamma_1 B_\mu)^2}
-\sum_{j=l}^{q} \gamma_{j}
+\mathsf{E}\sum_{j=l}^{q} \gamma_{j}
 \left(
@@
 \frac{A_{K}^{2}}{(1-\gamma_1 B_\mu)^3}
-\sum_{j=l}^{q} \gamma_{j}
+\mathsf{E}\sum_{j=l}^{q} \gamma_{j}
 \prod_{k=l}^{j} \alpha_{k}
@@
 \right)^{2}
-\\
-&\le
-\frac{A_{K}^{2}}{(1-\gamma_1 B_\mu)^3}C^3=B_K,
 \end{align*}
-where $C$ is the bound from~\eqref{eq:17042018a1}.
+
+%\som{Changed exponent from 1/2 to 2 here}
+Now, from Hoelder inequality, we obtain (with $\|X\|_p = (\mathsf{E}X^p)^{\frac{1}{p}}$)
+\begin{align*}
+\mathsf{E}\left[\sum_{j=l}^{q}\gamma_{j}\prod_{k=l}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\gamma_{p}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right] & \leq\sum_{j=l}^{q}\gamma_{j}\left\Vert \prod_{k=l}^{j}\alpha_{k}\right\Vert _{2}\left\Vert \sum_{p=j+1}^{q+1}\gamma_{p}\prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}^{2}\\
+ & \leq\sum_{j=l}^{q}\gamma_{j}\left\Vert \prod_{k=l}^{j}\alpha_{k}\right\Vert _{2}\left(\sum_{p=j+1}^{q+1}\gamma_{p}\left\Vert \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}\right)^{2}.
+\end{align*}
+Now using the fact that $\prod\limits_{k=j+1}^{p}\alpha_{k} \leq \exp\left(-\sum\limits_{k=j+1}^{p}\gamma_{k}\mu'(X_{k-1})\right),$
+we get 
+\[
+\mathsf{E}\left[\sum_{j=l}^{q}\gamma_{j}\prod_{k=l}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\gamma_{p}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]\leq\sum_{j=l}^{q}\gamma_{j} \zeta^{1/2}_{l,j}(2) \left(\sum_{p=j+1}^{q+1}\gamma_{p} \zeta^{1/4}_{j+1,p}(4)\right)^{2},
+\]
+where we denote
+\[
+\zeta_{l,j}(s)=\mathsf{E}\left[e^{-s\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})}\right],\quad s>0.
+\]
+From Theorem~\ref{exp_mom_ula} in Appendix~\ref{sec:appendix}, it follows that
+\[
+\zeta_{l,j}(s) \leq e^{s\kappa \Gamma^*}e^{-2s\Gamma_{l,j}\alpha}
+\]
+with $\Gamma^* = \sum\limits_{k=1}^{\infty}\gamma_k^2, \kappa = ..., \Gamma_{l,j} = \sum\limits_{k=l}^{j}\gamma_k, \alpha = \int \mu^2(x)\pi(x)\,dx$.
+Since $\sum\limits_{k=1}^{\infty}\gamma_k = \infty$, we may bound
+\begin{align*}
+\sum\limits_{p=j+1}^{\infty} \gamma_p e^{-2\alpha \Gamma_{j+1,p}} < \int\limits_{0}^{+\infty}e^{-2\alpha x}\,dx = \frac{1}{2\alpha}
+\end{align*}
+yielding 
+\begin{align*}
+\sum_{p=j+1}^{q+1}\gamma_{p} \zeta^{1/4}_{j+1,p}(4) \leq e^{\kappa \Gamma^*}\sum_{p=j+1}^{q+1}\gamma_{p}e^{-2\alpha \Gamma_{j+1,p}} \leq \frac{e^{\kappa \Gamma^*}}{2\alpha} 
+\end{align*}
+Using the same integral bound, 
+\begin{align*}
+\mathsf{E}\left[\sum_{j=l}^{q}\gamma_{j}\prod_{k=l}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\gamma_{p}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right] \leq
+\frac{e^{2\kappa \Gamma^*}}{2\alpha}\sum_{j=l}^{q}\gamma_{j}e^{-2\alpha \Gamma_{l,j}} \leq \frac{e^{2\kappa \Gamma^*}}{4\alpha^2}
+\end{align*}
+Plugging into bound for conditional variance, 
+\begin{align*}
+\mathsf{Var}\left[\sum_{p=l}^{q}\gamma_{p+1}\partial_{y_{l}}^{K}f\left(X_{p}\right)\Big|X_{l-1}\right] \leq
+\frac{A_{K}^{2}e^{2\kappa \Gamma^*}}{4\alpha^2(1-\gamma_1 B_\mu)^3}
+\end{align*}
+\par
 The proof is completed.
@@
 
-\section{Bounds for moments of  ULA}
+\section{Bounds for moments of  ULA}\label{sec:appendix}
 For $\gamma>0$, the Markov kernel $R_\gamma$
@@
 \begin{description}
-\item[(H1)] Assume that $\nabla U$ is Lipschitz, that is, there exists $L\geq 0$ such that 
+\item[{\bf (H1)}] Assume that $\nabla U$ is Lipschitz, that is, there exists $L\geq 0$ such that 
 \begin{eqnarray*}
@@
 
-\item[(H2)] There exist $K>0,$ $M>0$ and $m>0$ such that for any $x\not\in B_K(0)	$ and any $y\in \mathbb{R}^d$ it holds 
+\item[{\bf (H2)}] There exist $K>0,$ $M>0$ and $m>0$ such that for any $x\not\in B_K(0)	$ and any $y\in \mathbb{R}^d$ it holds 
 \begin{eqnarray*}
@@
 \end{eqnarray*}
-
 \end{description}
 
+Due to \cite[Lemma~14]{brosse:2018:diffusion}, it is known that conditions (H1) and (H2) imply the result
+
+\begin{description}
+\item[{\bf (H3)}] There exist $K_1 > 0, m > 0$, such that for any $x \not\in B_{K_1}(0)$ it holds
+\begin{eqnarray*}
+\langle \nabla U(x), x \rangle \geq \frac{m}{2}\|x\|^2
+\end{eqnarray*}
+\end{description}
+
+\begin{lem} Assume that the potential $U(x), x \in \mathbb{R}^D$ satisfies conditions {\bf (H1)} and {\bf (H3)} and without loss of generality consider $\nabla U(0) = 0$.
+Then the kernel \ref{} satisfies drift condition \ref{eq:drift} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L^2}$ with drift function $V(x) = 1 + \|x\|^2$, constants $\lambda = \exp{-\left(\frac{m}{2}\right)}$, $C = 3K_1^2 + 2D + C_1$ for some absolute constant $C_1 > 0$, , which does not depend on $\gamma, D, x$.
+\end{lem}
+
+\begin{proof} Consider $V(x) = 1 + \|x\|^2$, then
+$$
+P_{\gamma}V(x) = \int\limits_{\mathbb{R}^d}V(y)P_{\gamma}(x,dy) = \int\limits_{\mathbb{R}^d}(1+\|y\|^2)\frac{1}{(4\pi\gamma)^{\frac{d}{2}}}\exp{\left(-\frac{\|y-x+\gamma\nabla U(x)\|^2}{4\gamma}\right)}\,dy =
+$$
+
+$$
+=1 + \frac{1}{(4\pi\gamma)^{\frac{d}{2}}}\int\limits_{\mathbb{R}^d}\|z + x - \gamma \nabla U(x)\|^2 \exp{\left(-\frac{\|z\|^2}{4\gamma}\right)}\,dz
+$$
+Let us first consider the case $x \notin B(0,K_1)$.
+Note that $\|z+x-\gamma \nabla U(x)\|^2 = \|z\|^2 + 2 \langle z, x - \gamma \nabla U(x) \rangle + \gamma^2 \|x - \gamma \nabla U(x)\|^2$ and the linear term vanishes after integration, moreover, for $Z \sim \mathcal{N}(0,2\gamma I_d$, it holds $\mathbb{E}\|Z\|^2 = 2\gamma d$.
+It remains to notice that due to {\bf C1} and {\bf C2}
+$$
+\|x - \gamma \nabla U(x) \|^2 = \|x\|^2 - 2\gamma \langle \nabla U(x), x\rangle + \gamma^2 \|\nabla U(x)\|^2 \leq
+$$
+$$
+ \leq (1-\gamma m)\|x\|^2 + 2\gamma^2 L^2 \|x\|^2
+$$
+Thus, plugging everything into expression for $P_{\gamma}V$ and using $\gamma < \frac{m}{4L}$, we obtain
+$$
+P_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + 2\gamma d + (\gamma m - 2\gamma^2 L^2) \leq \exp^{-\frac{\gamma m}{2}}V(x) + 2\gamma d
+$$
+Now let $x \in B(0,K_1)$. Then simply using $\|x-\gamma \nabla U(x)\|^2 \leq 2(1+L\gamma)^2\|x\|^2$, we obtain
+$$
+P_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma\left((m - 2\gamma L^2)(1+\|x\|^2) + 2d + 2(1+L\gamma)^2\|x\|^d\right) \leq 
+$$
+
+$$
+\leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma(3K_1^2 + 2D + C_1)
+$$
+for some absolute constant $C_1$.
+\end{proof}
+
+Now the geometric ergodicity result readily follows from \cite[Theorem~19.4.1]{douc:moulines:priouret:soulier:2018}, namely, the following lemma holds:
+
+\begin{lem} Let the potential $U(x), x \in \mathbb{R}^D$ satisfy conditions {\bf H1} and {\bf H3}. Then for $0 < \gamma < \overline{\gamma}=\frac{m}{4L^2}$, for any $x \in X$ it holds
+$$
+d_V(\delta_xQ^n_\gamma,\pi_\gamma) \leq C\rho^n\left(V(x) + \pi_\gamma(V)\right)
+$$ 
+with $V(x) = 1 + \|x\|^2$ and constants
+$$
+b = \gamma(3K_1^2 + 2D + C_1); \quad \overline{b} = b\exp^{-\frac{m\gamma}{2}} + d; \quad \varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right); \quad
+$$
+
+$$
+C = \left(1 + \exp^{-\frac{m\gamma}{2}}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp^{-\frac{m\gamma}{2}} - \frac{2b}{1+d})}\right);
+$$
+
+$$
+\rho = \exp^{-\gamma\left(\frac{m}{2} - 2\frac{3K_1^2 + 2D + C_1}{d}\right)\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp^{-\frac{m\gamma}{2}} + \frac{2b}{d+1})}}}
+$$
+for some $C_1 > 0$, which does not depend on $d, \gamma$ and $n$.
+\end{lem}
+
+\begin{proof} Note that the condition {\bf H1} implies that the Markov kernel $Q^n_\gamma$ satisfies $(1,\varepsilon)$-Doeblin condition with $\varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right)$.
+Together with drift condition {\bf H2} it allows to apply \cite[Theorem~19.4.1]{douc:moulines:priouret:soulier:2018} with appropriate constants.
+\end{proof}
+
 \begin{thm}
+\label{exp_mom_ula}
 Suppose that
@@
 \end{eqnarray*}
-Under assumptions (H1) and (H2), then for any bounded function $g$ on $\mathbb{R}^d$ 
+Under assumptions {\bf (H1)} and {\bf (H2)}, for any bounded function $g$ on $\mathbb{R}^d$ 
 we have
@@
 \end{thm}
+\begin{proof}
+ Note that 
+\[
+\mathsf{E}\left[e^{-s\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})}\right]=\mathsf{E}\left[e^{-s\left(\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})-\Gamma_{l,j}\int\mu'(x)\pi(x)\,dx\right)}\right]e^{-s\Gamma_{l,j}\int\mu'(x)\pi(x)\,dx}
+\]
+ with $\Gamma_{l,j}=\sum_{k=l}^{j}\gamma_{k}.$ According to \cite[Theorem~3]{bms:exp_conc_note}, it holds
+\[
+\mathsf{E}\left[e^{-s\left(\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})-\Gamma_{l,j}\int\mu'(x)\pi(x)\,dx\right)}\right]\leq e^{s\varkappa\sum_{k=l}^{j}\gamma_{k}^{2}}
+\]
+ for some $\varkappa>0.$ Since $\sum_{k=1}^{\infty}\gamma_{k}^{2}\leq\Gamma^{\star},$we
+have 
+\[
+\mathsf{E}\left[e^{-s\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})}\right]\leq e^{s\varkappa\Gamma^{\star}}e^{-s\Gamma_{l,j}\int\mu'(x)\pi(x)\,dx}.
+\]
+Furthermore, since $\mu\pi',\mu'\pi\in L^{1}(\mathbb{R}),$ we have
+\[
+\int\mu'(x)\pi(x)\,dx=-\int\mu(x)\pi'(x)\,dx=2\int\mu^{2}(x)\pi(x)\,dx>0
+\]
+and as a result 
+\[
+\zeta_{l,j}(s)\leq e^{s\varkappa\Gamma^{\star}}e^{-2s\Gamma_{l,j}\int\mu^{2}(x)\pi(x)\,dx}
+\]
+ for any $l\leq j.$
+\end{proof}
+
+Suppose that now we use ULA with fixed step size, then we might use the following version of the previous theorem:
+\begin{thm} Suppose that we use ULA with constant step size $\gamma$.
+Under assumptions {\bf (H1)} and {\bf (H2)}, for any bounded function $g$ on $\mathbb{R}^d$ we have, given $0 < \gamma < \left(\frac{\int \mu^{\prime}(x)\pi(x)\,dx}{C_1 M}\right)^2$,
+\begin{eqnarray*}
+\mathsf{E}\left[\exp\left(\rho\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k})-\left(\sum_{k=l}^{p}\gamma_{k+1}\right)\pi(g)\right)\right)\right]\leq C
+\end{eqnarray*}
+for some absolute constant $C$ not depending on $l,p$.
+\end{thm}
+
+\begin{proof} Note that 
+\begin{eqnarray*} 
+\int \mu^{\prime}(x)\pi_{\gamma}(x)\,dx \geq \int \mu^{\prime}(x)\pi(x)\,dx - M d_{TV}(\pi,\pi_{\gamma}) \geq \int \mu^{\prime}(x)\pi(x)\,dx - C_1\sqrt{\gamma}M
+\end{eqnarray*}
+where the precise expression for constant $C_1$ might be found in \cite[Theorem~10]{durmus:moulines:aop:2016}.
+Thus, for $\gamma < \left(\frac{\int \mu^{\prime}(x)\pi(x)\,dx}{C_1 M}\right)^2$, we obtain along the lines of the proof of Theorem~16 that
+$$
+\int \mu^{\prime}(x)\pi_{\gamma}(x)\,dx > 0
+$$
+
+\end{proof}
 

### 2020-01-19 aap/Revision/vr-mcmc-revision.tex
update

@@
 
-%\som{Changed exponent from 1/2 to 2 here}
 Now, from Hoelder inequality, we obtain (with $\|X\|_p = (\mathsf{E}X^p)^{\frac{1}{p}}$)
@@
 \[
+\mathsf{E}\left[e^{-s\left(\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})-\Gamma_{l,j}\int\mu'(x)\pi(x)\,dx\right)}\right]\leq e^{s\kappa\sum_{k=l}^{j}\gamma_{k}^{2}}
+\]
+with constant $\kappa = ...$. We used the fact that $\mu^{\prime}(x)$ is bounded. Denote $\sum_{k=1}^{\infty}\gamma_{k}^{2} = \Gamma^{\star}$, then we have 
+\[
+\mathsf{E}\left[e^{-s\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})}\right]\leq e^{s\varkappa\Gamma^{\star}}e^{-s\Gamma_{l,j}\int\mu'(x)\pi(x)\,dx}.
+\]
+Furthermore, since $\mu\pi',\mu'\pi\in L^{1}(\mathbb{R})$, we have
+\[
+\int\mu'(x)\pi(x)\,dx=-\int\mu(x)\pi'(x)\,dx=2\int\mu^{2}(x)\pi(x)\,dx>0
+\]
+yielding the final bound
+\[
 \zeta_{l,j}(s) \leq e^{s\kappa \Gamma^*}e^{-2s\Gamma_{l,j}\alpha}
 \]
-with $\Gamma^* = \sum\limits_{k=1}^{\infty}\gamma_k^2, \kappa = ..., \Gamma_{l,j} = \sum\limits_{k=l}^{j}\gamma_k, \alpha = \int \mu^2(x)\pi(x)\,dx$.
-Since $\sum\limits_{k=1}^{\infty}\gamma_k = \infty$, we may bound
+where $\Gamma_{l,j} = \sum\limits_{k=l}^{j}\gamma_k, \alpha = \int \mu^2(x)\pi(x)\,dx$. Since $\sum\limits_{k=1}^{\infty}\gamma_k = \infty$, we may estimate
 \begin{align*}

### 2020-01-20 aap/Revision/vr-mcmc-revision.tex
update

@@
 is defined by
-$$
+\begin{equation}
+\label{eq:kernel}
 R_\gamma(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \nabla U(x)\right\|^2\right\}dy,\quad A\in \mathcal B(\mathbb R^d),
-$$
+\end{equation}
 while, for $k,l\in\mathbb N$, $k\le l$,
@@
 
-Due to \cite[Lemma~14]{brosse:2018:diffusion}, it is known that conditions (H1) and (H2) imply the result
-
-\begin{description}
-\item[{\bf (H3)}] There exist $K_1 > 0, m > 0$, such that for any $x \not\in B_{K_1}(0)$ it holds
-\begin{eqnarray*}
+Due to \cite[Lemma~14]{brosse:2018:diffusion}, conditions {\bf (H1)} and {\bf (H2)} imply that there exist $K_1 > 0, m > 0$, such that for any $x \not\in B_{K_1}(0)$ it holds
+\begin{equation}
+\label{eq:H3}
 \langle \nabla U(x), x \rangle \geq \frac{m}{2}\|x\|^2
-\end{eqnarray*}
-\end{description}
-
-\begin{lem} Assume that the potential $U(x), x \in \mathbb{R}^D$ satisfies conditions {\bf (H1)} and {\bf (H3)} and without loss of generality consider $\nabla U(0) = 0$.
-Then the kernel \ref{} satisfies drift condition \ref{eq:drift} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L^2}$ with drift function $V(x) = 1 + \|x\|^2$, constants $\lambda = \exp{-\left(\frac{m}{2}\right)}$, $C = 3K_1^2 + 2D + C_1$ for some absolute constant $C_1 > 0$, , which does not depend on $\gamma, D, x$.
+\end{equation}
+
+\begin{lem} Assume that the potential $U(x), x \in \mathbb{R}^d$ satisfies conditions {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$.
+Then the kernel $R_\gamma$ from \ref{eq:kernel} satisfies drift condition \ref{eq:drift} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L^2}$ with drift function $V(x) = 1 + \|x\|^2$, constants $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = \frac{25K_1^2}{8} + 2d + m$ with $K_1$ from \ref{eq:H3}, $m$ from {\bf (H2)}.
 \end{lem}
@@
 $$
-P_{\gamma}V(x) = \int\limits_{\mathbb{R}^d}V(y)P_{\gamma}(x,dy) = \int\limits_{\mathbb{R}^d}(1+\|y\|^2)\frac{1}{(4\pi\gamma)^{\frac{d}{2}}}\exp{\left(-\frac{\|y-x+\gamma\nabla U(x)\|^2}{4\gamma}\right)}\,dy =
+R_{\gamma}V(x) = \int\limits_{\mathbb{R}^d}V(y)P_{\gamma}(x,dy) = \int\limits_{\mathbb{R}^d}(1+\|y\|^2)\frac{1}{(4\pi\gamma)^{\frac{d}{2}}}\exp{\left(-\frac{\|y-x+\gamma\nabla U(x)\|^2}{4\gamma}\right)}\,dy =
 $$
@@
 $$
-Let us first consider the case $x \notin B(0,K_1)$.
-Note that $\|z+x-\gamma \nabla U(x)\|^2 = \|z\|^2 + 2 \langle z, x - \gamma \nabla U(x) \rangle + \gamma^2 \|x - \gamma \nabla U(x)\|^2$ and the linear term vanishes after integration, moreover, for $Z \sim \mathcal{N}(0,2\gamma I_d$, it holds $\mathbb{E}\|Z\|^2 = 2\gamma d$.
-It remains to notice that due to {\bf C1} and {\bf C2}
+Let us first consider the case $x \notin B(0,K_1)$. Note that 
+$$
+\|z+x-\gamma \nabla U(x)\|^2 = \|z\|^2 + 2 \langle z, x - \gamma \nabla U(x) \rangle + \gamma^2 \|x - \gamma \nabla U(x)\|^2
+$$ 
+and the linear term vanishes after integration, moreover, for $Z \sim \mathcal{N}(0,2\gamma I_d)$, it holds $\mathbb{E}\|Z\|^2 = 2\gamma d$.
+It remains to notice that due to {\bf H1} and {\bf H2} (which implies \ref{eq:H3}),
 $$
@@
 $$
+
 $$
@@
 $$
-Thus, plugging everything into expression for $P_{\gamma}V$ and using $\gamma < \frac{m}{4L}$, we obtain
-$$
-P_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + 2\gamma d + (\gamma m - 2\gamma^2 L^2) \leq \exp^{-\frac{\gamma m}{2}}V(x) + 2\gamma d
+Thus, plugging everything into expression for $R_{\gamma}V$ and using $\gamma < \frac{m}{4L}$, we obtain
+$$
+R_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + 2\gamma d + (\gamma m - 2\gamma^2 L^2) \leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma (2d + m)
 $$
@@
 $$
-P_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma\left((m - 2\gamma L^2)(1+\|x\|^2) + 2d + 2(1+L\gamma)^2\|x\|^d\right) \leq 
-$$
-
-$$
-\leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma(3K_1^2 + 2D + C_1)
+R_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma\left((m - 2\gamma L^2)(1+\|x\|^2) + 2d + 2(1+L\gamma)^2\|x\|^d\right) \leq 
+$$
+
+$$
+\leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma(\frac{25K_1^2}{8} + 2d + m)
 $$
@@
 $$
-for some $C_1 > 0$, which does not depend on $d, \gamma$ and $n$.
 \end{lem}

### 2020-01-20 aap/Revision/vr-mcmc-revision.tex
update

@@
 \newtheorem{thm}{Theorem}
+\newtheorem{definition}{Definition}
 \newtheorem{lem}[thm]{Lemma}
@@
 \newcommand{\jac}[1]{\operatorname{J} _{#1}}
+
+%added command for indicator - SS
+\newcommand{\indiacc}[1]{\mathbbm{1}_{\{#1\}}}
 
@@
 \section{Bounds for moments of  ULA}\label{sec:appendix}
-For $\gamma>0$, the Markov kernel $R_\gamma$
-on $(\mathbb R^d,\mathcal B(\mathbb R^d))$
-is defined by
+In this appendix we give some definitions and prove auxiliary results which are the used to bound the exponential moment of the additive functional of ULA chain (Theorem~\ref{th:exp_mom_ula}).
+\par
+For $\gamma>0$, the Markov kernel $R_\gamma$ on $(\mathbb R^d,\mathcal B(\mathbb R^d))$, corresponding to the Unadjusted Langevin Algorithm with step $\gamma$ is defined by
 \begin{equation}
@@
 
-Due to \cite[Lemma~14]{brosse:2018:diffusion}, conditions {\bf (H1)} and {\bf (H2)} imply that there exist $K_1 > 0, m > 0$, such that for any $x \not\in B_{K_1}(0)$ it holds
+Due to \cite[Lemma~14]{brosse:2018:diffusion}, conditions {\bf (H1)} and {\bf (H2)} implies that there exist constants $K_1 > 0, m > 0$, such that for any $x \not\in B_{K_1}(0)$ it holds
 \begin{equation}
@@
 
+We would show that conditions {\bf (H1)} and {\bf (H2)} are sufficient to show that ULA kernel $R_\gamma$ satisfies the so-called drift condition:
+\begin{definition}
+\label{def:drift}
+We say that the Markov kernel $R_\gamma$ satisfies Foster-Lyapunov drift condition if there exist a measurable function $V: X \rightarrow [1; +\infty)$, real numbers $\lambda \in (0,1)$ and $C > 0$ such that for any $x \in X$,
+\begin{eqnarray*}
+RV(x) \leq \lambda^{\gamma} V(x) + \gamma C
+\end{eqnarray*}
+\end{definition}
+
 \begin{lem} Assume that the potential $U(x), x \in \mathbb{R}^d$ satisfies conditions {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$.
-Then the kernel $R_\gamma$ from \ref{eq:kernel} satisfies drift condition \ref{eq:drift} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L^2}$ with drift function $V(x) = 1 + \|x\|^2$, constants $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = \frac{25K_1^2}{8} + 2d + m$ with $K_1$ from \ref{eq:H3}, $m$ from {\bf (H2)}.
+Then the kernel $R_\gamma$ from \ref{eq:kernel} satisfies drift condition \ref{def:drift} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L^2}$ with drift function $V(x) = 1 + \|x\|^2$, constants $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = \frac{25K_1^2}{8} + 2d + m$ where $K_1$ is from \ref{eq:H3}, $m$ is from {\bf (H2)}.
 \end{lem}
@@
 $$
-R_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma\left((m - 2\gamma L^2)(1+\|x\|^2) + 2d + 2(1+L\gamma)^2\|x\|^d\right) \leq 
+R_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma\left((m - 2\gamma L^2)(1+\|x\|^2) + 2d + 2(1+L\gamma)^2\|x\|^2\right) \leq 
 $$
@@
 $$
-for some absolute constant $C_1$.
 \end{proof}
 
-Now the geometric ergodicity result readily follows from \cite[Theorem~19.4.1]{douc:moulines:priouret:soulier:2018}, namely, the following lemma holds:
-
-\begin{lem} Let the potential $U(x), x \in \mathbb{R}^D$ satisfy conditions {\bf H1} and {\bf H3}. Then for $0 < \gamma < \overline{\gamma}=\frac{m}{4L^2}$, for any $x \in X$ it holds
+For now let us assume for simplicity that we run ULA with fixed step size $\gamma$.
+Then it is known that under assumption {\bf H1} corresponding Markov chain would have unique stationary distribution $\pi_\gamma$, which is not equal to $\pi$.
+Yet this chain will be $V-$geometrically ergodic due to \cite[Theorem~19.4.1]{douc:moulines:priouret:soulier:2018}.
+Namely, the following lemma holds:
+
+\begin{lem} Assume that the potential $U(x), x \in \mathbb{R}^d$ satisfies conditions {\bf H1} and {\bf H2}. Then for $0 < \gamma < \overline{\gamma}=\frac{m}{4L^2}$, for any $x \in X$ it holds
 $$
@@
 $$
-b = \gamma(3K_1^2 + 2D + C_1); \quad \overline{b} = b\exp^{-\frac{m\gamma}{2}} + d; \quad \varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right); \quad
-$$
-
-$$
 C = \left(1 + \exp^{-\frac{m\gamma}{2}}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp^{-\frac{m\gamma}{2}} - \frac{2b}{1+d})}\right);
@@
 $$
-\rho = \exp^{-\gamma\left(\frac{m}{2} - 2\frac{3K_1^2 + 2D + C_1}{d}\right)\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp^{-\frac{m\gamma}{2}} + \frac{2b}{d+1})}}}
+b = \gamma(\frac{25K_1^2}{8} + 2d + m); \quad \overline{b} = b\exp^{-\frac{m\gamma}{2}} + d; \quad \varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right); \quad
+$$
+
+$$
+\rho = \exp^{-\gamma\left(\frac{m}{2} - 2\frac{\frac{25K_1^2}{8} + 2d + m}{d}\right)\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp^{-\frac{m\gamma}{2}} + \frac{2b}{d+1})}}}
 $$
@@
 
-\begin{proof} Note that the condition {\bf H1} implies that the Markov kernel $Q^n_\gamma$ satisfies $(1,\varepsilon)$-Doeblin condition with $\varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right)$.
-Together with drift condition {\bf H2} it allows to apply \cite[Theorem~19.4.1]{douc:moulines:priouret:soulier:2018} with appropriate constants.
+\begin{proof} Note that the condition {\bf (H1)} implies that the Markov kernel $Q^n_\gamma$ satisfies $(1,\varepsilon)$-Doeblin condition with $\varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right)$.
+Together with drift condition \ref{eq:drift} it allows to apply \cite[Theorem~19.4.1]{douc:moulines:priouret:soulier:2018} with appropriate constants.
 \end{proof}
@@
 \begin{thm}
-\label{exp_mom_ula}
+\label{th:exp_mom_ula}
 Suppose that

### 2020-01-21 aap/Revision/vr-mcmc-revision.tex
update

@@
 \newcommand{\erici}[1]{\todo[inline,color=purple!10, author=Eric]{#1}}
-\newcommand{\denis}[1]{\todo[color=green!10, author=Eric]{#1}}
-\newcommand{\denisi}[1]{\todo[inline,color=green!10, author=Eric]{#1}}
+\newcommand{\denis}[1]{\todo[color=green!10, author=Denis]{#1}}
+\newcommand{\denisi}[1]{\todo[inline,color=green!10, author=Denis]{#1}}
+\newcommand{\sergey}[1]{\todo[color=red!10, author=Sergey]{#1}}
+\newcommand{\sergeyi}[1]{\todo[inline,color=red!10, author=Sergey]{#1}}
 
@@
 
-\begin{lem} Assume that the potential $U(x), x \in \mathbb{R}^d$ satisfies conditions {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$.
-Then the kernel $R_\gamma$ from \ref{eq:kernel} satisfies drift condition \ref{def:drift} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L^2}$ with drift function $V(x) = 1 + \|x\|^2$, constants $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = \frac{25K_1^2}{8} + 2d + m$ where $K_1$ is from \ref{eq:H3}, $m$ is from {\bf (H2)}.
+\begin{lem}
+\label{lem:drift}
+ Assume that the potential $U(x), x \in \mathbb{R}^d$ satisfies conditions {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$.
+Then the kernel $R_\gamma$ from \ref{eq:kernel} satisfies drift condition \ref{def:drift} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L^2}$ with drift function $V(x) = 1 + \|x\|^2$, constants $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = \frac{25K_1^2}{8} + 2d + m$ with $K_1$ from \ref{eq:H3}, $m$ from {\bf (H2)}.
 \end{lem}
@@
 For now let us assume for simplicity that we run ULA with fixed step size $\gamma$.
-Then it is known that under assumption {\bf H1} corresponding Markov chain would have unique stationary distribution $\pi_\gamma$, which is not equal to $\pi$.
+Then it is known that under assumption {\bf (H1)} corresponding Markov chain would have unique stationary distribution $\pi_\gamma$, which is different from $\pi$.
 Yet this chain will be $V-$geometrically ergodic due to \cite[Theorem~19.4.1]{douc:moulines:priouret:soulier:2018}.
@@
 
-\begin{lem} Assume that the potential $U(x), x \in \mathbb{R}^d$ satisfies conditions {\bf H1} and {\bf H2}. Then for $0 < \gamma < \overline{\gamma}=\frac{m}{4L^2}$, for any $x \in X$ it holds
-$$
-d_V(\delta_xQ^n_\gamma,\pi_\gamma) \leq C\rho^n\left(V(x) + \pi_\gamma(V)\right)
+\begin{lem}
+\label{lem:v_ergodicity}
+ Assume that the potential $U(x), x \in \mathbb{R}^d$ satisfies conditions {\bf (H1)} and {\bf (H2)}. Then for $0 < \gamma < \overline{\gamma}=\frac{m}{4L^2}$, for any $x \in X$ it holds
+$$
+d_V(\delta_xQ_{\boldsymbol{\gamma}}^{1,n},\pi_\gamma) \leq C\rho^n\left(V(x) + \pi_\gamma(V)\right)
 $$ 
@@
 $$
-C = \left(1 + \exp^{-\frac{m\gamma}{2}}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp^{-\frac{m\gamma}{2}} - \frac{2b}{1+d})}\right);
+C = \left(1 + \exp{\left(-\frac{m\gamma}{2}\right)}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp{\left(-\frac{m\gamma}{2}\right)} - \frac{2b}{1+d})}\right);
 $$
@@
 $$
-\rho = \exp^{-\gamma\left(\frac{m}{2} - 2\frac{\frac{25K_1^2}{8} + 2d + m}{d}\right)\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp^{-\frac{m\gamma}{2}} + \frac{2b}{d+1})}}}
+\rho = \exp{\left(-\gamma\left(\frac{m}{2} - 2\frac{\frac{25K_1^2}{8} + 2d + m}{d}\right)\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{m\gamma}{2}\right)} + \frac{2b}{d+1})}}\right)}
 $$
@@
 
-\begin{proof} Note that the condition {\bf (H1)} implies that the Markov kernel $Q^n_\gamma$ satisfies $(1,\varepsilon)$-Doeblin condition with $\varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right)$.
+\begin{proof} Note that the condition {\bf (H1)} implies that the Markov kernel $Q^{1,n}_{\boldsymbol{\gamma}}$ satisfies $(1,\varepsilon)$-Doeblin condition with $\varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right)$.
 Together with drift condition \ref{eq:drift} it allows to apply \cite[Theorem~19.4.1]{douc:moulines:priouret:soulier:2018} with appropriate constants.
 \end{proof}
+
+Aforementioned lemmas allow us to bound the exponential moment of the additive functional of ULA. Namely, the following theorem holds
+
+\begin{thm}
+\label{thm:ula_const_size}
+Suppose that ULA kernel \ref{eq:kernel} satisfies assumptions {\bf (H1)} and {\bf (H2)}, and let $X_1,\ldots,X_n,\ldots$ be generated by ULA with constant step size $\gamma$.
+Let $X_0 = x$ be fixed.
+Then for any bounded function $g(x) : \mathbb{R}^d \rightarrow \mathbb{R}, |g(x)| \leq M$, it holds
+\begin{eqnarray*}
+\mathsf{E}\left[\exp\left(-\rho\gamma\sum_{k=l}^{p}\left(g(X_{k})-\pi_{\gamma}(g)\right)\right)\right] \leq C_1\exp{\left(s^2 \kappa \gamma^2 (p-l+1)\right)}
+\end{eqnarray*}
+for some absolute constant $C_1$ not depending on $l,p$.
+\end{thm}
+
+\begin{proof} Conditions {\bf (H1)} and {\bf (H2)} imply that the Markov kernel $R_\gamma$ is $V-$ergodic (lemma \ref{lem:v_ergodicity}) and satisfies drift condition \ref{lem:drift}.
+Due to \cite[Theorem~3, Fact 3]{moulines:bound_dif:2019}, we obtain the following bound
+\begin{align*}
+\mathsf{E}_x\left[\exp\left(-s\gamma\sum_{k=l}^{p}\left(g(X_{k}) - \mathsf{E}_x g(X_k)\right)\right)\right] \leq \exp{\left(s^2 \kappa \gamma^2 (p-l+1)\right)}
+\end{align*}
+\sergey{I will put precise constant later}
+with constant $\kappa = ...$. Note that
+\begin{align*}
+\mathsf{E}\left[\exp\left(-s\gamma\sum_{k=l}^{p}\left(g(X_{k})-\pi_{\gamma}(g)\right)\right)\right] = 
+\end{align*}
+
+\begin{align*}
+=\mathsf{E}_x\left[\exp{\left(-s\gamma\sum_{k=l}^{p}\left(g(X_{k}) - \mathsf{E}_x g(X_k)\right)\right)}\right]
+\exp{\left(s\gamma\sum\limits_{k=l}^{p}\left(\pi_\gamma (g) - \mathsf{E}_x g(X_k)\right)\right)}
+\end{align*}
+Using lemma~\ref{lem:v_ergodicity},
+\begin{align*}
+|\mathsf{E} g(X_k) - \pi_\gamma(g)| \leq Md_V(\pi_\gamma, \delta_xQ^{1,n}_{\boldsymbol{\gamma}}) \leq CM\rho^k\left(V(x) + \pi_\gamma(V)\right)
+\end{align*}
+Hence,
+\begin{align*}
+\exp{\left(s\gamma\sum\limits_{k=l}^{p}\left(\pi_\gamma (g) - \mathsf{E}_x g(X_k)\right)\right)} \leq \exp{\left(s\gamma CM(V(x) + \pi_\gamma(V))\frac{\rho^l - \rho^{p+1}}{1-\rho}\right)}
+\end{align*}
+and the statement follows.
+\end{proof}
+
+Let us formulate and prove prove analogue of the previous theorem for inhomogeneous Markov chain with step size $\gamma_1,\ldots,\gamma_n,\ldots$.
+Note that under assumptions $\sum_{p=1}^{\infty} \gamma_p=\infty, \quad \sum_{p=1}^{\infty} \gamma^2_p<\infty$ and {\bf (H1)} the stationary distribution of ULA-based chain will be equal to $\pi$.
+We will need additional assumption
+\begin{description}
+\item[{\bf (H3)}] There exist such constants $\rho > 0, K_3 > 0$ that 
+\begin{eqnarray*}
+d_V(\delta_xQ_{\boldsymbol{\gamma}}^{1,n},\pi) \leq K_3 V(x) \rho^n 
+\end{eqnarray*}
+\end{description}
 
@@
 \end{eqnarray*}
-Under assumptions {\bf (H1)} and {\bf (H2)}, for any bounded function $g$ on $\mathbb{R}^d$ 
+Under assumptions {\bf (H1)}, {\bf (H2)} and {\bf (H3)}, for any bounded function $g$ on $\mathbb{R}^d$ 
 we have
-\begin{eqnarray*}
-\mathsf{E}\left[\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}g(X_{p})-\bar Q_l(X_l)\right)^{4}\right] \lesssim \frac{1}{\Gamma^4_{N+2,N+n+1}}
-\end{eqnarray*}
- and
-\begin{eqnarray*}
-\bar Q_l(X_l)=\mathsf{E}\left[\left.\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right|X_{l}\right].
-\end{eqnarray*}
-Moreover, 
-\begin{eqnarray*}
-\mathsf{E}\left[\exp\left(\rho\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k})-\left(\sum_{k=l}^{p}\gamma_{k+1}\right)\pi(g)\right)\right)\right]\leq C
-\end{eqnarray*}
-for some small enough $\rho>0,$  all natural $0<l\leq p<\infty$ and constant $C$ not depending on $l,p.$
+%\begin{eqnarray*}
+%\mathsf{E}\left[\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}g(X_{p})-\bar Q_l(X_l)\right)^{4}\right] \lesssim \frac{1}{\Gamma^4_{N+2,N+n+1}}
+%\end{eqnarray*}
+ %and
+%\begin{eqnarray*}
+%\bar Q_l(X_l)=\mathsf{E}\left[\left.\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right|X_{l}\right].
+%\end{eqnarray*}
+%Moreover, 
+\begin{eqnarray*}
+\mathsf{E}\left[\exp\left(-s\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k})-\left(\sum_{k=l}^{p}\gamma_{k+1}\right)\pi(g)\right)\right)\right]\leq C
+\end{eqnarray*}
+for all natural $0<l\leq p<\infty$ with constant $C$ not depending on $l,p.$
 \end{thm}
 \begin{proof}
- Note that 
-\[
-\mathsf{E}\left[e^{-s\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})}\right]=\mathsf{E}\left[e^{-s\left(\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})-\Gamma_{l,j}\int\mu'(x)\pi(x)\,dx\right)}\right]e^{-s\Gamma_{l,j}\int\mu'(x)\pi(x)\,dx}
-\]
- with $\Gamma_{l,j}=\sum_{k=l}^{j}\gamma_{k}.$ According to \cite[Theorem~3]{bms:exp_conc_note}, it holds
-\[
-\mathsf{E}\left[e^{-s\left(\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})-\Gamma_{l,j}\int\mu'(x)\pi(x)\,dx\right)}\right]\leq e^{s\varkappa\sum_{k=l}^{j}\gamma_{k}^{2}}
-\]
- for some $\varkappa>0.$ Since $\sum_{k=1}^{\infty}\gamma_{k}^{2}\leq\Gamma^{\star},$we
-have 
-\[
-\mathsf{E}\left[e^{-s\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})}\right]\leq e^{s\varkappa\Gamma^{\star}}e^{-s\Gamma_{l,j}\int\mu'(x)\pi(x)\,dx}.
-\]
-Furthermore, since $\mu\pi',\mu'\pi\in L^{1}(\mathbb{R}),$ we have
-\[
-\int\mu'(x)\pi(x)\,dx=-\int\mu(x)\pi'(x)\,dx=2\int\mu^{2}(x)\pi(x)\,dx>0
-\]
-and as a result 
-\[
-\zeta_{l,j}(s)\leq e^{s\varkappa\Gamma^{\star}}e^{-2s\Gamma_{l,j}\int\mu^{2}(x)\pi(x)\,dx}
-\]
- for any $l\leq j.$
+
+ Note that under assumptions {\bf (H1)}, {\bf (H2)} and {\bf (H3)} it holds due to \cite[Theorem~4]{moulines:bound_dif:2019} that
+ \begin{align*}
+\mathsf{E}_x\left[\exp\left(-s\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k}) - \mathsf{E}_x \left(\sum\limits_{k=l}^{p}\gamma_{k+1}g(X_k)\right)\right)\right)\right] \leq \exp{\left(s^2 \kappa \sum\limits_{k=l}^{p}\gamma_{k+1}^2\right)} \leq \exp{\left(s^2 \kappa \Gamma^*\right)}
+\end{align*}
+\sergey{This theorem is not yet written in the second paper}
+with $\Gamma^* = \sum\limits_{k=1}^{\infty}\gamma_k^2 < \infty$ and $\kappa = ...$. Using the same transformation as in theorem~\ref{thm:ula_const_size}, and using the bound
+\begin{align*}
+|\mathsf{E}_x f(X_k) - \pi(f)| \leq K_3 M \rho^{\Gamma_{1,k}}V(x)
+\end{align*}
+we obtain
+\begin{align*}
+\mathsf{E}\left[\exp\left(-s\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k})-\left(\sum_{k=l}^{p}\gamma_{k+1}\right)\pi(g)\right)\right)\right] \leq \exp{\left(s^2 \kappa \Gamma^*\right)}\exp{\left(K_3 M V(x) \sum\limits_{k=l}^{p}\rho^{\Gamma_{1,k}}\gamma_{k+1}\right)} \leq
+\end{align*}
+
+\begin{align*}
+\leq \exp{\left(s^2 \kappa \Gamma^*\right)}\exp{\left(K_3 M V(x) \rho^{\Gamma_{1,k}} \int\limits_{0}^{+\infty}\rho^y\,dy\right)} = \exp{\left(s^2 \kappa \Gamma^*\right)}\exp{\left(K_3 M V(x) \frac{\rho^{\Gamma_{1,k}}}{\ln{\rho}}\right)}
+\end{align*}
 \end{proof}
 
-Suppose that now we use ULA with fixed step size, then we might use the following version of the previous theorem:
-\begin{thm} Suppose that we use ULA with constant step size $\gamma$.
-Under assumptions {\bf (H1)} and {\bf (H2)}, for any bounded function $g$ on $\mathbb{R}^d$ we have, given $0 < \gamma < \left(\frac{\int \mu^{\prime}(x)\pi(x)\,dx}{C_1 M}\right)^2$,
-\begin{eqnarray*}
-\mathsf{E}\left[\exp\left(\rho\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k})-\left(\sum_{k=l}^{p}\gamma_{k+1}\right)\pi(g)\right)\right)\right]\leq C
-\end{eqnarray*}
-for some absolute constant $C$ not depending on $l,p$.
-\end{thm}
-
-\begin{proof} Note that 
-\begin{eqnarray*} 
-\int \mu^{\prime}(x)\pi_{\gamma}(x)\,dx \geq \int \mu^{\prime}(x)\pi(x)\,dx - M d_{TV}(\pi,\pi_{\gamma}) \geq \int \mu^{\prime}(x)\pi(x)\,dx - C_1\sqrt{\gamma}M
-\end{eqnarray*}
-where the precise expression for constant $C_1$ might be found in \cite[Theorem~10]{durmus:moulines:aop:2016}.
-Thus, for $\gamma < \left(\frac{\int \mu^{\prime}(x)\pi(x)\,dx}{C_1 M}\right)^2$, we obtain along the lines of the proof of Theorem~16 that
-$$
-\int \mu^{\prime}(x)\pi_{\gamma}(x)\,dx > 0
-$$
-
-\end{proof}
+The proof is completed.
 

### 2020-01-23 aap/Revision/vr-mcmc-revision.tex
update

@@
 %Moreover, 
-\begin{eqnarray*}
+\begin{eqnarray}
+\label{eq:exp_decrease_step}
 \mathsf{E}\left[\exp\left(-s\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k})-\left(\sum_{k=l}^{p}\gamma_{k+1}\right)\pi(g)\right)\right)\right]\leq C
-\end{eqnarray*}
-for all natural $0<l\leq p<\infty$ with constant $C$ not depending on $l,p.$
+\end{eqnarray}
+for all natural $0<l\leq p<\infty$ with constant $C$ not depending on $l,p.$. Moreover, 
+\begin{eqnarray}
+\label{eq:var_decrease_step}
+\mathsf{Var}_x{\left(\sum_{k=l}^{l+n-1}\frac{\gamma_{k+1}}{\Gamma_{l+1,l+n}}g(X_k) - \mathsf{E}_x \sum_{k=l}^{l+n-1}\frac{\gamma_{k+1}}{\Gamma_{l+1,l+n}}g(X_k)\right)} \leq \frac{C_1}{\Gamma_{l+1,l+n}}
+\end{eqnarray}
 \end{thm}
@@
  Note that under assumptions {\bf (H1)}, {\bf (H2)} and {\bf (H3)} it holds due to \cite[Theorem~4]{moulines:bound_dif:2019} that
- \begin{align*}
+ \begin{align}
+ \label{eq:exp_bounded_dif}
 \mathsf{E}_x\left[\exp\left(-s\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k}) - \mathsf{E}_x \left(\sum\limits_{k=l}^{p}\gamma_{k+1}g(X_k)\right)\right)\right)\right] \leq \exp{\left(s^2 \kappa \sum\limits_{k=l}^{p}\gamma_{k+1}^2\right)} \leq \exp{\left(s^2 \kappa \Gamma^*\right)}
-\end{align*}
+\end{align}
 \sergey{This theorem is not yet written in the second paper}
@@
 \end{proof}
-
-The proof is completed.
-
+which proves \ref{eq:exp_decrease_step}. Note that from \ref{eq:exp_bounded_dif} by Taylor expansion for small $s$ we obtain
+\begin{align*}
+1 + \frac{s^2}{2}\mathsf{Var}_x{\left(\sum_{k=l}^{l+n-1}\gamma_{k+1}g(X_k) - \mathsf{E}_x \sum_{k=l}^{l+n-1}\gamma_{k+1}g(X_k)\right)} + O(s^3) \leq 1 + s^2 \kappa \sum\limits_{k=l+1}^{l+n}\gamma_{k}^2 + O(s^4)
+\end{align*}
+Since it holds for arbitarily small $s$, the variance is bounded by
+\begin{align*}
+\mathsf{Var}_x{\left(\sum_{k=l}^{l+n-1}\gamma_{k+1}g(X_k) - \mathsf{E}_x \sum_{k=l}^{l+n-1}\gamma_{k+1}g(X_k)\right)} \leq 2\kappa \sum\limits_{k=l+1}^{l+n}\gamma_{k}^2 \leq 2\kappa \Gamma_{l+1,l+n}
+\end{align*}
+Now we need to divide both parts by $\Gamma^2_{l+1,l+n}$ to get \ref{eq:var_decrease_step}. The proof is completed.
 \bibliographystyle{plain}

### 2020-01-23 aap/Revision/vr-mcmc-revision.tex
update

@@
 \end{eqnarray*}
-Obviously \(\mathsf{E}[\widehat M_{K,n}^N(f)|\TrainSet]=0\) and \(\widehat M_{K,n}^N(f)\) is indeed a valid control variate in that it
+Obviously \(\mathsf{E}[\widehat M_{K,n}^N(f)|\TrainSet]=0\) and \(\widehat M_{K,n}^N(f)\) is indeed a valid control variate in a sense that it
 does not introduce any bias.

### 2020-01-24 aap/Revision/vr-mcmc-revision.tex
update

@@
 \end{proof}
-\begin{remark}\label{rem:p-norm}
-In fact in the proof of Lemma~\ref{lem:var_poincare} using \eqref{eq:part-part} we have shown that the expectations
-\begin{eqnarray*}
-\mathsf E\left[\left.
-\|\nabla_Z F(X_{l-1},Z_l,Z_{l+1},\ldots,Z_q)\|^r
-\right|X_{l-1}\right]
-\end{eqnarray*}
-are uniformly bounded for all natural \(l\leq q\) and any \(r\geq 2.\) Indeed it follows from \eqref{eq:17042018a1} that  the quantities
-\begin{eqnarray*}
-\left(\sum_{j=l}^{q} \gamma_{j}
-\prod_{k=l}^{j} \alpha_{k}\right)^{r/2}
-\left(
-\sum_{p=j+1}^{q+1} \gamma_{p} \prod_{k=j+1}^{p} \alpha_{k}
-\right)^{r}
-\end{eqnarray*}
-are bounded.
-This stronger result in combination with a \(L^r\)-Sobolev-type inequalities (see e.g.
-\cite{Adamczak}) implies that also higher order conditional (on \(X_{l-1}\)) central moments of the r.v.
-\begin{eqnarray*}
-\sum_{p=l}^{q}\gamma_{p+1}\partial_{y_{l}}^{K}f\left(X_{p}\right)
-\end{eqnarray*}
-are uniformly bounded in \(l\leq q\) and \(p.\)
-
-\end{remark}
 

### 2020-01-28 aap/Revision/vr-mcmc-revision-new.tex
constant step size

--- 2020-01-24 aap/Revision/vr-mcmc-revision.tex
@@
 \label{seq:mart_repr}
-In this section we give a general discrete-time martingale representation for the Markov chain \eqref{eq:chain_gen} which is  used below to construct an efficient variance reduction algorithm.
+In this section we give a general discrete-time martingale representation for  Markov chains of the type  \eqref{eq:chain_gen} which is  used below to construct an efficient variance reduction algorithm.
 Let \((\phi_k)_{k\in \mathbb{Z}_+}\) be a complete orthonormal system in \(L^2(\mathbb{R}^m, P_{\xi})\) with \(\phi_0\equiv 1\).
@@
  Let $f$ be a Borel function $\mathbb{R}^{d}\to\mathbb R$ such that
-$\mathsf{E}\left[\left|f(X_{p})\right|^{2}\right]<\infty$, with a Markov chain $(X_{p})_{p\geq 0}$ defined in \eqref{eq:chain_gen}.
+$\mathsf{E}\left[\left|f(X_{p})\right|^{2}\right]<\infty$,   where $(X_{p})_{p\geq 0}$ is a Markov chain of the form \eqref{eq:chain_gen} for some kernels $(\Phi_l)_{l\geq 1}.$  
 Then,
@@
 \end{thm}
+\begin{cor}
+If  all the kernels $\Phi_l,$ $l\geq 1,$ are equal,  then  the representation \eqref{eq:mart_repr} takes the form 
+\begin{eqnarray*}
+f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}\bar a_{p-l,k}(X_{l-1})\phi_k\left(\xi_{l}\right)
+\end{eqnarray*}
+with 
+\begin{eqnarray*}
+\bar a_{r,k}(x)=\mathsf{E}\left[\left.f(X_{r})\phi_k\left(\xi_{1}\right)\right|X_{0}=x\right], \quad r, k\in \mathbb{N}.
+\end{eqnarray*}
+
+\end{cor}
 \begin{proof}
@@
 %where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
-The functions \((Q_{p,l})_{l=0}^p\)  can be computed by the backward recurrence: $Q_{p,p}(x)=f(x)$ and for $l \in \{0,\cdots,p-1\}$
+The functions \((Q_{p,l})_{l=0}^p\)  can be computed by the backward recurrence: $Q_{p,p}(x)=f(x)$ and for $l \in \{0,\ldots,p-1\}$
 \begin{eqnarray}
@@
 \end{eqnarray}
+In the case $\Phi_l=\Phi$ for all $l\geq 1,$ we have 
+\begin{eqnarray*}
+\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
+\end{eqnarray*}
+with  \(Q_{r}(x)=\mathsf{E}\left[\left.f(X_{r})\right|X_{0}=x\right],\) $r\in \mathbb{N}.$
 \end{prop}
 \par
-Next we show how  the representation \eqref{eq:mart_repr} can be used to efficiently reduce the variance of MCMC algorithms.
-Let $(\gamma_{p})_{p\in\mathbb N}$ be a sequence of positive and non-increasing
-step sizes
-with $\sum_{p=1}^\infty \gamma_p=\infty$
-and, for $n,l\in\mathbb{N}$, $n\le l$, we set
-\[
-\Gamma_{n,l}=\sum_{p=n}^{l}\gamma_{p}.
-\]
- Consider a weighted average estimator $\pi_{n}^{N}(f)$ of the form
+Next we show how  the representation \eqref{eq:mart_repr} can be used to  reduce the variance of MCMC algorithms.
+Consider the case of time homogeneous transition kernels and introduce a weighted average estimator $\pi_{n}^{N}(f)$ of the form
 \begin{equation}\label{eq:29032018a2}
-\pi_{n}^{N}(f)=\sum_{p=N+1}^{N+n}\omega_{p,n}^{N}f(X_{p}),\quad\omega_{p,n}^{N}=\gamma_{p+1}\Gamma_{N+2,N+n+1}^{-1},
+\pi_{n}^{N}(f)=\frac{1}{n}\sum_{p=N+1}^{N+n} f(X_{p}),
 \end{equation}
@@
 Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
-\begin{equation}
-M_{K,n}^N(f) =\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\bar a_{l,k}(X_{l-1})\phi_k(\xi_{l}),
+\begin{eqnarray}
+\nonumber
+M_{K,n}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=1}^{p} \bar a_{p-l,k}(X_{l-1})\phi_k(\xi_{l})\right]
+\\
+&=& \sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left(1+\frac{N-l}{n}\right)A_{N+n-l,k}(X_{l-1})\phi_{k}\left(\xi_{l}\right),
 \label{eq:29032018a5}
-\end{equation}
+\end{eqnarray}
 where 
-\begin{align}
-\label{eq:first-expression-bar-a-l-k}
-\bar{a}_{l,k}(x)
-=\sum_{p=l}^{N+n}\omega_{p,n}^{N}a_{p,l,k}(x)=\mathsf{E}\left[\left.\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right)\phi_k(\xi_{l})\right|X_{l-1}=x\right].
-\end{align}
-Since \(X_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\geq 1,\)  the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
+\[
+A_{s,k}(x)=\frac{1}{s}\sum_{r=0}^{s}\bar a_{r,k}(x).
+\]
+Since \(X_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq l,\)  the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
 The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
@@
 \end{equation}
-should be small for \(K\) large enough. In this section   we provide  an analysis of the variance-reduced ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
+should be small for \(K\) large enough. Indeed, due to orthonormality  we have
+\[
+\mathsf{Var}[\pi_{K,n}^N(f)]\leq\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\mathsf{E}[A^2_{n-l,k}(X_{N+l-1})]
+\] 
+and $\mathsf{Var}[\pi_{K,n}^N(f)]$ is small provided that the coefficients $A_{s,k}$ decay fast enough as $k\to \infty.$
+In the next section   we provide a detailed   analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
 \section{Analysis of variance reduced ULA}
 \label{sec:ula_analysis}
-
+In this section we perform the convergence analysis of the ULA algorithm.
+For the sake of clarity and notational simplicity we restrict our attention to the constant time step, that is, we take $\gamma=\gamma_1=\gamma_2=\ldots.$
 We shall use the notations $\mathbb N=\{1,2,\ldots\}$ and $\mathbb N_0=\mathbb N\cup\{0\}$.
@@
 \begin{align}
-M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{p=N+1}^{N+n}\omega_{p,n}^{N}\sum_{l=N+1}^{p}a_{p,l,\mathbf{k}}(X_{l-1})\mathbf{H}_\mathbf{k}(Z_{l})
-\label{eq:29032018a5}\\
-&=\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{l=N+1}^{N+n}\bar{a}_{l,\mathbf{k}}(X_{l-1})\mathbf{H}_\mathbf{k}(Z_{l})
+M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{l=1}^{n}\Bigl(1-\frac{l}{n}\Bigr)A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})
+\label{eq:29032018a5}
 \notag
 \end{align}
-with \(\|\mathbf{k}\|=\max_{i} k_i\) and
-\begin{align*}
-\bar{a}_{l,\mathbf{k}}(x)
-& =\sum_{p=l}^{N+n}\omega_{p,n}^{N}a_{p,l,\mathbf{k}}(x) =\mathsf{E}\left[\left.\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right)\mathbf{H}_\mathbf{k}\left(Z_{l}\right)\right|X_{l-1}=x\right].
-\end{align*}
+with \(\|\mathbf{k}\|=\max_{i} k_i.\) 
 For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
@@
 \end{eqnarray*}
-\item[(H2)][\textbf{Convexity outside a ball}] There exist $K_U>0,$ $M_U>0$ and $m_U>0$ such that for any $x\not\in B_{K_U}(0)	$ and any $y\in \mathbb{R}^d$ it holds 
-\begin{eqnarray*}
-\langle D^2U(x)y,y\rangle \geq m_U\|y\|^2, \quad \|D^3 U(y)\| \leq M_U. 
+\item[(H2)][\textbf{Convexity outside a ball}] There exist $K_U>0,$ $M_U>0$ and $m_U>0$ such that for any $x\not\in B_{K_U}(0)$  it holds 
+\begin{eqnarray*}
+\langle D^2U(x),x\rangle \geq \frac{m_U}{2}\|x\|^2. 
 \end{eqnarray*}
@@
 &\le
-\mathrm{osc}(f)^2
-\sum_{p=N+1}^{N+n}
-\omega_{p,n}^N\,
-\|Q_\gamma^{p}(x_0,\cdot)-\pi(\cdot)\|_{TV}^2,
+\frac{\mathrm{osc}(f)^2}{n}\sum_{p=N+1}^{N+n}
+\|\delta_x Q_\gamma^{p}-\pi(\cdot)\|_{TV}^2,
 \notag
@@
 with density $\pi$ of~\eqref{eq:stationary_distr};
-for $\gamma>0$, the Markov kernel $R_\gamma$
+for $\gamma>0$, the Markov kernel $Q_\gamma$
 is defined by
 $$
-R_\gamma(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\}dy,
-$$
-while, for $k,l\in\mathbb N$, $k\le l$,
-the kernel $Q_\gamma^{k,l}$ is
-$
-Q_\gamma^{k,l}=R_{\gamma_l}\cdots R_{\gamma_k},
-$
-which, finally, provides the  measure
-$Q_\gamma^{p}(x_0,\cdot)=Q_\gamma^{p,1}$
-used in the right-hand side of~\eqref{eq:06042018a2}.
+\mathcal{Q}_\gamma(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\}dy.
+$$
 Different specific upper bounds for the squared bias
@@
 using results of Section~3 in~\cite{durmus:moulines:2017}
-on bounds in the total variation distance. For example, the following proposition holds.
-\begin{prop}
-Suppose that the assumption (H1) holds, then it holds
-\begin{eqnarray*}
-\|Q_\gamma^{p}(x,\cdot)-\pi(\cdot)\|^2_{TV}\lesssim \sum_{k=r+1}^{p} \gamma_k^2+\exp\left(-\varkappa\sum_{k=r+1}^{p} \gamma_k\right),
-\end{eqnarray*}
-for some $\varkappa>0,$ where $\lesssim$ stands for  inequality up to a constant depending on $L$ and $\mathcal G_N.$ In particular if $\gamma_k=k^{-\alpha}$ for some $\alpha\in (0.5,1], $ then 
-\begin{eqnarray*}
-\|Q_\gamma^{N+1,N+n}(x,\cdot)-\pi(\cdot)\|^2_{TV}\lesssim \frac{1}{(N+1)^{2\alpha-1}}
-\end{eqnarray*}
-provided that $N/n\to 0$ as $N,n\to \infty.$
-\end{prop}
-\subsection{Conditional variance}
+on bounds in the total variation distance.
+It is known  that under assumptions {\bf (H1)} and {\bf (H2)}  the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$.
+Moreover, as shown in Proposition (Appendix) \begin{eqnarray}
+\label{eq:bias}
+\|\delta_x\mathcal{Q}_{\gamma}^{n}-\pi_\gamma\|_{TV} \leq C_1\rho^{\gamma n}\left(V(x) + \pi_\gamma(V)\right), \quad \|\pi-\pi_\gamma\|_{TV} \leq C_2 \sqrt{\gamma}
+\end{eqnarray}
+for  $V(x) = 1 + \|x\|^2,$ some $\rho\in (0,1)$ and  constants $C_1,$ $C_2>0$ independent of $\gamma$  and $n.$
+\subsection{Variance}
 An upper bound for the variance of the classical estimator~\eqref{eq:29032018a2}
-is provided in  in~\cite[Theorem~17]{durmus:moulines:2017}.
-As for the estimator~\eqref{eq:29032018a3}, 
-it follows from~\eqref{eq:29032018a5}, Theorem~\ref{prop:29032018a1} and independence condition for the sequence $(Z_p)_{p \geq 1}$.
-\begin{prop}
-\label{cond_var_dec}
-	For variance-reduced estimator~\eqref{eq:29032018a3}, defined for the Markov Chain~\eqref{eq:chain_gen} the following representation holds
-\begin{equation}\label{eq:29032018a6}
-\mathsf{Var}\left[\pi_{K,n}^{N}(f)\right]
-%=\sum_{\min_i (k_i)\ge K+1}
-=\sum_{\mathbf k\colon \|\mathbf k\|\ge K+1}
-\sum_{l=N+1}^{N+n}
-\mathsf{E}\left[\left|\bar a_{l,\mathbf{k}}(X_{l-1})\right|^{2}\right].
-\end{equation}
-\end{prop}
-Now we present an upper bound for the right-hand sides of~\eqref{eq:29032018a6} in case of ULA algorithm.
+under {\bf (H1)} and {\bf (H2)} is derived in Section.  In particular, for any bounded function $f,$
+\begin{eqnarray}
+\label{eq:var-mc}
+\mathsf{Var}[\pi_n^N(f)]\lesssim \frac{1}{n\gamma},
+\end{eqnarray}
+where $\lesssim$ stands for inequality up to a constant depending on constants $L_U,$ $m_U,$  $M_U$ and $\|f\|_\infty.$ 
+One of the main results of this paper is the following  upper bound for the variance of $\pi_{K,n}^{N}(f).$ 
 \begin{thm}\label{th:mr}
-Assume (H1) and (H2).
-Suppose that $f$  is two times continuously differentiable
+Assume {\bf (H1)} and {\bf (H2)}.
+Suppose additionally that a bounded function $f$ and $\mu=\nabla U$ are $K\geq 2$ times continuously differentiable
 and it holds
-\begin{eqnarray*}
-|\partial^{\mathbf{k}} f(x)|\le  B_f, \quad
+\begin{eqnarray}
+\label{eq:smooth-mu}
+|\partial^{\mathbf{k}} f(x)|\le  B_f, \quad |\partial^{\mathbf{k}} \mu (x)|\leq B_\mu, \quad 
 x\in\mathbb R^d
-\end{eqnarray*}
-for all \(\mathbf{k}\) with \(0<\|\mathbf{k}\|\leq  2.\) Let $(\gamma_k)_{k\in\mathbb N}$ be a sequence of positive
-and non-increasing step sizes with $\sum_{k=1}^\infty \gamma_k=\infty$ and $\sum_{k=1}^\infty \gamma^2_k<\infty.$ 
-Then it holds under 
+\end{eqnarray}
+for all \(\mathbf{k}\) with \(0<\|\mathbf{k}\|\leq  K.\)  Then it holds 
 \begin{eqnarray}
 \label{eq:var-bound}
-\mathsf{Var}\left[\pi_{K,n}^{N}(f)\right]\lesssim \frac{1}{\Gamma^2_{N+2,N+n+1}},
+\mathsf{Var}\left[\pi_{K,n}^{N}(f)\right]\lesssim \frac{\gamma^{K-2}}{n},
 \end{eqnarray}
-where \(\lesssim\) stands for inequality up to a constant not depending  on \(n\) and \(N.\)
+where \(\lesssim\) stands for inequality up to a constant not depending  on $\gamma,$ \(n\) and \(N.\)
 \end{thm}
-\begin{cor}
-
-\end{cor}
-
-\section{Nonparametric regression}
-The functions \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. This amounts to solve a nonparametric regression problem.
-We present a generic regression algorithm. Algorithm starts with estimating the functions $\bar{Q}_{l}$ for $l=N+1,\ldots,N+n,$ where
-\begin{equation*}
-\bar{Q}_{l}(x)=\sum_{p=l}^{N+n} \omega_{p,n}^{N} Q_{p,l}\left(x\right)=\omega_{l,n}^{N}f(x)+\mathsf{E}\left[\left.\sum_{p=l+1}^{N+n}\omega_{p,n}^{N}f(X_{p})\right|X_{l}=x\right] .
-\end{equation*}
-We first generate \(\NtrainPath\)  paths conditionally independent of the $\sigma$-algebra generated by the burn-in sequence $X_1,\dots,X_N$:
-\[
-\TrainSet=\Bigl\{(X^{(s)}_{N+1},\ldots,X^{(s)}_{N+n}),\quad s=1,\ldots,\NtrainPath\Bigr\}
-\]
-of the chain \(X\)
-(the so-called ``training paths'').
-The regression algorithm proceeds with estimating the  functions \((\bar Q_{l})\) by solving  the least squares optimization problems
-\begin{equation}\label{eq:06042018a1}
-{\widehat  Q}_{l}=\argmin_{\psi\in \Psi}\sum_{s=1}^{\NtrainPath} \left | \sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X^{(s)}_{p})-\psi(X^{(s)}_{l})\right|^2
-\end{equation}
-for \(l=N+1,\ldots, N+n-1.\)
-where \(\Psi\) is a  class of  functions on \(\mathbb{R}^d.\)
-Next  we estimate the coefficients \( \bar a_{l,k}\) using
-the formula
-\begin{equation}
-\label{eq:a_est_int}
-\widehat a_{l,k}(x)=\mathsf{E}\left[\left.\phi_k\left(\xi\right)\widehat Q_{l}\left(\Phi_l(x,\xi)\right)\right | \TrainSet\right] \,,
-\end{equation}
-where $\xi$ is a random variable independent from $\mathcal{T}$ with law $P_\xi,$
-$\Phi_l$ is defined in \eqref{eq:chain_gen} and the expectation is taken according to known distribution \(P_\xi\).
-Conditioning on $\TrainSet$ underlines that estimation of $\widehat Q_{l}$ derived from training trajectories.
-In many cases integration can be done in closed analytical form.
-
- For example, in the case of ULA algorithm we have \(P_\xi$ is the multivariate normal distribution $ \mathcal{N}(0,I_d)\) and \(\Phi_l(x,\xi)=x-\gamma_{l}\mu(x)+\sqrt{\gamma_{l}}\xi.\) Therefore if we use polynomials to approximate \((\bar Q_{l}),\) then \((\widehat a_{l,k})\) can be even computed in closed form; implementations details are provided in  Section~\ref{sec:num}.
-
-Upon estimating  the coefficients \((\widehat a_{l,k})\), one can construct an empirical estimate  of \(M_{K,n}^N(f)\) in the form
-\begin{eqnarray*}
-\widehat M_{K,n}^N(f) =\sum_{1\leq  k \le K}\sum_{l=N+1}^{N+n}\widehat a_{l,k}(X_{l-1})\phi_k(\xi_{l}).
-\end{eqnarray*}
-Obviously \(\mathsf{E}[\widehat M_{K,n}^N(f)|\TrainSet]=0\) and \(\widehat M_{K,n}^N(f)\) is indeed a valid control variate in a sense that it
-does not introduce any bias.
-
-\section{Complexity analysis for ULA}
-\label{sec:coeff}
-
-The following result quantifies the error of estimating the functions \((\bar Q_l)\)
-via the algorithm and its proof follows from Theorem~2.2 in \cite{audibert2011robust}.
-\begin{thm}\label{berm:theorem:regression_cv}
- Suppose that for any \(l\in \{N+1,\ldots,N+n+1\},\)
-\[
-\mathsf{E}\left[\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})-\bar Q_l(X_l)\right)^4\right]\leq \sigma_l^4,
-\]
-for some finite positive numbers \(\sigma_{N+1},\ldots, \sigma_{N+n}. \)
-Furthermore, assume that \(\Psi=\mathrm{span}\{\psi_1,\ldots,\psi_D\},\) where the functions  \(\psi_1,\ldots,\psi_D\) are uniformly bounded and satisfy
-\[
-\max_{l}\sup_{g\in \Psi\setminus \{0\}} \|g\|^2_\infty/\mathsf{E}[g^2(X_l) | \mathcal{G}_N]\leq B<\infty.
-\]
-Then for any values of \(\varepsilon\) and \(\NtrainPath\) such that \(2/\NtrainPath\leq \varepsilon\leq 1\) and
-\begin{eqnarray*}
-\NtrainPath\gtrsim B^2\left[BD+\log(2/\varepsilon)+\frac{B^2D^2}{\NtrainPath}\right]
-\end{eqnarray*}
-it holds with probability at least \(1-\varepsilon,\)
-\begin{multline}
-\mathsf{E}\left[ \left.\Bigl|\bar Q_l(X_l)
--\widehat Q_l(X_l)\Bigr|^2\right | \TrainSet \vee \mathcal{G}_N\right]
-\lesssim  \sigma^2_l B
-\left(\frac{BD+\log(2/\varepsilon)}{\NtrainPath}+\frac{B^2D^2}{\NtrainPath^2}\right)
-\\
- +
-\inf_{\psi\in \Psi}
-\mathsf{E}\left[\left.\left|\bar Q_l(X_{l})-\psi(X_{l})\right|^{2}\right | \mathcal{G}_N \right], \label{eq:regr_error}
-\end{multline}
-for \(l=N+1,\ldots,N+n,
-\) with \(\lesssim\) standing for inequality up to a universal multiplicative constant.
-\end{thm}
-Now we are able to give  a bound for the difference between \( M_{K,n}^N\) and \(\widehat M_{K,n}^N.\)
-\begin{prop}
-\label{cor:dif_m}
-Under  conditions of Theorem~\ref{berm:theorem:regression_cv}, we have with probability at least \(1-\varepsilon,\)
-\begin{align}
-\label{eq:main_bound}
-&\mathsf{E}\left[\left.\left|\widehat{M}_{K,n}^{N}(f)-M_{K,n}^{N}(f)\right|^{2}\right | \TrainSet,\mathcal{G}_N\right] \\
-\nonumber
-&\phantom{\mathsf{E}\widehat{M}_{K,n}^{N}}  \lesssim K\left[\sum_{l=N+1}^{N+n} \sigma^2_l \right]
-B
-\left(\frac{BD+\log(2/\varepsilon)}{\NtrainPath}+\frac{B^2D^2}{\NtrainPath^2}\right)
-\\ \nonumber
-& \phantom{\mathsf{E}|\widehat{M}_{K,n}^{N}} + K \sum_{l=n+1}^{N+n}\inf_{\psi\in \Psi}
-\mathsf{E}\left[\left.\left|\bar Q_l(X_{l})-\psi(X_{l})\right|^{2}\right | \mathcal{G}_N \right].
-\end{align}
-\end{prop}
-\begin{proof}
-Using the conditional Cauchy-Schwarz inequality and  orthonormality of \((\phi_k)_{k\geq 0}\), we derive
-\begin{eqnarray*}
-\mathsf{E}\left[\left.\left|\widehat{a}_{l,k}(X_{l-1})-\bar a_{l,k}(X_{l-1})\right|^{2}\right | \TrainSet \vee \mathcal{G}_N\right]\leq\mathsf{E}\left[\left.\left|\widehat{Q}_{l}\left(X_{l}\right)-\bar Q_{l}\left(X_{l}\right)\right|^{2}\right | \TrainSet \vee \mathcal{G}_N\right]
-\end{eqnarray*}
-By the Jensen inequality and orthonormality of \((\phi_k)_{k\geq 0},\)
-\begin{multline*}
-\mathsf{E}\left[\left.\left|\widehat{M}_{K,n}^{N}(f)-M_{K,n}^{N}(f)\right|^{2}\right | \TrainSet \vee \mathcal{G}_N\right]
-\\
-\leq \sum_{1\leq k\le K}\sum_{l=N+1}^{N+n}\mathsf{E}\left[\left.\left|\widehat{a}_{l,k}(X_{l-1})-\bar a_{l,k}(X_{l-1})\right|^{2}\right | \TrainSet \vee \mathcal{G}_N\right].
-\end{multline*}
-\end{proof}
-\subsection{Complexity analysis of variance reduction for the ULA algorithm}
-In the case of ULA one can bound the quantities \(\sigma_l^2\) using  \(L^p\)-type Sobolev inequalities  (see \cite{Adamczak}), see Remark~\ref{rem:p-norm} in Section~\ref{sec:proofs}.
-In particular, we can derive that  under conditions of Theorem~\ref{th:mr} with \(K=1,\)
-\begin{eqnarray*}
-\sigma_l^2\lesssim \frac{1}{\Gamma^2_{N+2,N+n+1}},\quad l=N+1,\ldots,N+n+1,
-\end{eqnarray*}
-where \(\lesssim \) stands for the inequality up to a multiplicative constant not depending on \(n\) and \(N.\) Using this inequality and combining \eqref{eq:main_bound} with Theorem~\ref{th:mr}, we get for the variance of \(\widehat \pi_{K,n}^N(f)=\pi_n^N(f)-\widehat M_{K,n}^N(f),\) with probability at least \(1-\varepsilon,\)
+Let us sketch the main steps of the proof. First using integration by parts we prove that 
 \begin{eqnarray}
-\nonumber
-\mathsf{Var}\left[\left.\widehat \pi_{K,n}^N(f)\right | \TrainSet, \mathcal{G}_N\right]&\lesssim & \frac{n KB}{\Gamma^2_{N+2,N+n+1}}
-\left(\frac{BD+\log(2/\varepsilon)}{\NtrainPath}+\frac{B^2D^2}{\NtrainPath^2}\right)
-\\
-\nonumber
-&& + \frac{1}{\Gamma^2_{N+2,N+n+1}}\sum_{l=N+1}^{N+n}\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
-\left(\frac{\gamma_{l}}{2}\right)^{|I|K}
-\\
-\label{eq:ula_red_var}
-&& +
-K \sum_{l=n+1}^{N+n}\inf_{\psi\in \Psi}
-\mathsf{E}\left[\left.\left|\bar Q_l(X_{l})-\psi(X_{l})\right|^{2}\right | \TrainSet \vee \mathcal{G}_N \right].
+\label{eq:Ask-part}
+A_{s,\mathbf{k}}(x)=\gamma^{|\mathbf{k}'|/2}\frac{\sqrt{(\mathbf{k}-\mathbf{k}')!}}{\sqrt{\mathbf{k}!}}\mathsf{E}\left[ \partial_{Z_1}^{\mathbf{k}'} F(x,Z_1,\ldots,Z_s) \mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_{1})\right]
 \end{eqnarray}
-In order to assess the complexity of the proposed algorithm,  we first prove that under some conditions the coefficients \(a_{p,l,\mathbf{k}}\) decay exponentially fast as \(|p-l|\to \infty.\)
-\begin{lem}
-\label{a_decay}
-Suppose that \(f,\mu\in C^1(\mathbb{R}^d),\) \(0<b_\mu I\leq\jac{\mu}(x)\leq B_\mu I\) and \(B_\mu\gamma_{l}\leq 1,\) then
-\begin{eqnarray*}
-\|a_{p,l,\mathbf{k}}\|_{\infty}\leq \sqrt{\gamma_l}\, \|\nabla f\|_\infty\exp\left(-b_{\mu}\sum_{r=l+1}^p \gamma_{r}  \right),\quad \mathbf{k}\in \mathbb{N}^d\setminus \{\mathbf{0}\}.
-\end{eqnarray*}
-\end{lem}
-\begin{cor}
-Assume that
-\begin{eqnarray}
-\label{eq:gamk-decay}
-\gamma_k=\gamma_1 k^{-\alpha},\quad \alpha\in (0,1),
-\end{eqnarray}
- then
-\begin{eqnarray*}
-\|a_{p,l,\mathbf{k}}\|_{\infty}\leq \sqrt{\gamma_1} l^{-\alpha/2}\, \|\nabla f\|_\infty\exp\left(-c b_{\mu}\gamma_1 (p^{1-\alpha}-l^{1-\alpha}) \right)
-\end{eqnarray*}
-for some constant \(c>0.\)
-\end{cor}
-Suppose now that the chain is close to stationarity, then \(Q_{p,l}\) are functions of \(p-l\) only.
-Furthermore,  according to Lemma~\ref{a_decay}, \(a_{p,l,\mathbf{k}}\) are exponentially small for \(p-l\) large.
-As a result we only need to compute a logarithmic (in \(n\)) number of functions \(Q_{p,l}.\)
-Hence the cost of computing the estimates  \(\widehat {a}_{p,l,\mathbf{k}}(x)\) for \(l=N+1,\ldots,N+n,\) and \(\|\mathbf{k}\|\leq K\) using regression on \(\Psi,\) is of order
-\[
-\log^{1/(1-\alpha)} (n) K^d  \NtrainPath D^2.
-\]
-Suppose for simplicity that all functions \(\bar{Q}_l\) are in \(\Psi\) for some \(D>0,\) that is, the third term in \eqref{eq:ula_red_var} is zero.
-Then under \eqref{eq:gamk-decay} it is enough to take \(K=\lceil 1/\alpha \rceil+1\) to get
-\begin{eqnarray*}
-\sum_{l=N+1}^{N+n}\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
-\left(\frac{\gamma_{l}}{2}\right)^{|I|K}\leq C
-\end{eqnarray*}
-for some constant \(C\) not depending on \(d.\) Then we have with high probability
-\begin{eqnarray*}
-\mathsf{Var}\left[\left.\widehat \pi_{K,n}^N(f)\right | \TrainSet \vee \mathcal{G}_N\right]\lesssim \frac{1}{n^{2(1-\alpha)}}\left[1+\frac{n}{\NtrainPath}\right],
-\end{eqnarray*}
-with corresponding cost proportional to \(\NtrainPath\log^{1/(1-\alpha)} (n) ,\) provided \(N/n=o(1).\) We should compare this to the standard weighted estimator \(\pi_{n}^N(f)\) with variance
-\begin{eqnarray*}
-\mathsf{Var}\left[\left. \pi_{n}^N(f)\right |  \mathcal{G}_N\right]\lesssim \frac{1}{n^{1-\alpha}}
-\end{eqnarray*}
-and cost of order \(n.\) Thus while the cost of achieving
-\begin{eqnarray*}
-\mathsf{Var}\left[\left. \pi_{n}^N(f)\right | \TrainSet \vee \mathcal{G}_N\right]\leq \varepsilon^2
-\end{eqnarray*}
-is of order  \(\varepsilon^{-2/(1-\alpha)},\) we get the same bound for \(\pi_{K,n}^N(f)\) at a cost of order
-\[
-\varepsilon^{-1/(1-\alpha)}\log^{1/(1-\alpha)} (\varepsilon).
-\]
-In the presence of approximation errors we have to increase \(D\) with \(n\) to balance  the statistical and approximation errors resulting in a smaller complexity reduction.
-Note that our approach does not require the computation of \(\nabla \log(\pi)\) for the construction of control variates.
-This leads to an additional complexity reduction especially in statistical application where the computation of \(\nabla \log(\pi)\) can be very costly.
-\section{Numerical results}
-\label{sec:num}
-If each function  \(Q_{p,l}(x) \) can be well approximated by polynomials, then we can use polynomial basis functions in   \eqref{eq:06042018a1} to approximate \(\bar Q_l\).
-Suppose that we constructed a polynomial approximation  for each function \(\bar Q_{l}\)  in the form:
-\begin{equation*}
-\widehat{Q}_{l}(x) = \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} x^{\mathbf{s}},\quad s=(s_1,\ldots,s_d)
-\end{equation*}
-with some coefficients \(\beta_{\mathbf{s}}\in \mathbb{R}.\)
-Then using the identity
-\[
-\xi^j = j! \sum_{r = 0}^{\lfloor j/2 \rfloor} \frac{1}{2^r  r! \sqrt{(j-2r)!}} H_{j-2r}(\xi),\quad \xi \in \mathbb{R},
-\]
-we derive for all $x \in \mathbb{R}^d$,
-\begin{eqnarray*}
-\widehat a_{l,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(x) \widehat{Q}_{l}(x - \gamma_l \mu(x) + \sqrt{\gamma_l}\xi)}{\TrainSet \vee \mathcal{G}_N}
-= \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{i,k_i,s_i}(x),
-\end{eqnarray*}
-where for all integers $i,k,s_i$ and $x \in \rset^d$,
-\begin{align*}
-P_{l,k,s}(x) &= \mathsf{E} \left[ H_{k}(\xi_l) (x_l - \gamma_l \mu_l(x) + \sqrt{\gamma_l} \xi_l)^{s} \right]
-\end{align*}
-is a one-dimensional polynomial (in \(x\)) of degree at most \(s\) with analytically known coefficients.
-Alternatively, on practice we can work directly with functions \(Q_{p,l}.\) Due to (almost) time-homogenous Markov chain generated by ULA, each function $Q_{p,l}$ can be approximated by function, which depends only on the  $p-l$, that is,
-\[
- Q_{p,l}(x) = \mathsf{E} \left[f(X_p) | X_l = x\right] \approx Q^\circ_{p-l}(x)
- \]
-Thus we have
-\[
-  Q^\circ_{r}(x) = \mathsf{E} \left[f(X_{l+r}) | X_l = x \right].
- \]
-Consequently, the functions $Q^\circ_r$ can be estimated using a modified least squares criteria
-\begin{eqnarray}
-\label{eq:Qcircr}
-\widehat{Q}^\circ_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
-\end{eqnarray}
-for \(1 \leq r \leq n-1,\) where  $\widehat{Q}^\circ_0(x) = f(x)$ by definition.
-Due to Lemma~\ref{a_decay}, it is enough to estimate $Q^\circ_{r}$ for $|r| < n_{\mathrm{trunc}}$ for some truncation level \(n_{\mathrm{trunc}}\) depending on \(d\) and \(\gamma.\) It allows us to use a smaller amount of training trajectories to approximate conditional expectations $Q^\circ_{r}$.
-Finally  we construct a truncated version of our estimator:
-\[
-\pi_{K,n,n_{\mathrm{trunc}}}^N(f) = \pi_n^N(f) -  M_{K,n, n_{\mathrm{trunc}}}^{N}(f),
-\]
-where
-\begin{eqnarray*}
-{\widehat  M}_{K,n, n_{\mathrm{trunc}}}^N &=& \sum_{0<\left\| \mathbf{k}\right\| \leq K} \sum_{p=N+1}^{N+n} \omega_{p,n}^N
-\\
-&&\times
-\sum_{l=N+1}^p {\widehat  a}_{p-l,\mathbf{k}}(X_{l-1})\mathbf{H}_{\mathbf{k}}(Z_l) \mathbbm{1}\{ |p-l| < n_{\mathrm{trunc}} \}
-\end{eqnarray*}
-and
-\begin{eqnarray*}
-\widehat a_{p-l,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(x) \widehat{Q}^\circ_{p-l}(x - \gamma_l \mu(x) + \sqrt{\gamma_l}\xi)}{\TrainSet \vee \mathcal{G}_N}.
-\end{eqnarray*}
-\subsection{Gaussian mixtures}
- We consider a sample generated by ULA with $\pi$  given by the mixture of two Gaussian distributions  with equal weights:
-\[
-\pi (x) = \frac{1}{2(2\pi)^{d/2}} \left( \rme^{\frac{-\| x-a\|_2^2}{2}}  + \rme^{\frac{-\| x+a\|_2^2}{2}} \right), \quad x \in\mathbb R^d
-\]
-where $a \in \mathbb{R}^d$ is a given vector. The function $U(x)$ and its gradient are given by
-\[
-U(x) = \frac{1}{2} \|x - a\|_2^2 - \text{log}(1 + \rme^{-2x^\top a}), \,\,
-\nabla U(x) = x-a +2a(1 + \rme^{2 x^\top a})^{-1},
-\]
-respectively.
-In our experiments we considered dimensions $d = 2$ and $d = 8$ and take $a=( (2d)^{-1/2}, \dots, (2d)^{-1/2})^\top$.
-In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i,\)
- we have used constant step sizes $\gamma_i = \gamma=0.2$ and sampled $n = 10^4$ samples for $d = 2$ and $n = 2 \times 10^3$ for $d = 8$.
-We generated $\NtrainPath = 10$ independent "training" trajectories and solved the least squares  problems \eqref{eq:Qcircr} using the first order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section.
-The truncation level \(n_{\mathrm{trunc}}\) is chosen to be \(50.\) To test our variance reduction algorithm, we generated 100 independent trajectories.
- In Figure ~\ref{fig:1} we compare our approach to variance reduction methods of \cite{mira2013zero} and \cite{brosse2018diffusion}.
-
- \begin{figure}[tbh]
-\centering
-\subfloat[Subfigure 1 list of figures text][d = 2]{
-\includegraphics[width=0.45\textwidth]{../pictures/GM_2d_f1.png}
-\label{fig:subfig1}}
-\qquad
-\subfloat[Subfigure 2 list of figures text][d = 8]{
-\includegraphics[width=0.45\textwidth]{../pictures/GM_8d_f1.png}
-\label{fig:subfig2}}
-\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Gaussian mixture model.
-The compared estimators are the ordinary empirical average  (O), our estimator of control variates (CV-B), zero variance estimator (ZV) and the control variates using diffusion approximation (CV), along with the region obtained by the ordinary ULA of the length $10^2 \times n$ (green regions).
-\label{fig:1}}
-\end{figure}
-
-\subsection{Binary Logistic Regression}
-
-Second experiment considers the problem of logistic regression.
-Suppose we have i.i.d.
-sample $\left\{ (\mathbf{X}_i, Y_i)\right\}$ for $ i =1, \dots , m$ with features $\mathbf{X}_i \in \mathbb{R}^d $ and binary labels $Y_i \in \left\{0,1 \right\}$.
-The binary logistic regression model defines the conditional distribution of $Y$ given $X$ by a logistic function
-\begin{eqnarray*}
-r(\theta, x) = \frac{e^{\theta^T x}}{1 + e^{\theta^T x}},
-\end{eqnarray*}
-where $\theta$ is a parameter of model.
-In order to estimate $\theta$ according to given data, the Bayessian approach introduces prior distribution $\pi_0(\theta)$ and inferences the posterior density $\pi(\theta)$ using Bayes' rule.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=0.6\textwidth]{../pictures/BLR_banknotes_4d.png}
-\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Logistic Regression.
-The compared estimators are the ordinary empirical average  (O), our estimator of control variates (CV-B), zero variance estimator (ZV) and control variates using diffusion approximation (CV).
- \label{fig:2}}
-\end{center}
-\end{figure}
-In the case of  Gaussian prior $\pi_0$ with zero mean and covariance matrix $\sigma^2 I_d$, the posterior density takes the form:
-\begin{eqnarray*}
-\pi(\theta) \propto \exp \left\{ -  \mathbf{Y}^T \mathbf{X \theta} - \sum_{i=1}^m \text{log} (1 + e^{-\mathbf{\theta}^T \mathbf{X}_i}) - \frac{1}{2\sigma^2} \left\| \mathbf{\theta}\right\|_2^2 \right\},
-\end{eqnarray*}
-where $\mathbf{Y}$ is defined as $(Y_1, \dots, Y_m)^T \in \left\{ 0,1\right\}$ and $\sigma^2 $ is an additional parameter. We have
-\begin{eqnarray*}
-U( \mathbf{\theta}) = \mathbf{Y}^T \mathbf{X \theta} + \sum_{i=1}^m \text{log} (1 + e^{-\mathbf{\theta}^T \mathbf{X}_i}) + \frac{1}{2\sigma^2} \left\|\mathbf{\theta}\right\|_2^2, \,\,
-\nabla U(\mathbf{\theta}) = \mathbf{X}^T \mathbf{Y} - \sum_{i=1}^m \frac{\mathbf{X}_i}{1 + e^{\mathbf{\theta}^T \mathbf{X}_i}} + \frac{1}{\sigma^2} \mathbf{\theta}.
-\end{eqnarray*}
-To demonstrate the performance of the proposed control variates approach in the above Bayesian logistic regression model, we take a simple dataset from \cite{mira2013zero}, which contains the measurements of four variables on $m=200$ Swiss banknotes.
-Prior distribution of the regression parameter $\mathbf{\theta} = \left( \theta_1, \theta_2, \theta_3, \theta_4 \right)$ assumed to be normal with the covariance matrix $\sigma^2 I_4$, where $\sigma^2 = 100$.
-To construct trajectories of length $n = 5 \times 10^3, $ we take  constant step sizes $\gamma_i = 0.1$ for the ULA scheme with $N = 10^3$ burn-in steps.
-As in the previous experiment we use the first order polynomials approximations to analytically compute the coefficients $\hat{a}_{p-l,\mathbf{k}},$ where $T = 10$ and $n_{trunc} = 50$.
-The target function is taken to be $f(\theta) = \sum_{i=1}^{i=d}\theta_i$.
-In order to test our variance reduction algorithm, we generates \(100\) independent test trajectories.
-In Figure ~\ref{fig:2} we compare our approach to variance reduction methods of \cite{mira2013zero} and \cite{brosse2018diffusion}.
-
-\subsection{Binary Probit Regression}
-
-Finally we consider a Bayesian probit regression model in its ordinary form, following \cite{mira2013zero}. More precisely, the log-likelihood of the model looks as follows
-\begin{equation*}
-    \mathbf{L}(\mathbf{Y}| \theta, \mathbf{X}) = \sum_{i=1}^{m} \left[ \mathbf{Y}_i \text{log} (\Phi(\theta^T \mathbf{X}_i)) + (1-\mathbf{Y}_i) \text{log}(\Phi(-\theta^T \mathbf{X}_i))\right],
-\end{equation*}
-where $\theta \in \mathbb{R}^d$ is a parameter of the model and $\Phi$ is a cumulative distribution function of the standard normal distribution $\mathcal{N}(0,1)$.
-Then, if the prior $\pi(\mathbf{\theta}) = \mathcal{N}(0, \sigma^2 I_d)$, then the posterior distribution expresses as
-\begin{equation*}
-    \pi(\theta | \mathbf{Y}, \mathbf{X})  \propto \text{exp} \left\{\sum_{i=1}^{m} \left[\mathbf{Y}_i \text{log} (\Phi(\theta^T \mathbf{X}_i)) + (1-\mathbf{Y}_i) \text{log}(\Phi(-\theta^T \mathbf{X}_i))\right]- \frac{1}{2\sigma^2} \left\| \mathbf{\theta}\right\|_2^2  \right\}
-\end{equation*}
-Consequently,
-\begin{equation*}
-    U(\theta) =  - \sum_{i=1}^{m} \left[\mathbf{Y}_i \text{log} (\Phi(\theta^T \mathbf{X}_i)) + (1-\mathbf{Y}_i) \text{log}(\Phi(-\theta^T \mathbf{X}_i))\right] + \frac{1}{2\sigma^2} \left\| \mathbf{\theta}\right\|_2^2
-\end{equation*}
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=0.6\textwidth]{../pictures/Probit_banknotes_4d.png}
-\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Probit Regression.
-The compared estimators are the ordinary empirical average  (O), our modified estimator (CV-B), zero variance estimator (ZV) and the estimator with control variates using asymptotic variance minimization (CV).
- \label{fig:3}}
-\end{center}
-\end{figure}
-
-\begin{equation*}
-    \nabla U(\theta) = - \mathbf{X}^T s + \frac{1}{\sigma^2} \theta,
-\end{equation*}
-where the elements of the \(m\)-dimensional column vector $s$ are given by
-\begin{equation*}
-    s_i = \mathbf{Y}_i \Omega (\theta^T \mathbf{X}_i) - (1 - \mathbf{Y}_i) \Omega(- \theta^T \mathbf{X}_i), \quad i = 1, 2, \dots, m,
-\end{equation*}
-and $\Omega(\theta^T \mathbf{X}_i) : = \phi(\theta^T \mathbf{X}_i) / \Phi(\theta^T \mathbf{X}_i)$, where $\phi$ stands for the density of the standard normal distribution.
-As in the previous examples, to show the performance  of the proposed variance reduced estimator for the Bayesian probit regression model, banknotes dataset is used and the regularization parameter is taken to be $\sigma^2=100$.
-To generate trajectories of the length $n = 5 \times 10^3, $ we take a constant step size $\gamma_i = 0.1$ for the ULA scheme after $N = 10^3$ burn-in steps.
-We use again the first order polynomials approximations to  compute the coefficients $\hat{a}_{p-l,\mathbf{k}}$ and fixed $T = 10$, $n_{trunc} = 50$ and $K=1$.
-The target function is now $f(\theta) = \sum_{i=1}^{i=d}\theta_i$.
-In order to test our variance reduction algorithm, we generate \(100\) independent test trajectories.
-In Figure ~\ref{fig:3} we compare our approach to the variance reduction methods of \cite{mira2013zero} and \cite{brosse2018diffusion}.
-
+where $F_s(X_0,Z_1,\ldots,Z_s)=\frac{1}{s}\sum_{r=0}^s f(X_r)$ and $\partial_{Z_1}^{\mathbf{k}'}$
+stands for a weak partial derivative of the functional $F_s$ that also can be viewed as discretised version of Malliavin derivative. From \eqref{eq:Ask-part}, we derive
+\begin{eqnarray*}
+\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1} A^2_{s,\mathbf{k}}(x)\leq \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
+\left(\frac{\gamma}{2}\right)^{|I|K}
+\mathsf{Var}_x\left(\frac{1}{s}\sum_{p=1}^{s}\partial_{y_{1}}^{\mathbf{K}_I}f\left(X_{p}\right)
+\right),
+\end{eqnarray*}
+where the sum runs over all nonempty subsets $I$ of the set $\{1,\ldots,d\}$ and 
+\[
+\mathbf{\mathbf{K}}_{I}=(K1_{\{1\in I\}},\ldots,K1_{\{d\in I\}}).
+\]
+ Finally we show via  the Poincare inequality that under smoothness assumption \eqref{eq:smooth-mu} it holds
+\begin{eqnarray*}
+\mathsf{Var}_x\left(\frac{1}{s}\sum_{p=1}^{s}\partial_{y_{1}}^{\mathbf{K}_I}f\left(X_{p}\right)
+\right)\leq W_s(x)
+\end{eqnarray*}
+for all $x$ and some family of functions $(W_s)_{s\geq 1}$ such that  the sum $\sum_{s=1}^n \mathsf{E}[W_s(X_{N+n-s-1})]$ is bounded uniformly in $n\in \mathbb{N}.$
 \section{Proofs}
@@
 $$
-\partial^{\mathbf k}_{y_j} h(y_1,\ldots,y_m)
+\partial^{\mathbf k}_{j} h(y_1,\ldots,y_m)
 :=\partial^{k_d}_{y_j^d}
@@
 \[
-X_{p}=G_{p,l}(X_{l-1},\sqrt{\gamma_{l}}Z_{l},\ldots,\sqrt{\gamma_{p}}Z_{p}),
-\]
-where the function \(G_{p,l}:\) \(\mathbb{R}^{d\times(p-l+2)}\to \mathbb{R}^{d}\) is defined as
+X_{p}=G_{p}(X_0,\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p}),
+\]
+where the function \(G_{p}:\) \(\mathbb{R}^{d\times(p+1)}\to \mathbb{R}^{d}\) is defined as
 \begin{equation}
 \label{eq:definition-G-p-l}
-G_{p,l}(x,y_l,\ldots,y_p):=\Phi_{p}(\cdot,y_{p})\circ\Phi_{p-1}(\cdot,y_{p-1})\circ\ldots\circ\Phi_{l}(x,y_{l})
+G_{p}(x,y_1,\ldots,y_p):=\Phi(\cdot,y_{p})\circ\Phi(\cdot,y_{p-1})\circ\ldots\circ\Phi(x,y_{1})
 \end{equation}
@@
 \[
-\Phi_{j}(x,y)=x-\gamma_{j}\mu(x)+y,
+\Phi(x,y)=x-\gamma\mu(x)+y,
 \quad x,y\in\mathbb R^d.
@@
 $$
-f\left(X_{p}\right) =f\circ G_{p,l}(X_{l-1},\sqrt{\gamma_{l}}Z_{l},\ldots,\sqrt{\gamma_{p}}Z_{p}).
+f\left(X_{p}\right) =f\circ G_{p}(X_{0},\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p}).
 $$
@@
 \label{eq:definition-differential-f-p}
-\partial_{y_l}^{\mathbf k} f\left(X_{p}\right)
-:=\partial_{y_l}^{\mathbf k} [f\circ G_{p,l}](X_{l-1},\sqrt{\gamma_{l}}Z_{l},\ldots,\sqrt{\gamma_{p}}Z_{p})
+\partial_{1}^{\mathbf k} f\left(X_{p}\right)
+:=\partial_{1}^{\mathbf k} [f\circ G_{p}](X_{0},\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p})
 \end{equation}
-whenever the function $f\circ G_{p,l}$ is smooth enough
+whenever the function $f\circ G_{p}$ is smooth enough
 (that is, $f$ and $\mu$ need to be smooth enough).
@@
 \begin{lem}\label{eq:a_repr}
-Fix $l\le p$ and some \(\mathbf{k},\mathbf{k}'\in \mathbb{N}_0^d\) with \(\mathbf{k}'\le \mathbf{k}\) componentwise.  Then the following representation holds
-$$
-a_{p,l,\mathbf{k}}(X_{l-1})=\left(\gamma_{l}^{|\mathbf{k}'|}\frac{(\mathbf{k}-\mathbf{k}') !}{\mathbf{k}!}\right)^{1/2}
+Fix some \(\mathbf{k},\mathbf{k}'\in \mathbb{N}_0^d\) with \(\mathbf{k}'\le \mathbf{k}\) componentswise.  Then the following representation holds
+$$
+\bar a_{p,\mathbf{k}}(x)=\left(\gamma^{|\mathbf{k}'|}\frac{(\mathbf{k}-\mathbf{k}') !}{\mathbf{k}!}\right)^{1/2}
 \mathsf E\left[\left.
-\partial_{y_l}^{\mathbf{k}'}f(X_p)\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_l)\right|X_{l-1}\right].
+\partial_{1}^{\mathbf{k}'}f(X_p)\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_l)\right|X_{0}=x\right].
 $$
@@
 \begin{align*}
-&a_{p,l,\mathbf k}(x)
+&\bar a_{p,\mathbf{k}}(x)
  =
 \int_{\mathbb R^d}\ldots\int_{\mathbb R^d}
-f\circ G_{p,l}(x,\sqrt{\gamma_{l}}z_{l},\ldots,\sqrt{\gamma_{p}}z_{p})
-\mathbf{H}_{\mathbf{k}}(z_{l})\boldsymbol{\varphi}(z_l)
-\prod_{j=l+1}^p\boldsymbol{\varphi}(z_j)\, dz_{l}\ldots dz_{p}
+f\circ G_{p}(x,\sqrt{\gamma}z_{1},\ldots,\sqrt{\gamma}z_{p})
+\mathbf{H}_{\mathbf{k}}(z_{1})\boldsymbol{\varphi}(z_1)
+\prod_{j=2}^p\boldsymbol{\varphi}(z_j)\, dz_{1}\ldots dz_{p}
 \\
@@
 \int_{\mathbb R^d}\ldots\int_{\mathbb R^d}
-f\circ G_{p,l}(x,\sqrt{\gamma_{l}}z_{l},\ldots,\sqrt{\gamma_{p}}z_{p})
-(-1)^{|\mathbf{k}|}\partial^{\mathbf{k}} \boldsymbol{\varphi}(z_l)
-\prod_{j=l+1}^p\boldsymbol{\varphi}(z_j)\, dz_{l}\ldots dz_{p}
+f\circ G_{p}(x,\sqrt{\gamma}z_{1},\ldots,\sqrt{\gamma}z_{p})
+(-1)^{|\mathbf{k}|}\partial^{\mathbf{k}} \boldsymbol{\varphi}(z_1)
+\prod_{j=2}^p\boldsymbol{\varphi}(z_j)\, dz_{1}\ldots dz_{p}
 \\
 & =
-\gamma_l^{|\mathbf k'|/2}\frac{1}{\sqrt{\mathbf k!}}
+\frac{\gamma^{|\mathbf k'|/2}}{\sqrt{\mathbf k!}}
 \int_{\mathbb R^d}\ldots\int_{\mathbb R^d}
-\partial_{y_l}^{\mathbf k'}[f\circ G_{p,l}](x,\sqrt{\gamma_{l}}z_{l},\ldots,\sqrt{\gamma_{p}}z_{p})
-\\
-&\hspace{1em}\times
-(-1)^{|\mathbf{k}-\mathbf k'|}\partial^{\mathbf{k}-\mathbf k'} \boldsymbol{\varphi}(z_l)
-\prod_{j=l+1}^p\boldsymbol{\varphi}(z_j)\, dz_{l}\ldots dz_{p}
-\\
-%& =
-%\gamma_l^{|\mathbf k'|/2}\frac{\sqrt{(\mathbf k-\mathbf k')!}}{\sqrt{\mathbf k!}}
-%\int_{\mathbb R^d}\ldots\int_{\mathbb R^d}
-%\partial_{y_l}^{\mathbf k'}[f\circ G_{p,l}](x,\sqrt{\gamma_{l}}z_{l},\ldots,\sqrt{\gamma_{p}}z_{p})
-%\\
-&\hspace{1em}\times
-\mathbf{H}_{\mathbf{k}-\mathbf k'}(z_{l})\boldsymbol{\varphi}(z_l)
-\prod_{j=l+1}^p\boldsymbol{\varphi}(z_j)\, dz_{l}\ldots dz_{p}
-\\
- & =\gamma_{l}^{|\mathbf{k}'|/2}\frac{\sqrt{(\mathbf{k}-\mathbf{k}')!}}{\sqrt{\mathbf{k}!}}\mathsf{E}\left[\partial_{y_{l}}^{\mathbf{k}'}[f\circ G_{p,l}](x,\sqrt{\gamma_{l}}Z_{l},\ldots,\sqrt{\gamma_{p}}Z_{p})\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_{l})\right].
+\partial_{1}^{\mathbf k'}[f\circ G_{p}](x,\sqrt{\gamma}z_{1},\ldots,\sqrt{\gamma}z_{p})
+(-1)^{|\mathbf{k}-\mathbf k'|}\partial^{\mathbf{k}-\mathbf k'} \boldsymbol{\varphi}(z_1)
+\prod_{j=2}^p\boldsymbol{\varphi}(z_j)\, dz_{1}\ldots dz_{p}
+\\
+ & =\gamma^{|\mathbf{k}'|/2}\frac{\sqrt{(\mathbf{k}-\mathbf{k}')!}}{\sqrt{\mathbf{k}!}}\mathsf{E}\left[\partial_{y_{1}}^{\mathbf{k}'}[f\circ G_{p}](x,\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p})\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_{1})\right].
 \end{align*}
@@
 \begin{equation*}
-\begin{split}
-\bar a_{l,\mathbf{k}}(X_{l-1})
-&=\left(\gamma_{l}^{|\mathbf{k}'|}\frac{(\mathbf{k}-\mathbf{k}')!}{\mathbf{k}!}\right)^{1/2}
-\,\sum_{p=l}^{N+n}\omega_{p,n}^N
-\mathsf E\left[\left.
-\partial_{y_l}^{\mathbf{k}'}f(X_p)\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_l)\right|X_{l-1}\right]\\
-&=\left(\gamma_{l}^{|\mathbf{k}'|}\frac{(\mathbf{k}-\mathbf{k}')!}{\mathbf{k}!}\right)^{1/2}
-\\
-&\times \sum_{p=l}^{N+n}\omega_{p,n}^N
-\mathsf E\left[\left.
-\left(
-\partial_{y_l}^{\mathbf{k}'}f(X_p)-\mathsf E\left[\partial_{y_l}^{\mathbf{k}'}f(X_p)|X_{l-1}\right]
-\right)
-\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_l)\right|X_{l-1}\right].
-\end{split}
+A_{s,\mathbf{k}}(x)
+=\left(\gamma^{|\mathbf{k}'|}\frac{(\mathbf{k}-\mathbf{k}')!}{\mathbf{k}!}\right)^{1/2}
+\,
+\mathsf{E}_x\left[
+\frac{1}{s}\sum_{r=1}^{s}(\partial_{1}^{\mathbf{k}'} f(X_r)-\mathsf{E}_x\partial_{1}^{\mathbf{k}'}f(X_r))\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_1)\right]
 \end{equation*}
@@
 by taking $\mathbf k'= \mathbf k'(\mathbf k)
-=(K1_{\{k_1>K\}},\ldots,K1_{\{k_d>K\}})$,
-for each $l\in\{N+1,\ldots,N+n\}$, we get
+=(K1_{\{k_1>K\}},\ldots,K1_{\{k_d>K\}})$, we get
 \begin{align}
 \label{eq:sum_abar}
-\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}\left[\bar{a}_{l,\mathbf{k}}^{2}(X_{l-1})|\mathcal{G}_{N}\right] & =\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}\left(\gamma_{l}^{|\mathbf{k}'|}\frac{(\mathbf{k}-\mathbf{k}')!}{\mathbf{k}!}\right)Q(\mathbf{k}',\mathbf{k}-\mathbf{k}')\\
+\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}A^2_{s,\mathbf{k}}(x) & =\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}\left(\gamma^{|\mathbf{k}'|}\frac{(\mathbf{k}-\mathbf{k}')!}{\mathbf{k}!}\right)Q_s(\mathbf{k}',\mathbf{k}-\mathbf{k}')\\
 \nonumber
- & =\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}\gamma_{l}^{|I|K}\sum_{\mathbf{m}_{I}\in\mathbb{N}_{I}^{d}}\frac{\mathbf{m}_{I}!}{\left(\mathbf{m}_{I}+\mathbf{K}_{I}\right)!}
+ & =\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}\gamma^{|I|K}\sum_{\mathbf{m}_{I}\in\mathbb{N}_{I}^{d}}\frac{\mathbf{m}_{I}!}{\left(\mathbf{m}_{I}+\mathbf{K}_{I}\right)!}
  \\
  \nonumber
- & \hspace{1em}\times\sum_{\mathbf{m}_{I^c}\in \mathbb{N}^d_{0,I^c},\,\|\mathbf{m}_{I^c}\|\leq K}Q(\mathbf{\mathbf{K}}_{I},\mathbf{m}_{I}+\mathbf{m}_{I^c}),
+ & \hspace{1em}\times\sum_{\mathbf{m}_{I^c}\in \mathbb{N}^d_{0,I^c},\,\|\mathbf{m}_{I^c}\|\leq K}Q_s(\mathbf{\mathbf{K}}_{I},\mathbf{m}_{I}+\mathbf{m}_{I^c}),
 \end{align}
 where for any two multi-indices \(\mathbf{r},\) \(\mathbf{q}\) from \(\mathbb{N}_0^d\)
-\begin{multline*}
-Q(\mathbf{r},\mathbf{q})
-\\
-=\mathsf{E}\left\{
-\left(\mathsf{E}\left[\sum_{p=l}^{N+n}\omega_{p,n}^{N}\left(\partial_{y_{l}}^{\mathbf r}f\left(X_{p}\right)-\mathsf{E}\left[\partial_{y_{l}}^{\mathbf r}f\left(X_{p}\right)|X_{l-1}\right]\right)\mathbf{H}_{\mathbf{q}}(Z_{l})\Big|X_{l-1}\right]\right)^{2}
-\right\}.
-\end{multline*}
+\begin{eqnarray*}
+Q_s(\mathbf{r},\mathbf{q})
+=
+\left(\mathsf{E}_x\left[\frac{1}{s}\sum_{p=1}^{s}\left(\partial_{y_{1}}^{\mathbf r}f\left(X_{p}\right)-\mathsf{E}_x\left[\partial_{y_{1}}^{\mathbf r}f\left(X_{p}\right)\right]\right)\mathbf{H}_{\mathbf{q}}(Z_{1})\right]\right)^{2}.
+\end{eqnarray*}
 In \eqref{eq:sum_abar} the first sum runs over all nonempty subsets $I$ of the set $\{1,\ldots,d\}.$
@@
 \begin{eqnarray*}
-\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}\left[\bar{a}_{l,\mathbf{k}}^{2}(X_{l-1})|\mathcal{G}_{N}\right]&\leq &
+\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}A^2_{s,\mathbf{k}}(x)&\leq &
 \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset} (\gamma_{l}/2)^{|I|K}
@@
 Now using the consequence
-$\sum_{\mathbf{m}\in \mathbb{N}^d_0} \langle\xi,\mathbf{H}_\mathbf{m}(Z_l)\rangle^2\le\langle\xi,\xi\rangle$
-of Parseval's identity
-(the latter is used conditionally on $X_{l-1}$
-%equivalently, on $\mathcal G_{l-1}$,
-which is possible because the system
-$\{\mathbf{H}_\mathbf{m}(Z_l)\}_{m\in \mathbb{N}^d_0}$
-is orthonormal conditionally on $X_{l-1}$),
-%and the tower property $\mathsf E[\mathsf E(\cdot|\mathcal G_{l-1})|\mathcal G_N]=\mathsf E[\cdot|\mathcal G_N]$
+$\sum_{\mathbf{m}\in \mathbb{N}^d_0} \langle\xi,\mathbf{H}_\mathbf{m}(Z_1)\rangle^2\le\langle\xi,\xi\rangle$
+of the Parseval's identity,
 we derive
 \begin{eqnarray*}
-\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}\left[\bar{a}_{l,\mathbf{k}}^{2}(X_{l-1})|\mathcal{G}_{N}\right]
+\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}A^2_{s,\mathbf{k}}(x)
 &\leq & \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
 \left(\frac{\gamma_{l}}{2}\right)^{|I|K}
-\\
-&& \times\mathsf E\left[
-\mathsf{Var}\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}\partial_{y_{l}}^{\mathbf{K}_I}f\left(X_{p}\right)
-\Big|X_{l-1}\right)
-\Bigg|\mathcal G_N\right]
-\\
-&=& \frac{1}{\Gamma^2_{N+2,N+n+1}}\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
-\left(\frac{\gamma_{l}}{2}\right)^{|I|K}
-R_{l,n,N}^{I,K}
-\end{eqnarray*}
-with
-\begin{eqnarray*}
-R_{l,n,N}^{I,K}=\mathsf E\left[
-\mathsf{Var}\left(\sum_{p=l}^{N+n}\gamma_{p+1}\partial_{y_{l}}^{\mathbf{K}_I}f\left(X_{p}\right)
-\Big|X_{l-1}\right)\right].
-\end{eqnarray*}
-As a result
-\begin{eqnarray*}
-\mathsf{Var}\left[\left.\pi_{K,n}^{N}(f)\right|\mathcal G_N\right]\leq \frac{1}{\Gamma^2_{N+2,N+n+1}}\sum_{l=N+1}^{N+n}\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
-\left(\frac{\gamma_{l}}{2}\right)^{|I|K}
-R_{l,n,N}^{I,K}.
-\end{eqnarray*}
-Next we show that under the conditions  of Theorem~\ref{th:mr}, the quantity \(R_{l,n,N}^{I,K}\) is uniformly bounded in \(l,n,N,I.\) For the sake of simplicity we present the proof only in one-dimensional case.
-First we need to prove several auxiliary results.
+\mathsf{Var}_x\left(\frac{1}{s}\sum_{p=1}^{s}\partial_{1}^{\mathbf{K}_I}f\left(X_{p}\right)
+\right)
+\end{eqnarray*}
+Next we show that under the conditions  of Theorem~\ref{th:mr}  
+\begin{eqnarray*}
+\mathsf{Var}_x\left(\frac{1}{s}\sum_{p=1}^{s}\partial_{1}^{\mathbf{K}_I}f\left(X_{p}\right)
+\right)\leq W_s(x)
+\end{eqnarray*}
+for all $x$ and some family of functions $(W_s)_{s\geq 1}$ such that  the sum $\sum_{s=1}^n \mathsf{E}[W_s(X_{N+n-s-1})]$ is bounded uniformly in $n\in \mathbb{N}.$
+
+ For the sake of simplicity we present the proof only in one-dimensional case. First we need to prove several auxiliary results.
 \begin{lem}\label{lem:06062018a1}
@@
 \begin{align}
-0&\le x_p\le\alpha_p x_{p-1}+\gamma_p \epsilon_p,\quad p\in\mathbb N,
+0&\le x_p\le\alpha_p x_{p-1}+\gamma \epsilon_p,\quad p\in\mathbb N,
 \label{eq:06062018a1}\\
@@
 \end{align}
-where $\alpha_p,\gamma_p\in(0,1)$, $p\in\mathbb N$,
+where $\alpha_p,\gamma\in(0,1)$, $p\in\mathbb N$,
 and $\ol C_0,\ol C_1$ are some nonnegative constants. Assume
 \begin{equation}\label{eq:06062018a3}
-\sum_{r=1}^\infty\gamma_r \prod_{k=1}^r \alpha_k\le\ol C_2
+\gamma \sum_{r=1}^\infty \prod_{k=1}^r \alpha_k\le\ol C_2
 \end{equation}
@@
 x_p\le\ol C_0\prod_{k=1}^p \alpha_k
-+\sum_{r=1}^p \gamma_r\epsilon_r
++\gamma\sum_{r=1}^p \epsilon_r
 \prod_{k=r+1}^p \alpha_k,
@@
 x_p\le\left(\ol C_0+\ol C_1
-\sum_{r=1}^p \gamma_r \prod_{k=1}^r \alpha_k
+\gamma\sum_{r=1}^p  \prod_{k=1}^r \alpha_k
 \right)
@@
 \label{eq:definition-alpha}
-\alpha_k=1-\gamma_k \mu'(X_{k-1}),\quad k\in\mathbb N.
+\alpha_k=1-\gamma \mu'(X_{k-1}),\quad k\in\mathbb N.
 \end{equation}
 
-\begin{remark}\label{rem:10062018a1}
-
-Below this will allow us to apply
-Lemma~\ref{lem:06062018a1}
-to bound double indexed sequences
-$(x_{j,p})_{j\ge1,p\ge j}$
-satisfying
-$$
-0\le x_{j,p}\le\alpha_p x_{j,p-1}+\gamma_p \epsilon_{j,p},\quad p\ge j+1,
-$$
-with suitable $(\epsilon_{j,p})_{j\ge1,p\ge j+1}$.
-\end{remark}
-The assumption (H1) implies that  $|\mu'(x)|\leq B_\mu$ for some constant $B_\mu>0$ and all $x\in \mathbb{R}^d$. Without loss of generality we suppose that $\gamma_N B_\mu<1.$
+The assumption \eqref{eq:smooth-mu} implies that  $|\mu'(x)|\leq B_\mu$ for some constant $B_\mu>0$ and all $x\in \mathbb{R}^d$. Without loss of generality we suppose that $\gamma B_\mu<1.$
 \begin{lem}\label{lem:06062018a2}
@@
 \label{eq:08062018b2}
-\left|\partial_{y_l}^r X_p \right| \le C_r\prod_{k=l+1}^p (1-\gamma_k\mu'(X_{k-1}))\quad \text{a.s.} 
+\left|\partial_{y_l}^r X_p \right| \le C_r\prod_{k=l+1}^p (1-\gamma\mu'(X_{k-1}))\quad \text{a.s.} 
 \end{equation}
@@
 \left|\partial_{y_j} \partial_{y_l}^r X_p\right|
-\le c_r\prod_{k=l+1}^p (1-\gamma_k\mu'(X_{k-1})), \quad \text{a.s.}
+\le c_r\prod_{k=l+1}^p (1-\gamma\mu'(X_{k-1})), \quad \text{a.s.}
 \end{equation}
@@
 \partial_{y_{l}}X_{p}
-=\left[1-\gamma_{p}\mu'(X_{p-1})\right]\partial_{y_{l}}X_{p-1}
+=\left[1-\gamma\mu'(X_{p-1})\right]\partial_{y_{l}}X_{p-1}
 $$
@@
 \partial_{y_{j}}\partial_{y_{l}}X_{p}
-=\left[1-\gamma_{p}\mu'(X_{p-1})\right]\partial_{y_{j}}\partial_{y_{l}}X_{p-1}-\gamma_{p}\mu''(X_{p-1})\partial_{y_{j}}X_{p-1}\partial_{y_{l}}X_{p-1}.
+=\left[1-\gamma\mu'(X_{p-1})\right]\partial_{y_{j}}\partial_{y_{l}}X_{p-1}-\gamma\mu''(X_{p-1})\partial_{y_{j}}X_{p-1}\partial_{y_{l}}X_{p-1}.
 $$
@@
 &\le\alpha_p|\partial_{y_{j}}\partial_{y_{l}}X_{p-1}|
-+\gamma_{p}
++\gamma
 B_\mu
@@
 &\le\alpha_p|\partial_{y_{j}}\partial_{y_{l}}X_{p-1}|
-+\gamma_{p}
++\gamma
 \const
@@
 with a suitable constant.
-By Lemma~\ref{lem:06062018a1}
-together with Remark~\ref{rem:10062018a1} applied
+By Lemma~\ref{lem:06062018a1} applied
 to bound $|\partial_{y_{j}}\partial_{y_{l}}X_{p}|$
@@
 \partial_{y_{l}}^{r}X_{p}
-&=\left[1-\gamma_{p}\mu'(X_{p-1})\right]\partial_{y_{l}}^{r}X_{p-1}
+&=\left[1-\gamma\mu'(X_{p-1})\right]\partial_{y_{l}}^{r}X_{p-1}
 \label{eq:08062018a1}\\
-&\hspace{1em}-\gamma_{p}\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\mu^{(m_{1}+\ldots+m_{r-1})}(X_{p-1})\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X_{p-1}}{k!}\right)^{m_{k}},
+&\hspace{1em}-\gamma\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\mu^{(m_{1}+\ldots+m_{r-1})}(X_{p-1})\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X_{p-1}}{k!}\right)^{m_{k}},
 \notag
@@
 &\partial_{y_{j}}\partial_{y_{l}}^{r}X_{p}
-=\left[1-\gamma_{p}\mu'(X_{p-1})\right]\partial_{y_{j}}\partial_{y_{l}}^{r}X_{p-1}-\gamma_{p}\mu''(X_{p-1})\partial_{y_{l}}^{r}X_{p-1}\partial_{y_{j}}X_{p-1}
+=\left[1-\gamma_{p}\mu'(X_{p-1})\right]\partial_{y_{j}}\partial_{y_{l}}^{r}X_{p-1}-\gamma\mu''(X_{p-1})\partial_{y_{l}}^{r}X_{p-1}\partial_{y_{j}}X_{p-1}
 \label{eq:09062018a2}\\
-&\hspace{1em}-\gamma_{p}\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\mu^{(m_{1}+\ldots+m_{r-1}+1)}(X_{p-1})\partial_{y_{j}}X_{p-1}\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X_{p-1}}{k!}\right)^{m_{k}}
+&\hspace{1em}-\gamma\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\mu^{(m_{1}+\ldots+m_{r-1}+1)}(X_{p-1})\partial_{y_{j}}X_{p-1}\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X_{p-1}}{k!}\right)^{m_{k}}
 \notag\\
-&\hspace{1em}-\gamma_{p}\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\mu^{(m_{1}+\ldots+m_{r-1})}(X_{p-1})\partial_{y_{j}}\left[\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X_{p-1}}{k!}\right)^{m_{k}}\right]
+&\hspace{1em}-\gamma\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\mu^{(m_{1}+\ldots+m_{r-1})}(X_{p-1})\partial_{y_{j}}\left[\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X_{p-1}}{k!}\right)^{m_{k}}\right]
 \notag\\
-&\hspace{1em} =\left[1-\gamma_{p}\mu'(X_{p-1})\right]\partial_{y_{j}}\partial_{y_{l}}^{r}X_{p-1}+\gamma_{p}\epsilon_{l,j,p},
+&\hspace{1em} =\left[1-\gamma\mu'(X_{p-1})\right]\partial_{y_{j}}\partial_{y_{l}}^{r}X_{p-1}+\gamma\epsilon_{l,j,p},
 \notag
@@
 \le\alpha_p|\partial_{y_{j}}\partial_{y_{l}}^{r}X_{p-1}|
-+\gamma_p \const\prod_{k=l+1}^{j}\alpha_{k}\prod_{k=j+1}^{p}\alpha_{k}^{2},
++\gamma \,\const\prod_{k=l+1}^{j}\alpha_{k}\prod_{k=j+1}^{p}\alpha_{k}^{2},
 \quad p\ge j+1.
@@
 We can again apply Lemma~\ref{lem:06062018a1}
-and Remark~\ref{rem:10062018a1}
 to bound $|\partial_{y_{j}}\partial_{y_{l}}^r X_{p}|$
@@
 \label{lem:var_poincare}
-Under assumptions of Theorem~\ref{th:mr},
-for all natural $l\le q$, it holds
-$$
-\mathsf{Var}\left[\sum_{p=l}^{q}\gamma_{p+1}\partial_{y_{l}}^{K}f\left(X_{p}\right)\Big|X_{l-1}\right]\le B_K\quad\text{a.s.},
+Under assumptions of Theorem~\ref{th:mr}, it holds
+$$
+\mathsf{Var}_x\left[\sum_{p=1}^{s}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right]\le B_K\quad\text{a.s.},
 $$
@@
 The expression
-$\sum_{p=l}^q \gamma_{p+1} \partial_{y_l}^K f(X_p)$
+$\sum_{p=1}^q \partial_{y_1}^K f(X_p)$
 can be viewed as a deterministic function of
-$X_{l-1},Z_l,Z_{l+1},\ldots,Z_q$
-$$
-\sum_{p=l}^q \gamma_{p+1} \partial_{y_l}^K f(X_p)
-=F(X_{l-1},Z_l,Z_{l+1},\ldots,Z_q).
+$X_{0},Z_1,Z_{2},\ldots,Z_q$
+$$
+\sum_{p=1}^q  \partial_{y_1}^K f(X_p)
+=F(X_{0},Z_1,Z_{2},\ldots,Z_q).
 $$
@@
 $$
-\mathsf{Var}\left[\sum_{p=l}^{q}\gamma_{p+1}\partial_{y_{l}}^{K}f\left(X_{p}\right)\Big|X_{l-1}\right]
-\le\mathsf E\left[\left.
-\|\nabla_Z F(X_{l-1},Z_l,Z_{l+1},\ldots,Z_q)\|^2
-\right|X_{l-1}\right],
-$$
-where $\nabla_Z F=(\partial_{Z_l} F,\ldots,\partial_{Z_q} F)$,
+\mathsf{Var}_x\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right]
+\le\mathsf E_x\left[
+\|\nabla_Z F(X_{0},Z_1,Z_{2},\ldots,Z_q)\|^2
+\right],
+$$
+where $\nabla_Z F=(\partial_{Z_1} F,\ldots,\partial_{Z_q} F)$,
 and $\|\cdot\|$ denotes the Euclidean norm.
-Notice that \(\partial_{Z_j} F=\sqrt{\gamma_j}\,\partial_{y_j} F\) and
+Notice that \(\partial_{Z_j} F=\sqrt{\gamma}\,\partial_{y_j} F\) and
 hence
 $$
-\mathsf{Var}\left[\sum_{p=l}^{q}\gamma_{p+1}\partial_{y_{l}}^{K}f\left(X_{p}\right)\Big|X_{l-1}\right]\leq\sum_{j=l}^{q}\gamma_{j}\mathsf{E}\left[\left(\sum_{p=l}^{q}\gamma_{p+1}\partial_{y_{j}}\partial_{y_{l}}^{K}f\left(X_{p}\right)\right)^{2}\Big|X_{l-1}\right].
+\mathsf{Var}_x\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right]\leq\gamma\sum_{j=1}^{q}\mathsf{E}_x\left[\left(\sum_{p=1}^{q}\gamma\partial_{y_{j}}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right)^{2}\right].
 $$
 It is straightforward to check that
-$\partial_{y_j}\partial_{y_l}^K f(X_p)=0$
+$\partial_{y_j}\partial_{y_1}^K f(X_p)=0$
 whenever $p<j$. Therefore, we get
 \begin{equation}\label{eq:10062018a2}
-\mathsf{Var}\left[\sum_{p=l}^{q}\gamma_{p+1}\partial_{y_{l}}^{K}f\left(X_{p}\right)\Big|X_{l-1}\right]\leq\sum_{j=l}^{q}\gamma_{j}\mathsf{E}\left[\left(\sum_{p=j}^{q}\gamma_{p+1}\partial_{y_{j}}\partial_{y_{l}}^{K}f\left(X_{p}\right)\right)^{2}\Big|X_{l-1}\right].
+\mathsf{Var}_x\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right]\leq\gamma\sum_{j=1}^{q}\mathsf{E}_x\left[\left(\sum_{p=j}^{q}\gamma \partial_{y_{j}}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right)^{2}\right].
 \end{equation}
-Now fix $p$ and $j$, $p\ge j$, in $\{l,\ldots,q\}$.
+Now fix $p$ and $j$, $p\ge j$, in $\{1,\ldots,q\}$.
 By Fa\`a di Bruno's formula
 \[
-\partial_{y_{l}}^{K}f\left(X_{p}\right)=\sum\frac{K!}{m_{1}!\ldots m_{K}!}f^{(m_{1}+\ldots+m_{K})}(X_{p})\prod_{k=1}^{K}\left(\frac{\partial_{y_{l}}^{k}X_{p}}{k!}\right)^{m_{k}},
+\partial_{y_{1}}^{K}f\left(X_{p}\right)=\sum\frac{K!}{m_{1}!\ldots m_{K}!}f^{(m_{1}+\ldots+m_{K})}(X_{p})\prod_{k=1}^{K}\left(\frac{\partial_{y_{1}}^{k}X_{p}}{k!}\right)^{m_{k}},
 \]
@@
 \begin{eqnarray*}
-\partial_{y_{j}}\partial_{y_{l}}^{K}f\left(X_{p}\right)
-&=&\sum\frac{K!}{m_{1}!\ldots m_{K}!}f^{(m_{1}+\ldots+m_{K}+1)}(X_{p})\left[\partial_{y_{j}}X_{p}\right]\prod_{k=1}^{K}\left(\frac{\partial_{y_{l}}^{k}X_{p}}{k!}\right)^{m_{k}} \\
+\partial_{y_{j}}\partial_{y_{1}}^{K}f\left(X_{p}\right)
+&=&\sum\frac{K!}{m_{1}!\ldots m_{K}!}f^{(m_{1}+\ldots+m_{K}+1)}(X_{p})\left[\partial_{y_{j}}X_{p}\right]\prod_{k=1}^{K}\left(\frac{\partial_{y_{1}}^{k}X_{p}}{k!}\right)^{m_{k}} \\
 &&+\sum\frac{K!}{m_{1}!\ldots m_{K}!}f^{(m_{1}+\ldots+m_{K})}(X_{p})
 \sum_{s=1}^{K}\frac{m_{s}}{s!}
-\left(\frac{\partial_{y_{l}}^{s}X_{p}}{s!}\right)^{m_{s}-1}
-\\
-&& \times\left[\partial_{y_{j}}\partial_{y_{l}}^{s}X_{p}\right]
-\prod_{k\le K,\,k\neq s}\left(\frac{\partial_{y_{l}}^{k}X_{p}}{k!}\right)^{m_{k}}.
+\left(\frac{\partial_{y_{1}}^{s}X_{p}}{s!}\right)^{m_{s}-1}
+\\
+&& \times\left[\partial_{y_{j}}\partial_{y_{1}}^{s}X_{p}\right]
+\prod_{k\le K,\,k\neq s}\left(\frac{\partial_{y_{1}}^{k}X_{p}}{k!}\right)^{m_{k}}.
 \end{eqnarray*}
@@
 \label{eq:part-part}
-\left|\partial_{y_{j}}\partial_{y_{l}}^{K}f\left(X_{p}\right)\right|
-\leq A_{K}\prod_{k=l+1}^{p}\alpha_{k}
+\left|\partial_{y_{j}}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right|
+\leq A_{K}\prod_{k=2}^{p}\alpha_{k}
 \end{eqnarray}
@@
 \begin{align*}
-\mathsf{Var}\left[\sum_{p=l}^{q}\gamma_{p+1}\partial_{y_{l}}^{K}f\left(X_{p}\right)\Big|X_{l-1}\right]
+\mathsf{Var}_x\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right]
 &\le
-A_{K}^{2}\sum_{j=l}^{q}\gamma_{j}
-\mathsf{E}\left(\sum_{p=j}^{q}\gamma_{p+1}\prod_{k=l+1}^{p}\alpha_{k}\right)^{2}
+\gamma^3 A_{K}^{2}\sum_{j=1}^{q}
+\mathsf{E}\left(\sum_{p=j}^{q}\prod_{k=2}^{p}\alpha_{k}\right)^{2}
 \\
 &\le
-\frac{A_{K}^{2}}{(1-\gamma_1 B_\mu)^2}
-\mathsf{E}\sum_{j=l}^{q} \gamma_{j}
-\left(
-\sum_{p=j+1}^{q+1} \gamma_{p} \prod_{k=l+1}^{p} \alpha_{k}
+\frac{\gamma^3 A_{K}^{2}}{(1-\gamma B_\mu)^2}
+\mathsf{E}\sum_{j=1}^{q} 
+\left(\sum_{p=j+1}^{q+1}  \prod_{k=2}^{p} \alpha_{k}
 \right)^{2}
@@
 &\le
-\frac{A_{K}^{2}}{(1-\gamma_1 B_\mu)^3}
-\mathsf{E}\sum_{j=l}^{q} \gamma_{j}
-\prod_{k=l}^{j} \alpha_{k}
-\left(
-\sum_{p=j+1}^{q+1} \gamma_{p} \prod_{k=j+1}^{p} \alpha_{k}
+\frac{\gamma^3 A_{K}^{2}}{(1-\gamma B_\mu)^3}
+\mathsf{E}\sum_{j=1}^{q} 
+\prod_{k=1}^{j} \alpha_{k}
+\left(\sum_{p=j+1}^{q+1}  \prod_{k=j+1}^{p} \alpha_{k}
 \right)^{2}
@@
 
-Now, from Hoelder inequality, we obtain (with $\|X\|_p = (\mathsf{E}X^p)^{\frac{1}{p}}$)
-\begin{align*}
-\mathsf{E}\left[\sum_{j=l}^{q}\gamma_{j}\prod_{k=l}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\gamma_{p}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right] & \leq\sum_{j=l}^{q}\gamma_{j}\left\Vert \prod_{k=l}^{j}\alpha_{k}\right\Vert _{2}\left\Vert \sum_{p=j+1}^{q+1}\gamma_{p}\prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}^{2}\\
- & \leq\sum_{j=l}^{q}\gamma_{j}\left\Vert \prod_{k=l}^{j}\alpha_{k}\right\Vert _{2}\left(\sum_{p=j+1}^{q+1}\gamma_{p}\left\Vert \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}\right)^{2}.
-\end{align*}
-Now using the fact that $\prod\limits_{k=j+1}^{p}\alpha_{k} \leq \exp\left(-\sum\limits_{k=j+1}^{p}\gamma_{k}\mu'(X_{k-1})\right),$
+Now, from the H\"older inequality, we obtain (with $\|X\|_p = (\mathsf{E}X^p)^{\frac{1}{p}}$)
+\begin{align*}
+\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=l}^{j}\alpha_{k}\left( \sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right] & \leq\sum_{j=1}^{q}\left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left\Vert \sum_{p=j+1}^{q+1} \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}^{2}\\
+ & \leq\sum_{j=1}^{q} \left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left(\sum_{p=j+1}^{q+1}\left\Vert \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}\right)^{2}.
+\end{align*}
+Now using the fact that $\prod\limits_{k=j+1}^{p}\alpha_{k} \leq \exp\left(-\sum\limits_{k=j+1}^{p}\gamma\mu'(X_{k-1})\right),$
 we get 
 \[
-\mathsf{E}\left[\sum_{j=l}^{q}\gamma_{j}\prod_{k=l}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\gamma_{p}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]\leq\sum_{j=l}^{q}\gamma_{j} \zeta^{1/2}_{l,j}(2) \left(\sum_{p=j+1}^{q+1}\gamma_{p} \zeta^{1/4}_{j+1,p}(4)\right)^{2},
+\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=1}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]\leq\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \left(\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4)\right)^{2},
 \]
@@
 \[
-\zeta_{l,j}(s)=\mathsf{E}\left[e^{-s\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})}\right],\quad s>0.
-\]
-From Theorem~\ref{exp_mom_ula} in Appendix~\ref{sec:appendix}, it follows that
-\[
-\mathsf{E}\left[e^{-s\left(\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})-\Gamma_{l,j}\int\mu'(x)\pi(x)\,dx\right)}\right]\leq e^{s\kappa\sum_{k=l}^{j}\gamma_{k}^{2}}
-\]
-with constant $\kappa = ...$. We used the fact that $\mu^{\prime}(x)$ is bounded. Denote $\sum_{k=1}^{\infty}\gamma_{k}^{2} = \Gamma^{\star}$, then we have 
-\[
-\mathsf{E}\left[e^{-s\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})}\right]\leq e^{s\varkappa\Gamma^{\star}}e^{-s\Gamma_{l,j}\int\mu'(x)\pi(x)\,dx}.
-\]
+\zeta_{l,j}(u)=\mathsf{E}\left[e^{-u\sum_{k=l}^{j}\gamma\mu'(X_{k-1})}\right],\quad u>0.
+\]
+From Theorem~\ref{thm:ula_const_size} in Appendix~\ref{sec:appendix}, it follows that
+\[
+\mathsf{E}\left[e^{-u\gamma\left(\sum_{k=l}^{j}[\mu'(X_{k-1})-\pi_\gamma(\mu')]\right)}\right]\leq e^{u^2\varkappa (j-l-1)\gamma^{2}}
+\]
+with constant $\varkappa = ...$, where we used the fact that $\mu^{\prime}(x)$ is bounded. Then we have 
+\[
+\mathsf{E}\left[e^{-s\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})}\right]\leq e^{s^2\varkappa \gamma^2(j-l-1)}e^{-s\gamma (j-l-1)\pi_\gamma(\mu')}.
+\]
+\denis{Please finish}
 Furthermore, since $\mu\pi',\mu'\pi\in L^{1}(\mathbb{R})$, we have
 \[
-\int\mu'(x)\pi(x)\,dx=-\int\mu(x)\pi'(x)\,dx=2\int\mu^{2}(x)\pi(x)\,dx>0
+\pi(\mu')=\int\mu'(x)\pi(x)\,dx=-\int\mu(x)\pi'(x)\,dx=2\int\mu^{2}(x)\pi(x)\,dx>0
 \]
@@
 \[
-\zeta_{l,j}(s) \leq e^{s\kappa \Gamma^*}e^{-2s\Gamma_{l,j}\alpha}
-\]
-where $\Gamma_{l,j} = \sum\limits_{k=l}^{j}\gamma_k, \alpha = \int \mu^2(x)\pi(x)\,dx$. Since $\sum\limits_{k=1}^{\infty}\gamma_k = \infty$, we may estimate
-\begin{align*}
-\sum\limits_{p=j+1}^{\infty} \gamma_p e^{-2\alpha \Gamma_{j+1,p}} < \int\limits_{0}^{+\infty}e^{-2\alpha x}\,dx = \frac{1}{2\alpha}
-\end{align*}
-yielding 
-\begin{align*}
-\sum_{p=j+1}^{q+1}\gamma_{p} \zeta^{1/4}_{j+1,p}(4) \leq e^{\kappa \Gamma^*}\sum_{p=j+1}^{q+1}\gamma_{p}e^{-2\alpha \Gamma_{j+1,p}} \leq \frac{e^{\kappa \Gamma^*}}{2\alpha} 
-\end{align*}
-Using the same integral bound, 
-\begin{align*}
-\mathsf{E}\left[\sum_{j=l}^{q}\gamma_{j}\prod_{k=l}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\gamma_{p}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right] \leq
-\frac{e^{2\kappa \Gamma^*}}{2\alpha}\sum_{j=l}^{q}\gamma_{j}e^{-2\alpha \Gamma_{l,j}} \leq \frac{e^{2\kappa \Gamma^*}}{4\alpha^2}
+\zeta_{l,j}(s) \leq e^{s^2\kappa (j-l-1) \gamma^2}e^{-2s\gamma(j-l-1)\alpha}
+\]
+where $ \alpha = \int \mu^2(x)\pi(x)\,dx$. 
+Finally 
+\begin{align*}
+\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=1}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]  \lesssim 
 \end{align*}
@@
 
-\subsection{Proof of Lemma~\ref{a_decay}}
-
-We have for any \(\mathbf{k}\neq \mathbf{0}\)
-\begin{eqnarray*}
-a_{p,l,\mathbf{k}}(x)=\mathsf{E}\left[\mathbf{H}_\mathbf{k}\left(Z_l\right)[Q_{p,l}\left(\Phi_l(x,Z_l)\right)-Q_{p,l}\left(\Phi_l(x,0)\right)]\right].
-\end{eqnarray*}
-Hence
-\begin{eqnarray*}
-\|a_{p,l,\mathbf{k}}\|_{\infty}\leq \sqrt{\gamma_l}\|\nabla Q_{p,l}\|_{\infty}.
-\end{eqnarray*}
-Since \(f,\mu\in C^1(\mathbb{R}^d),\)  we have
-\begin{eqnarray*}
-\nabla Q_{p,l}(x)=\mathsf{E}\left[(I-\gamma_{l+1}\nabla\mu(x))\nabla Q_{p,l+1}(x-\gamma_{l+1}\mu(x)+\sqrt{\gamma_{l+1}}\xi)\right].
-\end{eqnarray*}
-Hence
-\begin{eqnarray*}
-\|\nabla Q_{p,l}\|_\infty\leq (1-b_{\mu}\gamma_{l+1})\|\nabla Q_{p,l+1}\|_\infty
-\end{eqnarray*}
-and
-\begin{eqnarray*}
-\|\nabla Q_{p,l}\|_\infty&\leq & \|\nabla f\|_\infty\prod_{r=l+1}^p (1-b_{\mu}\gamma_{r})
-\\
-&\leq & \|\nabla f\|_\infty\exp\left(-b_{\mu}\sum_{r=l+1}^p \gamma_{r}  \right).
-\end{eqnarray*}
-As a result
-\begin{eqnarray*}
-\|a_{p,l,\mathbf{k}}\|_{\infty}\leq \sqrt{\gamma_l}\, \|\nabla f\|_\infty\exp\left(-b_{\mu}\sum_{r=l+1}^p \gamma_{r}  \right).
-\end{eqnarray*}
-
 \appendix
-
 \section{Bounds for moments of  ULA}\label{sec:appendix}
-In this appendix we give some definitions and prove auxiliary results which are the used to bound the exponential moment of the additive functional of ULA chain (Theorem~\ref{th:exp_mom_ula}).
-\par
-For $\gamma>0$, the Markov kernel $R_\gamma$ on $(\mathbb R^d,\mathcal B(\mathbb R^d))$, corresponding to the Unadjusted Langevin Algorithm with step $\gamma$ is defined by
-\begin{equation}
-\label{eq:kernel}
-R_\gamma(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \nabla U(x)\right\|^2\right\}dy,\quad A\in \mathcal B(\mathbb R^d),
-\end{equation}
-while, for $k,l\in\mathbb N$, $k\le l$,
-the kernel $Q_{\boldsymbol{\gamma}}^{k,l}$ is
-$
-Q_{\boldsymbol{\gamma}}^{k,l}=R_{\gamma_l}\cdots R_{\gamma_k}.
-$
-We make the following assumptions.
-
-\begin{description}
-\item[{\bf (H1)}] Assume that $\nabla U$ is Lipschitz, that is, there exists $L\geq 0$ such that 
-\begin{eqnarray*}
-\|\nabla U(x)-\nabla U(y)\|\leq L\|x-y\|, \quad x,y\in \mathbb{R}^d.
-\end{eqnarray*}
-
-\item[{\bf (H2)}] There exist $K>0,$ $M>0$ and $m>0$ such that for any $x\not\in B_K(0)	$ and any $y\in \mathbb{R}^d$ it holds 
-\begin{eqnarray*}
-\langle D^2U(x)y,y\rangle \geq m\|y\|^2, \quad \|D^3 U(y)\| \leq M. 
-\end{eqnarray*}
-\end{description}
-
-Due to \cite[Lemma~14]{brosse:2018:diffusion}, conditions {\bf (H1)} and {\bf (H2)} implies that there exist constants $K_1 > 0, m > 0$, such that for any $x \not\in B_{K_1}(0)$ it holds
-\begin{equation}
-\label{eq:H3}
-\langle \nabla U(x), x \rangle \geq \frac{m}{2}\|x\|^2
-\end{equation}
-
-We would show that conditions {\bf (H1)} and {\bf (H2)} are sufficient to show that ULA kernel $R_\gamma$ satisfies the so-called drift condition:
+
+We shall show that conditions {\bf (H1)} and {\bf (H2)} are sufficient to show that ULA kernel $R_\gamma$ satisfies the so-called drift condition:
 \begin{definition}

### 2020-01-28 aap/Revision/vr-mcmc-revision.tex
fixed misprint

@@
 \label{sec:ula_analysis}
-
+In this section we perform the convergence analysis of the ULA algorithm. For the sake of simplicity we restrict our attention to the constant time step.
 We shall use the notations $\mathbb N=\{1,2,\ldots\}$ and $\mathbb N_0=\mathbb N\cup\{0\}$.
@@
 \omega_{p,n}^N\,
-\|Q_\gamma^{p}(x_0,\cdot)-\pi(\cdot)\|_{TV}^2,
+\|\delta_x Q_\gamma^{p}-\pi(\cdot)\|_{TV}^2,
 \notag
@@
 $$
-R_\gamma(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\}dy,
+\mathcal{R}_\gamma(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\}dy,
 $$
 while, for $k,l\in\mathbb N$, $k\le l$,
-the kernel $Q_\gamma^{k,l}$ is
+the kernel $\mathcal{Q}_{\boldsymbol{\gamma}}^{k,l}$ is
 $
-Q_\gamma^{k,l}=R_{\gamma_l}\cdots R_{\gamma_k},
+\mathcal{Q}_{\boldsymbol{\gamma}}^{k,l}=\mathcal{R}_{\gamma_l}\cdots \mathcal{R}_{\gamma_k},
 $
 which, finally, provides the  measure
-$Q_\gamma^{p}(x_0,\cdot)=Q_\gamma^{p,1}$
+$\mathcal{Q}_{\boldsymbol{\gamma}}^{p}=\mathcal{Q}_{\boldsymbol{\gamma}}^{p,1}$
 used in the right-hand side of~\eqref{eq:06042018a2}.
@@
 using results of Section~3 in~\cite{durmus:moulines:2017}
-on bounds in the total variation distance. For example, the following proposition holds.
-\begin{prop}
-Suppose that the assumption (H1) holds, then it holds
-\begin{eqnarray*}
-\|Q_\gamma^{p}(x,\cdot)-\pi(\cdot)\|^2_{TV}\lesssim \sum_{k=r+1}^{p} \gamma_k^2+\exp\left(-\varkappa\sum_{k=r+1}^{p} \gamma_k\right),
-\end{eqnarray*}
-for some $\varkappa>0,$ where $\lesssim$ stands for  inequality up to a constant depending on $L$ and $\mathcal G_N.$ In particular if $\gamma_k=k^{-\alpha}$ for some $\alpha\in (0.5,1], $ then 
-\begin{eqnarray*}
-\|Q_\gamma^{N+1,N+n}(x,\cdot)-\pi(\cdot)\|^2_{TV}\lesssim \frac{1}{(N+1)^{2\alpha-1}}
-\end{eqnarray*}
-provided that $N/n\to 0$ as $N,n\to \infty.$
-\end{prop}
+on bounds in the total variation distance.
+For the case of a constant time step $\gamma_1=\ldots=\gamma_p=\gamma$ it is known that under assumption {\bf (H1)} the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$.
+Moreover, we shall show in Section that  under {\bf (H1)} and {\bf (H2)}
+\begin{eqnarray}
+\label{eq:bias}
+\|\delta_x\mathcal{Q}_{\boldsymbol{\gamma}}^{n}-\pi_\gamma\|_{TV} \leq C\rho^{\gamma n}\left(V(x) + \pi_\gamma(V)\right)
+\end{eqnarray}
+for  $V(x) = 1 + \|x\|^2,$ some $\rho\in (0,1)$ independent of $\gamma$ and an absolute constant $C>0.$
 \subsection{Conditional variance}
 An upper bound for the variance of the classical estimator~\eqref{eq:29032018a2}
-is provided in  in~\cite[Theorem~17]{durmus:moulines:2017}.
+under {\bf (H1)} and {\bf (H2)} is derived in Section.  In particular, if $\sum_{j=1}^\infty \gamma_j=\infty$ and $\sum_{j=1}^\infty \gamma^2_j<\infty,$ then for any bounded function $f,$
+\begin{eqnarray}
+\label{eq:var-mc}
+\mathsf{Var}[\pi_n^N(f)]\lesssim 1/\Gamma_{N+2,N+n+1},
+\end{eqnarray}
+where $\lesssim$ stands for inequality up to a constant depending on constants $L_U,$ $m_U,$  $M_U$ and $\|f\|_\infty.$ For a constant time step,  we derive from \eqref{eq:var-mc} that
+$\mathsf{Var}[\pi_n^N(f)]\lesssim 1/(n\gamma).$
+
 As for the estimator~\eqref{eq:29032018a3}, 
@@
 \end{prop}
-Now we present an upper bound for the right-hand sides of~\eqref{eq:29032018a6} in case of ULA algorithm.
+One of the main results of this paper is the following  upper bound for the right-hand sides of~\eqref{eq:29032018a6} in the case of ULA algorithm.
 \begin{thm}\label{th:mr}
-Assume (H1) and (H2).
+Assume {\bf (H1)} and {\bf (H2)}.
 Suppose that $f$  is two times continuously differentiable
@@
 \begin{eqnarray*}
-d_V(\delta_xQ_{\boldsymbol{\gamma}}^{1,n},\pi) \leq K_3 V(x) \rho^n 
+d_V(\delta_xQ_{\boldsymbol{\gamma}}^{1,n},\pi) \leq K_3 V(x) \rho^{\Gamma_{1,n}} 
 \end{eqnarray*}

### 2020-02-17 aap/vr-mcmc.tex
update

--- 2020-01-23 aap/Revision/vr-mcmc-revision.tex
@@
 \newcommand{\erici}[1]{\todo[inline,color=purple!10, author=Eric]{#1}}
-\newcommand{\denis}[1]{\todo[color=green!10, author=Denis]{#1}}
-\newcommand{\denisi}[1]{\todo[inline,color=green!10, author=Denis]{#1}}
-\newcommand{\sergey}[1]{\todo[color=red!10, author=Sergey]{#1}}
-\newcommand{\sergeyi}[1]{\todo[inline,color=red!10, author=Sergey]{#1}}
+\newcommand{\denis}[1]{\todo[color=green!10, author=Eric]{#1}}
+\newcommand{\denisi}[1]{\todo[inline,color=green!10, author=Eric]{#1}}
 
@@
 \newtheorem{thm}{Theorem}
-\newtheorem{definition}{Definition}
 \newtheorem{lem}[thm]{Lemma}
@@
 
-%added command for indicator - SS
-\newcommand{\indiacc}[1]{\mathbbm{1}_{\{#1\}}}
-
 \newcommand*{\argmin}{\operatornamewithlimits{arg\,min}}
@@
 \and
-\author{\snm{S. Samsonov}\thanksref{b}\ead[label=e3]{svsamsonov@hse.ru}}
+\author{\snm{N. Shagadatov}\thanksref{b}\ead[label=e3]{nshagadatov@edu.hse.ru}}
 \and
-\author{\snm{N. Shagadatov}\thanksref{b}\ead[label=e4]{nshagadatov@edu.hse.ru}}
+\author{\snm{M. Urusov}\thanksref{a}\ead[label=e4]{murusov@uni-due.de}}
 
@@
 $\mathcal{X}\to\mathbb{R}$ with $f\in L^2(\pi)$.
-The idea of the so-called control variates variance reduction method
+The idea of the control variates variance reduction method
  is to find a cheaply computable random variable $\zeta$ with $\mathsf{E}[\zeta]=0$ and \(\mathsf{E}[\zeta^2]<\infty,\)
@@
 Alternatively, if an orthonormal system in \(L^2(\pi)\) is analytically available, one can build control variates \(\zeta\) as a linear combination of the corresponding basis functions, see \cite{GobetCV}.
-Furthermore, if \(\pi\) is known only up to a normalizing constant (which is often the case in Bayesian statistics), one can apply the recent approach of control variates  depending only on the gradient \(\nabla \log \pi\)  using Schr\(\ddot{\text{o}}\)dinger-type Hamiltonian operator in  \cite{mira2013zero} and so-called Stein operator in \cite{brosse2018diffusion}.
+Furthermore, if \(\pi\) is known only up to a normalizing constant (which is often the case in Bayesian statistics), one can apply the recent approach of control variates  depending only on the gradient \(\nabla \log \pi\)  using Schr\(\ddot{\text{o}}\)dinger-type Hamiltonian operator in  \cite{mira2013zero} and Stein operator in \cite{brosse2018diffusion}.
 In some situations \(\pi\) is not known analytically, but \(X\) can be represented as a function of  simple random variables with known distribution.
@@
 Markov chain \(X_p,\) \(p=0,1,2,\ldots,\) whose marginal distribution  converges  to \(\pi\) as time grows.
-One important class of the variance reduction methods in this case  is based on the so-called Poisson equation for the corresponding  Markov chain.
-It was observed in Henderson~\cite{henderson1997variance}  that if a time-homogeneous Markov chain \((X_p)\) is stationary with stationary distribution \(\pi,\) then for any real-valued function \(G \in L^1(\pi) \) defined on the state space of the Markov chain \((X_p),\)  the function \(U(x) := G(x)-\mathsf{E}[G(X_{1})|X_0 = x]\) has zero mean with respect to \(\pi\).
-The best choice for the function \(G\) corresponds to a solution of the so-called Poisson equation  \(\mathsf{E}[G(X_{1})|X_0 = x]-G(x)=-f(x)+\pi(f)\).
+One important class of the variance reduction methods in this case  is based on the Poisson equation for the corresponding  Markov chain.
+It was observed in~\cite{henderson1997variance}  that if a time-homogeneous Markov chain \((X_p)\) is stationary with stationary distribution \(\pi,\) then for any real-valued function \(G \in L^1(\pi) \) defined on the state space of the Markov chain \((X_p),\)  the function \(U(x) := G(x)-\mathsf{E}[G(X_{1})|X_0 = x]\) has zero mean with respect to \(\pi\).
+The best choice for the function \(G\) corresponds to a solution of the Poisson equation  \(\mathsf{E}[G(X_{1})|X_0 = x]-G(x)=-f(x)+\pi(f)\).
 Moreover, it is also related to the minimal asymptotic variance in the corresponding central limit theorem, see \cite{duncan2016variance} and \cite{mira2013zero}.
 Although the Poisson equation involves the quantity of interest \(\pi(f)\)  and can not be  solved explicitly in most cases, this idea still can be used to construct some  approximations for the optimal zero-variance control variates.
-For example,  Henderson~\cite{henderson1997variance} proposed to compute approximations for the solution of the Poisson equation for specific Markov chains with particular emphasis on models arising in stochastic network theory.
+For example,  \cite{henderson1997variance} proposed to compute approximations for the solution of the Poisson equation for specific Markov chains with particular emphasis on models arising in stochastic network theory.
 In \cite{dellaportas2012control} and \cite{brosse2018diffusion}  series-type control variates are introduced and studied for reversible Markov chains.
 It is assumed in \cite{dellaportas2012control}  that the one-step conditional expectations  can be computed explicitly  for a set of basis functions.
-The authors in \cite{brosse2018diffusion} proposed another approach tailored to diffusion setting which doesn't require the computation of integrals of basis functions and only involves  applications of the underlying generator.
+The authors in \cite{brosse2018diffusion} proposed another approach tailored to diffusion setting which does not require the computation of integrals of basis functions and only involves  applications of the underlying generator.
 \par
@@
 standard Gaussian random vectors. If
-${\mu=\frac{\nabla U}{2}}$
+$\mu=(1/2) \nabla U$
 for some continuously differentiable function $U,$ then Markov chain~\eqref{eq:chain} can be used to approximately sample from the density
@@
 \end{algorithm}
-Then, as shown in Metropolis et al.~\cite{metropolis1953equation}, this transition is reversible with respect to \(\pi\) and therefore preserves the stationary density \(\pi\).
-If \(q_p\) have a wide enough
+This transition is reversible with respect to \(\pi\) and therefore preserves the stationary density \(\pi\); see \cite[Chapter~2]{moulines2018}. If \(q_p\) have a wide enough
 support to eventually reach any region
@@
 \label{eq:sde-inv}
-dX_t=b(X_t)\,dt+\sigma(X_t)\,dW_t,\quad t\geq 0,	
+dX_t=b(X_t)\, \rmd t+\sigma(X_t)\, \rmd W_t,\quad t\geq 0,	
 \end{eqnarray}
@@
 \end{prop}
-\par
 Next we show how  the representation \eqref{eq:mart_repr} can be used to efficiently reduce the variance of MCMC algorithms.
@@
 Since \(X_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\geq 1,\)  the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
+
+\section{Analysis of variance reduced ULA}
+\label{sec:ula_analysis}
 The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
@@
 should be small for \(K\) large enough. In this section   we provide  an analysis of the variance-reduced ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
-\section{Analysis of variance reduced ULA}
-\label{sec:ula_analysis}
-
 We shall use the notations $\mathbb N=\{1,2,\ldots\}$ and $\mathbb N_0=\mathbb N\cup\{0\}$.
@@
 For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
-of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we shall be interested in the Mean Squared Error (MSE),
-which can be decomposed as the sum of the squared  bias
-and the  variance:
+of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we are interested in its conditional
+Mean Squared Error (MSE),
+which can be decomposed as the sum of the squared conditional bias
+and the conditional variance:
 \begin{eqnarray}\label{eq:29032018a4}
-\mathrm{MSE}\left[\rho(f)\right]
-&=&\mathsf{E}\left[(\rho(f)-\pi(f))^2\right]
+\mathrm{MSE}\left[\rho(f)|\mathcal G_N\right]
+&=&\mathsf{E}\left[(\rho(f)-\pi(f))^2|\mathcal G_N\right]
 \\
 \nonumber
-&=&\left(\mathsf{E}[\rho(f)]-\pi(f)\right)^2
-+\mathsf{Var}[\rho(f)].
+&=&\left(\mathsf{E}[\rho(f)|\mathcal G_N]-\pi(f)\right)^2
++\mathsf{Var}\left[\rho(f)|\mathcal G_N\right].
 \end{eqnarray}
-Our analysis will be done under the following two assumptions.
-
-\begin{description}
-\item[(H1)][\textbf{Lipschitz continuity}] Assume that $\nabla U$ is Lipschitz, that is, there exists $L_U\geq 0$ such that 
-\begin{eqnarray*}
-|\nabla U(x)-\nabla U(y)|\leq L_U|x-y|, \quad x,y\in \mathbb{R}^d.
-\end{eqnarray*}
-\item[(H2)][\textbf{Convexity outside a ball}] There exist $K_U>0,$ $M_U>0$ and $m_U>0$ such that for any $x\not\in B_{K_U}(0)	$ and any $y\in \mathbb{R}^d$ it holds 
-\begin{eqnarray*}
-\langle D^2U(x)y,y\rangle \geq m_U\|y\|^2, \quad \|D^3 U(y)\| \leq M_U. 
-\end{eqnarray*}
-\end{description}
-First we analyse the biases of the estimators
-$\pi_n^N(f)$ and $\pi_{K,n}^N(f).$
-\subsection{Squared bias}
+The quantities in~\eqref{eq:29032018a4}
+are conditioned on $\mathcal G_N$,
+as it reflects the way the estimators are used for MCMC:
+the path of the Markov chain is simulated only once,
+and we start to use the realized values
+of the Markov chain to construct
+our estimate only after the burn-in period of length~$N$.
+We also notice that, due to the Markovian structure,
+the conditioning on $\mathcal G_N$
+in~\eqref{eq:29032018a4}
+is equivalent to conditioning on $X_N$ only
+(this is particularly clear in the case $\rho(f)=\pi_n^N(f)$
+but requires some calculation
+in the remaining case $\rho(f)=\pi_{K,n}^N(f)$).
+
+\subsection{Squared conditional bias}
 Due to the martingale transform structure
@@
 \[
-\mathsf E\left[M_{K,n}^N(f)\right]=0,
+\mathsf E\left[\left.M_{K,n}^N(f)\right|\mathcal G_N\right]=0,
 \]
@@
 \begin{align}
-\left(\mathsf{E}[\pi_{K,n}^N(f)]-\pi(f)\right)^2
-&=\left(\mathsf{E}[\pi_{n}^N(f)]-\pi(f)\right)^2
+\left(\mathsf{E}[\pi_{K,n}^N(f)|\mathcal G_N]-\pi(f)\right)^2
+&=\left(\mathsf{E}[\pi_{n}^N(f)|\mathcal G_N]-\pi(f)\right)^2
 \label{eq:06042018a2}\\
@@
 \omega_{p,n}^N\,
-\|Q_\gamma^{p}(x_0,\cdot)-\pi(\cdot)\|_{TV}^2,
+\|Q_\gamma^{N+1,p}(X_N,\cdot)-\pi(\cdot)\|_{TV}^2,
 \notag
@@
 for $\gamma>0$, the Markov kernel $R_\gamma$
+from $(\mathbb R^d,\mathcal B(\mathbb R^d))$
+into $(\mathbb R^d,\mathcal B(\mathbb R^d))$
 is defined by
@@
 $
-which, finally, provides the  measure
-$Q_\gamma^{p}(x_0,\cdot)=Q_\gamma^{p,1}$
+which, finally, provides the (random) measure
+$Q_\gamma^{N+1,p}(X_N,\cdot)$
 used in the right-hand side of~\eqref{eq:06042018a2}.
@@
 using results of Section~3 in~\cite{durmus:moulines:2017}
-on bounds in the total variation distance. For example, the following proposition holds.
-\begin{prop}
-Suppose that the assumption (H1) holds, then it holds
-\begin{eqnarray*}
-\|Q_\gamma^{p}(x,\cdot)-\pi(\cdot)\|^2_{TV}\lesssim \sum_{k=r+1}^{p} \gamma_k^2+\exp\left(-\varkappa\sum_{k=r+1}^{p} \gamma_k\right),
-\end{eqnarray*}
-for some $\varkappa>0,$ where $\lesssim$ stands for  inequality up to a constant depending on $L$ and $\mathcal G_N.$ In particular if $\gamma_k=k^{-\alpha}$ for some $\alpha\in (0.5,1], $ then 
-\begin{eqnarray*}
-\|Q_\gamma^{N+1,N+n}(x,\cdot)-\pi(\cdot)\|^2_{TV}\lesssim \frac{1}{(N+1)^{2\alpha-1}}
-\end{eqnarray*}
-provided that $N/n\to 0$ as $N,n\to \infty.$
-\end{prop}
+on bounds in the total variation distance.
+
 \subsection{Conditional variance}
@@
 is provided in  in~\cite[Theorem~17]{durmus:moulines:2017}.
-As for the estimator~\eqref{eq:29032018a3}, 
-it follows from~\eqref{eq:29032018a5}, Theorem~\ref{prop:29032018a1} and independence condition for the sequence $(Z_p)_{p \geq 1}$.
-\begin{prop}
+As for the estimator~\eqref{eq:29032018a3}, we introduce lemma of conditional variance decomposition.
+It follows from~\eqref{eq:29032018a5}, Theorem~\ref{prop:29032018a1} and independence condition for the sequence $(Z_p)_{p \geq 1}$.
+\begin{lem}
 \label{cond_var_dec}
@@
 \begin{equation}\label{eq:29032018a6}
-\mathsf{Var}\left[\pi_{K,n}^{N}(f)\right]
+\mathsf{Var}\left[\left.\pi_{K,n}^{N}(f)\right|\mathcal G_N\right]
 %=\sum_{\min_i (k_i)\ge K+1}
@@
 \sum_{l=N+1}^{N+n}
-\mathsf{E}\left[\left|\bar a_{l,\mathbf{k}}(X_{l-1})\right|^{2}\right].
-\end{equation}
-\end{prop}
+\mathsf{E}\left[\left.\left|\bar a_{l,\mathbf{k}}(X_{l-1})\right|^{2}
+\right|\mathcal G_N\right].
+\end{equation}
+\end{lem}
 Now we present an upper bound for the right-hand sides of~\eqref{eq:29032018a6} in case of ULA algorithm.
 \begin{thm}\label{th:mr}
-Assume (H1) and (H2).
-Suppose that $f$  is two times continuously differentiable
+Fix $K\in\mathbb N$.
+Suppose that $f$  and \(\mu\) are \(K+1\) times continuously differentiable
 and it holds
 \begin{eqnarray*}
-|\partial^{\mathbf{k}} f(x)|\le  B_f, \quad
+|\partial^{\mathbf{k}} f(x)|\le  B_f,\quad |\partial^{\mathbf{k}}\mu(x)|\le  B_\mu, \quad
 x\in\mathbb R^d
 \end{eqnarray*}
-for all \(\mathbf{k}\) with \(0<\|\mathbf{k}\|\leq  2.\) Let $(\gamma_k)_{k\in\mathbb N}$ be a sequence of positive
-and non-increasing step sizes with $\sum_{k=1}^\infty \gamma_k=\infty$ and $\sum_{k=1}^\infty \gamma^2_k<\infty.$ 
-Then it holds under 
+for all \(\mathbf{k}\) with \(0<\|\mathbf{k}\|\leq  K+1\),
+\[
+\jac{\mu}(x)\ge b_\mu I ,\quad x\in\mathbb R^d,
+\]
+for some positive number $b_\mu\in(0,B_\mu]$  where  \(\jac{\mu}\) stands for Jacobian of \(\mu.\)
+Let $(\gamma_k)_{k\in\mathbb N}$ be a sequence of positive
+and non-increasing step sizes with $\sum_{k=1}^\infty \gamma_k=\infty$.
+We  assume that $\gamma_1<\frac1{B_\mu}$ and that
+\begin{equation}\label{eq:17042018a1}
+\sum_{r=j}^\infty \gamma_r\prod_{k=j}^{r}\left[1-\gamma_{k}b_\mu\right]\leq C,\quad\text{for all }j\in\mathbb N,
+\end{equation}
+with some constant \(C\). Then it holds
 \begin{eqnarray}
 \label{eq:var-bound}
-\mathsf{Var}\left[\pi_{K,n}^{N}(f)\right]\lesssim \frac{1}{\Gamma^2_{N+2,N+n+1}},
+\mathsf{Var}\left[\left.\pi_{K,n}^{N}(f)\right|\mathcal G_N\right]\lesssim \frac{1}{\Gamma^2_{N+2,N+n+1}}\sum_{l=N+1}^{N+n}\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
+\left(\frac{\gamma_{l}}{2}\right)^{|I|K},
 \end{eqnarray}
-where \(\lesssim\) stands for inequality up to a constant not depending  on \(n\) and \(N.\)
+where the sum in \eqref{eq:var-bound} is taken over all subsets \(I\) of \(\{1,\ldots,d\}\) and \(\lesssim\) stands for inequality up to a constant not depending  on \(n\) and \(N.\)
 \end{thm}
-\begin{cor}
-
-\end{cor}
+\begin{remark}
+Note that in  Theorem~\eqref{th:mr} we don't require that  the function \(\mu\) itself is bounded.
+Assumption~\eqref{eq:17042018a1} is not restrictive.
+For instance, a straightforward calculation shows
+that \eqref{eq:17042018a1} is satisfied in most interesting case
+$\gamma_k=\const/k^\alpha$ with $\alpha\in (0,1)$.
+\end{remark}
 
@@
 \]
-of the chain \(X\)
-(the so-called ``training paths'').
+of the chain \(X\) (referred to as \emph{training paths}).
 The regression algorithm proceeds with estimating the  functions \((\bar Q_{l})\) by solving  the least squares optimization problems
@@
 \end{eqnarray*}
-Obviously \(\mathsf{E}[\widehat M_{K,n}^N(f)|\TrainSet]=0\) and \(\widehat M_{K,n}^N(f)\) is indeed a valid control variate in a sense that it
+Obviously \(\mathsf{E}[\widehat M_{K,n}^N(f)|\TrainSet]=0\) and \(\widehat M_{K,n}^N(f)\) is indeed a valid control variate in that it
 does not introduce any bias.
@@
 \subfloat[Subfigure 1 list of figures text][d = 2]{
-\includegraphics[width=0.45\textwidth]{../pictures/GM_2d_f1.png}
+\includegraphics[width=0.45\textwidth]{pictures/GM_2d_f1.png}
 \label{fig:subfig1}}
@@
 \subfloat[Subfigure 2 list of figures text][d = 8]{
-\includegraphics[width=0.45\textwidth]{../pictures/GM_8d_f1.png}
+\includegraphics[width=0.45\textwidth]{pictures/GM_8d_f1.png}
 \label{fig:subfig2}}
@@
 \begin{center}
-\includegraphics[width=0.6\textwidth]{../pictures/BLR_banknotes_4d.png}
+\includegraphics[width=0.6\textwidth]{pictures/BLR_banknotes_4d.png}
 \caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Logistic Regression.
@@
 \begin{center}
-\includegraphics[width=0.6\textwidth]{../pictures/Probit_banknotes_4d.png}
+\includegraphics[width=0.6\textwidth]{pictures/Probit_banknotes_4d.png}
 \caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Probit Regression.
@@
 \left(\mathsf{E}\left[\sum_{p=l}^{N+n}\omega_{p,n}^{N}\left(\partial_{y_{l}}^{\mathbf r}f\left(X_{p}\right)-\mathsf{E}\left[\partial_{y_{l}}^{\mathbf r}f\left(X_{p}\right)|X_{l-1}\right]\right)\mathbf{H}_{\mathbf{q}}(Z_{l})\Big|X_{l-1}\right]\right)^{2}
-\right\}.
+\Bigg|\mathcal G_N \right\}.
 \end{multline*}
@@
 \mathsf{Var}\left(\sum_{p=l}^{N+n}\gamma_{p+1}\partial_{y_{l}}^{\mathbf{K}_I}f\left(X_{p}\right)
-\Big|X_{l-1}\right)\right].
+\Big|X_{l-1}\right)
+\Bigg|\mathcal G_N\right].
 \end{eqnarray*}
@@
 \label{eq:definition-alpha}
-\alpha_k=1-\gamma_k \mu'(X_{k-1}),\quad k\in\mathbb N.
+\alpha_k=1-\gamma_k b_\mu,\quad k\in\mathbb N.
 \end{equation}
@@
 \begin{remark}\label{rem:10062018a1}
-
+Notice that, under~\eqref{eq:17042018a1},
+not only~\eqref{eq:06062018a3} but also
+\begin{equation}\label{eq:10062018a1}
+\sum_{r=j}^\infty\gamma_r \prod_{k=j}^r \alpha_k\le\ol C_2
+\end{equation}
+is satisfied with the same constant~$\ol C_2$
+(which is $C$ of~\eqref{eq:17042018a1})
+simultaneously for all $j\in\mathbb N$.
 Below this will allow us to apply
@@
 $$
-with suitable $(\epsilon_{j,p})_{j\ge1,p\ge j+1}$.
+with suitable $(\epsilon_{j,p})_{j\ge1,p\ge j+1}$
+and the constant $\ol C_2$ in~\eqref{eq:10062018a1}
+being independent of~$j$.
 \end{remark}
-The assumption (H1) implies that  $|\mu'(x)|\leq B_\mu$ for some constant $B_\mu>0$ and all $x\in \mathbb{R}^d$. Without loss of generality we suppose that $\gamma_N B_\mu<1.$
+
 \begin{lem}\label{lem:06062018a2}
@@
 \label{eq:08062018b2}
-\left|\partial_{y_l}^r X_p \right| \le C_r\prod_{k=l+1}^p (1-\gamma_k\mu'(X_{k-1}))\quad \text{a.s.} 
+\left|\partial_{y_l}^r X_p \right| \le C_r\prod_{k=l+1}^p \alpha_k \,,
 \end{equation}
@@
 \end{lem}
+\begin{proof}
+The proof is along the same lines as Lemma~\ref{lem:06062018a3}.
+\end{proof}
 \begin{lem}\label{lem:06062018a3}
@@
 \left|\partial_{y_j} \partial_{y_l}^r X_p\right|
-\le c_r\prod_{k=l+1}^p (1-\gamma_k\mu'(X_{k-1})), \quad \text{a.s.}
+\le c_r\prod_{k=l+1}^p \alpha_k
 \end{equation}
@@
 \end{align*}
-with a suitable constant.
+with a suitable constant
+(we can take, e.g., $\const=\frac{B_\mu}{(1-\gamma_1 b_\mu)^2}$).
 By Lemma~\ref{lem:06062018a1}
@@
 \end{align*}
-with some constant ``$\const$'' depending on
+with some constant ``$\const$'',
+which is, in fact, $\frac1{(1-\gamma_1 b_\mu)^2}$ times
+the expression involving
 $B_\mu,r,C_1,\ldots,C_r,c_1,\ldots,c_{r-1}$.
@@
 A_{K}^{2}\sum_{j=l}^{q}\gamma_{j}
-\mathsf{E}\left(\sum_{p=j}^{q}\gamma_{p+1}\prod_{k=l+1}^{p}\alpha_{k}\right)^{2}
+\left(\sum_{p=j}^{q}\gamma_{p+1}\prod_{k=l+1}^{p}\alpha_{k}\right)^{2}
 \\
 &\le
-\frac{A_{K}^{2}}{(1-\gamma_1 B_\mu)^2}
-\mathsf{E}\sum_{j=l}^{q} \gamma_{j}
+\frac{A_{K}^{2}}{(1-\gamma_1 b_\mu)^2}
+\sum_{j=l}^{q} \gamma_{j}
 \left(
@@
 &\le
-\frac{A_{K}^{2}}{(1-\gamma_1 B_\mu)^3}
-\mathsf{E}\sum_{j=l}^{q} \gamma_{j}
+\frac{A_{K}^{2}}{(1-\gamma_1 b_\mu)^3}
+\sum_{j=l}^{q} \gamma_{j}
 \prod_{k=l}^{j} \alpha_{k}
@@
 \right)^{2}
+\\
+&\le
+\frac{A_{K}^{2}}{(1-\gamma_1 b_\mu)^3}C^3=B_K,
 \end{align*}
-
-Now, from Hoelder inequality, we obtain (with $\|X\|_p = (\mathsf{E}X^p)^{\frac{1}{p}}$)
-\begin{align*}
-\mathsf{E}\left[\sum_{j=l}^{q}\gamma_{j}\prod_{k=l}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\gamma_{p}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right] & \leq\sum_{j=l}^{q}\gamma_{j}\left\Vert \prod_{k=l}^{j}\alpha_{k}\right\Vert _{2}\left\Vert \sum_{p=j+1}^{q+1}\gamma_{p}\prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}^{2}\\
- & \leq\sum_{j=l}^{q}\gamma_{j}\left\Vert \prod_{k=l}^{j}\alpha_{k}\right\Vert _{2}\left(\sum_{p=j+1}^{q+1}\gamma_{p}\left\Vert \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}\right)^{2}.
-\end{align*}
-Now using the fact that $\prod\limits_{k=j+1}^{p}\alpha_{k} \leq \exp\left(-\sum\limits_{k=j+1}^{p}\gamma_{k}\mu'(X_{k-1})\right),$
-we get 
-\[
-\mathsf{E}\left[\sum_{j=l}^{q}\gamma_{j}\prod_{k=l}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\gamma_{p}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]\leq\sum_{j=l}^{q}\gamma_{j} \zeta^{1/2}_{l,j}(2) \left(\sum_{p=j+1}^{q+1}\gamma_{p} \zeta^{1/4}_{j+1,p}(4)\right)^{2},
-\]
-where we denote
-\[
-\zeta_{l,j}(s)=\mathsf{E}\left[e^{-s\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})}\right],\quad s>0.
-\]
-From Theorem~\ref{exp_mom_ula} in Appendix~\ref{sec:appendix}, it follows that
-\[
-\mathsf{E}\left[e^{-s\left(\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})-\Gamma_{l,j}\int\mu'(x)\pi(x)\,dx\right)}\right]\leq e^{s\kappa\sum_{k=l}^{j}\gamma_{k}^{2}}
-\]
-with constant $\kappa = ...$. We used the fact that $\mu^{\prime}(x)$ is bounded. Denote $\sum_{k=1}^{\infty}\gamma_{k}^{2} = \Gamma^{\star}$, then we have 
-\[
-\mathsf{E}\left[e^{-s\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})}\right]\leq e^{s\varkappa\Gamma^{\star}}e^{-s\Gamma_{l,j}\int\mu'(x)\pi(x)\,dx}.
-\]
-Furthermore, since $\mu\pi',\mu'\pi\in L^{1}(\mathbb{R})$, we have
-\[
-\int\mu'(x)\pi(x)\,dx=-\int\mu(x)\pi'(x)\,dx=2\int\mu^{2}(x)\pi(x)\,dx>0
-\]
-yielding the final bound
-\[
-\zeta_{l,j}(s) \leq e^{s\kappa \Gamma^*}e^{-2s\Gamma_{l,j}\alpha}
-\]
-where $\Gamma_{l,j} = \sum\limits_{k=l}^{j}\gamma_k, \alpha = \int \mu^2(x)\pi(x)\,dx$. Since $\sum\limits_{k=1}^{\infty}\gamma_k = \infty$, we may estimate
-\begin{align*}
-\sum\limits_{p=j+1}^{\infty} \gamma_p e^{-2\alpha \Gamma_{j+1,p}} < \int\limits_{0}^{+\infty}e^{-2\alpha x}\,dx = \frac{1}{2\alpha}
-\end{align*}
-yielding 
-\begin{align*}
-\sum_{p=j+1}^{q+1}\gamma_{p} \zeta^{1/4}_{j+1,p}(4) \leq e^{\kappa \Gamma^*}\sum_{p=j+1}^{q+1}\gamma_{p}e^{-2\alpha \Gamma_{j+1,p}} \leq \frac{e^{\kappa \Gamma^*}}{2\alpha} 
-\end{align*}
-Using the same integral bound, 
-\begin{align*}
-\mathsf{E}\left[\sum_{j=l}^{q}\gamma_{j}\prod_{k=l}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\gamma_{p}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right] \leq
-\frac{e^{2\kappa \Gamma^*}}{2\alpha}\sum_{j=l}^{q}\gamma_{j}e^{-2\alpha \Gamma_{l,j}} \leq \frac{e^{2\kappa \Gamma^*}}{4\alpha^2}
-\end{align*}
-Plugging into bound for conditional variance, 
-\begin{align*}
-\mathsf{Var}\left[\sum_{p=l}^{q}\gamma_{p+1}\partial_{y_{l}}^{K}f\left(X_{p}\right)\Big|X_{l-1}\right] \leq
-\frac{A_{K}^{2}e^{2\kappa \Gamma^*}}{4\alpha^2(1-\gamma_1 B_\mu)^3}
-\end{align*}
-\par
+where $C$ is the bound from~\eqref{eq:17042018a1}.
 The proof is completed.
@@
 
-\appendix
-
-\section{Bounds for moments of  ULA}\label{sec:appendix}
-In this appendix we give some definitions and prove auxiliary results which are the used to bound the exponential moment of the additive functional of ULA chain (Theorem~\ref{th:exp_mom_ula}).
-\par
-For $\gamma>0$, the Markov kernel $R_\gamma$ on $(\mathbb R^d,\mathcal B(\mathbb R^d))$, corresponding to the Unadjusted Langevin Algorithm with step $\gamma$ is defined by
-\begin{equation}
-\label{eq:kernel}
-R_\gamma(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \nabla U(x)\right\|^2\right\}dy,\quad A\in \mathcal B(\mathbb R^d),
-\end{equation}
-while, for $k,l\in\mathbb N$, $k\le l$,
-the kernel $Q_{\boldsymbol{\gamma}}^{k,l}$ is
-$
-Q_{\boldsymbol{\gamma}}^{k,l}=R_{\gamma_l}\cdots R_{\gamma_k}.
-$
-We make the following assumptions.
-
-\begin{description}
-\item[{\bf (H1)}] Assume that $\nabla U$ is Lipschitz, that is, there exists $L\geq 0$ such that 
-\begin{eqnarray*}
-\|\nabla U(x)-\nabla U(y)\|\leq L\|x-y\|, \quad x,y\in \mathbb{R}^d.
-\end{eqnarray*}
-
-\item[{\bf (H2)}] There exist $K>0,$ $M>0$ and $m>0$ such that for any $x\not\in B_K(0)	$ and any $y\in \mathbb{R}^d$ it holds 
-\begin{eqnarray*}
-\langle D^2U(x)y,y\rangle \geq m\|y\|^2, \quad \|D^3 U(y)\| \leq M. 
-\end{eqnarray*}
-\end{description}
-
-Due to \cite[Lemma~14]{brosse:2018:diffusion}, conditions {\bf (H1)} and {\bf (H2)} implies that there exist constants $K_1 > 0, m > 0$, such that for any $x \not\in B_{K_1}(0)$ it holds
-\begin{equation}
-\label{eq:H3}
-\langle \nabla U(x), x \rangle \geq \frac{m}{2}\|x\|^2
-\end{equation}
-
-We would show that conditions {\bf (H1)} and {\bf (H2)} are sufficient to show that ULA kernel $R_\gamma$ satisfies the so-called drift condition:
-\begin{definition}
-\label{def:drift}
-We say that the Markov kernel $R_\gamma$ satisfies Foster-Lyapunov drift condition if there exist a measurable function $V: X \rightarrow [1; +\infty)$, real numbers $\lambda \in (0,1)$ and $C > 0$ such that for any $x \in X$,
-\begin{eqnarray*}
-RV(x) \leq \lambda^{\gamma} V(x) + \gamma C
-\end{eqnarray*}
-\end{definition}
-
-\begin{lem}
-\label{lem:drift}
- Assume that the potential $U(x), x \in \mathbb{R}^d$ satisfies conditions {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$.
-Then the kernel $R_\gamma$ from \ref{eq:kernel} satisfies drift condition \ref{def:drift} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L^2}$ with drift function $V(x) = 1 + \|x\|^2$, constants $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = \frac{25K_1^2}{8} + 2d + m$ with $K_1$ from \ref{eq:H3}, $m$ from {\bf (H2)}.
-\end{lem}
-
-\begin{proof} Consider $V(x) = 1 + \|x\|^2$, then
-$$
-R_{\gamma}V(x) = \int\limits_{\mathbb{R}^d}V(y)P_{\gamma}(x,dy) = \int\limits_{\mathbb{R}^d}(1+\|y\|^2)\frac{1}{(4\pi\gamma)^{\frac{d}{2}}}\exp{\left(-\frac{\|y-x+\gamma\nabla U(x)\|^2}{4\gamma}\right)}\,dy =
-$$
-
-$$
-=1 + \frac{1}{(4\pi\gamma)^{\frac{d}{2}}}\int\limits_{\mathbb{R}^d}\|z + x - \gamma \nabla U(x)\|^2 \exp{\left(-\frac{\|z\|^2}{4\gamma}\right)}\,dz
-$$
-Let us first consider the case $x \notin B(0,K_1)$. Note that 
-$$
-\|z+x-\gamma \nabla U(x)\|^2 = \|z\|^2 + 2 \langle z, x - \gamma \nabla U(x) \rangle + \gamma^2 \|x - \gamma \nabla U(x)\|^2
-$$ 
-and the linear term vanishes after integration, moreover, for $Z \sim \mathcal{N}(0,2\gamma I_d)$, it holds $\mathbb{E}\|Z\|^2 = 2\gamma d$.
-It remains to notice that due to {\bf H1} and {\bf H2} (which implies \ref{eq:H3}),
-$$
-\|x - \gamma \nabla U(x) \|^2 = \|x\|^2 - 2\gamma \langle \nabla U(x), x\rangle + \gamma^2 \|\nabla U(x)\|^2 \leq
-$$
-
-$$
- \leq (1-\gamma m)\|x\|^2 + 2\gamma^2 L^2 \|x\|^2
-$$
-Thus, plugging everything into expression for $R_{\gamma}V$ and using $\gamma < \frac{m}{4L}$, we obtain
-$$
-R_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + 2\gamma d + (\gamma m - 2\gamma^2 L^2) \leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma (2d + m)
-$$
-Now let $x \in B(0,K_1)$. Then simply using $\|x-\gamma \nabla U(x)\|^2 \leq 2(1+L\gamma)^2\|x\|^2$, we obtain
-$$
-R_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma\left((m - 2\gamma L^2)(1+\|x\|^2) + 2d + 2(1+L\gamma)^2\|x\|^2\right) \leq 
-$$
-
-$$
-\leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma(\frac{25K_1^2}{8} + 2d + m)
-$$
-\end{proof}
-
-For now let us assume for simplicity that we run ULA with fixed step size $\gamma$.
-Then it is known that under assumption {\bf (H1)} corresponding Markov chain would have unique stationary distribution $\pi_\gamma$, which is different from $\pi$.
-Yet this chain will be $V-$geometrically ergodic due to \cite[Theorem~19.4.1]{douc:moulines:priouret:soulier:2018}.
-Namely, the following lemma holds:
-
-\begin{lem}
-\label{lem:v_ergodicity}
- Assume that the potential $U(x), x \in \mathbb{R}^d$ satisfies conditions {\bf (H1)} and {\bf (H2)}. Then for $0 < \gamma < \overline{\gamma}=\frac{m}{4L^2}$, for any $x \in X$ it holds
-$$
-d_V(\delta_xQ_{\boldsymbol{\gamma}}^{1,n},\pi_\gamma) \leq C\rho^n\left(V(x) + \pi_\gamma(V)\right)
-$$ 
-with $V(x) = 1 + \|x\|^2$ and constants
-$$
-C = \left(1 + \exp{\left(-\frac{m\gamma}{2}\right)}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp{\left(-\frac{m\gamma}{2}\right)} - \frac{2b}{1+d})}\right);
-$$
-
-$$
-b = \gamma(\frac{25K_1^2}{8} + 2d + m); \quad \overline{b} = b\exp^{-\frac{m\gamma}{2}} + d; \quad \varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right); \quad
-$$
-
-$$
-\rho = \exp{\left(-\gamma\left(\frac{m}{2} - 2\frac{\frac{25K_1^2}{8} + 2d + m}{d}\right)\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{m\gamma}{2}\right)} + \frac{2b}{d+1})}}\right)}
-$$
-\end{lem}
-
-\begin{proof} Note that the condition {\bf (H1)} implies that the Markov kernel $Q^{1,n}_{\boldsymbol{\gamma}}$ satisfies $(1,\varepsilon)$-Doeblin condition with $\varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right)$.
-Together with drift condition \ref{eq:drift} it allows to apply \cite[Theorem~19.4.1]{douc:moulines:priouret:soulier:2018} with appropriate constants.
-\end{proof}
-
-Aforementioned lemmas allow us to bound the exponential moment of the additive functional of ULA. Namely, the following theorem holds
-
-\begin{thm}
-\label{thm:ula_const_size}
-Suppose that ULA kernel \ref{eq:kernel} satisfies assumptions {\bf (H1)} and {\bf (H2)}, and let $X_1,\ldots,X_n,\ldots$ be generated by ULA with constant step size $\gamma$.
-Let $X_0 = x$ be fixed.
-Then for any bounded function $g(x) : \mathbb{R}^d \rightarrow \mathbb{R}, |g(x)| \leq M$, it holds
-\begin{eqnarray*}
-\mathsf{E}\left[\exp\left(-\rho\gamma\sum_{k=l}^{p}\left(g(X_{k})-\pi_{\gamma}(g)\right)\right)\right] \leq C_1\exp{\left(s^2 \kappa \gamma^2 (p-l+1)\right)}
-\end{eqnarray*}
-for some absolute constant $C_1$ not depending on $l,p$.
-\end{thm}
-
-\begin{proof} Conditions {\bf (H1)} and {\bf (H2)} imply that the Markov kernel $R_\gamma$ is $V-$ergodic (lemma \ref{lem:v_ergodicity}) and satisfies drift condition \ref{lem:drift}.
-Due to \cite[Theorem~3, Fact 3]{moulines:bound_dif:2019}, we obtain the following bound
-\begin{align*}
-\mathsf{E}_x\left[\exp\left(-s\gamma\sum_{k=l}^{p}\left(g(X_{k}) - \mathsf{E}_x g(X_k)\right)\right)\right] \leq \exp{\left(s^2 \kappa \gamma^2 (p-l+1)\right)}
-\end{align*}
-\sergey{I will put precise constant later}
-with constant $\kappa = ...$. Note that
-\begin{align*}
-\mathsf{E}\left[\exp\left(-s\gamma\sum_{k=l}^{p}\left(g(X_{k})-\pi_{\gamma}(g)\right)\right)\right] = 
-\end{align*}
-
-\begin{align*}
-=\mathsf{E}_x\left[\exp{\left(-s\gamma\sum_{k=l}^{p}\left(g(X_{k}) - \mathsf{E}_x g(X_k)\right)\right)}\right]
-\exp{\left(s\gamma\sum\limits_{k=l}^{p}\left(\pi_\gamma (g) - \mathsf{E}_x g(X_k)\right)\right)}
-\end{align*}
-Using lemma~\ref{lem:v_ergodicity},
-\begin{align*}
-|\mathsf{E} g(X_k) - \pi_\gamma(g)| \leq Md_V(\pi_\gamma, \delta_xQ^{1,n}_{\boldsymbol{\gamma}}) \leq CM\rho^k\left(V(x) + \pi_\gamma(V)\right)
-\end{align*}
-Hence,
-\begin{align*}
-\exp{\left(s\gamma\sum\limits_{k=l}^{p}\left(\pi_\gamma (g) - \mathsf{E}_x g(X_k)\right)\right)} \leq \exp{\left(s\gamma CM(V(x) + \pi_\gamma(V))\frac{\rho^l - \rho^{p+1}}{1-\rho}\right)}
-\end{align*}
-and the statement follows.
-\end{proof}
-
-Let us formulate and prove prove analogue of the previous theorem for inhomogeneous Markov chain with step size $\gamma_1,\ldots,\gamma_n,\ldots$.
-Note that under assumptions $\sum_{p=1}^{\infty} \gamma_p=\infty, \quad \sum_{p=1}^{\infty} \gamma^2_p<\infty$ and {\bf (H1)} the stationary distribution of ULA-based chain will be equal to $\pi$.
-We will need additional assumption
-\begin{description}
-\item[{\bf (H3)}] There exist such constants $\rho > 0, K_3 > 0$ that 
-\begin{eqnarray*}
-d_V(\delta_xQ_{\boldsymbol{\gamma}}^{1,n},\pi) \leq K_3 V(x) \rho^n 
-\end{eqnarray*}
-\end{description}
-
-\begin{thm}
-\label{th:exp_mom_ula}
-Suppose that
-\begin{eqnarray*}
-\sum_{p=1}^{\infty} \gamma_p=\infty, \quad \sum_{p=1}^{\infty} \gamma^2_p<\infty.
-\end{eqnarray*}
-Under assumptions {\bf (H1)}, {\bf (H2)} and {\bf (H3)}, for any bounded function $g$ on $\mathbb{R}^d$ 
-we have
-%\begin{eqnarray*}
-%\mathsf{E}\left[\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}g(X_{p})-\bar Q_l(X_l)\right)^{4}\right] \lesssim \frac{1}{\Gamma^4_{N+2,N+n+1}}
-%\end{eqnarray*}
- %and
-%\begin{eqnarray*}
-%\bar Q_l(X_l)=\mathsf{E}\left[\left.\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right|X_{l}\right].
-%\end{eqnarray*}
-%Moreover, 
-\begin{eqnarray}
-\label{eq:exp_decrease_step}
-\mathsf{E}\left[\exp\left(-s\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k})-\left(\sum_{k=l}^{p}\gamma_{k+1}\right)\pi(g)\right)\right)\right]\leq C
-\end{eqnarray}
-for all natural $0<l\leq p<\infty$ with constant $C$ not depending on $l,p.$. Moreover, 
-\begin{eqnarray}
-\label{eq:var_decrease_step}
-\mathsf{Var}_x{\left(\sum_{k=l}^{l+n-1}\frac{\gamma_{k+1}}{\Gamma_{l+1,l+n}}g(X_k) - \mathsf{E}_x \sum_{k=l}^{l+n-1}\frac{\gamma_{k+1}}{\Gamma_{l+1,l+n}}g(X_k)\right)} \leq \frac{C_1}{\Gamma_{l+1,l+n}}
-\end{eqnarray}
-\end{thm}
-\begin{proof}
-
- Note that under assumptions {\bf (H1)}, {\bf (H2)} and {\bf (H3)} it holds due to \cite[Theorem~4]{moulines:bound_dif:2019} that
- \begin{align}
- \label{eq:exp_bounded_dif}
-\mathsf{E}_x\left[\exp\left(-s\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k}) - \mathsf{E}_x \left(\sum\limits_{k=l}^{p}\gamma_{k+1}g(X_k)\right)\right)\right)\right] \leq \exp{\left(s^2 \kappa \sum\limits_{k=l}^{p}\gamma_{k+1}^2\right)} \leq \exp{\left(s^2 \kappa \Gamma^*\right)}
-\end{align}
-\sergey{This theorem is not yet written in the second paper}
-with $\Gamma^* = \sum\limits_{k=1}^{\infty}\gamma_k^2 < \infty$ and $\kappa = ...$. Using the same transformation as in theorem~\ref{thm:ula_const_size}, and using the bound
-\begin{align*}
-|\mathsf{E}_x f(X_k) - \pi(f)| \leq K_3 M \rho^{\Gamma_{1,k}}V(x)
-\end{align*}
-we obtain
-\begin{align*}
-\mathsf{E}\left[\exp\left(-s\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k})-\left(\sum_{k=l}^{p}\gamma_{k+1}\right)\pi(g)\right)\right)\right] \leq \exp{\left(s^2 \kappa \Gamma^*\right)}\exp{\left(K_3 M V(x) \sum\limits_{k=l}^{p}\rho^{\Gamma_{1,k}}\gamma_{k+1}\right)} \leq
-\end{align*}
-
-\begin{align*}
-\leq \exp{\left(s^2 \kappa \Gamma^*\right)}\exp{\left(K_3 M V(x) \rho^{\Gamma_{1,k}} \int\limits_{0}^{+\infty}\rho^y\,dy\right)} = \exp{\left(s^2 \kappa \Gamma^*\right)}\exp{\left(K_3 M V(x) \frac{\rho^{\Gamma_{1,k}}}{\ln{\rho}}\right)}
-\end{align*}
-\end{proof}
-which proves \ref{eq:exp_decrease_step}. Note that from \ref{eq:exp_bounded_dif} by Taylor expansion for small $s$ we obtain
-\begin{align*}
-1 + \frac{s^2}{2}\mathsf{Var}_x{\left(\sum_{k=l}^{l+n-1}\gamma_{k+1}g(X_k) - \mathsf{E}_x \sum_{k=l}^{l+n-1}\gamma_{k+1}g(X_k)\right)} + O(s^3) \leq 1 + s^2 \kappa \sum\limits_{k=l+1}^{l+n}\gamma_{k}^2 + O(s^4)
-\end{align*}
-Since it holds for arbitarily small $s$, the variance is bounded by
-\begin{align*}
-\mathsf{Var}_x{\left(\sum_{k=l}^{l+n-1}\gamma_{k+1}g(X_k) - \mathsf{E}_x \sum_{k=l}^{l+n-1}\gamma_{k+1}g(X_k)\right)} \leq 2\kappa \sum\limits_{k=l+1}^{l+n}\gamma_{k}^2 \leq 2\kappa \Gamma_{l+1,l+n}
-\end{align*}
-Now we need to divide both parts by $\Gamma^2_{l+1,l+n}$ to get \ref{eq:var_decrease_step}. The proof is completed.
 \bibliographystyle{plain}

### 2020-02-17 aap/Revision/vr-mcmc-revision-new.tex
update vr-mcmc-revision-new.tex

@@
 \newcommand*{\ol}{\overline}
+
+\newcommand{\tcr}[1]{\textcolor{red}{#1}}
 
@@
 \label{eq:sde-inv}
-dX_t=b(X_t)\, \rmd t+\sigma(X_t)\, \rmd W_t,\quad t\geq 0,	
+\rmd X_t=b(X_t)\, \rmd t+\sigma(X_t)\, \rmd W_t,\quad t\geq 0,	
 \end{eqnarray}
@@
 process and let \(L\) denote its infinitesimal generator defined by
-\begin{eqnarray*}
-Lg=b^\top \nabla g+\frac{1}{2}\text{Tr}(\sigma^\top D^2g\sigma)
-\end{eqnarray*}
+\begin{equation*}
+Lg=b^\top \nabla g+\frac{1}{2}\operatorname{Tr}(\sigma^\top D^2g\sigma)
+\end{equation*}
+\eric{do we need a trace here ?}
 for any \(g\in C^2(\mathbb{R}^d).\)
@@
 In particular, we have
-\begin{eqnarray*}
+\begin{equation*}
 \mathsf{E}[\phi_i(\xi)\phi_j(\xi)]=\delta_{ij},\quad i,j\in  \mathbb{Z}_{+}
-\end{eqnarray*}
+\end{equation*}
 with \(\xi \sim P_{\xi}.\)
@@
 $\phi_k(\xi)$, $k\ge1$, are centered.
-As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of Shifted Legendre polynomials for "uniform part" and Hermite polynomials for "gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the Shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\) In the sequel, we denote by $(\mathcal{G}_p)_{p\in \mathbb{Z}_+}$  the filtration
+As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of Shifted Legendre polynomials for "uniform part" and Hermite polynomials for "Gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the Shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\) In the sequel, we denote by $(\mathcal{G}_p)_{p\in \mathbb{Z}_+}$  the filtration
 generated by generated by $(\xi_p)_{p \in \mathbb{N}^*}$ with the convention $\mathcal{G}_0=\mathrm{triv}$.
-
+\erici{It would be perhaps interesting to state the conditions satisfied by the functions $(\Phi_l)_{l \geq 1}$ here}
 \begin{thm}\label{prop:29032018a1}
  Let $f$ be a Borel function $\mathbb{R}^{d}\to\mathbb R$ such that
-$\mathsf{E}\left[\left|f(X_{p})\right|^{2}\right]<\infty$,   where $(X_{p})_{p\geq 0}$ is a Markov chain of the form \eqref{eq:chain_gen} for some kernels $(\Phi_l)_{l\geq 1}.$
+$\mathsf{E}\left[\left|f(X_{p})\right|^{2}\right]<\infty$,   where $(X_{p})_{p\geq 0}$ is a Markov chain of the form \eqref{eq:chain_gen} for some sequence of functions $(\Phi_l)_{l\geq 1}.$
 Then,
@@
 provided $\mathsf{E}\left[\left|f(X_{1})\right|^{2}\right]<\infty.$
-Recall that $\mathcal{G}_l=\sigma(\xi_{1},\ldots,\xi_{l}),$
+\tcr{Recall that $\mathcal{G}_l=\sigma(\xi_{1},\ldots,\xi_{l}),$
 $l=1,2,\ldots, $
-and $\mathcal{G}_0=\mathrm{triv}$.
+and $\mathcal{G}_0=\mathrm{triv}$.}
+\eric{do we need this, this is recalled ten lines above}
 Suppose that (\ref{eq:mart_repr})

### 2020-02-17 aap/Revision/vr-mcmc-revision-new.tex
update vr-mcmc-revision-new.tex

@@
 
+\def\PE{\mathsf{E}}
 \newcommandx{\CPE}[3][1=]{{\mathsf E}_{#1}\left[\left. #2 \, \right| #3 \right]}
@@
 for $p>j$, the following representation holds in \(L^2(P)\)
+\eric{I would write \tcr{$L^2(P_\xi)$} to be coherent with the previous notations} 
 \begin{eqnarray}
@@
 \end{cor}
-\begin{proof}
+\begin{proof}[Proof of Theorem~\ref{prop:29032018a1}]
 The expansion obviously holds for $p=1$ and $j=0$.
-Indeed, due to the orthonormality and completeness
-of the system $\left(\phi_{k}\right)$, we have
-\[
-f(X_{1})=\mathsf{E}\left[f(X_{1})\right]+\sum_{k\geq1}a_{1,1,k}(X_0)\phi_{k}(\xi_{1})
-\]
-with
-\[
-a_{1,1,k}(x_0)=\CPE{f(X_{1})\phi_{k}\left(\xi_{1}\right)}{X_0=x_0},
-\]
-provided $\mathsf{E}\left[\left|f(X_{1})\right|^{2}\right]<\infty.$
+Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system of , we have for all $x \in \rset^d$,
+\eric{it is clear, but we must say in which sense this equality should be understood}
+\[
+f(X^x_{1})=\PE[f(X_{1}^x)]+\sum_{k\geq1}a_{1,1,k}(x)\phi_{k}(\xi_{1})
+\]
+with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$, 
+provided $\PE[|f(X^x_{1})|^{2}]<\infty.$
 \tcr{Recall that $\mathcal{G}_l=\sigma(\xi_{1},\ldots,\xi_{l}),$

### 2020-02-18 aap/Revision/vr-mcmc-revision-new.tex
update vr-mcmc-revision-new.tex
update

@@
 \def\PE{\mathsf{E}}
+\def\PVar{\mathsf{Var}}
 \newcommandx{\CPE}[3][1=]{{\mathsf E}_{#1}\left[\left. #2 \, \right| #3 \right]}
+\newcommand{\ps}[2]{\langle #1, #2 \rangle}
+\def\nset{\mathbb{N}}
+\def\rset{\mathbb{R}}
 \usepackage[textsize=footnotesize]{todonotes} % pour voir les commentaires
@@
 %added command for indicator - SS
+\newcommand{\indi}[1]{\mathbbm{1}_{#1}}
 \newcommand{\indiacc}[1]{\mathbbm{1}_{\{#1\}}}
@@
 \newcommand{\tcr}[1]{\textcolor{red}{#1}}
-
+\newcommand{\tvnorm}[1]{\| #1 \|_{\operatorname{TV}}}
+\def\kerULA{\operatorname{Q}}
+\def\eqsp{\,}
 \begin{document}
@@
 \end{equation*}
-\eric{do we need a trace here ?}
+\eric{do we need a trace here ? I think also that $D^2$ is not defined }
 for any \(g\in C^2(\mathbb{R}^d).\)
@@
 for $p>j$, the following representation holds in \(L^2(P)\)
-\eric{I would write \tcr{$L^2(P_\xi)$} to be coherent with the previous notations} 
+\eric{I would write \tcr{$L^2(\mathsf{P})$} to be coherent with the notation $\mathsf{E}$ and we need to define what is the underlying probability space}
 \begin{eqnarray}
@@
 \]
-Since \(X_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq l,\)  the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
+Since \(X_{l-1}\) is independent of \(\xi_{l}\) and \tcr{\(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 1,\)} \eric{I do not understand this}  the r.v.
+\(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
 The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
@@
 \end{equation}
-should be small for \(K\) large enough. Indeed, due to orthonormality  we have
+should be small for \(K\) large enough. Indeed, since $\PE[ \phi_k(\xi_l) \phi_k'(\xi_l)]=0$ if $k \ne k'$, we obtain
 \[
@@
 For the sake of clarity and notational simplicity we restrict our attention to the constant time step, that is, we take $\gamma=\gamma_1=\gamma_2=\ldots.$
-We shall use the notations $\mathbb N=\{1,2,\ldots\}$ and $\mathbb N_0=\mathbb N\cup\{0\}$.
+\tcr{We shall use the notations $\nset=\{1,2,\ldots\}$ and $\nset_0=\mathbb N\cup\{0\}$.} \eric{these notations are used before; it is best to put all the notations in a short paragraph just after the introduction}
 By $H_k$, $k\in\mathbb N_0$,
@@
 and the  variance:
-\begin{eqnarray}\label{eq:29032018a4}
-\mathrm{MSE}\left[\rho(f)\right]
-&=&\mathsf{E}\left[(\rho(f)-\pi(f))^2\right]
-\\
-\nonumber
-&=&\left(\mathsf{E}[\rho(f)]-\pi(f)\right)^2
-+\mathsf{Var}[\rho(f)].
-\end{eqnarray}
-Our analysis will be done under the following two assumptions.
-
+\begin{equation}
+\label{eq:29032018a4}
+\mathrm{MSE}\left[\rho(f)\right] = \PE\left[\left\{\rho(f)-\pi(f)\right\}^2\right]
+=\left\{\PE[\rho(f)]-\pi(f)\right\}^2 +\PVar[\rho(f)].
+\end{equation}
+Our analysis is carried out under the following two assumptions:
 \begin{description}
-\item[(H1)][\textbf{Lipschitz continuity}] Assume that $\nabla U$ is Lipschitz, that is, there exists $L_U\geq 0$ such that
-\begin{eqnarray*}
-|\nabla U(x)-\nabla U(y)|\leq L_U|x-y|, \quad x,y\in \mathbb{R}^d.
+\item[(H1)][\textbf{Lipschitz continuity}] The potential $U$ is differentiable and $\nabla U$ is Lipschitz, that is, there exists $L_U < \infty$ such that
+\begin{eqnarray*}
+|\nabla U(x)-\nabla U(y)|\leq L_U|x-y|, \quad x,y\in \rset^d.
 \end{eqnarray*}
 \item[(H2)][\textbf{Convexity outside a ball}] There exist $K_U>0,$ $M_U>0$ and $m_U>0$ such that for any $x\not\in B_{K_U}(0)$  it holds
-\begin{eqnarray*}
-\langle D^2U(x),x\rangle \geq \frac{m_U}{2}\|x\|^2.
-\end{eqnarray*}
+\begin{equation*}
+\ps{D^2U(x)}{x} \geq (m_U/2)\|x\|^2.
+\end{equation*}
 \end{description}
@@
 $\pi_n^N(f)$ and $\pi_{K,n}^N(f).$
-\subsection{Squared bias}
-Due to the martingale transform structure
+\paragraph{Squared bias:} Due to the martingale transform structure
 of $M_{K,n}^N(f)$,
@@
 the coefficients $a_{p,l,\mathbf{k}}$ in~\eqref{eq:29032018a5}
-with some independent approximations $\widehat a_{p,l,\mathbf{k}}.$ For a bounded Borel function~$f$,
+with some independent approximations $\widehat a_{p,l,\mathbf{k}}.$
+
+Denote by $\pi($ denotes the probability measure on $\mathbb R^d$
+with density $\pi$ of~\eqref{eq:stationary_distr};
+for $\gamma>0$, define the Markov kernel $Q_\gamma$ associated to one-step of the ULA algorithm by
+\[
+\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\} \rmd y.
+\]
+\eric{what is the notation convention here; I do not like much $\mathcal{Q}$ but $Q$ is already define and it is perhaps not a great idea to supersede the previous notation}
+For a bounded Borel function~$f$,
 we can estimate the conditional bias
-similarly to the beginning of ~\cite[Section~4]{durmus:moulines:2017}:
-\begin{align}
-\left(\mathsf{E}[\pi_{K,n}^N(f)]-\pi(f)\right)^2
-&=\left(\mathsf{E}[\pi_{n}^N(f)]-\pi(f)\right)^2
-\label{eq:06042018a2}\\
-&\le
+similarly to~\cite[Section~4]{durmus:moulines:2017}:
+\begin{equation}
+\label{eq:06042018a2}
+\left\{\mathsf{E}[\pi_{K,n}^N(f)]-\pi(f)\right\}^2
+=\left(\mathsf{E}[\pi_{n}^N(f)]-\pi(f)\right)^2
+\le
 \frac{\mathrm{osc}(f)^2}{n}\sum_{p=N+1}^{N+n}
-\|\delta_x Q_\gamma^{p}-\pi(\cdot)\|_{TV}^2,
-\notag
-\end{align}
+\tvnorm{\delta_x \kerULA_\gamma^{p}-\pi(\cdot)}^2,
+\end{equation}
 where
 $\mathrm{osc}(f):=\sup_{x\in\mathbb R^d}f(x)-\inf_{x\in\mathbb R^d}f(x)$,
-$\|\mu-\nu\|_{\TV}$ denotes the total variation distance
+$\tvnorm{\mu-\nu}$ denotes the total variation distance
 between probability measures $\mu$ and $\nu$, that is,
-$$
-\|\mu-\nu\|_{TV}=\sup_{A\in\mathcal B(\mathbb R^d)}
-|\mu(A)-\nu(A)|,
-$$
-$\pi(\cdot)$ denotes the probability measure on $\mathbb R^d$
-with density $\pi$ of~\eqref{eq:stationary_distr};
-for $\gamma>0$, the Markov kernel $Q_\gamma$
-is defined by
-$$
-\mathcal{Q}_\gamma(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\}dy.
-$$
+\[
+\tvnorm{\mu-\nu}=\sup_{A\in\mathcal B(\mathbb R^d)}
+|\mu(A)-\nu(A)| \eqsp.
+\]
 Different specific upper bounds for the squared bias
@@
 on bounds in the total variation distance.
-It is known  that under assumptions {\bf (H1)} and {\bf (H2)}  the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$.
-Moreover, as shown in Proposition (Appendix) \begin{eqnarray}
+It is known  that under assumptions {\bf (H1)} and {\bf (H2)},   the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$.
+Moreover, as shown in Proposition, \tcr{(Appendix)}
+\eric{put a precise reference here} 
+ \begin{equation}
 \label{eq:bias}
-\|\delta_x\mathcal{Q}_{\gamma}^{n}-\pi_\gamma\|_{TV} \leq C_1\rho^{\gamma n}\left(V(x) + \pi_\gamma(V)\right), \quad \|\pi-\pi_\gamma\|_{TV} \leq C_2 \sqrt{\gamma}
-\end{eqnarray}
+\tvnorm{\delta_x\kerULA_{\gamma}^{n}-\pi_\gamma} \leq C_1\rho^{\gamma n}\left(V(x) + \pi_\gamma(V)\right), \quad \tvnorm{\pi-\pi_\gamma} \leq C_2 \sqrt{\gamma}
+\end{equation}
 for  $V(x) = 1 + \|x\|^2,$ some $\rho\in (0,1)$ and  constants $C_1,$ $C_2>0$ independent of $\gamma$  and $n.$
-\subsection{Variance}
-An upper bound for the variance of the classical estimator~\eqref{eq:29032018a2}
+\paragraph{Variance:} An upper bound for the variance of the classical estimator~\eqref{eq:29032018a2}
 under {\bf (H1)} and {\bf (H2)} is derived in Section.  In particular, for any bounded function $f,$
-\begin{eqnarray}
+\begin{equation}
 \label{eq:var-mc}
-\mathsf{Var}[\pi_n^N(f)]\lesssim \frac{1}{n\gamma},
-\end{eqnarray}
+\PVar[\pi_n^N(f)]\lesssim (n\gamma)^{-1} \eqsp,
+\end{equation}
 where $\lesssim$ stands for inequality up to a constant depending on constants $L_U,$ $m_U,$  $M_U$ and $\|f\|_\infty.$
 One of the main results of this paper is the following  upper bound for the variance of $\pi_{K,n}^{N}(f).$
+\erici{the notation $\partial^{\mathbf{k}}$ should be defined here}
 \begin{thm}\label{th:mr}
@@
 and it holds
-\begin{eqnarray}
+\begin{equation}
 \label{eq:smooth-mu}
-|\partial^{\mathbf{k}} f(x)|\le  B_f, \quad |\partial^{\mathbf{k}} \mu (x)|\leq B_\mu, \quad
+|\partial^{\mathbf{k}} f(x)|\le  B_f, \quad |\partial^{\mathbf{k}} U (x)|\leq B_\mu, \quad
 x\in\mathbb R^d
-\end{eqnarray}
+\end{equation}
+\eric{I have replaced $\mu$ by $U$; in this section we use mostly $U$ and not $\mu$; we could use $\mu$ but in such case it would be better to recall that $\mu= \nabla U/2$}
+\eric{Is it required to assume that the gradient is bounded ? It contradicts assumption (H2) which assumes a quadratic growth at infinity - coercivity of the operator ?}
 for all \(\mathbf{k}\) with \(0<\|\mathbf{k}\|\leq  K.\)  Then it holds
@@
 \label{eq:var-bound}
-\mathsf{Var}\left[\pi_{K,n}^{N}(f)\right]\lesssim \frac{\gamma^{K-2}}{n},
+\PVar\left(\pi_{K,n}^{N}(f)\right)\lesssim n^{-1} \gamma^{K-2} \eqsp,
 \end{eqnarray}
-where \(\lesssim\) stands for inequality up to a constant not depending  on $\gamma,$ \(n\) and \(N.\)
+where \tcr{\(\lesssim\) stands for inequality up to a constant not depending  on $\gamma,$ \(n\) and \(N.\)}
+\eric{this is defined less precisely just above}
 \end{thm}
@@
 \end{eqnarray}
-where $F_s(X_0,Z_1,\ldots,Z_s)=\frac{1}{s}\sum_{r=0}^s f(X_r)$ and $\partial_{Z_1}^{\mathbf{k}'}$
+where $F_s(X_0,Z_1,\ldots,Z_s)=s^{-1}\sum_{r=0}^s f(X_r)$ and $\partial_{Z_1}^{\mathbf{k}'}$
 stands for a weak partial derivative of the functional $F_s$ that also can be viewed as discretised version of Malliavin derivative. From \eqref{eq:Ask-part}, we derive
@@
 \left(\frac{\gamma}{2}\right)^{|I|K}
-\mathsf{Var}_x\left(\frac{1}{s}\sum_{p=1}^{s}\partial_{y_{1}}^{\mathbf{K}_I}f\left(X_{p}\right)
+\PVar_x\left( s^{-1} \sum\nolimits_{p=1}^{s}\partial_{y_{1}}^{\mathbf{K}_I}f\left(X_{p}\right)
 \right),
@@
 \[
-\mathbf{\mathbf{K}}_{I}=(K1_{\{1\in I\}},\ldots,K1_{\{d\in I\}}).
+\mathbf{\mathbf{K}}_{I}=K(\indi{I}(1),\ldots, \indi{I}(d)).
 \]
  Finally we show via  the Poincare inequality that under smoothness assumption \eqref{eq:smooth-mu} it holds
-\begin{eqnarray*}
-\mathsf{Var}_x\left(\frac{1}{s}\sum_{p=1}^{s}\partial_{y_{1}}^{\mathbf{K}_I}f\left(X_{p}\right)
+ \eric{Since we do note state was $W_s$ is, it is a bit difficult to guess why we are using a Poincare inequality there}
+\begin{equation*}
+\PVar_x\left(s^{-1} \sum\nolimits_{p=1}^{s}\partial_{y_{1}}^{\mathbf{K}_I}f\left(X_{p}\right)
 \right)\leq W_s(x)
-\end{eqnarray*}
+\end{equation*}
 for all $x$ and some family of functions $(W_s)_{s\geq 1}$ such that  the sum $\sum_{s=1}^n \mathsf{E}[W_s(X_{N+n-s-1})]$ is bounded uniformly in $n\in \mathbb{N}.$

### 2020-02-18 aap/Revision/vr-mcmc-revision-new.tex
update vr-mcmc-revision-new.tex

@@
 \begin{equation}\label{eq:chain}
-X_{p+1}=X_{p}-\gamma_{p+1}\mu(X_{p})+\sqrt{\gamma_{p+1}}Z_{p+1},\end{equation}
+X_{p+1}=X_{p}-\gamma_{p+1}\mu(X_{p})+\sqrt{2 \gamma_{p+1}}Z_{p+1},\end{equation}
 where $\left(Z_{p}\right)_{p\geq1}$ is an i.i.d.\ sequence of $d$-dimensional
 standard Gaussian random vectors. If
-$\mu=(1/2) \nabla U$
+$\mu= \nabla U$
 for some continuously differentiable function $U,$ then Markov chain~\eqref{eq:chain} can be used to approximately sample from the density
@@
 One of the main results of this paper is the following  upper bound for the variance of $\pi_{K,n}^{N}(f).$
-\erici{the notation $\partial^{\mathbf{k}}$ should be defined here}
+We start by introducing some notations.
+For $m\in\mathbb N$, a smooth function
+$h\colon\mathbb R^{d\times m}\to\mathbb R$
+with arguments being denoted
+$(y_1,\ldots,y_m)$, $y_i\in\mathbb R^d$, $i=1,\ldots,m$,
+a multi-index $\mathbf k=(k_i)\in\mathbb N_0^d$,
+and $j\in\{1,\ldots,m\}$,
+we use the notation $\partial^{\mathbf k}_{y_j} h$ for the multiple derivative of $h$
+with respect to the components of~$y_j$:
+\[
+\partial^{\mathbf k}_{j} h(y_1,\ldots,y_m)
+:=\partial^{k_d}_{y_j^d}
+\ldots
+\partial^{k_1}_{y_j^1}
+h(y_1,\ldots,z_m),
+\quad y_j=(y_j^1,\ldots,y_j^d).
+\]
+In the particular case $m=1$ we can drop
+the subscript $y_1$ in that notation.
 \begin{thm}\label{th:mr}
@@
 Suppose additionally that a bounded function $f$ and $\mu=\nabla U$ are $K\geq 2$ times continuously differentiable
-and it holds
+and for all $x\in\mathbb R^d$, $j \in \{1,\dots,d\}$ and  \(\mathbf{k}\) satisfying \(0<\|\mathbf{k}\|\leq  K,\)
 \begin{equation}
 \label{eq:smooth-mu}
-|\partial^{\mathbf{k}} f(x)|\le  B_f, \quad |\partial^{\mathbf{k}} U (x)|\leq B_\mu, \quad
-x\in\mathbb R^d
-\end{equation}
-\eric{I have replaced $\mu$ by $U$; in this section we use mostly $U$ and not $\mu$; we could use $\mu$ but in such case it would be better to recall that $\mu= \nabla U/2$}
-\eric{Is it required to assume that the gradient is bounded ? It contradicts assumption (H2) which assumes a quadratic growth at infinity - coercivity of the operator ?}
-for all \(\mathbf{k}\) with \(0<\|\mathbf{k}\|\leq  K.\)  Then it holds
+|\partial_j^{\mathbf{k}} f(x)|\le  B_f, \quad |\partial_j^{\mathbf{k}} \mu (x)|\leq B_\mu \eqsp.
+\end{equation}
+Then it holds
 \begin{eqnarray}
@@
 \subsection{Proof of Theorem~\ref{th:mr}}
-We start with introducing some notations.
-For $m\in\mathbb N$, a smooth function
-$h\colon\mathbb R^{d\times m}\to\mathbb R$
-with arguments being denoted
-$(y_1,\ldots,y_m)$, $y_i\in\mathbb R^d$, $i=1,\ldots,m$,
-a multi-index $\mathbf k=(k_i)\in\mathbb N_0^d$,
-and $j\in\{1,\ldots,m\}$,
-we use the notation $\partial^{\mathbf k}_{y_j} h$ for the multiple derivative of $h$
-with respect to the components of~$y_j$:
-$$
-\partial^{\mathbf k}_{j} h(y_1,\ldots,y_m)
-:=\partial^{k_d}_{y_j^d}
-\ldots
-\partial^{k_1}_{y_j^1}
-h(y_1,\ldots,z_m),
-\quad y_j=(y_j^1,\ldots,y_j^d).
-$$
-In the particular case $m=1$ we can drop
-the subscript $y_1$ in that notation.
 For $l\le p$, we have the representation
@@
 \end{equation}
-with
-\[
-\Phi(x,y)=x-\gamma\mu(x)+y,
-\quad x,y\in\mathbb R^d.
-\]
-As a consequence,
-for a function $f\colon\mathbb R^d\to\mathbb R$
+with, for $x,y\in\mathbb R^d$,   $\Phi(x,y)=x-\gamma\mu(x)+y$. 
+As a consequence, for a function $f\colon\mathbb R^d\to\mathbb R$
 as in Section~\ref{sec:setup}, we have
@@
 (that is, $f$ and $\mu$ need to be smooth enough).
+\eric{this notation is a bit confusing, $f \circ G_p$ is defined on $p (d+1)$ coordinates}
 Finally, for a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$, we use the notation

### 2020-02-18 aap/Revision/vr-mcmc-revision-new.tex
update vr-mcmc-revision-new.tex

@@
 Suppose additionally that a bounded function $f$ and $\mu=\nabla U$ are $K\geq 2$ times continuously differentiable
-and for all $x\in\mathbb R^d$, $j \in \{1,\dots,d\}$ and  \(\mathbf{k}\) satisfying \(0<\|\mathbf{k}\|\leq  K,\)
+and for all $x\in\mathbb R^d$ and  \(\mathbf{k}\) satisfying \(0<\|\mathbf{k}\|\leq  K,\)
 \begin{equation}
 \label{eq:smooth-mu}
-|\partial_j^{\mathbf{k}} f(x)|\le  B_f, \quad |\partial_j^{\mathbf{k}} \mu (x)|\leq B_\mu \eqsp.
+|\partial^{\mathbf{k}} f(x)|\le  B_f, \quad |\partial^{\mathbf{k}} \mu (x)|\leq B_\mu \eqsp.
 \end{equation}
@@
 \subsection{Proof of Theorem~\ref{th:mr}}
-For $l\le p$, we have the representation
-\[
-X_{p}=G_{p}(X_0,\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p}),
+For $l\le p$ and $x \in \rset^d$, we have the representation
+\[
+X^x_{p}=G_{p}(x,\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p}),
 \]
@@
 (that is, $f$ and $\mu$ need to be smooth enough).
-\eric{this notation is a bit confusing, $f \circ G_p$ is defined on $p (d+1)$ coordinates}
 Finally, for a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$, we use the notation
@@
 \begin{lem}\label{eq:a_repr}
-Fix some \(\mathbf{k},\mathbf{k}'\in \mathbb{N}_0^d\) with \(\mathbf{k}'\le \mathbf{k}\) componentswise.  Then the following representation holds
-$$
+Fix some \(\mathbf{k},\mathbf{k}'\in \mathbb{N}_0^d\) with \(\mathbf{k}'\le \mathbf{k}\) componentwise.  Then the following representation holds
+\[
 \bar a_{p,\mathbf{k}}(x)=\left(\gamma^{|\mathbf{k}'|}\frac{(\mathbf{k}-\mathbf{k}') !}{\mathbf{k}!}\right)^{1/2}
-\mathsf E\left[\left.
-\partial_{1}^{\mathbf{k}'}f(X_p)\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_l)\right|X_{0}=x\right].
-$$
+\PE\left[
+\partial_{1}^{\mathbf{k}'}f(X^x_p)\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_l)\right] \eqsp.
+\]
 \end{lem}

### 2020-02-18 Code/baselines.py
code update: multidimensional case

@@ -69,12 +69,15 @@ def set_function(f_type,traj,inds_arr,params):
         for traj_ind in range(len(traj)):
             for point_ind in range(len(inds_arr)):
                 f_vals[traj_ind,:,point_ind] = copy.deepcopy(traj[traj_ind][:,inds_arr[point_ind]]**2)
-    
     elif f_type == "3rd_moment": #params is ignored in this case
         f_vals = np.zeros((len(traj),len(traj[0]),len(inds_arr)),dtype = float)
         for traj_ind in range(len(traj)):
             for point_ind in range(len(inds_arr)):
                 f_vals[traj_ind,:,point_ind] = copy.deepcopy(traj[traj_ind][:,inds_arr[point_ind]]**3)
+    elif f_type == "sum": #params is ignored in this case
+        f_vals = np.zeros((len(traj),len(traj[0]),len(inds_arr)),dtype = float)
+        for traj_ind in range(len(traj)):
+            f_vals[traj_ind,:,0] = copy.deepcopy(np.sum(traj[traj_ind],axis=1))   
                 
     elif f_type == "exp_linear":
         f_vals = np.zeros((len(traj),len(traj[0]),1),dtype = float)

### 2020-02-18 Code/martingale.py

@@ -1,5 +1,48 @@
 import numpy as np
 import numpy.polynomial as P
+from sklearn.preprocessing import PolynomialFeatures
+from samplers import ULA
+from potentials import GaussPotential,GaussMixture,GausMixtureIdent,GausMixtureSame
+import copy
+
+def H(k, x):
+    if k==0:
+        return 1.0
+    if k ==1:
+        return x
+    if k==2:
+        return (x**2 - 1)/np.sqrt(2)
+    h = hermitenorm(k)(x) /  np.sqrt(math.factorial(k))
+    return h
+
+def split_index(k,d,max_deg):
+    """
+    transforms single index k into d-dimensional multi-index d with max_deg degree each coordinate at most;
+    """
+    k_vec = np.zeros(d,dtype = int)
+    for i in range(d):
+        k_vec[-(i+1)] = k % (max_deg + 1)
+        k = k // (max_deg + 1)
+    return k_vec
+
+def Hermite_val(k_vec,x_vec):
+    P = 1.0
+    d = x_vec.shape[0]
+    for i in range(d):
+        P = P * H(k_vec[i],x_vec[i])
+    return P
+
+def Eval_Hermite(k,x_vec,max_deg):
+    """
+    Evaluates Hermite polynomials at component x_vec by multi-index obtained from single integer k;
+    Args:
+        max_deg - integer, maximal degree of a polynomial at fixed dimension component;
+        k - integer, number of given basis function; 1 <= k <= (max_deg+1)**d
+        x_vec - np.array of shape(d,N), where d - dimension, N - Train (or Test) sample size
+    """
+    k_vec = split_index(k,len(x_vec),max_deg)
+    #now we initialised k_vec
+    return Hermite_val(k_vec,x_vec)
 
 def approx_q(X_train,Y_train,N_traj_train,lag,max_deg):
     """
@@ -11,7 +54,9 @@ def approx_q(X_train,Y_train,N_traj_train,lag,max_deg):
         lag - truncation point for coefficients, those for |p-l| > lag are set to 0;
         max_deg - maximum degree of polynomial in regression
     """
-    coefs_poly = np.zeros((lag,max_deg+1),dtype = float)
+    dim = X_train[0,:].shape[0]
+    print("dimension = ",dim)
+    coefs_poly = np.array([])
     for i in range(lag):
         x_all = np.array([])
         y_all = np.array([])
@@ -21,8 +66,73 @@ def approx_q(X_train,Y_train,N_traj_train,lag,max_deg):
[…]

### 2020-02-18 Code/samplers.py

@@ -12,18 +12,14 @@ def MC_sampler(intseed,Potential,N_test,d):
     traj,traj_grad = Potential.sample(intseed,N_test)
     return traj,traj_grad
 
-def ULA(r_seed,Potential,step, N, n, d, burn_type = "SGLD",main_type = "SGLDFP"):
+def ULA(r_seed,Potential,step, N, n, d, return_noise = False):
     """ MCMC ULA
     Args:
         Potential - one of objects from potentials.py
         step: stepsize of the algorithm
         N: burn-in period
         n: number of samples after the burn-in
-        d: dimensionality of the problem
-        burn_type: type of gradient updates during burn-in period;
-                    allowed values: "full","SGLD","SGLDFP","SAGA"
-        main_type: type of gradient updates during main loop;
-                    allowed values: "full", "SGLD", "SGLDFP", "SAGA"    
+        d: dimensionality of the problem   
     Returns:
         traj: a numpy array of size (n, d), where the trajectory is stored
         traj_grad: numpy array of size (n, d), where the gradients of the 
@@ -31,63 +27,32 @@ def ULA(r_seed,Potential,step, N, n, d, burn_type = "SGLD",main_type = "SGLDFP")
     """
     np.random.seed(r_seed)
     #select method for gradient updates during burn-in
-    if burn_type == "full":
-        grad_burn = Potential.gradpotential
-    elif burn_type == "SGLD":
-        grad_burn = Potential.stoch_grad
-    elif burn_type == "SGLDFP":
-        grad_burn = Potential.stoch_grad_fixed_point
-    #elif burn_type == "SAGA":
-        #grad_burn = Potential.stoch_grad_SAGA
-    else:
-        raise "Not implemented error: invalid value in ULA sampler, in burn_type"
+    grad_burn = Potential.gradpotential
     #select method for gradient updates during main loop   
     traj = np.zeros((n, d))
-    traj_grad = np.zeros((n, d))
+    #traj_grad = np.zeros((n, d))
+    if return_noise:
+        noise = np.zeros((n, d))
     x = np.random.normal(scale=5.0, size=d) # initial value X_0
     for k in np.arange(N): # burn-in period
         grad_burn_val = grad_burn(x)
         x = x + step * grad_burn_val +\
             np.sqrt(2*step)*np.random.normal(size=d)
     #burn-in ended
-    if main_type == "SAGA":#compute SAGA updates here
-        grads_SAGA = Potential.update_gradients(np.arange(Potential.p),x)
[…]

### 2020-02-18 aap/Revision/vr-mcmc-revision-new.tex
update vr-mcmc-revision-new.tex

@@
 In the case $\Phi_l=\Phi$ for all $l\geq 1,$ we have
-\begin{eqnarray*}
+\begin{equation}
+\label{eq:definition-bar-a}
 \bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
-\end{eqnarray*}
+\end{equation}
 with  \(Q_{r}(x)=\mathsf{E}\left[\left.f(X_{r})\right|X_{0}=x\right],\) $r\in \mathbb{N}.$
@@
 where
-\[
-A_{s,k}(x)=\frac{1}{s}\sum_{r=0}^{s}\bar a_{r,k}(x).
-\]
+\begin{equation}
+\label{eq:definition-A-s,k}
+A_{s,k}(x)=\frac{1}{s}\sum_{r=1}^{s}\bar a_{r,k}(x).
+\end{equation}
 Since \(X_{l-1}\) is independent of \(\xi_{l}\) and \tcr{\(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 1,\)} \eric{I do not understand this}  the r.v.
@@
 \partial^{k_1}_{y_j^1}
-h(y_1,\ldots,z_m),
+h(y_1,\ldots,y_m),
 \quad y_j=(y_j^1,\ldots,y_j^d).
@@
 Assume {\bf (H1)} and {\bf (H2)}.
-Suppose additionally that a bounded function $f$ and $\mu=\nabla U$ are $K\geq 2$ times continuously differentiable
+Suppose additionally that a bounded function $f$ and $\mu=\nabla U$ are $K \times d \geq 2$ times continuously differentiable
 and for all $x\in\mathbb R^d$ and  \(\mathbf{k}\) satisfying \(0<\|\mathbf{k}\|\leq  K,\)
@@
 \PE\left[
-\partial_{1}^{\mathbf{k}'}f(X^x_p)\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_l)\right] \eqsp.
-\]
+\partial_{1}^{\mathbf{k}'}f(X^x_p)\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_1)\right] \eqsp,
+\]
+where $\bar a_{p,\mathbf{k}}$ is defined in \eqref{eq:definition-bar-a}.
 \end{lem}
@@
 \begin{proof}
-Let $\boldsymbol{\varphi}(z)=\frac{1}{(2\pi)^{d/2}} \exp\{-|z|^2/2\}$,
+Let $\boldsymbol{\varphi}(z)=(2\pi)^{-d/2} \exp(-|z|^2/2)$,
 $z\in\mathbb R^d$, denote the density of a $d$-dimensional
@@
 \mathbf{H}_{\mathbf{k}}(z_{1})\boldsymbol{\varphi}(z_1)
-\prod_{j=2}^p\boldsymbol{\varphi}(z_j)\, dz_{1}\ldots dz_{p}
+\prod_{j=2}^p\boldsymbol{\varphi}(z_j)\, \rmd z_{1}\ldots \rmd z_{p}
 \\
@@
 (-1)^{|\mathbf{k}|}\partial^{\mathbf{k}} \boldsymbol{\varphi}(z_1)
-\prod_{j=2}^p\boldsymbol{\varphi}(z_j)\, dz_{1}\ldots dz_{p}
+\prod_{j=2}^p\boldsymbol{\varphi}(z_j)\, \rmd z_{1}\ldots  \rmd z_{p}
 \\
@@
 (-1)^{|\mathbf{k}-\mathbf k'|}\partial^{\mathbf{k}-\mathbf k'} \boldsymbol{\varphi}(z_1)
-\prod_{j=2}^p\boldsymbol{\varphi}(z_j)\, dz_{1}\ldots dz_{p}
+\prod_{j=2}^p\boldsymbol{\varphi}(z_j)\, \rmd z_{1}\ldots \rmd z_{p}
 \\
@@
 and $\mathbf k'\ne\mathbf k$,
-we get applying first Lemma~\ref{eq:a_repr}
+we get applying first Lemma~\ref{eq:a_repr},
 \begin{equation*}
@@
 \,
-\mathsf{E}_x\left[
-\frac{1}{s}\sum_{r=1}^{s}(\partial_{1}^{\mathbf{k}'} f(X_r)-\mathsf{E}_x\partial_{1}^{\mathbf{k}'}f(X_r))\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_1)\right]
+\PE\left[
+s^{-1} \sum\nolimits_{r=1}^{s}\{\partial_{1}^{\mathbf{k}'} f(X^x_r)-\PE[\partial_{1}^{\mathbf{k}'}f(X^x_r)]\}\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_1)\right]
 \end{equation*}
-
+where $A_{s,\mathbf{k}}$ is defined in \eqref{eq:definition-A-s,k}.
 Assume that $\mu$ and $f$ are $K\times d$ times continuously differentiable.

### 2020-02-19 Code/martingale.py
code update

@@ -109,7 +109,8 @@ def test_traj(r_seed,max_deg):
             a_cur = np.zeros(...,dtype=float)
             for m in range(len(coefs_poly[0])):
                 poly_vspom = np.zeros((max_deg_regr+1)*d,dtype=float)
-                poly_indices = split_index(k,
+                poly_indices = split_index(m,d,max_deg_regr)
+                print(poly_indices)
                 for u in range(m+1):
                     deg = u % d
                     poly_vspom[u] = ((x-gamma*Cur_pot.gradpotential(x))**(m-deg))*((np.sqrt(2*gamma))**deg)*sp.special.binom(m,deg)

### 2020-02-19 aap/Revision/vr-mcmc-revision-new.tex

@@
 \end{equation}
-should be small for \(K\) large enough. Indeed, since $\PE[ \phi_k(\xi_l) \phi_k'(\xi_l)]=0$ if $k \ne k'$, we obtain
-\[
-\mathsf{Var}[\pi_{K,n}^N(f)]\leq\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\mathsf{E}[A^2_{n-l,k}(X_{N+l-1})]
-\]
+should be small for \(K\) large enough. Indeed, since $\PE[ \phi_k(\xi_l) \phi_{k'}(\xi_l)]=0$ if $k \ne k'$, we obtain
+\begin{equation}
+\label{eq:expression-variance}
+\PVar[\pi_{K,n}^N(f)]\leq\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X_{N+l-1})]
+\end{equation}
 and $\mathsf{Var}[\pi_{K,n}^N(f)]$ is small provided that the coefficients $A_{s,k}$ decay fast enough as $k\to \infty.$
@@
 \begin{lem}\label{eq:a_repr}
-Fix some \(\mathbf{k},\mathbf{k}'\in \mathbb{N}_0^d\) with \(\mathbf{k}'\le \mathbf{k}\) componentwise.  Then the following representation holds
+For any \(\mathbf{k},\mathbf{k}'\in \mathbb{N}_0^d\) such that \(\mathbf{k}'\le \mathbf{k}\) componentwise and $\| \mathbf{k}' \| \leq K$, the following representation holds
 \[
@@
 with $\mathbf k'\le\mathbf k$ componentwise
-and $\mathbf k'\ne\mathbf k$,
+and $\mathbf k'\ne\mathbf k$, $\| k' \| \leq K$, 
 we get applying first Lemma~\ref{eq:a_repr},
@@
 by taking $\mathbf k'= \mathbf k'(\mathbf k)
-=(K1_{\{k_1>K\}},\ldots,K1_{\{k_d>K\}})$, we get
-\begin{align}
+=K(\indiacc{k_1>K}\,\ldots, \indiacc{k_d>K})$, we get
+\begin{multline}
 \label{eq:sum_abar}
-\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}A^2_{s,\mathbf{k}}(x) & =\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}\left(\gamma^{|\mathbf{k}'|}\frac{(\mathbf{k}-\mathbf{k}')!}{\mathbf{k}!}\right)Q_s(\mathbf{k}',\mathbf{k}-\mathbf{k}')\\
-\nonumber
- & =\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}\gamma^{|I|K}\sum_{\mathbf{m}_{I}\in\mathbb{N}_{I}^{d}}\frac{\mathbf{m}_{I}!}{\left(\mathbf{m}_{I}+\mathbf{K}_{I}\right)!}
- \\
- \nonumber
- & \hspace{1em}\times\sum_{\mathbf{m}_{I^c}\in \mathbb{N}^d_{0,I^c},\,\|\mathbf{m}_{I^c}\|\leq K}Q_s(\mathbf{\mathbf{K}}_{I},\mathbf{m}_{I}+\mathbf{m}_{I^c}),
-\end{align}
+\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}A^2_{s,\mathbf{k}}(x)  
+=\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}\left(\gamma^{|\mathbf{k}'|}\frac{(\mathbf{k}-\mathbf{k}')!}{\mathbf{k}!}\right)Q_s(\mathbf{k}',\mathbf{k}-\mathbf{k}')\\
+= \left\{ \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}\gamma^{|I|K}\sum_{\mathbf{m}_{I}\in\mathbb{N}_{I}^{d}}\frac{\mathbf{m}_{I}!}{\left(\mathbf{m}_{I}+\mathbf{K}_{I}\right)!} \right\}
+\left\{ \sum_{\mathbf{m}_{I^c}\in \mathbb{N}^d_{0,I^c},\,\|\mathbf{m}_{I^c}\|\leq K}Q_s(\mathbf{\mathbf{K}}_{I},\mathbf{m}_{I}+\mathbf{m}_{I^c}) \right\},
+\end{multline}
 where for any two multi-indices \(\mathbf{r},\) \(\mathbf{q}\) from \(\mathbb{N}_0^d\)
-\begin{eqnarray*}
+\begin{equation*}
 Q_s(\mathbf{r},\mathbf{q})
 =
-\left(\mathsf{E}_x\left[\frac{1}{s}\sum_{p=1}^{s}\left(\partial_{y_{1}}^{\mathbf r}f\left(X_{p}\right)-\mathsf{E}_x\left[\partial_{y_{1}}^{\mathbf r}f\left(X_{p}\right)\right]\right)\mathbf{H}_{\mathbf{q}}(Z_{1})\right]\right)^{2}.
-\end{eqnarray*}
+\left\{\PE\left[\frac{1}{s}\sum_{p=1}^{s}\left\{\partial_{1}^{\mathbf r}f\left(X^x_{p}\right)-\PE\left[\partial_{1}^{\mathbf r}f\left(X^x_{p}\right)\right]\right\}\mathbf{H}_{\mathbf{q}}(Z_{1})\right]\right\}^{2}.
+\end{equation*}
 In \eqref{eq:sum_abar} the first sum runs over all nonempty subsets $I$ of the set $\{1,\ldots,d\}.$
@@
 \sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}A^2_{s,\mathbf{k}}(x)&\leq &
-\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset} (\gamma_{l}/2)^{|I|K}
+\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset} (\gamma/2)^{|I|K}
 \\
@@
 &\leq &
-\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset} (\gamma_{l}/2)^{|I|K} \sum_{\mathbf{m}\in\mathbb{N}_0^{d}} Q(\mathbf{\mathbf{K}}_{I},\mathbf{m}).
+\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset} (\gamma/2)^{|I|K} \sum_{\mathbf{m}\in\mathbb{N}_0^{d}} Q(\mathbf{\mathbf{K}}_{I},\mathbf{m}).
 \end{eqnarray*}
@@
 &\leq & \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
-\left(\frac{\gamma_{l}}{2}\right)^{|I|K}
-\mathsf{Var}_x\left(\frac{1}{s}\sum_{p=1}^{s}\partial_{1}^{\mathbf{K}_I}f\left(X_{p}\right)
+\left(\frac{\gamma}{2}\right)^{|I|K}
+\PVar\left(s^{-1}\sum\nolimits_{p=1}^{s}\partial_{1}^{\mathbf{K}_I}f\left(X^x_{p}\right)
 \right)
@@
 \begin{eqnarray*}
-\mathsf{Var}_x\left(\frac{1}{s}\sum_{p=1}^{s}\partial_{1}^{\mathbf{K}_I}f\left(X_{p}\right)
+\PVar\left(s^{-1}\sum\nolimits_{p=1}^{s}\partial_{1}^{\mathbf{K}_I}f\left(X^x_{p}\right)
 \right)\leq W_s(x)

### 2020-02-19 aap/Revision/vr-mcmc-revision-new.tex
update vr-

@@
 we get
-\begin{eqnarray*}
-\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}A^2_{s,\mathbf{k}}(x)&\leq &
+\begin{align}
+\label{eq:sumA}
+\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}A^2_{s,\mathbf{k}}(x)
+&\leq 
 \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset} (\gamma/2)^{|I|K}
 \\
-&& \times\sum_{\mathbf{m}_{I}\in\mathbb{N}_{I}^{d}} \sum_{\mathbf{m}_{I^c}\in \mathbb{N}^d_{0,I^c},\,\|\mathbf{m}_{I^c}\|\leq K} Q(\mathbf{\mathbf{K}}_{I},\mathbf{m}_{I}+\mathbf{m}_{I^c})
+\nonumber
+& \times\sum_{\mathbf{m}_{I}\in\mathbb{N}_{I}^{d}} \sum_{\mathbf{m}_{I^c}\in \mathbb{N}^d_{0,I^c},\,\|\mathbf{m}_{I^c}\|\leq K} Q(\mathbf{\mathbf{K}}_{I},\mathbf{m}_{I}+\mathbf{m}_{I^c})
 \\
-&\leq &
+\nonumber
+&\leq 
 \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset} (\gamma/2)^{|I|K} \sum_{\mathbf{m}\in\mathbb{N}_0^{d}} Q(\mathbf{\mathbf{K}}_{I},\mathbf{m}).
-\end{eqnarray*}
-Now using the consequence
-$\sum_{\mathbf{m}\in \mathbb{N}^d_0} \langle\xi,\mathbf{H}_\mathbf{m}(Z_1)\rangle^2\le\langle\xi,\xi\rangle$
-of the Parseval's identity,
-we derive
+\end{align}
+The Parseval identity implies that for any function $\varphi: \rset^d \to \rset$ satisfying $\PE[\varphi^2(Z_1)] < \infty$,
+\[
+\sum_{\mathbf{m}\in \nset^d_0} \{\PE[\varphi(Z_1) \mathbf{H}_\mathbf{m}(Z_1)] \}^2 \le \PE[\{\varphi(Z_1)\}^2]
+\]
+Using this identity in \eqref{eq:sumA} implies 
 \begin{eqnarray*}
@@
 
- For the sake of simplicity we present the proof only in one-dimensional case. First we need to prove several auxiliary results.
+To keep the notational burden at a reasonable level, we present the proof only in one-dimensional case. 
+Multidimensional extension is straightforward but requires involved notations. 
+First we need to prove several auxiliary results.
 \begin{lem}\label{lem:06062018a1}

### 2020-02-19 aap/Revision/vr-mcmc-revision-new.tex
notations update

--- 2020-02-19 aap/Revision/vr-mcmc-revision-new.tex
@@
 \begin{equation}\label{eq:chain}
-X_{p+1}=X_{p}-\gamma_{p+1}\mu(X_{p})+\sqrt{2 \gamma_{p+1}}Z_{p+1},\end{equation}
+X_{p+1}=X_{p}-\gamma_{p+1}\mu(X_{p})+\sqrt{\gamma_{p+1}}Z_{p+1},\end{equation}
 where $\left(Z_{p}\right)_{p\geq1}$ is an i.i.d.\ sequence of $d$-dimensional
@@
 \begin{equation}\label{eq:stationary_distr}
-\pi(x)= \mathrm{const}\,\rme^{-U(x)}, \quad \mathrm{const}= \left. 1 \middle /{\int_{\mathbb{R}^{d}} \rme^{-U(x)}\, \rmd x,} \right.
+\pi(x)= \mathrm{const}\,\rme^{-\frac{U(x)}{2}}, \quad \mathrm{const}= \left. 1 \middle /{\int_{\mathbb{R}^{d}} \rme^{-\frac{U(x)}{2}}\, \rmd x,} \right.
 \end{equation}
@@
 $\phi_k(\xi)$, $k\ge1$, are centered.
-As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of Shifted Legendre polynomials for "uniform part" and Hermite polynomials for "Gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the Shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\) In the sequel, we denote by $(\mathcal{G}_p)_{p\in \mathbb{Z}_+}$  the filtration
-generated by generated by $(\xi_p)_{p \in \mathbb{N}^*}$ with the convention $\mathcal{G}_0=\mathrm{triv}$.
-\erici{It would be perhaps interesting to state the conditions satisfied by the functions $(\Phi_l)_{l \geq 1}$ here}
+As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of Shifted Legendre polynomials for "uniform part" and Hermite polynomials for "Gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the Shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\) In the sequel, we denote by $(\mathcal{G}_p)_{p\in \mathbb{Z}_+}$  the filtration generated by independent random variables $(\xi_p)_{p \in \mathbb{N}^*}$ with the convention $\mathcal{G}_0=\mathrm{triv}$.
+For $l \le p$ and $x \in \rset^d$, put for $x,y \in \rset^d$ functions $\Phi_k(x,y)=x-\gamma_k\mu(x)+y$.
+Then we can define
+\[
+X^x_{l,p}=G_{l,p}(x,\sqrt{\gamma_l}\xi_{l},\ldots,\sqrt{\gamma_p}\xi_{p}),
+\]
+with the function \(G_{l,p}:\) \(\mathbb{R}^{d\times(p-l+2)}\to \mathbb{R}^{d}\) is defined as
+\begin{equation}
+\label{eq:definition-G-p-l}
+G_{l,p}(x,y_l,\ldots,y_p):=\Phi_p(\cdot,y_{p})\circ\Phi_{p-1}(\cdot,y_{p-1})\circ\ldots\circ\Phi_{l}(x,y_{l})
+\end{equation}
+For simplicity we will write $X^x_{p}$ and $G_{p}$ instead of $X^x_{0,p}$ and $G_{0,p}$. As a consequence, for a function $f\colon \rset^d \to \rset$ we have
+$$
+f\left(X^x_{p}\right) =%f\circ G_{p}(x,\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p}).
+f\circ G_{p}(x,\sqrt{\gamma}\xi_{1},\ldots,\sqrt{\gamma}\xi_{p})
+$$
+%\erici{It would be perhaps interesting to state the conditions satisfied by the functions $(\Phi_l)_{l \geq 1}$ here}
 \begin{thm}\label{prop:29032018a1}
- Let $f$ be a Borel function $\mathbb{R}^{d}\to\mathbb R$ such that
-$\mathsf{E}\left[\left|f(X_{p})\right|^{2}\right]<\infty$,   where $(X_{p})_{p\geq 0}$ is a Markov chain of the form \eqref{eq:chain_gen} for some sequence of functions $(\Phi_l)_{l\geq 1}.$
+ Let $f$ be a Borel function $\rset^{d}\to\mathbb R$ such that
+$\mathsf{E}\left[\left |f(X_{p})\right|^{2}\right]<\infty$,   where $(X_{p})_{p\geq 0}$ is a Markov chain of the form \eqref{eq:chain_gen} for some sequence of functions $(\Phi_l)_{l\geq 1}.$
 Then,
@@
 \label{eq:mart_repr}
-f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}a_{p,l,k}(X_{l-1})\phi_k\left(\xi_{l}\right),
+f(X^x_{p})=\mathsf{E}\left[\left.f(X^x_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}a_{p,l,k}(X^x_{l-1})\phi_k\left(\xi_{l}\right),
 \end{eqnarray}
@@
 \label{eq:coeff_mart}
-a_{p,l,k}(x)=\mathsf{E}\left[\left.f(X_{p})\phi_k\left(\xi_{l}\right)\right|X_{l-1}=x\right], \quad p\geq l, \quad k\in \mathbb{N}.
+%a_{p,l,k}(x)=\mathsf{E}\left[\left.f(X_{p})\phi_k\left(\xi_{l}\right)\right|X_{l-1}=x\right], \quad p\geq l, \quad k\in \mathbb{N}.
+a_{p,l,k}(x)=\mathsf{E}\left[f(X^x_{l-1,p})\phi_k\left(\xi_{l}\right)\right], \quad p\geq l, \quad k\in \mathbb{N}
 \end{eqnarray}
 \end{thm}
-\begin{cor}
-If  all the kernels $\Phi_l,$ $l\geq 1,$ are equal,  then  the representation \eqref{eq:mart_repr} takes the form
-\begin{eqnarray*}
-f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}\bar a_{p-l,k}(X_{l-1})\phi_k\left(\xi_{l}\right)
-\end{eqnarray*}
-with
-\begin{eqnarray*}
-\bar a_{r,k}(x)=\mathsf{E}\left[\left.f(X_{r})\phi_k\left(\xi_{1}\right)\right|X_{0}=x\right], \quad r, k\in \mathbb{N}.
-\end{eqnarray*}
-
-\end{cor}
-\begin{proof}[Proof of Theorem~\ref{prop:29032018a1}]
+
+\begin{proof}%[Proof of Theorem~\ref{prop:29032018a1}]
 The expansion obviously holds for $p=1$ and $j=0$.
-Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system of , we have for all $x \in \rset^d$,
+Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system in \(L^2(\mathbb{R}^d, P_{\xi})\), we have for all $x \in \rset^d$,
 \eric{it is clear, but we must say in which sense this equality should be understood}
@@
 \end{proof}
+
+\begin{cor}
+If  all the kernels $\Phi_l,$ $l\geq 1,$ are equal,  then  the representation \eqref{eq:mart_repr} takes the form
+\begin{eqnarray*}
+f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}\bar a_{p-l,k}(X_{l-1})\phi_k\left(\xi_{l}\right)
+\end{eqnarray*}
+with
+\begin{eqnarray*}
+\bar a_{r,k}(x)=\mathsf{E}\left[f(X^x_{r})\phi_k\left(\xi_{1}\right)\right], \quad r, k\in \mathbb{N}.
+\end{eqnarray*}
+\end{cor}
+
 From  numerical point of view another representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
@@
 \end{eqnarray*}
-with \(Q_{p,l}(x)=\mathsf{E}\left[\left.f(X_{p})\right|X_{l}=x\right],\) \(p\geq l.\)
+with \(Q_{p,l}(x)=\mathsf{E}\left[f(X^x_{l,p})\right],\) \(p\geq l.\)
 %where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
@@
 \label{eq:qpl}
-Q_{p,l}(x)=\mathsf{E}\left[\left.Q_{p,l+1}(X_{l+1})\right|X_{l}=x\right].
+Q_{p,l}(x)=\mathsf{E}\left[Q_{p,l+1}(X^x_{l+1})\right].
 %=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))]
@@
 \end{equation}
-with  \(Q_{r}(x)=\mathsf{E}\left[\left.f(X_{r})\right|X_{0}=x\right],\) $r\in \mathbb{N}.$
+with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$
 \end{prop}

### 2020-02-19 aap/Revision/vr-mcmc-revision.tex
notations update

@@
 \begin{equation}\label{eq:chain}
-X_{p+1}=X_{p}-\gamma_{p+1}\mu(X_{p})+\sqrt{\gamma_{p+1}}Z_{p+1},\end{equation}
+X_{p+1}=X_{p}-\gamma_{p+1}\mu(X_{p})+\sqrt{2\gamma_{p+1}}Z_{p+1},\end{equation}
 where $\left(Z_{p}\right)_{p\geq1}$ is an i.i.d.\ sequence of $d$-dimensional
@@
 \begin{eqnarray*}
-\Phi_p(x,(u,z)^\top)=x+\mathbbm{1}\bigl(u\leq \alpha(x,x-\gamma_{p}\mu(x)+\sqrt{\gamma_{p}}z)\bigr)(-\gamma_{p}\mu(x)+\sqrt{\gamma_{p}}z).
+\Phi_p(x,(u,z)^\top)=x+\mathbbm{1}\bigl(u\leq \alpha(x,x-\gamma_{p}\mu(x)+\sqrt{2\gamma_{p}}z)\bigr)(-\gamma_{p}\mu(x)+\sqrt{2\gamma_{p}}z).
 \end{eqnarray*}

### 2020-02-19 aap/Revision/vr-mcmc-revision-new.tex
update

@@
 \newcommand{\denisi}[1]{\todo[inline,color=green!10, author=Denis]{#1}}
-\newcommand{\sergey}[1]{\todo[color=red!10, author=Sergey]{#1}}
+\newcommand{\sergey}[1]{\todo[color=blue!10, author=Sergey]{#1}}
 \newcommand{\sergeyi}[1]{\todo[inline,color=red!10, author=Sergey]{#1}}
@@
 $\phi_k(\xi)$, $k\ge1$, are centered.
-As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of Shifted Legendre polynomials for "uniform part" and Hermite polynomials for "Gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the Shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\) In the sequel, we denote by $(\mathcal{G}_p)_{p\in \mathbb{Z}_+}$  the filtration generated by independent random variables $(\xi_p)_{p \in \mathbb{N}^*}$ with the convention $\mathcal{G}_0=\mathrm{triv}$.
+As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of Shifted Legendre polynomials for "uniform part" and Hermite polynomials for "Gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the Shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\) We shall use the notations $\nset=\{1,2,\ldots\}$ and $\nset_0=\mathbb N\cup\{0\}$.
+We denote by $(\mathcal{G}_p)_{p\in \mathbb{Z}_+}$  the filtration generated by independent random variables $(\xi_p)_{p \in \mathbb{N}^*}$ with the convention $\mathcal{G}_0=\mathrm{triv}$.
 For $l \le p$ and $x \in \rset^d$, put for $x,y \in \rset^d$ functions $\Phi_k(x,y)=x-\gamma_k\mu(x)+y$.
@@
 \nonumber
-M_{K,n}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=1}^{p} \bar a_{p-l,k}(X_{l-1})\phi_k(\xi_{l})\right]
+M_{K,n}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=N+1}^{p} \bar a_{p-l,k}(X_{l-1})\phi_k(\xi_{l})\right]
 \\
@@
 \end{equation}
-Since \(X_{l-1}\) is independent of \(\xi_{l}\) and \tcr{\(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 1,\)} \eric{I do not understand this}  the r.v.
-\(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
+Since \(X_{l-1}\) is independent of \(\xi_{l}\) and \tcr{\(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0\)}, we obtain 
+\[\mathsf{E}[A_{N+n-l,k}(X_{l-1})\phi_{k}\left(\xi_{l}\right)] = \mathsf{E}[A_{N+n-l,k}(X_{l-1})\mathsf{E}[\left.\phi_{k}\left(\xi_{l}\right)\right|\mathcal{G}_{l-1}]] = 0\]
+\eric{I do not understand this} and hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
 The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
@@
 In this section we perform the convergence analysis of the ULA algorithm.
-For the sake of clarity and notational simplicity we restrict our attention to the constant time step, that is, we take $\gamma=\gamma_1=\gamma_2=\ldots.$
-\tcr{We shall use the notations $\nset=\{1,2,\ldots\}$ and $\nset_0=\mathbb N\cup\{0\}$.} \eric{these notations are used before; it is best to put all the notations in a short paragraph just after the introduction}
+For the sake of clarity and notational simplicity we restrict our attention to the constant time step, that is, we take $\gamma_k =\gamma$ for any $k \in \nset$.
+
+\sergey{N and N0 were defined here before}
+\eric{these notations are used before; it is best to put all the notations in a short paragraph just after the introduction}
+
 By $H_k$, $k\in\mathbb N_0$,
@@
 \[
-\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\} \rmd y.
-\]
+\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(2 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{2\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\} \rmd y.
+\]
+\sergey{Changed notation here, put variance of the noise equal to $\gamma$}
 \eric{what is the notation convention here; I do not like much $\mathcal{Q}$ but $Q$ is already define and it is perhaps not a great idea to supersede the previous notation}
@@
 \end{eqnarray}
-where \tcr{\(\lesssim\) stands for inequality up to a constant not depending  on $\gamma,$ \(n\) and \(N.\)}
+%where \tcr{\(\lesssim\) stands for inequality up to a constant not depending  on $\gamma,$ \(n\) and \(N.\)}
+\sergey{removed repeating explanation of $\lesssim$}
 \eric{this is defined less precisely just above}

### 2020-02-19 aap/Revision/vr-mcmc-revision-new.tex
update

@@
 \label{eq:mart_repr}
-f(X^x_{p})=\mathsf{E}\left[\left.f(X^x_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}a_{p,l,k}(X^x_{l-1})\phi_k\left(\xi_{l}\right),
+f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}a_{p,l,k}(X_{l-1})\phi_k\left(\xi_{l}\right),
 \end{eqnarray}
@@
 \[
-f(X^x_{1})=\PE[f(X_{1}^x)]+\sum_{k\geq1}a_{1,1,k}(x)\phi_{k}(\xi_{1})
+f(X_{1})=\PE[f(X_{1})]+\sum_{k\geq1}a_{1,1,k}(x)\phi_{k}(\xi_{1})
 \]
 with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$,
-provided $\PE[|f(X^x_{1})|^{2}]<\infty.$
+provided $\PE[|f(X_{1})|^{2}]<\infty.$
 \tcr{Recall that $\mathcal{G}_l=\sigma(\xi_{1},\ldots,\xi_{l}),$
@@
 for all $x$ and some family of functions $(W_s)_{s\geq 1}$ such that  the sum $\sum_{s=1}^n \mathsf{E}[W_s(X_{N+n-s-1})]$ is bounded uniformly in $n\in \mathbb{N}.$
+
+\section{Numerical results}
+\label{sec:num}
+If each function  \(Q_{p,l}(x) \) can be well approximated by polynomials, then we can use polynomial basis functions in   \eqref{eq:06042018a1} to approximate \(\bar Q_l\).
+Suppose that we constructed a polynomial approximation  for each function \(\bar Q_{l}\)  in the form:
+\begin{equation*}
+\widehat{Q}_{l}(x) = \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} x^{\mathbf{s}},\quad s=(s_1,\ldots,s_d)
+\end{equation*}
+with some coefficients \(\beta_{\mathbf{s}}\in \mathbb{R}.\)
+Then using the identity
+\[
+\xi^j = j! \sum_{r = 0}^{\lfloor j/2 \rfloor} \frac{1}{2^r  r! \sqrt{(j-2r)!}} H_{j-2r}(\xi),\quad \xi \in \mathbb{R},
+\]
+we derive for all $x \in \mathbb{R}^d$,
+\begin{eqnarray*}
+\widehat a_{l,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(x) \widehat{Q}_{l}(x - \gamma_l \mu(x) + \sqrt{\gamma_l}\xi)}{\TrainSet \vee \mathcal{G}_N}
+= \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{i,k_i,s_i}(x),
+\end{eqnarray*}
+where for all integers $i,k,s_i$ and $x \in \rset^d$,
+\begin{align*}
+P_{l,k,s}(x) &= \mathsf{E} \left[ H_{k}(\xi_l) (x_l - \gamma_l \mu_l(x) + \sqrt{\gamma_l} \xi_l)^{s} \right]
+\end{align*}
+is a one-dimensional polynomial (in \(x\)) of degree at most \(s\) with analytically known coefficients.
+Alternatively, on practice we can work directly with functions \(Q_{p,l}.\) Due to (almost) time-homogenous Markov chain generated by ULA, each function $Q_{p,l}$ can be approximated by function, which depends only on the  $p-l$, that is,
+\[
+ Q_{p,l}(x) = \mathsf{E} \left[f(X_p) | X_l = x\right] \approx Q^\circ_{p-l}(x)
+ \]
+Thus we have
+\[
+  Q^\circ_{r}(x) = \mathsf{E} \left[f(X_{l+r}) | X_l = x \right].
+ \]
+Consequently, the functions $Q^\circ_r$ can be estimated using a modified least squares criteria
+\begin{eqnarray}
+\label{eq:Qcircr}
+\widehat{Q}^\circ_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
+\end{eqnarray}
+for \(1 \leq r \leq n-1,\) where  $\widehat{Q}^\circ_0(x) = f(x)$ by definition.
+Due to Lemma~\ref{a_decay}, it is enough to estimate $Q^\circ_{r}$ for $|r| < n_{\mathrm{trunc}}$ for some truncation level \(n_{\mathrm{trunc}}\) depending on \(d\) and \(\gamma.\) It allows us to use a smaller amount of training trajectories to approximate conditional expectations $Q^\circ_{r}$.
+Finally  we construct a truncated version of our estimator:
+\[
+\pi_{K,n,n_{\mathrm{trunc}}}^N(f) = \pi_n^N(f) -  M_{K,n, n_{\mathrm{trunc}}}^{N}(f),
+\]
+where
+\begin{eqnarray*}
+{\widehat  M}_{K,n, n_{\mathrm{trunc}}}^N &=& \sum_{0<\left\| \mathbf{k}\right\| \leq K} \sum_{p=N+1}^{N+n} \omega_{p,n}^N
+\\
+&&\times
+\sum_{l=N+1}^p {\widehat  a}_{p-l,\mathbf{k}}(X_{l-1})\mathbf{H}_{\mathbf{k}}(Z_l) \mathbbm{1}\{ |p-l| < n_{\mathrm{trunc}} \}
+\end{eqnarray*}
+and
+\begin{eqnarray*}
+\widehat a_{p-l,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(x) \widehat{Q}^\circ_{p-l}(x - \gamma_l \mu(x) + \sqrt{\gamma_l}\xi)}{\TrainSet \vee \mathcal{G}_N}.
+\end{eqnarray*}
+\subsection{Gaussian mixtures}
+ We consider a sample generated by ULA with $\pi$  given by the mixture of two Gaussian distributions  with equal weights:
+\[
+\pi (x) = \frac{1}{2(2\pi)^{d/2}} \left( \rme^{\frac{-\| x-a\|_2^2}{2}}  + \rme^{\frac{-\| x+a\|_2^2}{2}} \right), \quad x \in\mathbb R^d
+\]
+where $a \in \mathbb{R}^d$ is a given vector. The function $U(x)$ and its gradient are given by
+\[
+U(x) = \frac{1}{2} \|x - a\|_2^2 - \text{log}(1 + \rme^{-2x^\top a}), \,\,
+\nabla U(x) = x-a +2a(1 + \rme^{2 x^\top a})^{-1},
+\]
+respectively.
+In our experiments we considered dimensions $d = 2$ and $d = 8$ and take $a=( (2d)^{-1/2}, \dots, (2d)^{-1/2})^\top$.
+In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i,\)
+ we have used constant step sizes $\gamma_i = \gamma=0.2$ and sampled $n = 10^4$ samples for $d = 2$ and $n = 2 \times 10^3$ for $d = 8$.
+We generated $\NtrainPath = 10$ independent "training" trajectories and solved the least squares  problems \eqref{eq:Qcircr} using the first order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section.
+The truncation level \(n_{\mathrm{trunc}}\) is chosen to be \(50.\) To test our variance reduction algorithm, we generated 100 independent trajectories.
+ In Figure ~\ref{fig:1} we compare our approach to variance reduction methods of \cite{mira2013zero} and \cite{brosse2018diffusion}.
+
 \section{Proofs}

### 2020-02-20 aap/Revision/vr-mcmc-revision-new.tex
update

@@
 \label{sec:num}
-If each function  \(Q_{p,l}(x) \) can be well approximated by polynomials, then we can use polynomial basis functions in   \eqref{eq:06042018a1} to approximate \(\bar Q_l\).
-Suppose that we constructed a polynomial approximation  for each function \(\bar Q_{l}\)  in the form:
+We will perform numerical experiments based on ULA with constant step size $\gamma$.
+Suppose that each function  \(Q_{r}(x)\) from \ref{eq:definition-bar-a} can be well approximated by polynomials, and we constructed a polynomial approximation  for each \(Q_r(x)\)  in the form:
 \begin{equation*}
-\widehat{Q}_{l}(x) = \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} x^{\mathbf{s}},\quad s=(s_1,\ldots,s_d)
+\widehat{Q}_{r}(x) = \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} x^{\mathbf{s}},\quad s=(s_1,\ldots,s_d)
 \end{equation*}
@@
 \]
-we derive for all $x \in \mathbb{R}^d$,
-\begin{eqnarray*}
-\widehat a_{l,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(x) \widehat{Q}_{l}(x - \gamma_l \mu(x) + \sqrt{\gamma_l}\xi)}{\TrainSet \vee \mathcal{G}_N}
+we obtain closed-form expression for the estimates $\widehat a_{r,k}(x)$ of functions $\bar a_{r,k}(x)$ from \ref{eq:definition-bar-a}. Namely, for all $x \in \mathbb{R}^d$, 
+\begin{eqnarray*}
+\widehat a_{r,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(x) \widehat{Q}_{r}(x - \gamma \mu(x) + \sqrt{\gamma}\xi)}{\TrainSet \vee \mathcal{G}_N}
 = \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{i,k_i,s_i}(x),
 \end{eqnarray*}
-where for all integers $i,k,s_i$ and $x \in \rset^d$,
-\begin{align*}
-P_{l,k,s}(x) &= \mathsf{E} \left[ H_{k}(\xi_l) (x_l - \gamma_l \mu_l(x) + \sqrt{\gamma_l} \xi_l)^{s} \right]
+where for all integers $i,k_i,s_i$ and $x \in \rset^d$,
+\begin{align*}
+P_{r,k,s}(x) &= \mathsf{E} \left[ H_{k}(\xi_l) (x - \gamma \mu(x) + \sqrt{\gamma} \xi)^{s} \right]
 \end{align*}
 is a one-dimensional polynomial (in \(x\)) of degree at most \(s\) with analytically known coefficients.
-Alternatively, on practice we can work directly with functions \(Q_{p,l}.\) Due to (almost) time-homogenous Markov chain generated by ULA, each function $Q_{p,l}$ can be approximated by function, which depends only on the  $p-l$, that is,
-\[
- Q_{p,l}(x) = \mathsf{E} \left[f(X_p) | X_l = x\right] \approx Q^\circ_{p-l}(x)
- \]
-Thus we have
-\[
-  Q^\circ_{r}(x) = \mathsf{E} \left[f(X_{l+r}) | X_l = x \right].
- \]
-Consequently, the functions $Q^\circ_r$ can be estimated using a modified least squares criteria
+In consequence, it is enough to estimate functions \(Q_{r}(x)\).
+It can be done using a modified least-squares criteria based on $T$ training trajectories \(\left(X_1^{(s)},\ldots,X_{N+n}^{(s)}\right), s = 1,\ldots,T\) with $N$ being the size of burn-in period:
 \begin{eqnarray}
 \label{eq:Qcircr}
-\widehat{Q}^\circ_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
+\widehat{Q}_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
 \end{eqnarray}
-for \(1 \leq r \leq n-1,\) where  $\widehat{Q}^\circ_0(x) = f(x)$ by definition.
-Due to Lemma~\ref{a_decay}, it is enough to estimate $Q^\circ_{r}$ for $|r| < n_{\mathrm{trunc}}$ for some truncation level \(n_{\mathrm{trunc}}\) depending on \(d\) and \(\gamma.\) It allows us to use a smaller amount of training trajectories to approximate conditional expectations $Q^\circ_{r}$.
-Finally  we construct a truncated version of our estimator:
-\[
-\pi_{K,n,n_{\mathrm{trunc}}}^N(f) = \pi_n^N(f) -  M_{K,n, n_{\mathrm{trunc}}}^{N}(f),
+for \(1 \leq r \leq n-1,\), $\Psi$ being the class of polynomials $\Psi = \{\psi(x) | \psi(x) = \sum_{\|\mathbf{s}\|\leq m} \alpha_{\mathbf{s}} x^{\mathbf{s}} \}$ and $\widehat{Q}_0(x) = f(x)$ by definition.
+Due to Lemma~\ref{a_decay},
+\sergey{we need to put some lemma about the truncation point here, I will do it}
+it is enough to estimate $Q_{r}$ for $r < n_{\mathrm{trunc}}$ for some truncation level \(n_{\mathrm{trunc}}\) depending on \(d\) and \(\gamma.\) It allows us to use a smaller amount of training trajectories to approximate $Q_{r}$.
+
+Finally  we construct a truncated version of the estimator from \ref{}:
+\[
+\widehat\pi_{K,n,n_{\mathrm{trunc}}}^N(f) = \widehat\pi_n^N(f) -  \widehat M_{K,n, n_{\mathrm{trunc}}}^{N}(f),
 \]
@@
 \begin{eqnarray*}
-{\widehat  M}_{K,n, n_{\mathrm{trunc}}}^N &=& \sum_{0<\left\| \mathbf{k}\right\| \leq K} \sum_{p=N+1}^{N+n} \omega_{p,n}^N
-\\
-&&\times
-\sum_{l=N+1}^p {\widehat  a}_{p-l,\mathbf{k}}(X_{l-1})\mathbf{H}_{\mathbf{k}}(Z_l) \mathbbm{1}\{ |p-l| < n_{\mathrm{trunc}} \}
+\widehat M_{K,n,n_{\mathrm{trunc}}}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=N+1}^{p} \bar a_{p-l,k}(X_{l-1})\phi_k(\xi_{l})\right]
 \end{eqnarray*}
@@
 \end{eqnarray*}
+
 \subsection{Gaussian mixtures}

### 2020-02-20 aap/Revision/vr-mcmc-revision-new.tex
update

@@
 \end{eqnarray}
-for \(1 \leq r \leq n-1,\), $\Psi$ being the class of polynomials $\Psi = \{\psi(x) | \psi(x) = \sum_{\|\mathbf{s}\|\leq m} \alpha_{\mathbf{s}} x^{\mathbf{s}} \}$ and $\widehat{Q}_0(x) = f(x)$ by definition.
+for \(1 \leq r \leq n-1,\) $\Psi$ being the class of polynomials $\Psi = \{\psi(x) | \psi(x) = \sum_{\|\mathbf{s}\|\leq m} \alpha_{\mathbf{s}} x^{\mathbf{s}} \}$ and $\widehat{Q}_0(x) = f(x)$ by definition.
 Due to Lemma~\ref{a_decay},
 \sergey{we need to put some lemma about the truncation point here, I will do it}
-it is enough to estimate $Q_{r}$ for $r < n_{\mathrm{trunc}}$ for some truncation level \(n_{\mathrm{trunc}}\) depending on \(d\) and \(\gamma.\) It allows us to use a smaller amount of training trajectories to approximate $Q_{r}$.
+it is enough to estimate $Q_{r}$ for $r < n_{\mathrm{trunc}}$ for some truncation level \(n_{\mathrm{trunc}}\) depending on \(d\) and \(\gamma.\) It allows us to use a smaller amount of training trajectories to approximate $Q_{r}(x)$.
 
@@
 \[
-\widehat\pi_{K,n,n_{\mathrm{trunc}}}^N(f) = \widehat\pi_n^N(f) -  \widehat M_{K,n, n_{\mathrm{trunc}}}^{N}(f),
+\pi_{K,n,n_{\mathrm{trunc}}}^N(f) = \pi_n^N(f) -  \widehat M_{K,n, n_{\mathrm{trunc}}}^{N}(f),
 \]
@@
 \begin{eqnarray*}
-\widehat M_{K,n,n_{\mathrm{trunc}}}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=N+1}^{p} \bar a_{p-l,k}(X_{l-1})\phi_k(\xi_{l})\right]
-\end{eqnarray*}
-and
-\begin{eqnarray*}
-\widehat a_{p-l,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(x) \widehat{Q}^\circ_{p-l}(x - \gamma_l \mu(x) + \sqrt{\gamma_l}\xi)}{\TrainSet \vee \mathcal{G}_N}.
+\widehat M_{K,n,n_{\mathrm{trunc}}}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{0 < \|\mathbf{k}\| < K}\sum_{l=N+1}^{p} \widehat a_{p-l,\mathbf{k}}(X_{l-1})\phi_k(\xi_{l})\mathbbm{1}\{ | p-l | < n_{\mathrm{trunc}}\right]
 \end{eqnarray*}

### 2020-02-20 aap/Revision/vr-mcmc-revision-new.tex
update vr

@@
 \newcommand{\denisi}[1]{\todo[inline,color=green!10, author=Denis]{#1}}
-\newcommand{\sergey}[1]{\todo[color=blue!10, author=Sergey]{#1}}
+\newcommand{\sergey}[1]{\todo[color=red!10, author=Sergey]{#1}}
 \newcommand{\sergeyi}[1]{\todo[inline,color=red!10, author=Sergey]{#1}}
@@
 \begin{equation}\label{eq:chain}
-X_{p+1}=X_{p}-\gamma_{p+1}\mu(X_{p})+\sqrt{\gamma_{p+1}}Z_{p+1},\end{equation}
+X_{p+1}=X_{p}-\gamma_{p+1}\mu(X_{p})+\sqrt{2 \gamma_{p+1}}Z_{p+1},\end{equation}
 where $\left(Z_{p}\right)_{p\geq1}$ is an i.i.d.\ sequence of $d$-dimensional
@@
 \begin{equation}\label{eq:stationary_distr}
-\pi(x)= \mathrm{const}\,\rme^{-\frac{U(x)}{2}}, \quad \mathrm{const}= \left. 1 \middle /{\int_{\mathbb{R}^{d}} \rme^{-\frac{U(x)}{2}}\, \rmd x,} \right.
+\pi(x)= \mathrm{const}\,\rme^{-U(x)}, \quad \mathrm{const}= \left. 1 \middle /{\int_{\mathbb{R}^{d}} \rme^{-U(x)}\, \rmd x,} \right.
 \end{equation}
@@
 $\phi_k(\xi)$, $k\ge1$, are centered.
-As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of Shifted Legendre polynomials for "uniform part" and Hermite polynomials for "Gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the Shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\) We shall use the notations $\nset=\{1,2,\ldots\}$ and $\nset_0=\mathbb N\cup\{0\}$.
-We denote by $(\mathcal{G}_p)_{p\in \mathbb{Z}_+}$  the filtration generated by independent random variables $(\xi_p)_{p \in \mathbb{N}^*}$ with the convention $\mathcal{G}_0=\mathrm{triv}$.
-For $l \le p$ and $x \in \rset^d$, put for $x,y \in \rset^d$ functions $\Phi_k(x,y)=x-\gamma_k\mu(x)+y$.
-Then we can define
-\[
-X^x_{l,p}=G_{l,p}(x,\sqrt{\gamma_l}\xi_{l},\ldots,\sqrt{\gamma_p}\xi_{p}),
-\]
-with the function \(G_{l,p}:\) \(\mathbb{R}^{d\times(p-l+2)}\to \mathbb{R}^{d}\) is defined as
-\begin{equation}
-\label{eq:definition-G-p-l}
-G_{l,p}(x,y_l,\ldots,y_p):=\Phi_p(\cdot,y_{p})\circ\Phi_{p-1}(\cdot,y_{p-1})\circ\ldots\circ\Phi_{l}(x,y_{l})
-\end{equation}
-For simplicity we will write $X^x_{p}$ and $G_{p}$ instead of $X^x_{0,p}$ and $G_{0,p}$. As a consequence, for a function $f\colon \rset^d \to \rset$ we have
-$$
-f\left(X^x_{p}\right) =%f\circ G_{p}(x,\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p}).
-f\circ G_{p}(x,\sqrt{\gamma}\xi_{1},\ldots,\sqrt{\gamma}\xi_{p})
-$$
-%\erici{It would be perhaps interesting to state the conditions satisfied by the functions $(\Phi_l)_{l \geq 1}$ here}
+As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of Shifted Legendre polynomials for "uniform part" and Hermite polynomials for "Gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the Shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\) In the sequel, we denote by $(\mathcal{G}_p)_{p\in \mathbb{Z}_+}$  the filtration
+generated by generated by $(\xi_p)_{p \in \mathbb{N}^*}$ with the convention $\mathcal{G}_0=\mathrm{triv}$.
+\erici{It would be perhaps interesting to state the conditions satisfied by the functions $(\Phi_l)_{l \geq 1}$ here}
 \begin{thm}\label{prop:29032018a1}
- Let $f$ be a Borel function $\rset^{d}\to\mathbb R$ such that
-$\mathsf{E}\left[\left |f(X_{p})\right|^{2}\right]<\infty$,   where $(X_{p})_{p\geq 0}$ is a Markov chain of the form \eqref{eq:chain_gen} for some sequence of functions $(\Phi_l)_{l\geq 1}.$
+ Let $f$ be a Borel function $\mathbb{R}^{d}\to\mathbb R$ such that
+$\mathsf{E}\left[\left|f(X_{p})\right|^{2}\right]<\infty$,   where $(X_{p})_{p\geq 0}$ is a Markov chain of the form \eqref{eq:chain_gen} for some sequence of functions $(\Phi_l)_{l\geq 1}.$
 Then,
@@
 \label{eq:coeff_mart}
-%a_{p,l,k}(x)=\mathsf{E}\left[\left.f(X_{p})\phi_k\left(\xi_{l}\right)\right|X_{l-1}=x\right], \quad p\geq l, \quad k\in \mathbb{N}.
-a_{p,l,k}(x)=\mathsf{E}\left[f(X^x_{l-1,p})\phi_k\left(\xi_{l}\right)\right], \quad p\geq l, \quad k\in \mathbb{N}
+a_{p,l,k}(x)=\mathsf{E}\left[\left.f(X_{p})\phi_k\left(\xi_{l}\right)\right|X_{l-1}=x\right], \quad p\geq l, \quad k\in \mathbb{N}.
 \end{eqnarray}
 \end{thm}
-
-\begin{proof}%[Proof of Theorem~\ref{prop:29032018a1}]
+\begin{cor}
+If  all the kernels $\Phi_l,$ $l\geq 1,$ are equal,  then  the representation \eqref{eq:mart_repr} takes the form
+\begin{eqnarray*}
+f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}\bar a_{p-l,k}(X_{l-1})\phi_k\left(\xi_{l}\right)
+\end{eqnarray*}
+with
+\begin{eqnarray*}
+\bar a_{r,k}(x)=\mathsf{E}\left[\left.f(X_{r})\phi_k\left(\xi_{1}\right)\right|X_{0}=x\right], \quad r, k\in \mathbb{N}.
+\end{eqnarray*}
+
+\end{cor}
+\begin{proof}[Proof of Theorem~\ref{prop:29032018a1}]
 The expansion obviously holds for $p=1$ and $j=0$.
-Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system in \(L^2(\mathbb{R}^d, P_{\xi})\), we have for all $x \in \rset^d$,
+Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system of , we have for all $x \in \rset^d$,
 \eric{it is clear, but we must say in which sense this equality should be understood}
 \[
-f(X_{1})=\PE[f(X_{1})]+\sum_{k\geq1}a_{1,1,k}(x)\phi_{k}(\xi_{1})
+f(X^x_{1})=\PE[f(X_{1}^x)]+\sum_{k\geq1}a_{1,1,k}(x)\phi_{k}(\xi_{1})
 \]
 with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$,
-provided $\PE[|f(X_{1})|^{2}]<\infty.$
+provided $\PE[|f(X^x_{1})|^{2}]<\infty.$
 \tcr{Recall that $\mathcal{G}_l=\sigma(\xi_{1},\ldots,\xi_{l}),$
@@
 \end{proof}
-
-\begin{cor}
-If  all the kernels $\Phi_l,$ $l\geq 1,$ are equal,  then  the representation \eqref{eq:mart_repr} takes the form
-\begin{eqnarray*}
-f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}\bar a_{p-l,k}(X_{l-1})\phi_k\left(\xi_{l}\right)
-\end{eqnarray*}
-with
-\begin{eqnarray*}
-\bar a_{r,k}(x)=\mathsf{E}\left[f(X^x_{r})\phi_k\left(\xi_{1}\right)\right], \quad r, k\in \mathbb{N}.
-\end{eqnarray*}
-\end{cor}
-
 From  numerical point of view another representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
@@
 \end{eqnarray*}
-with \(Q_{p,l}(x)=\mathsf{E}\left[f(X^x_{l,p})\right],\) \(p\geq l.\)
+with \(Q_{p,l}(x)=\mathsf{E}\left[\left.f(X_{p})\right|X_{l}=x\right],\) \(p\geq l.\)
 %where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
@@
 \label{eq:qpl}
-Q_{p,l}(x)=\mathsf{E}\left[Q_{p,l+1}(X^x_{l+1})\right].
+Q_{p,l}(x)=\mathsf{E}\left[\left.Q_{p,l+1}(X_{l+1})\right|X_{l}=x\right].
 %=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))]
@@
 \end{equation}
-with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$
+with  \(Q_{r}(x)=\mathsf{E}\left[\left.f(X_{r})\right|X_{0}=x\right],\) $r\in \mathbb{N}.$
 \end{prop}
@@
 \nonumber
-M_{K,n}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=N+1}^{p} \bar a_{p-l,k}(X_{l-1})\phi_k(\xi_{l})\right]
+M_{K,n}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=1}^{p} \bar a_{p-l,k}(X_{l-1})\phi_k(\xi_{l})\right]
 \\
@@
 \end{equation}
-Since \(X_{l-1}\) is independent of \(\xi_{l}\) and \tcr{\(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0\)}, we obtain 
-\[\mathsf{E}[A_{N+n-l,k}(X_{l-1})\phi_{k}\left(\xi_{l}\right)] = \mathsf{E}[A_{N+n-l,k}(X_{l-1})\mathsf{E}[\left.\phi_{k}\left(\xi_{l}\right)\right|\mathcal{G}_{l-1}]] = 0\]
-\eric{I do not understand this} and hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
+Since \(X_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0,\)   the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
 The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
@@
 In this section we perform the convergence analysis of the ULA algorithm.
-For the sake of clarity and notational simplicity we restrict our attention to the constant time step, that is, we take $\gamma_k =\gamma$ for any $k \in \nset$.
-
-\sergey{N and N0 were defined here before}
-\eric{these notations are used before; it is best to put all the notations in a short paragraph just after the introduction}
-
+For the sake of clarity and notational simplicity we restrict our attention to the constant time step, that is, we take $\gamma=\gamma_1=\gamma_2=\ldots.$
+\tcr{We shall use the notations $\nset=\{1,2,\ldots\}$ and $\nset_0=\mathbb N\cup\{0\}$.} \eric{these notations are used before; it is best to put all the notations in a short paragraph just after the introduction}
 By $H_k$, $k\in\mathbb N_0$,
@@
 \[
-\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(2 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{2\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\} \rmd y.
-\]
-\sergey{Changed notation here, put variance of the noise equal to $\gamma$}
+\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\} \rmd y.
+\]
 \eric{what is the notation convention here; I do not like much $\mathcal{Q}$ but $Q$ is already define and it is perhaps not a great idea to supersede the previous notation}
@@
 \end{eqnarray}
-%where \tcr{\(\lesssim\) stands for inequality up to a constant not depending  on $\gamma,$ \(n\) and \(N.\)}
-\sergey{removed repeating explanation of $\lesssim$}
+where \tcr{\(\lesssim\) stands for inequality up to a constant not depending  on $\gamma,$ \(n\) and \(N.\)}
 \eric{this is defined less precisely just above}
@@
 for all $x$ and some family of functions $(W_s)_{s\geq 1}$ such that  the sum $\sum_{s=1}^n \mathsf{E}[W_s(X_{N+n-s-1})]$ is bounded uniformly in $n\in \mathbb{N}.$
-
-\section{Numerical results}
-\label{sec:num}
-We will perform numerical experiments based on ULA with constant step size $\gamma$.
-Suppose that each function  \(Q_{r}(x)\) from \ref{eq:definition-bar-a} can be well approximated by polynomials, and we constructed a polynomial approximation  for each \(Q_r(x)\)  in the form:
-\begin{equation*}
-\widehat{Q}_{r}(x) = \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} x^{\mathbf{s}},\quad s=(s_1,\ldots,s_d)
-\end{equation*}
-with some coefficients \(\beta_{\mathbf{s}}\in \mathbb{R}.\)
-Then using the identity
-\[
-\xi^j = j! \sum_{r = 0}^{\lfloor j/2 \rfloor} \frac{1}{2^r  r! \sqrt{(j-2r)!}} H_{j-2r}(\xi),\quad \xi \in \mathbb{R},
-\]
-we obtain closed-form expression for the estimates $\widehat a_{r,k}(x)$ of functions $\bar a_{r,k}(x)$ from \ref{eq:definition-bar-a}. Namely, for all $x \in \mathbb{R}^d$, 
-\begin{eqnarray*}
-\widehat a_{r,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(x) \widehat{Q}_{r}(x - \gamma \mu(x) + \sqrt{\gamma}\xi)}{\TrainSet \vee \mathcal{G}_N}
-= \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{i,k_i,s_i}(x),
-\end{eqnarray*}
-where for all integers $i,k_i,s_i$ and $x \in \rset^d$,
-\begin{align*}
-P_{r,k,s}(x) &= \mathsf{E} \left[ H_{k}(\xi_l) (x - \gamma \mu(x) + \sqrt{\gamma} \xi)^{s} \right]
-\end{align*}
-is a one-dimensional polynomial (in \(x\)) of degree at most \(s\) with analytically known coefficients.
-In consequence, it is enough to estimate functions \(Q_{r}(x)\).
-It can be done using a modified least-squares criteria based on $T$ training trajectories \(\left(X_1^{(s)},\ldots,X_{N+n}^{(s)}\right), s = 1,\ldots,T\) with $N$ being the size of burn-in period:
-\begin{eqnarray}
-\label{eq:Qcircr}
-\widehat{Q}_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
-\end{eqnarray}
-for \(1 \leq r \leq n-1,\) $\Psi$ being the class of polynomials $\Psi = \{\psi(x) | \psi(x) = \sum_{\|\mathbf{s}\|\leq m} \alpha_{\mathbf{s}} x^{\mathbf{s}} \}$ and $\widehat{Q}_0(x) = f(x)$ by definition.
-Due to Lemma~\ref{a_decay},
-\sergey{we need to put some lemma about the truncation point here, I will do it}
-it is enough to estimate $Q_{r}$ for $r < n_{\mathrm{trunc}}$ for some truncation level \(n_{\mathrm{trunc}}\) depending on \(d\) and \(\gamma.\) It allows us to use a smaller amount of training trajectories to approximate $Q_{r}(x)$.
-
-Finally  we construct a truncated version of the estimator from \ref{}:
-\[
-\pi_{K,n,n_{\mathrm{trunc}}}^N(f) = \pi_n^N(f) -  \widehat M_{K,n, n_{\mathrm{trunc}}}^{N}(f),
-\]
-where
-\begin{eqnarray*}
-\widehat M_{K,n,n_{\mathrm{trunc}}}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{0 < \|\mathbf{k}\| < K}\sum_{l=N+1}^{p} \widehat a_{p-l,\mathbf{k}}(X_{l-1})\phi_k(\xi_{l})\mathbbm{1}\{ | p-l | < n_{\mathrm{trunc}}\right]
-\end{eqnarray*}
-
-\subsection{Gaussian mixtures}
- We consider a sample generated by ULA with $\pi$  given by the mixture of two Gaussian distributions  with equal weights:
-\[
-\pi (x) = \frac{1}{2(2\pi)^{d/2}} \left( \rme^{\frac{-\| x-a\|_2^2}{2}}  + \rme^{\frac{-\| x+a\|_2^2}{2}} \right), \quad x \in\mathbb R^d
-\]
-where $a \in \mathbb{R}^d$ is a given vector. The function $U(x)$ and its gradient are given by
-\[
-U(x) = \frac{1}{2} \|x - a\|_2^2 - \text{log}(1 + \rme^{-2x^\top a}), \,\,
-\nabla U(x) = x-a +2a(1 + \rme^{2 x^\top a})^{-1},
-\]
-respectively.
-In our experiments we considered dimensions $d = 2$ and $d = 8$ and take $a=( (2d)^{-1/2}, \dots, (2d)^{-1/2})^\top$.
-In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i,\)
- we have used constant step sizes $\gamma_i = \gamma=0.2$ and sampled $n = 10^4$ samples for $d = 2$ and $n = 2 \times 10^3$ for $d = 8$.
-We generated $\NtrainPath = 10$ independent "training" trajectories and solved the least squares  problems \eqref{eq:Qcircr} using the first order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section.
-The truncation level \(n_{\mathrm{trunc}}\) is chosen to be \(50.\) To test our variance reduction algorithm, we generated 100 independent trajectories.
- In Figure ~\ref{fig:1} we compare our approach to variance reduction methods of \cite{mira2013zero} and \cite{brosse2018diffusion}.
-
 \section{Proofs}

### 2020-02-20 aap/Revision/vr-mcmc-revision-new.tex
update vr-mcmc-revision-new.tex
theorem 1 update (not final)

--- 2020-02-20 aap/Revision/vr-mcmc-revision-new.tex
@@
 As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of Shifted Legendre polynomials for "uniform part" and Hermite polynomials for "Gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the Shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\) We shall use the notations $\nset=\{1,2,\ldots\}$ and $\nset_0=\mathbb N\cup\{0\}$.
-We denote by $(\mathcal{G}_p)_{p\in \mathbb{Z}_+}$  the filtration generated by independent random variables $(\xi_p)_{p \in \mathbb{N}^*}$ with the convention $\mathcal{G}_0=\mathrm{triv}$.
+We denote by $(\mathcal{G}_p)_{p\in \mathbb{Z}_+}$  the filtration generated by independent random variables $(\xi_p)_{p \in \nset}$ with the convention $\mathcal{G}_0=\mathrm{triv}$.
 For $l \le p$ and $x \in \rset^d$, put for $x,y \in \rset^d$ functions $\Phi_k(x,y)=x-\gamma_k\mu(x)+y$.
 Then we can define
-\[
+\begin{equation}
+\label{eq:ula_new}
 X^x_{l,p}=G_{l,p}(x,\sqrt{\gamma_l}\xi_{l},\ldots,\sqrt{\gamma_p}\xi_{p}),
-\]
-with the function \(G_{l,p}:\) \(\mathbb{R}^{d\times(p-l+2)}\to \mathbb{R}^{d}\) is defined as
+\end{equation}
+with the function \(G_{l,p}:\) \(\mathbb{R}^{d\times(p-l+2)}\to \mathbb{R}^{d}\) defined as
 \begin{equation}
@@
 \end{equation}
-For simplicity we will write $X^x_{p}$ and $G_{p}$ instead of $X^x_{0,p}$ and $G_{0,p}$. As a consequence, for a function $f\colon \rset^d \to \rset$ we have
+Note that $\left(X^x_{0,p}\right)_{p \in \nset_0}$ is a Markov chain of the form \eqref{eq:chain_gen}, starting at $X_0 = x$.
+For simplicity we write $X^x_{p}$ and $G_{p}$ instead of $X^x_{0,p}$ and $G_{0,p}$.
+As a consequence, for a function $f\colon \rset^d \to \rset$ we have
 $$
 f\left(X^x_{p}\right) =%f\circ G_{p}(x,\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p}).
-f\circ G_{p}(x,\sqrt{\gamma}\xi_{1},\ldots,\sqrt{\gamma}\xi_{p})
+f\circ G_{p}(x,\sqrt{\gamma_1}\xi_{1},\ldots,\sqrt{\gamma_p}\xi_{p})
 $$
@@
 \begin{thm}\label{prop:29032018a1}
- Let $f$ be a Borel function $\rset^{d}\to\mathbb R$ such that
-$\mathsf{E}\left[\left |f(X_{p})\right|^{2}\right]<\infty$,   where $(X_{p})_{p\geq 0}$ is a Markov chain of the form \eqref{eq:chain_gen} for some sequence of functions $(\Phi_l)_{l\geq 1}.$
-Then,
-for $p>j$, the following representation holds in \(L^2(P)\)
+Let \((X^x_{p})_{p\geq 0}\) be a Markov chain of the form \eqref{eq:ula_new}  and $f$ be a Borel function $\rset^{d}\to\mathbb R$ such that
+$\mathsf{E}\left[\left |f(X^x_{p})\right|^{2}\right]<\infty$. Then, for $p>j$, the following representation holds in \(L^2(P)\)
 \eric{I would write \tcr{$L^2(\mathsf{P})$} to be coherent with the notation $\mathsf{E}$ and we need to define what is the underlying probability space}
@@
 %a_{p,l,k}(x)=\mathsf{E}\left[\left.f(X_{p})\phi_k\left(\xi_{l}\right)\right|X_{l-1}=x\right], \quad p\geq l, \quad k\in \mathbb{N}.
-a_{p,l,k}(x)=\mathsf{E}\left[f(X^x_{l-1,p})\phi_k\left(\xi_{l}\right)\right], \quad p\geq l, \quad k\in \mathbb{N}
+a_{p,l,k}(y)=\mathsf{E}\left[f(X^y_{l-1,p})\phi_k\left(\xi_{l}\right)\right], \quad p\geq l, \quad k\in \mathbb{N}
 \end{eqnarray}
@@
 The expansion obviously holds for $p=1$ and $j=0$.
-Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system in \(L^2(\mathbb{R}^d, P_{\xi})\), we have for all $x \in \rset^d$,
+Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system in \(L^2(\mathbb{R}^d, P_{\xi})\), we have for arbitary $x \in \rset^d$,
 \eric{it is clear, but we must say in which sense this equality should be understood}
 \[
-f(X_{1})=\PE[f(X_{1})]+\sum_{k\geq1}a_{1,1,k}(x)\phi_{k}(\xi_{1})
+f(X^x_{1})=\PE[f(X^x_{1})]+\sum_{k\geq1}a_{1,1,k}(x)\phi_{k}(\xi_{1})
 \]
@@
 \begin{eqnarray}
-\label{eq:Qcircr}
+\label{eq:Qregr}
 \widehat{Q}_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
@@
 for \(1 \leq r \leq n-1,\) $\Psi$ being the class of polynomials $\Psi = \{\psi(x) | \psi(x) = \sum_{\|\mathbf{s}\|\leq m} \alpha_{\mathbf{s}} x^{\mathbf{s}} \}$ and $\widehat{Q}_0(x) = f(x)$ by definition.
+Certainly one may use another functional class $\Psi$ in $\label{eq:Qregr}$, but using polynomials allows to compute closed-form representation of $\widehat a_{r,\mathbf{k}} (x)$.
+\par
 Due to Lemma~\ref{a_decay},
@@
 
-Finally  we construct a truncated version of the estimator from \ref{}:
+Finally  we construct a truncated version of the estimator \ref{eq:29032018a3}:
 \[
@@
 In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i,\)
- we have used constant step sizes $\gamma_i = \gamma=0.2$ and sampled $n = 10^4$ samples for $d = 2$ and $n = 2 \times 10^3$ for $d = 8$.
+ we have used constant step size $\gamma=0.2$ and sampled $n = 10^4$ samples for $d = 2$ and $n = 2 \times 10^3$ for $d = 8$.
 We generated $\NtrainPath = 10$ independent "training" trajectories and solved the least squares  problems \eqref{eq:Qcircr} using the first order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section.

### 2020-02-20 aap/Revision/vr-mcmc-revision-new.tex
update

@@
 
+\paragraph{Notations. }\label{par:notations}
+We use the notations $\nset=\{1,2,\ldots\}$ and $\nset_0=\mathbb N\cup\{0\}$.
+We denote $\boldsymbol{\varphi}(z)=(2\pi)^{-d/2} \exp\{-|z|^2/2\}$, $z\in\mathbb R^d$ probability density function of the $d-$dimensional standard normal distribution.
+For $x \in \rset^d$ and $r>0$ let $B_r(x) = \{y \in \rset^d | \|y - x\| < r\}$ where $\|\cdot\|$ is a standard Euclidean norm.
+For the twice differentiable function $g: \rset^d \rightarrow \rset$ we denote by $D^2g$ its Hessian at point $x$.
+For $m\in\mathbb N$, a smooth function
+$h\colon\mathbb R^{d\times m}\to\mathbb R$
+with arguments being denoted
+$(y_1,\ldots,y_m)$, $y_i\in\mathbb R^d$, $i=1,\ldots,m$,
+a multi-index $\mathbf k=(k_i)\in\mathbb N_0^d$,
+and $j\in\{1,\ldots,m\}$,
+we use the notation $\partial^{\mathbf k}_{y_j} h$ for the multiple derivative of $h$
+with respect to the components of~$y_j$:
+\[
+\partial^{\mathbf k}_{j} h(y_1,\ldots,y_m)
+:=\partial^{k_d}_{y_j^d}
+\ldots
+\partial^{k_1}_{y_j^1}
+h(y_1,\ldots,y_m),
+\quad y_j=(y_j^1,\ldots,y_j^d).
+\]
+In the particular case $m=1$ we drop the subscript $y_1$ in that notation.
+
 \section{Setup}\label{sec:setup}
@@
 \label{eq:chain_gen}
-X_{p}=\Phi_{p}(X_{p-1},\xi_{p}),\quad p=1,2,\ldots ,\quad X_{0}=x_0,
+X_{p}=\Phi_{p}(X_{p-1},\xi_{p}),\quad p=1,2,\ldots ,\quad X_{0}=x,
 \end{equation}
@@
 \end{equation}
-provided that \(\int_{\mathbb{R}^{d}} \rme^{-U(x)}\,\rmd x\) is finite. This method is usually referred to as
+provided that \(\int_{\mathbb{R}^{d}} \rme^{-\frac{U(x)}{2}}\,\rmd x\) is finite. This method is usually referred to as
 Unadjusted Langevin Algorithm (ULA).
@@
 \begin{equation*}
-Lg=b^\top \nabla g+\frac{1}{2}\operatorname{Tr}(\sigma^\top D^2g\sigma)
+Lg=b^\top \nabla g+\frac{1}{2}\sigma^\top D^2g\sigma
 \end{equation*}
-\eric{do we need a trace here ? I think also that $D^2$ is not defined }
 for any \(g\in C^2(\mathbb{R}^d).\)
@@
 $\phi_k(\xi)$, $k\ge1$, are centered.
-As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of Shifted Legendre polynomials for "uniform part" and Hermite polynomials for "Gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the Shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\) We shall use the notations $\nset=\{1,2,\ldots\}$ and $\nset_0=\mathbb N\cup\{0\}$.
-We denote by $(\mathcal{G}_p)_{p\in \mathbb{Z}_+}$  the filtration generated by independent random variables $(\xi_p)_{p \in \nset}$ with the convention $\mathcal{G}_0=\mathrm{triv}$.
-For $l \le p$ and $x \in \rset^d$, put for $x,y \in \rset^d$ functions $\Phi_k(x,y)=x-\gamma_k\mu(x)+y$.
-Then we can define
+As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of Shifted Legendre polynomials for "uniform part" and Hermite polynomials for "Gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the Shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\)  Let $\left(\Omega,\mathcal{F},\mathsf{P}\right)$ be a probability space, $(\xi_p)_{p \in \nset}$ be i.i.d.
+$d_{\xi}-$dimensional random vectors.
+We denote by $(\mathcal{G}_p)_{p \in \nset_0}$  the filtration generated by $(\xi_p)_{p \in \nset}$ with the convention $\mathcal{G}_0=\mathrm{triv}$.
+For  $x \in \rset^d$ and $y \in \rset^{d_\xi}$ let $\Phi_k(x,y)$ be a function mapping $\rset^{d+d_{\xi}}$ to $\rset^d$.
+Then we can define for $l \le p$
 \begin{equation}
 \label{eq:ula_new}
-X^x_{l,p}=G_{l,p}(x,\sqrt{\gamma_l}\xi_{l},\ldots,\sqrt{\gamma_p}\xi_{p}),
-\end{equation}
-with the function \(G_{l,p}:\) \(\mathbb{R}^{d\times(p-l+2)}\to \mathbb{R}^{d}\) defined as
+X^x_{l,p}=G_{l,p}(x,\xi_{l},\ldots,\xi_{p}),
+\end{equation}
+with the function \(G_{l,p}:\) \(\rset^{d + d_{\xi}\times(p-l+1)}\to \rset^{d}\) defined as
 \begin{equation}
 \label{eq:definition-G-p-l}
-G_{l,p}(x,y_l,\ldots,y_p):=\Phi_p(\cdot,y_{p})\circ\Phi_{p-1}(\cdot,y_{p-1})\circ\ldots\circ\Phi_{l}(x,y_{l})
-\end{equation}
-Note that $\left(X^x_{0,p}\right)_{p \in \nset_0}$ is a Markov chain of the form \eqref{eq:chain_gen}, starting at $X_0 = x$.
-For simplicity we write $X^x_{p}$ and $G_{p}$ instead of $X^x_{0,p}$ and $G_{0,p}$.
-As a consequence, for a function $f\colon \rset^d \to \rset$ we have
-$$
-f\left(X^x_{p}\right) =%f\circ G_{p}(x,\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p}).
-f\circ G_{p}(x,\sqrt{\gamma_1}\xi_{1},\ldots,\sqrt{\gamma_p}\xi_{p})
-$$
-%\erici{It would be perhaps interesting to state the conditions satisfied by the functions $(\Phi_l)_{l \geq 1}$ here}
+G_{l,p}(x,y_l,\ldots,y_p):=\Phi_p(\cdot,y_{p})\circ\Phi_{p-1}(\cdot,y_{p-1})\circ\dots\circ\Phi_{l}(x,y_{l})
+\end{equation}
+Note that $\left(X^x_{0,p}\right)_{p \in \nset_0}$ is a Markov chain with values in $\rset^d$ of the form \eqref{eq:chain_gen}, starting at $X_0 = x$.
+We write $X^x_{p}$ and $G_{p}$ as a shorthand notation for $X^x_{0,p}$ and $G_{0,p}$, respectively.
+
+%For a function $f\colon \rset^d \to \rset$ we define
+%$$
+%f\left(X^x_{p}\right) =%f\circ G_{p}(x,\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p}).
+%f\circ G_{p}(x,\xi_{1},\ldots,\xi_{p})
+%$$
 \begin{thm}\label{prop:29032018a1}
-Let \((X^x_{p})_{p\geq 0}\) be a Markov chain of the form \eqref{eq:ula_new}  and $f$ be a Borel function $\rset^{d}\to\mathbb R$ such that
-$\mathsf{E}\left[\left |f(X^x_{p})\right|^{2}\right]<\infty$. Then, for $p>j$, the following representation holds in \(L^2(P)\)
-\eric{I would write \tcr{$L^2(\mathsf{P})$} to be coherent with the notation $\mathsf{E}$ and we need to define what is the underlying probability space}
+Let \((X^x_{p})_{p\geq 0}\) be a Markov chain of the form \eqref{eq:ula_new}.
+Then, for $p>j$ and for all Borel functions $f: \rset^{d}\to \rset$ such that $\mathsf{E}\left[\left |f(X^x_{p})\right|^{2}\right]<\infty$, the following representation holds in \(L^2(\mathsf{P})\)
 \begin{eqnarray}
@@
 \end{eqnarray}
-where
+where for $y \in \rset^d$
 \begin{eqnarray}
@@
 The expansion obviously holds for $p=1$ and $j=0$.
-Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system in \(L^2(\mathbb{R}^d, P_{\xi})\), we have for arbitary $x \in \rset^d$,
-\eric{it is clear, but we must say in which sense this equality should be understood}
+Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system in \(L^2(\mathbb{R}^d, P_{\xi})\), it holds in \(L^2(\mathsf{P})\) that
 \[
@@
 \]
-with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$,
-provided $\PE[|f(X_{1})|^{2}]<\infty.$
-\tcr{Recall that $\mathcal{G}_l=\sigma(\xi_{1},\ldots,\xi_{l}),$
-$l=1,2,\ldots, $
-and $\mathcal{G}_0=\mathrm{triv}$.}
-\eric{do we need this, this is recalled ten lines above}
+with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$.
 Suppose that (\ref{eq:mart_repr})
 holds for $p=q$, all $j<q$, and all Borel-measurable functions
-$f$ with $\mathsf{E}\left[|f(X_{q})|^2\right]<\infty$.
+$f$ with $\mathsf{E}\left[|f(X^x_{q})|^2\right]<\infty$.
 Let us prove it for $p=q+1$.
-Given $f$ with $\mathsf{E}\left[|f(X_{p})|^2\right]<\infty$,
+Given $f$ with $\mathsf{E}\left[|f(X^x_{p})|^2\right]<\infty$,
 due to the orthonormality and completeness
@@
 \[
-f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|{\mathcal{G}}_q\right]+\sum_{k\geq1}\alpha_{p,q+1,k}\phi_{k}(\xi_{q+1}),
+f(X^x_{p})=\mathsf{E}\left[\left.f(X^x_{p})\right|{\mathcal{G}}_q\right]+\sum_{k\geq1}\alpha_{p,q+1,k}\phi_{k}(\xi_{q+1}),
 \]
@@
 \alpha_{p,q+1,k}
-=\mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{q+1})\right|\mathcal G_q\right].
+=\mathsf{E}\left[\left.f(X^x_{p})\phi_{k}(\xi_{q+1})\right|\mathcal G_q\right].
 \end{equation*}
-By the Markov property of $(X_{l})$,
+By the Markov property of $(X^x_{l})$,
 we have
-$\mathsf{E}[f(X_{p})|\mathcal{G}_q]
-=\mathsf{E}[f(X_{p})|X_{q}]$.
+$\mathsf{E}[f(X^x_{p})|\mathcal{G}_q]
+=\mathsf{E}[f(X^x_{p})|X^x_{q}]$.
 Furthermore, a calculation involving
@@
 \begin{eqnarray*}
-f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}\bar a_{p-l,k}(X_{l-1})\phi_k\left(\xi_{l}\right)
+f(X^x_{p})=\mathsf{E}\left[\left.f(X^x_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}\bar a_{p-l,k}(X^x_{l-1})\phi_k\left(\xi_{l}\right)
 \end{eqnarray*}
@@
 \begin{eqnarray*}
-\bar a_{r,k}(x)=\mathsf{E}\left[f(X^x_{r})\phi_k\left(\xi_{1}\right)\right], \quad r, k\in \mathbb{N}.
+\bar a_{r,k}(y)=\mathsf{E}\left[f(X^y_{r})\phi_k\left(\xi_{1}\right)\right], \quad r, k\in \mathbb{N}.
 \end{eqnarray*}
@@
 Since \(X_{l-1}\) is independent of \(\xi_{l}\) and \tcr{\(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0\)}, we obtain
-\[\mathsf{E}[A_{N+n-l,k}(X_{l-1})\phi_{k}\left(\xi_{l}\right)] = \mathsf{E}[A_{N+n-l,k}(X_{l-1})\mathsf{E}[\left.\phi_{k}\left(\xi_{l}\right)\right|\mathcal{G}_{l-1}]] = 0\]
-\eric{I do not understand this} and hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
+\[
+\PE[A_{N+n-l,k}(X_{l-1})\phi_{k}\left(\xi_{l}\right)] = \mathsf{E}[A_{N+n-l,k}(X_{l-1})\CPE{\phi_{k}\left(\xi_{l}\right)}{\mathcal{G}_{l-1}]} = 0
+\]
+and hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
 The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
@@
 \sergey{Changed notation here, put variance of the noise equal to $\gamma$}
-\eric{what is the notation convention here; I do not like much $\mathcal{Q}$ but $Q$ is already define and it is perhaps not a great idea to supersede the previous notation}
+%\eric{what is the notation convention here; I do not like much $\mathcal{Q}$ but $Q$ is already define and it is perhaps not a great idea to supersede the previous notation}
 For a bounded Borel function~$f$,
@@
 One of the main results of this paper is the following  upper bound for the variance of $\pi_{K,n}^{N}(f).$
-We start by introducing some notations.
-For $m\in\mathbb N$, a smooth function
-$h\colon\mathbb R^{d\times m}\to\mathbb R$
-with arguments being denoted
-$(y_1,\ldots,y_m)$, $y_i\in\mathbb R^d$, $i=1,\ldots,m$,
-a multi-index $\mathbf k=(k_i)\in\mathbb N_0^d$,
-and $j\in\{1,\ldots,m\}$,
-we use the notation $\partial^{\mathbf k}_{y_j} h$ for the multiple derivative of $h$
-with respect to the components of~$y_j$:
-\[
-\partial^{\mathbf k}_{j} h(y_1,\ldots,y_m)
-:=\partial^{k_d}_{y_j^d}
-\ldots
-\partial^{k_1}_{y_j^1}
-h(y_1,\ldots,y_m),
-\quad y_j=(y_j^1,\ldots,y_j^d).
-\]
-In the particular case $m=1$ we can drop
-the subscript $y_1$ in that notation.
+
 \begin{thm}\label{th:mr}
@@
 \sergey{removed repeating explanation of $\lesssim$}
-\eric{this is defined less precisely just above}
+%\eric{this is defined less precisely just above}
 \end{thm}

### 2020-02-20 aap/Revision/vr-mcmc-revision-new.tex
theorem 1: update

@@
 Let us prove it for $p=q+1$.
-Given $f$ with $\mathsf{E}\left[|f(X^x_{p})|^2\right]<\infty$,
+Given $f$ with $\mathsf{E}\left[|f(X^x_{q+1})|^2\right]<\infty$,
 due to the orthonormality and completeness
@@
 \[
-f(X^x_{p})=\mathsf{E}\left[\left.f(X^x_{p})\right|{\mathcal{G}}_q\right]+\sum_{k\geq1}\alpha_{p,q+1,k}\phi_{k}(\xi_{q+1}),
+f(X^x_{q+1})=\mathsf{E}\left[\left.f(X^x_{q+1})\right|{\mathcal{G}}_q\right]+\sum_{k\geq1}\alpha_{q+1,q+1,k}\phi_{k}(\xi_{q+1}),
 \]
@@
 \begin{equation*}
-\alpha_{p,q+1,k}
-=\mathsf{E}\left[\left.f(X^x_{p})\phi_{k}(\xi_{q+1})\right|\mathcal G_q\right].
+\alpha_{q+1,q+1,k}
+=\mathsf{E}\left[\left.f(X^x_{q+1})\phi_{k}(\xi_{q+1})\right|\mathcal G_q\right].
 \end{equation*}
-By the Markov property of $(X^x_{l})$,
-we have
-$\mathsf{E}[f(X^x_{p})|\mathcal{G}_q]
-=\mathsf{E}[f(X^x_{p})|X^x_{q}]$.
-Furthermore, a calculation involving
-intermediate conditioning on $\mathcal G_{q+1}$
-and the recurrence relation
-$X_{q+1}=\Phi_{q+1}(X_{q},\xi_{q+1})$
-verifies that
-\[
-\alpha_{p,q+1,k}=\mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{q+1})\right|X_{q}\right]
-=a_{p,q+1,k}(X_{q})
-\]
-for suitably chosen
-Borel-measurable functions $a_{p,q+1,k}$.
+By the Markov property of $(X^x_{l})$, it holds $\mathsf{E}[f(X^x_{q+1})|\mathcal{G}_q] = \mathsf{E}[f(X^x_{q+1})| X_q^x]$.
+Using Markov property together with conditioning on $\mathcal{G}_{q+1}$, we obtain
+\[
+\alpha_{q+1,q+1,k}=\mathsf{E}\left[\left.f(X^x_{q+1})\phi_{k}(\xi_{q+1})\right|X^x_{q}\right]
+=a_{q+1,q+1,k}(X^x_{q})
+\]
+where for $y \in \rset^d$
+\[
+a_{q+1,q+1,k}(y) = \mathsf{E}\left[f(X^y_{q,q+1})\phi_k\left(\xi_{q+1}\right)\right]
+\]
+
 We thus arrive at
@@
 \label{berm:sig_X}
-f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|X_{q}\right]+\sum_{k\geq1}a_{p,q+1,k}(X_{q})\phi_{k}(\xi_{q+1}),
+f(X^x_{q+1})=\mathsf{E}\left[\left.f(X^x_{q+1})\right|X^x_{q}\right]+\sum_{k\geq1}a_{q+1,q+1,k}(X^x_{q})\phi_{k}(\xi_{q+1}),
 \end{align}
@@
 The random variable
-$\mathsf{E}\left[\left.f(X_{p})\right|X_{q}\right]$
+$\mathsf{E}\left[\left.f(X^x_{q+1})\right|X^x_{q}\right]$
 is square integrable and has the form
-$g(X_{q})$,
+$g(X^x_{q})$,
 hence the induction hypothesis applies, and we get
 \begin{equation}\label{eq:28082017a1}
-\mathsf{E}\left[\left.f(X_{p})\right|X_{q}\right]=\mathsf{E}\left[\left.f(X_{p})\right|X_{j}\right]+\sum_{k\geq1}\sum_{l=j+1}^{q}a_{p,l,k}(X_{l-1})\phi_{k}(\xi_{l})
+\mathsf{E}\left[\left.f(X^x_{q+1})\right|X^x_{q}\right]=\mathsf{E}\left[\left.f(X^x_{q+1})\right|X^x_{j}\right]+\sum_{k\geq1}\sum_{l=j+1}^{q}a_{q+1,l,k}(X^x_{l-1})\phi_{k}(\xi_{l})
 \end{equation}
@@
 \begin{align*}
-a_{p,l,k}(X_{l-1}) &= \mathsf{E}\left[\left.\mathsf{E}\left[\left.f(X_{p})\right|\mathcal{G}_{q}\right]\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]
-= \mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]\\
-&=\mathsf{E}\left[\left.f(X_{p})\phi_{k}(\xi_{l})\right|X_{l-1}\right].
-\end{align*}
+a_{q+1,l,k}(X^x_{l-1}) &= \mathsf{E}\left[\left.\mathsf{E}\left[\left.f(X^x_{q+1})\right|\mathcal{G}_{q}\right]\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]
+ = \mathsf{E}\left[\left.f(X^x_{q+1})\phi_{k}(\xi_{l})\right|X^x_{l-1}\right].
+\end{align*}
+where for $y \in \rset^d$,
+\[
+a_{q+1,l,k}(y) = \mathsf{E}\left[f(X^y_{l-1,q+1})\phi_{k}(\xi_{l})\right]
+\]
 Formulas \eqref{berm:sig_X}
-and~\eqref{eq:28082017a1} conclude the proof.
+and~\eqref{eq:28082017a1} conclude the induction step for $p = q+1$ and hence the proof.
 \end{proof}
@@
 \end{eqnarray*}
-with
+where for $y \in \rset^d$
 \begin{eqnarray*}
@@
 \begin{equation}\label{eq:29032018a2}
-\pi_{n}^{N}(f)=\frac{1}{n}\sum_{p=N+1}^{N+n} f(X_{p}),
+\pi_{n}^{N}(f)=\frac{1}{n}\sum_{p=N+1}^{N+n} f(X^x_{p}),
 \end{equation}

### 2020-02-20 aap/Revision/vr-mcmc-revision-new.tex
1st theorem proof update

@@
 
-%For a function $f\colon \rset^d \to \rset$ we define
-%$$
-%f\left(X^x_{p}\right) =%f\circ G_{p}(x,\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p}).
-%f\circ G_{p}(x,\xi_{1},\ldots,\xi_{p})
-%$$
 \begin{thm}\label{prop:29032018a1}
@@
 with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$.
-Suppose that (\ref{eq:mart_repr})
-holds for $p=q$, all $j<q$, and all Borel-measurable functions
+Assume now that (\ref{eq:mart_repr})
+holds for $p=q$, all $j<q$, and Borel-measurable functions
 $f$ with $\mathsf{E}\left[|f(X^x_{q})|^2\right]<\infty$.
-Let us prove it for $p=q+1$.
+Let us prove that the induction assumption holds for $p=q+1$.
 Given $f$ with $\mathsf{E}\left[|f(X^x_{q+1})|^2\right]<\infty$,
 due to the orthonormality and completeness
-of the system $\left(\phi_{k}\right)$, we get by
-conditioning on $\mathcal{G}_{q}$,
-\[
-f(X^x_{q+1})=\mathsf{E}\left[\left.f(X^x_{q+1})\right|{\mathcal{G}}_q\right]+\sum_{k\geq1}\alpha_{q+1,q+1,k}\phi_{k}(\xi_{q+1}),
-\]
+of the system $\left(\phi_{k}\right)$, we get that for any $y \in \rset^d$ it holds in \(L^2(\mathsf{P})\) 
+\[
+f(X^y_{q,q+1}) = \mathsf{E} f(X^y_{q,q+1}) + \sum\limits_{k=1}^{\infty}\mathsf{E}[f(X^y_{q,q+1})\phi_{k}(\xi_{q+1})]\phi_k(\xi_{q+1})
+\]
+which means that
+\begin{equation}
+\label{eq:parseval}
+\lim_{n \rightarrow \infty}\mathsf{E}[|f(X^y_{q,q+1}) - \mathsf{E} f(X^y_{q,q+1}) - \sum\limits_{k=1}^{n}\mathsf{E}[f(X^y_{q,q+1})\phi_{k}(\xi_{q+1})]\phi_k(\xi_{q+1})|^2] = 0
+\end{equation}
+Let us define sequence $\psi_{q+1,n}(y) = \mathsf{E}[|f(X^y_{q,q+1}) - \mathsf{E} f(X^y_{q,q+1})|^2] - \sum\limits_{k=1}^{n}\mathsf{E}^2[f(X^y_{q,q+1})\phi_k(\xi_{q+1})]$.
+Note that $\psi_{q+1,n}(y)$ coincides with the left side of \ref{eq:parseval}, and due to Parseval identity $\Psi_{q+1,n}(y) \rightarrow 0, n \rightarrow \infty$ for any $y \in \rset^d$.
+Note that $\psi_n(y) < \mathsf{E}[|f(X^y_{q,q+1}) - \mathsf{E} f(X^y_{q,q+1})|^2] \leq \frac{\|f\|^2_{\infty}}{4}$.
+Hence, by Lebesgue dominated convergence theorem, $\mathsf{E}\psi_n(X_q^x)^2 \rightarrow 0$ and since $\mathsf{E}[f(X^y_{q,q+1})\phi_{k}(\xi_{q+1})]$ is a version of conditional expectation $\mathsf{E}[f(X^x_{q+1}|X^x_q = y]$, it holds in \(L^2(\mathsf{P})\)
+\begin{equation}
+\label{eq:1-step-decomp}
+f(X^x_{q+1}) = \mathsf{E}[f(X^x_{q+1})|\mathcal{G}_k] + \sum\limits_{k=1}^{\infty}a_{q+1,q+1,k}(X^x_{q})\phi_{k}(\xi_{q+1})
+\end{equation}
 where
-\begin{equation*}
-\alpha_{q+1,q+1,k}
-=\mathsf{E}\left[\left.f(X^x_{q+1})\phi_{k}(\xi_{q+1})\right|\mathcal G_q\right].
-\end{equation*}
-By the Markov property of $(X^x_{l})$, it holds $\mathsf{E}[f(X^x_{q+1})|\mathcal{G}_q] = \mathsf{E}[f(X^x_{q+1})| X_q^x]$.
+\[
+a_{q+1,q+1,k}(y) = \mathsf{E}\left[f(X^y_{q,q+1})\phi_k\left(\xi_{q+1}\right)\right]
+\]
+
+%we get by
+%conditioning on $\mathcal{G}_{q}$,
+%\[
+%f(X^x_{q+1})=\mathsf{E}\left[\left.f(X^x_{q+1})\right|{\mathcal{G}}_q\right]+\sum_{k\geq1}\alpha_{q+1,q+1,k}\phi_{k}(\xi_{q+1}),
+%\]
+%where
+%\begin{equation*}
+%\alpha_{q+1,q+1,k}
+%=\mathsf{E}\left[\left.f(X^x_{q+1})\phi_{k}(\xi_{q+1})\right|\mathcal G_q\right].
+%\end{equation*}
+%By the Markov property of $(X^x_{l})$, it holds $\mathsf{E}[f(X^x_{q+1})|\mathcal{G}_q] = \mathsf{E}[f(X^x_{q+1})| X_q^x]$.
 Using Markov property together with conditioning on $\mathcal{G}_{q+1}$, we obtain
-\[
-\alpha_{q+1,q+1,k}=\mathsf{E}\left[\left.f(X^x_{q+1})\phi_{k}(\xi_{q+1})\right|X^x_{q}\right]
-=a_{q+1,q+1,k}(X^x_{q})
-\]
-where for $y \in \rset^d$
-\[
-a_{q+1,q+1,k}(y) = \mathsf{E}\left[f(X^y_{q,q+1})\phi_k\left(\xi_{q+1}\right)\right]
-\]
+%\[
+%\alpha_{q+1,q+1,k}=\mathsf{E}\left[\left.f(X^x_{q+1})\phi_{k}(\xi_{q+1})\right|X^x_{q}\right]
+%=a_{q+1,q+1,k}(X^x_{q})
+%\]
+%where for $y \in \rset^d$
+%\[
+%a_{q+1,q+1,k}(y) = \mathsf{E}\left[f(X^y_{q,q+1})\phi_k\left(\xi_{q+1}\right)\right]
+%\]
 

### 2020-02-20 aap/Revision/vr-mcmc-revision-new.tex
1st theorem proof update

@@
 \end{equation}
-Let us define sequence $\psi_{q+1,n}(y) = \mathsf{E}[|f(X^y_{q,q+1}) - \mathsf{E} f(X^y_{q,q+1})|^2] - \sum\limits_{k=1}^{n}\mathsf{E}^2[f(X^y_{q,q+1})\phi_k(\xi_{q+1})]$.
-Note that $\psi_{q+1,n}(y)$ coincides with the left side of \ref{eq:parseval}, and due to Parseval identity $\Psi_{q+1,n}(y) \rightarrow 0, n \rightarrow \infty$ for any $y \in \rset^d$.
-Note that $\psi_n(y) < \mathsf{E}[|f(X^y_{q,q+1}) - \mathsf{E} f(X^y_{q,q+1})|^2] \leq \frac{\|f\|^2_{\infty}}{4}$.
-Hence, by Lebesgue dominated convergence theorem, $\mathsf{E}\psi_n(X_q^x)^2 \rightarrow 0$ and since $\mathsf{E}[f(X^y_{q,q+1})\phi_{k}(\xi_{q+1})]$ is a version of conditional expectation $\mathsf{E}[f(X^x_{q+1}|X^x_q = y]$, it holds in \(L^2(\mathsf{P})\)
+Let us define sequence of functions
+\[
+\psi_{q+1,n}(y) = \mathsf{E}[|f(X^y_{q,q+1}) - \mathsf{E} f(X^y_{q,q+1})|^2] - \sum\limits_{k=1}^{n}\mathsf{E}^2[f(X^y_{q,q+1})\phi_k(\xi_{q+1})].
+\]
+which coincide with the left side of \ref{eq:parseval}.
+Due to Parseval identity $\Psi_{q+1,n}(y) \rightarrow 0, n \rightarrow \infty$ for any $y \in \rset^d$ and $\psi_n(y) < \mathsf{E}[|f(X^y_{q,q+1}) - \mathsf{E} f(X^y_{q,q+1})|^2] \leq \frac{\|f\|^2_{\infty}}{4}$.
+Hence, by Lebesgue dominated convergence theorem, $\mathsf{E}\psi^2_n(X_q^x) \rightarrow 0$ and since $\mathsf{E}[f(X^y_{q,q+1})]$ is a version of the conditional expectation $\mathsf{E}[f(X^x_{q+1}|X^x_q = y]$, it holds in \(L^2(\mathsf{P})\) that
 \begin{equation}
 \label{eq:1-step-decomp}
-f(X^x_{q+1}) = \mathsf{E}[f(X^x_{q+1})|\mathcal{G}_k] + \sum\limits_{k=1}^{\infty}a_{q+1,q+1,k}(X^x_{q})\phi_{k}(\xi_{q+1})
+f(X^x_{q+1}) = \mathsf{E}[f(X^x_{q+1})|X^x_{q+1}] + \sum\limits_{k=1}^{\infty}a_{q+1,q+1,k}(X^x_{q})\phi_{k}(\xi_{q+1})
 \end{equation}
@@
 \]
-
 %we get by
@@
 %\]
-
-We thus arrive at
-\begin{align}
-\label{berm:sig_X}
-f(X^x_{q+1})=\mathsf{E}\left[\left.f(X^x_{q+1})\right|X^x_{q}\right]+\sum_{k\geq1}a_{q+1,q+1,k}(X^x_{q})\phi_{k}(\xi_{q+1}),
-\end{align}
+%We thus arrive at
+%\begin{align}
+%\label{berm:sig_X}
+%f(X^x_{q+1})=\mathsf{E}\left[\left.f(X^x_{q+1})\right|X^x_{q}\right]+\sum_{k\geq1}a_{q+1,q+1,k}(X^x_{q})\phi_{k}(\xi_{q+1}),
+%\end{align}
 which is the required statement in the case $j=q$.
@@
 \]
-Formulas \eqref{berm:sig_X}
+Formulas \eqref{eq:1-step-decomp}
 and~\eqref{eq:28082017a1} conclude the induction step for $p = q+1$ and hence the proof.

### 2020-02-20 aap/Revision/vr-mcmc-revision-new.tex
1st theorem proof update

@@
 Due to Parseval identity $\Psi_{q+1,n}(y) \rightarrow 0, n \rightarrow \infty$ for any $y \in \rset^d$ and $\psi_n(y) < \mathsf{E}[|f(X^y_{q,q+1}) - \mathsf{E} f(X^y_{q,q+1})|^2] \leq \frac{\|f\|^2_{\infty}}{4}$.
-Hence, by Lebesgue dominated convergence theorem, $\mathsf{E}\psi^2_n(X_q^x) \rightarrow 0$ and since $\mathsf{E}[f(X^y_{q,q+1})]$ is a version of the conditional expectation $\mathsf{E}[f(X^x_{q+1}|X^x_q = y]$, it holds in \(L^2(\mathsf{P})\) that
+Hence, by Lebesgue dominated convergence theorem, $\mathsf{E}\psi^2_n(X_q^x) \rightarrow 0$ and since $\mathsf{E}[f(X^{X^x_q}_{q,q+1})|X^x_{q}] = \mathsf{E}[f(X^{x}_{q+1})|X^x_{q}]$, it holds in \(L^2(\mathsf{P})\) that
 \begin{equation}
 \label{eq:1-step-decomp}
-f(X^x_{q+1}) = \mathsf{E}[f(X^x_{q+1})|X^x_{q+1}] + \sum\limits_{k=1}^{\infty}a_{q+1,q+1,k}(X^x_{q})\phi_{k}(\xi_{q+1})
+f(X^x_{q+1}) = \mathsf{E}[f(X^x_{q+1})|X^x_{q}] + \sum\limits_{k=1}^{\infty}a_{q+1,q+1,k}(X^x_{q})\phi_{k}(\xi_{q+1})
 \end{equation}
@@
 \nonumber
-M_{K,n}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=N+1}^{p} \bar a_{p-l,k}(X_{l-1})\phi_k(\xi_{l})\right]
+M_{K,n}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=N+1}^{p} \bar a_{p-l,k}(X^x_{l-1})\phi_k(\xi_{l})\right]
 \\
-&=& \sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left(1+\frac{N-l}{n}\right)A_{N+n-l,k}(X_{l-1})\phi_{k}\left(\xi_{l}\right),
+&=& \sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left(1+\frac{N-l}{n}\right)A_{N+n-l,k}(X^x_{l-1})\phi_{k}\left(\xi_{l}\right),
 \label{eq:29032018a5}
@@
 \end{equation}
-Since \(X_{l-1}\) is independent of \(\xi_{l}\) and \tcr{\(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0\)}, we obtain
-\[
-\PE[A_{N+n-l,k}(X_{l-1})\phi_{k}\left(\xi_{l}\right)] = \mathsf{E}[A_{N+n-l,k}(X_{l-1})\CPE{\phi_{k}\left(\xi_{l}\right)}{\mathcal{G}_{l-1}]} = 0
+Since \(X^x_{l-1}\) is independent of \(\xi_{l}\) and \tcr{\(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0\)}, we obtain
+\[
+\PE[A_{N+n-l,k}(X^x_{l-1})\phi_{k}\left(\xi_{l}\right)] = \mathsf{E}[A_{N+n-l,k}(X^x_{l-1})\CPE{\phi_{k}\left(\xi_{l}\right)}{\mathcal{G}_{l-1}]} = 0
 \]
@@
 \label{eq:expression-variance}
-\PVar[\pi_{K,n}^N(f)]\leq\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X_{N+l-1})]
+\PVar[\pi_{K,n}^N(f)]\leq\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})]
 \end{equation}

### 2020-02-21 aap/Revision/vr-mcmc-revision-new.tex
update

@@
 \end{eqnarray}
-where for $y \in \rset^d$
+where for all $y \in \rset^d$,
 \begin{eqnarray}
@@
 Let us prove that the induction assumption holds for $p=q+1$.
-Given $f$ with $\mathsf{E}\left[|f(X^x_{q+1})|^2\right]<\infty$,
-due to the orthonormality and completeness
-of the system $\left(\phi_{k}\right)$, we get that for any $y \in \rset^d$ it holds in \(L^2(\mathsf{P})\) 
-\[
-f(X^y_{q,q+1}) = \mathsf{E} f(X^y_{q,q+1}) + \sum\limits_{k=1}^{\infty}\mathsf{E}[f(X^y_{q,q+1})\phi_{k}(\xi_{q+1})]\phi_k(\xi_{q+1})
-\]
-which means that
-\begin{equation}
-\label{eq:parseval}
-\lim_{n \rightarrow \infty}\mathsf{E}[|f(X^y_{q,q+1}) - \mathsf{E} f(X^y_{q,q+1}) - \sum\limits_{k=1}^{n}\mathsf{E}[f(X^y_{q,q+1})\phi_{k}(\xi_{q+1})]\phi_k(\xi_{q+1})|^2] = 0
-\end{equation}
-Let us define sequence of functions
-\[
-\psi_{q+1,n}(y) = \mathsf{E}[|f(X^y_{q,q+1}) - \mathsf{E} f(X^y_{q,q+1})|^2] - \sum\limits_{k=1}^{n}\mathsf{E}^2[f(X^y_{q,q+1})\phi_k(\xi_{q+1})].
-\]
-which coincide with the left side of \ref{eq:parseval}.
-Due to Parseval identity $\Psi_{q+1,n}(y) \rightarrow 0, n \rightarrow \infty$ for any $y \in \rset^d$ and $\psi_n(y) < \mathsf{E}[|f(X^y_{q,q+1}) - \mathsf{E} f(X^y_{q,q+1})|^2] \leq \frac{\|f\|^2_{\infty}}{4}$.
+For any function $f$ such that $\mathsf{E}\left[|f(X^x_{q+1})|^2\right]<\infty$,
+the orthonormality and completeness
+of the system $\left(\phi_{k}\right)_{k=0}^\infty$ implies that for any $y \in \rset^d$, 
+$\lim_{n \rightarrow \infty}\Psi_{n,q,1}(y)= 0$ with $\Psi_{n,q,1}(y)= \PE[|f(X^y_{q,1}) - f_{n,q,1}(y)|^2]$ and
+\[
+f_{n,q,1}(y) = \PE[f(X^y_{q,q+1})] +  \sum\limits_{k=1}^{n}\PE[f(X^y_{q,q+1})\phi_{k}(\xi_{q+1})]\phi_k(\xi_{q+1}).
+\]
+$\psi_n(y) < \mathsf{E}[|f(X^y_{q,q+1}) - \mathsf{E} f(X^y_{q,q+1})|^2] \leq \frac{\|f\|^2_{\infty}}{4}$.
 Hence, by Lebesgue dominated convergence theorem, $\mathsf{E}\psi^2_n(X_q^x) \rightarrow 0$ and since $\mathsf{E}[f(X^{X^x_q}_{q,q+1})|X^x_{q}] = \mathsf{E}[f(X^{x}_{q+1})|X^x_{q}]$, it holds in \(L^2(\mathsf{P})\) that

### 2020-02-21 aap/Revision/vr-mcmc-revision-new.tex
update vr-mcmc-revision-new.tex

@@
 of the system $\left(\phi_{k}\right)_{k=0}^\infty$ implies that for any $y \in \rset^d$,
-$\lim_{n \rightarrow \infty}\Psi_{n,q,1}(y)= 0$ with $\Psi_{n,q,1}(y)= \PE[|f(X^y_{q,1}) - f_{n,q,1}(y)|^2]$ and
+$\lim_{n \rightarrow \infty}\Psi_{n,q,1}(y)= 0$ with $\Psi_{n,q,1}(y)= \PE[|f(X^y_{q,q+1}) - f_{n,q,1}(y)|^2]$ and
 \[

### 2020-02-21 aap/Revision/vr-mcmc-revision-new.tex
thm 1 update

--- 2020-02-21 aap/Revision/vr-mcmc-revision-new.tex
@@
 We write $X^x_{p}$ and $G_{p}$ as a shorthand notation for $X^x_{0,p}$ and $G_{0,p}$, respectively.
+Let us also define by $\mathcal{F}_2^p$ class of Borel functions $f: \rset^d \to \rset$ such that for any $1 \leq l \leq q \leq p$ and $y \in \rset^d$, $\mathsf{E}\left[\left |f(X^y_{l,q})\right|^{2}\right]<\infty$.
+Note that for $q \leq p$ it holds $\mathcal{F}_2^p \subseteq \mathcal{F}_2^q$.
 
 \begin{thm}\label{prop:29032018a1}
-Let \((X^x_{p})_{p\geq 0}\) be a Markov chain of the form \eqref{eq:ula_new}.
-Then, for $p>j$ and for all Borel functions $f: \rset^{d}\to \rset$ such that $\mathsf{E}\left[\left |f(X^x_{p})\right|^{2}\right]<\infty$, the following representation holds in \(L^2(\mathsf{P})\)
+For all $p \in \nset,$ $q \leq p$, $j < q \leq p$, $f \in \mathcal{F}_2^p$ and $x \in \rset^d$ the following representation holds in \(L^2(\mathsf{P})\)
 \begin{eqnarray}
 \label{eq:mart_repr}
-f(X^x_{p})=\mathsf{E}\left[\left.f(X^x_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}a_{p,l,k}(X^x_{l-1})\phi_k\left(\xi_{l}\right),
+f(X^x_{q})=\mathsf{E}\left[\left.f(X^x_{q})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{q}a_{q,l,k}(X^x_{l-1})\phi_k\left(\xi_{l}\right),
 \end{eqnarray}
-where for all $y \in \rset^d$,
+where \((X^x_{p})_{p\geq 0}\) is given in \eqref{eq:ula_new} and for all $y \in \rset^d$
 \begin{eqnarray}
@@
 %a_{p,l,k}(x)=\mathsf{E}\left[\left.f(X_{p})\phi_k\left(\xi_{l}\right)\right|X_{l-1}=x\right], \quad p\geq l, \quad k\in \mathbb{N}.
-a_{p,l,k}(y)=\mathsf{E}\left[f(X^y_{l-1,p})\phi_k\left(\xi_{l}\right)\right], \quad p\geq l, \quad k\in \mathbb{N}
+a_{q,l,k}(y)=\mathsf{E}\left[f(X^y_{l-1,q})\phi_k\left(\xi_{l}\right)\right], \quad q\geq l, \quad k\in \mathbb{N}
 \end{eqnarray}
@@
 \begin{proof}%[Proof of Theorem~\ref{prop:29032018a1}]
-The expansion obviously holds for $p=1$ and $j=0$.
+The expansion obviously holds for $p = q=1$ and $j=0$.
 Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system in \(L^2(\mathbb{R}^d, P_{\xi})\), it holds in \(L^2(\mathsf{P})\) that
@@
 \]
-with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$.
-Assume now that (\ref{eq:mart_repr})
-holds for $p=q$, all $j<q$, and Borel-measurable functions
-$f$ with $\mathsf{E}\left[|f(X^x_{q})|^2\right]<\infty$.
-Let us prove that the induction assumption holds for $p=q+1$.
-For any function $f$ such that $\mathsf{E}\left[|f(X^x_{q+1})|^2\right]<\infty$,
+for any $f \in \mathcal{F}_2^p$  with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$.
+Assume now that (\ref{eq:mart_repr}) holds for some $m < p$, any $q \leq m$, $j < q \leq m$ and $f$ from $\mathcal{F}_2^p$.
+Let us prove that the induction assumption holds for $q=m+1$ and all $j < q$.
+For any function $f$ from $\mathcal{F}_2^p$,
 the orthonormality and completeness
 of the system $\left(\phi_{k}\right)_{k=0}^\infty$ implies that for any $y \in \rset^d$, 
-$\lim_{n \rightarrow \infty}\Psi_{n,q,1}(y)= 0$ with $\Psi_{n,q,1}(y)= \PE[|f(X^y_{q,1}) - f_{n,q,1}(y)|^2]$ and
-\[
-f_{n,q,1}(y) = \PE[f(X^y_{q,q+1})] +  \sum\limits_{k=1}^{n}\PE[f(X^y_{q,q+1})\phi_{k}(\xi_{q+1})]\phi_k(\xi_{q+1}).
+$\lim_{n \rightarrow \infty}\Psi_{n,m+1,1}(y)= 0$ with $\Psi_{n,m+1,1}(y)= \PE[|f(X^y_{m,m+1}) - f_{n,m+1,1}(y)|^2]$ and
+\[
+f_{n,m+1,1}(y) = \PE[f(X^y_{m,m+1})] +  \sum\limits_{k=1}^{n}\PE[f(X^y_{m,m+1})\phi_{k}(\xi_{m+1})]\phi_k(\xi_{m+1}).
 \]

### 2020-02-22 aap/Revision/vr-mcmc-revision-new.tex
thm 1 update

@@
 \]
-$\psi_n(y) < \mathsf{E}[|f(X^y_{q,q+1}) - \mathsf{E} f(X^y_{q,q+1})|^2] \leq \frac{\|f\|^2_{\infty}}{4}$.
-Hence, by Lebesgue dominated convergence theorem, $\mathsf{E}\psi^2_n(X_q^x) \rightarrow 0$ and since $\mathsf{E}[f(X^{X^x_q}_{q,q+1})|X^x_{q}] = \mathsf{E}[f(X^{x}_{q+1})|X^x_{q}]$, it holds in \(L^2(\mathsf{P})\) that
+Note that  for any $y \in \rset^d$ and $n \in \nset$ it holds $\Psi_{n,m+1,1}(y) \leq \Psi_{0,m+1,1}(y)$ and hence
+\begin{eqnarray*}
+\PE \psi_{0,m+1,1}(X_m^x) &=& \PE \left[ \left.
+\PE \left[ \psi_{0,m+1,1}(X_m^x) \right| \mathcal{G}_m\right]\right] = \PE \left[ \left.
+\PE \left[ |f(X^{X_m^x}_{m,m+1}) - \PE f(X^{X_m^x}_{m,m+1})|^2 \right| X_m^x\right]\right]
+\\
+&=& \PE \left[ |f(X_{m+1}^x - \PE f(X_{m+1}^x))|^2 \right] = \PVar{f(X_{m+1}^x)} < \infty
+\end{eqnarray*}
+Hence, by Lebesgue dominated convergence theorem, $\PE \psi_{n,m+1,1}(X_m^x) \rightarrow 0$ and since for all $y \in \rset^d$ the expectation $\PE[f(X^y_{m,m+1})]$ is a version of $\PE \left[f(X^y_{m,m+1}) | \mathcal{G}_m \right]$, it holds in \(L^2(\mathsf{P})\) that
 \begin{equation}
 \label{eq:1-step-decomp}
-f(X^x_{q+1}) = \mathsf{E}[f(X^x_{q+1})|X^x_{q}] + \sum\limits_{k=1}^{\infty}a_{q+1,q+1,k}(X^x_{q})\phi_{k}(\xi_{q+1})
+f(X^x_{m+1}) = \PE[f(X^x_{m+1})|\mathcal{G}_m] + \sum\limits_{k=1}^{\infty}a_{m+1,m+1,k}(X^x_{m})\phi_{k}(\xi_{m+1})
 \end{equation}
@@
 \[
-a_{q+1,q+1,k}(y) = \mathsf{E}\left[f(X^y_{q,q+1})\phi_k\left(\xi_{q+1}\right)\right]
+a_{m+1,m+1,k}(y) = \PE \left[f(X^y_{m,m+1})\phi_k\left(\xi_{m+1}\right)\right]
 \]
@@
 %\end{align}
-which is the required statement in the case $j=q$.
-Now assume $j<q$.
+which is the required statement in the case $q = m+1$ and $j=m$.
+Now assume that $j<m$.
 The random variable
-$\mathsf{E}\left[\left.f(X^x_{q+1})\right|X^x_{q}\right]$
+$\mathsf{E}\left[\left.f(X^x_{m+1})\right|X^x_{m}\right]$
 is square integrable and has the form
-$g(X^x_{q})$,
+$g(X^x_{m})$,
 hence the induction hypothesis applies, and we get

### 2020-02-22 aap/Revision/vr-mcmc-revision-new.tex
thm 1 update

@@
 The random variable
-$\mathsf{E}\left[\left.f(X^x_{m+1})\right|X^x_{m}\right]$
-is square integrable and has the form
-$g(X^x_{m})$,
-hence the induction hypothesis applies, and we get
+$\PE \left[\left.f(X^x_{m+1})\right| \mathcal{G}_m \right]$ has the form $g(X^x_{m})$ and is square integrable since 
+\[
+\PE \left[ \PE \left[\left.f(X^x_{m+1})\right| \mathcal{G}_m \right]^2 \right] \leq \PE \left[ \PE \left[\left.f^2(X^x_{m+1})\right| \mathcal{G}_m \right] \right] = \PE \left[f^2(X^x_{m+1})\right] < \infty
+\]
+Hence the induction hypothesis applies, and we get
 \begin{equation}\label{eq:28082017a1}
-\mathsf{E}\left[\left.f(X^x_{q+1})\right|X^x_{q}\right]=\mathsf{E}\left[\left.f(X^x_{q+1})\right|X^x_{j}\right]+\sum_{k\geq1}\sum_{l=j+1}^{q}a_{q+1,l,k}(X^x_{l-1})\phi_{k}(\xi_{l})
+\mathsf{E}\left[\left.f(X^x_{m+1})\right| \mathcal{G}_m \right]=\mathsf{E}\left[\left.f(X^x_{m+1})\right| \mathcal{G}_j \right]+\sum_{k\geq1}\sum_{l=j+1}^{m}a_{m+1,l,k}(X^x_{l-1})\phi_{k}(\xi_{l})
 \end{equation}
@@
 \begin{align*}
-a_{q+1,l,k}(X^x_{l-1}) &= \mathsf{E}\left[\left.\mathsf{E}\left[\left.f(X^x_{q+1})\right|\mathcal{G}_{q}\right]\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]
- = \mathsf{E}\left[\left.f(X^x_{q+1})\phi_{k}(\xi_{l})\right|X^x_{l-1}\right].
+a_{m+1,l,k}(X^x_{l-1}) &= \mathsf{E}\left[\left.\mathsf{E}\left[\left.f(X^x_{m+1})\right|\mathcal{G}_{m}\right]\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]
+ = \mathsf{E}\left[\left.f(X^x_{m+1})\phi_{k}(\xi_{l})\right|X^x_{l-1}\right].
 \end{align*}
@@
 \[
-a_{q+1,l,k}(y) = \mathsf{E}\left[f(X^y_{l-1,q+1})\phi_{k}(\xi_{l})\right]
+a_{m+1,l,k}(y) = \mathsf{E}\left[f(X^y_{l-1,m+1})\phi_{k}(\xi_{l})\right]
 \]
 Formulas \eqref{eq:1-step-decomp}
-and~\eqref{eq:28082017a1} conclude the induction step for $p = q+1$ and hence the proof.
+and~\eqref{eq:28082017a1} conclude the induction step for $q = m+1$ and all $j < q$ and hence the proof.
 \end{proof}
@@
 \begin{cor}
-If  all the kernels $\Phi_l,$ $l\geq 1,$ are equal,  then  the representation \eqref{eq:mart_repr} takes the form
-\begin{eqnarray*}
-f(X^x_{p})=\mathsf{E}\left[\left.f(X^x_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}\bar a_{p-l,k}(X^x_{l-1})\phi_k\left(\xi_{l}\right)
+If  all the kernels $\Phi_l,$ $l\geq 1,$ in \eqref{eq:ula_new} are equal,  then  the representation \eqref{eq:mart_repr} for all $p \in \nset,$ $q \leq p$, $j < q \leq p$, $f \in \mathcal{F}_2^p$ and $x \in \rset^d$ takes the form
+\begin{eqnarray*}
+f(X^x_{q})=\mathsf{E}\left[\left.f(X^x_{q})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{q}\bar a_{q-l,k}(X^x_{l-1})\phi_k\left(\xi_{l}\right)
 \end{eqnarray*}
@@
 \begin{prop}
-The coefficients \(a_{p,l,k}\) in \eqref{eq:coeff_mart}  can be alternatively represented as
-\begin{eqnarray*}
-a_{p,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{p,l}\left(\Phi_l(x,\xi)\right)\right]
-\end{eqnarray*}
-with \(Q_{p,l}(x)=\mathsf{E}\left[f(X^x_{l,p})\right],\) \(p\geq l.\)
+The coefficients \(a_{q,l,k}\) in \eqref{eq:coeff_mart}  can be alternatively represented as
+\begin{eqnarray*}
+a_{q,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{q,l}\left(\Phi_l(x,\xi)\right)\right]
+\end{eqnarray*}
+with \(Q_{q,l}(x)=\mathsf{E}\left[f(X^x_{l,q})\right],\) \(p\geq l.\)
 %where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
-The functions \((Q_{p,l})_{l=0}^p\)  can be computed by the backward recurrence: $Q_{p,p}(x)=f(x)$ and for $l \in \{0,\ldots,p-1\}$
+The functions \((Q_{q,l})_{l=0}^p\)  can be computed by the backward recurrence: $Q_{q,q}(x)=f(x)$ and for $l \in \{0,\ldots,q-1\}$
 \begin{eqnarray}
 \label{eq:qpl}
-Q_{p,l}(x)=\mathsf{E}\left[Q_{p,l+1}(X^x_{l+1})\right].
+Q_{q,l}(x)=\mathsf{E}\left[Q_{q,l+1}(X^x_{l+1})\right].
 %=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))]
@@
 \\
-&=& \sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left(1+\frac{N-l}{n}\right)A_{N+n-l,k}(X^x_{l-1})\phi_{k}\left(\xi_{l}\right),
+&=& \sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left(1+\frac{N-l+1}{n}\right)A_{N+n-l,k}(X^x_{l-1})\phi_{k}\left(\xi_{l}\right),
 \label{eq:29032018a5}

### 2020-02-22 aap/Revision/vr-mcmc-revision-new.tex
update
update part 1

--- 2020-02-21 aap/Revision/vr-mcmc-revision-new.tex
@@
 As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of Shifted Legendre polynomials for "uniform part" and Hermite polynomials for "Gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the Shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\)  Let $\left(\Omega,\mathcal{F},\mathsf{P}\right)$ be a probability space, $(\xi_p)_{p \in \nset}$ be i.i.d.
-$d_{\xi}-$dimensional random vectors.
+$m-$dimensional random vectors.
 We denote by $(\mathcal{G}_p)_{p \in \nset_0}$  the filtration generated by $(\xi_p)_{p \in \nset}$ with the convention $\mathcal{G}_0=\mathrm{triv}$.
-For  $x \in \rset^d$ and $y \in \rset^{d_\xi}$ let $\Phi_k(x,y)$ be a function mapping $\rset^{d+d_{\xi}}$ to $\rset^d$.
+For  $x \in \rset^d, y \in \rset^{m}$ and $k \in \nset$ let $\Phi_k(x,y)$ be a function mapping $\rset^{d+m}$ to $\rset^d$.
 Then we can define for $l \le p$
@@
 We write $X^x_{p}$ and $G_{p}$ as a shorthand notation for $X^x_{0,p}$ and $G_{0,p}$, respectively.
-Let us also define by $\mathcal{F}_2^p$ class of Borel functions $f: \rset^d \to \rset$ such that for any $1 \leq l \leq q \leq p$ and $y \in \rset^d$, $\mathsf{E}\left[\left |f(X^y_{l,q})\right|^{2}\right]<\infty$.
-Note that for $q \leq p$ it holds $\mathcal{F}_2^p \subseteq \mathcal{F}_2^q$.
 
 \begin{thm}\label{prop:29032018a1}
-For all $p \in \nset,$ $q \leq p$, $j < q \leq p$, $f \in \mathcal{F}_2^p$ and $x \in \rset^d$ the following representation holds in \(L^2(\mathsf{P})\)
+For all $p \in \nset,$ $q \leq p$, $j < q \leq p$, Borel bounded functions $f: \rset^d \rightarrow \rset$ and $x \in \rset^d$ the following representation holds in \(L^2(\mathsf{P})\)
 \begin{eqnarray}
@@
 %a_{p,l,k}(x)=\mathsf{E}\left[\left.f(X_{p})\phi_k\left(\xi_{l}\right)\right|X_{l-1}=x\right], \quad p\geq l, \quad k\in \mathbb{N}.
-a_{q,l,k}(y)=\mathsf{E}\left[f(X^y_{l-1,q})\phi_k\left(\xi_{l}\right)\right], \quad q\geq l, \quad k\in \mathbb{N}
+a_{q,l,k}(y)=\mathsf{E}\left[f(X^y_{l-1,q})\phi_k\left(\xi_{l}\right)\right], \quad q\geq l, \quad k\in \nset
 \end{eqnarray}
@@
 \]
-for any $f \in \mathcal{F}_2^p$  with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$.
-Assume now that (\ref{eq:mart_repr}) holds for some $m < p$, any $q \leq m$, $j < q \leq m$ and $f$ from $\mathcal{F}_2^p$.
-Let us prove that the induction assumption holds for $q=m+1$ and all $j < q$.
-For any function $f$ from $\mathcal{F}_2^p$,
-the orthonormality and completeness
-of the system $\left(\phi_{k}\right)_{k=0}^\infty$ implies that for any $y \in \rset^d$, 
+for any bounded $f$ with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$.
+Assume now that (\ref{eq:mart_repr}) holds for some $m < p$, any $q \leq m$, $j < q \leq m$ and all bounded $f$. Let us prove that the induction assumption holds for $q=m+1$ and all $j < q$.
+The orthonormality and completeness
+of the system $\left(\phi_{k}\right)_{k=0}^\infty$ implies that for any bounded $f$ and $y \in \rset^d$, 
 $\lim_{n \rightarrow \infty}\Psi_{n,m+1,1}(y)= 0$ with $\Psi_{n,m+1,1}(y)= \PE[|f(X^y_{m,m+1}) - f_{n,m+1,1}(y)|^2]$ and
@@
 \]
-$\psi_n(y) < \mathsf{E}[|f(X^y_{q,q+1}) - \mathsf{E} f(X^y_{q,q+1})|^2] \leq \frac{\|f\|^2_{\infty}}{4}$.
-Hence, by Lebesgue dominated convergence theorem, $\mathsf{E}\psi^2_n(X_q^x) \rightarrow 0$ and since $\mathsf{E}[f(X^{X^x_q}_{q,q+1})|X^x_{q}] = \mathsf{E}[f(X^{x}_{q+1})|X^x_{q}]$, it holds in \(L^2(\mathsf{P})\) that
+Note that  for any $y \in \rset^d$ and $n \in \nset$ it holds $\Psi_{n,m+1,1}(y) \leq \Psi_{0,m+1,1}(y)$ and
+\begin{eqnarray*}
+\PE \left[ \psi_{0,m+1,1}(X_m^x) \right] &=& \PE \left[ \left.
+\PE \left[ \psi_{0,m+1,1}(X_m^x) \right| \mathcal{G}_m\right]\right] = \PE \left[ \left.
+\PE \left[ |f(X^{X_m^x}_{m,m+1}) - \PE f(X^{X_m^x}_{m,m+1})|^2 \right| X_m^x\right]\right]
+\\
+&=& \PE \left[ |f(X_{m+1}^x) - \PE f(X_{m+1}^x)|^2 \right] = \PVar{f(X_{m+1}^x)} < \infty
+\end{eqnarray*}
+Hence, by Lebesgue dominated convergence theorem, $\lim_{n \rightarrow \infty}\PE \left[\psi_{n,m+1,1}(X_m^x)\right] = 0$ and since for all $y \in \rset^d$ the expectation $\PE[f(X^y_{m,m+1})]$ is a version of $\PE \left[f(X^y_{m,m+1}) | \mathcal{G}_m \right]$, it holds in \(L^2(\mathsf{P})\) that
 \begin{equation}
 \label{eq:1-step-decomp}
-f(X^x_{q+1}) = \mathsf{E}[f(X^x_{q+1})|X^x_{q}] + \sum\limits_{k=1}^{\infty}a_{q+1,q+1,k}(X^x_{q})\phi_{k}(\xi_{q+1})
+f(X^x_{m+1}) = \PE[f(X^x_{m+1})|\mathcal{G}_m] + \sum\limits_{k=1}^{\infty}a_{m+1,m+1,k}(X^x_{m})\phi_{k}(\xi_{m+1})
 \end{equation}
@@
 \[
-a_{q+1,q+1,k}(y) = \mathsf{E}\left[f(X^y_{q,q+1})\phi_k\left(\xi_{q+1}\right)\right]
-\]
-%we get by
-%conditioning on $\mathcal{G}_{q}$,
-%\[
-%f(X^x_{q+1})=\mathsf{E}\left[\left.f(X^x_{q+1})\right|{\mathcal{G}}_q\right]+\sum_{k\geq1}\alpha_{q+1,q+1,k}\phi_{k}(\xi_{q+1}),
-%\]
-%where
-%\begin{equation*}
-%\alpha_{q+1,q+1,k}
-%=\mathsf{E}\left[\left.f(X^x_{q+1})\phi_{k}(\xi_{q+1})\right|\mathcal G_q\right].
-%\end{equation*}
-%By the Markov property of $(X^x_{l})$, it holds $\mathsf{E}[f(X^x_{q+1})|\mathcal{G}_q] = \mathsf{E}[f(X^x_{q+1})| X_q^x]$.
-Using Markov property together with conditioning on $\mathcal{G}_{q+1}$, we obtain
-%\[
-%\alpha_{q+1,q+1,k}=\mathsf{E}\left[\left.f(X^x_{q+1})\phi_{k}(\xi_{q+1})\right|X^x_{q}\right]
-%=a_{q+1,q+1,k}(X^x_{q})
-%\]
-%where for $y \in \rset^d$
-%\[
-%a_{q+1,q+1,k}(y) = \mathsf{E}\left[f(X^y_{q,q+1})\phi_k\left(\xi_{q+1}\right)\right]
-%\]
-%We thus arrive at
-%\begin{align}
-%\label{berm:sig_X}
-%f(X^x_{q+1})=\mathsf{E}\left[\left.f(X^x_{q+1})\right|X^x_{q}\right]+\sum_{k\geq1}a_{q+1,q+1,k}(X^x_{q})\phi_{k}(\xi_{q+1}),
-%\end{align}
-which is the required statement in the case $j=q$.
-Now assume $j<q$.
-The random variable
-$\mathsf{E}\left[\left.f(X^x_{q+1})\right|X^x_{q}\right]$
-is square integrable and has the form
-$g(X^x_{q})$,
-hence the induction hypothesis applies, and we get
+a_{m+1,m+1,k}(y) = \PE \left[f(X^y_{m,m+1})\phi_k\left(\xi_{m+1}\right)\right]
+\]
+which is the required statement in the case $q = m+1$ and $j=m$.
+Now assume that $j<m$.
+Set $g(y) = \PE \left[f(X_{m,m+1}^y)\right]$.
+Note that $\mathsf{P}$-a.s.
+it holds $g(X_m^x) = \PE\left[\left.f(X_m^x) \right| \mathcal{G}_m\right]$ and $g$ is bounded by construction.
+Hence the induction hypothesis applies, and we get
 \begin{equation}\label{eq:28082017a1}
-\mathsf{E}\left[\left.f(X^x_{q+1})\right|X^x_{q}\right]=\mathsf{E}\left[\left.f(X^x_{q+1})\right|X^x_{j}\right]+\sum_{k\geq1}\sum_{l=j+1}^{q}a_{q+1,l,k}(X^x_{l-1})\phi_{k}(\xi_{l})
+\mathsf{E}\left[\left.f(X^x_{m+1})\right| \mathcal{G}_m \right]=\mathsf{E}\left[\left.f(X^x_{m+1})\right| \mathcal{G}_j \right]+\sum_{k\geq1}\sum_{l=j+1}^{m}a_{m+1,l,k}(X^x_{l-1})\phi_{k}(\xi_{l})
 \end{equation}
@@
 \[
-a_{q+1,l,k}(y) = \mathsf{E}\left[f(X^y_{l-1,q+1})\phi_{k}(\xi_{l})\right]
+a_{m+1,l,k}(y) = \mathsf{E}\left[f(X^y_{l-1,m+1})\phi_{k}(\xi_{l})\right]
 \]
 Formulas \eqref{eq:1-step-decomp}
-and~\eqref{eq:28082017a1} conclude the induction step for $p = q+1$ and hence the proof.
+and~\eqref{eq:28082017a1} conclude the induction step for $q = m+1$ and all $j < q$ and hence the proof.
 \end{proof}
 
+\begin{remark}\label{rem:general_moments} Denote by $\mathcal{F}_2^p$ class of Borel functions $f: \rset^d \to \rset$ such that for any $n \leq p$, any indices $1 \leq i_1 < \ldots < i_n \leq p$ and $y \in \rset^d$,
+\begin{equation}
+\label{def:f_2_p}
+\mathsf{E}\left[\left |f(\Phi_{i_n}(\cdot,\xi_{i_n})\circ\Phi_{i_{n-1}}(\cdot,\xi_{i_{n-1}})\circ\dots\circ\Phi_{i_{1}}(y,\xi_{i_{1}}))\right|^{2}\right]<\infty
+\end{equation}
+Then the statement of Theorem~\ref{prop:29032018a1} remains valid for $f \in \mathcal{F}_2^p$. 
+\end{remark}
+
+If  all the functions $\Phi_l,$ $l\geq 1,$ in \eqref{eq:ula_new} are equal, then the condition \eqref{def:f_2_p} reduces to $\PE \left[ f^2(X_q^y)\right] < \infty$ for all $q \leq p$ and $y \in \rset^d$.
+Let us denote this class of functions by $\mathcal{F}_{2,\text{hom}}^p$.
+
 \begin{cor}
-If  all the kernels $\Phi_l,$ $l\geq 1,$ are equal,  then  the representation \eqref{eq:mart_repr} takes the form
-\begin{eqnarray*}
-f(X^x_{p})=\mathsf{E}\left[\left.f(X^x_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}\bar a_{p-l,k}(X^x_{l-1})\phi_k\left(\xi_{l}\right)
+Let \((X^x_{p})_{p\geq 0}\) be a homogeneous Markov chain of form \eqref{eq:ula_new} with $\Phi_l = \Phi,$ $l\geq 1$.
+Then for all $p \in \nset,$ $q \leq p$, $j < q \leq p$, $f \in \mathcal{F}_{2,\text{hom}}^p$ and $x \in \rset^d$ it holds in \(L^2(\mathsf{P})\)
+\begin{eqnarray*}
+f(X^x_{q})=\mathsf{E}\left[\left.f(X^x_{q})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{q}\bar a_{q-l,k}(X^x_{l-1})\phi_k\left(\xi_{l}\right)
 \end{eqnarray*}
@@
 \begin{prop}
-The coefficients \(a_{p,l,k}\) in \eqref{eq:coeff_mart}  can be alternatively represented as
-\begin{eqnarray*}
-a_{p,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{p,l}\left(\Phi_l(x,\xi)\right)\right]
-\end{eqnarray*}
-with \(Q_{p,l}(x)=\mathsf{E}\left[f(X^x_{l,p})\right],\) \(p\geq l.\)
+Let $q\geq l, k\in \nset$. Then the coefficients \(a_{q,l,k}(x)\) in \eqref{eq:coeff_mart}  can be alternatively represented as
+\begin{eqnarray*}
+a_{q,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{q,l}\left(\Phi_l(x,\xi)\right)\right]
+\end{eqnarray*}
+with \(Q_{q,l}(x)=\mathsf{E}\left[f(X^x_{l,q})\right],\) \(q\geq l.\)
 %where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
-The functions \((Q_{p,l})_{l=0}^p\)  can be computed by the backward recurrence: $Q_{p,p}(x)=f(x)$ and for $l \in \{0,\ldots,p-1\}$
+The functions \((Q_{q,l})_{l=0}^q\)  can be computed by the backward recurrence: $Q_{q,q}(x)=f(x)$ and for $l \in \{0,\ldots,q-1\}$
 \begin{eqnarray}
 \label{eq:qpl}
-Q_{p,l}(x)=\mathsf{E}\left[Q_{p,l+1}(X^x_{l+1})\right].
+Q_{q,l}(x)=\mathsf{E}\left[Q_{q,l+1}(X^x_{l+1})\right].
 %=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))]
 \end{eqnarray}
-In the case $\Phi_l=\Phi$ for all $l\geq 1,$ we have
+In the case $\Phi_l=\Phi$ in \eqref{eq:ula_new} for all $l\geq 1,$ we have
 \begin{equation}
@@
 Next we show how  the representation \eqref{eq:mart_repr} can be used to  reduce the variance of MCMC algorithms.
-Consider the case of time homogeneous transition kernels and introduce a weighted average estimator $\pi_{n}^{N}(f)$ of the form
+Consider the time homogeneous case ($\Phi_l = \Phi$ for all $l \in \nset$) and the sample average
 \begin{equation}\label{eq:29032018a2}
@@
 \end{equation}
-Since \(X^x_{l-1}\) is independent of \(\xi_{l}\) and \tcr{\(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0\)}, we obtain
+Since \(X^x_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0\), we obtain
 \[
@@
 In this section we perform the convergence analysis of the ULA algorithm.
+We use the notations of example \ref{exam:langevin-algorithm}.
 For the sake of clarity and notational simplicity we restrict our attention to the constant time step, that is, we take $\gamma_k =\gamma$ for any $k \in \nset$.
-
-\sergey{N and N0 were defined here before}
-\eric{these notations are used before; it is best to put all the notations in a short paragraph just after the introduction}
 
@@
 $\pi_n^N(f)$ and $\pi_{K,n}^N(f)$ have the same conditional bias.
-Notice that this remains true also when we substitute
-the coefficients $a_{p,l,\mathbf{k}}$ in~\eqref{eq:29032018a5}
-with some independent approximations $\widehat a_{p,l,\mathbf{k}}.$
+%Notice that this remains true also when we substitute
+%the coefficients $a_{p,l,\mathbf{k}}$ in~\eqref{eq:29032018a5}
+%with some independent approximations $\widehat a_{p,l,\mathbf{k}}.$
 
@@
 \]
-\sergey{Changed notation here, put variance of the noise equal to $\gamma$}
-%\eric{what is the notation convention here; I do not like much $\mathcal{Q}$ but $Q$ is already define and it is perhaps not a great idea to supersede the previous notation}
 For a bounded Borel function~$f$,
@@
 \label{eq:06042018a2}
-\left\{\mathsf{E}[\pi_{K,n}^N(f)]-\pi(f)\right\}^2
+\left( \mathsf{E}[\pi_{K,n}^N(f)]-\pi(f)\right)^2
 =\left(\mathsf{E}[\pi_{n}^N(f)]-\pi(f)\right)^2
@@
 %where \tcr{\(\lesssim\) stands for inequality up to a constant not depending  on $\gamma,$ \(n\) and \(N.\)}
-\sergey{removed repeating explanation of $\lesssim$}
-%\eric{this is defined less precisely just above}
 \end{thm}
@@
 \end{eqnarray}
-where $F_s(X_0,Z_1,\ldots,Z_s)=s^{-1}\sum_{r=0}^s f(X_r)$ and $\partial_{Z_1}^{\mathbf{k}'}$
+where $F_s(x,Z_1,\ldots,Z_s)=s^{-1}\sum_{r=0}^s f(X^x_r)$ and $\partial_{Z_1}^{\mathbf{k}'}$
 stands for a weak partial derivative of the functional $F_s$ that also can be viewed as discretised version of Malliavin derivative. From \eqref{eq:Ask-part}, we derive
@@
 \left(\frac{\gamma}{2}\right)^{|I|K}
-\PVar_x\left( s^{-1} \sum\nolimits_{p=1}^{s}\partial_{y_{1}}^{\mathbf{K}_I}f\left(X_{p}\right)
+\PVar \left( s^{-1} \sum\nolimits_{p=1}^{s}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X^x_{p}\right)
 \right),
@@
  Finally we show via  the Poincare inequality that under smoothness assumption \eqref{eq:smooth-mu} it holds
- \eric{Since we do note state was $W_s$ is, it is a bit difficult to guess why we are using a Poincare inequality there}
+ \eric{Since we do not state was $W_s$ is, it is a bit difficult to guess why we are using a Poincare inequality there}
 \begin{equation*}
-\PVar_x\left(s^{-1} \sum\nolimits_{p=1}^{s}\partial_{y_{1}}^{\mathbf{K}_I}f\left(X_{p}\right)
+\PVar \left(s^{-1} \sum\nolimits_{p=1}^{s}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X^x_{p}\right)
 \right)\leq W_s(x)
@@
 for all $x$ and some family of functions $(W_s)_{s\geq 1}$ such that  the sum $\sum_{s=1}^n \mathsf{E}[W_s(X_{N+n-s-1})]$ is bounded uniformly in $n\in \mathbb{N}.$
+\eric{Speak about truncation here.}
 
@@
 \label{sec:num}
-We will perform numerical experiments based on ULA with constant step size $\gamma$.
-Suppose that each function  \(Q_{r}(x)\) from \ref{eq:definition-bar-a} can be well approximated by polynomials, and we constructed a polynomial approximation  for each \(Q_r(x)\)  in the form:
+We perform numerical experiments based on ULA algorithm \ref{exam:langevin-algorithm} with constant step size $\gamma$.
+We construct a polynomial approximation  for each \(Q_r(x)\) in \eqref{eq:definition-bar-a} of the form:
 \begin{equation*}
-\widehat{Q}_{r}(x) = \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} x^{\mathbf{s}},\quad s=(s_1,\ldots,s_d)
+\widehat{Q}_{r}(x) = \sum_{\|\mathbf{s}\|\leq m} \widehat\beta_{\mathbf{s}} x^{\mathbf{s}},\quad s=(s_1,\ldots,s_d) \in \nset_0^d
 \end{equation*}
-with some coefficients \(\beta_{\mathbf{s}}\in \mathbb{R}.\)
+with some coefficients \(\widehat \beta_{\mathbf{s}}\in \mathbb{R}\) obtained using a modified least-squares criteria based on $T$ independent training trajectories \(\left\{\left(X_1^{(s)},\ldots,X_{N+n}^{(s)}\right)\right\}_{s=1}^{T}\) with $N$ being the size of burn-in period:
+\begin{eqnarray}
+\label{eq:Qregr}
+\widehat{Q}_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
+\end{eqnarray}
+for \(1 \leq r \leq n-1,\) $\Psi$ being the class of polynomials $\Psi = \{\psi(x) | \psi(x) = \sum_{\|\mathbf{s}\|\leq m} \alpha_{\mathbf{s}} x^{\mathbf{s}} \}$ and $\widehat{Q}_0(x) = f(x)$ by definition.
 Then using the identity
@@
 \]
-we obtain closed-form expression for the estimates $\widehat a_{r,k}(x)$ of functions $\bar a_{r,k}(x)$ from \ref{eq:definition-bar-a}. Namely, for all $x \in \mathbb{R}^d$,
+we obtain closed-form expression for the estimates $\widehat a_{r,k}(x)$ of functions $\bar a_{r,k}(x)$ in \eqref{eq:definition-bar-a}. Namely, for all $x \in \mathbb{R}^d$,
 \begin{eqnarray*}

### 2020-02-24 Code/martingale.py
code update: multidimensional case

@@ -84,12 +84,48 @@ def approx_q(X_train,Y_train,N_traj_train,lag,max_deg):
             coefs_poly = np.concatenate((coefs_poly,coefs),axis=0)
     return coefs_poly
 
-def test_traj(r_seed,max_deg):
+def get_indices_poly(ind,K_max,S_max):
+    """
+    Transforms 1d index into 2d index
+    """
+    S = ind % (S_max + 1)
+    K = ind // (S_max + 1) 
+    return K,S
+
+def init_basis_polynomials(K_max,S_max,st_norm_moments,gamma):
+    """
+    Represents E[H_k(\xi)*(x-\gamma \mu(x) + \sqrt{2\gamma}\xi)^s] as a polynomial of variable $y$, where y = x - \gamma \mu(x)
+    Args:
+        K_max - maximal degree of Hermite polynomial;
+        S_max - maximal degree of regressor polynomial;
+        st_norm_moments - array containing moments of standard normal distribution;
+    Return:
+        Polynomial coefficients
+    """
+    Poly_coefs_regression = np.zeros((K_max+1,S_max+1,S_max+1),dtype = float)
+    for k in range(Poly_coefs_regression.shape[0]):
+        for s in range(Poly_coefs_regression.shape[1]):
+            herm_poly = np.zeros(K_max+1, dtype = float)
+            herm_poly[k] = 1.0
+            herm_k = P.hermite_e.herme2poly(herm_poly)
+            herm_k = herm_k / np.sqrt(sp.special.factorial(k))
+            c = np.zeros(S_max+1, dtype = float)
+            for deg in range(s+1):
+                c[deg] = (np.sqrt(2*gamma)**(s-deg))*sp.special.binom(s,deg)*np.dot(herm_k,st_norm_moments[(s-deg):(s - deg+len(herm_k))])
+            Poly_coefs_regression[k,s,:] = c
+    return Poly_coefs_regression
+
+def compute_a_val(y,coefs,k):
+    """
+    compute values of a scalar product of a given polynomial with  
+    """
+
+def test_traj(r_seed,K_max):
     """
     """
     X_test,Noise = ULA(r_seed,Potential,step, N, n, d, return_noise = True)
     #compute number of basis polynomials
-    num_basis_funcs = (max_deg+1)**d - 1
+    num_basis_funcs = (K_max+1)**d
     print("number of basis functions = ",num_basis_funcs)
     #compute polynomials of noise variables Z_l
[…]

### 2020-02-24 Code/martingale.py
code update: multidimensional case

@@ -1,5 +1,6 @@
 import numpy as np
 import numpy.polynomial as P
+import scipy as sp
 from sklearn.preprocessing import PolynomialFeatures
 from samplers import ULA
 from potentials import GaussPotential,GaussMixture,GausMixtureIdent,GausMixtureSame
@@ -25,14 +26,14 @@ def split_index(k,d,max_deg):
         k = k // (max_deg + 1)
     return k_vec
 
-def Hermite_val(k_vec,x_vec):
+def hermite_val(k_vec,x_vec):
     P = 1.0
     d = x_vec.shape[0]
     for i in range(d):
         P = P * H(k_vec[i],x_vec[i])
     return P
 
-def Eval_Hermite(k,x_vec,max_deg):
+def eval_hermite(k,x_vec,max_deg):
     """
     Evaluates Hermite polynomials at component x_vec by multi-index obtained from single integer k;
     Args:
@@ -42,7 +43,7 @@ def Eval_Hermite(k,x_vec,max_deg):
     """
     k_vec = split_index(k,len(x_vec),max_deg)
     #now we initialised k_vec
-    return Hermite_val(k_vec,x_vec)
+    return hermite_val(k_vec,x_vec)
 
 def approx_q(X_train,Y_train,N_traj_train,lag,max_deg):
     """
@@ -94,7 +95,7 @@ def get_indices_poly(ind,K_max,S_max):
 
 def init_basis_polynomials(K_max,S_max,st_norm_moments,gamma):
     """
-    Represents E[H_k(\xi)*(x-\gamma \mu(x) + \sqrt{2\gamma}\xi)^s] as a polynomial of variable $y$, where y = x - \gamma \mu(x)
+    Represents E[H_k(xi)*(x-gamma mu(x) + sqrt{2gamma}xi)^s] as a polynomial of variable $y$, where y = x - gamma*mu(x)
     Args:
         K_max - maximal degree of Hermite polynomial;
         S_max - maximal degree of regressor polynomial;
@@ -115,47 +116,77 @@ def init_basis_polynomials(K_max,S_max,st_norm_moments,gamma):
             Poly_coefs_regression[k,s,:] = c
     return Poly_coefs_regression
 
-def compute_a_val(y,coefs,k):
+def init_moments(order):
     """
-    compute values of a scalar product of a given polynomial with  
+    Compute moments of standard normal distribution
     """
+    moments_stand_norm = np.zeros(2*order+1,dtype = float)
+    moments_stand_norm[0] = 1.0
[…]

### 2020-03-01 Code/martingale.py
code update

@@ -166,6 +166,7 @@ def get_representations(k,s,d,K_max):
                               [0,2,0,0],[0,1,1,0],[0,1,0,1],
                               [0,0,2,0],[0,0,1,1],
                               [0,0,0,2]])
+        s_vec = vec_table[s,:]
     else:
         vec_table = np.zeros((d+1,d),dtype = int)
         for ind in range(1,len(vec_table)):
@@ -218,10 +219,13 @@ def test_traj(Potential,coefs_poly_regr,step,r_seed,lag,K_max,S_max,N_burn,N_tes
                 a_cur = np.ones(coefs_poly_regr.shape[1], dtype = float)
                 for s in range(len(a_cur)):
                     k_vect,s_vect = get_representations(k,s,d,K_max)
+                    #print("K = ",k_vect)
+                    #print("S = ",s_vect)
                     for dim_ind in range(d):
                         a_cur[s] = a_cur[s]*P.polynomial.polyval(x_next[dim_ind],table_coefs[k_vect[dim_ind],s_vect[dim_ind],:])
                 a_vals[-(func_order+1),k] = np.dot(a_cur,coefs_poly_regr[func_order,:])
             #OK, now I have coefficients of the polynomial, and I need to integrate it w.r.t. Gaussian measure
+        #print("sum of coefficients",np.sum(np.abs(a_vals)))
         cvfs[i] = np.sum(a_vals*(poly_vals[:,i-num_lags+1:i+1].T))
         #save results
         test_stat_vanilla[i] = np.mean(f_vals_vanilla[1:(i+1)])

### 2020-03-03 aap/Revision/vr-mcmc-revision-new.tex
update

@@
 \subsection{Gaussian mixtures}
- We consider a sample generated by ULA with $\pi$  given by the mixture of two Gaussian distributions  with equal weights:
-\[
-\pi (x) = \frac{1}{2(2\pi)^{d/2}} \left( \rme^{\frac{-\| x-a\|_2^2}{2}}  + \rme^{\frac{-\| x+a\|_2^2}{2}} \right), \quad x \in\mathbb R^d
-\]
-where $a \in \mathbb{R}^d$ is a given vector. The function $U(x)$ and its gradient are given by
+ We consider a sample generated by ULA with $\pi$  given by the mixture of two equally-weighted $d$-dimensional Gaussian distributions of the following form
+ \[
+ \pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}|\Sigma|}} \left( \rme^{-\frac{(x-\mu)^T\Sigma^{-1}(x-\mu)}{2}} + \rme^{-\frac{(x+\mu)^T\Sigma^{-1}(x+\mu)}{2}}\right)
+ \]
+where $\rho \in (0,1), \mu \in \rset^d, \Sigma \in \rset^{d \times d}$. The function $U(x)$ and its gradient are given by
 \[
@@
 respectively.
-In our experiments we considered dimensions $d = 2$ and $d = 8$ and take $a=( (2d)^{-1/2}, \dots, (2d)^{-1/2})^\top$.
-In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i,\)
- we have used constant step size $\gamma=0.2$ and sampled $n = 10^4$ samples for $d = 2$ and $n = 2 \times 10^3$ for $d = 8$.
-We generated $\NtrainPath = 10$ independent "training" trajectories and solved the least squares  problems \eqref{eq:Qcircr} using the first order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section.
-The truncation level \(n_{\mathrm{trunc}}\) is chosen to be \(50.\) To test our variance reduction algorithm, we generated 100 independent trajectories.
- In Figure ~\ref{fig:1} we compare our approach to variance reduction methods of \cite{mira2013zero} and \cite{brosse2018diffusion}.
+In our experiments we considered $d = 2$, $\mu = \left(0.5,0.5\right)$ and $\Sigma = \Sigma_0$ with randomly generated positive-definite matrix $\Sigma_0$ (heterogeneous structure).
+In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i,\) and \(f(x) = \sum_{i=1}^d x^2_i\) we have used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 10$ independent "training" trajectories of length $n = 10^4$.
+Then we solve the least squares  problems \eqref{eq:Qcircr} using the first order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section, hence we set $K = 1$.
+We set the truncation level \(n_{\mathrm{trunc}} = 50\).
+To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 10^3$ and plot boxplots of the ergodic averages in Figure ~\ref{fig:GMM}.
+Our approach (MDCV, Martingale decomposition control variates) is compared to other variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}.
+
+ \begin{figure}[tbh]
+\centering
+\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
+\includegraphics[width=0.45\textwidth]{../pictures/GMM_2d_lin_sum.png}
+\label{fig:subfig1}}
+\qquad
+\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
+\includegraphics[width=0.45\textwidth]{../pictures/GMM_2d_lin_sum_squares.png}
+\label{fig:subfig2}}
+\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Gaussian mixture model.
+The compared estimators are the ordinary empirical average  (Vanilla), our estimator of control variates (MDCV), zero variance estimator (ZV) and control variates obtained with empirical spectral variance minimisation (ESVM)
+\label{fig:GMM}}
+\end{figure}
+
+\subsection{Binary Logistic Regression}
+
+Second experiment considers the problem of logistic regression.
+Let $\mathsf{Y} = \left(\mathsf{Y}_1,\ldots,\mathsf{Y}_n\right) \in \{0,1\}^m$ be binary response variables, $\mathsf{X} \in \rset^{m \times d}$ be a feature matrix and $\theta \in \rset^d$ - vector of regression parameters.
+We define log-likelihood of $i$-th observation as
+\[
+\ell(\mathsf{Y}_i | \theta, \mathsf{X}_i) = \mathsf{Y}_i \mathsf{X}_i^T\theta - \ln{\left(1 + \rme^{\mathsf{X}_i^T\theta}\right)}
+\]
+In order to estimate $\theta$ according to given data, the Bayesian approach introduces prior distribution $\pi_0(\theta)$ and inferences the posterior density $\pi(\theta| \mathsf{Y}, \mathsf{X})$ using Bayes' rule.
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=0.6\textwidth]{../pictures/logreg_1st_order.png}
+\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Logistic Regression.
+The compared estimators are the ordinary empirical average (Vanilla), our estimator of control variates (MDCV), zero variance estimator (ZV) control variates obtained with empirical spectral variance minimisation (ESVM).
+ \label{fig:2}}
+\end{center}
+\end{figure}
+
+In the case of  Gaussian prior $\pi_0(\theta) \sim \mathcal{N}(0,\sigma^2 I_d)$, the unnormalized posterior density takes the form:
+\begin{eqnarray*}
+\pi(\theta | \mathsf{Y}, \mathsf{X}) \propto \exp \left\{ \mathsf{Y}^T \mathsf{X} \theta - \sum\limits_{i=1}^m \ln{\left(1 + \rme^{\mathsf{X}_i \theta}\right)} - \frac{1}{2\sigma^2} \left\|\theta\right\|_2^2 \right\},
+\end{eqnarray*}
+Thus we obtain
+\begin{eqnarray*}
+U(\theta) = \mathsf{Y}^T \mathsf{X} \theta + \sum_{i=1}^m \ln{\left(1 + \rme^{-\theta^T \mathsf{X}_i})} + \frac{1}{2\sigma^2} \left\|\mathbf{\theta}\right\|_2^2, \,\,
+\nabla U(\mathbf{\theta}) = \mathbf{X}^T \mathbf{Y} - \sum_{i=1}^m \frac{\mathbf{X}_i}{1 + e^{\mathbf{\theta}^T \mathbf{X}_i}} + \frac{1}{\sigma^2} \mathbf{\theta}.
+\end{eqnarray*}
+To demonstrate the performance of the proposed control variates approach in the above Bayesian logistic regression model, we take a simple dataset from \cite{mira2013zero}, which contains the measurements of four variables on $m=200$ Swiss banknotes.
+Prior distribution of the regression parameter $\mathbf{\theta} = \left( \theta_1, \theta_2, \theta_3, \theta_4 \right)$ assumed to be normal with the covariance matrix $\sigma^2 I_4$, where $\sigma^2 = 100$.
+To construct trajectories of length $n = 1 \times 10^4, $ we take step size $\gamma = 0.1$ for the ULA scheme with $N = 10^3$ burn-in steps.
+As in the previous experiment we use the first order polynomials approximations to analytically compute the coefficients $\hat{a}_{p-l,\mathbf{k}},$ based on $T = 10$ training trajectories.
+We put $n_{trunc} = 50$ and $K = 1$.
+The target function is taken to be $f(\theta) = \sum_{i=1}^{i=d}\theta_i$.
+In order to test our variance reduction algorithm, we generate \(100\) independent test trajectories of length $n = 10^3$.
+In Figure ~\ref{fig:2} we compare our approach to the variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}.
 

### 2020-03-04 aap/Revision/vr-mcmc-revision-new.tex
experiments description

@@
 \[
-U(x) = \frac{1}{2} \|x - a\|_2^2 - \text{log}(1 + \rme^{-2x^\top a}), \,\,
-\nabla U(x) = x-a +2a(1 + \rme^{2 x^\top a})^{-1},
+U(x) = \frac{1}{2}(x-\mu)^T\Sigma^{-1}(x-\mu) - \ln{\left(1 + \rme^{-2\mu^\top \Sigma^{-1}x}\right)}, \quad
+\]
+\[
+\nabla U(x) = \Sigma^{-1}(x-\mu) +2\mu^\top\Sigma^{-1}\left(1 + \rme^{2 \mu^\top\Sigma^{-1} x}\right)^{-1},
 \]
@@
 In our experiments we considered $d = 2$, $\mu = \left(0.5,0.5\right)$ and $\Sigma = \Sigma_0$ with randomly generated positive-definite matrix $\Sigma_0$ (heterogeneous structure).
-In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i,\) and \(f(x) = \sum_{i=1}^d x^2_i\) we have used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 10$ independent "training" trajectories of length $n = 10^4$.
+In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i\) or \(f(x) = \sum_{i=1}^d x^2_i\) we used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 10$ independent training trajectories, each one of size $n = 10^4$.
 Then we solve the least squares  problems \eqref{eq:Qcircr} using the first order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section, hence we set $K = 1$.
@@
 Our approach (MDCV, Martingale decomposition control variates) is compared to other variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}.
+In the baselines we use first order polynomials, since they are used as regressors during computations of $\hat{a}_{p-l,\mathbf{k}}$.
 
@@
 \begin{eqnarray*}
-U(\theta) = \mathsf{Y}^T \mathsf{X} \theta + \sum_{i=1}^m \ln{\left(1 + \rme^{-\theta^T \mathsf{X}_i})} + \frac{1}{2\sigma^2} \left\|\mathbf{\theta}\right\|_2^2, \,\,
-\nabla U(\mathbf{\theta}) = \mathbf{X}^T \mathbf{Y} - \sum_{i=1}^m \frac{\mathbf{X}_i}{1 + e^{\mathbf{\theta}^T \mathbf{X}_i}} + \frac{1}{\sigma^2} \mathbf{\theta}.
+U(\theta) = -\mathsf{Y}^T \mathsf{X} \theta + \sum_{i=1}^m \ln{\left(1 + \rme^{\theta^T \mathsf{X}_i}\right)} + \frac{1}{2\sigma^2} \left\| \theta\right\|_2^2, \quad
+\nabla U(\theta) = -\mathsf{Y}^T \mathsf{X} + \sum_{i=1}^m \frac{\mathsf{X}_i}{1 + \rme^{-\mathsf{\theta}^T \mathsf{X}_i}} + \frac{1}{\sigma^2} \theta.
 \end{eqnarray*}
@@
 In Figure ~\ref{fig:2} we compare our approach to the variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}.
+In the baselines we use first order polynomials for the same reasons as in the Gaussian mixtures example.
 

### 2020-03-04 aap/Revision/vr-mcmc-revision-new.tex
lemma 8 update

@@
  \]
-where $\rho \in (0,1), \mu \in \rset^d, \Sigma \in \rset^{d \times d}$. The function $U(x)$ and its gradient are given by
+where $\rho \in (0,1), \mu \in \rset^d, \Sigma \in \rset^{d \times d}$ and $|\Sigma|$ is its determinant. The function $U(x)$ and its gradient are given by
 \[
@@
 \[
-\zeta_{l,j}(u)=\mathsf{E}\left[e^{-u\sum_{k=l}^{j}\gamma\mu'(X_{k-1})}\right],\quad u>0.
+\zeta_{l,j}(u)=\mathsf{E}\left[e^{-u\gamma\sum_{k=l}^{j}\mu'(X_{k-1})}\right],\quad u>0.
 \]
@@
 \[
-\mathsf{E}\left[e^{-u\gamma\left(\sum_{k=l}^{j}[\mu'(X_{k-1})-\pi_\gamma(\mu')]\right)}\right]\leq e^{u^2\varkappa (j-l-1)\gamma^{2}}
+\mathsf{E}\left[e^{-u\gamma\left(\sum_{k=l}^{j}[\mu'(X_{k-1})-\pi_\gamma(\mu')]\right)}\right]\leq e^{u^2\varkappa (j-l+1)\gamma^{2}}
 \]
@@
 \[
-\mathsf{E}\left[e^{-s\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})}\right]\leq e^{s^2\varkappa \gamma^2(j-l-1)}e^{-s\gamma (j-l-1)\pi_\gamma(\mu')}.
-\]
-\denis{Please finish}
-Furthermore, since $\mu\pi',\mu'\pi\in L^{1}(\mathbb{R})$, we have
-\[
-\pi(\mu')=\int\mu'(x)\pi(x)\,dx=-\int\mu(x)\pi'(x)\,dx=2\int\mu^{2}(x)\pi(x)\,dx>0
-\]
-yielding the final bound
-\[
-\zeta_{l,j}(s) \leq e^{s^2\kappa (j-l-1) \gamma^2}e^{-2s\gamma(j-l-1)\alpha}
-\]
-where $ \alpha = \int \mu^2(x)\pi(x)\,dx$.
-Finally
-\begin{align*}
-\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=1}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]  \lesssim
-\end{align*}
-Plugging into bound for conditional variance,
-\begin{align*}
-\mathsf{Var}\left[\sum_{p=l}^{q}\gamma_{p+1}\partial_{y_{l}}^{K}f\left(X_{p}\right)\Big|X_{l-1}\right] \leq
+\mathsf{E}\left[e^{-s\gamma\sum_{k=l}^{j}\mu'(X_{k-1})}\right]\leq e^{s^2\varkappa \gamma^2(j-l+1)}e^{-s\gamma (j-l+1)\pi_\gamma(\mu')}.
+\]
+Furthermore, since $\mu\pi',\mu'\pi\in L^{1}(\mathbb{R})$ and $\pi^{\prime}(x) = -\frac{1}{2}\pi(x)\mu(x)$, we have
+\[
+\pi(\mu')=\int\mu'(x)\pi(x)\,dx=-\int\mu(x)\pi'(x)\,dx=\frac{1}{2}\int\mu^{2}(x)\pi(x)\,dx>0
+\]
+Note also that \(\pi_\gamma(\mu^{\prime}) \geq \pi_\gamma(\mu^{\prime}) - C_1\sqrt{\gamma}\)
+yielding the bound
+\[
+\zeta_{l,j}(s) \leq e^{s^2\kappa (j-l+1) \gamma^2}e^{-s\gamma(j-l+1)\alpha}
+\]
+where $ \alpha = \frac{1}{2}\int \mu^2(x)\pi(x)\,dx - C_1\sqrt{\gamma} > 0$ for sufficiently small $\gamma$. Hence we obtain
+\begin{align*}
+\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4) \leq \sum_{p=j+1}^{q+1}e^{4\kappa(p-j)\gamma^2}e^{-2\gamma(p-j)\alpha} < \frac{e^{4\kappa\gamma^2 - 2\gamma\alpha}}{1-e^{4\kappa\gamma^2 - 2\gamma\alpha}}
+\end{align*}
+and thus the final bound follows:
+\begin{align*}
+\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=1}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]  \leq \frac{e^{6\kappa\gamma^2 - 3\gamma\alpha}}{\left(1-e^{4\kappa\gamma^2 - 2\gamma\alpha}\right)\left(1-e^{2\kappa\gamma^2 - \gamma\alpha}\right)}
+\end{align*}
+Plugging into variance bound,
+\begin{align*}
+\mathsf{Var}_x\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right] \leq
 \frac{A_{K}^{2}e^{2\kappa \Gamma^*}}{4\alpha^2(1-\gamma_1 B_\mu)^3}

### 2020-03-05 aap/Revision/vr-mcmc-revision-new.tex
lemma 8 update
update: squared bias

@@
 In the particular case $m=1$ we drop the subscript $y_1$ in that notation.
+For probability measures $\mu$ and $\nu$ on $\rset^d$ denote by $\tvnorm{\mu-\nu}$ the total variation distance between $\mu$ and $\nu$, that is,
+\[
+\tvnorm{\mu-\nu}=\sup_{A \in \mathcal{B}(\rset^d)}
+|\mu(A)-\nu(A)| \eqsp.
+\]
+where $\mathcal{B}(\rset^d)$ is a Borel $\sigma$-algebra of $\rset^d$.
+For a bounded Borel function $f: \rset^d \rightarrow \rset$ denote $\mathrm{osc}(f):=\sup_{x\in\mathbb R^d}f(x)-\inf_{x\in\mathbb R^d}f(x)$.
 
@@
 \end{description}
-First we analyse the biases of the estimators
-$\pi_n^N(f)$ and $\pi_{K,n}^N(f).$
+Let $\pi$ be the probability measure on $\rset^d$ with density $\pi(x)$ of the form \eqref{eq:stationary_distr};
+for $\gamma>0$, define the Markov kernel $Q_\gamma$ associated to one step of the ULA algorithm by
+\[
+\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(2 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{2\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\} \rmd y
+\]
+for any $A \in \mathcal{B}(\rset^d)$.
+Note that for a homogeneous Markov chain $\left(X^x_{p}\right)_{p \in \nset_0}$ of the form \eqref{eq:chain} with constant step size $\gamma$, $\mathsf{P}(X_n^x \in A) = \kerULA^n_{\gamma}(x,A)$.
+
 \paragraph{Squared bias:} Due to the martingale transform structure
@@
 $\pi_n^N(f)$ and $\pi_{K,n}^N(f)$ have the same conditional bias.
-%Notice that this remains true also when we substitute
-%the coefficients $a_{p,l,\mathbf{k}}$ in~\eqref{eq:29032018a5}
-%with some independent approximations $\widehat a_{p,l,\mathbf{k}}.$
-
-Denote by $\pi($ denotes the probability measure on $\mathbb R^d$
-with density $\pi$ of~\eqref{eq:stationary_distr};
-for $\gamma>0$, define the Markov kernel $Q_\gamma$ associated to one-step of the ULA algorithm by
-\[
-\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(2 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{2\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\} \rmd y.
-\]
-For a bounded Borel function~$f$,
-we can estimate the conditional bias
-similarly to~\cite[Section~4]{durmus:moulines:2017}:
+
+\begin{lem} 
+\label{lem:bias}
+Let $f$ be a bounded Borel function and $(X^x_{p})_{p \in \nset_0}$ be a Markov chain generated by ULA with constant step size $\gamma$. Then the squared bias satisfies
+\begin{equation}
+\label{eq:sq_bias}
+\left( \mathsf{E}[\rho(f)]-\pi(f)\right)^2 \lesssim \mathrm{osc}(f)^2 \left[\gamma + \frac{1}{n}\left(V(x) + \pi_{\gamma}(V)\right)\right]
+\end{equation}
+where $\lesssim$ means less up to a constant not depending on $n$ and $\gamma$.
+\end{lem}
+
+\begin{proof} We can estimate the conditional bias similarly to~\cite[Section~3]{durmus:moulines:2017}:
 \begin{equation}
@@
 \end{equation}
-where
-$\mathrm{osc}(f):=\sup_{x\in\mathbb R^d}f(x)-\inf_{x\in\mathbb R^d}f(x)$,
-$\tvnorm{\mu-\nu}$ denotes the total variation distance
-between probability measures $\mu$ and $\nu$, that is,
-\[
-\tvnorm{\mu-\nu}=\sup_{A\in\mathcal B(\mathbb R^d)}
-|\mu(A)-\nu(A)| \eqsp.
-\]
-Different specific upper bounds for the squared bias
-can be deduced from~\eqref{eq:06042018a2}
-using results of Section~3 in~\cite{durmus:moulines:2017}
-on bounds in the total variation distance.
-It is known  that under assumptions {\bf (H1)} and {\bf (H2)},   the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$.
-Moreover, as shown in Proposition, \tcr{(Appendix)}
-\eric{put a precise reference here}
+Under assumptions {\bf (H1)} and {\bf (H2)}, the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$.
+From \cite[Theorem~10]{durmus:moulines:2017} and Lemma~\ref{lem:v_ergodicity} it follows that
 \begin{equation}
 \label{eq:bias}
-\tvnorm{\delta_x\kerULA_{\gamma}^{n}-\pi_\gamma} \leq C_1\rho^{\gamma n}\left(V(x) + \pi_\gamma(V)\right), \quad \tvnorm{\pi-\pi_\gamma} \leq C_2 \sqrt{\gamma}
-\end{equation}
-for  $V(x) = 1 + \|x\|^2,$ some $\rho\in (0,1)$ and  constants $C_1,$ $C_2>0$ independent of $\gamma$  and $n.$
+\tvnorm{\pi - \pi_\gamma} < C_1\sqrt{\gamma}, \quad \tvnorm{\delta_x\kerULA_{\gamma}^{n}-\pi_\gamma} \leq C_2\rho^{\gamma n}\left(V(x) + \pi_\gamma(V)\right)
+\end{equation}
+for  $V(x) = 1 + \|x\|^2,$ some $\rho\in (0,1)$ and constants $C_1,C_2>0$ independent of $\gamma$  and $n$. Combining \ref{eq:bias} with the triangle inequality, we obtain \ref{eq:sq_bias}.
+\end{proof}
+
 \paragraph{Variance:} An upper bound for the variance of the classical estimator~\eqref{eq:29032018a2}
@@
 \begin{align*}
-\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=1}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]  \leq \frac{e^{6\kappa\gamma^2 - 3\gamma\alpha}}{\left(1-e^{4\kappa\gamma^2 - 2\gamma\alpha}\right)\left(1-e^{2\kappa\gamma^2 - \gamma\alpha}\right)}
+\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=1}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]  \leq \frac{e^{10\kappa\gamma^2 - 5\gamma\alpha}}{\left(1-e^{4\kappa\gamma^2 - 2\gamma\alpha}\right)^2\left(1-e^{2\kappa\gamma^2 - \gamma\alpha}\right)}
 \end{align*}
@@
 \mathsf{Var}_x\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right] \leq
-\frac{A_{K}^{2}e^{2\kappa \Gamma^*}}{4\alpha^2(1-\gamma_1 B_\mu)^3}
-\end{align*}
-\par
-The proof is completed.
+\frac{\gamma^3 A_{K}^{2} e^{10\kappa\gamma^2 - 5\gamma\alpha}}{\left(1-e^{4\kappa\gamma^2 - 2\gamma\alpha}\right)^2\left(1-e^{2\kappa\gamma^2 - \gamma\alpha}\right)(1-\gamma B_\mu)^3} \leq B_K
+\end{align*}
+for $\gamma$ small enough with $B_K$ not depending on $p,q$. The proof is completed.
 \end{proof}
@@
 
-We shall show that conditions {\bf (H1)} and {\bf (H2)} are sufficient to show that ULA kernel $R_\gamma$ satisfies the so-called drift condition:
+We shall show that conditions {\bf (H1)} and {\bf (H2)} are sufficient to show that ULA kernel $Q_\gamma$ satisfies the so-called drift condition:
 \begin{definition}
 \label{def:drift}
-We say that the Markov kernel $R_\gamma$ satisfies Foster-Lyapunov drift condition if there exist a measurable function $V: X \rightarrow [1; +\infty)$, real numbers $\lambda \in (0,1)$ and $C > 0$ such that for any $x \in X$,
-\begin{eqnarray*}
-RV(x) \leq \lambda^{\gamma} V(x) + \gamma C
+We say that the Markov kernel $Q_{\gamma}$ satisfies Foster-Lyapunov drift condition if there exist a measurable function $V: X \rightarrow [1; +\infty)$, real numbers $\lambda \in (0,1)$ and $C > 0$ such that for any $x \in X$,
+\begin{eqnarray*}
+Q_{\gamma}V(x) \leq \lambda^{\gamma} V(x) + \gamma C
 \end{eqnarray*}
@@
  Assume that the potential $U(x), x \in \mathbb{R}^d$ satisfies conditions {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$.
-Then the kernel $R_\gamma$ from \ref{eq:kernel} satisfies drift condition \ref{def:drift} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L^2}$ with drift function $V(x) = 1 + \|x\|^2$, constants $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = \frac{25K_1^2}{8} + 2d + m$ with $K_1$ from \ref{eq:H3}, $m$ from {\bf (H2)}.
+Then the kernel $Q_\gamma$ from \ref{eq:kernel} satisfies drift condition \ref{def:drift} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L^2}$ with drift function $V(x) = 1 + \|x\|^2$, constants $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = \frac{25K_1^2}{8} + 2d + m$ with $K_1$ from \ref{eq:H3}, $m$ from {\bf (H2)}.
 \end{lem}
@@
 $$
-R_{\gamma}V(x) = \int\limits_{\mathbb{R}^d}V(y)P_{\gamma}(x,dy) = \int\limits_{\mathbb{R}^d}(1+\|y\|^2)\frac{1}{(4\pi\gamma)^{\frac{d}{2}}}\exp{\left(-\frac{\|y-x+\gamma\nabla U(x)\|^2}{4\gamma}\right)}\,dy =
+Q_{\gamma}V(x) = \int\limits_{\rset^d}V(y)Q_{\gamma}(x,dy) = \int\limits_{\rset^d}(1+\|y\|^2)\frac{1}{(4\pi\gamma)^{\frac{d}{2}}}\exp{\left(-\frac{\|y-x+\gamma\nabla U(x)\|^2}{4\gamma}\right)}\,dy =
 $$

### 2020-03-05 aap/Revision/vr-mcmc-revision-new.tex
lemma 10 update

@@
 \begin{description}
-\item[(H1)][\textbf{Lipschitz continuity}] The potential $U$ is differentiable and $\nabla U$ is Lipschitz, that is, there exists $L_U < \infty$ such that
-\begin{eqnarray*}
-|\nabla U(x)-\nabla U(y)|\leq L_U|x-y|, \quad x,y\in \rset^d.
-\end{eqnarray*}
-\item[(H2)][\textbf{Convexity outside a ball}] There exist $K_U>0,$ $M_U>0$ and $m_U>0$ such that for any $x\not\in B_{K_U}(0)$  it holds
+\item[(H1)][\textbf{Lipschitz continuity}] The potential $U$ is differentiable and $\nabla U$ is Lipschitz, that is, there exists $L < \infty$ such that
+\begin{eqnarray*}
+\| \nabla U(x)-\nabla U(y) \| \leq L \|x-y\|, \quad x,y\in \rset^d.
+\end{eqnarray*}
+\item[(H2)][\textbf{Convexity outside a ball}] There exist $K>0$ and $m>0$ such that for any $x\not\in B_{K}(0)$  it holds
 \begin{equation*}
-\ps{D^2U(x)}{x} \geq (m_U/2)\|x\|^2.
+\ps{D^2U(x)}{x} \geq m\|x\|^2.
 \end{equation*}
@@
 for $\gamma>0$, define the Markov kernel $Q_\gamma$ associated to one step of the ULA algorithm by
-\[
+\begin{equation}
+\label{eq:ula_kernel}
 \kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(2 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{2\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\} \rmd y
-\]
+\end{equation}
 for any $A \in \mathcal{B}(\rset^d)$.
@@
 \end{equation}
-where $\lesssim$ means less up to a constant not depending on $n$ and $\gamma$.
+for $V(x) = 1 + \|x\|^2$, where $\lesssim$ means less up to a constant not depending on $n$ and $\gamma$.
 \end{lem}
@@
 \end{equation}
-for  $V(x) = 1 + \|x\|^2,$ some $\rho\in (0,1)$ and constants $C_1,C_2>0$ independent of $\gamma$  and $n$. Combining \ref{eq:bias} with the triangle inequality, we obtain \ref{eq:sq_bias}.
+for  $V(x) = 1 + \|x\|^2,$ some $\rho\in (0,1)$ and constants $C_1,C_2>0$ independent of $\gamma$  and $n$. Combining \eqref{eq:bias} with the triangle inequality, we obtain \eqref{eq:sq_bias}.
 \end{proof}
@@
 \section{Bounds for moments of  ULA}\label{sec:appendix}
-
-We shall show that conditions {\bf (H1)} and {\bf (H2)} are sufficient to show that ULA kernel $Q_\gamma$ satisfies the so-called drift condition:
+First let us notice that conditions {\bf (H1)} and {\bf (H2)} imply (\cite[Lemma~16]{brosse2018diffusion}) that for some $K_2 > 0$ it holds
+\begin{equation}
+\label{eq:H3}
+\ps{\nabla U(x)}{x} \geq \frac{m}{2}\|x\|^2, \quad \|x\| \geq K_2
+\end{equation}
+
+Now we shall show that conditions {\bf (H1)} and {\bf (H2)} are sufficient to show that ULA kernel $Q_\gamma$ satisfies the so-called drift condition:
 \begin{definition}
@@
 We say that the Markov kernel $Q_{\gamma}$ satisfies Foster-Lyapunov drift condition if there exist a measurable function $V: X \rightarrow [1; +\infty)$, real numbers $\lambda \in (0,1)$ and $C > 0$ such that for any $x \in X$,
-\begin{eqnarray*}
+\begin{equation}
+\label{eq:lyapunov}
 Q_{\gamma}V(x) \leq \lambda^{\gamma} V(x) + \gamma C
-\end{eqnarray*}
+\end{equation}
 \end{definition}
@@
  Assume that the potential $U(x), x \in \mathbb{R}^d$ satisfies conditions {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$.
-Then the kernel $Q_\gamma$ from \ref{eq:kernel} satisfies drift condition \ref{def:drift} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L^2}$ with drift function $V(x) = 1 + \|x\|^2$, constants $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = \frac{25K_1^2}{8} + 2d + m$ with $K_1$ from \ref{eq:H3}, $m$ from {\bf (H2)}.
+Then the kernel $Q_\gamma$ from \eqref{eq:ula_kernel} satisfies drift condition \eqref{eq:lyapunov} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L}$ with drift function $V(x) = 1 + \|x\|^2$,   $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = \frac{25K_2^2}{8} + d + m$ with $K_2$ from \eqref{eq:H3}, $m_U$ from {\bf (H2)}.
 \end{lem}
 
-\begin{proof} Consider $V(x) = 1 + \|x\|^2$, then
-$$
-Q_{\gamma}V(x) = \int\limits_{\rset^d}V(y)Q_{\gamma}(x,dy) = \int\limits_{\rset^d}(1+\|y\|^2)\frac{1}{(4\pi\gamma)^{\frac{d}{2}}}\exp{\left(-\frac{\|y-x+\gamma\nabla U(x)\|^2}{4\gamma}\right)}\,dy =
-$$
-
-$$
-=1 + \frac{1}{(4\pi\gamma)^{\frac{d}{2}}}\int\limits_{\mathbb{R}^d}\|z + x - \gamma \nabla U(x)\|^2 \exp{\left(-\frac{\|z\|^2}{4\gamma}\right)}\,dz
-$$
-Let us first consider the case $x \notin B(0,K_1)$. Note that
-$$
-\|z+x-\gamma \nabla U(x)\|^2 = \|z\|^2 + 2 \langle z, x - \gamma \nabla U(x) \rangle + \gamma^2 \|x - \gamma \nabla U(x)\|^2
-$$
-and the linear term vanishes after integration, moreover, for $Z \sim \mathcal{N}(0,2\gamma I_d)$, it holds $\mathbb{E}\|Z\|^2 = 2\gamma d$.
-It remains to notice that due to {\bf H1} and {\bf H2} (which implies \ref{eq:H3}),
-$$
-\|x - \gamma \nabla U(x) \|^2 = \|x\|^2 - 2\gamma \langle \nabla U(x), x\rangle + \gamma^2 \|\nabla U(x)\|^2 \leq
-$$
-
-$$
- \leq (1-\gamma m)\|x\|^2 + 2\gamma^2 L^2 \|x\|^2
-$$
-Thus, plugging everything into expression for $R_{\gamma}V$ and using $\gamma < \frac{m}{4L}$, we obtain
-$$
-R_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + 2\gamma d + (\gamma m - 2\gamma^2 L^2) \leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma (2d + m)
-$$
-Now let $x \in B(0,K_1)$. Then simply using $\|x-\gamma \nabla U(x)\|^2 \leq 2(1+L\gamma)^2\|x\|^2$, we obtain
-$$
-R_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma\left((m - 2\gamma L^2)(1+\|x\|^2) + 2d + 2(1+L\gamma)^2\|x\|^2\right) \leq
-$$
-
-$$
-\leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma(\frac{25K_1^2}{8} + 2d + m)
+\begin{proof} Let $V(x) = 1 + \|x\|^2$, then
+$$
+Q_{\gamma}V(x) = \int\limits_{\rset^d}V(y)Q_{\gamma}(x,dy) = \int\limits_{\rset^d}\frac{(1+\|y\|^2)}{(2\pi\gamma)^{\frac{d}{2}}}\exp{\left(-\frac{\|y-x+\gamma\nabla U(x)\|^2}{2\gamma}\right)}\,dy =
+$$
+
+$$
+=1 + \frac{1}{(2\pi\gamma)^{\frac{d}{2}}}\int\limits_{\mathbb{R}^d}\|z + x - \gamma \nabla U(x)\|^2 \exp{\left(-\frac{\|z\|^2}{2\gamma}\right)}\,dz
+$$
+Let us first consider the case $x \notin B(0,K_2)$. Note that
+$$
+\|z+x-\gamma \nabla U(x)\|^2 = \|z\|^2 + 2 \langle z, x - \gamma \nabla U(x) \rangle + \|x - \gamma \nabla U(x)\|^2
+$$
+and the linear term vanishes after integration, moreover, for $Z \sim \mathcal{N}(0,\gamma I_d)$, it holds $\mathbb{E}\|Z\|^2 = \gamma d$. It remains to notice that due to \eqref{eq:H3},
+\[
+\|x - \gamma \nabla U(x) \|^2 = \|x\|^2 - 2\gamma \langle \nabla U(x), x\rangle + \gamma^2 \|\nabla U(x)\|^2 \leq \left(1 - \gamma m + 2\gamma^2 L^2\right)\|x\|^2
+\]
+Since $\gamma < \frac{m}{4L}$, we obtain
+$$
+Q_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma d + (\gamma m - 2\gamma^2 L^2) \leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma (d + m)
+$$
+Now let $x \in B(0,K_2)$. Then simply using $\|x-\gamma \nabla U(x)\|^2 \leq 2(1+L\gamma)^2\|x\|^2$, we obtain
+$$
+Q_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma\left((m - 2\gamma L^2)(1+\|x\|^2) + d + 2(1+L\gamma)^2\|x\|^2\right) \leq
+$$
+
+$$
+\leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma(\frac{25K_2^2}{8} + d + m)
 $$

### 2020-03-10 aap/Revision/vr-mcmc-revision-new.tex
theorem 4 update

@@
 \\
-&=& \sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left(1+\frac{N-l}{n}\right)A_{N+n-l,k}(X^x_{l-1})\phi_{k}\left(\xi_{l}\right),
+&=& \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\bar{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
 \label{eq:29032018a5}
 \end{eqnarray}
-where
-\begin{equation}
-\label{eq:definition-A-s,k}
-A_{s,k}(x)=\frac{1}{s}\sum_{r=1}^{s}\bar a_{r,k}(x).
-\end{equation}
 Since \(X^x_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0\), we obtain
 \[
-\PE[A_{N+n-l,k}(X^x_{l-1})\phi_{k}\left(\xi_{l}\right)] = \mathsf{E}[A_{N+n-l,k}(X^x_{l-1})\CPE{\phi_{k}\left(\xi_{l}\right)}{\mathcal{G}_{l-1}]} = 0
-\]
-and hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
+\PE[g(X^x_{l-1})\phi_{k}\left(\xi_{l}\right)] = \mathsf{E}[g(X^x_{l-1})\CPE{\phi_{k}\left(\xi_{l}\right)}{\mathcal{G}_{l-1}]} = 0
+\]
+for any measurable function $g$  and hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
 The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
@@
 \label{eq:expression-variance}
-\PVar[\pi_{K,n}^N(f)]\leq\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})]
-\end{equation}
+\PVar_x[\pi_{K,n}^N(f)]\leq\sum_{k=K+1}^{\infty}\frac{1}{n^2}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})],
+\end{equation}
+where 
+\begin{eqnarray}
+\label{eq:definition-A-s,k}
+A_{q,k}(y)=\sum_{r=0}^q \bar{a}_{r,k}(y),\quad q\in \mathbb{N}
+\end{eqnarray}
 and $\mathsf{Var}[\pi_{K,n}^N(f)]$ is small provided that the coefficients $A_{s,k}$ decay fast enough as $k\to \infty.$
@@
 \begin{align}
-M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{l=1}^{n}\Bigl(1-\frac{l}{n}\Bigr)A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})
+M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\frac{1}{n}\sum_{l=1}^{n}A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})
 \label{eq:29032018a5}
@@
 \label{eq:var-bound}
-\PVar\left(\pi_{K,n}^{N}(f)\right)\lesssim n^{-1} \gamma^{K-2} \eqsp,
+\PVar\left(\pi_{K,n}^{N}(f)\right)\lesssim n^{-1} \gamma^{K-1} \eqsp,
 \end{eqnarray}
+provided $n\gamma^2$ is bounded.
 %where \tcr{\(\lesssim\) stands for inequality up to a constant not depending  on $\gamma,$ \(n\) and \(N.\)}
@@
 \end{eqnarray}
-where $F_s(x,Z_1,\ldots,Z_s)=s^{-1}\sum_{r=0}^s f(X^x_r)$ and $\partial_{Z_1}^{\mathbf{k}'}$
+where $F_s(x,Z_1,\ldots,Z_s)=\sum_{r=0}^s f(X^x_r)$ and $\partial_{Z_1}^{\mathbf{k}'}$
 stands for a weak partial derivative of the functional $F_s$ that also can be viewed as discretised version of Malliavin derivative. From \eqref{eq:Ask-part}, we derive
@@
 \left(\frac{\gamma}{2}\right)^{|I|K}
-\PVar \left( s^{-1} \sum\nolimits_{p=1}^{s}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X^x_{p}\right)
+\PVar_x \left( \sum_{p=1}^{s}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X_{p}\right)
 \right),
@@
 \]
- Finally we show via  the Poincare inequality that under smoothness assumption \eqref{eq:smooth-mu} it holds
- \eric{Since we do not state was $W_s$ is, it is a bit difficult to guess why we are using a Poincare inequality there}
+ Finally, we show  that under our smoothness assumption \eqref{eq:smooth-mu} it holds
 \begin{equation*}
-\PVar \left(s^{-1} \sum\nolimits_{p=1}^{s}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X^x_{p}\right)
-\right)\leq W_s(x)
+\PVar \left(\sum_{p=1}^{q}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X^x_{p}\right)
+\right)\leq C(1+V(x))
 \end{equation*}
-for all $x$ and some family of functions $(W_s)_{s\geq 1}$ such that  the sum $\sum_{s=1}^n \mathsf{E}[W_s(X_{N+n-s-1})]$ is bounded uniformly in $n\in \mathbb{N}.$
-\eric{Speak about truncation here.}
-
+for some constant $C$ not depending on $q$ and $V(x) = 1 + \|x\|^2$.
 \section{Numerical results}
@@
 \,
-\PE\left[
-s^{-1} \sum\nolimits_{r=1}^{s}\{\partial_{1}^{\mathbf{k}'} f(X^x_r)-\PE[\partial_{1}^{\mathbf{k}'}f(X^x_r)]\}\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_1)\right]
+\PE\left[ \sum\nolimits_{r=1}^{s}\{\partial_{1}^{\mathbf{k}'} f(X^x_r)-\PE[\partial_{1}^{\mathbf{k}'}f(X^x_r)]\}\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_1)\right]
 \end{equation*}
@@
 =
-\left\{\PE\left[\frac{1}{s}\sum_{p=1}^{s}\left\{\partial_{1}^{\mathbf r}f\left(X^x_{p}\right)-\PE\left[\partial_{1}^{\mathbf r}f\left(X^x_{p}\right)\right]\right\}\mathbf{H}_{\mathbf{q}}(Z_{1})\right]\right\}^{2}.
+\left\{\PE\left[\sum_{p=1}^{s}\left\{\partial_{1}^{\mathbf r}f\left(X^x_{p}\right)-\PE\left[\partial_{1}^{\mathbf r}f\left(X^x_{p}\right)\right]\right\}\mathbf{H}_{\mathbf{q}}(Z_{1})\right]\right\}^{2}.
 \end{equation*}
@@
 \left(\frac{\gamma}{2}\right)^{|I|K}
-\PVar\left(s^{-1}\sum\nolimits_{p=1}^{s}\partial_{1}^{\mathbf{K}_I}f\left(X^x_{p}\right)
+\PVar\left(\sum_{p=1}^{s}\partial_{1}^{\mathbf{K}_I}f\left(X^x_{p}\right)
 \right)
@@
 \begin{eqnarray*}
-\PVar\left(s^{-1}\sum\nolimits_{p=1}^{s}\partial_{1}^{\mathbf{K}_I}f\left(X^x_{p}\right)
-\right)\leq W_s(x)
-\end{eqnarray*}
-for all $x$ and some family of functions $(W_s)_{s\geq 1}$ such that  the sum $\sum_{s=1}^n \mathsf{E}[W_s(X_{N+n-s-1})]$ is bounded uniformly in $n\in \mathbb{N}.$
-
+\PVar\left(\sum_{p=1}^{q}\partial_{1}^{\mathbf{K}_I}f\left(X^x_{p}\right)
+\right)\leq C(1+V(x))
+\end{eqnarray*}
+for all $x$ and some constant $C$  not depending on $q.$
 To keep the notational burden at a reasonable level, we present the proof only in one-dimensional case.
 Multidimensional extension is straightforward but requires involved notations.
-First we need to prove several auxiliary results.
+First, we need to prove several auxiliary results.
 \begin{lem}\label{lem:06062018a1}
@@
 \end{proof}
-
 In what follows, we use the notation
@@
 \end{equation}
-
 The assumption \eqref{eq:smooth-mu} implies that  $|\mu'(x)|\leq B_\mu$ for some constant $B_\mu>0$ and all $x\in \mathbb{R}^d$. Without loss of generality we suppose that $\gamma B_\mu<1.$
@@
 $$
-\mathsf{Var}_x\left[\sum_{p=1}^{s}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right]\le B_K\quad\text{a.s.},
-$$
-where $B_K$ is a deterministic bound that does not depend on $l$ and~$q$.
+\mathsf{Var}_x\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right]\le  \gamma^{-1} C_K (1+V(x)),
+$$
+where $C_K$ is a constant that does not depend on $q$ and $\gamma.$ 
 \end{lem}
-
 \begin{proof}
@@
 can be viewed as a deterministic function of
-$X_{0},Z_1,Z_{2},\ldots,Z_q$
+$x,Z_1,Z_{2},\ldots,Z_q$
 $$
 \sum_{p=1}^q  \partial_{y_1}^K f(X_p)
-=F(X_{0},Z_1,Z_{2},\ldots,Z_q).
+=F(x,Z_1,Z_{2},\ldots,Z_q).
 $$
@@
 \le\mathsf E_x\left[
-\|\nabla_Z F(X_{0},Z_1,Z_{2},\ldots,Z_q)\|^2
+\|\nabla_Z F(x,Z_1,Z_{2},\ldots,Z_q)\|^2
 \right],
@@
 $$
-\mathsf{Var}_x\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right]\leq\gamma\sum_{j=1}^{q}\mathsf{E}_x\left[\left(\sum_{p=1}^{q}\gamma\partial_{y_{j}}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right)^{2}\right].
+\mathsf{Var}_x\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right]\leq\gamma^2\sum_{j=1}^{q}\mathsf{E}_x\left[\left(\sum_{p=1}^{q}\partial_{y_{j}}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right)^{2}\right].
 $$
@@
 \begin{equation}\label{eq:10062018a2}
-\mathsf{Var}_x\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right]\leq\gamma\sum_{j=1}^{q}\mathsf{E}_x\left[\left(\sum_{p=j}^{q}\gamma \partial_{y_{j}}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right)^{2}\right].
+\mathsf{Var}_x\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right]\leq\gamma^2\sum_{j=1}^{q}\mathsf{E}_x\left[\left(\sum_{p=j}^{q}\partial_{y_{j}}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right)^{2}\right].
 \end{equation}
@@
 &\le
-\gamma^3 A_{K}^{2}\sum_{j=1}^{q}
+\gamma^2 A_{K}^{2}\sum_{j=1}^{q}
 \mathsf{E}\left(\sum_{p=j}^{q}\prod_{k=2}^{p}\alpha_{k}\right)^{2}
@@
 &\le
-\frac{\gamma^3 A_{K}^{2}}{(1-\gamma B_\mu)^2}
+\frac{\gamma^2 A_{K}^{2}}{(1-\gamma B_\mu)^2}
 \mathsf{E}\sum_{j=1}^{q}
@@
 &\le
-\frac{\gamma^3 A_{K}^{2}}{(1-\gamma B_\mu)^3}
+\frac{\gamma^2 A_{K}^{2}}{(1-\gamma B_\mu)^3}
 \mathsf{E}\sum_{j=1}^{q}
@@
 \begin{align*}
-\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=l}^{j}\alpha_{k}\left( \sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right] & \leq\sum_{j=1}^{q}\left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left\Vert \sum_{p=j+1}^{q+1} \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}^{2}\\
+\mathsf{E}_x\left[\sum_{j=1}^{q}\prod_{k=l}^{j}\alpha_{k}\left( \sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right] & \leq\sum_{j=1}^{q}\left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left\Vert \sum_{p=j+1}^{q+1} \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}^{2}\\
  & \leq\sum_{j=1}^{q} \left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left(\sum_{p=j+1}^{q+1}\left\Vert \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}\right)^{2}.
@@
 \[
-\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=1}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]\leq\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \left(\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4)\right)^{2},
+\mathsf{E}_x\left[\sum_{j=1}^{q}\prod_{k=1}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]\leq\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \left(\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4)\right)^{2},
 \]
@@
 \[
-\zeta_{l,j}(u)=\mathsf{E}\left[e^{-u\gamma\sum_{k=l}^{j}\mu'(X_{k-1})}\right],\quad u>0.
-\]
-From Theorem~\ref{thm:ula_const_size} in Appendix~\ref{sec:appendix}, it follows that
-\[
-\mathsf{E}\left[e^{-u\gamma\left(\sum_{k=l}^{j}[\mu'(X_{k-1})-\pi_\gamma(\mu')]\right)}\right]\leq e^{u^2\varkappa (j-l+1)\gamma^{2}}
-\]
-with constant $\varkappa = ...$, where we used the fact that $\mu^{\prime}(x)$ is bounded. Then we have
-\[
-\mathsf{E}\left[e^{-s\gamma\sum_{k=l}^{j}\mu'(X_{k-1})}\right]\leq e^{s^2\varkappa \gamma^2(j-l+1)}e^{-s\gamma (j-l+1)\pi_\gamma(\mu')}.
+\zeta_{l,j}(u)=\mathsf{E}_x\left[e^{-u\gamma\sum_{k=l}^{j}\mu'(X_{k-1})}\right],\quad u>0.
+\]
+Note that $\mu^{\prime}(x)$ is bounded due  our assumptions and from Theorem~1 in \cite{delyon1999small} and Lemma~\ref{lem:v_ergodicity} in Appendix~\ref{sec:appendix} it follows that
+\[
+\mathsf{E}_x\left[e^{-u\gamma\left(\sum_{k=l}^{j}[\mu'(X_{k-1})-\pi_\gamma(\mu')]\right)}\right]\leq C(1+V(x)),\quad u\gamma\leq \gamma_0
+\]
+for some constants $\gamma_0$ and $C>0$  depending only on $\gamma_0$, provided that $n\gamma^2$ stays bounded. Then we have
+\[
+\zeta_{l,j}(u)=\mathsf{E}\left[e^{-s\gamma\sum_{k=l}^{j}\mu'(X_{k-1})}\right]\leq  C(1+V(x))e^{-s\gamma (j-l+1)\pi_\gamma(\mu')}.
 \]
@@
 \[
-\zeta_{l,j}(s) \leq e^{s^2\kappa (j-l+1) \gamma^2}e^{-s\gamma(j-l+1)\alpha}
+\zeta_{l,j}(s) \leq  C(1+V(x))e^{-s\gamma(j-l+1)\alpha}
 \]
@@
 \begin{align*}
-\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4) \leq \sum_{p=j+1}^{q+1}e^{4\kappa(p-j)\gamma^2}e^{-2\gamma(p-j)\alpha} < \frac{e^{4\kappa\gamma^2 - 2\gamma\alpha}}{1-e^{4\kappa\gamma^2 - 2\gamma\alpha}}
-\end{align*}
-and thus the final bound follows:
-\begin{align*}
-\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=1}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]  \leq \frac{e^{10\kappa\gamma^2 - 5\gamma\alpha}}{\left(1-e^{4\kappa\gamma^2 - 2\gamma\alpha}\right)^2\left(1-e^{2\kappa\gamma^2 - \gamma\alpha}\right)}
-\end{align*}
-Plugging into variance bound,
-\begin{align*}
-\mathsf{Var}_x\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right] \leq
-\frac{\gamma^3 A_{K}^{2} e^{10\kappa\gamma^2 - 5\gamma\alpha}}{\left(1-e^{4\kappa\gamma^2 - 2\gamma\alpha}\right)^2\left(1-e^{2\kappa\gamma^2 - \gamma\alpha}\right)(1-\gamma B_\mu)^3} \leq B_K
-\end{align*}
-for $\gamma$ small enough with $B_K$ not depending on $p,q$. The proof is completed.
+\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4) \leq (C(1+V(x)))^{1/4}\sum_{p=j+1}^{q+1}e^{-\gamma(p-j)\alpha} \leq  (C(1+V(x)))^{1/4}\frac{1}{1-e^{- \gamma\alpha}}
+\end{align*}
+and
+\begin{align*}
+\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \leq (C(1+V(x)))^{1/2}\sum_{j=0}^{q}e^{-\gamma j\alpha} \leq  (C(1+V(x)))^{1/2}\frac{ 1}{1-e^{- \gamma\alpha}}.
+\end{align*}
+Thus the final bound follows:
+\begin{align*}
+\mathsf{Var}_x\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right] \leq \gamma^{-1} C_K (1+V(x))
+\end{align*}
+for $\gamma$ small enough with $C_K$ not depending on $q$. The proof is completed.
 \end{proof}
@@
 $$
-d_V(\delta_xQ_{\boldsymbol{\gamma}}^{1,n},\pi_\gamma) \leq C\rho^n\left(V(x) + \pi_\gamma(V)\right)
+d_V(\delta_xQ_{\gamma}^{1,n},\pi_\gamma) \leq C\rho^n\left(V(x) + \pi_\gamma(V)\right)
 $$

### 2020-03-10 aap/Revision/vr-mcmc-revision-new.tex
theorem 4 update

@@
 \paragraph{Variance:} An upper bound for the variance of the classical estimator~\eqref{eq:29032018a2}
-under {\bf (H1)} and {\bf (H2)} is derived in Section.  In particular, for any bounded function $f,$
+under {\bf (H1)} and {\bf (H2)} is derived in Section ??. \denis{Insert} In particular, for any bounded function $f,$
 \begin{equation}
@@
 \end{equation}
-where $\lesssim$ stands for inequality up to a constant depending on constants $L_U,$ $m_U,$  $M_U$ and $\|f\|_\infty.$
+where $\lesssim$ stands for inequality up to a constant depending on constants $L,$ $m,$  $M$ and $\|f\|_\infty.$
 One of the main results of this paper is the following  upper bound for the variance of $\pi_{K,n}^{N}(f).$
@@
 \label{eq:Ask-part}
-A_{s,\mathbf{k}}(x)=\gamma^{|\mathbf{k}'|/2}\frac{\sqrt{(\mathbf{k}-\mathbf{k}')!}}{\sqrt{\mathbf{k}!}}\mathsf{E}\left[ \partial_{Z_1}^{\mathbf{k}'} F(x,Z_1,\ldots,Z_s) \mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_{1})\right]
+A_{q,\mathbf{k}}(x)=\gamma^{|\mathbf{k}'|/2}\frac{\sqrt{(\mathbf{k}-\mathbf{k}')!}}{\sqrt{\mathbf{k}!}}\mathsf{E}\left[ \partial_{Z_1}^{\mathbf{k}'} F(x,Z_1,\ldots,Z_q) \mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_{1})\right]
 \end{eqnarray}
-where $F_s(x,Z_1,\ldots,Z_s)=\sum_{r=0}^s f(X^x_r)$ and $\partial_{Z_1}^{\mathbf{k}'}$
+where $F_q(x,Z_1,\ldots,Z_q)=\sum_{r=0}^q f(X^x_r)$ and $\partial_{Z_1}^{\mathbf{k}'}$
 stands for a weak partial derivative of the functional $F_s$ that also can be viewed as discretised version of Malliavin derivative. From \eqref{eq:Ask-part}, we derive
 \begin{eqnarray*}
-\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1} A^2_{s,\mathbf{k}}(x)\leq \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
+\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1} A^2_{q,\mathbf{k}}(x)\leq \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
 \left(\frac{\gamma}{2}\right)^{|I|K}
-\PVar_x \left( \sum_{p=1}^{s}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X_{p}\right)
+\PVar \left( \sum_{p=1}^{q}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X_{p}^x\right)
 \right),
@@
 \PVar \left(\sum_{p=1}^{q}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X^x_{p}\right)
-\right)\leq C(1+V(x))
+\right)\leq C_{K}\gamma^{-1}(1+V(x)), \quad V(x) = 1 + \|x\|^2
 \end{equation*}
-for some constant $C$ not depending on $q$ and $V(x) = 1 + \|x\|^2$.
+for some constant $C_K$ not depending on $q$ and $\gamma$.
 \section{Numerical results}

### 2020-03-11 aap/Revision/vr-mcmc-revision-new.tex
update

@@
 \label{eq:expression-variance}
-\PVar_x[\pi_{K,n}^N(f)]\leq\sum_{k=K+1}^{\infty}\frac{1}{n^2}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})],
+\PVar_x[\pi_{K,n}^N(f)]=\sum_{k=K+1}^{\infty}\frac{1}{n^2}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})],
 \end{equation}
@@
 In the next section   we provide a detailed   analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
+\section{Estimation of coefficients}
+The coefficients \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. This amounts to solve a nonparametric regression problem.
+We present a generic regression algorithm.
+Algorithm starts with estimating the functions \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) for $r=N+1,\ldots,N+n.$ We first generate \(\NtrainPath\)  paths  of the chain \(X\)
+(the so-called ``training paths''):
+\[
+\TrainSet=\Bigl\{(X^{(s)}_{0},\ldots,X^{(s)}_{N+n}),\quad s=1,\ldots,\NtrainPath\Bigr\}
+\]
+with $X^{(s)}_{0}=x,$ $s=1,\ldots,T.$
+The regression algorithm proceeds with estimating the  functions \((\bar Q_{l})\) by solving  the least squares optimization problems
+\begin{equation}\label{eq:06042018a1}
+\widehat{Q}_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
+\end{equation}
+for \(r=1,\ldots, n-1,\)
+where \(\Psi\) is a  class of  functions on \(\mathbb{R}^d.\)
+Next  we estimate the coefficients \( \bar a_{l,k}\) using
+the formula
+\begin{equation}
+\label{eq:a_est_int}
+\widehat a_{l,k}(x)=\mathsf{E}\left[\left.\phi_k\left(\xi\right)\widehat Q_{l}\left(\Phi(x,\xi)\right)\right | \TrainSet\right] \,,
+\end{equation}
+where $\xi$ is a random variable independent from $\mathcal{T}$ with law $P_\xi,$
+$\Phi_l$ is defined in \eqref{eq:chain_gen} and the expectation is taken according to known distribution \(P_\xi\).
+Conditioning on $\TrainSet$ underlines that estimation of $\widehat Q_{l}$ derived from training trajectories.
+In many cases integration can be done in closed analytical form.
+For example, in the case of ULA algorithm we have \(P_\xi$ is the multivariate normal distribution $ \mathcal{N}(0,I_d)\) and \(\Phi_l(x,\xi)=x-\gamma_{l}\mu(x)+\sqrt{\gamma_{l}}\xi.\) Therefore if we use polynomials to approximate \((\bar Q_{l}),\) then \((\widehat a_{l,k})\) can be even computed in closed form; implementations details are provided in  Section~\ref{sec:num}.
+Upon estimating  the coefficients \((\widehat a_{l,k})\), one can construct an empirical estimate  of \(M_{K,n}^N(f)\) in the form
+\begin{eqnarray*}
+\widehat M_{K,n}^N(f) =\frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\widehat{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
+\end{eqnarray*}
+Obviously, \(\mathsf{E}[\widehat M_{K,n}^N(f)|\TrainSet]=0\) and \(\widehat M_{K,n}^N(f)\) is indeed a valid control variate in that it
+does not introduce any bias. By the Jensen inequality and orthonormality of \((\phi_k)_{k\geq 0},\)
+\begin{multline*}
+\mathsf{E}_x\left[\left.\left|\widehat{M}_{K,n}^{N}(f)-M_{K,n}^{N}(f)\right|^{2}\right | \TrainSet \right]
+\\
+\leq \sum_{k=1}^{K}\frac{1}{n^2}\sum_{l=1}^{n}\PE\left[\left.|A_{n-l,k}(X^x_{N+l-1})-\widehat A_{n-l,k}(X^x_{N+l-1})|^2\right | \TrainSet \right].
+\end{multline*}
+
 \section{Analysis of variance reduced ULA}
@@
 \end{thm}
-Let us sketch the main steps of the proof. First using integration by parts we prove that
+Let us sketch the main steps of the proof. First using integration by parts, we prove that
 \begin{eqnarray}
 \label{eq:Ask-part}
-A_{q,\mathbf{k}}(x)=\gamma^{|\mathbf{k}'|/2}\frac{\sqrt{(\mathbf{k}-\mathbf{k}')!}}{\sqrt{\mathbf{k}!}}\mathsf{E}\left[ \partial_{Z_1}^{\mathbf{k}'} F(x,Z_1,\ldots,Z_q) \mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_{1})\right]
+A_{q,\mathbf{k}}(x)=\gamma^{|\mathbf{k}'|/2}\frac{\sqrt{(\mathbf{k}-\mathbf{k}')!}}{\sqrt{\mathbf{k}!}}\mathsf{E}_x\left[ \partial_{Z_1}^{\mathbf{k}'} F(x,Z_1,\ldots,Z_q) \mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_{1})\right],
 \end{eqnarray}
 where $F_q(x,Z_1,\ldots,Z_q)=\sum_{r=0}^q f(X^x_r)$ and $\partial_{Z_1}^{\mathbf{k}'}$
-stands for a weak partial derivative of the functional $F_s$ that also can be viewed as discretised version of Malliavin derivative. From \eqref{eq:Ask-part}, we derive
+stands for a weak partial derivative of the functional $F_s$ that also can be viewed as discretised version of Malliavin derivative. From \eqref{eq:Ask-part} we derive
 \begin{eqnarray*}

### 2020-03-11 aap/Revision/vr-mcmc-revision-new.tex
update

@@
 In many cases integration can be done in closed analytical form.
-For example, in the case of ULA algorithm we have \(P_\xi$ is the multivariate normal distribution $ \mathcal{N}(0,I_d)\) and \(\Phi_l(x,\xi)=x-\gamma_{l}\mu(x)+\sqrt{\gamma_{l}}\xi.\) Therefore if we use polynomials to approximate \((\bar Q_{l}),\) then \((\widehat a_{l,k})\) can be even computed in closed form; implementations details are provided in  Section~\ref{sec:num}.
+For example, in the case of ULA algorithm we have \(P_\xi$ is the multivariate normal distribution $ \mathcal{N}(0,I_d)\) and \(\Phi(x,\xi)=x-\gamma\mu(x)+\sqrt{\gamma}\xi.\) Therefore if we use polynomials to approximate \((\bar Q_{l}),\) then \((\widehat a_{l,k})\) can be even computed in closed form; implementations details are provided in  Section~\ref{sec:num}.
 Upon estimating  the coefficients \((\widehat a_{l,k})\), one can construct an empirical estimate  of \(M_{K,n}^N(f)\) in the form

### 2020-03-11 aap/Revision/vr-mcmc-revision-new.tex
update

@@
 \author{\snm{S. Samsonov}\thanksref{b}\ead[label=e3]{svsamsonov@hse.ru}}
-\and
-\author{\snm{N. Shagadatov}\thanksref{b}\ead[label=e4]{nshagadatov@edu.hse.ru}}
+%\and
+%\author{\snm{N. Shagadatov}\thanksref{b}\ead[label=e4]{nshagadatov@edu.hse.ru}}
 
@@
 \label{eq:expression-variance}
-\PVar_x[\pi_{K,n}^N(f)]=\sum_{k=K+1}^{\infty}\frac{1}{n^2}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})],
+\PVar_x[\pi_{K,n}^N(f)]=\frac{1}{n^2}\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})],
 \end{equation}
@@
 for \(r=1,\ldots, n-1,\)
-where \(\Psi\) is a  class of  functions on \(\mathbb{R}^d.\)
+where \(\Psi\) is a  class of  real-valued functions on \(\mathbb{R}^d.\)
 Next  we estimate the coefficients \( \bar a_{l,k}\) using
@@
 where $\xi$ is a random variable independent from $\mathcal{T}$ with law $P_\xi,$
-$\Phi_l$ is defined in \eqref{eq:chain_gen} and the expectation is taken according to known distribution \(P_\xi\).
+$\Phi$ is defined in \eqref{eq:chain_gen} and the expectation is taken according to known distribution \(P_\xi\).
 Conditioning on $\TrainSet$ underlines that estimation of $\widehat Q_{l}$ derived from training trajectories.
@@
 \\
-\leq \sum_{k=1}^{K}\frac{1}{n^2}\sum_{l=1}^{n}\PE\left[\left.|A_{n-l,k}(X^x_{N+l-1})-\widehat A_{n-l,k}(X^x_{N+l-1})|^2\right | \TrainSet \right].
+\leq \frac{1}{n^2}\sum_{k=1}^{K}\sum_{l=1}^{n}\PE\left[\left.|A_{n-l,k}(X^x_{N+l-1})-\widehat A_{n-l,k}(X^x_{N+l-1})|^2\right | \TrainSet \right].
 \end{multline*}
@@
 $$
-\mathsf{Var}_x\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right]\le  \gamma^{-1} C_K (1+V(x)),
+\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right]\le  \gamma^{-1} C_K (1+V(x)),
 $$
@@
 The expression
-$\sum_{p=1}^q \partial_{y_1}^K f(X_p)$
+$\sum_{p=1}^q \partial_{y_1}^K f(X_p^x)$
 can be viewed as a deterministic function of
@@
 $$
-\sum_{p=1}^q  \partial_{y_1}^K f(X_p)
+\sum_{p=1}^q  \partial_{y_1}^K f(X_p^x)
 =F(x,Z_1,Z_{2},\ldots,Z_q).
@@
 $$
-\mathsf{Var}_x\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right]
+\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right]
 \le\mathsf E_x\left[
@@
 $$
-\mathsf{Var}_x\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right]\leq\gamma^2\sum_{j=1}^{q}\mathsf{E}_x\left[\left(\sum_{p=1}^{q}\partial_{y_{j}}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right)^{2}\right].
+\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right]\leq\gamma^2\sum_{j=1}^{q}\mathsf{E}\left[\left(\sum_{p=1}^{q}\partial_{y_{j}}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right)^{2}\right].
 $$
 It is straightforward to check that
-$\partial_{y_j}\partial_{y_1}^K f(X_p)=0$
+$\partial_{y_j}\partial_{y_1}^K f(X_p^x)=0$
 whenever $p<j$. Therefore, we get
 \begin{equation}\label{eq:10062018a2}
-\mathsf{Var}_x\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right]\leq\gamma^2\sum_{j=1}^{q}\mathsf{E}_x\left[\left(\sum_{p=j}^{q}\partial_{y_{j}}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right)^{2}\right].
+\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right]\leq\gamma^2\sum_{j=1}^{q}\mathsf{E}\left[\left(\sum_{p=j}^{q}\partial_{y_{j}}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right)^{2}\right].
 \end{equation}
@@
 \[
-\partial_{y_{1}}^{K}f\left(X_{p}\right)=\sum\frac{K!}{m_{1}!\ldots m_{K}!}f^{(m_{1}+\ldots+m_{K})}(X_{p})\prod_{k=1}^{K}\left(\frac{\partial_{y_{1}}^{k}X_{p}}{k!}\right)^{m_{k}},
+\partial_{y_{1}}^{K}f\left(X_{p}^x\right)=\sum\frac{K!}{m_{1}!\ldots m_{K}!}f^{(m_{1}+\ldots+m_{K})}(X_{p}^x)\prod_{k=1}^{K}\left(\frac{\partial_{y_{1}}^{k}X^x_{p}}{k!}\right)^{m_{k}},
 \]
@@
 \begin{eqnarray*}
-\partial_{y_{j}}\partial_{y_{1}}^{K}f\left(X_{p}\right)
-&=&\sum\frac{K!}{m_{1}!\ldots m_{K}!}f^{(m_{1}+\ldots+m_{K}+1)}(X_{p})\left[\partial_{y_{j}}X_{p}\right]\prod_{k=1}^{K}\left(\frac{\partial_{y_{1}}^{k}X_{p}}{k!}\right)^{m_{k}} \\
-&&+\sum\frac{K!}{m_{1}!\ldots m_{K}!}f^{(m_{1}+\ldots+m_{K})}(X_{p})
+\partial_{y_{j}}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)
+&=&\sum\frac{K!}{m_{1}!\ldots m_{K}!}f^{(m_{1}+\ldots+m_{K}+1)}(X^x_{p})\left[\partial_{y_{j}}X^x_{p}\right]\prod_{k=1}^{K}\left(\frac{\partial_{y_{1}}^{k}X_{p}^x}{k!}\right)^{m_{k}} \\
+&&+\sum\frac{K!}{m_{1}!\ldots m_{K}!}f^{(m_{1}+\ldots+m_{K})}(X^x_{p})
 \sum_{s=1}^{K}\frac{m_{s}}{s!}
-\left(\frac{\partial_{y_{1}}^{s}X_{p}}{s!}\right)^{m_{s}-1}
+\left(\frac{\partial_{y_{1}}^{s}X^x_{p}}{s!}\right)^{m_{s}-1}
 \\
-&& \times\left[\partial_{y_{j}}\partial_{y_{1}}^{s}X_{p}\right]
-\prod_{k\le K,\,k\neq s}\left(\frac{\partial_{y_{1}}^{k}X_{p}}{k!}\right)^{m_{k}}.
+&& \times\left[\partial_{y_{j}}\partial_{y_{1}}^{s}X^x_{p}\right]
+\prod_{k\le K,\,k\neq s}\left(\frac{\partial_{y_{1}}^{k}X^x_{p}}{k!}\right)^{m_{k}}.
 \end{eqnarray*}
@@
 \label{eq:part-part}
-\left|\partial_{y_{j}}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right|
+\left|\partial_{y_{j}}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right|
 \leq A_{K}\prod_{k=2}^{p}\alpha_{k}
@@
 \begin{align*}
-\mathsf{Var}_x\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right]
+\mathsf{Var}_x\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right]
 &\le
@@
 \begin{align*}
-\mathsf{E}_x\left[\sum_{j=1}^{q}\prod_{k=l}^{j}\alpha_{k}\left( \sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right] & \leq\sum_{j=1}^{q}\left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left\Vert \sum_{p=j+1}^{q+1} \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}^{2}\\
+\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=l}^{j}\alpha_{k}\left( \sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right] & \leq\sum_{j=1}^{q}\left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left\Vert \sum_{p=j+1}^{q+1} \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}^{2}\\
  & \leq\sum_{j=1}^{q} \left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left(\sum_{p=j+1}^{q+1}\left\Vert \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}\right)^{2}.
 \end{align*}
-Now using the fact that $\prod\limits_{k=j+1}^{p}\alpha_{k} \leq \exp\left(-\sum\limits_{k=j+1}^{p}\gamma\mu'(X_{k-1})\right),$
+Now using the fact that $\prod\limits_{k=j+1}^{p}\alpha_{k} \leq \exp\left(-\sum\limits_{k=j+1}^{p}\gamma\mu'(X_{k-1}^x)\right),$
 we get
 \[
-\mathsf{E}_x\left[\sum_{j=1}^{q}\prod_{k=1}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]\leq\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \left(\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4)\right)^{2},
+\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=1}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]\leq\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \left(\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4)\right)^{2},
 \]
@@
 \[
-\zeta_{l,j}(u)=\mathsf{E}_x\left[e^{-u\gamma\sum_{k=l}^{j}\mu'(X_{k-1})}\right],\quad u>0.
+\zeta_{l,j}(u)=\mathsf{E}\left[e^{-u\gamma\sum_{k=l}^{j}\mu'(X^x_{k-1})}\right],\quad u>0.
 \]
@@
 \[
-\mathsf{E}_x\left[e^{-u\gamma\left(\sum_{k=l}^{j}[\mu'(X_{k-1})-\pi_\gamma(\mu')]\right)}\right]\leq C(1+V(x)),\quad u\gamma\leq \gamma_0
+\mathsf{E}\left[e^{-u\gamma\left(\sum_{k=l}^{j}[\mu'(X^x_{k-1})-\pi_\gamma(\mu')]\right)}\right]\leq C(1+V(x)),\quad u\gamma\leq \gamma_0
 \]
@@
 \[
-\zeta_{l,j}(u)=\mathsf{E}\left[e^{-s\gamma\sum_{k=l}^{j}\mu'(X_{k-1})}\right]\leq  C(1+V(x))e^{-s\gamma (j-l+1)\pi_\gamma(\mu')}.
+\zeta_{l,j}(u)=\mathsf{E}\left[e^{-s\gamma\sum_{k=l}^{j}\mu'(X^x_{k-1})}\right]\leq  C(1+V(x))e^{-s\gamma (j-l+1)\pi_\gamma(\mu')}.
 \]
@@
 \begin{align*}
-\mathsf{Var}_x\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right] \leq \gamma^{-1} C_K (1+V(x))
+\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X^x_{p}\right)\right] \leq \gamma^{-1} C_K (1+V(x))
 \end{align*}

### 2020-03-11 aap/Revision/vr-mcmc-revision-new.tex
update

@@
 \label{eq:definition-alpha}
-\alpha_k=1-\gamma \mu'(X_{k-1}),\quad k\in\mathbb N.
+\alpha_k=1-\gamma \mu'(X_{k-1}^x),\quad k\in\mathbb N.
 \end{equation}
@@
 \label{eq:08062018b2}
-\left|\partial_{y_l}^r X_p \right| \le C_r\prod_{k=l+1}^p (1-\gamma\mu'(X_{k-1}))\quad \text{a.s.}
-\end{equation}
-where $\partial_{y_l}^r X_p$ is defined in \eqref{eq:definition-differential-f-p}. Moreover, we can choose $C_1=1$.
+\left|\partial_{y_l}^r X^x_p \right| \le C_r\prod_{k=l+1}^p (1-\gamma\mu'(X^x_{k-1}))\quad \text{a.s.}
+\end{equation}
+where $\partial_{y_l}^r X^x_p$ is defined in \eqref{eq:definition-differential-f-p}. Moreover, we can choose $C_1=1$.
 \end{lem}
@@
 \begin{equation}\label{eq:08062018b3}
-\left|\partial_{y_j} \partial_{y_l}^r X_p\right|
-\le c_r\prod_{k=l+1}^p (1-\gamma\mu'(X_{k-1})), \quad \text{a.s.}
+\left|\partial_{y_j} \partial_{y_l}^r X^x_p\right|
+\le c_r\prod_{k=l+1}^p (1-\gamma\mu'(X^x_{k-1})), \quad \text{a.s.}
 \end{equation}
@@
 while, for $p\le j$, it holds
-$\partial_{y_{j}}\partial_{y_{l}}^{r}X_{p}=0$.
+$\partial_{y_{j}}\partial_{y_{l}}^{r}X^x_{p}=0$.
 \end{lem}
@@
 $$
-\partial_{y_{l}}X_{p}
-=\left[1-\gamma\mu'(X_{p-1})\right]\partial_{y_{l}}X_{p-1}
+\partial_{y_{l}}X^x_{p}
+=\left[1-\gamma\mu'(X^x_{p-1})\right]\partial_{y_{l}}X^x_{p-1}
 $$
@@
 $$
-\partial_{y_{j}}\partial_{y_{l}}X_{p}
-=\left[1-\gamma\mu'(X_{p-1})\right]\partial_{y_{j}}\partial_{y_{l}}X_{p-1}-\gamma\mu''(X_{p-1})\partial_{y_{j}}X_{p-1}\partial_{y_{l}}X_{p-1}.
+\partial_{y_{j}}\partial_{y_{l}}X^x_{p}
+=\left[1-\gamma\mu'(X^x_{p-1})\right]\partial_{y_{j}}\partial_{y_{l}}X^x_{p-1}-\gamma\mu''(X^x_{p-1})\partial_{y_{j}}X^x_{p-1}\partial_{y_{l}}X^x_{p-1}.
 $$
@@
 \begin{align*}
-|\partial_{y_{j}}\partial_{y_{l}}X_{p}|
-&\le\alpha_p|\partial_{y_{j}}\partial_{y_{l}}X_{p-1}|
+|\partial_{y_{j}}\partial_{y_{l}}X^x_{p}|
+&\le\alpha_p|\partial_{y_{j}}\partial_{y_{l}}X^x_{p-1}|
 +\gamma
@@
 \prod_{k=j+1}^{p-1}\alpha_k\\
-&\le\alpha_p|\partial_{y_{j}}\partial_{y_{l}}X_{p-1}|
+&\le\alpha_p|\partial_{y_{j}}\partial_{y_{l}}X^x_{p-1}|
 +\gamma
@@
 By Lemma~\ref{lem:06062018a1} applied
-to bound $|\partial_{y_{j}}\partial_{y_{l}}X_{p}|$
+to bound $|\partial_{y_{j}}\partial_{y_{l}}X^x_{p}|$
 for $p\ge j+1$
-(notice that $\partial_{y_j}\partial_{y_l}X_j=0$, that is,
+(notice that $\partial_{y_j}\partial_{y_l}X^x_j=0$, that is,
 $\ol C_0$ in Lemma~\ref{lem:06062018a1} is zero,
@@
 \begin{equation}\label{eq:induc-derive}
-\left|\partial_{y_{j}}\partial_{y_{l}}^{k}X_{p}\right|
+\left|\partial_{y_{j}}\partial_{y_{l}}^{k}X^x_{p}\right|
 \leq c_{k}\prod_{s=l+1}^{p}\alpha_{s}
@@
 \begin{align}
-\partial_{y_{l}}^{r}X_{p}
-&=\left[1-\gamma\mu'(X_{p-1})\right]\partial_{y_{l}}^{r}X_{p-1}
+\partial_{y_{l}}^{r}X^x_{p}
+&=\left[1-\gamma\mu'(X^x_{p-1})\right]\partial_{y_{l}}^{r}X^x_{p-1}
 \label{eq:08062018a1}\\
-&\hspace{1em}-\gamma\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\mu^{(m_{1}+\ldots+m_{r-1})}(X_{p-1})\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X_{p-1}}{k!}\right)^{m_{k}},
+&\hspace{1em}-\gamma\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\mu^{(m_{1}+\ldots+m_{r-1})}(X^x_{p-1})\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X^x_{p-1}}{k!}\right)^{m_{k}},
 \notag
@@
 because the term containing
-$\partial_{y_l}^r X_{p-1}$
+$\partial_{y_l}^r X^x_{p-1}$
 on the right-hand side of~\eqref{eq:08062018a1}
@@
 \begin{align}
-&\partial_{y_{j}}\partial_{y_{l}}^{r}X_{p}
-=\left[1-\gamma_{p}\mu'(X_{p-1})\right]\partial_{y_{j}}\partial_{y_{l}}^{r}X_{p-1}-\gamma\mu''(X_{p-1})\partial_{y_{l}}^{r}X_{p-1}\partial_{y_{j}}X_{p-1}
+&\partial_{y_{j}}\partial_{y_{l}}^{r}X^x_{p}
+=\left[1-\gamma_{p}\mu'(X^x_{p-1})\right]\partial_{y_{j}}\partial_{y_{l}}^{r}X^x_{p-1}-\gamma\mu''(X^x_{p-1})\partial_{y_{l}}^{r}X^x_{p-1}\partial_{y_{j}}X^x_{p-1}
 \label{eq:09062018a2}\\
-&\hspace{1em}-\gamma\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\mu^{(m_{1}+\ldots+m_{r-1}+1)}(X_{p-1})\partial_{y_{j}}X_{p-1}\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X_{p-1}}{k!}\right)^{m_{k}}
+&\hspace{1em}-\gamma\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\mu^{(m_{1}+\ldots+m_{r-1}+1)}(X^x_{p-1})\partial_{y_{j}}X^x_{p-1}\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X^x_{p-1}}{k!}\right)^{m_{k}}
 \notag\\
-&\hspace{1em}-\gamma\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\mu^{(m_{1}+\ldots+m_{r-1})}(X_{p-1})\partial_{y_{j}}\left[\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X_{p-1}}{k!}\right)^{m_{k}}\right]
+&\hspace{1em}-\gamma\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\mu^{(m_{1}+\ldots+m_{r-1})}(X^x_{p-1})\partial_{y_{j}}\left[\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X^x_{p-1}}{k!}\right)^{m_{k}}\right]
 \notag\\
-&\hspace{1em} =\left[1-\gamma\mu'(X_{p-1})\right]\partial_{y_{j}}\partial_{y_{l}}^{r}X_{p-1}+\gamma\epsilon_{l,j,p},
+&\hspace{1em} =\left[1-\gamma\mu'(X^x_{p-1})\right]\partial_{y_{j}}\partial_{y_{l}}^{r}X^x_{p-1}+\gamma\epsilon_{l,j,p},
 \notag
@@
 \begin{eqnarray*}
-\partial_{y_{j}}\left[\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X_{p-1}}{k!}\right)^{m_{k}}\right]&=&\sum_{s=1}^{r-1}\frac{m_{s}}{s!}\left(\frac{\partial_{y_{l}}^{s}X_{p-1}}{s!}\right)^{m_{s}-1}\partial_{y_{j}}\partial_{y_{l}}^{s}X_{p-1}
+\partial_{y_{j}}\left[\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X^x_{p-1}}{k!}\right)^{m_{k}}\right]&=&\sum_{s=1}^{r-1}\frac{m_{s}}{s!}\left(\frac{\partial_{y_{l}}^{s}X^x_{p-1}}{s!}\right)^{m_{s}-1}\partial_{y_{j}}\partial_{y_{l}}^{s}X^x_{p-1}
 \\
-&& \times \prod_{k\le r-1,\,k\neq s}\left(\frac{\partial_{y_{l}}^{k}X_{p-1}}{k!}\right)^{m_{k}}.
+&& \times \prod_{k\le r-1,\,k\neq s}\left(\frac{\partial_{y_{l}}^{k}X^x_{p-1}}{k!}\right)^{m_{k}}.
 \end{eqnarray*}
@@
 $$
-|\partial_{y_{j}}\partial_{y_{l}}^{r}X_{p}|
-\le\alpha_p|\partial_{y_{j}}\partial_{y_{l}}^{r}X_{p-1}|
+|\partial_{y_{j}}\partial_{y_{l}}^{r}X^x_{p}|
+\le\alpha_p|\partial_{y_{j}}\partial_{y_{l}}^{r}X^x_{p-1}|
 +\gamma \,\const\prod_{k=l+1}^{j}\alpha_{k}\prod_{k=j+1}^{p}\alpha_{k}^{2},
@@
 We can again apply Lemma~\ref{lem:06062018a1}
-to bound $|\partial_{y_{j}}\partial_{y_{l}}^r X_{p}|$
+to bound $|\partial_{y_{j}}\partial_{y_{l}}^r X^x_{p}|$
 for $p\ge j+1$
-(notice that $\partial_{y_j}\partial_{y_l}^r X_j=0$, that is,
+(notice that $\partial_{y_j}\partial_{y_l}^r X^x_j=0$, that is,
 $\ol C_0$ in Lemma~\ref{lem:06062018a1} is zero,

### 2020-03-11 aap/Revision/vr-mcmc-revision-new.tex
update: variance of vanilla estimator

--- 2020-03-11 aap/Revision/vr-mcmc-revision-new.tex
@@
 
+\def\P{\mathsf{P}}
 \def\PE{\mathsf{E}}
 \def\PVar{\mathsf{Var}}
+\def\PCov{\mathsf{cov}}
 \newcommandx{\CPE}[3][1=]{{\mathsf E}_{#1}\left[\left. #2 \, \right| #3 \right]}
@@
 \label{eq:expression-variance}
-\PVar_x[\pi_{K,n}^N(f)]=\sum_{k=K+1}^{\infty}\frac{1}{n^2}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})],
+\PVar_x[\pi_{K,n}^N(f)]\leq\sum_{k=K+1}^{\infty}\frac{1}{n^2}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})],
 \end{equation}
@@
 In the next section   we provide a detailed   analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
-\section{Estimation of coefficients}
-The coefficients \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. This amounts to solve a nonparametric regression problem.
-We present a generic regression algorithm.
-Algorithm starts with estimating the functions \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) for $r=N+1,\ldots,N+n.$ We first generate \(\NtrainPath\)  paths  of the chain \(X\)
-(the so-called ``training paths''):
-\[
-\TrainSet=\Bigl\{(X^{(s)}_{0},\ldots,X^{(s)}_{N+n}),\quad s=1,\ldots,\NtrainPath\Bigr\}
-\]
-with $X^{(s)}_{0}=x,$ $s=1,\ldots,T.$
-The regression algorithm proceeds with estimating the  functions \((\bar Q_{l})\) by solving  the least squares optimization problems
-\begin{equation}\label{eq:06042018a1}
-\widehat{Q}_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
-\end{equation}
-for \(r=1,\ldots, n-1,\)
-where \(\Psi\) is a  class of  functions on \(\mathbb{R}^d.\)
-Next  we estimate the coefficients \( \bar a_{l,k}\) using
-the formula
-\begin{equation}
-\label{eq:a_est_int}
-\widehat a_{l,k}(x)=\mathsf{E}\left[\left.\phi_k\left(\xi\right)\widehat Q_{l}\left(\Phi(x,\xi)\right)\right | \TrainSet\right] \,,
-\end{equation}
-where $\xi$ is a random variable independent from $\mathcal{T}$ with law $P_\xi,$
-$\Phi_l$ is defined in \eqref{eq:chain_gen} and the expectation is taken according to known distribution \(P_\xi\).
-Conditioning on $\TrainSet$ underlines that estimation of $\widehat Q_{l}$ derived from training trajectories.
-In many cases integration can be done in closed analytical form.
-For example, in the case of ULA algorithm we have \(P_\xi$ is the multivariate normal distribution $ \mathcal{N}(0,I_d)\) and \(\Phi(x,\xi)=x-\gamma\mu(x)+\sqrt{\gamma}\xi.\) Therefore if we use polynomials to approximate \((\bar Q_{l}),\) then \((\widehat a_{l,k})\) can be even computed in closed form; implementations details are provided in  Section~\ref{sec:num}.
-Upon estimating  the coefficients \((\widehat a_{l,k})\), one can construct an empirical estimate  of \(M_{K,n}^N(f)\) in the form
-\begin{eqnarray*}
-\widehat M_{K,n}^N(f) =\frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\widehat{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
-\end{eqnarray*}
-Obviously, \(\mathsf{E}[\widehat M_{K,n}^N(f)|\TrainSet]=0\) and \(\widehat M_{K,n}^N(f)\) is indeed a valid control variate in that it
-does not introduce any bias. By the Jensen inequality and orthonormality of \((\phi_k)_{k\geq 0},\)
-\begin{multline*}
-\mathsf{E}_x\left[\left.\left|\widehat{M}_{K,n}^{N}(f)-M_{K,n}^{N}(f)\right|^{2}\right | \TrainSet \right]
-\\
-\leq \sum_{k=1}^{K}\frac{1}{n^2}\sum_{l=1}^{n}\PE\left[\left.|A_{n-l,k}(X^x_{N+l-1})-\widehat A_{n-l,k}(X^x_{N+l-1})|^2\right | \TrainSet \right].
-\end{multline*}
-
 \section{Analysis of variance reduced ULA}
@@
 
-\paragraph{Variance:} An upper bound for the variance of the classical estimator~\eqref{eq:29032018a2}
-under {\bf (H1)} and {\bf (H2)} is derived in Section ??. \denis{Insert} In particular, for any bounded function $f,$
+\paragraph{Variance:} In this section we derive an upper bound for the variance of the classical estimator~\eqref{eq:29032018a2} for ULA-based chain.
+
+\begin{lem} 
+\label{lem:variance}
+Let $f$ be a bounded Borel function and $(X^x_{p})_{p \in \nset_0}$ be a Markov chain generated by ULA with constant step size $\gamma$, satisfying {\bf (H1)} and {\bf (H2)} and assume also $n\gamma > C_0$ for some fixed $C_0$.
+Then it holds
 \begin{equation}
 \label{eq:var-mc}
-\PVar[\pi_n^N(f)]\lesssim (n\gamma)^{-1} \eqsp,
-\end{equation}
-where $\lesssim$ stands for inequality up to a constant depending on constants $L,$ $m,$  $M$ and $\|f\|_\infty.$
-One of the main results of this paper is the following  upper bound for the variance of $\pi_{K,n}^{N}(f).$
-
+\PVar[\pi_n^N(f)]\lesssim \frac{V(x)}{n\gamma} \eqsp,
+\end{equation}
+with $V(x) = 1+ \|x\|^2$ where $\lesssim$ stands for inequality up to a constant depending on constants $L,$ $m,$  $M$ and $\mathrm{osc}(f)$.
+\end{lem}
+\begin{proof}
+Under assumptions {\bf (H1)} and {\bf (H2)} the Markov chain $(X^x_{p})_{p \in \nset_0}$ is $V-$geometrically ergodic with $V(x) = 1+\|x\|^2$ and $\rho = \rme^{-\kappa\gamma}$ with $\kappa$ specified in lemma~\ref{lem:v_ergodicity}.
+Due to lemma~\ref{lem:covariance},
+\[
+\PVar[\pi_n^N(f)]\lesssim \frac{1}{n(1-\rho^{1/2})}
+\]
+and the result of lemma follows from Taylor expansion of denominator.
+\end{proof}
+
+One of the main results of this paper is the following  upper bound for the variance of variance-reduced estimator $\pi_{K,n}^{N}(f).$
 \begin{thm}\label{th:mr}
@@
 \end{thm}
-Let us sketch the main steps of the proof. First using integration by parts, we prove that
+Let us sketch the main steps of the proof. First using integration by parts we prove that
 \begin{eqnarray}
 \label{eq:Ask-part}
-A_{q,\mathbf{k}}(x)=\gamma^{|\mathbf{k}'|/2}\frac{\sqrt{(\mathbf{k}-\mathbf{k}')!}}{\sqrt{\mathbf{k}!}}\mathsf{E}_x\left[ \partial_{Z_1}^{\mathbf{k}'} F(x,Z_1,\ldots,Z_q) \mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_{1})\right],
+A_{s,\mathbf{k}}(x)=\gamma^{|\mathbf{k}'|/2}\frac{\sqrt{(\mathbf{k}-\mathbf{k}')!}}{\sqrt{\mathbf{k}!}}\mathsf{E}\left[ \partial_{Z_1}^{\mathbf{k}'} F(x,Z_1,\ldots,Z_s) \mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_{1})\right]
 \end{eqnarray}
-where $F_q(x,Z_1,\ldots,Z_q)=\sum_{r=0}^q f(X^x_r)$ and $\partial_{Z_1}^{\mathbf{k}'}$
-stands for a weak partial derivative of the functional $F_s$ that also can be viewed as discretised version of Malliavin derivative. From \eqref{eq:Ask-part} we derive
-\begin{eqnarray*}
-\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1} A^2_{q,\mathbf{k}}(x)\leq \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
+where $F_s(x,Z_1,\ldots,Z_s)=\sum_{r=0}^s f(X^x_r)$ and $\partial_{Z_1}^{\mathbf{k}'}$
+stands for a weak partial derivative of the functional $F_s$ that also can be viewed as discretised version of Malliavin derivative. From \eqref{eq:Ask-part}, we derive
+\begin{eqnarray*}
+\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1} A^2_{s,\mathbf{k}}(x)\leq \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
 \left(\frac{\gamma}{2}\right)^{|I|K}
-\PVar \left( \sum_{p=1}^{q}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X_{p}^x\right)
+\PVar_x \left( \sum_{p=1}^{s}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X_{p}\right)
 \right),
@@
 \PVar \left(\sum_{p=1}^{q}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X^x_{p}\right)
-\right)\leq C_{K}\gamma^{-1}(1+V(x)), \quad V(x) = 1 + \|x\|^2
+\right)\leq C(1+V(x))
 \end{equation*}
-for some constant $C_K$ not depending on $q$ and $\gamma$.
+for some constant $C$ not depending on $q$ and $V(x) = 1 + \|x\|^2$.
 \section{Numerical results}
@@
 
-Aforementioned lemmas allow us to bound the exponential moment of the additive functional of ULA. Namely, the following theorem holds
-
-\begin{thm}
-\label{thm:ula_const_size}
-Suppose that ULA kernel \ref{eq:kernel} satisfies assumptions {\bf (H1)} and {\bf (H2)}, and let $X_1,\ldots,X_n,\ldots$ be generated by ULA with constant step size $\gamma$.
+\section{Covariance estimation for $V-$geometrically ergodic Markov chains}\label{sec:appendix_moments}
+In this section we assume that $(X_p)_{p \in \nset_0}$ is a $V-$geometrically ergodic Markov chain and prove bounds on the variance of ergodic average $\pi_N^n(f)$ of the form \eqref{eq:29032018a2}.
+We use the same technique as in \cite{belomestny2019esvm} to control autocovariances for a given Markov chain.
+We start from auxiliary lemma:
+\begin{lem} 
+\label{lem:fp_covariance}
+Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$.
+Then it holds for some constant $C > 0$
+\begin{equation}
+|\mathsf{E}_x[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq C\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x)
+\end{equation}
+For the stationary distribution it holds
+\begin{equation}
+|\mathsf{E}_{\pi}[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq C\pi(V)\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}
+\end{equation}
+\end{lem}
+\begin{proof}
+Note that
+\[
+|\mathsf{E}_x[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq |\tilde{f}(x)| \mathsf{E}_x|\tilde{f}(X_s)| \leq \|\tilde{f}\|_{V^{\frac{1}{2}}}V^{\frac{1}{2}}(x)\int\limits_{\rset^d} V^{\frac{1}{2}}(y)|\mathsf{P}^{s}(x,\,dy)-\pi(dy)|
+\]
+By Hoelder inequality, 
+\begin{eqnarray*}
+\int\limits_{\rset^d} V^{\frac{1}{2}}(y)|\mathsf{P}^{s}(x,\,dy)-\pi(dy)| & \leq & \left(\int\limits_{\rset^d} V(y)|\mathsf{P}^{s}(x,\,dy)-\pi(dy)|\right)^{1/2}\left(\int\limits_{\rset^d}|\mathsf{P}^{s}(x,\,dy)-\pi(dy)|\right)^{1/2} \leq \\
+& \leq & 2\rho^{s/2}V^{1/2}(x)
+\end{eqnarray*}
+yielding the first statement of lemma. The second statement can be obtained from the first one by integration with respect to $\pi$. 
+\end{proof}
+
+\begin{lem}
+\label{lem:covariance}
+Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Then
+\begin{equation}
+\label{eq:cov_v_ergodic}
+\PCov_x \left[f(X_k),f(X_{k+s})\right] \leq C^2V^2(x)\rho^{k+s} + C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x) + C\pi(V)\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}
+\end{equation}
+\end{lem}
+\begin{proof} 
+Note that
+\[
+\PCov_x \left[f(X_k),f(X_{k+s})\right] = \PE_x\left[f(X_k) - \pi(f)\right]\left[f(X_{k+s})-\pi(f)\right] + \left[\pi(f) - \PE_xf(X_s)\right]\left[\PE_xf(X_k) - \pi(f)\right]
+\]
+Due to $V-$ergodicity, 
+\[
+\left|\left[\pi(f) - \PE_xf(X_s)\right]\left[\PE_xf(X_k) - \pi(f)\right]\right| \leq C^2V^2(x)\rho^{k+s}
+\]
+To bound the first term note that
+\begin{eqnarray*}
+\left|\PE_x\left[f(X_k) - \pi(f)\right]\left[f(X_{k+s})-\pi(f)\right] - \PCov_{\pi}\left[f(X_k),f(X_{k+s})\right]\right| \leq \int\limits_{\rset^d}\left|\PE_y\left[\tilde{f}(X_0)\tilde{f}(X_{s})\right]\right| |\P^k(x,dy)-\pi(dy)| \leq  \\
+\leq C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x)
+\end{eqnarray*}
+where the last inequality is due to lemma~\ref{lem:fp_covariance}, which implies \eqref{eq:cov_v_ergodic}.
+\end{proof}
+Now we state and prove the main result of this section on the variance bound for the estimator $\pi_N^n(f)$ for $V-$geometrically ergodic Markov chain.
+\begin{lem}
+Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$.
+Assume also that $n(1-\rho^{1/2}) > 1$.
+Then
+\begin{equation}
+\label{eq:vanilla_var}
+\PVar_x \left[ \pi_{N}^n(f) \right] \leq \frac{C\pi(V)\|\tilde{f}\|_{V^{1/2}}}{n(1-\rho^{1/2})} + \mathcal{O}(\frac{1}{n^2(1-\rho^{1/2})}) 
+\end{equation}
+\end{lem}
+\begin{proof}
+Note that 
+\begin{eqnarray*}
+\PVar_x \left[ \frac{1}{n}\sum\limits_{k=N+1}^{N+n}f(X_k) \right] = \underbrace{\frac{1}{n^2}\sum\limits_{k=N+1}^{N+n}\PVar_x \left[ f(X_k) \right]}_{S_1} + \underbrace{\frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\sum\limits_{s=1}^{n-k-1}\PCov_x\left[f(X_k),f(X_{k+s})\right]}_{S_2}
+\end{eqnarray*}
+Now we bound first and second sum using lemma~\ref{lem:covariance}:
+\[
+S_1 \leq \frac{1}{n}C\pi(V)\|\tilde{f}\|_{V^{1/2}} + \frac{\rho^N C^2\left(V^2(x) + V(x)\|\tilde{f}\|_{V^{1/2}}\right)}{n^2(1-\rho)}
+\]
+\begin{eqnarray*}
+S_2 & \leq & \frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\left[C^2V^2(x)\rho^{k+1}\frac{1}{1-\rho} + C^2V(x)\rho^{k+1/2}\frac{1}{1-\rho^{1/2}} + C\pi(V)\frac{1}{1-\rho^{1/2}}\|\tilde{f}\|_{V^{1/2}}\right] \leq \\
+& \leq & \frac{2\rho^NC^2V^2(x)}{n^2(1-\rho)^2} + \frac{2\rho^NC^2V(x)}{n^2(1-\rho)(1-\rho^{1/2})} + \frac{C\pi(V)\|\tilde{f}\|_{V^{1/2}}}{n(1-\rho^{1/2})}
+\end{eqnarray*}
+Hence \eqref{eq:vanilla_var} follows.
+\end{proof}
+
+%Aforementioned lemmas allow us to bound the exponential moment of the additive functional of ULA. Namely, the following theorem holds
+%
+%\begin{thm}
+%\label{thm:ula_const_size}
+%Suppose that ULA kernel \ref{eq:kernel} satisfies assumptions {\bf (H1)} and {\bf (H2)}, and let $X_1,\ldots,X_n,\ldots$ be generated by ULA with constant step size $\gamma$.
 Let $X_0 = x$ be fixed.
 Then for any bounded function $g(x) : \mathbb{R}^d \rightarrow \mathbb{R}, |g(x)| \leq M$, it holds
-\begin{eqnarray*}
-\mathsf{E}\left[\exp\left(-\rho\gamma\sum_{k=l}^{p}\left(g(X_{k})-\pi_{\gamma}(g)\right)\right)\right] \leq C_1\exp{\left(s^2 \kappa \gamma^2 (p-l+1)\right)}
-\end{eqnarray*}
-for some absolute constant $C_1$ not depending on $l,p$.
-\end{thm}
-
-\begin{proof} Conditions {\bf (H1)} and {\bf (H2)} imply that the Markov kernel $R_\gamma$ is $V-$ergodic (lemma \ref{lem:v_ergodicity}) and satisfies drift condition \ref{lem:drift}.
+%\begin{eqnarray*}
+%\mathsf{E}\left[\exp\left(-\rho\gamma\sum_{k=l}^{p}\left(g(X_{k})-\pi_{\gamma}(g)\right)\right)\right] \leq C_1\exp{\left(s^2 \kappa \gamma^2 (p-l+1)\right)}
+%\end{eqnarray*}
+%for some absolute constant $C_1$ not depending on $l,p$.
+%\end{thm}
+%
+%\begin{proof} Conditions {\bf (H1)} and {\bf (H2)} imply that the Markov kernel $R_\gamma$ is $V-$ergodic (lemma \ref{lem:v_ergodicity}) and satisfies drift condition \ref{lem:drift}.
 Due to \cite[Theorem~3, Fact 3]{moulines:bound_dif:2019}, we obtain the following bound
-\begin{align*}
-\mathsf{E}_x\left[\exp\left(-s\gamma\sum_{k=l}^{p}\left(g(X_{k}) - \mathsf{E}_x g(X_k)\right)\right)\right] \leq \exp{\left(s^2 \kappa \gamma^2 (p-l+1)\right)}
-\end{align*}
-\sergey{I will put precise constant later}
-with constant $\kappa = ...$. Note that
-\begin{align*}
-\mathsf{E}\left[\exp\left(-s\gamma\sum_{k=l}^{p}\left(g(X_{k})-\pi_{\gamma}(g)\right)\right)\right] =
-\end{align*}
-
-\begin{align*}
-=\mathsf{E}_x\left[\exp{\left(-s\gamma\sum_{k=l}^{p}\left(g(X_{k}) - \mathsf{E}_x g(X_k)\right)\right)}\right]
-\exp{\left(s\gamma\sum\limits_{k=l}^{p}\left(\pi_\gamma (g) - \mathsf{E}_x g(X_k)\right)\right)}
-\end{align*}
-Using lemma~\ref{lem:v_ergodicity},
-\begin{align*}
-|\mathsf{E} g(X_k) - \pi_\gamma(g)| \leq Md_V(\pi_\gamma, \delta_xQ^{1,n}_{\boldsymbol{\gamma}}) \leq CM\rho^k\left(V(x) + \pi_\gamma(V)\right)
-\end{align*}
-Hence,
-\begin{align*}
-\exp{\left(s\gamma\sum\limits_{k=l}^{p}\left(\pi_\gamma (g) - \mathsf{E}_x g(X_k)\right)\right)} \leq \exp{\left(s\gamma CM(V(x) + \pi_\gamma(V))\frac{\rho^l - \rho^{p+1}}{1-\rho}\right)}
-\end{align*}
-and the statement follows.
-\end{proof}
-
-Let us formulate and prove prove analogue of the previous theorem for inhomogeneous Markov chain with step size $\gamma_1,\ldots,\gamma_n,\ldots$.
+%\begin{align*}
+%\mathsf{E}_x\left[\exp\left(-s\gamma\sum_{k=l}^{p}\left(g(X_{k}) - \mathsf{E}_x g(X_k)\right)\right)\right] \leq \exp{\left(s^2 \kappa \gamma^2 (p-l+1)\right)}
+%\end{align*}
+%\sergey{I will put precise constant later}
+%with constant $\kappa = ...$. Note that
+%\begin{align*}
+%\mathsf{E}\left[\exp\left(-s\gamma\sum_{k=l}^{p}\left(g(X_{k})-\pi_{\gamma}(g)\right)\right)\right] =
+%\end{align*}
+%
+%\begin{align*}
+%=\mathsf{E}_x\left[\exp{\left(-s\gamma\sum_{k=l}^{p}\left(g(X_{k}) - \mathsf{E}_x g(X_k)\right)\right)}\right]
+%\exp{\left(s\gamma\sum\limits_{k=l}^{p}\left(\pi_\gamma (g) - \mathsf{E}_x g(X_k)\right)\right)}
+%\end{align*}
+%Using lemma~\ref{lem:v_ergodicity},
+%\begin{align*}
+%|\mathsf{E} g(X_k) - \pi_\gamma(g)| \leq Md_V(\pi_\gamma, \delta_xQ^{1,n}_{\boldsymbol{\gamma}}) \leq CM\rho^k\left(V(x) + \pi_\gamma(V)\right)
+%\end{align*}
+%Hence,
+%\begin{align*}
+%\exp{\left(s\gamma\sum\limits_{k=l}^{p}\left(\pi_\gamma (g) - \mathsf{E}_x g(X_k)\right)\right)} \leq \exp{\left(s\gamma CM(V(x) + \pi_\gamma(V))\frac{\rho^l - \rho^{p+1}}{1-\rho}\right)}
+%\end{align*}
+%and the statement follows.
+%\end{proof}
+%
+%Let us formulate and prove prove analogue of the previous theorem for inhomogeneous Markov chain with step size $\gamma_1,\ldots,\gamma_n,\ldots$.
 Note that under assumptions $\sum_{p=1}^{\infty} \gamma_p=\infty, \quad \sum_{p=1}^{\infty} \gamma^2_p<\infty$ and {\bf (H1)} the stationary distribution of ULA-based chain will be equal to $\pi$.
 We will need additional assumption
-\begin{description}
-\item[{\bf (H3)}] There exist such constants $\rho > 0, K_3 > 0$ that
-\begin{eqnarray*}
-d_V(\delta_xQ_{\boldsymbol{\gamma}}^{1,n},\pi) \leq K_3 V(x) \rho^n
-\end{eqnarray*}
-\end{description}
-
-\begin{thm}
-\label{th:exp_mom_ula}
-Suppose that
-\begin{eqnarray*}
-\sum_{p=1}^{\infty} \gamma_p=\infty, \quad \sum_{p=1}^{\infty} \gamma^2_p<\infty.
-\end{eqnarray*}
-Under assumptions {\bf (H1)}, {\bf (H2)} and {\bf (H3)}, for any bounded function $g$ on $\mathbb{R}^d$
-we have
+%\begin{description}
+%\item[{\bf (H3)}] There exist such constants $\rho > 0, K_3 > 0$ that
 %\begin{eqnarray*}
-%\mathsf{E}\left[\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}g(X_{p})-\bar Q_l(X_l)\right)^{4}\right] \lesssim \frac{1}{\Gamma^4_{N+2,N+n+1}}
+%d_V(\delta_xQ_{\boldsymbol{\gamma}}^{1,n},\pi) \leq K_3 V(x) \rho^n
 %\end{eqnarray*}
- %and
+%\end{description}
+%
+%\begin{thm}
+%\label{th:exp_mom_ula}
+%Suppose that
 %\begin{eqnarray*}
-%\bar Q_l(X_l)=\mathsf{E}\left[\left.\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right|X_{l}\right].
+%\sum_{p=1}^{\infty} \gamma_p=\infty, \quad \sum_{p=1}^{\infty} \gamma^2_p<\infty.
 %\end{eqnarray*}
-%Moreover,
-\begin{eqnarray}
-\label{eq:exp_decrease_step}
-\mathsf{E}\left[\exp\left(-s\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k})-\left(\sum_{k=l}^{p}\gamma_{k+1}\right)\pi(g)\right)\right)\right]\leq C
-\end{eqnarray}
-for all natural $0<l\leq p<\infty$ with constant $C$ not depending on $l,p.$. Moreover,
-\begin{eqnarray}
-\label{eq:var_decrease_step}
-\mathsf{Var}_x{\left(\sum_{k=l}^{l+n-1}\frac{\gamma_{k+1}}{\Gamma_{l+1,l+n}}g(X_k) - \mathsf{E}_x \sum_{k=l}^{l+n-1}\frac{\gamma_{k+1}}{\Gamma_{l+1,l+n}}g(X_k)\right)} \leq \frac{C_1}{\Gamma_{l+1,l+n}}
-\end{eqnarray}
-\end{thm}
-\begin{proof}
-
- Note that under assumptions {\bf (H1)}, {\bf (H2)} and {\bf (H3)} it holds due to \cite[Theorem~4]{moulines:bound_dif:2019} that
- \begin{align}
- \label{eq:exp_bounded_dif}
-\mathsf{E}_x\left[\exp\left(-s\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k}) - \mathsf{E}_x \left(\sum\limits_{k=l}^{p}\gamma_{k+1}g(X_k)\right)\right)\right)\right] \leq \exp{\left(s^2 \kappa \sum\limits_{k=l}^{p}\gamma_{k+1}^2\right)} \leq \exp{\left(s^2 \kappa \Gamma^*\right)}
-\end{align}
-\sergey{This theorem is not yet written in the second paper}
-with $\Gamma^* = \sum\limits_{k=1}^{\infty}\gamma_k^2 < \infty$ and $\kappa = ...$. Using the same transformation as in theorem~\ref{thm:ula_const_size}, and using the bound
-\begin{align*}
-|\mathsf{E}_x f(X_k) - \pi(f)| \leq K_3 M \rho^{\Gamma_{1,k}}V(x)
-\end{align*}
-we obtain
-\begin{align*}
-\mathsf{E}\left[\exp\left(-s\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k})-\left(\sum_{k=l}^{p}\gamma_{k+1}\right)\pi(g)\right)\right)\right] \leq \exp{\left(s^2 \kappa \Gamma^*\right)}\exp{\left(K_3 M V(x) \sum\limits_{k=l}^{p}\rho^{\Gamma_{1,k}}\gamma_{k+1}\right)} \leq
-\end{align*}
-
-\begin{align*}
-\leq \exp{\left(s^2 \kappa \Gamma^*\right)}\exp{\left(K_3 M V(x) \rho^{\Gamma_{1,k}} \int\limits_{0}^{+\infty}\rho^y\,dy\right)} = \exp{\left(s^2 \kappa \Gamma^*\right)}\exp{\left(K_3 M V(x) \frac{\rho^{\Gamma_{1,k}}}{\ln{\rho}}\right)}
-\end{align*}
-\end{proof}
-which proves \ref{eq:exp_decrease_step}. Note that from \ref{eq:exp_bounded_dif} by Taylor expansion for small $s$ we obtain
-\begin{align*}
-1 + \frac{s^2}{2}\mathsf{Var}_x{\left(\sum_{k=l}^{l+n-1}\gamma_{k+1}g(X_k) - \mathsf{E}_x \sum_{k=l}^{l+n-1}\gamma_{k+1}g(X_k)\right)} + O(s^3) \leq 1 + s^2 \kappa \sum\limits_{k=l+1}^{l+n}\gamma_{k}^2 + O(s^4)
-\end{align*}
-Since it holds for arbitarily small $s$, the variance is bounded by
-\begin{align*}
-\mathsf{Var}_x{\left(\sum_{k=l}^{l+n-1}\gamma_{k+1}g(X_k) - \mathsf{E}_x \sum_{k=l}^{l+n-1}\gamma_{k+1}g(X_k)\right)} \leq 2\kappa \sum\limits_{k=l+1}^{l+n}\gamma_{k}^2 \leq 2\kappa \Gamma_{l+1,l+n}
-\end{align*}
-Now we need to divide both parts by $\Gamma^2_{l+1,l+n}$ to get \ref{eq:var_decrease_step}. The proof is completed.
+%Under assumptions {\bf (H1)}, {\bf (H2)} and {\bf (H3)}, for any bounded function $g$ on $\mathbb{R}^d$
+%we have
+%%\begin{eqnarray*}
+%%\mathsf{E}\left[\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}g(X_{p})-\bar Q_l(X_l)\right)^{4}\right] \lesssim \frac{1}{\Gamma^4_{N+2,N+n+1}}
+%%\end{eqnarray*}
+% %and
+%%\begin{eqnarray*}
+%%\bar Q_l(X_l)=\mathsf{E}\left[\left.\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right|X_{l}\right].
+%%\end{eqnarray*}
+%%Moreover,
+%\begin{eqnarray}
+%\label{eq:exp_decrease_step}
+%\mathsf{E}\left[\exp\left(-s\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k})-\left(\sum_{k=l}^{p}\gamma_{k+1}\right)\pi(g)\right)\right)\right]\leq C
+%\end{eqnarray}
+%for all natural $0<l\leq p<\infty$ with constant $C$ not depending on $l,p.$. Moreover,
+%\begin{eqnarray}
+%\label{eq:var_decrease_step}
+%\mathsf{Var}_x{\left(\sum_{k=l}^{l+n-1}\frac{\gamma_{k+1}}{\Gamma_{l+1,l+n}}g(X_k) - \mathsf{E}_x \sum_{k=l}^{l+n-1}\frac{\gamma_{k+1}}{\Gamma_{l+1,l+n}}g(X_k)\right)} \leq \frac{C_1}{\Gamma_{l+1,l+n}}
+%\end{eqnarray}
+%\end{thm}
+%\begin{proof}
+%
+% Note that under assumptions {\bf (H1)}, {\bf (H2)} and {\bf (H3)} it holds due to \cite[Theorem~4]{moulines:bound_dif:2019} that
+% \begin{align}
+% \label{eq:exp_bounded_dif}
+%\mathsf{E}_x\left[\exp\left(-s\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k}) - \mathsf{E}_x \left(\sum\limits_{k=l}^{p}\gamma_{k+1}g(X_k)\right)\right)\right)\right] \leq \exp{\left(s^2 \kappa \sum\limits_{k=l}^{p}\gamma_{k+1}^2\right)} \leq \exp{\left(s^2 \kappa \Gamma^*\right)}
+%\end{align}
+%\sergey{This theorem is not yet written in the second paper}
+%with $\Gamma^* = \sum\limits_{k=1}^{\infty}\gamma_k^2 < \infty$ and $\kappa = ...$. Using the same transformation as in theorem~\ref{thm:ula_const_size}, and using the bound
+%\begin{align*}
+%|\mathsf{E}_x f(X_k) - \pi(f)| \leq K_3 M \rho^{\Gamma_{1,k}}V(x)
+%\end{align*}
+%we obtain
+%\begin{align*}
+%\mathsf{E}\left[\exp\left(-s\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k})-\left(\sum_{k=l}^{p}\gamma_{k+1}\right)\pi(g)\right)\right)\right] \leq \exp{\left(s^2 \kappa \Gamma^*\right)}\exp{\left(K_3 M V(x) \sum\limits_{k=l}^{p}\rho^{\Gamma_{1,k}}\gamma_{k+1}\right)} \leq
+%\end{align*}
+%
+%\begin{align*}
+%\leq \exp{\left(s^2 \kappa \Gamma^*\right)}\exp{\left(K_3 M V(x) \rho^{\Gamma_{1,k}} \int\limits_{0}^{+\infty}\rho^y\,dy\right)} = \exp{\left(s^2 \kappa \Gamma^*\right)}\exp{\left(K_3 M V(x) \frac{\rho^{\Gamma_{1,k}}}{\ln{\rho}}\right)}
+%\end{align*}
+%\end{proof}
+%which proves \ref{eq:exp_decrease_step}. Note that from \ref{eq:exp_bounded_dif} by Taylor expansion for small $s$ we obtain
+%\begin{align*}
+%1 + \frac{s^2}{2}\mathsf{Var}_x{\left(\sum_{k=l}^{l+n-1}\gamma_{k+1}g(X_k) - \mathsf{E}_x \sum_{k=l}^{l+n-1}\gamma_{k+1}g(X_k)\right)} + O(s^3) \leq 1 + s^2 \kappa \sum\limits_{k=l+1}^{l+n}\gamma_{k}^2 + O(s^4)
+%\end{align*}
+%Since it holds for arbitarily small $s$, the variance is bounded by
+%\begin{align*}
+%\mathsf{Var}_x{\left(\sum_{k=l}^{l+n-1}\gamma_{k+1}g(X_k) - \mathsf{E}_x \sum_{k=l}^{l+n-1}\gamma_{k+1}g(X_k)\right)} \leq 2\kappa \sum\limits_{k=l+1}^{l+n}\gamma_{k}^2 \leq 2\kappa \Gamma_{l+1,l+n}
+%\end{align*}
+%Now we need to divide both parts by $\Gamma^2_{l+1,l+n}$ to get \ref{eq:var_decrease_step}. The proof is completed.
 \bibliographystyle{plain}

### 2020-03-12 aap/Revision/vr-mcmc-revision-new.tex
truncation

@@
 \begin{eqnarray*}
-\widehat M_{K,n}^N(f) =\frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\widehat{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
+\widehat M_{K,n}^N(f) = \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\widehat{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
 \end{eqnarray*}
@@
 \end{multline*}
+Yet it is impractical to use estimator of the form $\widehat M_{K,n}^N(f)$, since it may be too expensive to compute for $n$ large enough.
+Instead we introduce the truncated version of this control variate
+\begin{eqnarray*}
+\tilde M_{K,n}^N(f) = \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\widehat{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\mathbbm{1}\{ |p-k| < n_{\mathrm{trunc}} \}\right]
+\end{eqnarray*}
+where $n_{\mathrm{trunc}} \in \nset$ is a hyperparameter. Validity of this approach follows from the following lemma
+
+\begin{lem}
+\label{lem:coef_decrease}
+Let $(X^x_{p})_{p \in \nset_0}$ be a $V-$geometrically ergodic homogeneous Markov chain with stationary distribution $\pi$ and $f$ be a Borel function with $\|f\|_{V} < \infty$. Then it holds
+\begin{equation}
+|\bar a_{r,k}(x)| \leq C\rho^r\|f\|_{V} (\PE V^2(\Phi(x,\xi)))^{1/2}
+\end{equation}
+where $\bar a_{r,k}(x)$ is defined in \eqref{eq:definition-bar-a}.
+\end{lem}
+\begin{proof}
+Note that 
+\[
+\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
+\]
+with  $Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right], r\in \mathbb{N}$. $V-$ergodicity of $(X^x_{p})_{p \in \nset_0}$ implies that
+\[
+|Q_{r}(x) - \pi(f)| \leq |\int\limits_{\rset^d} f(y)\left(\P^r(x,dy)-\pi(dy)\right)| \leq C\rho^r V(x) \|f\|_{V}
+\]
+Thus, since $\PE\phi_k\left(\xi\right) = 0$, we obtain from Cauchy-Schwarz inequality
+\begin{eqnarray*}
+|\bar a_{r,k}(x)| \leq \left(\PE \phi^2_k\left(\xi\right)\right)^{1/2}\left(\PE |Q_{r}(\Phi(x,\xi)) - \pi(f)|^{2}\right)^{1/2} \leq C\rho^r\|f\|_{V} (\PE V^2(\Phi(x,\xi)))^{1/2}
+\end{eqnarray*}
+and the statement follows.
+\end{proof}
 

### 2020-03-12 aap/Revision/vr-mcmc-revision-new.tex
local update
numerics update:banana-shape

@@
 
+\subsection{Banana shape distribution}
+The “Banana-shape” distribution, proposed by \cite{Haario1999}, can be obtained from a d-dimensional Gaussian vector with zero mean and covariance $\mathrm{diag}(p,1,\ldots,1)$ by applying transformation $\varphi_b(x) : \rset^d \rightarrow \rset^d$ of the form
+\[
+\varphi(x_1,\ldots,x_n) = (x_1, x_2 + bx^2_1 - pb,x_3,\ldots,x_d)
+\]
+where $p > 0$ and $b > 0$ are parameters; $b$ accounts for the curvature of density’s level sets. The potential $U$ is given by
+\[
+U(x_1, . . . , x_d) = \frac{x_1^2}{2} + (x_2 + bx_1^2 - pb)^2 + \sum\limits_{k=3}^{d}\frac{x_k^2}{2}
+\]
+The quantity of interest is $\PE_{\pi}[X_2]$.
+We set $p = 100, b = 0.1$ and consider $d=8$.
+We solve the least squares  problems \eqref{eq:Qcircr} using the second-order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section, hence we set $K = 2$.
+We set the truncation level \(n_{\mathrm{trunc}} = 50\).
+To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 10^3$ and plot boxplots of the ergodic averages in Figure~\ref{fig:banana_shape}.
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=0.6\textwidth]{../pictures/banana_shape.png}
+\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Banana shape density.
+The compared estimators are the ordinary empirical average (Vanilla), our estimator of control variates (MDCV), zero variance estimator (ZV) control variates obtained with empirical spectral variance minimisation (ESVM).
+\label{fig:banana_shape}}
+\end{center}
+\end{figure}
+
 \subsection{Binary Logistic Regression}
-
 Second experiment considers the problem of logistic regression.

### 2020-03-12 aap/Revision/vr-mcmc-revision-new.tex
update: bias

@@
 \label{eq:expression-variance}
-\PVar_x[\pi_{K,n}^N(f)]=\frac{1}{n^2}\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})],
+\PVar[\pi_{K,n}^N(f)]=\frac{1}{n^2}\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})],
 \end{equation}
@@
 \label{eq:definition-A-s,k}
-A_{q,k}(y)=\sum_{r=0}^q \bar{a}_{r,k}(y),\quad q\in \mathbb{N}
+A_{q,k}(y)=\sum_{r=0}^q \bar{a}_{r,k}(y),\quad q\in \mathbb{N},
 \end{eqnarray}
@@
 \label{eq:sq_bias}
-\left( \mathsf{E}[\rho(f)]-\pi(f)\right)^2 \lesssim \mathrm{osc}(f)^2 \left[\gamma + \frac{1}{n}\left(V(x) + \pi_{\gamma}(V)\right)\right]
-\end{equation}
-for $V(x) = 1 + \|x\|^2$, where $\lesssim$ means less up to a constant not depending on $n$ and $\gamma$.
+\left( \mathsf{E}[\rho(f)]-\pi(f)\right)^2 \lesssim \|f\|^2_{\infty}\left(\gamma + \frac{\rho^{2N}}{n(1-\rho^2)}\right) \lesssim \|f\|^2_{\infty}\left(\gamma + \frac{e^{-\gamma \kappa N}}{2\gamma n}\right)
+\end{equation}
+for $V(x) = 1 + \|x\|^2$, $\rho$ defined in Lemma~\ref{lem:v_ergodicity}, $\kappa$ independent of $N,n,\gamma$ and $\lesssim$ means less up to a constant not depending on $N,n$ and $\gamma$.
 \end{lem}
 
-\begin{proof} We can estimate the conditional bias similarly to~\cite[Section~3]{durmus:moulines:2017}:
-\begin{equation}
-\label{eq:06042018a2}
+\begin{proof} Under assumptions {\bf (H1)} and {\bf (H2)}, the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$.
+Thus we represent the squared bias as
+\begin{eqnarray}
+\label{eq:bias_1}
 \left( \mathsf{E}[\pi_{K,n}^N(f)]-\pi(f)\right)^2
-=\left(\mathsf{E}[\pi_{n}^N(f)]-\pi(f)\right)^2
-\le
-\frac{\mathrm{osc}(f)^2}{n}\sum_{p=N+1}^{N+n}
-\tvnorm{\delta_x \kerULA_\gamma^{p}-\pi(\cdot)}^2,
-\end{equation}
-Under assumptions {\bf (H1)} and {\bf (H2)}, the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$.
+&=&\left(\mathsf{E}[\pi_{n}^N(f)]-\pi(f)\right)^2
+=
+\frac{1}{n^2}\left[\sum\limits_{k=N+1}^{N+k}\left(f(X_k^x)-\pi(f)\right)\right]^2 \leq \\
+&\leq& 
+2|\pi_{\gamma}(f) - \pi(f)|^2 + \frac{2}{n}\sum\limits_{k=N+1}^{N+n}\left[\PE f(X_k^x) - \pi_{\gamma}(f)\right]^2
+\end{eqnarray}
  From \cite[Theorem~10]{durmus:moulines:2017} and Lemma~\ref{lem:v_ergodicity} it follows that
 \begin{equation}
-\label{eq:bias}
-\tvnorm{\pi - \pi_\gamma} < C_1\sqrt{\gamma}, \quad \tvnorm{\delta_x\kerULA_{\gamma}^{n}-\pi_\gamma} \leq C_2\rho^{\gamma n}\left(V(x) + \pi_\gamma(V)\right)
-\end{equation}
-for  $V(x) = 1 + \|x\|^2,$ some $\rho\in (0,1)$ and constants $C_1,C_2>0$ independent of $\gamma$  and $n$. Combining \eqref{eq:bias} with the triangle inequality, we obtain \eqref{eq:sq_bias}.
+\label{eq:bias_2}
+\tvnorm{\pi - \pi_\gamma} < C_1\sqrt{\gamma}, \quad |\PE f(X_k^x) - \pi_{\gamma}(f)| \leq C_2 \rho^n\|f\|_{V}V(x)
+\end{equation}
+for  $V(x) = 1 + \|x\|^2,$ $\rho\in (0,1)$ specified in Lemma~\ref{13} and constants $C_1,C_2>0$ independent of $\gamma$  and $n$.
+Since $f$ is bounded, $\|f\|_V \leq \|f\|_{\infty}$.
+Hence, combining \eqref{eq:bias_1} and \eqref{eq:bias_2}, we obtain \eqref{eq:sq_bias}.
 \end{proof}

### 2020-03-13 aap/Revision/vr-mcmc-revision-new.tex
update

@@
 In the next section   we provide a detailed   analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
+In many situations the coefficients $\bar{a}_{r,k}$ also decay exponentially fast as $r$ increases.
+The following result can be easily established.
+\begin{lem}
+\label{lem:coef_decrease}
+Let $(X^x_{p})_{p \in \nset_0}$ be a $V-$geometrically ergodic homogeneous Markov chain with stationary distribution $\pi$ and $f$ be a Borel function with $\|f\|_{V} < \infty$. Then it holds
+\begin{equation}
+|\bar a_{r,k}(x)| \leq C\rho^r\|f\|_{V} (\PE V^2(\Phi(x,\xi)))^{1/2}
+\end{equation}
+where $\rho\in (0,1)$ and $C$ is an absolute constant.
+\end{lem}
+
 \section{Estimation of coefficients}
@@
 \begin{equation}\label{eq:06042018a1}
-\widehat{Q}_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
-\end{equation}
-for \(r=1,\ldots, n-1,\)
-where \(\Psi\) is a  class of  real-valued functions on \(\mathbb{R}^d.\)
+\widehat{Q}_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n_{tr}-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
+\end{equation}
+for \(r=1,\ldots, n_{tr}-1,\)
+where \(\Psi\) is a  class of  real-valued functions on \(\mathbb{R}^d\) and $n_tr$ is a fixed natural number. 
 Next  we estimate the coefficients \( \bar a_{l,k}\) using
@@
 where $n_{\mathrm{trunc}} \in \nset$ is a hyperparameter. Validity of this approach follows from the following lemma
-
-\begin{lem}
-\label{lem:coef_decrease}
-Let $(X^x_{p})_{p \in \nset_0}$ be a $V-$geometrically ergodic homogeneous Markov chain with stationary distribution $\pi$ and $f$ be a Borel function with $\|f\|_{V} < \infty$. Then it holds
-\begin{equation}
-|\bar a_{r,k}(x)| \leq C\rho^r\|f\|_{V} (\PE V^2(\Phi(x,\xi)))^{1/2}
-\end{equation}
-where $\bar a_{r,k}(x)$ is defined in \eqref{eq:definition-bar-a}.
-\end{lem}
-\begin{proof}
-Note that 
-\[
-\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
-\]
-with  $Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right], r\in \mathbb{N}$. $V-$ergodicity of $(X^x_{p})_{p \in \nset_0}$ implies that
-\[
-|Q_{r}(x) - \pi(f)| \leq |\int\limits_{\rset^d} f(y)\left(\P^r(x,dy)-\pi(dy)\right)| \leq C\rho^r V(x) \|f\|_{V}
-\]
-Thus, since $\PE\phi_k\left(\xi\right) = 0$, we obtain from Cauchy-Schwarz inequality
-\begin{eqnarray*}
-|\bar a_{r,k}(x)| \leq \left(\PE \phi^2_k\left(\xi\right)\right)^{1/2}\left(\PE |Q_{r}(\Phi(x,\xi)) - \pi(f)|^{2}\right)^{1/2} \leq C\rho^r\|f\|_{V} (\PE V^2(\Phi(x,\xi)))^{1/2}
-\end{eqnarray*}
-and the statement follows.
-\end{proof}
 
@@
 \end{proof} 
+\section{Proofs}
+\subsection{Proof of Lemma~\ref{lem:coef_decrease}}
+Note that 
+\[
+\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
+\]
+with  $Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right], r\in \mathbb{N}$. $V-$ergodicity of $(X^x_{p})_{p \in \nset_0}$ implies that
+\[
+|Q_{r}(x) - \pi(f)| \leq \left|\int\limits_{\rset^d} f(y)\left(\P^r(x,dy)-\pi(dy)\right)\right| \leq C\rho^r V(x) \|f\|_{V}
+\]
+Thus, since $\PE\phi_k\left(\xi\right) = 0$, we obtain from Cauchy-Schwarz inequality
+\begin{eqnarray*}
+|\bar a_{r,k}(x)| \leq \left(\PE \phi^2_k\left(\xi\right)\right)^{1/2}\left(\PE |Q_{r}(\Phi(x,\xi)) - \pi(f)|^{2}\right)^{1/2} \leq C\rho^r\|f\|_{V} (\PE V^2(\Phi(x,\xi)))^{1/2}
+\end{eqnarray*}
+and the statement follows.
 

### 2020-03-19 aap/Revision/vr-mcmc-revision-new-new.tex
update

--- 2020-03-12 aap/Revision/vr-mcmc-revision-new.tex
@@
 \def\NtrainPath{T}
-\def\TrainSet{\mathcal{T}}
+\def\TrainSet{\mathcal{D}}
 \modulolinenumbers[5]
@@
 In Section~\ref{sec:ula_analysis} we analyze the performance of the proposed variance reduction algorithm in the case of Unadjusted Langevin Algorithm (ULA).
-Section~\ref{sec:coeff} studies the complexity of the variance reduced ULA.
 Finally, numerical examples are presented in Section~\ref{sec:num}.
@@
 \end{example}
-\section{Martingale representation and variance reduction}
+\section{Martingale representation}
 \label{seq:mart_repr}
@@
 $\phi_k(\xi)$, $k\ge1$, are centered.
-As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of Shifted Legendre polynomials for "uniform part" and Hermite polynomials for "Gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the Shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\)  Let $\left(\Omega,\mathcal{F},\mathsf{P}\right)$ be a probability space, $(\xi_p)_{p \in \nset}$ be i.i.d.
+As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of shifted Legendre polynomials for "uniform part" and Hermite polynomials for "Gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\)
+\par
+Let  $(\xi_p)_{p \in \nset}$ be i.i.d.
 $m-$dimensional random vectors.
 We denote by $(\mathcal{G}_p)_{p \in \nset_0}$  the filtration generated by $(\xi_p)_{p \in \nset}$ with the convention $\mathcal{G}_0=\mathrm{triv}$.
-For  $x \in \rset^d, y \in \rset^{m}$ and $k \in \nset$ let $\Phi_k(x,y)$ be a function mapping $\rset^{d+m}$ to $\rset^d$.
-Then we can define for $l \le p$
+For  $x \in \rset^d, y \in \rset^{m}$ and $k \in \nset,$ let $\Phi_k(x,y)$ be a function mapping $\rset^{d+m}$ to $\rset^d$.
+Then we set for $l \le p$
 \begin{equation}
 \label{eq:ula_new}
-X^x_{l,p}=G_{l,p}(x,\xi_{l},\ldots,\xi_{p}),
-\end{equation}
-with the function \(G_{l,p}:\) \(\rset^{d + d_{\xi}\times(p-l+1)}\to \rset^{d}\) defined as
+X^x_{l,p}:=G_{l,p}(x,\xi_{l},\ldots,\xi_{p}),
+\end{equation}
+with the functions \(G_{l,p}:\) \(\rset^{d + m\times(p-l+1)}\to \rset^{d}\) defined as
 \begin{equation}
 \label{eq:definition-G-p-l}
-G_{l,p}(x,y_l,\ldots,y_p):=\Phi_p(\cdot,y_{p})\circ\Phi_{p-1}(\cdot,y_{p-1})\circ\dots\circ\Phi_{l}(x,y_{l})
+G_{l,p}(x,y_l,\ldots,y_p):=\Phi_p(\cdot,y_{p})\circ\Phi_{p-1}(\cdot,y_{p-1})\circ\dots\circ\Phi_{l}(x,y_{l}).
 \end{equation}
 Note that $\left(X^x_{0,p}\right)_{p \in \nset_0}$ is a Markov chain with values in $\rset^d$ of the form \eqref{eq:chain_gen}, starting at $X_0 = x$.
-We write $X^x_{p}$ and $G_{p}$ as a shorthand notation for $X^x_{0,p}$ and $G_{0,p}$, respectively.
-
-\begin{thm}\label{prop:29032018a1}
-For all $p \in \nset,$ $q \leq p$, $j < q \leq p$, Borel bounded functions $f: \rset^d \rightarrow \rset$ and $x \in \rset^d$ the following representation holds in \(L^2(\mathsf{P})\)
+In the sequel we write $X^x_{p}$ and $G_{p}$ as a shorthand notation for $X^x_{0,p}$ and $G_{0,p}$, respectively.
+
+\begin{thm}\label{thm:main-repr}
+For all $p \in \nset,$ $q \leq p$, $j < q \leq p$, any Borel bounded functions $f: \rset^d \rightarrow \rset$ and all $x \in \rset^d$ the following representation holds in \(L^2(\mathsf{P})\)
 \begin{eqnarray}
@@
 %a_{p,l,k}(x)=\mathsf{E}\left[\left.f(X_{p})\phi_k\left(\xi_{l}\right)\right|X_{l-1}=x\right], \quad p\geq l, \quad k\in \mathbb{N}.
-a_{q,l,k}(y)=\mathsf{E}\left[f(X^y_{l-1,q})\phi_k\left(\xi_{l}\right)\right], \quad q\geq l, \quad k\in \nset
+a_{q,l,k}(y)=\mathsf{E}\left[f(X^y_{l-1,q})\phi_k\left(\xi_{l}\right)\right], \quad q\geq l, \quad k\in \nset.
 \end{eqnarray}
@@
 
-\begin{proof}%[Proof of Theorem~\ref{prop:29032018a1}]
-The expansion obviously holds for $p = q=1$ and $j=0$.
-Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system in \(L^2(\mathbb{R}^d, P_{\xi})\), it holds in \(L^2(\mathsf{P})\) that
-\[
-f(X^x_{1})=\PE[f(X^x_{1})]+\sum_{k\geq1}a_{1,1,k}(x)\phi_{k}(\xi_{1})
-\]
-for any bounded $f$ with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$.
-Assume now that (\ref{eq:mart_repr}) holds for some $m < p$, any $q \leq m$, $j < q \leq m$ and all bounded $f$. Let us prove that the induction assumption holds for $q=m+1$ and all $j < q$.
-The orthonormality and completeness
-of the system $\left(\phi_{k}\right)_{k=0}^\infty$ implies that for any bounded $f$ and $y \in \rset^d$, 
-$\lim_{n \rightarrow \infty}\Psi_{n,m+1,1}(y)= 0$ with $\Psi_{n,m+1,1}(y)= \PE[|f(X^y_{m,m+1}) - f_{n,m+1,1}(y)|^2]$ and
-\[
-f_{n,m+1,1}(y) = \PE[f(X^y_{m,m+1})] +  \sum\limits_{k=1}^{n}\PE[f(X^y_{m,m+1})\phi_{k}(\xi_{m+1})]\phi_k(\xi_{m+1}).
-\]
-Note that  for any $y \in \rset^d$ and $n \in \nset$ it holds $\Psi_{n,m+1,1}(y) \leq \Psi_{0,m+1,1}(y)$ and
-\begin{eqnarray*}
-\PE \left[ \psi_{0,m+1,1}(X_m^x) \right] &=& \PE \left[ \left.
-\PE \left[ \psi_{0,m+1,1}(X_m^x) \right| \mathcal{G}_m\right]\right] = \PE \left[ \left.
-\PE \left[ |f(X^{X_m^x}_{m,m+1}) - \PE f(X^{X_m^x}_{m,m+1})|^2 \right| X_m^x\right]\right]
-\\
-&=& \PE \left[ |f(X_{m+1}^x) - \PE f(X_{m+1}^x)|^2 \right] = \PVar{f(X_{m+1}^x)} < \infty
-\end{eqnarray*}
-Hence, by Lebesgue dominated convergence theorem, $\lim_{n \rightarrow \infty}\PE \left[\psi_{n,m+1,1}(X_m^x)\right] = 0$ and since for all $y \in \rset^d$ the expectation $\PE[f(X^y_{m,m+1})]$ is a version of $\PE \left[f(X^y_{m,m+1}) | \mathcal{G}_m \right]$, it holds in \(L^2(\mathsf{P})\) that
-\begin{equation}
-\label{eq:1-step-decomp}
-f(X^x_{m+1}) = \PE[f(X^x_{m+1})|\mathcal{G}_m] + \sum\limits_{k=1}^{\infty}a_{m+1,m+1,k}(X^x_{m})\phi_{k}(\xi_{m+1})
-\end{equation}
-where
-\[
-a_{m+1,m+1,k}(y) = \PE \left[f(X^y_{m,m+1})\phi_k\left(\xi_{m+1}\right)\right]
-\]
-which is the required statement in the case $q = m+1$ and $j=m$.
-Now assume that $j<m$.
-Set $g(y) = \PE \left[f(X_{m,m+1}^y)\right]$.
-Note that $\mathsf{P}$-a.s.
-it holds $g(X_m^x) = \PE\left[\left.f(X_m^x) \right| \mathcal{G}_m\right]$ and $g$ is bounded by construction.
-Hence the induction hypothesis applies, and we get
-\begin{equation}\label{eq:28082017a1}
-\mathsf{E}\left[\left.f(X^x_{m+1})\right| \mathcal{G}_m \right]=\mathsf{E}\left[\left.f(X^x_{m+1})\right| \mathcal{G}_j \right]+\sum_{k\geq1}\sum_{l=j+1}^{m}a_{m+1,l,k}(X^x_{l-1})\phi_{k}(\xi_{l})
-\end{equation}
-with
-\begin{align*}
-a_{m+1,l,k}(X^x_{l-1}) &= \mathsf{E}\left[\left.\mathsf{E}\left[\left.f(X^x_{m+1})\right|\mathcal{G}_{m}\right]\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]
- = \mathsf{E}\left[\left.f(X^x_{m+1})\phi_{k}(\xi_{l})\right|X^x_{l-1}\right].
-\end{align*}
-where for $y \in \rset^d$,
-\[
-a_{m+1,l,k}(y) = \mathsf{E}\left[f(X^y_{l-1,m+1})\phi_{k}(\xi_{l})\right]
-\]
-Formulas \eqref{eq:1-step-decomp}
-and~\eqref{eq:28082017a1} conclude the induction step for $q = m+1$ and all $j < q$ and hence the proof.
-\end{proof}
-
 \begin{remark}\label{rem:general_moments} Denote by $\mathcal{F}_2^p$ class of Borel functions $f: \rset^d \to \rset$ such that for any $n \leq p$, any indices $1 \leq i_1 < \ldots < i_n \leq p$ and $y \in \rset^d$,
@@
 \end{equation}
-Then the statement of Theorem~\ref{prop:29032018a1} remains valid for $f \in \mathcal{F}_2^p$. 
+Then the statement of Theorem~\ref{thm:main-repr} remains valid for $f \in \mathcal{F}_2^p$. 
 \end{remark}
@@
 \begin{cor}
-Let \((X^x_{p})_{p\geq 0}\) be a homogeneous Markov chain of form \eqref{eq:ula_new} with $\Phi_l = \Phi,$ $l\geq 1$.
+Let \((X^x_{p})_{p\geq 0}\) be a homogeneous Markov chain of the form \eqref{eq:ula_new} with $\Phi_l = \Phi,$ $l\geq 1$.
 Then for all $p \in \nset,$ $q \leq p$, $j < q \leq p$, $f \in \mathcal{F}_{2,\text{hom}}^p$ and $x \in \rset^d$ it holds in \(L^2(\mathsf{P})\)
@@
 \end{eqnarray*}
-where for $y \in \rset^d$
+where for all $y \in \rset^d$
 \begin{eqnarray*}
@@
 
-From  numerical point of view another representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
-\begin{prop}
-Let $q\geq l, k\in \nset$. Then the coefficients \(a_{q,l,k}(x)\) in \eqref{eq:coeff_mart}  can be alternatively represented as
-\begin{eqnarray*}
-a_{q,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{q,l}\left(\Phi_l(x,\xi)\right)\right]
-\end{eqnarray*}
-with \(Q_{q,l}(x)=\mathsf{E}\left[f(X^x_{l,q})\right],\) \(q\geq l.\)
-%where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
-The functions \((Q_{q,l})_{l=0}^q\)  can be computed by the backward recurrence: $Q_{q,q}(x)=f(x)$ and for $l \in \{0,\ldots,q-1\}$
-\begin{eqnarray}
-\label{eq:qpl}
-Q_{q,l}(x)=\mathsf{E}\left[Q_{q,l+1}(X^x_{l+1})\right].
-%=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))]
-\end{eqnarray}
-In the case $\Phi_l=\Phi$ in \eqref{eq:ula_new} for all $l\geq 1,$ we have
-\begin{equation}
-\label{eq:definition-bar-a}
-\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
-\end{equation}
-with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$
-\end{prop}
-\par
 Next we show how  the representation \eqref{eq:mart_repr} can be used to  reduce the variance of MCMC algorithms.
-Consider the time homogeneous case ($\Phi_l = \Phi$ for all $l \in \nset$) and the sample average
+For the sake of clarity, in the sequel we consider only the time homogeneous case ($\Phi_l = \Phi$ for all $l \in \nset$).
+Define
 \begin{equation}\label{eq:29032018a2}
@@
 the number of effective samples.
-Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
+Fix some $K\in\mathbb N$ and denote
 \begin{eqnarray}
@@
 \]
-for any measurable function $g$  and hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
-The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
+for any measurable function $g.$  Hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
+The representation \eqref{eq:coeff_mart} suggests that the variance of  the   estimator
 \begin{equation}
@@
 \label{eq:definition-A-s,k}
-A_{q,k}(y)=\sum_{r=0}^q \bar{a}_{r,k}(y),\quad q\in \mathbb{N},
+A_{q,k}(y)=\sum_{r=0}^q \bar{a}_{r,k}(y),\quad q\in \mathbb{N}.
 \end{eqnarray}
-and $\mathsf{Var}[\pi_{K,n}^N(f)]$ is small provided that the coefficients $A_{s,k}$ decay fast enough as $k\to \infty.$
-In the next section   we provide a detailed   analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
+Hence $\mathsf{Var}[\pi_{K,n}^N(f)]$ is small provided that the coefficients $A_{s,k}$ decay fast enough as $k\to \infty.$
+In   Section~\ref{sec:ula_analysis}   we provide a detailed  theoretical  analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).   
 \section{Estimation of coefficients}
-The coefficients \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. This amounts to solve a nonparametric regression problem.
-We present a generic regression algorithm.
-Algorithm starts with estimating the functions \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) for $r=N+1,\ldots,N+n.$ We first generate \(\NtrainPath\)  paths  of the chain \(X\)
+The coefficients \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. This amounts to solving a nonparametric regression problem.
+We first present a generic regression algorithm and then in Section~\ref{sec:num} give further implementation details. Our algorithm starts with estimating the functions 
+$A_{q,k}$ for $q=N+1,\ldots,N+n,$ $k=1,\ldots,K. $ We first generate \(\NtrainPath\)  paths  of the chain \(X\)
 (the so-called ``training paths''):
 \[
-\TrainSet=\Bigl\{(X^{(s)}_{0},\ldots,X^{(s)}_{N+n}),\quad s=1,\ldots,\NtrainPath\Bigr\}
+\TrainSet=\Bigl\{(X^{(s)}_{N+1},\ldots,X^{(s)}_{N+n}),\quad s=1,\ldots,\NtrainPath\Bigr\}
 \]
 with $X^{(s)}_{0}=x,$ $s=1,\ldots,T.$
-The regression algorithm proceeds with estimating the  functions \((\bar Q_{l})\) by solving  the least squares optimization problems
+Then we solve   the least squares optimization problems
 \begin{equation}\label{eq:06042018a1}
-\widehat{Q}_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
-\end{equation}
-for \(r=1,\ldots, n-1,\)
-where \(\Psi\) is a  class of  real-valued functions on \(\mathbb{R}^d.\)
-Next  we estimate the coefficients \( \bar a_{l,k}\) using
-the formula
-\begin{equation}
-\label{eq:a_est_int}
-\widehat a_{l,k}(x)=\mathsf{E}\left[\left.\phi_k\left(\xi\right)\widehat Q_{l}\left(\Phi(x,\xi)\right)\right | \TrainSet\right] \,,
-\end{equation}
-where $\xi$ is a random variable independent from $\mathcal{T}$ with law $P_\xi,$
-$\Phi$ is defined in \eqref{eq:chain_gen} and the expectation is taken according to known distribution \(P_\xi\).
-Conditioning on $\TrainSet$ underlines that estimation of $\widehat Q_{l}$ derived from training trajectories.
-In many cases integration can be done in closed analytical form.
-For example, in the case of ULA algorithm we have \(P_\xi$ is the multivariate normal distribution $ \mathcal{N}(0,I_d)\) and \(\Phi(x,\xi)=x-\gamma\mu(x)+\sqrt{\gamma}\xi.\) Therefore if we use polynomials to approximate \((\bar Q_{l}),\) then \((\widehat a_{l,k})\) can be even computed in closed form; implementations details are provided in  Section~\ref{sec:num}.
-Upon estimating  the coefficients \((\widehat a_{l,k})\), one can construct an empirical estimate  of \(M_{K,n}^N(f)\) in the form
-\begin{eqnarray*}
-\widehat M_{K,n}^N(f) = \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\widehat{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
-\end{eqnarray*}
-Obviously, \(\mathsf{E}[\widehat M_{K,n}^N(f)|\TrainSet]=0\) and \(\widehat M_{K,n}^N(f)\) is indeed a valid control variate in that it
+\widetilde{A}_{q,k} \in \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \left| \sum_{r = 0}^{q}  f(X_{r+L}^{(s)})\phi_k(\xi^{(s)}_{L+1}) - \psi(X^{(s)}_L)\right|^2
+\end{equation}
+for some $L$ with $L+q\leq N+n,$
+where \(\Psi\) is a  class of  real-valued functions on \(\mathbb{R}^d\) and $\xi^{(s)}_{N+1},\ldots,\xi^{(s)}_{N+n}$ are iid random variables with distribution $P_\xi$ used to construct the $s$-th trajectory of $X.$   As usual in regression analysis (see Section~11 in \cite{gyorfi2006distribution})  we  truncate  the least squares estimates $(\widetilde{A}_{q,k})$ and set
+\begin{eqnarray}
+\label{eq:trunc}
+\widehat{A}_{q,k}(y)=
+\begin{cases}
+\widetilde{A}_{q,k}(y), & |\widetilde{A}_{q,k}(y)|\leq U_k,
+\\
+U_k \,\mathrm{sign}(\widetilde{A}_{q,k}(y)), & \mathrm{otherwise}. 
+\end{cases}
+\end{eqnarray}
+ for a sequence of positive real numbers $U_k,$ $k=1,\ldots,K.$
+Upon estimating  the functions \((A_{q,k})\), one can construct the empirical estimate  of \(M_{K,n}^N(f)\) in the form
+\begin{eqnarray*}
+\widehat M_{K,n}^N(f) = \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\widehat{A}_{N+n-l,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
+\end{eqnarray*}
+Obviously, \(\mathsf{E}[\widehat M_{K,n}^N(f)|\TrainSet]=0\) and \(\widehat M_{K,n}^N(f)\) is a valid control variate in that it
 does not introduce any bias. By the Jensen inequality and orthonormality of \((\phi_k)_{k\geq 0},\)
 \begin{multline*}
-\mathsf{E}_x\left[\left.\left|\widehat{M}_{K,n}^{N}(f)-M_{K,n}^{N}(f)\right|^{2}\right | \TrainSet \right]
+\mathsf{E}\left[\left.\left|\widehat{M}_{K,n}^{N}(f)-M_{K,n}^{N}(f)\right|^{2}\right | \TrainSet \right]
 \\
@@
 \end{multline*}
-Yet it is impractical to use estimator of the form $\widehat M_{K,n}^N(f)$, since it may be too expensive to compute for $n$ large enough.
-Instead we introduce the truncated version of this control variate
-\begin{eqnarray*}
-\tilde M_{K,n}^N(f) = \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\widehat{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\mathbbm{1}\{ |p-k| < n_{\mathrm{trunc}} \}\right]
-\end{eqnarray*}
-where $n_{\mathrm{trunc}} \in \nset$ is a hyperparameter. Validity of this approach follows from the following lemma
-
-\begin{lem}
-\label{lem:coef_decrease}
-Let $(X^x_{p})_{p \in \nset_0}$ be a $V-$geometrically ergodic homogeneous Markov chain with stationary distribution $\pi$ and $f$ be a Borel function with $\|f\|_{V} < \infty$. Then it holds
-\begin{equation}
-|\bar a_{r,k}(x)| \leq C\rho^r\|f\|_{V} (\PE V^2(\Phi(x,\xi)))^{1/2}
-\end{equation}
-where $\bar a_{r,k}(x)$ is defined in \eqref{eq:definition-bar-a}.
-\end{lem}
-\begin{proof}
-Note that 
-\[
-\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
-\]
-with  $Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right], r\in \mathbb{N}$. $V-$ergodicity of $(X^x_{p})_{p \in \nset_0}$ implies that
-\[
-|Q_{r}(x) - \pi(f)| \leq |\int\limits_{\rset^d} f(y)\left(\P^r(x,dy)-\pi(dy)\right)| \leq C\rho^r V(x) \|f\|_{V}
-\]
-Thus, since $\PE\phi_k\left(\xi\right) = 0$, we obtain from Cauchy-Schwarz inequality
-\begin{eqnarray*}
-|\bar a_{r,k}(x)| \leq \left(\PE \phi^2_k\left(\xi\right)\right)^{1/2}\left(\PE |Q_{r}(\Phi(x,\xi)) - \pi(f)|^{2}\right)^{1/2} \leq C\rho^r\|f\|_{V} (\PE V^2(\Phi(x,\xi)))^{1/2}
-\end{eqnarray*}
-and the statement follows.
-\end{proof}
-
+Set
+\begin{eqnarray*}
+\widehat\pi_{K,n}^N(f):=\pi_n^N(f)-\widehat{M}_{K,n}^N(f)
+\end{eqnarray*}
+Combining this with \eqref{eq:expression-variance} results in 
+\begin{multline*}
+\PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]=\frac{1}{n^2}\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})]
+\\
++\frac{1}{n^2}\sum_{k=1}^{K}\sum_{l=1}^{n}\PE\left[\left.|A_{n-l,k}(X^x_{N+l-1})-\widehat A_{n-l,k}(X^x_{N+l-1})|^2\right | \TrainSet \right].
+\end{multline*}
+In the next section we analyze the case of ULA and show that under some rather mild conditions (convexity of the potential $U$ outside a ball around zero)
+\begin{eqnarray*}
+\PE[A^2_{l,k}(X^x_{N+l-1})]\leq R^2_k,\quad k=1,2,\ldots, 
+\end{eqnarray*}
+for a sequence  $(R_k)$ satisfying $\sum_{k=1}^\infty R^2_k\leq C$ with constant $C$  depending neither on $\gamma$ nor on $n$ (see \eqref{eq:Aqk2}).
+Hence if we take $U_k=R_k$ as truncation parameters in \eqref{eq:trunc}, then $\widehat A_{n-l,k}$ converges to $A_{n-l,k}$ as $T\to \infty.$ Moreover,
+\begin{eqnarray*}
+\PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]\leq \frac{2C}{n}, 
+\end{eqnarray*}
+for a constant $C$ not depending on $n$ and $\gamma.$
+If the class of functions $\Psi$ is linear of dimension $\mathrm{D}_\Psi$ As can be easily seen the cost of computing $\widehat \pi_{K,n}^N(f)$ is of order $ K T^2 n$.
+At the same time the variance of the standard estimate $\widehat \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved in general.
+Thus the ratio of cost-variances is of order
+\begin{eqnarray*}
+\frac{\mathrm{cost}(\pi_{K,n}^N) \PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]}{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}\leq C \gamma K T^2 L^2 \mathrm{Dim}(\Psi). 
+\end{eqnarray*}
+If $K,$ $L$ and $\mathrm{Dim}(\Psi)$ are fixed, then our variance reduction method is advantageous if $\gamma=O(1/T^2).$ 
 \section{Analysis of variance reduced ULA}
@@
 \end{align}
-with \(\|\mathbf{k}\|=\max_{i} k_i.\)
+with \(\|\mathbf{k}\|=\max_{i} k_i\) and 
+\begin{eqnarray*}
+A_{q,\mathbf{k}}(y)=\sum_{r=0}^{ q } \bar{a}_{r,\mathbf{k}}(y).
+\end{eqnarray*}
 For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
@@
 \label{eq:bias_1}
+\nonumber
 \left( \mathsf{E}[\pi_{K,n}^N(f)]-\pi(f)\right)^2
@@
 =
-\frac{1}{n^2}\left[\sum\limits_{k=N+1}^{N+k}\left(f(X_k^x)-\pi(f)\right)\right]^2 \leq \\
+\frac{1}{n^2}\left[\sum\limits_{k=N+1}^{N+k}\left(f(X_k^x)-\pi(f)\right)\right]^2 \\
 &\leq& 
@@
 \end{equation}
-for  $V(x) = 1 + \|x\|^2,$ $\rho\in (0,1)$ specified in Lemma~\ref{13} and constants $C_1,C_2>0$ independent of $\gamma$  and $n$.
+for  $V(x) = 1 + \|x\|^2,$ $\rho\in (0,1)$ specified in Lemma~\ref{lem:v_ergodicity} and constants $C_1,C_2>0$ independent of $\gamma$  and $n$.
 Since $f$ is bounded, $\|f\|_V \leq \|f\|_{\infty}$.
@@
 Then it holds
+
 \begin{equation}
@@
 \end{equation}
-Then  it holds
+Suppose that $n\gamma^2\leq \Gamma.$ Set $V(x) = 1 + \|x\|^2,$ then  it holds
+\begin{eqnarray}
+\label{eq:Aqk2}
+\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K} A^2_{q,\mathbf{k}}(x)\leq C_{K}\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
+\left(\frac{\gamma}{2}\right)^{|I|K-1}
+(1+V(x))
+\end{eqnarray}
+for some constant $C_K$ not depending on $q$ and depending on $\gamma$ only through $\Gamma$. 
+\end{thm}
+\begin{cor}
+Under assumptions of Theorem~\ref{th:mr},
 \begin{eqnarray}
 \label{eq:var-bound}
-\PVar\left(\pi_{K,n}^{N}(f)\right)\lesssim n^{-1} \gamma^{K-1} \eqsp,
+\PVar\left(\pi_{K,n}^{N}(f)\right)\lesssim n^{-1} \gamma^{K-1} 
+\eqsp,
 \end{eqnarray}
-provided $n\gamma^2$ is bounded.
-%where \tcr{\(\lesssim\) stands for inequality up to a constant not depending  on $\gamma,$ \(n\) and \(N.\)}
-\end{thm}
+where $\lesssim$ stands for inequality up to a constant depending only on $K$ and $\Gamma$.
+\end{cor}
 Let us sketch the main steps of the proof. First using integration by parts, we prove that
@@
 \begin{eqnarray*}
-\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1} A^2_{q,\mathbf{k}}(x)\leq \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
+\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K} A^2_{q,\mathbf{k}}(x)\leq \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
 \left(\frac{\gamma}{2}\right)^{|I|K}
@@
  Finally, we show  that under our smoothness assumption \eqref{eq:smooth-mu} it holds
-\begin{equation*}
+\begin{equation}
+\label{eq:varboundV}
 \PVar \left(\sum_{p=1}^{q}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X^x_{p}\right)
-\right)\leq C_{K}\gamma^{-1}(1+V(x)), \quad V(x) = 1 + \|x\|^2
-\end{equation*}
+\right)\leq C_{K}\gamma^{-1}(1+V(x))
+\end{equation}
 for some constant $C_K$ not depending on $q$ and $\gamma$.
-\section{Numerical results}
+\section{Numerical analysis}
 \label{sec:num}
@@
 \begin{eqnarray*}
-\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}A^2_{s,\mathbf{k}}(x)
+\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K}A^2_{s,\mathbf{k}}(x)
 &\leq & \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
@@
 \end{proof} 
+\section{Proofs}
+\subsection{Proof of Theorem~\ref{thm:main-repr}}
+The expansion obviously holds for $p = q=1$ and $j=0$.
+Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system in \(L^2(\mathbb{R}^d, P_{\xi})\), it holds in \(L^2(\mathsf{P})\) that
+\[
+f(X^x_{1})=\PE[f(X^x_{1})]+\sum_{k\geq1}a_{1,1,k}(x)\phi_{k}(\xi_{1})
+\]
+for any bounded $f$ with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$.
+Assume now that (\ref{eq:mart_repr}) holds for some $m < p$, any $q \leq m$, $j < q \leq m$ and all bounded $f$. Let us prove that the induction assumption holds for $q=m+1$ and all $j < q$.
+The orthonormality and completeness
+of the system $\left(\phi_{k}\right)_{k=0}^\infty$ implies that for any bounded $f$ and $y \in \rset^d$, 
+$\lim_{n \rightarrow \infty}\Psi_{n,m+1,1}(y)= 0$ with $\Psi_{n,m+1,1}(y)= \PE[|f(X^y_{m,m+1}) - f_{n,m+1,1}(y)|^2]$ and
+\[
+f_{n,m+1,1}(y) = \PE[f(X^y_{m,m+1})] +  \sum\limits_{k=1}^{n}\PE[f(X^y_{m,m+1})\phi_{k}(\xi_{m+1})]\phi_k(\xi_{m+1}).
+\]
+Note that  for any $y \in \rset^d$ and $n \in \nset$ it holds $\Psi_{n,m+1,1}(y) \leq \Psi_{0,m+1,1}(y)$ and
+\begin{eqnarray*}
+\PE \left[ \psi_{0,m+1,1}(X_m^x) \right] &=& \PE \left[ \left.
+\PE \left[ \psi_{0,m+1,1}(X_m^x) \right| \mathcal{G}_m\right]\right] = \PE \left[ \left.
+\PE \left[ |f(X^{X_m^x}_{m,m+1}) - \PE f(X^{X_m^x}_{m,m+1})|^2 \right| X_m^x\right]\right]
+\\
+&=& \PE \left[ |f(X_{m+1}^x) - \PE f(X_{m+1}^x)|^2 \right] = \PVar{f(X_{m+1}^x)} < \infty
+\end{eqnarray*}
+Hence, by Lebesgue dominated convergence theorem, $\lim_{n \rightarrow \infty}\PE \left[\psi_{n,m+1,1}(X_m^x)\right] = 0$ and since for all $y \in \rset^d$ the expectation $\PE[f(X^y_{m,m+1})]$ is a version of $\PE \left[f(X^y_{m,m+1}) | \mathcal{G}_m \right]$, it holds in \(L^2(\mathsf{P})\) that
+\begin{equation}
+\label{eq:1-step-decomp}
+f(X^x_{m+1}) = \PE[f(X^x_{m+1})|\mathcal{G}_m] + \sum\limits_{k=1}^{\infty}a_{m+1,m+1,k}(X^x_{m})\phi_{k}(\xi_{m+1})
+\end{equation}
+where
+\[
+a_{m+1,m+1,k}(y) = \PE \left[f(X^y_{m,m+1})\phi_k\left(\xi_{m+1}\right)\right]
+\]
+which is the required statement in the case $q = m+1$ and $j=m$.
+Now assume that $j<m$.
+Set $g(y) = \PE \left[f(X_{m,m+1}^y)\right]$.
+Note that $\mathsf{P}$-a.s.
+it holds $g(X_m^x) = \PE\left[\left.f(X_m^x) \right| \mathcal{G}_m\right]$ and $g$ is bounded by construction.
+Hence the induction hypothesis applies, and we get
+\begin{equation}\label{eq:28082017a1}
+\mathsf{E}\left[\left.f(X^x_{m+1})\right| \mathcal{G}_m \right]=\mathsf{E}\left[\left.f(X^x_{m+1})\right| \mathcal{G}_j \right]+\sum_{k\geq1}\sum_{l=j+1}^{m}a_{m+1,l,k}(X^x_{l-1})\phi_{k}(\xi_{l})
+\end{equation}
+with
+\begin{align*}
+a_{m+1,l,k}(X^x_{l-1}) &= \mathsf{E}\left[\left.\mathsf{E}\left[\left.f(X^x_{m+1})\right|\mathcal{G}_{m}\right]\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]
+ = \mathsf{E}\left[\left.f(X^x_{m+1})\phi_{k}(\xi_{l})\right|X^x_{l-1}\right].
+\end{align*}
+where for $y \in \rset^d$,
+\[
+a_{m+1,l,k}(y) = \mathsf{E}\left[f(X^y_{l-1,m+1})\phi_{k}(\xi_{l})\right]
+\]
+Formulas \eqref{eq:1-step-decomp}
+and~\eqref{eq:28082017a1} conclude the induction step for $q = m+1$ and all $j < q$ and hence the proof.
+
+\subsection{Proof of Lemma~\ref{lem:coef_decrease}}
+Note that 
+\[
+\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
+\]
+with  $Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right], r\in \mathbb{N}$. $V-$ergodicity of $(X^x_{p})_{p \in \nset_0}$ implies that
+\[
+|Q_{r}(x) - \pi(f)| \leq \left|\int\limits_{\rset^d} f(y)\left(\P^r(x,dy)-\pi(dy)\right)\right| \leq C\rho^r V(x) \|f\|_{V}
+\]
+Thus, since $\PE\phi_k\left(\xi\right) = 0$, we obtain from Cauchy-Schwarz inequality
+\begin{eqnarray*}
+|\bar a_{r,k}(x)| \leq \left(\PE \phi^2_k\left(\xi\right)\right)^{1/2}\left(\PE |Q_{r}(\Phi(x,\xi)) - \pi(f)|^{2}\right)^{1/2} \leq C\rho^r\|f\|_{V} (\PE V^2(\Phi(x,\xi)))^{1/2}
+\end{eqnarray*}
+and the statement follows.
 
@@
 \end{document}
+
+From  numerical point of view another representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
+\begin{prop}
+Let $q\geq l, k\in \nset$. Then the coefficients \(a_{q,l,k}(x)\) in \eqref{eq:coeff_mart}  can be alternatively represented as
+\begin{eqnarray*}
+a_{q,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{q,l}\left(\Phi_l(x,\xi)\right)\right]
+\end{eqnarray*}
+with \(Q_{q,l}(x)=\mathsf{E}\left[f(X^x_{l,q})\right],\) \(q\geq l.\)
+%where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
+The functions \((Q_{q,l})_{l=0}^q\)  can be computed by the backward recurrence: $Q_{q,q}(x)=f(x)$ and for $l \in \{0,\ldots,q-1\}$
+\begin{eqnarray}
+\label{eq:qpl}
+Q_{q,l}(x)=\mathsf{E}\left[Q_{q,l+1}(X^x_{l+1})\right].
+%=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))]
+\end{eqnarray}
+In the case $\Phi_l=\Phi$ in \eqref{eq:ula_new} for all $l\geq 1,$ we have
+\begin{equation}
+\label{eq:definition-bar-a}
+\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
+\end{equation}
+with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$
+\end{prop}
+\par

### 2020-03-19 aap/Revision/vr-mcmc-revision-new.tex
update
lower bound

@@
 \end{equation}
-where $\rho\in (0,1)$ and $C$ is an absolute constant.
+where $\rho\in (0,1)$ and $C$ is an absolute constant. In particular, if $f$ is uniformly bounded we can take $V(x)\equiv 1$ to get
+\begin{eqnarray*}
+|\bar a_{r,k}(x)| \leq C\rho^r\|f\|_{\infty}.
+\end{eqnarray*}
+
 \end{lem}
-
-\section{Estimation of coefficients}
+This lemma shows that in fact  only a logarithmic number (in $n$) of coefficients need to be computed.
+
+\section{Estimation of coefficients and variance reduction}
 The coefficients \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. This amounts to solve a nonparametric regression problem.
-We present a generic regression algorithm.
-Algorithm starts with estimating the functions \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) for $r=N+1,\ldots,N+n.$ We first generate \(\NtrainPath\)  paths  of the chain \(X\)
+We present a generic regression algorithm. Algorithm starts with estimating the functions 
+\[
+S_{q}(x)=\mathsf{E}\left[\sum_{r=0}^q f(X^x_{r})\right]
+\]
+for $q=N+1,\ldots,N+n.$ We first generate \(\NtrainPath\)  paths  of the chain \(X\)
 (the so-called ``training paths''):
 \[
-\TrainSet=\Bigl\{(X^{(s)}_{0},\ldots,X^{(s)}_{N+n}),\quad s=1,\ldots,\NtrainPath\Bigr\}
+\TrainSet=\Bigl\{(X^{(s)}_{N+1},\ldots,X^{(s)}_{N+n}),\quad s=1,\ldots,\NtrainPath\Bigr\}
 \]
@@
 \begin{equation}\label{eq:06042018a1}
-\widehat{Q}_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n_{tr}-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
-\end{equation}
-for \(r=1,\ldots, n_{tr}-1,\)
-where \(\Psi\) is a  class of  real-valued functions on \(\mathbb{R}^d\) and $n_tr$ is a fixed natural number. 
-Next  we estimate the coefficients \( \bar a_{l,k}\) using
+\widehat{S}_q = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l=N+1}^{L}\left| \sum_{r = 0}^{q}  f(X_{r+l}^{(s)}) - \psi(X^{(s)}_l)\right|^2
+\end{equation}
+for some $L$ with $L+q\leq N+n,$
+where \(\Psi\) is a  class of  real-valued functions on \(\mathbb{R}^d\) and set  $\widehat{S}_{R,q}=\max\{\widehat{S}_q,R\}$ for some $R>0.$
+Next  we estimate the coefficients \( A_{q,k}\) using
 the formula
@@
 \label{eq:a_est_int}
-\widehat a_{l,k}(x)=\mathsf{E}\left[\left.\phi_k\left(\xi\right)\widehat Q_{l}\left(\Phi(x,\xi)\right)\right | \TrainSet\right] \,,
+\widehat A_{q,k}(x)=\mathsf{E}\left[\left.\phi_k\left(\xi\right)\widehat S_{R,q}\left(\Phi(x,\xi)\right)\right | \TrainSet\right] \,,
 \end{equation}
@@
 $\Phi$ is defined in \eqref{eq:chain_gen} and the expectation is taken according to known distribution \(P_\xi\).
-Conditioning on $\TrainSet$ underlines that estimation of $\widehat Q_{l}$ derived from training trajectories.
+Conditioning on $\TrainSet$ underlines that estimation of $\widehat S_{R,q}$ derived from training trajectories.
 In many cases integration can be done in closed analytical form.
-For example, in the case of ULA algorithm we have \(P_\xi$ is the multivariate normal distribution $ \mathcal{N}(0,I_d)\) and \(\Phi(x,\xi)=x-\gamma\mu(x)+\sqrt{\gamma}\xi.\) Therefore if we use polynomials to approximate \((\bar Q_{l}),\) then \((\widehat a_{l,k})\) can be even computed in closed form; implementations details are provided in  Section~\ref{sec:num}.
-Upon estimating  the coefficients \((\widehat a_{l,k})\), one can construct an empirical estimate  of \(M_{K,n}^N(f)\) in the form
-\begin{eqnarray*}
-\widehat M_{K,n}^N(f) = \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\widehat{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
+For example, in the case of ULA algorithm we have \(P_\xi$ is the multivariate normal distribution $ \mathcal{N}(0,I_d)\) and \(\Phi(x,\xi)=x-\gamma\mu(x)+\sqrt{\gamma}\xi.\) Therefore if we use polynomials to approximate \((S_{q}),\) then \((\widehat A_{q,k})\) can be even computed in closed form; implementations details are provided in  Section~\ref{sec:num}.
+Upon estimating  the functions \((\widehat A_{q,k})\), one can construct an empirical estimate  of \(M_{K,n}^N(f)\) in the form
+\begin{eqnarray*}
+\widehat M_{K,n}^N(f) = \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\widehat{A}_{N+n-l,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
 \end{eqnarray*}
@@
 \begin{multline*}
-\mathsf{E}_x\left[\left.\left|\widehat{M}_{K,n}^{N}(f)-M_{K,n}^{N}(f)\right|^{2}\right | \TrainSet \right]
+\mathsf{E}\left[\left.\left|\widehat{M}_{K,n}^{N}(f)-M_{K,n}^{N}(f)\right|^{2}\right | \TrainSet \right]
 \\
@@
 \end{multline*}
-Yet it is impractical to use estimator of the form $\widehat M_{K,n}^N(f)$, since it may be too expensive to compute for $n$ large enough.
-Instead we introduce the truncated version of this control variate
-\begin{eqnarray*}
-\tilde M_{K,n}^N(f) = \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\widehat{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\mathbbm{1}\{ |p-k| < n_{\mathrm{trunc}} \}\right]
-\end{eqnarray*}
-where $n_{\mathrm{trunc}} \in \nset$ is a hyperparameter. Validity of this approach follows from the following lemma
-
+Combining this with \eqref{eq:expression-variance} results in 
+\begin{multline*}
+\PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]=\frac{1}{n^2}\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})]
+\\
++\frac{1}{n^2}\sum_{k=1}^{K}\sum_{l=1}^{n}\PE\left[\left.|A_{n-l,k}(X^x_{N+l-1})-\widehat A_{n-l,k}(X^x_{N+l-1})|^2\right | \TrainSet \right]
+\end{multline*}
+
+Suppose that $\|f\|_\infty\leq F$ and $\|A_{l,k}\|_{\infty}\leq R$ for all $k=1,\ldots,K,$ $l=1,2,\ldots $ and some constants $F,K>0.$ The standard regression estimates imply that 
+\begin{eqnarray*}
+\PE\left[\left.|A_{n-l,k}(X^x_{N+l-1})-\widehat A_{n-l,k}(X^x_{N+l-1})|^2\right | \TrainSet \right]&\lesssim &  \frac{(R^2 \vee F^2)\,\mathrm{Dim}(\Psi)}{T}
+\\
+&&+\sup_{\psi\in\Psi}\|A_{n-l,k}-\psi\|_{L^2(P^x_{N+l-1})}
+\end{eqnarray*}
+with high probability.
 \section{Analysis of variance reduced ULA}
@@
 
+Note that the bound of lemma~\ref{lem:variance} is sharp, namely, the following statement holds:
+\begin{lem} 
+\label{lem:ula_var_lower_bound}
+Let $f(x) = x$ and $(X^x_{p})_{p \in \nset_0}$ be a Markov chain generated by ULA with constant step size $\gamma$, with $\pi(x) = \frac{1}{\sqrt{2\pi}}e^{-\frac{x^2}{2}}$ and assume also $n\gamma > C_0$ for some fixed $C_0$.
+Then it holds
+\begin{equation}
+\label{eq:lower_bound}
+\PVar[\pi_n^0(f)] \geq \frac{2}{n\gamma} - \mathcal{O}\left(\frac{1}{n^2\gamma^2}\right) \eqsp,
+\end{equation}
+\end{lem}
+
 One of the main results of this paper is the following  upper bound for the variance of variance-reduced estimator $\pi_{K,n}^{N}(f).$
@@
 \label{eq:var-bound}
-\PVar\left(\pi_{K,n}^{N}(f)\right)\lesssim n^{-1} \gamma^{K-1} \eqsp,
+\PVar\left(\pi_{K,n}^{N}(f)\right)\lesssim n^{-1} \gamma^{K-1} 
+\eqsp,
 \end{eqnarray}
@@
 \end{proof}
+
 \section{Proofs}
@@
 and the statement follows.
+
+\subsection{Proof of Lemma~\ref{lem:ula_var_lower_bound}}
+Note that ULA for the standard normal distribution takes form
+\begin{eqnarray*}
+X_{k+1} = (1-\gamma)X_k + \sqrt{2\gamma}\xi_{k+1},\, \xi_{k+1} \sim \mathcal{N}(0,1)
+\end{eqnarray*}
+Thus, setting $X_0 = 0$, we obtain
+\begin{eqnarray*}
+\PVar_{0} \pi_n^0(f) &=& \PVar_{0}\left[\frac{1}{n}\sum\limits_{i=1}^{n}X_i\right] = \frac{1}{n^2}\PVar\left[\sum\limits_{i=1}^{n}\sum\limits_{j=1}^{i}\sqrt{2\gamma}(1-\gamma)^{i-j}\xi_{j}\right] = \frac{1}{n^2}\PVar\left[\sum\limits_{j=1}^{n}\xi_j\frac{\sqrt{2\gamma}(1-(1-\gamma)^{n-j+1}}{\gamma}\right] = \\
+&=&\frac{2}{n^2\gamma}\sum\limits_{j=1}^{n}\left(1-(1-\gamma)^{n-j+1}\right)^2 = \frac{2}{n\gamma} - \frac{4(1-\gamma)\left[1-(1-\gamma)^{n+1}\right]}{n^2\gamma^2} + \frac{2(1-\gamma)^2\left[1-(1-\gamma)^{2n+2}\right]}{n^2\gamma^2(2-\gamma)} \geq \\
+& \geq & \frac{2}{n\gamma} - \frac{4(1-\gamma)}{n^2\gamma^2} \geq \frac{2}{n\gamma} - \frac{4}{n^2\gamma^2}
+\end{eqnarray*}
 

### 2020-03-19 Code/martingale.py

@@ -54,7 +54,7 @@ def approx_q(X_train,Y_train,N_traj_train,lag,max_deg):
     Args:
         X_train - train tralectory;
         Y_train - function values;
-        N_traj_train - number of teraining trajectories;
+        N_traj_train - number of training trajectories;
         lag - truncation point for coefficients, those for |p-l| > lag are set to 0;
         max_deg - maximum degree of polynomial in regression
     """
@@ -88,6 +88,31 @@ def approx_q(X_train,Y_train,N_traj_train,lag,max_deg):
         else:
             coefs_poly = np.concatenate((coefs_poly,coefs),axis=0)
     return coefs_poly
+
+def approx_q_independent(X_train,Y_train,N_traj_train,lag,max_deg):
+    """
+    Function to regress q on a polynomial basis based on independent trajectories
+    """
+    dim = X_train[0,:].shape[0]
+    print("dimension = ",dim)
+    coefs_poly = np.array([])
+    for i in range(lag):
+        x_all = X_train[:,0,:]
+        y_all = Y_train[:,i,:]
+        #should use polyfeatures here
+        poly = PolynomialFeatures(max_deg)
+        X_features = poly.fit_transform(x_all)
+        print(X_features.shape)
+        lstsq_results = np.linalg.lstsq(X_features,y_all,rcond = None)
+        coefs = copy.deepcopy(lstsq_results[0])
+        coefs.resize((1,X_features.shape[1]))           
+        if coefs_poly.size == 0:
+            coefs_poly = copy.deepcopy(coefs)
+        else:
+            coefs_poly = np.concatenate((coefs_poly,coefs),axis=0)
+    return coefs_poly
+    
+    
 
 def get_indices_poly(ind,K_max,S_max):
     """

### 2020-03-19 Code/samplers.py

@@ -110,7 +110,7 @@ def ULA_light(r_seed,Potential,step, N, n, d, return_noise = False, x0 = None, f
     #traj_grad = np.zeros((n, d))
     if return_noise:
         noise = np.zeros((n, d))
-    x = np.random.normal(scale=5.0, size=d) # initial value X_0
+    x = np.random.normal(scale=10.0, size=d) # initial value X_0
     for k in np.arange(N): # burn-in period
         grad_burn_val = grad_burn(x)
         x = x + step * grad_burn_val +\
@@ -118,7 +118,7 @@ def ULA_light(r_seed,Potential,step, N, n, d, return_noise = False, x0 = None, f
     #burn-in ended
     if fixed_start: #start all test trajectories from the same point
         x = x0
-    traj[0,] = x0
+    traj[0,] = x
     grad_main = Potential.gradpotential
     for k in np.arange(1,n): # samples
         grad = grad_main(x)

### 2020-03-21 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 Monte Carlo and Markov Chain Monte Carlo methods.
-Introduction to many of the variance reduction techniques can be found in \cite{christian1999monte}, \cite{rubinstein2016simulation}, \cite{GobetBook} and \cite{glasserman2013monte}.
+Introduction to many of the variance reduction techniques can be found in \cite{rubinstein2016simulation}, \cite{GobetBook} and \cite{glasserman2013monte}.
 Recently one witnessed a revival of interest in efficient variance reduction methods  for MCMC algorithms, see for example \cite{dellaportas2012control}, \cite{mira2013zero}, \cite{brosse2018diffusion} and references therein.
@@
 \end{cor}
-
+\section{Variance reduction}
 Next we show how  the representation \eqref{eq:mart_repr} can be used to  reduce the variance of MCMC algorithms.
@@
 In   Section~\ref{sec:ula_analysis}   we provide a detailed  theoretical  analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).   
-\section{Estimation of coefficients}
+\par
 The coefficients \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. This amounts to solving a nonparametric regression problem.
@@
 \begin{equation}\label{eq:06042018a1}
-\widetilde{A}_{q,k} \in \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \left| \sum_{r = 0}^{q}  f(X_{r+L}^{(s)})\phi_k(\xi^{(s)}_{L+1}) - \psi(X^{(s)}_L)\right|^2
-\end{equation}
-for some $L$ with $L+q\leq N+n,$
+\widetilde{A}_{q,k} \in \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \left| \sum_{r = 0}^{q}  f(X_{r+l_0}^{(s)})\phi_k(\xi^{(s)}_{l_0+1}) - \psi(X^{(s)}_{l_0})\right|^2
+\end{equation}
+for some $l_0$ with $l_0+q\leq N+n,$
 where \(\Psi\) is a  class of  real-valued functions on \(\mathbb{R}^d\) and $\xi^{(s)}_{N+1},\ldots,\xi^{(s)}_{N+n}$ are iid random variables with distribution $P_\xi$ used to construct the $s$-th trajectory of $X.$   As usual in regression analysis (see Section~11 in \cite{gyorfi2006distribution})  we  truncate  the least squares estimates $(\widetilde{A}_{q,k})$ and set
@@
 \begin{eqnarray*}
-\PE[A^2_{l,k}(X^x_{N+l-1})]\leq R^2_k,\quad k=1,2,\ldots, 
-\end{eqnarray*}
-for a sequence  $(R_k)$ satisfying $\sum_{k=1}^\infty R^2_k\leq C$ with constant $C$  depending neither on $\gamma$ nor on $n$ (see \eqref{eq:Aqk2}).
-Hence if we take $U_k=R_k$ as truncation parameters in \eqref{eq:trunc}, then $\widehat A_{n-l,k}$ converges to $A_{n-l,k}$ as $T\to \infty.$ Moreover,
-\begin{eqnarray*}
-\PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]\leq \frac{2C}{n}, 
-\end{eqnarray*}
-for a constant $C$ not depending on $n$ and $\gamma.$
-If the class of functions $\Psi$ is linear of dimension $\mathrm{D}_\Psi$ As can be easily seen the cost of computing $\widehat \pi_{K,n}^N(f)$ is of order $ K T^2 n$.
-At the same time the variance of the standard estimate $\widehat \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved in general.
-Thus the ratio of cost-variances is of order
-\begin{eqnarray*}
-\frac{\mathrm{cost}(\pi_{K,n}^N) \PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]}{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}\leq C \gamma K T^2 L^2 \mathrm{Dim}(\Psi). 
-\end{eqnarray*}
-If $K,$ $L$ and $\mathrm{Dim}(\Psi)$ are fixed, then our variance reduction method is advantageous if $\gamma=O(1/T^2).$ 
+\PE[A^2_{l,k}(X^x_{N+l-1})]\leq e^{\varkappa n\gamma^2}R^2_k ,\quad k=1,2,\ldots, \quad l=1,\ldots,n,
+\end{eqnarray*}
+for a sequence  $(R_k)$ satisfying $\sum_{k=1}^\infty R^2_k\leq C$ with constant $C$  not depending  on $\gamma$ and $n$ (see \eqref{eq:Aqk2}).
+Hence if we take $W_k=R_k$ as truncation parameters in \eqref{eq:trunc}, then $\widehat A_{n-l,k}$ converges to
+\[
+\bar A_{q,k}=\argmin_{\psi\in\Psi}\mathsf{E}[|A_{q,k}(X_L)-\psi(X_L)|]^2
+\]
+ as $T\to \infty,$ see Section~11 in \cite{gyorfi2006distribution}.  Moreover,
+\begin{eqnarray*}
+\PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]\leq \frac{2Ce^{\varkappa n\gamma^2}}{n}. 
+\end{eqnarray*}
+ Note that the above estimate is rather rough as we ignore here the fact that the difference between $\widehat A_{n-l,k}$ and $A_{n-l,k}$ is small.
+If the class of functions $\Psi$ is a linear one of dimension $\mathtt{D}_\Psi,$ then the cost of computing $\widehat \pi_{K,n}^N(f)$ is of order $ \mathtt{D}_\Psi K T^2 n$.
+At the same time the variance of the standard estimate $ \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved, see Remark~\ref{rem:var-low}.
+Thus, the ratio of cost-variances is of order
+\begin{eqnarray*}
+\frac{\mathrm{cost}(\pi_{K,n}^N) \PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]}{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}\leq C e^{\varkappa n\gamma^2} K \gamma \, T^2 \mathtt{D}_\Psi. 
+\end{eqnarray*}
+Thus, for any fixed $K\geq 1,$  our variance reduction method is advantageous if $\gamma=\min\{1/(\mathtt{D}_\Psi T^2),1/\sqrt{n}\}$  
 \section{Analysis of variance reduced ULA}
@@
 \label{eq:ula_kernel}
-\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(2 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{2\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\} \rmd y
+\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(2 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{2\gamma} \left\| y - x + \gamma \nabla U(x)\right\|^2\right\} \rmd y
 \end{equation}
@@
 Note that for a homogeneous Markov chain $\left(X^x_{p}\right)_{p \in \nset_0}$ of the form \eqref{eq:chain} with constant step size $\gamma$, $\mathsf{P}(X_n^x \in A) = \kerULA^n_{\gamma}(x,A)$.
-
-\paragraph{Squared bias:} Due to the martingale transform structure
+Due to the martingale transform structure
 of $M_{K,n}^N(f)$,
@@
 \[
-\mathsf E\left[M_{K,n}^N(f)\right]=0,
+\mathsf E\left[M_{K,n}^N(f)\right]=0.
 \]
 Hence both estimators
-$\pi_n^N(f)$ and $\pi_{K,n}^N(f)$ have the same conditional bias.
-
-\begin{lem} 
-\label{lem:bias}
-Let $f$ be a bounded Borel function and $(X^x_{p})_{p \in \nset_0}$ be a Markov chain generated by ULA with constant step size $\gamma$. Then the squared bias satisfies
-\begin{equation}
-\label{eq:sq_bias}
-\left( \mathsf{E}[\rho(f)]-\pi(f)\right)^2 \lesssim \|f\|^2_{\infty}\left(\gamma + \frac{\rho^{2N}}{n(1-\rho^2)}\right) \lesssim \|f\|^2_{\infty}\left(\gamma + \frac{e^{-\gamma \kappa N}}{2\gamma n}\right)
-\end{equation}
-for $V(x) = 1 + \|x\|^2$, $\rho$ defined in Lemma~\ref{lem:v_ergodicity}, $\kappa$ independent of $N,n,\gamma$ and $\lesssim$ means less up to a constant not depending on $N,n$ and $\gamma$.
-\end{lem}
-
-\begin{proof} Under assumptions {\bf (H1)} and {\bf (H2)}, the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$.
-Thus we represent the squared bias as
-\begin{eqnarray}
-\label{eq:bias_1}
-\nonumber
-\left( \mathsf{E}[\pi_{K,n}^N(f)]-\pi(f)\right)^2
-&=&\left(\mathsf{E}[\pi_{n}^N(f)]-\pi(f)\right)^2
-=
-\frac{1}{n^2}\left[\sum\limits_{k=N+1}^{N+k}\left(f(X_k^x)-\pi(f)\right)\right]^2 \\
-&\leq& 
-2|\pi_{\gamma}(f) - \pi(f)|^2 + \frac{2}{n}\sum\limits_{k=N+1}^{N+n}\left[\PE f(X_k^x) - \pi_{\gamma}(f)\right]^2
-\end{eqnarray}
+$\pi_n^N(f)$ and $\pi_{K,n}^N(f)$ have the same  bias.
+Under assumptions {\bf (H1)} and {\bf (H2)}, the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$.
 From \cite[Theorem~10]{durmus:moulines:2017} and Lemma~\ref{lem:v_ergodicity} it follows that
@@
 \label{eq:bias_2}
-\tvnorm{\pi - \pi_\gamma} < C_1\sqrt{\gamma}, \quad |\PE f(X_k^x) - \pi_{\gamma}(f)| \leq C_2 \rho^n\|f\|_{V}V(x)
-\end{equation}
-for  $V(x) = 1 + \|x\|^2,$ $\rho\in (0,1)$ specified in Lemma~\ref{lem:v_ergodicity} and constants $C_1,C_2>0$ independent of $\gamma$  and $n$.
-Since $f$ is bounded, $\|f\|_V \leq \|f\|_{\infty}$.
-Hence, combining \eqref{eq:bias_1} and \eqref{eq:bias_2}, we obtain \eqref{eq:sq_bias}.
-\end{proof}
-
-\paragraph{Variance:} In this section we derive an upper bound for the variance of the classical estimator~\eqref{eq:29032018a2} for ULA-based chain.
+\tvnorm{\pi - \pi_\gamma} < C_1\sqrt{\gamma}.
+\end{equation}
+Let us now derive an upper bound for the variance of the classical estimator~\eqref{eq:29032018a2} for ULA-based chain.
 
@@
 \label{lem:variance}
-Let $f$ be a bounded Borel function and $(X^x_{p})_{p \in \nset_0}$ be a Markov chain generated by ULA with constant step size $\gamma$, satisfying {\bf (H1)} and {\bf (H2)} and assume also $n\gamma > C_0$ for some fixed $C_0$.
-Then it holds
+Let $f$ be a bounded Borel function and $(X^x_{p})_{p \in \nset_0}$ be a Markov chain generated by ULA with constant step size $\gamma$, satisfying {\bf (H1)} and {\bf (H2)}.
+Assume also that $n\gamma > C_0$ for some fixed $C_0,$ then it holds
 
@@
 \end{lem}
-\begin{proof}
-Under assumptions {\bf (H1)} and {\bf (H2)} the Markov chain $(X^x_{p})_{p \in \nset_0}$ is $V-$geometrically ergodic with $V(x) = 1+\|x\|^2$ and $\rho = \rme^{-\kappa\gamma}$ with $\kappa$ specified in lemma~\ref{lem:v_ergodicity}.
-Due to lemma~\ref{lem:covariance},
-\[
-\PVar[\pi_n^N(f)]\lesssim \frac{1}{n(1-\rho^{1/2})}
-\]
-and the result of lemma follows from Taylor expansion of denominator.
-\end{proof}
-
-One of the main results of this paper is the following  upper bound for the variance of variance-reduced estimator $\pi_{K,n}^{N}(f).$
+\begin{remark}
+\label{rem:var-low}
+The bound in Lemma~\ref{lem:variance} is sharp and can not be improved even in the case of a normal distribution $\pi.$
+Indeed,  ULA for the standard normal distribution takes form
+\begin{eqnarray*}
+X_{k+1} = (1-\gamma)X_k + \sqrt{2\gamma}\xi_{k+1},\quad \xi_{k+1} \sim \mathcal{N}(0,1)
+\end{eqnarray*}
+Thus, setting $X_0 = 0$, we obtain
+\begin{eqnarray*}
+\PVar_{0} \pi_n^0(f) &=& \PVar_{0}\left[\frac{1}{n}\sum\limits_{i=1}^{n}X_i\right] 
+= \frac{1}{n^2}\PVar\left[\sum\limits_{i=1}^{n}\sum\limits_{j=1}^{i}\sqrt{2\gamma}(1-\gamma)^{i-j}\xi_{j}\right]
+\geq  \frac{2}{n\gamma} - \frac{4(1-\gamma)}{n^2\gamma^2}.
+\end{eqnarray*}
+\end{remark}
+One of the main results of this paper is the following  upper bound for the functions $A_{q,\mathbf{k}}.$
 \begin{thm}\label{th:mr}
 Assume {\bf (H1)} and {\bf (H2)}.
-Suppose additionally that a bounded function $f$ and $\mu=\nabla U$ are $K \times d \geq 2$ times continuously differentiable
+Fix some $K\geq 1$ and suppose additionally that a bounded function $f$ and $\mu=\nabla U$ are $K \times d $ times continuously differentiable
 and for all $x\in\mathbb R^d$ and  \(\mathbf{k}\) satisfying \(0<\|\mathbf{k}\|\leq  K,\)
@@
 \end{equation}
-Suppose that $n\gamma^2\leq \Gamma.$ Set $V(x) = 1 + \|x\|^2,$ then  it holds
+ Set $V(x) = 1 + \|x\|^2,$ then  it holds for $q=1,\ldots,n,$ and all $x$
 \begin{eqnarray}
 \label{eq:Aqk2}
-\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K} A^2_{q,\mathbf{k}}(x)\leq C_{K}\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
+A^2_{q,\mathbf{k}}(x)\leq C_Ke^{\varkappa n\gamma^2}(1+V(x)),\quad 1\leq |\mathbf{k}|\leq K,
+\end{eqnarray}
+ and 
+\begin{eqnarray}
+\label{eq:Aqk2sum}
+\sum_{\|\mathbf{k}\|\geq K+1} A^2_{q,\mathbf{k}}(x)\leq C_{K}e^{\varkappa n\gamma^2}\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
 \left(\frac{\gamma}{2}\right)^{|I|K-1}
@@
 \end{eqnarray}
-for some constant $C_K$ not depending on $q$ and depending on $\gamma$ only through $\Gamma$. 
+with some positive constants $\varkappa, C_K$ not depending on $n$ and  $\gamma.$  The sum in \eqref{eq:Aqk2} runs over all nonempty subsets $I$ of the set $\{1,\ldots,d\}$ 
 \end{thm}
@@
 \label{eq:var-bound}
-\PVar\left(\pi_{K,n}^{N}(f)\right)\lesssim n^{-1} \gamma^{K-1} 
+\PVar\left(\pi_{K,n}^{N}(f)\right)\lesssim n^{-1} \gamma^{K-1}  e^{\varkappa n\gamma^2}
 \eqsp,
 \end{eqnarray}
-where $\lesssim$ stands for inequality up to a constant depending only on $K$ and $\Gamma$.
+where $\lesssim$ stands for inequality up to a constant not depending  on $n$ and $\gamma$.
 \end{cor}
@@
 \label{eq:Ask-part}
-A_{q,\mathbf{k}}(x)=\gamma^{|\mathbf{k}'|/2}\frac{\sqrt{(\mathbf{k}-\mathbf{k}')!}}{\sqrt{\mathbf{k}!}}\mathsf{E}_x\left[ \partial_{Z_1}^{\mathbf{k}'} F(x,Z_1,\ldots,Z_q) \mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_{1})\right],
+A_{q,\mathbf{k}}(x)=\gamma^{|\mathbf{k}'|/2}\frac{\sqrt{(\mathbf{k}-\mathbf{k}')!}}{\sqrt{\mathbf{k}!}}\mathsf{E}\left[ \partial_{Z_1}^{\mathbf{k}'} F(x,Z_1,\ldots,Z_q) \mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_{1})\right],
 \end{eqnarray}
 where $F_q(x,Z_1,\ldots,Z_q)=\sum_{r=0}^q f(X^x_r)$ and $\partial_{Z_1}^{\mathbf{k}'}$
-stands for a weak partial derivative of the functional $F_s$ that also can be viewed as discretised version of Malliavin derivative. From \eqref{eq:Ask-part} we derive
-\begin{eqnarray*}
-\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K} A^2_{q,\mathbf{k}}(x)\leq \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
+stands for a weak partial derivative of the functional $F_s$ that also can be viewed as discretised version of Malliavin derivative. 
+Now by taking $\mathbf{k}'=\mathbf{k},$ we get
+\begin{eqnarray*}
+A^2_{q,\mathbf{k}}(x)\leq \gamma^{|\mathbf{k}|} \PVar \left( \sum_{p=1}^{q}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X_{p}^x\right)
+\right).
+\end{eqnarray*}
+Also from \eqref{eq:Ask-part} we can derive
+\begin{eqnarray*}
+\sum_{\|\mathbf{k}\|\geq K+1} A^2_{q,\mathbf{k}}(x)\leq \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
 \left(\frac{\gamma}{2}\right)^{|I|K}
@@
 \end{eqnarray*}
-where the sum runs over all nonempty subsets $I$ of the set $\{1,\ldots,d\}$ and
+where 
 \[
@@
 \PVar \left(\sum_{p=1}^{q}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X^x_{p}\right)
-\right)\leq C_{K}\gamma^{-1}(1+V(x))
-\end{equation}
-for some constant $C_K$ not depending on $q$ and $\gamma$.
+\right)\leq C_{K}\gamma^{-1}e^{\varkappa n\gamma^2}(1+V(x)),\quad q=1,\ldots,n,
+\end{equation}
+for some positive constants $\varkappa, C_K$ not depending on $n$ and  $\gamma.$ 
 \section{Numerical analysis}
 \label{sec:num}
-We perform numerical experiments based on ULA algorithm \ref{exam:langevin-algorithm} with constant step size $\gamma$.
-We construct a polynomial approximation  for each \(Q_r(x)\) in \eqref{eq:definition-bar-a} of the form:
+In this section we illustrate the performance of the proposed variance reduction method for ULA.
+First we note that from  numerical point of view another representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
+\begin{prop}
+Let $q\geq l, k\in \nset$. Then the coefficients \(a_{q,l,k}(x)\) in \eqref{eq:coeff_mart}  can be alternatively represented as
+\begin{eqnarray*}
+a_{q,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{q,l}\left(\Phi_l(x,\xi)\right)\right]
+\end{eqnarray*}
+with \(Q_{q,l}(x)=\mathsf{E}\left[f(X^x_{l,q})\right],\) \(q\geq l.\)
+In the case $\Phi_l=\Phi$ in \eqref{eq:ula_new} for all $l\geq 1,$ we have
+\begin{equation}
+\label{eq:definition-bar-a}
+\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
+\end{equation}
+with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$
+\end{prop}
+ We construct a polynomial approximation  for each \(Q_r(x)\) in \eqref{eq:definition-bar-a} in the form:
 \begin{equation*}
-\widehat{Q}_{r}(x) = \sum_{\|\mathbf{s}\|\leq m} \widehat\beta_{\mathbf{s}} x^{\mathbf{s}},\quad s=(s_1,\ldots,s_d) \in \nset_0^d
+\widehat{Q}_{r}(x) = \sum_{\|\mathbf{s}\|\leq m} \widehat\beta_{\mathbf{s}} x^{\mathbf{s}},\quad s=(s_1,\ldots,s_d) \in \nset_0^d.
 \end{equation*}
-with some coefficients \(\widehat \beta_{\mathbf{s}}\in \mathbb{R}\) obtained using a modified least-squares criteria based on $T$ independent training trajectories \(\left\{\left(X_1^{(s)},\ldots,X_{N+n}^{(s)}\right)\right\}_{s=1}^{T}\) with $N$ being the size of burn-in period:
+The coefficients \(\widehat \beta_{\mathbf{s}}\in \mathbb{R}\) are obtained using a modified least-squares criteria based on $T$ independent training trajectories \(\bigl\{\bigl(X_1^{(s)},\ldots,X_{N+n}^{(s)}\bigr)\bigr\}_{s=1}^{T}\).
+More precisely  we define $\widehat{Q}_0(x) = f(x)$ and
 \begin{eqnarray}
 \label{eq:Qregr}
-\widehat{Q}_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
+\widehat{Q}_r = \argmin_{\psi \in \Psi_m} \sum_{s = 1}^{T}  \left| f(X_{l_0+r}^{(s)}) - \psi(X_{l_0}^{(s)})\right|^2
 \end{eqnarray}
-for \(1 \leq r \leq n-1,\) $\Psi$ being the class of polynomials $\Psi = \{\psi(x) | \psi(x) = \sum_{\|\mathbf{s}\|\leq m} \alpha_{\mathbf{s}} x^{\mathbf{s}} \}$ and $\widehat{Q}_0(x) = f(x)$ by definition.
+for \(1 \leq l_0+r \leq n-1,\) where $\Psi_m$ is a class of polynomials 
+\[
+ \psi(x) = \sum_{\|\mathbf{s}\|\leq m} \alpha_{\mathbf{s}} x^{\mathbf{s}}, \quad \alpha_{\mathbf{s}}\in \mathbb{R}.
+\] 
 Then using the identity
@@
 \begin{eqnarray*}
-\widehat a_{r,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(x) \widehat{Q}_{r}(x - \gamma \mu(x) + \sqrt{\gamma}\xi)}{\TrainSet \vee \mathcal{G}_N}
-= \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{i,k_i,s_i}(x),
-\end{eqnarray*}
-where for all integers $i,k_i,s_i$ and $x \in \rset^d$,
-\begin{align*}
-P_{r,k,s}(x) &= \mathsf{E} \left[ H_{k}(\xi_l) (x - \gamma \mu(x) + \sqrt{\gamma} \xi)^{s} \right]
-\end{align*}
+\widehat a_{r,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(\xi) \widehat{Q}_{r}(x - \gamma \mu(x) + \sqrt{\gamma}\xi)}{\TrainSet}
+= \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{k_i,s_i}(x_i),
+\end{eqnarray*}
+where for any integer $k,s$ $P_{k,s}$ 
 is a one-dimensional polynomial (in \(x\)) of degree at most \(s\) with analytically known coefficients.
-In consequence, it is enough to estimate functions \(Q_{r}(x)\).
-It can be done using a modified least-squares criteria based on $T$ training trajectories \(\left(X_1^{(s)},\ldots,X_{N+n}^{(s)}\right), s = 1,\ldots,T\) with $N$ being the size of burn-in period:
-\begin{eqnarray}
-\label{eq:Qregr}
-\widehat{Q}_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
-\end{eqnarray}
-for \(1 \leq r \leq n-1,\) $\Psi$ being the class of polynomials $\Psi = \{\psi(x) | \psi(x) = \sum_{\|\mathbf{s}\|\leq m} \alpha_{\mathbf{s}} x^{\mathbf{s}} \}$ and $\widehat{Q}_0(x) = f(x)$ by definition.
-Certainly one may use another functional class $\Psi$ in $\label{eq:Qregr}$, but using polynomials allows to compute closed-form representation of $\widehat a_{r,\mathbf{k}} (x)$.
-\par
-Due to Lemma~\ref{a_decay},
-\sergey{we need to put some lemma about the truncation point here, I will do it}
-it is enough to estimate $Q_{r}$ for $r < n_{\mathrm{trunc}}$ for some truncation level \(n_{\mathrm{trunc}}\) depending on \(d\) and \(\gamma.\) It allows us to use a smaller amount of training trajectories to approximate $Q_{r}(x)$.
-
-Finally  we construct a truncated version of the estimator \ref{eq:29032018a3}:
+We estimate $Q_{r}$ only for $r < n_{\mathrm{trunc}}$ where the truncation level \(n_{\mathrm{trunc}}\) may depend on \(d\) and \(\gamma.\) It allows us to use a smaller amount of training trajectories to approximate $Q_{r}(x)$.
+Finally  we construct a truncated version of the estimator \eqref{eq:29032018a3}:
 \[
@@
 \begin{eqnarray*}
-\widehat M_{K,n,n_{\mathrm{trunc}}}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{0 < \|\mathbf{k}\| < K}\sum_{l=N+1}^{p} \widehat a_{p-l,\mathbf{k}}(X_{l-1})\phi_k(\xi_{l})\mathbbm{1}\{ | p-l | < n_{\mathrm{trunc}}\right]
+\widehat M_{K,n,n_{\mathrm{trunc}}}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{0 < \|\mathbf{k}\| < K}\sum_{l=N+1}^{p} \widehat a_{p-l,\mathbf{k}}(X_{l-1})\phi_k(\xi_{l})\mathbbm{1}\{ | p-l | < n_{\mathrm{trunc}}\right].
 \end{eqnarray*}
@@
 In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i\) or \(f(x) = \sum_{i=1}^d x^2_i\) we used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 10$ independent training trajectories, each one of size $n = 10^4$.
-Then we solve the least squares  problems \eqref{eq:Qcircr} using the first order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section, hence we set $K = 1$.
+Then we solve the least squares  problems \eqref{eq:Qregr} using the first order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section, hence we set $K = 1$.
 We set the truncation level \(n_{\mathrm{trunc}} = 50\).
@@
 We set $p = 100, b = 0.1$ and consider $d=8$.
-We solve the least squares  problems \eqref{eq:Qcircr} using the second-order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section, hence we set $K = 2$.
+We solve the least squares  problems \eqref{eq:Qregr} using the second-order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section, hence we set $K = 2$.
 We set the truncation level \(n_{\mathrm{trunc}} = 50\).
@@
 \begin{eqnarray*}
-U(\theta) = -\mathsf{Y}^T \mathsf{X} \theta + \sum_{i=1}^m \ln{\left(1 + \rme^{\theta^T \mathsf{X}_i}\right)} + \frac{1}{2\sigma^2} \left\| \theta\right\|_2^2, \quad
-\nabla U(\theta) = -\mathsf{Y}^T \mathsf{X} + \sum_{i=1}^m \frac{\mathsf{X}_i}{1 + \rme^{-\mathsf{\theta}^T \mathsf{X}_i}} + \frac{1}{\sigma^2} \theta.
+U(\theta) &=& -\mathsf{Y}^T \mathsf{X} \theta + \sum_{i=1}^m \ln{\left(1 + \rme^{\theta^T \mathsf{X}_i}\right)} + \frac{1}{2\sigma^2} \left\| \theta\right\|_2^2, 
+\\
+\nabla U(\theta) &=& -\mathsf{Y}^T \mathsf{X} + \sum_{i=1}^m \frac{\mathsf{X}_i}{1 + \rme^{-\mathsf{\theta}^T \mathsf{X}_i}} + \frac{1}{\sigma^2} \theta.
 \end{eqnarray*}
@@
 \label{sec:proofs}
+\subsection{Proof of Theorem~\ref{thm:main-repr}}
+The expansion obviously holds for $p = q=1$ and $j=0$.
+Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system in \(L^2(\mathbb{R}^d, P_{\xi})\), it holds in \(L^2(\mathsf{P})\) that
+\[
+f(X^x_{1})=\PE[f(X^x_{1})]+\sum_{k\geq1}a_{1,1,k}(x)\phi_{k}(\xi_{1})
+\]
+for any bounded $f$ with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$.
+Assume now that (\ref{eq:mart_repr}) holds for some $m < p$, any $q \leq m$, $j < q \leq m$ and all bounded $f$. Let us prove that the induction assumption holds for $q=m+1$ and all $j < q$.
+The orthonormality and completeness
+of the system $\left(\phi_{k}\right)_{k=0}^\infty$ implies that for any bounded $f$ and $y \in \rset^d$, 
+$\lim_{n \rightarrow \infty}\Psi_{n,m+1,1}(y)= 0$ with $\Psi_{n,m+1,1}(y)= \PE[|f(X^y_{m,m+1}) - f_{n,m+1,1}(y)|^2]$ and
+\[
+f_{n,m+1,1}(y) = \PE[f(X^y_{m,m+1})] +  \sum\limits_{k=1}^{n}\PE[f(X^y_{m,m+1})\phi_{k}(\xi_{m+1})]\phi_k(\xi_{m+1}).
+\]
+Note that  for any $y \in \rset^d$ and $n \in \nset$ it holds $\Psi_{n,m+1,1}(y) \leq \Psi_{0,m+1,1}(y)$ and
+\begin{eqnarray*}
+\PE \left[ \psi_{0,m+1,1}(X_m^x) \right] &=& \PE \left[ \left.
+\PE \left[ \psi_{0,m+1,1}(X_m^x) \right| \mathcal{G}_m\right]\right] = \PE \left[ \left.
+\PE \left[ |f(X^{X_m^x}_{m,m+1}) - \PE f(X^{X_m^x}_{m,m+1})|^2 \right| X_m^x\right]\right]
+\\
+&=& \PE \left[ |f(X_{m+1}^x) - \PE f(X_{m+1}^x)|^2 \right] = \PVar{f(X_{m+1}^x)} < \infty
+\end{eqnarray*}
+Hence, by Lebesgue dominated convergence theorem, $\lim_{n \rightarrow \infty}\PE \left[\psi_{n,m+1,1}(X_m^x)\right] = 0$ and since for all $y \in \rset^d$ the expectation $\PE[f(X^y_{m,m+1})]$ is a version of $\PE \left[f(X^y_{m,m+1}) | \mathcal{G}_m \right]$, it holds in \(L^2(\mathsf{P})\) that
+\begin{equation}
+\label{eq:1-step-decomp}
+f(X^x_{m+1}) = \PE[f(X^x_{m+1})|\mathcal{G}_m] + \sum\limits_{k=1}^{\infty}a_{m+1,m+1,k}(X^x_{m})\phi_{k}(\xi_{m+1})
+\end{equation}
+where
+\[
+a_{m+1,m+1,k}(y) = \PE \left[f(X^y_{m,m+1})\phi_k\left(\xi_{m+1}\right)\right]
+\]
+which is the required statement in the case $q = m+1$ and $j=m$.
+Now assume that $j<m$.
+Set $g(y) = \PE \left[f(X_{m,m+1}^y)\right]$.
+Note that $\mathsf{P}$-a.s.
+it holds $g(X_m^x) = \PE\left[\left.f(X_m^x) \right| \mathcal{G}_m\right]$ and $g$ is bounded by construction.
+Hence the induction hypothesis applies, and we get
+\begin{equation}\label{eq:28082017a1}
+\mathsf{E}\left[\left.f(X^x_{m+1})\right| \mathcal{G}_m \right]=\mathsf{E}\left[\left.f(X^x_{m+1})\right| \mathcal{G}_j \right]+\sum_{k\geq1}\sum_{l=j+1}^{m}a_{m+1,l,k}(X^x_{l-1})\phi_{k}(\xi_{l})
+\end{equation}
+with
+\begin{align*}
+a_{m+1,l,k}(X^x_{l-1}) &= \mathsf{E}\left[\left.\mathsf{E}\left[\left.f(X^x_{m+1})\right|\mathcal{G}_{m}\right]\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]
+ = \mathsf{E}\left[\left.f(X^x_{m+1})\phi_{k}(\xi_{l})\right|X^x_{l-1}\right].
+\end{align*}
+where for $y \in \rset^d$,
+\[
+a_{m+1,l,k}(y) = \mathsf{E}\left[f(X^y_{l-1,m+1})\phi_{k}(\xi_{l})\right]
+\]
+Formulas \eqref{eq:1-step-decomp}
+and~\eqref{eq:28082017a1} conclude the induction step for $q = m+1$ and all $j < q$ and hence the proof.
+
+\subsection{Proof of Lemma~\ref{lem:variance}}
+Under assumptions {\bf (H1)} and {\bf (H2)} the Markov chain $(X^x_{p})_{p \in \nset_0}$ is $V-$geometrically ergodic with $V(x) = 1+\|x\|^2$ and $\rho = \rme^{-\kappa\gamma}$ with $\kappa$ specified in Lemma~\ref{lem:v_ergodicity}.
+Due to lemma~\ref{lem:covariance},
+\[
+\PVar[\pi_n^N(f)]\lesssim \frac{1}{n(1-\rho^{1/2})}
+\]
+and the result of lemma follows from Taylor expansion of denominator.
+
 \subsection{Proof of Theorem~\ref{th:mr}}
@@
 \begin{eqnarray*}
-\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K}A^2_{s,\mathbf{k}}(x)
+\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}A^2_{s,\mathbf{k}}(x)
 &\leq & \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
@@
 \PVar\left(\sum_{p=1}^{q}\partial_{1}^{\mathbf{K}_I}f\left(X^x_{p}\right)
-\right)\leq C(1+V(x))
-\end{eqnarray*}
-for all $x$ and some constant $C$  not depending on $q.$
+\right)\leq C\gamma^{-1}e^{\varkappa n\gamma^2}(1+V(x)),\quad q=1,\ldots,n,
+\end{eqnarray*}
+for all $x$ and some constants $C,\varkappa>0$  not depending on $n$ and $\gamma.$
 To keep the notational burden at a reasonable level, we present the proof only in one-dimensional case.
@@
 $$
-\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right]\le  \gamma^{-1} C_K (1+V(x)),
-$$
-where $C_K$ is a constant that does not depend on $q$ and $\gamma.$ 
+\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right]\le  \gamma^{-1} C e^{\varkappa n\gamma^2}(1+V(x)), \quad q=1,\ldots,n,
+$$
+where $C,\varkappa>0$ are constants that do not depend on $n$ and $\gamma.$ 
 \end{lem}
@@
 \[
-\mathsf{E}\left[e^{-u\gamma\left(\sum_{k=l}^{j}[\mu'(X^x_{k-1})-\pi_\gamma(\mu')]\right)}\right]\leq C(1+V(x)),\quad u\gamma\leq \gamma_0
-\]
-for some constants $\gamma_0$ and $C>0$  depending only on $\gamma_0$, provided that $n\gamma^2$ stays bounded. Then we have
-\[
-\zeta_{l,j}(u)=\mathsf{E}\left[e^{-s\gamma\sum_{k=l}^{j}\mu'(X^x_{k-1})}\right]\leq  C(1+V(x))e^{-s\gamma (j-l+1)\pi_\gamma(\mu')}.
+\mathsf{E}\left[e^{-u\gamma\left(\sum_{k=l}^{j}[\mu'(X^x_{k-1})-\pi_\gamma(\mu')]\right)}\right]\leq C_1e^{\varkappa n\gamma^2}(1+V(x)),\quad u\gamma\leq \gamma_0,
+\]
+for some constants $\gamma_0$ and $C_1,\varkappa>0$. Then we have
+\[
+\zeta_{l,j}(u)=\mathsf{E}\left[e^{-s\gamma\sum_{k=l}^{j}\mu'(X^x_{k-1})}\right]\leq  C_1 e^{\varkappa n\gamma^2}(1+V(x))e^{-s\gamma (j-l+1)\pi_\gamma(\mu')}.
 \]
@@
 \]
-Note also that \(\pi_\gamma(\mu^{\prime}) \geq \pi_\gamma(\mu^{\prime}) - C_1\sqrt{\gamma}\)
+Note also that \(\pi_\gamma(\mu^{\prime}) \geq \pi_\gamma(\mu^{\prime}) - C_2\sqrt{\gamma}\)
 yielding the bound
 \[
-\zeta_{l,j}(s) \leq  C(1+V(x))e^{-s\gamma(j-l+1)\alpha}
+\zeta_{l,j}(s) \leq  C_1e^{\varkappa n\gamma^2}(1+V(x))e^{-s\gamma(j-l+1)\alpha}
 \]
@@
 \begin{align*}
-\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4) \leq (C(1+V(x)))^{1/4}\sum_{p=j+1}^{q+1}e^{-\gamma(p-j)\alpha} \leq  (C(1+V(x)))^{1/4}\frac{1}{1-e^{- \gamma\alpha}}
+\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4) &\leq (C_1e^{\varkappa n\gamma^2}(1+V(x)))^{1/4}\sum_{p=j+1}^{q+1}e^{-\gamma(p-j)\alpha} 
+\\
+&\leq  (C_1e^{\varkappa n\gamma^2}(1+V(x)))^{1/4}\frac{1}{1-e^{- \gamma\alpha}}
 \end{align*}
@@
 \begin{align*}
-\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \leq (C(1+V(x)))^{1/2}\sum_{j=0}^{q}e^{-\gamma j\alpha} \leq  (C(1+V(x)))^{1/2}\frac{ 1}{1-e^{- \gamma\alpha}}.
+\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \leq (C_1e^{\varkappa n\gamma^2}(1+V(x)))^{1/2}\sum_{j=0}^{q}e^{-\gamma j\alpha} 
+\\
+\leq  (C_1e^{\varkappa n\gamma^2}(1+V(x)))^{1/2}\frac{ 1}{1-e^{- \gamma\alpha}}.
 \end{align*}
@@
 \begin{align*}
-\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X^x_{p}\right)\right] \leq \gamma^{-1} C_K (1+V(x))
+\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X^x_{p}\right)\right] \leq \gamma^{-1} C_K e^{\varkappa n\gamma^2} (1+V(x))
 \end{align*}
-for $\gamma$ small enough with $C_K$ not depending on $q$. The proof is completed.
+ with $C_K$  depending neither on $q$ nor on $\gamma$. The proof is completed.
 \end{proof}
@@
 Then it is known that under assumption {\bf (H1)} corresponding Markov chain would have unique stationary distribution $\pi_\gamma$, which is different from $\pi$.
-Yet this chain will be $V-$geometrically ergodic due to \cite[Theorem~19.4.1]{douc:moulines:priouret:soulier:2018}.
+Yet this chain will be $V-$geometrically ergodic due to \cite[Theorem~19.4.1]{moulines2018}.
 Namely, the following lemma holds:
@@
 \begin{proof} Note that the condition {\bf (H1)} implies that the Markov kernel $Q^{1,n}_{\boldsymbol{\gamma}}$ satisfies $(1,\varepsilon)$-Doeblin condition with $\varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right)$.
-Together with drift condition \ref{eq:drift} it allows to apply \cite[Theorem~19.4.1]{douc:moulines:priouret:soulier:2018} with appropriate constants.
+Together with drift condition \ref{eq:drift} it allows to apply \cite[Theorem~19.4.1]{moulines2018} with appropriate constants.
 \end{proof}
@@
 To bound the first term note that
-\begin{eqnarray*}
-\left|\PE_x\left[f(X_k) - \pi(f)\right]\left[f(X_{k+s})-\pi(f)\right] - \PCov_{\pi}\left[f(X_k),f(X_{k+s})\right]\right| \leq \int\limits_{\rset^d}\left|\PE_y\left[\tilde{f}(X_0)\tilde{f}(X_{s})\right]\right| |\P^k(x,dy)-\pi(dy)| \leq  \\
-\leq C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x)
-\end{eqnarray*}
+\begin{multline*}
+\left|\PE_x\left[f(X_k) - \pi(f)\right]\left[f(X_{k+s})-\pi(f)\right] - \PCov_{\pi}\left[f(X_k),f(X_{k+s})\right]\right| 
+\\
+\leq \int\limits_{\rset^d}\left|\PE_y\left[\tilde{f}(X_0)\tilde{f}(X_{s})\right]\right| |\P^k(x,dy)-\pi(dy)|  
+\leq C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x),
+\end{multline*}
 where the last inequality is due to lemma~\ref{lem:fp_covariance}, which implies \eqref{eq:cov_v_ergodic}.
@@
 \end{proof}
-\section{Proofs}
-\subsection{Proof of Theorem~\ref{thm:main-repr}}
-The expansion obviously holds for $p = q=1$ and $j=0$.
-Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system in \(L^2(\mathbb{R}^d, P_{\xi})\), it holds in \(L^2(\mathsf{P})\) that
-\[
-f(X^x_{1})=\PE[f(X^x_{1})]+\sum_{k\geq1}a_{1,1,k}(x)\phi_{k}(\xi_{1})
-\]
-for any bounded $f$ with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$.
-Assume now that (\ref{eq:mart_repr}) holds for some $m < p$, any $q \leq m$, $j < q \leq m$ and all bounded $f$. Let us prove that the induction assumption holds for $q=m+1$ and all $j < q$.
-The orthonormality and completeness
-of the system $\left(\phi_{k}\right)_{k=0}^\infty$ implies that for any bounded $f$ and $y \in \rset^d$, 
-$\lim_{n \rightarrow \infty}\Psi_{n,m+1,1}(y)= 0$ with $\Psi_{n,m+1,1}(y)= \PE[|f(X^y_{m,m+1}) - f_{n,m+1,1}(y)|^2]$ and
-\[
-f_{n,m+1,1}(y) = \PE[f(X^y_{m,m+1})] +  \sum\limits_{k=1}^{n}\PE[f(X^y_{m,m+1})\phi_{k}(\xi_{m+1})]\phi_k(\xi_{m+1}).
-\]
-Note that  for any $y \in \rset^d$ and $n \in \nset$ it holds $\Psi_{n,m+1,1}(y) \leq \Psi_{0,m+1,1}(y)$ and
-\begin{eqnarray*}
-\PE \left[ \psi_{0,m+1,1}(X_m^x) \right] &=& \PE \left[ \left.
-\PE \left[ \psi_{0,m+1,1}(X_m^x) \right| \mathcal{G}_m\right]\right] = \PE \left[ \left.
-\PE \left[ |f(X^{X_m^x}_{m,m+1}) - \PE f(X^{X_m^x}_{m,m+1})|^2 \right| X_m^x\right]\right]
-\\
-&=& \PE \left[ |f(X_{m+1}^x) - \PE f(X_{m+1}^x)|^2 \right] = \PVar{f(X_{m+1}^x)} < \infty
-\end{eqnarray*}
-Hence, by Lebesgue dominated convergence theorem, $\lim_{n \rightarrow \infty}\PE \left[\psi_{n,m+1,1}(X_m^x)\right] = 0$ and since for all $y \in \rset^d$ the expectation $\PE[f(X^y_{m,m+1})]$ is a version of $\PE \left[f(X^y_{m,m+1}) | \mathcal{G}_m \right]$, it holds in \(L^2(\mathsf{P})\) that
-\begin{equation}
-\label{eq:1-step-decomp}
-f(X^x_{m+1}) = \PE[f(X^x_{m+1})|\mathcal{G}_m] + \sum\limits_{k=1}^{\infty}a_{m+1,m+1,k}(X^x_{m})\phi_{k}(\xi_{m+1})
-\end{equation}
-where
-\[
-a_{m+1,m+1,k}(y) = \PE \left[f(X^y_{m,m+1})\phi_k\left(\xi_{m+1}\right)\right]
-\]
-which is the required statement in the case $q = m+1$ and $j=m$.
-Now assume that $j<m$.
-Set $g(y) = \PE \left[f(X_{m,m+1}^y)\right]$.
-Note that $\mathsf{P}$-a.s.
-it holds $g(X_m^x) = \PE\left[\left.f(X_m^x) \right| \mathcal{G}_m\right]$ and $g$ is bounded by construction.
-Hence the induction hypothesis applies, and we get
-\begin{equation}\label{eq:28082017a1}
-\mathsf{E}\left[\left.f(X^x_{m+1})\right| \mathcal{G}_m \right]=\mathsf{E}\left[\left.f(X^x_{m+1})\right| \mathcal{G}_j \right]+\sum_{k\geq1}\sum_{l=j+1}^{m}a_{m+1,l,k}(X^x_{l-1})\phi_{k}(\xi_{l})
-\end{equation}
-with
-\begin{align*}
-a_{m+1,l,k}(X^x_{l-1}) &= \mathsf{E}\left[\left.\mathsf{E}\left[\left.f(X^x_{m+1})\right|\mathcal{G}_{m}\right]\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]
- = \mathsf{E}\left[\left.f(X^x_{m+1})\phi_{k}(\xi_{l})\right|X^x_{l-1}\right].
-\end{align*}
-where for $y \in \rset^d$,
-\[
-a_{m+1,l,k}(y) = \mathsf{E}\left[f(X^y_{l-1,m+1})\phi_{k}(\xi_{l})\right]
-\]
-Formulas \eqref{eq:1-step-decomp}
-and~\eqref{eq:28082017a1} conclude the induction step for $q = m+1$ and all $j < q$ and hence the proof.
-
-\subsection{Proof of Lemma~\ref{lem:coef_decrease}}
-Note that 
-\[
-\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
-\]
-with  $Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right], r\in \mathbb{N}$. $V-$ergodicity of $(X^x_{p})_{p \in \nset_0}$ implies that
-\[
-|Q_{r}(x) - \pi(f)| \leq \left|\int\limits_{\rset^d} f(y)\left(\P^r(x,dy)-\pi(dy)\right)\right| \leq C\rho^r V(x) \|f\|_{V}
-\]
-Thus, since $\PE\phi_k\left(\xi\right) = 0$, we obtain from Cauchy-Schwarz inequality
-\begin{eqnarray*}
-|\bar a_{r,k}(x)| \leq \left(\PE \phi^2_k\left(\xi\right)\right)^{1/2}\left(\PE |Q_{r}(\Phi(x,\xi)) - \pi(f)|^{2}\right)^{1/2} \leq C\rho^r\|f\|_{V} (\PE V^2(\Phi(x,\xi)))^{1/2}
-\end{eqnarray*}
-and the statement follows.
 
@@
 
-From  numerical point of view another representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
-\begin{prop}
-Let $q\geq l, k\in \nset$. Then the coefficients \(a_{q,l,k}(x)\) in \eqref{eq:coeff_mart}  can be alternatively represented as
-\begin{eqnarray*}
-a_{q,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{q,l}\left(\Phi_l(x,\xi)\right)\right]
-\end{eqnarray*}
-with \(Q_{q,l}(x)=\mathsf{E}\left[f(X^x_{l,q})\right],\) \(q\geq l.\)
-%where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
-The functions \((Q_{q,l})_{l=0}^q\)  can be computed by the backward recurrence: $Q_{q,q}(x)=f(x)$ and for $l \in \{0,\ldots,q-1\}$
-\begin{eqnarray}
-\label{eq:qpl}
-Q_{q,l}(x)=\mathsf{E}\left[Q_{q,l+1}(X^x_{l+1})\right].
-%=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))]
-\end{eqnarray}
-In the case $\Phi_l=\Phi$ in \eqref{eq:ula_new} for all $l\geq 1,$ we have
-\begin{equation}
-\label{eq:definition-bar-a}
-\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
-\end{equation}
-with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$
-\end{prop}
 \par

### 2020-03-21 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 \label{seq:mart_repr}
-In this section we give a general discrete-time martingale representation for  Markov chains of the type  \eqref{eq:chain_gen} which is  used below to construct an efficient variance reduction algorithm.
+In this section we give a general discrete-time martingale representation for  Markov chains of the type  \eqref{eq:chain_gen} which is  used later to construct an efficient variance reduction algorithm.
 Let \((\phi_k)_{k\in \mathbb{Z}_+}\) be a complete orthonormal system in \(L^2(\mathbb{R}^m, P_{\xi})\) with \(\phi_0\equiv 1\).
@@
 \[
-\bar A_{q,k}=\argmin_{\psi\in\Psi}\mathsf{E}[|A_{q,k}(X_L)-\psi(X_L)|]^2
+\bar A_{q,k}=\argmin_{\psi\in\Psi}\mathsf{E}[|A_{q,k}(X_{l_0})-\psi(X_{l_0})|]^2
 \]
@@
 At the same time the variance of the standard estimate $ \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved, see Remark~\ref{rem:var-low}.
-Thus, the ratio of cost-variances is of order
+Thus, the ratio of the corresponding cost-variances is of order
 \begin{eqnarray*}
@@
 \end{eqnarray*}
-Thus, for any fixed $K\geq 1,$  our variance reduction method is advantageous if $\gamma=\min\{1/(\mathtt{D}_\Psi T^2),1/\sqrt{n}\}$  
+Thus, for any fixed $K\geq 1,$  our variance reduction method is advantageous if $\gamma<\min\{1/(\mathtt{D}_\Psi T^2),1/\sqrt{n}\}.$  
 \section{Analysis of variance reduced ULA}
@@
 \label{eq:bias_2}
-\tvnorm{\pi - \pi_\gamma} < C_1\sqrt{\gamma}.
+\tvnorm{\pi - \pi_\gamma} < C\sqrt{\gamma}.
 \end{equation}

### 2020-03-21 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 \end{equation}
-where $N\in\mathbb N_0$ is the length of the burn-in period and $n\in\mathbb N$
+where $N\in\mathbb N_0$ is the length of the burn-in period and $n\in\mathbb N$ is
 the number of effective samples.
@@
 \]
-for any measurable function $g.$  Hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
+for any integrable function $g.$  Hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
 The representation \eqref{eq:coeff_mart} suggests that the variance of  the   estimator
@@
 \label{eq:definition-A-s,k}
-A_{q,k}(y)=\sum_{r=0}^q \bar{a}_{r,k}(y),\quad q\in \mathbb{N}.
+A_{q,k}(y)=\sum_{r=0}^q \bar{a}_{r,k}(y),\quad q=0,\ldots,n-1.
 \end{eqnarray}
 Hence $\mathsf{Var}[\pi_{K,n}^N(f)]$ is small provided that the coefficients $A_{s,k}$ decay fast enough as $k\to \infty.$
-In   Section~\ref{sec:ula_analysis}   we provide a detailed  theoretical  analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).   
+In   Section~\ref{sec:ula_analysis}   we provide a detailed  theoretical  analysis of  this decay  for   ULA   (see Example~\ref{exam:langevin-algorithm}).   
 \par
@@
 We first present a generic regression algorithm and then in Section~\ref{sec:num} give further implementation details. Our algorithm starts with estimating the functions 
-$A_{q,k}$ for $q=N+1,\ldots,N+n,$ $k=1,\ldots,K. $ We first generate \(\NtrainPath\)  paths  of the chain \(X\)
+$A_{q,k}$ for $q=0,\ldots,n-1,$ $k=1,\ldots,K. $ We first generate \(\NtrainPath\)  paths  of the chain \(X\)
 (the so-called ``training paths''):
 \[
-\TrainSet=\Bigl\{(X^{(s)}_{N+1},\ldots,X^{(s)}_{N+n}),\quad s=1,\ldots,\NtrainPath\Bigr\}
+\TrainSet=\Bigl\{(X^{(s)}_{1},\ldots,X^{(s)}_{N+n}),\quad s=1,\ldots,\NtrainPath\Bigr\}
 \]
@@
 for some $l_0$ with $l_0+q\leq N+n,$
-where \(\Psi\) is a  class of  real-valued functions on \(\mathbb{R}^d\) and $\xi^{(s)}_{N+1},\ldots,\xi^{(s)}_{N+n}$ are iid random variables with distribution $P_\xi$ used to construct the $s$-th trajectory of $X.$   As usual in regression analysis (see Section~11 in \cite{gyorfi2006distribution})  we  truncate  the least squares estimates $(\widetilde{A}_{q,k})$ and set
+where \(\Psi\) is a  class of  real-valued functions on \(\mathbb{R}^d\) and $\xi^{(s)}_{1},\ldots,\xi^{(s)}_{N+n}$ are iid random variables with distribution $P_\xi$ used to construct the $s$-th trajectory of $X.$   As usual in regression analysis (see Section~11 in \cite{gyorfi2006distribution}),  we  truncate  the least squares estimates $(\widetilde{A}_{q,k})$ and set
 \begin{eqnarray}
@@
 \begin{cases}
-\widetilde{A}_{q,k}(y), & |\widetilde{A}_{q,k}(y)|\leq U_k,
-\\
-U_k \,\mathrm{sign}(\widetilde{A}_{q,k}(y)), & \mathrm{otherwise}. 
+\widetilde{A}_{q,k}(y), & |\widetilde{A}_{q,k}(y)|\leq W_k,
+\\
+W_k \,\mathrm{sign}(\widetilde{A}_{q,k}(y)), & \mathrm{otherwise}. 
 \end{cases}
 \end{eqnarray}
- for a sequence of positive real numbers $U_k,$ $k=1,\ldots,K.$
+ for a sequence of positive real numbers $W_k,$ $k=1,\ldots,K.$
 Upon estimating  the functions \((A_{q,k})\), one can construct the empirical estimate  of \(M_{K,n}^N(f)\) in the form
@@
 \begin{eqnarray*}
-\PE[A^2_{l,k}(X^x_{N+l-1})]\leq e^{\varkappa n\gamma^2}R^2_k ,\quad k=1,2,\ldots, \quad l=1,\ldots,n,
+\PE[A^2_{l,k}(X^x_{N+l-1})]\leq e^{\varkappa n\gamma^2}R^2_k ,\quad k=1,2,\ldots, \quad l=0,\ldots,n-1,
 \end{eqnarray*}
@@
 \end{eqnarray*}
- Note that the above estimate is rather rough as we ignore here the fact that the difference between $\widehat A_{n-l,k}$ and $A_{n-l,k}$ is small.
+ Note that the above estimate is rather rough, as we ignore here the fact that the difference between $\widehat A_{n-l,k}$ and $A_{n-l,k}$ is small.
 If the class of functions $\Psi$ is a linear one of dimension $\mathtt{D}_\Psi,$ then the cost of computing $\widehat \pi_{K,n}^N(f)$ is of order $ \mathtt{D}_\Psi K T^2 n$.
-At the same time the variance of the standard estimate $ \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved, see Remark~\ref{rem:var-low}.
+At the same time the variance of the standard estimate $ \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved, see Lemma~\ref{lem:variance} and Remark~\ref{rem:var-low}.
 Thus, the ratio of the corresponding cost-variances is of order
@@
 Let $f$ be a bounded Borel function and $(X^x_{p})_{p \in \nset_0}$ be a Markov chain generated by ULA with constant step size $\gamma$, satisfying {\bf (H1)} and {\bf (H2)}.
-Assume also that $n\gamma > C_0$ for some fixed $C_0,$ then it holds
+Assume also that $n\gamma > c_0$ for some fixed $c_0>0,$ then it holds
 
@@
 \end{equation}
- Set $V(x) = 1 + \|x\|^2,$ then  it holds for $q=1,\ldots,n,$ and all $x$
+ Set $V(x) = 1 + \|x\|^2,$ then  it holds for $q=0,\ldots,n-1,$ and all $x$
 \begin{eqnarray}
@@
 \PVar \left(\sum_{p=1}^{q}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X^x_{p}\right)
-\right)\leq C_{K}\gamma^{-1}e^{\varkappa n\gamma^2}(1+V(x)),\quad q=1,\ldots,n,
+\right)\leq C_{K}\gamma^{-1}e^{\varkappa n\gamma^2}(1+V(x)),\quad q=0,\ldots,n-1,
 \end{equation}

### 2020-03-21 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 In this section we perform the convergence analysis of the ULA algorithm.
-We use the notations of example \ref{exam:langevin-algorithm}.
+We use the notations of Example~\ref{exam:langevin-algorithm}.
 For the sake of clarity and notational simplicity we restrict our attention to the constant time step, that is, we take $\gamma_k =\gamma$ for any $k \in \nset$.
-
 By $H_k$, $k\in\mathbb N_0$,
@@
 \begin{align}
-M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\frac{1}{n}\sum_{l=1}^{n}A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})
+M_{K,n}^N(f) & =\sum_{ 0<\|\mathbf{k}\|\le K}\frac{1}{n}\sum_{l=1}^{n}A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})
 \label{eq:29032018a5}
@@
 \label{eq:bias_2}
-\tvnorm{\pi - \pi_\gamma} < C\sqrt{\gamma}.
-\end{equation}
-Let us now derive an upper bound for the variance of the classical estimator~\eqref{eq:29032018a2} for ULA-based chain.
+\tvnorm{\pi - \pi_\gamma} \leq  C\sqrt{\gamma}
+\end{equation}
+for some constant $C>0.$ Let us now derive an upper bound for the variance of the classical estimator~\eqref{eq:29032018a2} for ULA-based chain.
 
@@
 \end{equation}
-with $V(x) = 1+ \|x\|^2$ where $\lesssim$ stands for inequality up to a constant depending on constants $L,$ $m,$  $M$ and $\mathrm{osc}(f)$.
+with $V(x) = 1+ \|x\|^2$ where $\lesssim$ stands for inequality up to a constant not depending on $\gamma$ and $n$.
 \end{lem}
@@
 \end{remark}
-One of the main results of this paper is the following  upper bound for the functions $A_{q,\mathbf{k}}.$
+One of the main results of this paper is the following  upper bound for the functions $(A_{q,\mathbf{k}}).$
 \begin{thm}\label{th:mr}
@@
 \end{eqnarray}
-where $\lesssim$ stands for inequality up to a constant not depending  on $n$ and $\gamma$.
+where $\lesssim$ stands for inequality up to a constant not depending  on $n,$ $N$ and $\gamma$.
 \end{cor}
@@
 A^2_{q,\mathbf{k}}(x)\leq \gamma^{|\mathbf{k}|} \PVar \left( \sum_{p=1}^{q}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X_{p}^x\right)
-\right).
+\right),\quad q=0,\ldots,n-1.
 \end{eqnarray*}

### 2020-03-21 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 \subsection{Banana shape distribution}
-The “Banana-shape” distribution, proposed by \cite{Haario1999}, can be obtained from a d-dimensional Gaussian vector with zero mean and covariance $\mathrm{diag}(p,1,\ldots,1)$ by applying transformation $\varphi_b(x) : \rset^d \rightarrow \rset^d$ of the form
+The “Banana-shape” distribution, proposed by \cite{haario1999adaptive}, can be obtained from a d-dimensional Gaussian vector with zero mean and covariance $\mathrm{diag}(p,1,\ldots,1)$ by applying transformation $\varphi_b(x) : \rset^d \rightarrow \rset^d$ of the form
 \[

### 2020-03-25 aap/Revision/vr-mcmc-revision-new-new.tex
update

--- 2020-03-21 aap/Revision/vr-mcmc-revision-new-new.tex
@@
 The situation becomes much more difficult in the case of MCMC algorithms, where one  has to work with a
-Markov chain \(X_p,\) \(p=0,1,2,\ldots,\) whose marginal distribution  converges  to \(\pi\) as time grows.
+Markov chain \((X_p),\) \(p=0,1,2,\ldots,\) whose marginal distribution  converges  to \(\pi\) as time grows.
 One important class of the variance reduction methods in this case  is based on the Poisson equation for the corresponding  Markov chain.
@@
 For $x \in \rset^d$ and $r>0$ let $B_r(x) = \{y \in \rset^d | \|y - x\| < r\}$ where $\|\cdot\|$ is a standard Euclidean norm.
-For the twice differentiable function $g: \rset^d \rightarrow \rset$ we denote by $D^2g$ its Hessian at point $x$.
+For the twice differentiable function $g: \rset^d \rightarrow \rset$ we denote by $D^2g(x)$ its Hessian at point $x$.
 For $m\in\mathbb N$, a smooth function
@@
 \[
-\partial^{\mathbf k}_{j} h(y_1,\ldots,y_m)
+\partial^{\mathbf k}_{y_j} h(y_1,\ldots,y_m)
 :=\partial^{k_d}_{y_j^d}
@@
 For a bounded Borel function $f: \rset^d \rightarrow \rset$ denote $\mathrm{osc}(f):=\sup_{x\in\mathbb R^d}f(x)-\inf_{x\in\mathbb R^d}f(x)$.
+Given a function $V: \rset^d \rightarrow \rset$, for a function $f: \rset^d \rightarrow \rset$ we define $\|f\|_{V} = \sup_{x \in \rset^d}\frac{|f(x)|}{V(x)}$ and the corresponding $V-$norm between probability measures $\mu$ and $\nu$ on $\mathcal{B}(\rset^d)$ as
+\[
+d_V(\mu, \nu) = \|\mu - \nu\|_{V} = \sup_{\|f\|_{V} \leq 1} \left[\int\limits_{\rset^d}f(x)\,d\mu(x) - \int\limits_{\rset^d}f(x)\,d\nu(x)\right]  \eqsp.
+\]
 
@@
 However, in many practical situations  direct sampling from \(\pi\) is impossible and this precludes the use of plain Monte Carlo methods in this case.
-One popular alternative to Monte Carlo  is Markov Chain Monte Carlo (MCMC), where one is looking for a discrete time  (possibly non-homogeneous) Markov chain   \((X_p)_{p\geq 0}\) such that \(\pi\) is its unique invariant measure.
-In this paper we study a class of MCMC algorithms with \((X_p)_{p\geq 0}\) satisfying the  the following recurrence relation:
+One popular alternative to Monte Carlo  is Markov Chain Monte Carlo (MCMC), where one is looking for a discrete time  (possibly non-homogeneous) Markov chain   \((X_p)_{p \in \nset_0}\) such that \(\pi\) is its unique invariant measure.
+In this paper we study a class of MCMC algorithms with \((X_p)_{p \in \nset_0}\) satisfying the  the following recurrence relation:
 \begin{equation}
@@
 In fact, this is quite general class of Markov chains (see Theorem~1.3.6 in ~\cite{moulines2018})
-and many well-known MCMC algorithms can be represented in form \eqref{eq:chain_gen}.
+and many well-known MCMC algorithms can be represented in the form \eqref{eq:chain_gen}.
 Let us consider two popular examples.
@@
 Under assumptions {\bf (H1)} and {\bf (H2)}, the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$.
-From \cite[Theorem~10]{durmus:moulines:2017} and Lemma~\ref{lem:v_ergodicity} it follows that
+From \cite[Theorem~10]{durmus:moulines:2017}
+%and Lemma~\ref{lem:v_ergodicity} 
+it follows that
 \begin{equation}
@@
 \begin{eqnarray*}
-X_{k+1} = (1-\gamma)X_k + \sqrt{2\gamma}\xi_{k+1},\quad \xi_{k+1} \sim \mathcal{N}(0,1)
+X_{k+1} = (1-2\gamma)X_k + \sqrt{\gamma}\xi_{k+1},\quad \xi_{k+1} \sim \mathcal{N}(0,1)
 \end{eqnarray*}
@@
 \PVar_{0} \pi_n^0(f) &=& \PVar_{0}\left[\frac{1}{n}\sum\limits_{i=1}^{n}X_i\right] 
-= \frac{1}{n^2}\PVar\left[\sum\limits_{i=1}^{n}\sum\limits_{j=1}^{i}\sqrt{2\gamma}(1-\gamma)^{i-j}\xi_{j}\right]
-\geq  \frac{2}{n\gamma} - \frac{4(1-\gamma)}{n^2\gamma^2}.
+= \frac{1}{n^2}\PVar\left[\sum\limits_{i=1}^{n}\sum\limits_{j=1}^{i}\sqrt{\gamma}(1-2\gamma)^{i-j}\xi_{j}\right]
+\geq  \frac{1}{n\gamma} - \frac{2(1-2\gamma)}{n^2\gamma^2}.
 \end{eqnarray*}
@@
 respectively.
-In our experiments we considered $d = 2$, $\mu = \left(0.5,0.5\right)$ and $\Sigma = \Sigma_0$ with randomly generated positive-definite matrix $\Sigma_0$ (heterogeneous structure).
+In our experiments we considered $d = 2$, $\mu = \left(0.5,0.5\right)$ and randomly generated positive-definite matrix $\Sigma$ (heterogeneous structure).
 In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i\) or \(f(x) = \sum_{i=1}^d x^2_i\) we used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 10$ independent training trajectories, each one of size $n = 10^4$.
-Then we solve the least squares  problems \eqref{eq:Qregr} using the first order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section, hence we set $K = 1$.
+Then we solve the least squares  problems \eqref{eq:Qregr} using the first order polynomial approximations for the coefficients \(\hat{a}_{p,\mathbf{k}}\) as described in the previous section, hence we set $K = 1$.
 We set the truncation level \(n_{\mathrm{trunc}} = 50\).
@@
 \]
-The quantity of interest is $\PE_{\pi}[X_2]$.
+The quantity of interest is the expectation of $f(x) = x_2$.
 We set $p = 100, b = 0.1$ and consider $d=8$.
-We solve the least squares  problems \eqref{eq:Qregr} using the second-order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section, hence we set $K = 2$.
+We solve the least squares  problems \eqref{eq:Qregr} using the second-order polynomial approximations for the coefficients \(\hat{a}_{p,\mathbf{k}}\) as described in the previous section, hence we set $K = 2$.
+We use $T=10$ independent training trajectories, each of size $n=10^4$.
 We set the truncation level \(n_{\mathrm{trunc}} = 50\).
@@
 \section{Bounds for moments of  ULA}\label{sec:appendix}
-First let us notice that conditions {\bf (H1)} and {\bf (H2)} imply (\cite[Lemma~16]{brosse2018diffusion}) that for some $K_2 > 0$ it holds
-\begin{equation}
-\label{eq:H3}
-\ps{\nabla U(x)}{x} \geq \frac{m}{2}\|x\|^2, \quad \|x\| \geq K_2
-\end{equation}
-
-Now we shall show that conditions {\bf (H1)} and {\bf (H2)} are sufficient to show that ULA kernel $Q_\gamma$ satisfies the so-called drift condition:
+First let us introduce some definitions 
 \begin{definition}
@@
 
+\begin{definition}
+\label{def:v_ergodicity}
+We say that the Markov chain $(X_p)_{p \in \nset_0}$  taking values in some space $X$ with the Markov kernel $Q$ is $V-$geometrically ergodic for a given function $V: X \rightarrow [1; +\infty)$ if there exist real numbers $C > 0$ and $0 < \rho < 1$ such that for any $n \in \nset$,
+\begin{equation}
+\label{eq:v_ergodicity}
+d_V(\delta_xQ^{1,n},\pi_\gamma) \leq C\rho^n V(x)
+\end{equation}
+\end{definition}
+
+Note that conditions {\bf (H1)} and {\bf (H2)} imply (\cite[Lemma~16]{brosse2018diffusion}) that for some $K_2 > 0$ it holds
+\begin{equation}
+\label{eq:H3}
+\ps{\nabla U(x)}{x} \geq \frac{m}{2}\|x\|^2, \quad \|x\| \geq K_2
+\end{equation}
+
+Now we shall show that conditions {\bf (H1)} and {\bf (H2)} are sufficient to show that ULA kernel $Q_\gamma$ satisfies the drift condition \eqref{eq:drift}
+
 \begin{lem}
@@
  Assume that the potential $U(x), x \in \mathbb{R}^d$ satisfies conditions {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$.
-Then the kernel $Q_\gamma$ from \eqref{eq:ula_kernel} satisfies drift condition \eqref{eq:lyapunov} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L}$ with drift function $V(x) = 1 + \|x\|^2$,   $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = \frac{25K_2^2}{8} + d + m$ with $K_2$ from \eqref{eq:H3}, $m_U$ from {\bf (H2)}.
+Then the kernel $Q_\gamma$ from \eqref{eq:ula_kernel} satisfies drift condition \eqref{eq:lyapunov} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L}$ with drift function $V(x) = 1 + \|x\|^2$,   $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = \frac{25K_2^2}{8} + d + m$ with $K_2$ from \eqref{eq:H3}, $m$ from {\bf (H2)}.
 \end{lem}
@@
 
-For now let us assume for simplicity that we run ULA with fixed step size $\gamma$.
-Then it is known that under assumption {\bf (H1)} corresponding Markov chain would have unique stationary distribution $\pi_\gamma$, which is different from $\pi$.
+It is known that under assumption {\bf (H1)} the Markov chain generated by ULA with constant step size $\gamma$ would have unique stationary distribution $\pi_\gamma$, which is different from $\pi$.
 Yet this chain will be $V-$geometrically ergodic due to \cite[Theorem~19.4.1]{moulines2018}.
@@
 \begin{proof} Note that the condition {\bf (H1)} implies that the Markov kernel $Q^{1,n}_{\boldsymbol{\gamma}}$ satisfies $(1,\varepsilon)$-Doeblin condition with $\varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right)$.
-Together with drift condition \ref{eq:drift} it allows to apply \cite[Theorem~19.4.1]{moulines2018} with appropriate constants.
+Together with drift condition \eqref{eq:lyapunov} it allows to apply \cite[Theorem~19.4.1]{moulines2018} with appropriate constants.
 \end{proof}

### 2020-03-25 aap/Revision/vr-mcmc-revision-new-new.tex

@@
 \begin{eqnarray*}
-\frac{\mathrm{cost}(\pi_{K,n}^N) \PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]}{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}\leq C e^{\varkappa n\gamma^2} K \gamma \, T^2 \mathtt{D}_\Psi. 
-\end{eqnarray*}
-Thus, for any fixed $K\geq 1,$  our variance reduction method is advantageous if $\gamma<\min\{1/(\mathtt{D}_\Psi T^2),1/\sqrt{n}\}.$  
+\frac{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]}{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}\leq C e^{\varkappa n\gamma^2} K \gamma \, T^2 \mathtt{D}_\Psi. 
+\end{eqnarray*}
+Thus, for any fixed $K\geq 1,$  our variance reduction method is advantageous if $\gamma\ll\min\{1/(\mathtt{D}_\Psi T^2),1/\sqrt{n}\}.$  
 \section{Analysis of variance reduced ULA}

### 2020-03-25 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 \label{def:v_ergodicity}
-We say that the Markov chain $(X_p)_{p \in \nset_0}$  taking values in some space $X$ with the Markov kernel $Q$ is $V-$geometrically ergodic for a given function $V: X \rightarrow [1; +\infty)$ if there exist real numbers $C > 0$ and $0 < \rho < 1$ such that for any $n \in \nset$,
+Let $(X_p)_{p \in \nset_0}$ be a Markov chain taking values in some space $X$ with the Markov kernel $Q$ and stationary distribution $\pi$.
+We say that $(X_p)_{p \in \nset_0}$ is $V-$geometrically ergodic for a given function $V: X \rightarrow [1; +\infty)$ if there exist real numbers $C > 0$ and $0 < \rho < 1$ such that for any $n \in \nset$,
 \begin{equation}
 \label{eq:v_ergodicity}
-d_V(\delta_xQ^{1,n},\pi_\gamma) \leq C\rho^n V(x)
+d_V(\delta_xQ^{1,n},\pi) \leq C\rho^n V(x)
 \end{equation}
@@
 $$
-b = \gamma(\frac{25K_1^2}{8} + 2d + m); \quad \overline{b} = b\exp^{-\frac{m\gamma}{2}} + d; \quad \varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right); \quad
+b = \gamma(\frac{25K_1^2}{8} + 2d + m); \quad \overline{b} = b\exp{\left(-\frac{m\gamma}{2}\right)} + d; \quad \varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right); \quad
 $$

### 2020-03-25 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 
-Now we shall show that conditions {\bf (H1)} and {\bf (H2)} are sufficient to show that ULA kernel $Q_\gamma$ satisfies the drift condition \eqref{eq:drift}
+Now we shall show that conditions {\bf (H1)} and {\bf (H2)} are sufficient to show that ULA kernel $Q_\gamma$ satisfies the drift condition \eqref{eq:lyapunov}
 

### 2020-03-28 aap/Revision/Response_1.tex
update

\documentclass{article}%
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[authoryear]{natbib}
\usepackage{color}%
\setcounter{MaxMatrixCols}{30}
%TCIDATA{OutputFilter=latex2.dll}
%TCIDATA{Version=5.50.0.2890}
%TCIDATA{CSTFile=40 LaTeX article.cst}
%TCIDATA{Created=Thursday, September 14, 2017 13:22:20}
%TCIDATA{LastRevised=Saturday, December 07, 2019 12:32:10}
%TCIDATA{<META NAME="GraphicsSave" CONTENT="32">}
%TCIDATA{<META NAME="SaveForMode" CONTENT="1">}
%TCIDATA{BibliographyScheme=Manual}
%TCIDATA{<META NAME="DocumentShell" CONTENT="Standard LaTeX\Blank - Standard LaTeX Article">}
%BeginMSIPreambleData
\providecommand{\U}[1]{\protect\rule{.1in}{.1in}}
%EndMSIPreambleData
\newtheorem{theorem}{Theorem}
\newtheorem{acknowledgement}[theorem]{Acknowledgement}
\newtheorem{algorithm}[theorem]{Algorithm}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{case}[theorem]{Case}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conclusion}[theorem]{Conclusion}
\newtheorem{condition}[theorem]{Condition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{criterion}[theorem]{Criterion}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{solution}[theorem]{Solution}
\newtheorem{summary}[theorem]{Summary}
\newenvironment{proof}[1][Proof]{\noindent\textbf{#1.} }{\ \rule{0.5em}{0.5em}}
\begin{document}

\title{\textbf{Response to the Report by Referee 1}}
\author{{\normalsize \vspace{-1cm} Denis Belomestny, Eric Moulines, and Sergei Samsonov}}
\date{~ }
\maketitle

\vspace{-0.5cm}

We thank the referee for carefully reading our manuscript and for pertinent
comments which helped us to improve the paper. In the new version of our
paper, we have taken all comments and suggestion into account.  Below are some additional responses to the
comments by Referee 1

\section*{Main remarks}

\begin{itemize}

\item \textit{... The authors make an attempt to quantify the variance reduction obtained by the proposed approach, as well as the computational complexity to achieve this variance reduction. On these topics several results are presented as Theorems and Propositions  ... I would recommend the authors to phrase this results with clear conditions in such a way that it becomes directly accessible to the reader ... where the bound is phrased in an easily readible way involving constants that are defined in the statement of the theorem and which should not depend on infinite series and conditional expectations.}
\par
Let us stress that  current literature on MCMC  mainly focuses on  convergence of MCMC algorithms in Total Variation and Wasserstein distances thus  ignoring a sample error of these algorithms. Only few references studied MSE error of the Langevin algorithm (see e.g.  Durmus and  Moulines (2017)). In this work we propose a novel variance reduction approach which is able to reduce the statistical error of MCMC algorithms. The main advantage of our method is provable increase of efficiency for Langevin-based MCMC algorithms as measured by  cost-times-variance criterion. Our approach is based on a new discrete time martingale representation for Markov chains which is of its own interest. Therefore we first present this basic result and then discuss in Section 4 variance reduction method based on the derived martingale representation. In  Section~4 we  explain in accessible way why our approach leads to reduced cost-times-variance criterion. These results are based on deep analysis of the variance reduced estimate postponed to Section 5. In the revised version we significantly weaken  assumptions and prove our main result (Theorem~3 and Corollary~2) for ULA under smoothness  and convexity of the underlying potential outside a ball (assumption  (H2)).   Note that all assumptions needed are now collected in Theorem~3. For convenience of the reader we also sketch the main steps of the proof right after 
Theorem~3. We decided to present our main theoretical results as bounds for the functions $A_{q,k}$ since these functions play a crucial role in the variance bound (13). Given these estimates it is easy to see how to get bounds for the variance. 
\item \textit{... It is good to see that the technique has beneficial effects on the variance, but it does not become clear what was the computational effort required to achieve this benefit. The reader is left wondering whether or not the same variance reduction could have been achieved simply by drawing more MCMC samples ...}
\item \textit{... here the setting \(\gamma_k = c k^{-\alpha}, \alpha\in  (0, 1)\) is considered, and presented as  ``the most interesting case''. For many practitioners, $\alpha=0$ would be rather natural. What can be said in this important case? This is also applies to Section 6 ...}
\par
In the revised version we decided to switch to a constant time step and present all our bounds for this case. On the one hand, this allows us to formulate our main results in a simpler form depending only on a single parameter $\gamma.$ On the other hand, this is indeed the most relevant choice for practitioners.  
\item \textit{ ... here the functions g are assumed uniformly bounded, whereas earlier (end of Section 5) elements in $\Psi$ were proposed to be polynomials. In particular the bound B would be equal to infinity. How can this be reconciled? ...}
\par
In the revised version we skip this result because of size restrictions. 
\end{itemize}

\end{document}

### 2020-03-28 aap/Revision/Response_2.tex
update

--- 2020-03-28 aap/Revision/Response_1.tex
@@
 
-\title{\textbf{Response to the Report by Referee 1}}
+\title{\textbf{Response to the Report by Referee 2}}
 \author{{\normalsize \vspace{-1cm} Denis Belomestny, Eric Moulines, and Sergei Samsonov}}
@@
 paper, we have taken all comments and suggestion into account.  Below are some additional responses to the
-comments by Referee 1
+comments by Referee 2
 
@@
 \item \textit{...
-The authors make an attempt to quantify the variance reduction obtained by the proposed approach, as well as the computational complexity to achieve this variance reduction.
-On these topics several results are presented as Theorems and Propositions  ...
-I would recommend the authors to phrase this results with clear conditions in such a way that it becomes directly accessible to the reader ...
-where the bound is phrased in an easily readible way involving constants that are defined in the statement of the theorem and which should not depend on infinite series and conditional expectations.}
+Restrictive assumptions for the theoretical analysis.
+The main theorems of the paper, which seem to be Theorem 4 and the results of section 6.1, have the restrictive assumption that the drift vector field  needs to satisfy the strong convexity assumption ...
+In typically interesting cases, such as Bayesian inverse problems and machine learning, this assumption is not satisfied.
+Unfortunately, even in the numerical examples provided in Section 7, this assumption is not fulfilled ?
+it already doesn't hold for a two-component Gaussian mixture in one dimension.
+This is a serious restriction ...
+}
 \par
-Let us stress that  current literature on MCMC  mainly focuses on  convergence of MCMC algorithms in Total Variation and Wasserstein distances thus  ignoring a sample error of these algorithms.
-Only few references studied MSE error of the Langevin algorithm (see e.g.
-Durmus and  Moulines (2017)).
-In this work we propose a novel variance reduction approach which is able to reduce the statistical error of MCMC algorithms.
-The main advantage of our method is provable increase of efficiency for Langevin-based MCMC algorithms as measured by  cost-times-variance criterion.
-Our approach is based on a new discrete time martingale representation for Markov chains which is of its own interest.
-Therefore we first present this basic result and then discuss in Section 4 variance reduction method based on the derived martingale representation.
-In  Section~4 we  explain in accessible way why our approach leads to reduced cost-times-variance criterion.
-These results are based on deep analysis of the variance reduced estimate postponed to Section 5.
-In the revised version we significantly weaken  assumptions and prove our main result (Theorem~3 and Corollary~2) for ULA under smoothness  and convexity of the underlying potential outside a ball (assumption  (H2)).
+In the revised version we significantly weaken  the assumptions and prove our main result (Theorem~3 and Corollary~2) for ULA under smoothness  and convexity of the underlying potential outside a ball (assumption  (H2)).
+Thus we avoid the strong convexity assumption.
 Note that all assumptions needed are now collected in Theorem~3.
-For convenience of the reader we also sketch the main steps of the proof right after
-Theorem~3.
-We decided to present our main theoretical results as bounds for the functions $A_{q,k}$ since these functions play a crucial role in the variance bound (13).
-Given these estimates it is easy to see how to get bounds for the variance.
+These  assumptions can now be validated for a large class of examples including example of Gaussian mixtures considered in Section~6.1.
 \item \textit{...
-It is good to see that the technique has beneficial effects on the variance, but it does not become clear what was the computational effort required to achieve this benefit.
-The reader is left wondering whether or not the same variance reduction could have been achieved simply by drawing more MCMC samples ...}
-\item \textit{...
-here the setting \(\gamma_k = c k^{-\alpha}, \alpha\in  (0, 1)\) is considered, and presented as  ``the most interesting case''.
-For many practitioners, $\alpha=0$ would be rather natural.
-What can be said in this important case?
-This is also applies to Section 6 ...}
+in complicated models such as PDE-related Bayesian inverse problems or machine learning, the computation of \(\nabla \log(\pi)\) is expensive ?
+but these examples are typically ones where \(\log(\pi)\) does not satisfy the needed concavity assumptions ...}
 \par
-In the revised version we decided to switch to a constant time step and present all our bounds for this case.
-On the one hand, this allows us to formulate our main results in a simpler form depending only on a single parameter $\gamma.$ On the other hand, this is indeed the most relevant choice for practitioners.
-\item \textit{ ...
-here the functions g are assumed uniformly bounded, whereas earlier (end of Section 5) elements in $\Psi$ were proposed to be polynomials.
-In particular the bound B would be equal to infinity.
-How can this be reconciled?
-...}
-\par
-In the revised version we skip this result because of size restrictions. 
+Now we replaced the strong convexity assumption by a much weaker one (H2) (convexity outside a ball).
+Hence our method may gain additional efficiency as compared to other variance reduction approaches due to the fact that \(\nabla \log(\pi)\) doesn't need to be computed to construct control variates.
 \end{itemize}

### 2020-04-04 Code/martingale.py
code update

@@ -6,6 +6,7 @@ from samplers import ULA_light
 from potentials import GaussPotential,GaussMixture,GausMixtureIdent,GausMixtureSame
 import copy
 from baselines import set_function
+import time
 
 def H(k, x):
     if k==0:
@@ -77,6 +78,8 @@ def approx_q(X_train,Y_train,N_traj_train,lag,max_deg):
                 x_all = np.concatenate((x_all,x),axis = 0)
             y_all = np.concatenate([y_all,y])
         #should use polyfeatures here
+        print("variance: ",np.var(y_all))
+        print(y_all[:50])
         poly = PolynomialFeatures(max_deg)
         X_features = poly.fit_transform(x_all)
         print(X_features.shape)
@@ -88,6 +91,33 @@ def approx_q(X_train,Y_train,N_traj_train,lag,max_deg):
         else:
             coefs_poly = np.concatenate((coefs_poly,coefs),axis=0)
     return coefs_poly
+
+def approx_q_independent(X_train,Y_train,N_traj_train,lag,max_deg):
+    """
+    Function to regress q functions bases on a polynomial basis and big number of short independent trajectories
+    """
+    dim = X_train[0,:].shape[0]
+    print("dimension = ",dim)
+    coefs_poly = np.array([])
+    for i in range(lag):
+        x_all = X_train[:,0,:]
+        y_all = Y_train[:,i,0]
+        print(y_all[:50])
+        print("variance: ",np.var(y_all))
+        #should use polyfeatures here
+        poly = PolynomialFeatures(max_deg)
+        X_features = poly.fit_transform(x_all)
+        print(X_features.shape)
+        lstsq_results = np.linalg.lstsq(X_features,y_all,rcond = None)
+        coefs = copy.deepcopy(lstsq_results[0])
+        coefs.resize((1,X_features.shape[1]))           
+        if coefs_poly.size == 0:
+            coefs_poly = copy.deepcopy(coefs)
+        else:
+            coefs_poly = np.concatenate((coefs_poly,coefs),axis=0)
+    return coefs_poly
+    
+        
 
 def get_indices_poly(ind,K_max,S_max):
     """
@@ -158,7 +188,7 @@ def get_representations(k,s,d,K_max):
[…]

### 2020-04-04 Code/optimize.py

@@ -8,6 +8,7 @@ from VR_methods import Train_1st_order,Train_2nd_order,Train_1st_order_grad,Trai
 from samplers import ULA,MALA,RWM,ULA_SAGA,MC_sampler
 from multiprocessing import Pool
 import multiprocessing
+import time
 
 """
 def optim_1var(i,crit,inds_arr,traj,traj_grad,W_train_spec,n_restarts,tol,sigma):
@@ -466,6 +467,7 @@ def Run_eval_test(intseed,degree,sampler,methods,vars_arr,Potential,test_dict,CV
     N_burn = test_dict["burn_in"]
     N_test = test_dict["n_test"]
     d = test_dict["dim"]
+    start_time = time.time()
     if sampler_type == "ULA":
         traj,traj_grad = ULA(intseed,Potential,step, N_burn, N_test, d, burn_type, main_type)
     elif sampler_type == "MALA":
@@ -478,13 +480,16 @@ def Run_eval_test(intseed,degree,sampler,methods,vars_arr,Potential,test_dict,CV
     #lists to save the results of the trajectory
     res_dict = {"Vanilla":[],"ESVM":[],"EVM":[],"LS":[],"MAX":[]}
     vars_dict = {"Vanilla":[],"ESVM":[],"EVM":[],"LS":[],"MAX":[]}
+    time_dict = {"Vanilla":[],"ESVM":[],"EVM":[],"LS":[],"MAX":[]}
     #initialize function values
     f_vals = set_function(f_type,[traj],vars_arr,params_test)
     #kill dimension which is not needed
     f_vals = f_vals[0]
     integrals,vars_spec = eval_samples("Vanilla",f_vals,traj,traj_grad,1,W_spec,N_test,d,vars_arr) #usual samples, without variance reduction
+    vanilla_time = time.time() - start_time
     res_dict["Vanilla"].append(integrals)
     vars_dict["Vanilla"].append(vars_spec)
+    time_dict["Vanilla"].append(vanilla_time)
     #set flag based upon the polynomial degree
     if degree == 2:
         flag = "2nd_order"
@@ -492,11 +497,14 @@ def Run_eval_test(intseed,degree,sampler,methods,vars_arr,Potential,test_dict,CV
         flag = "kth_order"
     #main loop
     for ind in range(len(methods)):
+        new_time = time.time()
         Coef_matr = CV_dict[methods[ind]]
[…]

### 2020-04-04 Code/potentials.py

@@ -272,13 +272,12 @@ class GausMixtureSame:
         returns gradient of log-density at point x
         """
         return self.Sigma_inv @ (self.mu - x) - 2*(1-self.p)*self.Sigma_inv @ self.mu/(1-self.p + self.p*np.exp(2*np.dot(self.mu,self.Sigma_inv @ x)))
-#################################################################################################################################
-    
+#################################################################################################################################    
 class BananaShape:
     """
     Implements Banana-shaped density potential function in R^2 for density f(x,y) = \exp{-\frac{x^2}{2M} - \frac{1}{2}(y+Bx^2-100B)^2}
     """
-    def __init__(self,B,M,d=2):
+    def __init__(self,B,M,d):
         self.B = B
         self.M = M
         self.d = d
@@ -311,6 +310,170 @@ class BananaShape:
             grad[i] = -z[i]
         return grad
 ################################################################################################################################ #### 
+class PotentialOne:
+    """
+    Implements denisty exp(-1/0.32*(x_2 - sin(pi/2*x_1))**2)
+    """
+    def __init__(self,B,M,d):
+        self.B = B
+        self.M = M
+        self.d = d
+        
+    def potential(self,z):
+        """
+        returns log-density at z
+        """
+        x = z[0]
+        y = z[1]
+        M = self.M
+        B = self.B
+        exponent = -1./(2*M)*x**2 - 1./2*(y-np.sin(B*x))**2
+        for i in range(2,self.d):
+            exponent -= 1./2*z[i]**2
+        return exponent
+    
+    def gradpotential(self,z):
+        """
+        returns gradient of log-density at point z
+        """
+        x = z[0]
+        y = z[1]
+        M = self.M
+        B = self.B
+        grad = np.zeros(self.d, dtype = float)
+        grad[0] = -x/M + B*(y-np.sin(B*x))*np.cos(B*x)
+        grad[1] = -(y - np.sin(B*x))
+        for i in range(2,self.d):
+            grad[i] = -z[i]
+        return grad
+################################################################################################################################ #### 
+class PotentialOnePrime:
+    """
[…]

### 2020-04-04 Code/samplers.py

@@ -118,7 +118,7 @@ def ULA_light(r_seed,Potential,step, N, n, d, return_noise = False, x0 = None, f
     #burn-in ended
     if fixed_start: #start all test trajectories from the same point
         x = x0
-    traj[0,] = x0
+    traj[0,] = x
     grad_main = Potential.gradpotential
     for k in np.arange(1,n): # samples
         grad = grad_main(x)

### 2020-04-04 Code/zv_cv.py

@@ -1,0 +1,147 @@
+import numpy as np
+from baselines import set_function
+
+def ZVpolyOne(traj, traj_grad, f_target, params = None):
+    n, d = traj.shape
+    if f_target == "sum":
+        samples = traj.sum(axis = 1).reshape(-1,1)
+    elif f_target == "sum_squared":
+        samples = np.square(traj).sum(axis = 1).reshape(-1,1)
+    elif f_target == "sum_4th":
+        samples = ((traj)**4).sum(axis = 1).reshape(-1,1)
+    elif f_target == "exp_sum":
+        samples = np.exp(traj.sum(axis = 1)).reshape(-1,1)
+    else:
+        traj = np.expand_dims(traj, axis = 0)
+        samples = set_function(f_target,traj,[0],params)
+        traj = traj[0]
+        samples = samples[0]
+    cov1 = np.cov(traj_grad, rowvar=False)
+    A = np.linalg.inv(cov1)
+    covariance = np.cov(np.concatenate((-traj_grad, samples), axis=1), rowvar=False)
+    paramZV1 = -np.dot(A,covariance[:d, d:])
+    ZV1 = samples - np.dot(traj_grad, paramZV1)
+    mean_ZV1 = np.mean(ZV1, axis = 0)
+    return mean_ZV1
+
+def ZVpolyTwo(traj, traj_grad, f_target, params = None):
+    n, d = traj.shape
+    if f_target == "sum":
+        samples = traj.sum(axis = 1).reshape(-1,1)
+    elif f_target == "sum_squared":
+        samples = np.square(traj).sum(axis = 1).reshape(-1,1)
+    elif f_target == "sum_4th":
+        samples = ((traj)**4).sum(axis = 1).reshape(-1,1)
+    elif f_target == "exp_sum":
+        samples = np.exp(traj.sum(axis = 1)).reshape(-1,1)
+    else:
+        traj = np.expand_dims(traj, axis = 0)
+        samples = set_function(f_target,traj,[0],params)
+        traj = traj[0]
+        samples = samples[0]
+    Lpoisson = np.zeros((n,int(d*(d+3)/2)))
+    Lpoisson[:,np.arange(d)] = - traj_grad
+    Lpoisson[:,np.arange(d, 2*d)] = 2*(1. - np.multiply(traj, traj_grad))
+    k=2*d
+    for j in np.arange(d-1):
[…]

### 2020-04-06 Code/samplers.py
update

@@ -179,7 +179,7 @@ def MALA(r_seed,Potential, step, N, n, d, burn_type = "SGLD",main_type = "SGLDFP
     grad_main = Potential.gradpotential
     traj = np.zeros((n, d))
     traj_grad = np.zeros((n, d))
-    x = np.random.normal(scale=5.0, size=d)
+    x = np.random.normal(scale=1.0, size=d)
     for k in np.arange(N):
         y = x + step * grad_burn(x) + np.sqrt(2*step)*np.random.normal(size=d)
         #if full gradient computed during burn-in, do acceptance-rejection step

### 2020-04-07 Code/martingale.py
update

@@ -253,14 +253,12 @@ def test_traj(Potential,coefs_poly_regr,step,r_seed,lag,K_max,S_max,N_burn,N_tes
                 a_vals[-(func_order+1),k] = np.dot(a_cur,coefs_poly_regr[func_order,:])
             #OK, now I have coefficients of the polynomial, and I need to integrate it w.r.t. Gaussian measure
         #print("sum of coefficients",np.sum(np.abs(a_vals)))
-        if (i%100 == 0):
-            print(a_vals)
         cvfs[i] = np.sum(a_vals*(poly_vals[:,i-num_lags+1:i+1].T))
         #save results
         test_stat_vanilla[i] = np.mean(f_vals_vanilla[1:(i+1)])
         test_stat_vr[i] = test_stat_vanilla[i] - np.sum(cvfs[1:(i+1)])/i
     end_time = time.time() - start_time
-    return test_stat_vanilla, test_stat_vr, end_time
+    return test_stat_vanilla, test_stat_vr
 
 def test_monte_carlo(r_seed,Potential,step,N_burn,N,d,return_noise, x0, fixed_start):
     """

### 2020-04-09 Code/martingale.py
update

@@ -253,6 +253,7 @@ def test_traj(Potential,coefs_poly_regr,step,r_seed,lag,K_max,S_max,N_burn,N_tes
                 a_vals[-(func_order+1),k] = np.dot(a_cur,coefs_poly_regr[func_order,:])
             #OK, now I have coefficients of the polynomial, and I need to integrate it w.r.t. Gaussian measure
         #print("sum of coefficients",np.sum(np.abs(a_vals)))
+        #print(a_vals)
         cvfs[i] = np.sum(a_vals*(poly_vals[:,i-num_lags+1:i+1].T))
         #save results
         test_stat_vanilla[i] = np.mean(f_vals_vanilla[1:(i+1)])

### 2020-04-09 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 we obtain closed-form expression for the estimates $\widehat a_{r,k}(x)$ of functions $\bar a_{r,k}(x)$ in \eqref{eq:definition-bar-a}. Namely, for all $x \in \mathbb{R}^d$,
-\begin{eqnarray*}
+\begin{eqnarray}
+\nonumber
 \widehat a_{r,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(\xi) \widehat{Q}_{r}(x - \gamma \mu(x) + \sqrt{\gamma}\xi)}{\TrainSet}
-= \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{k_i,s_i}(x_i),
-\end{eqnarray*}
+\\
+&=& 
+\label{eq:a-pol}
+\sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{k_i,s_i}(x_i - \gamma \mu_i(x)),
+\end{eqnarray}
 where for any integer $k,s$ $P_{k,s}$ 
@@
 \begin{eqnarray*}
-\widehat M_{K,n,n_{\mathrm{trunc}}}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{0 < \|\mathbf{k}\| < K}\sum_{l=N+1}^{p} \widehat a_{p-l,\mathbf{k}}(X_{l-1})\phi_k(\xi_{l})\mathbbm{1}\{ | p-l | < n_{\mathrm{trunc}}\right].
+\widehat M_{K,n,n_{\mathrm{trunc}}}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{0 < \|\mathbf{k}\|\leq K}\sum_{l=N+1}^{p} \widehat a_{p-l,\mathbf{k}}(X_{l-1})\mathbf{H}_\mathbf{k}(\xi_{l})\mathbbm{1}\{ | p-l | < n_{\mathrm{trunc}}\}\right].
 \end{eqnarray*}
@@
 \subsection{Gaussian mixtures}
- We consider a sample generated by ULA with $\pi$  given by the mixture of two equally-weighted $d$-dimensional Gaussian distributions of the following form
- \[
+ We consider  ULA with $\pi$  given by the mixture of two equally-weighted $2$-dimensional Gaussian distributions of the following form
+\begin{eqnarray}
+\label{eq:gmm}
 \pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}|\Sigma|}} \left( \rme^{-\frac{(x-\mu)^T\Sigma^{-1}(x-\mu)}{2}} + \rme^{-\frac{(x+\mu)^T\Sigma^{-1}(x+\mu)}{2}}\right)
- \]
-where $\rho \in (0,1), \mu \in \rset^d, \Sigma \in \rset^{d \times d}$ and $|\Sigma|$ is its determinant. The function $U(x)$ and its gradient are given by
-\[
-U(x) = \frac{1}{2}(x-\mu)^T\Sigma^{-1}(x-\mu) - \ln{\left(1 + \rme^{-2\mu^\top \Sigma^{-1}x}\right)}, \quad
-\]
+\end{eqnarray}
+where $\rho \in (0,1), \mu \in \rset^d, \Sigma \in \rset^{2 \times 2}$ and $|\Sigma|$ is its determinant. The function $U(x)$ and its gradient are given by
+\[
+U(x) = \frac{1}{2}(x-\mu)^T\Sigma^{-1}(x-\mu) - \ln{\left(1 + \rme^{-2\mu^\top \Sigma^{-1}x}\right)}
+\]
+and
 \[
@@
 respectively.
-In our experiments we considered $d = 2$, $\mu = \left(0.5,0.5\right)$ and randomly generated positive-definite matrix $\Sigma$ (heterogeneous structure).
-In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i\) or \(f(x) = \sum_{i=1}^d x^2_i\) we used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 10$ independent training trajectories, each one of size $n = 10^4$.
-Then we solve the least squares  problems \eqref{eq:Qregr} using the first order polynomial approximations for the coefficients \(\hat{a}_{p,\mathbf{k}}\) as described in the previous section, hence we set $K = 1$.
+In our experiments we take $\mu = \left(0.5,0.5\right)$ and randomly generated positive-definite matrix $\Sigma$ (heterogeneous structure).
+In order to approximate the expectation \(\pi(f)\) for \(f(x)=\sum_{i=1}^d x_i\) or \(f(x) = \sum_{i=1}^d x^2_i,\) we use ULA with constant step size $\gamma=0.1$ and sampled $\NtrainPath = 10$ independent training trajectories, each one of size $n = 10^4$ to estimate coefficients $a_{p-l,\mathbf{k}}.$ Then we solve the least squares  problems \eqref{eq:Qregr} using the first or second order polynomial basis  depending on $f$.
+Respectively, we set $K=1$ for linear functionals $f$ and $K=2$ for quadratic functionals.
 We set the truncation level \(n_{\mathrm{trunc}} = 50\).
-To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 10^3$ and plot boxplots of the ergodic averages in Figure ~\ref{fig:GMM}.
+To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 10^3$ and depict boxplots of the ergodic averages in Figure ~\ref{fig:GMM}.
 Our approach (MDCV, Martingale decomposition control variates) is compared to other variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}.
-In the baselines we use first order polynomials, since they are used as regressors during computations of $\hat{a}_{p-l,\mathbf{k}}$.
-
+In the baselines we use first order polynomials, as they were used as regressors during computations of $\hat{a}_{p-l,\mathbf{k}}$.
+Next we compute the cost-variance ratios
+\[
+\mathcal{R}(K,N,n,n_{\mathrm{trunc}})=\frac{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n,n_{\mathrm{trunc}} }^N(f)|\TrainSet]} 
+\]
+where variances in the above definition are computed using $24$ independent trajectories of the length $n=2\cdot10^3.$ Note that the cost of $\widehat \pi_{K,n,n_{\mathrm{trunc}}} $ is proportional to the number of nonzero coefficients in \eqref{eq:a-pol} times $n_{\mathrm{trunc}}$.
+In Figure~\ref{fig:G-costvar} cost-variances ratios are shown for for the case of  Gaussian distribution with parameters.
+In particular, in Figure~\ref{fig:subfig3}  a function $\mathcal{R}(1,2\cdot 10^3,2\cdot 10^3,n_{\mathrm{trunc}})$ is shown in dependence on $n_{\mathrm{trunc}}$ for the case of a linear functional $f$.
+In Figure~\ref{fig:subfig4} a similar plot of $\mathcal{R}(2,2\cdot 10^3,2\cdot 10^3,n_{\mathrm{trunc}})$ is shown for the case of a quadratic functional $f.$
+Figure~\ref{fig:GMM-costvar} presents the corresponding coast-variance ratios   for a mixture of two two-dimensional normal distributions as in \eqref{eq:gmm} with $\Sigma=I$ and $\mu=(0.5,0.5).$ 
  \begin{figure}[tbh]
@@
 \label{fig:GMM}}
+\end{figure}
+
+\begin{figure}[tbh]
+\centering
+\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
+\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_linear_gaussian.png}
+\label{fig:subfig3}}
+\qquad
+\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
+\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_quadratic_gaussian.png}
+\label{fig:subfig4}}
+\caption{ Cost-variance ratios as a function of the truncation level for  two-dimensional standard Gaussian distribution and different test functions. \label{fig:G-costvar}}
+\end{figure}
+
+\begin{figure}[tbh]
+\centering
+\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
+\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_linear_gmm.png}
+\label{fig:subfig5}}
+\qquad
+\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
+\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_quadratic_gmm.png}
+\label{fig:subfig6}}
+\caption{ Cost-variance ratios as a function of the truncation level for  a mixture of two-dimensional  Gaussian distribution and different test functions.
+\label{fig:GMM-costvar}}
 \end{figure}
@@
 Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$.
+Let $\tilde{f}(x) = f(x) - \pi(f)$.
 Then it holds for some constant $C > 0$
@@
 \label{lem:covariance}
-Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$. Then
+Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$.
+Let $\tilde{f}(x) = f(x) - \pi(f)$.
+Then
 \begin{equation}

### 2020-04-10 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 \begin{cases}
-Y_p, & \texttt{ with probability } \alpha(x,Y_p),
-\\
-x, &  \texttt{ with probability } 1-\alpha(x,Y_p),
+Y_p, & \texttt{ with probability } \alpha_p(x,Y_p),
+\\
+x, &  \texttt{ with probability } 1-\alpha_p(x,Y_p),
 \end{cases}
@@
 \begin{eqnarray*}
-\alpha(x,y)=\min\left\{1,\frac{\pi(y)}{\pi(x)}\frac{q_p(x|y)}{q_p(y|x)}\right\}.
+\alpha_p(x,y)=\min\left\{1,\frac{\pi(y)}{\pi(x)}\frac{q_p(x|y)}{q_p(y|x)}\right\}.
 \end{eqnarray*}
@@
 \cite{mira2013zero}) impossible in this case.
-On the contrary, our method can be directly used to reduce the variance of the ergodic estimator \(\pi_n^\gamma\) without explicit knowledge of \(\pi.\)
+%On the contrary, our method can be directly used to reduce the variance of the ergodic estimator \(\pi_n^\gamma\) without explicit knowledge of \(\pi.\)
 \end{example}
@@
 \\
-&=& \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\bar{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
+&=& \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\sum_{p=0}^{N+n-l}\bar{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l}).
 \label{eq:29032018a5}
@@
 \]
-for any integrable function $g.$  Hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
-The representation \eqref{eq:coeff_mart} suggests that the variance of  the   estimator
+for any  function $g.$  Hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
+The representation \eqref{eq:coeff_mart} suggests also that the variance of  the   estimator
 \begin{equation}
@@
 \par
-The coefficients \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. This amounts to solving a nonparametric regression problem.
+The coefficients \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. One way to estimate them is to use nonparametric regression.
 We first present a generic regression algorithm and then in Section~\ref{sec:num} give further implementation details. Our algorithm starts with estimating the functions 
@@
 \end{equation}
-for some $l_0$ with $l_0+q\leq N+n,$
+for $k=1,\ldots,K,$ and some $l_0$ with $l_0+q\leq N+n,$
 where \(\Psi\) is a  class of  real-valued functions on \(\mathbb{R}^d\) and $\xi^{(s)}_{1},\ldots,\xi^{(s)}_{N+n}$ are iid random variables with distribution $P_\xi$ used to construct the $s$-th trajectory of $X.$   As usual in regression analysis (see Section~11 in \cite{gyorfi2006distribution}),  we  truncate  the least squares estimates $(\widetilde{A}_{q,k})$ and set
@@
 \label{eq:trunc}
-\widehat{A}_{q,k}(y)=
+\widehat{A}_{q,k}(y):=
 \begin{cases}
@@
 \begin{eqnarray*}
-\widehat M_{K,n}^N(f) = \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\widehat{A}_{N+n-l,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
+\widehat M_{K,n}^N(f) := \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\widehat{A}_{N+n-l,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
 \end{eqnarray*}
@@
 Combining this with \eqref{eq:expression-variance} results in 
-\begin{multline*}
+\begin{multline}
+\label{eq:PVarA}
 \PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]=\frac{1}{n^2}\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})]
@@
 +\frac{1}{n^2}\sum_{k=1}^{K}\sum_{l=1}^{n}\PE\left[\left.|A_{n-l,k}(X^x_{N+l-1})-\widehat A_{n-l,k}(X^x_{N+l-1})|^2\right | \TrainSet \right].
-\end{multline*}
+\end{multline}
 In the next section we analyze the case of ULA and show that under some rather mild conditions (convexity of the potential $U$ outside a ball around zero)
@@
 \[
-\bar A_{q,k}=\argmin_{\psi\in\Psi}\mathsf{E}[|A_{q,k}(X_{l_0})-\psi(X_{l_0})|]^2
-\]
- as $T\to \infty,$ see Section~11 in \cite{gyorfi2006distribution}.  Moreover,
+\bar A_{q,k}:=\argmin_{\psi\in\Psi}\mathsf{E}[|A_{q,k}(X_{l_0})-\psi(X_{l_0})|]^2
+\]
+ as $T\to \infty,$ see Section~11 in \cite{gyorfi2006distribution}.  Moreover, it follows from \eqref{eq:PVarA} that
 \begin{eqnarray*}
@@
  Note that the above estimate is rather rough, as we ignore here the fact that the difference between $\widehat A_{n-l,k}$ and $A_{n-l,k}$ is small.
-If the class of functions $\Psi$ is a linear one of dimension $\mathtt{D}_\Psi,$ then the cost of computing $\widehat \pi_{K,n}^N(f)$ is of order $ \mathtt{D}_\Psi K T^2 n$.
-At the same time the variance of the standard estimate $ \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved, see Lemma~\ref{lem:variance} and Remark~\ref{rem:var-low}.
+If the class of functions $\Psi$ is a linear one of dimension $\mathtt{D}_\Psi,$ then the cost of computing the coefficients $\widehat A_{l,k}$ for all $l=1,\ldots,n$ and $k=1,\ldots,K$ is of order  $ \mathtt{D}_\Psi K T^2 n$.
+Given $(\widehat A_{l,k})$ the cost of computing  $\widehat \pi_{K,n}^N$ is proportional to $\mathtt{D}_\Psi K n.$ At the same time the variance of the standard estimate $ \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved, see Lemma~\ref{lem:variance} and Remark~\ref{rem:var-low}.
 Thus, the ratio of the corresponding cost-variances is of order
@@
 \end{eqnarray*}
-Thus, for any fixed $K\geq 1,$  our variance reduction method is advantageous if $\gamma\ll\min\{1/(\mathtt{D}_\Psi T^2),1/\sqrt{n}\}.$  
-\section{Analysis of variance reduced ULA}
+Thus, for any fixed $K\geq 1,$  our variance reduction method is advantageous if $\gamma\ll\min\{1/(\mathtt{D}_\Psi T^2),1/\sqrt{n}\}.$  Note that in order to achieve convergence of the corresponding MSE to zero, we need to let $\gamma\to 0$ as the invariant measure of the ULA with a constant time step $\gamma$ is not equal to $\pi.$\section{Analysis of variance reduced ULA}
 \label{sec:ula_analysis}
@@
 $$
-H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
+H_k(x):=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
 \quad x\in\mathbb R.
@@
 $$
-\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
+\mathbf{H}_\mathbf{k}(\mathbf{x}):=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
 $$
@@
 \begin{align}
-M_{K,n}^N(f) & =\sum_{ 0<\|\mathbf{k}\|\le K}\frac{1}{n}\sum_{l=1}^{n}A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})
+M_{K,n}^N(f) & :=\frac{1}{n}\sum_{ 0<\|\mathbf{k}\|\le K}\sum_{l=1}^{n}A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})
 \label{eq:29032018a5}
@@
 \begin{eqnarray*}
-A_{q,\mathbf{k}}(y)=\sum_{r=0}^{ q } \bar{a}_{r,\mathbf{k}}(y).
+A_{q,\mathbf{k}}(y):=\sum_{r=0}^{ q } \bar{a}_{r,\mathbf{k}}(y).
 \end{eqnarray*}
@@
 \begin{description}
-\item[(H1)][\textbf{Lipschitz continuity}] The potential $U$ is differentiable and $\nabla U$ is Lipschitz, that is, there exists $L < \infty$ such that
+\item[(H1)][\textbf{Lipschitz continuity}] The potential $U$ is differentiable and $\nabla U$ is Lipschitz continuous, that is, there exists $L < \infty$ such that
 \begin{eqnarray*}

### 2020-04-10 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 with \(\|\mathbf{k}\|=\max_{i} k_i\) and 
-\begin{eqnarray*}
+\begin{eqnarray}
+\label{eq:A-ula}
 A_{q,\mathbf{k}}(y):=\sum_{r=0}^{ q } \bar{a}_{r,\mathbf{k}}(y).
-\end{eqnarray*}
+\end{eqnarray}
 For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
@@
 \begin{eqnarray*}
-\PVar_{0} \pi_n^0(f) &=& \PVar_{0}\left[\frac{1}{n}\sum\limits_{i=1}^{n}X_i\right] 
+\PVar_{0} [\pi_n^0(f)] &=& \PVar_{0}\left[\frac{1}{n}\sum\limits_{i=1}^{n}X_i\right] 
 = \frac{1}{n^2}\PVar\left[\sum\limits_{i=1}^{n}\sum\limits_{j=1}^{i}\sqrt{\gamma}(1-2\gamma)^{i-j}\xi_{j}\right]
@@
 \end{remark}
-One of the main results of this paper is the following  upper bound for the functions $(A_{q,\mathbf{k}}).$
+One of the main results of this paper is the following  upper bound for the functions $(A_{q,\mathbf{k}})$ in \eqref{eq:A-ula}.
 \begin{thm}\label{th:mr}
@@
 \end{eqnarray}
-where $F_q(x,Z_1,\ldots,Z_q)=\sum_{r=0}^q f(X^x_r)$ and $\partial_{Z_1}^{\mathbf{k}'}$
+where $F_q(x,Z_1,\ldots,Z_q):=\sum_{r=0}^q f(X^x_r)$ and $\partial_{Z_1}^{\mathbf{k}'}$
 stands for a weak partial derivative of the functional $F_s$ that also can be viewed as discretised version of Malliavin derivative. 
@@
 \end{eqnarray*}
-Also from \eqref{eq:Ask-part} we can derive
+Also from \eqref{eq:Ask-part} we can derive that
 \begin{eqnarray*}
@@
 \]
-Finally, we show  that under our smoothness assumption \eqref{eq:smooth-mu} it holds
+Finally, using the Gaussian Poincare inequality, we show  that under our smoothness assumption \eqref{eq:smooth-mu} it holds
 \begin{equation}
@@
 In this section we illustrate the performance of the proposed variance reduction method for ULA.
-First we note that from  numerical point of view another representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
+First we note that from  numerical point of view another equivalent representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
 \begin{prop}
@@
 \end{prop}
- We construct a polynomial approximation  for each \(Q_r(x)\) in \eqref{eq:definition-bar-a} in the form:
+First we construct a polynomial approximation  for each \(Q_r(x)\)  in the form:
 \begin{equation*}

### 2020-04-10 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 In our experiments we take $\mu = \left(0.5,0.5\right)$ and randomly generated positive-definite matrix $\Sigma$ (heterogeneous structure).
-In order to approximate the expectation \(\pi(f)\) for \(f(x)=\sum_{i=1}^d x_i\) or \(f(x) = \sum_{i=1}^d x^2_i,\) we use ULA with constant step size $\gamma=0.1$ and sampled $\NtrainPath = 10$ independent training trajectories, each one of size $n = 10^4$ to estimate coefficients $a_{p-l,\mathbf{k}}.$ Then we solve the least squares  problems \eqref{eq:Qregr} using the first or second order polynomial basis  depending on $f$.
+In order to approximate the expectation \(\pi(f)\) for \(f(x)=\sum_{i=1}^d x_i\) or \(f(x) = \sum_{i=1}^d x^2_i,\) we use ULA with constant step size $\gamma=0.1$ and sampled $\NtrainPath = 10^4$ independent training trajectories to estimate coefficients $a_{p-l,\mathbf{k}}.$ Then we solve the least squares  problems \eqref{eq:Qregr} using the first or second order polynomial basis  depending on $f$.
 Respectively, we set $K=1$ for linear functionals $f$ and $K=2$ for quadratic functionals.
@@
 In the baselines we use first order polynomials, as they were used as regressors during computations of $\hat{a}_{p-l,\mathbf{k}}$.
+\par
 Next we compute the cost-variance ratios
-\[
+\begin{eqnarray}
+\label{eq:cvR}
 \mathcal{R}(K,N,n,n_{\mathrm{trunc}})=\frac{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n,n_{\mathrm{trunc}} }^N(f)|\TrainSet]} 
-\]
-where variances in the above definition are computed using $24$ independent trajectories of the length $n=2\cdot10^3.$ Note that the cost of $\widehat \pi_{K,n,n_{\mathrm{trunc}}} $ is proportional to the number of nonzero coefficients in \eqref{eq:a-pol} times $n_{\mathrm{trunc}}$.
+\end{eqnarray}
+where variances in the above definition are computed using $24$ independent trajectories each of  length $n=2\cdot10^3.$ Note that the cost of $\widehat \pi_{K,n,n_{\mathrm{trunc}}} $ is proportional to the number of nonzero coefficients in \eqref{eq:a-pol} times $n_{\mathrm{trunc}}$.
 In Figure~\ref{fig:G-costvar} cost-variances ratios are shown for for the case of  Gaussian distribution with parameters.
-In particular, in Figure~\ref{fig:subfig3}  a function $\mathcal{R}(1,2\cdot 10^3,2\cdot 10^3,n_{\mathrm{trunc}})$ is shown in dependence on $n_{\mathrm{trunc}}$ for the case of a linear functional $f$.
-In Figure~\ref{fig:subfig4} a similar plot of $\mathcal{R}(2,2\cdot 10^3,2\cdot 10^3,n_{\mathrm{trunc}})$ is shown for the case of a quadratic functional $f.$
-Figure~\ref{fig:GMM-costvar} presents the corresponding coast-variance ratios   for a mixture of two two-dimensional normal distributions as in \eqref{eq:gmm} with $\Sigma=I$ and $\mu=(0.5,0.5).$ 
+In particular, in Figure~\ref{fig:subfig3}  a function $\mathcal{R}(1,2\cdot 10^3,2\cdot 10^3,n_{\mathrm{trunc}})$ is shown in dependence on $n_{\mathrm{trunc}}$ for the case of  linear  $f$.
+In Figure~\ref{fig:subfig4} a similar plot of $\mathcal{R}(2,2\cdot 10^3,2\cdot 10^3,n_{\mathrm{trunc}})$ is shown for the case of  quadratic $f.$
+Figure~\ref{fig:GMM-costvar} presents the corresponding coast-variance ratios   for the mixture of two two-dimensional normal distributions as in \eqref{eq:gmm} with $\Sigma=I$ and $\mu=(0.5,0.5).$  In each case, in order estimate functions $(Q_r),$ we have used $T=10^4$ independent trajectories in the least-squares regression and polynomials of the same order as the order of $f.$
  \begin{figure}[tbh]
@@
 \label{fig:subfig4}}
-\caption{ Cost-variance ratios as a function of the truncation level for  two-dimensional standard Gaussian distribution and different test functions. \label{fig:G-costvar}}
+\caption{ Cost-variance ratios \eqref{eq:cvR} as  functions of the truncation level $n_{\mathrm{trunc}}$ for  two-dimensional standard Gaussian distribution and different test functions.
+\label{fig:G-costvar}}
 \end{figure}
@@
 \label{fig:subfig6}}
-\caption{ Cost-variance ratios as a function of the truncation level for  a mixture of two-dimensional  Gaussian distribution and different test functions.
+\caption{ Cost-variance ratios  \eqref{eq:cvR} as  functions of the truncation level $n_{\mathrm{trunc}}$ for  a mixture of two-dimensional  Gaussian distribution and different test functions.
 \label{fig:GMM-costvar}}

### 2020-04-11 aap/Revision/vr-mcmc-revision-new-new.tex
update

--- 2020-04-10 aap/Revision/vr-mcmc-revision-new-new.tex
@@
 
-Monte Carlo integration typically has an error variance of the form
-$\sigma^{2}/n,$ where $n$ is a sample size and \(\sigma^2\) is the variance of  integrand. We can make the variance smaller
-by using a larger value of $n$.  Alternatively,
- we can  reduce $\sigma^2$ instead of increasing the sample size \(n\). To this end, one can try to construct
-a new Monte Carlo experiment with the same expectation as the original one
-but with a lower variance $\sigma^2$. Methods to achieve this are known as variance
-reduction techniques. Variance reduction plays an important role in
-Monte Carlo and Markov Chain Monte Carlo methods.
-Introduction to many of the variance reduction techniques can be found in \cite{rubinstein2016simulation}, \cite{GobetBook} and \cite{glasserman2013monte}.
+Monte Carlo and Markov Chain Monte Carlo algorithms play crucial role in modern numerical analysis, finding various applications in such research areas as Bayesian Statistics.
+Suppose that we aim at computing $\pi(f) = \pi(f):= \PE\left[f(X)\right]$, where $X$ is a random vector in $\mathcal{X} \subseteq \rset^d$ admitting a density $\pi$ and $f$ is a square-integrable function $f:\mathcal{X} \to \rset$, $f \in L^2(\pi)$.
+Typically it is not possible to compute $\pi(f)$ analytically, leading to the Monte Carlo methods as a popular solution.
+Given an independent identically distributed observations $X_1,\ldots,X_n$ form $\pi$, we might estimate $\pi(f)$ by $\hat{\pi}(f):=1/n\sum_{k=1}^{n}f(X_k)$.
+The variance of such estimate equals $\sigma^2(f)/n$  with $\sigma^2(f)$ being the variance of the integrand.
+The first way to obtain tighter estimate $\hat{\pi}(f)$ is simply to increase the sample size $n$.
+Unfortunately, this solution might be prohibitively costly, especially when the dimension $d$ is large enough or sampling from $\pi$ is complicated.
+An alternative approach is to decrease $\sigma^2(f)$ by constructing a new Monte-Carlo experiment with the same expectation as the original one, but with a lower variance.
+Methods to achieve this are known as variance
+reduction techniques.
+Introduction to many of them can be found in \cite{rubinstein2016simulation}, \cite{GobetBook} and \cite{glasserman2013monte}.
 Recently one witnessed a revival of interest in efficient variance reduction methods  for MCMC algorithms, see for example \cite{dellaportas2012control}, \cite{mira2013zero}, \cite{brosse2018diffusion} and references therein.
 \par
-Suppose that we wish to compute $\pi(f):=\mathsf{E}\left[f(X)\right]$, where $X$
-is a random vector-valued in $\mathcal{X}\subseteq\mathbb{R}^{d}$ with a density $\pi$ and $f:$
-$\mathcal{X}\to\mathbb{R}$ with $f\in L^2(\pi)$.
-The idea of the control variates variance reduction method
- is to find a cheaply computable random variable $\zeta$ with $\mathsf{E}[\zeta]=0$ and \(\mathsf{E}[\zeta^2]<\infty,\)
-such that the variance of the r.v.
-$f(X)-\zeta$ is small.
+
+One of the popular approaches to variance reduction is the so called control variates method.
+We aim at constructing cheaply computable random variable $\zeta$ (control variate) with $\PE[\zeta] = 0$ and $\PE[\zeta^2] < \infty$, such that the variance of the r.v.
+$f(X) - \zeta$ is small.
 The complexity of the problem of constructing classes $Z$ of control variates \(\zeta\) satisfying   $\mathsf{E}[\zeta]=0$ essentially depends on the degree of our knowledge on \(\pi.\)
@@
 \begin{cases}
-Y_p, & \texttt{ with probability } \alpha_p(x,Y_p),
-\\
-x, &  \texttt{ with probability } 1-\alpha_p(x,Y_p),
+Y_p, & \texttt{ with probability } \alpha(x,Y_p),
+\\
+x, &  \texttt{ with probability } 1-\alpha(x,Y_p),
 \end{cases}
@@
 \begin{eqnarray*}
-\alpha_p(x,y)=\min\left\{1,\frac{\pi(y)}{\pi(x)}\frac{q_p(x|y)}{q_p(y|x)}\right\}.
+\alpha(x,y)=\min\left\{1,\frac{\pi(y)}{\pi(x)}\frac{q_p(x|y)}{q_p(y|x)}\right\}.
 \end{eqnarray*}
@@
 \cite{mira2013zero}) impossible in this case.
-%On the contrary, our method can be directly used to reduce the variance of the ergodic estimator \(\pi_n^\gamma\) without explicit knowledge of \(\pi.\)
+On the contrary, our method can be directly used to reduce the variance of the ergodic estimator \(\pi_n^\gamma\) without explicit knowledge of \(\pi.\)
 \end{example}
@@
 \\
-&=& \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\sum_{p=0}^{N+n-l}\bar{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l}).
+&=& \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\bar{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
 \label{eq:29032018a5}
@@
 \]
-for any  function $g.$  Hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
-The representation \eqref{eq:coeff_mart} suggests also that the variance of  the   estimator
+for any integrable function $g.$  Hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
+The representation \eqref{eq:coeff_mart} suggests that the variance of  the   estimator
 \begin{equation}
@@
 \par
-The coefficients \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. One way to estimate them is to use nonparametric regression.
+The coefficients \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. This amounts to solving a nonparametric regression problem.
 We first present a generic regression algorithm and then in Section~\ref{sec:num} give further implementation details. Our algorithm starts with estimating the functions 
@@
 \end{equation}
-for $k=1,\ldots,K,$ and some $l_0$ with $l_0+q\leq N+n,$
+for some $l_0$ with $l_0+q\leq N+n,$
 where \(\Psi\) is a  class of  real-valued functions on \(\mathbb{R}^d\) and $\xi^{(s)}_{1},\ldots,\xi^{(s)}_{N+n}$ are iid random variables with distribution $P_\xi$ used to construct the $s$-th trajectory of $X.$   As usual in regression analysis (see Section~11 in \cite{gyorfi2006distribution}),  we  truncate  the least squares estimates $(\widetilde{A}_{q,k})$ and set
@@
 \label{eq:trunc}
-\widehat{A}_{q,k}(y):=
+\widehat{A}_{q,k}(y)=
 \begin{cases}
@@
 \begin{eqnarray*}
-\widehat M_{K,n}^N(f) := \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\widehat{A}_{N+n-l,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
+\widehat M_{K,n}^N(f) = \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\widehat{A}_{N+n-l,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
 \end{eqnarray*}
@@
 Combining this with \eqref{eq:expression-variance} results in 
-\begin{multline}
-\label{eq:PVarA}
+\begin{multline*}
 \PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]=\frac{1}{n^2}\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})]
@@
 +\frac{1}{n^2}\sum_{k=1}^{K}\sum_{l=1}^{n}\PE\left[\left.|A_{n-l,k}(X^x_{N+l-1})-\widehat A_{n-l,k}(X^x_{N+l-1})|^2\right | \TrainSet \right].
-\end{multline}
+\end{multline*}
 In the next section we analyze the case of ULA and show that under some rather mild conditions (convexity of the potential $U$ outside a ball around zero)
@@
 \[
-\bar A_{q,k}:=\argmin_{\psi\in\Psi}\mathsf{E}[|A_{q,k}(X_{l_0})-\psi(X_{l_0})|]^2
-\]
- as $T\to \infty,$ see Section~11 in \cite{gyorfi2006distribution}.  Moreover, it follows from \eqref{eq:PVarA} that
+\bar A_{q,k}=\argmin_{\psi\in\Psi}\mathsf{E}[|A_{q,k}(X_{l_0})-\psi(X_{l_0})|]^2
+\]
+ as $T\to \infty,$ see Section~11 in \cite{gyorfi2006distribution}.  Moreover,
 \begin{eqnarray*}
@@
  Note that the above estimate is rather rough, as we ignore here the fact that the difference between $\widehat A_{n-l,k}$ and $A_{n-l,k}$ is small.
-If the class of functions $\Psi$ is a linear one of dimension $\mathtt{D}_\Psi,$ then the cost of computing the coefficients $\widehat A_{l,k}$ for all $l=1,\ldots,n$ and $k=1,\ldots,K$ is of order  $ \mathtt{D}_\Psi K T^2 n$.
-Given $(\widehat A_{l,k})$ the cost of computing  $\widehat \pi_{K,n}^N$ is proportional to $\mathtt{D}_\Psi K n.$ At the same time the variance of the standard estimate $ \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved, see Lemma~\ref{lem:variance} and Remark~\ref{rem:var-low}.
+If the class of functions $\Psi$ is a linear one of dimension $\mathtt{D}_\Psi,$ then the cost of computing $\widehat \pi_{K,n}^N(f)$ is of order $ \mathtt{D}_\Psi K T^2 n$.
+At the same time the variance of the standard estimate $ \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved, see Lemma~\ref{lem:variance} and Remark~\ref{rem:var-low}.
 Thus, the ratio of the corresponding cost-variances is of order
-\begin{eqnarray*}
+\begin{equation}
+\label{eq:cost_theoretical}
 \frac{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]}{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}\leq C e^{\varkappa n\gamma^2} K \gamma \, T^2 \mathtt{D}_\Psi. 
-\end{eqnarray*}
-Thus, for any fixed $K\geq 1,$  our variance reduction method is advantageous if $\gamma\ll\min\{1/(\mathtt{D}_\Psi T^2),1/\sqrt{n}\}.$  Note that in order to achieve convergence of the corresponding MSE to zero, we need to let $\gamma\to 0$ as the invariant measure of the ULA with a constant time step $\gamma$ is not equal to $\pi.$\section{Analysis of variance reduced ULA}
+\end{equation}
+Thus, for any fixed $K\geq 1,$  our variance reduction method is advantageous if $\gamma\ll\min\{1/(\mathtt{D}_\Psi T^2),1/\sqrt{n}\}.$  
+\section{Analysis of variance reduced ULA}
 \label{sec:ula_analysis}
@@
 $$
-H_k(x):=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
+H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
 \quad x\in\mathbb R.
@@
 $$
-\mathbf{H}_\mathbf{k}(\mathbf{x}):=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
+\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
 $$
@@
 \begin{align}
-M_{K,n}^N(f) & :=\frac{1}{n}\sum_{ 0<\|\mathbf{k}\|\le K}\sum_{l=1}^{n}A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})
+M_{K,n}^N(f) & =\sum_{ 0<\|\mathbf{k}\|\le K}\frac{1}{n}\sum_{l=1}^{n}A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})
 \label{eq:29032018a5}
@@
 \begin{eqnarray*}
-A_{q,\mathbf{k}}(y):=\sum_{r=0}^{ q } \bar{a}_{r,\mathbf{k}}(y).
+A_{q,\mathbf{k}}(y)=\sum_{r=0}^{ q } \bar{a}_{r,\mathbf{k}}(y).
 \end{eqnarray*}
@@
 \begin{description}
-\item[(H1)][\textbf{Lipschitz continuity}] The potential $U$ is differentiable and $\nabla U$ is Lipschitz continuous, that is, there exists $L < \infty$ such that
+\item[(H1)][\textbf{Lipschitz continuity}] The potential $U$ is differentiable and $\nabla U$ is Lipschitz, that is, there exists $L < \infty$ such that
 \begin{eqnarray*}
@@
 we obtain closed-form expression for the estimates $\widehat a_{r,k}(x)$ of functions $\bar a_{r,k}(x)$ in \eqref{eq:definition-bar-a}. Namely, for all $x \in \mathbb{R}^d$,
-\begin{eqnarray}
-\nonumber
-\widehat a_{r,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(\xi) \widehat{Q}_{r}(x - \gamma \mu(x) + \sqrt{\gamma}\xi)}{\TrainSet}
-\\
-&=& 
-\label{eq:a-pol}
-\sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{k_i,s_i}(x_i - \gamma \mu_i(x)),
-\end{eqnarray}
+\begin{equation}
+\label{eq:a_hat}
+\widehat a_{r,\mathbf{k}} (x) = \CPE{\mathbf{H}_\mathbf{k}(\xi) \widehat{Q}_{r}(x - \gamma \mu(x) + \sqrt{\gamma}\xi)}{\TrainSet}
+= \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{k_i,s_i}(x_i - \gamma \mu_i(x)),
+\end{equation}
 where for any integer $k,s$ $P_{k,s}$ 
@@
 \begin{eqnarray*}
-\widehat M_{K,n,n_{\mathrm{trunc}}}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{0 < \|\mathbf{k}\|\leq K}\sum_{l=N+1}^{p} \widehat a_{p-l,\mathbf{k}}(X_{l-1})\mathbf{H}_\mathbf{k}(\xi_{l})\mathbbm{1}\{ | p-l | < n_{\mathrm{trunc}}\}\right].
-\end{eqnarray*}
+\widehat M_{K,n,n_{\mathrm{trunc}}}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{0 < \|\mathbf{k}\|\leq K}\sum_{l=N+1}^{p} \widehat a_{p-l,\mathbf{k}}(X_{l-1})\phi_k(\xi_{l})\mathbbm{1}\{ | p-l | < n_{\mathrm{trunc}}\}\right].
+\end{eqnarray*}
+
+\subsection{Comparison with vanilla ULA}
+In this subsection we aim at comparing cost of variance reduction, achieved by the proposed algorithm, to the cost of vanilla ULA algorithm.
+We consider samples, generated by ULA with $\pi$ being either the standard normal distribution in dimension $d$ or the mixture of two $d-$dimensional standard Gaussians of the form
+ \[
+ \pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}}} \left( \rme^{-\frac{\|x-\mu\|^2}{2}} + \rme^{-\frac{\|x+\mu\|^2}{2}}\right)
+ \]
+ where $d = 2$ and $\mu = (0.5,0.5)$.
+For both examples, we aimed at estimating $\pi(f)$ with $f(x)=\sum_{i=1}^d x_i$ and $f(x) = \sum_{i=1}^d x^2_i$.
+We used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 5 \times 10^4$ independent training trajectories, each one of size $n = 50$ with the burn-in period $n_{\text{burn}} = 100$.
+Then we solve the least squares  problems \eqref{eq:Qregr} using the first order polynomial approximations for $f(x)=\sum_{i=1}^d x_i$ and second order approximations for $f(x) = \sum_{i=1}^d x^2_i$ as described in the previous section.
+We set $K=1$ or $K=2$ for $f(x)=\sum_{i=1}^d x_i$ or $f(x)=\sum_{i=1}^d x^2_i$ respectively.
+Then we estimate the quantity
+\begin{eqnarray*}
+\frac{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]}{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}
+\end{eqnarray*}
+by its empirical counterpart, computed over $24$ independent trajectories with $n = 2\times 10^3$ and $N = 2 \times 10^3$.
+For different $n_{\mathrm{trunc}}$ and $K$ we estimate $\mathrm{cost}(\widehat \pi_{K,n}^N)$ as
+ \[
+ \mathrm{cost}(\widehat \pi_{K,n}^N) = \mathrm{cost}(\pi_{n}^N) \times n_{\mathrm{trunc}} \times d^{K}
+ \]
+ since the representations of $a_{r,\mathbf{k}} (x)$ might be pre-computed using formulas \eqref{eq:a_hat}.
+Variance reduction costs for Gaussian potential and different $n_{\mathrm{trunc}}$  are summarized in Figure~\ref{fig:gaus_cost}, and for the Gaussian mixture - in Figure~\ref{fig:gmm_cost}.
+Note that for both examples the proposed algorithm allows us to obtain a sufficient gain in variance reduction cost.
+
+\begin{figure}[tbh]
+\centering
+\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
+\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_linear_gaussian.png}
+\label{fig:subfig1}}
+\qquad
+\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
+\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_quadratic_gaussian.png}
+\label{fig:subfig2}}
+\caption{VR cost of MDCV algorithm compared to vanilla ULA on Gaussian potential
+\label{fig:gaus_cost}}
+\end{figure}
+
+\begin{figure}[tbh]
+\centering
+\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
+\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_linear_gmm.png}
+\label{fig:subfig1}}
+\qquad
+\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
+\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_quadratic_gmm.png}
+\label{fig:subfig2}}
+\caption{VR cost of MDCV algorithm compared to vanilla ULA on Gaussian mixture potential
+\label{fig:gmm_cost}}
+\end{figure}
 
 \subsection{Gaussian mixtures}
- We consider  ULA with $\pi$  given by the mixture of two equally-weighted $2$-dimensional Gaussian distributions of the following form
-\begin{eqnarray}
-\label{eq:gmm}
+ We consider a sample generated by ULA with $\pi$  given by the mixture of two equally-weighted $d$-dimensional Gaussian distributions of the following form
+ \[
  \pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}|\Sigma|}} \left( \rme^{-\frac{(x-\mu)^T\Sigma^{-1}(x-\mu)}{2}} + \rme^{-\frac{(x+\mu)^T\Sigma^{-1}(x+\mu)}{2}}\right)
-\end{eqnarray}
-where $\rho \in (0,1), \mu \in \rset^d, \Sigma \in \rset^{2 \times 2}$ and $|\Sigma|$ is its determinant. The function $U(x)$ and its gradient are given by
-\[
-U(x) = \frac{1}{2}(x-\mu)^T\Sigma^{-1}(x-\mu) - \ln{\left(1 + \rme^{-2\mu^\top \Sigma^{-1}x}\right)}
-\]
-and
+ \]
+where $\mu \in \rset^d, \Sigma \in \rset^{d \times d}$ and $|\Sigma|$ is its determinant. The function $U(x)$ and its gradient are given by
+\[
+U(x) = \frac{1}{2}(x-\mu)^T\Sigma^{-1}(x-\mu) - \ln{\left(1 + \rme^{-2\mu^\top \Sigma^{-1}x}\right)}, \quad
+\]
 \[
@@
 respectively.
-In our experiments we take $\mu = \left(0.5,0.5\right)$ and randomly generated positive-definite matrix $\Sigma$ (heterogeneous structure).
-In order to approximate the expectation \(\pi(f)\) for \(f(x)=\sum_{i=1}^d x_i\) or \(f(x) = \sum_{i=1}^d x^2_i,\) we use ULA with constant step size $\gamma=0.1$ and sampled $\NtrainPath = 10$ independent training trajectories, each one of size $n = 10^4$ to estimate coefficients $a_{p-l,\mathbf{k}}.$ Then we solve the least squares  problems \eqref{eq:Qregr} using the first or second order polynomial basis  depending on $f$.
-Respectively, we set $K=1$ for linear functionals $f$ and $K=2$ for quadratic functionals.
+In our experiments we considered $d = 2$, $\mu = \left(0.5,0.5\right)$ and randomly generated positive-definite matrix $\Sigma$ (heterogeneous structure).
+In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i\) or \(f(x) = \sum_{i=1}^d x^2_i\) we used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 5 \times 10^4$ independent training trajectories, each one of size $n = 50$ with the burn-in period $n_{\text{burn}} = 100$.
+Then we solve the least squares  problems \eqref{eq:Qregr} using the first order polynomial approximations for $f(x)=\sum_{i=1}^d x_i$ and second order approximations for $f(x) = \sum_{i=1}^d x^2_i$ as described in the previous section.
+Hence we set $K = 1$ for $f(x)=\sum_{i=1}^d x_i$ or $K = 2$ for $f(x) = \sum_{i=1}^d x^2_i$.
 We set the truncation level \(n_{\mathrm{trunc}} = 50\).
-To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 10^3$ and depict boxplots of the ergodic averages in Figure ~\ref{fig:GMM}.
+To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 2 \times 10^3$ and plot boxplots of the corresponding ergodic averages in Figure ~\ref{fig:GMM}.
 Our approach (MDCV, Martingale decomposition control variates) is compared to other variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}.
-In the baselines we use first order polynomials, as they were used as regressors during computations of $\hat{a}_{p-l,\mathbf{k}}$.
-Next we compute the cost-variance ratios
-\[
-\mathcal{R}(K,N,n,n_{\mathrm{trunc}})=\frac{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n,n_{\mathrm{trunc}} }^N(f)|\TrainSet]} 
-\]
-where variances in the above definition are computed using $24$ independent trajectories of the length $n=2\cdot10^3.$ Note that the cost of $\widehat \pi_{K,n,n_{\mathrm{trunc}}} $ is proportional to the number of nonzero coefficients in \eqref{eq:a-pol} times $n_{\mathrm{trunc}}$.
-In Figure~\ref{fig:G-costvar} cost-variances ratios are shown for for the case of  Gaussian distribution with parameters.
-In particular, in Figure~\ref{fig:subfig3}  a function $\mathcal{R}(1,2\cdot 10^3,2\cdot 10^3,n_{\mathrm{trunc}})$ is shown in dependence on $n_{\mathrm{trunc}}$ for the case of a linear functional $f$.
-In Figure~\ref{fig:subfig4} a similar plot of $\mathcal{R}(2,2\cdot 10^3,2\cdot 10^3,n_{\mathrm{trunc}})$ is shown for the case of a quadratic functional $f.$
-Figure~\ref{fig:GMM-costvar} presents the corresponding coast-variance ratios   for a mixture of two two-dimensional normal distributions as in \eqref{eq:gmm} with $\Sigma=I$ and $\mu=(0.5,0.5).$ 
+In the baselines we use first order polynomials in case of $K=1$ and second order polynomials for $K = 2$.
+
  \begin{figure}[tbh]
@@
 
-\begin{figure}[tbh]
-\centering
-\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
-\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_linear_gaussian.png}
-\label{fig:subfig3}}
-\qquad
-\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
-\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_quadratic_gaussian.png}
-\label{fig:subfig4}}
-\caption{ Cost-variance ratios as a function of the truncation level for  two-dimensional standard Gaussian distribution and different test functions. \label{fig:G-costvar}}
-\end{figure}
-
-\begin{figure}[tbh]
-\centering
-\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
-\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_linear_gmm.png}
-\label{fig:subfig5}}
-\qquad
-\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
-\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_quadratic_gmm.png}
-\label{fig:subfig6}}
-\caption{ Cost-variance ratios as a function of the truncation level for  a mixture of two-dimensional  Gaussian distribution and different test functions.
-\label{fig:GMM-costvar}}
-\end{figure}
-
 \subsection{Banana shape distribution}
@@
 \[
-U(x_1, . . . , x_d) = \frac{x_1^2}{2} + (x_2 + bx_1^2 - pb)^2 + \sum\limits_{k=3}^{d}\frac{x_k^2}{2}
+U(x_1,\ldots, x_d) = \frac{x_1^2}{2} + \frac{(x_2 + bx_1^2 - pb)^2}{2} + \sum\limits_{k=3}^{d}\frac{x_k^2}{2}
 \]
@@
 We solve the least squares  problems \eqref{eq:Qregr} using the second-order polynomial approximations for the coefficients \(\hat{a}_{p,\mathbf{k}}\) as described in the previous section, hence we set $K = 2$.
-We use $T=10$ independent training trajectories, each of size $n=10^4$.
+We use $T= 5 \times 10^4$ independent training trajectories, each of size $n=50$ with the burn-in $n_{\text{burn}} = 100$.
 We set the truncation level \(n_{\mathrm{trunc}} = 50\).
-To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 10^3$ and plot boxplots of the ergodic averages in Figure~\ref{fig:banana_shape}.
+To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 2 \times 10^3$ and plot boxplots of the ergodic averages in Figure~\ref{fig:banana_shape}.
 

### 2020-04-11 aap/Revision/vr-mcmc-revision-new-new.tex
merge conflict

@@
 \end{equation}
-Thus, for any fixed $K\geq 1,$  our variance reduction method is advantageous if $\gamma\ll\min\{1/(\mathtt{D}_\Psi T^2),1/\sqrt{n}\}.$  
-\section{Analysis of variance reduced ULA}
-\label{sec:ula_analysis}
+Thus, for any fixed $K\geq 1,$  our variance reduction method is advantageous if $\gamma\ll\min\{1/(\mathtt{D}_\Psi T^2),1/\sqrt{n}\}.$  Note that in order to achieve convergence of the corresponding MSE to zero, we need to let $\gamma\to 0$ as the invariant measure of the ULA with a constant time step $\gamma$ is not equal to $\pi.$
+\section{Analysis of variance reduced ULA}\label{sec:ula_analysis}
 In this section we perform the convergence analysis of the ULA algorithm.
@@
 with \(\|\mathbf{k}\|=\max_{i} k_i\) and 
-\begin{eqnarray*}
-A_{q,\mathbf{k}}(y)=\sum_{r=0}^{ q } \bar{a}_{r,\mathbf{k}}(y).
-\end{eqnarray*}
+\begin{eqnarray}
+\label{eq:A-ula}
+A_{q,\mathbf{k}}(y):=\sum_{r=0}^{ q } \bar{a}_{r,\mathbf{k}}(y).
+\end{eqnarray}
 For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
@@
 we obtain closed-form expression for the estimates $\widehat a_{r,k}(x)$ of functions $\bar a_{r,k}(x)$ in \eqref{eq:definition-bar-a}. Namely, for all $x \in \mathbb{R}^d$,
-\begin{equation}
-\label{eq:a_hat}
-\widehat a_{r,\mathbf{k}} (x) = \CPE{\mathbf{H}_\mathbf{k}(\xi) \widehat{Q}_{r}(x - \gamma \mu(x) + \sqrt{\gamma}\xi)}{\TrainSet}
-= \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{k_i,s_i}(x_i - \gamma \mu_i(x)),
-\end{equation}
+\begin{eqnarray}
+\nonumber
+\widehat a_{r,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(\xi) \widehat{Q}_{r}(x - \gamma \mu(x) + \sqrt{\gamma}\xi)}{\TrainSet}
+\\
+&=& 
+\label{eq:a-pol}
+\sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{k_i,s_i}(x_i - \gamma \mu_i(x)),
+\end{eqnarray}
 where for any integer $k,s$ $P_{k,s}$ 
@@
 \begin{eqnarray*}
-\widehat M_{K,n,n_{\mathrm{trunc}}}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{0 < \|\mathbf{k}\|\leq K}\sum_{l=N+1}^{p} \widehat a_{p-l,\mathbf{k}}(X_{l-1})\phi_k(\xi_{l})\mathbbm{1}\{ | p-l | < n_{\mathrm{trunc}}\}\right].
+\widehat M_{K,n,n_{\mathrm{trunc}}}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{0 < \|\mathbf{k}\|\leq K}\sum_{l=N+1}^{p} \widehat a_{p-l,\mathbf{k}}(X_{l-1})\mathbf{H}_\mathbf{k}(\xi_{l})\mathbbm{1}\{ | p-l | < n_{\mathrm{trunc}}\}\right].
 \end{eqnarray*}
@@
 \subsection{Gaussian mixtures}
- We consider a sample generated by ULA with $\pi$  given by the mixture of two equally-weighted $d$-dimensional Gaussian distributions of the following form
- \[
+ We consider  ULA with $\pi$  given by the mixture of two equally-weighted $2$-dimensional Gaussian distributions of the following form
+\begin{eqnarray}
+\label{eq:gmm}
 \pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}|\Sigma|}} \left( \rme^{-\frac{(x-\mu)^T\Sigma^{-1}(x-\mu)}{2}} + \rme^{-\frac{(x+\mu)^T\Sigma^{-1}(x+\mu)}{2}}\right)
- \]
-where $\mu \in \rset^d, \Sigma \in \rset^{d \times d}$ and $|\Sigma|$ is its determinant. The function $U(x)$ and its gradient are given by
-\[
-U(x) = \frac{1}{2}(x-\mu)^T\Sigma^{-1}(x-\mu) - \ln{\left(1 + \rme^{-2\mu^\top \Sigma^{-1}x}\right)}, \quad
-\]
+\end{eqnarray}
+where $\rho \in (0,1), \mu \in \rset^d, \Sigma \in \rset^{2 \times 2}$ and $|\Sigma|$ is its determinant. The function $U(x)$ and its gradient are given by
+\[
+U(x) = \frac{1}{2}(x-\mu)^T\Sigma^{-1}(x-\mu) - \ln{\left(1 + \rme^{-2\mu^\top \Sigma^{-1}x}\right)}
+\]
+and
 \[
@@
 In the baselines we use first order polynomials in case of $K=1$ and second order polynomials for $K = 2$.
-
+\par
+
+%Next we compute the cost-variance ratios
+%\begin{eqnarray}
+%\label{eq:cvR}
+%\mathcal{R}(K,N,n,n_{\mathrm{trunc}})=\frac{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n,n_{\mathrm{trunc}} }^N(f)|\TrainSet]} 
+%\end{eqnarray}
+%where variances in the above definition are computed using $24$ independent trajectories each of  length $n=2\cdot10^3.$ Note that the cost of $\widehat \pi_{K,n,n_{\mathrm{trunc}}} $ is proportional to the number of nonzero coefficients in \eqref{eq:a-pol} times $n_{\mathrm{trunc}}$.
+In Figure~\ref{fig:G-costvar} cost-variances ratios are shown for for the case of  Gaussian distribution with parameters.
+In particular, in Figure~\ref{fig:subfig3}  a function $\mathcal{R}(1,2\cdot 10^3,2\cdot 10^3,n_{\mathrm{trunc}})$ is shown in dependence on $n_{\mathrm{trunc}}$ for the case of  linear  $f$.
+In Figure~\ref{fig:subfig4} a similar plot of $\mathcal{R}(2,2\cdot 10^3,2\cdot 10^3,n_{\mathrm{trunc}})$ is shown for the case of  quadratic $f.$
+%Figure~\ref{fig:GMM-costvar} presents the corresponding coast-variance ratios   for the mixture of two two-dimensional normal distributions as in \eqref{eq:gmm} with $\Sigma=I$ and $\mu=(0.5,0.5).$  In each case, in order estimate functions $(Q_r),$ we have used $T=10^4$ independent trajectories in the least-squares regression and polynomials of the same order as the order of $f.$
  \begin{figure}[tbh]
@@
 \end{figure}
+
+%\begin{figure}[tbh]
+%\centering
+%\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
+%\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_linear_gaussian.png}
+%\label{fig:subfig3}}
+%\qquad
+%\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
+%\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_quadratic_gaussian.png}
+%\label{fig:subfig4}}
+%\caption{ Cost-variance ratios \eqref{eq:cvR} as  functions of the truncation level $n_{\mathrm{trunc}}$ for  two-dimensional standard Gaussian distribution and different test functions.
+\label{fig:G-costvar}}
+%\end{figure}
+
+%\begin{figure}[tbh]
+%\centering
+%\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
+%\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_linear_gmm.png}
+%\label{fig:subfig5}}
+%\qquad
+%\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
+%\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_quadratic_gmm.png}
+%\label{fig:subfig6}}
+%\caption{ Cost-variance ratios  \eqref{eq:cvR} as  functions of the truncation level $n_{\mathrm{trunc}}$ for  a mixture of two-dimensional  Gaussian distribution and different test functions.
+%\label{fig:GMM-costvar}}
+%\end{figure}
 

### 2020-04-11 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 Monte Carlo and Markov Chain Monte Carlo algorithms play crucial role in modern numerical analysis, finding various applications in such research areas as Bayesian Statistics.
-Suppose that we aim at computing $\pi(f) = \pi(f):= \PE\left[f(X)\right]$, where $X$ is a random vector in $\mathcal{X} \subseteq \rset^d$ admitting a density $\pi$ and $f$ is a square-integrable function $f:\mathcal{X} \to \rset$, $f \in L^2(\pi)$.
+Suppose that we aim at computing $\pi(f):= \PE\left[f(X)\right]$, where $X$ is a random vector in $\mathcal{X} \subseteq \rset^d$ admitting a density $\pi$ and $f$ is a square-integrable function $f:\mathcal{X} \to \rset$, $f \in L^2(\pi)$.
 Typically it is not possible to compute $\pi(f)$ analytically, leading to the Monte Carlo methods as a popular solution.
@@
 For both examples, we aimed at estimating $\pi(f)$ with $f(x)=\sum_{i=1}^d x_i$ and $f(x) = \sum_{i=1}^d x^2_i$.
-We used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 5 \times 10^4$ independent training trajectories, each one of size $n = 50$ with the burn-in period $n_{\text{burn}} = 100$.
-Then we solve the least squares  problems \eqref{eq:Qregr} using the first order polynomial approximations for $f(x)=\sum_{i=1}^d x_i$ and second order approximations for $f(x) = \sum_{i=1}^d x^2_i$ as described in the previous section.
+We used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 5 \times 10^4$ independent training trajectories, each one with the burn-in period $n_{\text{burn}} = 100$.
+Then we solve the least squares  problems \eqref{eq:Qregr} using the first order polynomial approximations for $f(x)=\sum_{i=1}^d x_i$ and second order polynomial approximations for $f(x) = \sum_{i=1}^d x^2_i$ as described in the previous section.
 We set $K=1$ or $K=2$ for $f(x)=\sum_{i=1}^d x_i$ or $f(x)=\sum_{i=1}^d x^2_i$ respectively.
-Then we estimate the quantity
+Then we estimate the cost-variance ratio
 \begin{eqnarray*}
@@
 For different $n_{\mathrm{trunc}}$ and $K$ we estimate $\mathrm{cost}(\widehat \pi_{K,n}^N)$ as
- \[
+ \begin{equation}
+ \label{eq:cvR}
+ %check this
+ %add n_{truncation} here
  \mathrm{cost}(\widehat \pi_{K,n}^N) = \mathrm{cost}(\pi_{n}^N) \times n_{\mathrm{trunc}} \times d^{K}
- \]
+ \end{equation}
  since the representations of $a_{r,\mathbf{k}} (x)$ might be pre-computed using formulas \eqref{eq:a_hat}.
@@
 Note that for both examples the proposed algorithm allows us to obtain a sufficient gain in variance reduction cost.
+
+%Next we compute the cost-variance ratios
+%\begin{eqnarray}
+%\label{eq:cvR}
+%\mathcal{R}(K,N,n,n_{\mathrm{trunc}})=\frac{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n,n_{\mathrm{trunc}} }^N(f)|\TrainSet]} 
+%\end{eqnarray}
+%where variances in the above definition are computed using $24$ independent trajectories each of  length $n=2\cdot10^3.$ Note that the cost of $\widehat \pi_{K,n,n_{\mathrm{trunc}}} $ is proportional to the number of nonzero coefficients in \eqref{eq:a-pol} times $n_{\mathrm{trunc}}$.
+In Figure~\ref{fig:G-costvar} cost-variances ratios are shown for for the case of  Gaussian distribution with parameters.
+In particular, in Figure~\ref{fig:subfig3}  a function $\mathcal{R}(1,2\cdot 10^3,2\cdot 10^3,n_{\mathrm{trunc}})$ is shown in dependence on $n_{\mathrm{trunc}}$ for the case of  linear  $f$.
+In Figure~\ref{fig:subfig4} a similar plot of $\mathcal{R}(2,2\cdot 10^3,2\cdot 10^3,n_{\mathrm{trunc}})$ is shown for the case of  quadratic $f.$
+%Figure~\ref{fig:GMM-costvar} presents the corresponding coast-variance ratios   for the mixture of two two-dimensional normal distributions as in \eqref{eq:gmm} with $\Sigma=I$ and $\mu=(0.5,0.5).$  In each case, in order estimate functions $(Q_r),$ we have used $T=10^4$ independent trajectories in the least-squares regression and polynomials of the same order as the order of $f.$
 
@@
 \label{fig:subfig2}}
-\caption{VR cost of MDCV algorithm compared to vanilla ULA on Gaussian potential
+\caption{Cost-variance ratios \eqref{eq:cvR} as  functions of the truncation level $n_{\mathrm{trunc}}$ for  two-dimensional standard Gaussian distribution and different test functions.
 \label{fig:gaus_cost}}
@@
 \label{fig:subfig2}}
-\caption{VR cost of MDCV algorithm compared to vanilla ULA on Gaussian mixture potential
+\caption{Cost-variance ratios  \eqref{eq:cvR} as  functions of the truncation level $n_{\mathrm{trunc}}$ for a mixture of two-dimensional Gaussian distributions and different test functions.
 \label{fig:gmm_cost}}
@@
 
-%Next we compute the cost-variance ratios
-%\begin{eqnarray}
-%\label{eq:cvR}
-%\mathcal{R}(K,N,n,n_{\mathrm{trunc}})=\frac{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n,n_{\mathrm{trunc}} }^N(f)|\TrainSet]} 
-%\end{eqnarray}
-%where variances in the above definition are computed using $24$ independent trajectories each of  length $n=2\cdot10^3.$ Note that the cost of $\widehat \pi_{K,n,n_{\mathrm{trunc}}} $ is proportional to the number of nonzero coefficients in \eqref{eq:a-pol} times $n_{\mathrm{trunc}}$.
-In Figure~\ref{fig:G-costvar} cost-variances ratios are shown for for the case of  Gaussian distribution with parameters.
-In particular, in Figure~\ref{fig:subfig3}  a function $\mathcal{R}(1,2\cdot 10^3,2\cdot 10^3,n_{\mathrm{trunc}})$ is shown in dependence on $n_{\mathrm{trunc}}$ for the case of  linear  $f$.
-In Figure~\ref{fig:subfig4} a similar plot of $\mathcal{R}(2,2\cdot 10^3,2\cdot 10^3,n_{\mathrm{trunc}})$ is shown for the case of  quadratic $f.$
-%Figure~\ref{fig:GMM-costvar} presents the corresponding coast-variance ratios   for the mixture of two two-dimensional normal distributions as in \eqref{eq:gmm} with $\Sigma=I$ and $\mu=(0.5,0.5).$  In each case, in order estimate functions $(Q_r),$ we have used $T=10^4$ independent trajectories in the least-squares regression and polynomials of the same order as the order of $f.$
  \begin{figure}[tbh]

### 2020-04-12 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 
-Monte Carlo and Markov Chain Monte Carlo algorithms play crucial role in modern numerical analysis, finding various applications in such research areas as Bayesian Statistics.
+Monte Carlo and Markov Chain Monte Carlo algorithms play crucial role in modern numerical analysis, finding various applications in such research areas as Bayesian statistics.
 Suppose that we aim at computing $\pi(f):= \PE\left[f(X)\right]$, where $X$ is a random vector in $\mathcal{X} \subseteq \rset^d$ admitting a density $\pi$ and $f$ is a square-integrable function $f:\mathcal{X} \to \rset$, $f \in L^2(\pi)$.
@@
 \par  
-
 One of the popular approaches to variance reduction is the so called control variates method.
 We aim at constructing cheaply computable random variable $\zeta$ (control variate) with $\PE[\zeta] = 0$ and $\PE[\zeta^2] < \infty$, such that the variance of the r.v.
-$f(X) - \zeta$ is small.
-The complexity of the problem of constructing classes $Z$ of control variates \(\zeta\) satisfying   $\mathsf{E}[\zeta]=0$ essentially depends on the degree of our knowledge on \(\pi.\)
+$f(X) + \zeta$ is small.
+The complexity of the problem of constructing classes $Z$ of control variates \(\zeta\) satisfying   $\PE[\zeta]=0$ essentially depends on the degree of our knowledge on \(\pi.\)
 For example, if \(\pi\) is analytically known and satisfies some regularity conditions, one can apply the well-known technique of  polynomial interpolation to construct control variates enjoying  some optimality properties, see, for example, Section~3.2 in \cite{dimov2008monte}.
@@
 Note that in order to compare different  variance reduction approaches, one has to analyze their complexity, that is, the number of numerical operations required to achieve a prescribed magnitude of the resulting variance.
-
-The situation becomes much more difficult in the case of MCMC algorithms, where one  has to work with a
-Markov chain \((X_p),\) \(p=0,1,2,\ldots,\) whose marginal distribution  converges  to \(\pi\) as time grows.
-One important class of the variance reduction methods in this case  is based on the Poisson equation for the corresponding  Markov chain.
-It was observed in~\cite{henderson1997variance}  that if a time-homogeneous Markov chain \((X_p)\) is stationary with stationary distribution \(\pi,\) then for any real-valued function \(G \in L^1(\pi) \) defined on the state space of the Markov chain \((X_p),\)  the function \(U(x) := G(x)-\mathsf{E}[G(X_{1})|X_0 = x]\) has zero mean with respect to \(\pi\).
-The best choice for the function \(G\) corresponds to a solution of the Poisson equation  \(\mathsf{E}[G(X_{1})|X_0 = x]-G(x)=-f(x)+\pi(f)\).
-Moreover, it is also related to the minimal asymptotic variance in the corresponding central limit theorem, see \cite{duncan2016variance} and \cite{mira2013zero}.
-Although the Poisson equation involves the quantity of interest \(\pi(f)\)  and can not be  solved explicitly in most cases, this idea still can be used to construct some  approximations for the optimal zero-variance control variates.
+\par
+Unfortunately it is not always possible to generate independent observations distributed according to $\pi$.
+To overcome this issue one might consider MCMC algorithms, where the exact samples from $\pi$ are replaced by \((X_p),\) \(p=0,1,2,\ldots,\) forming a Markov chain with a marginal distribution of $X_n$ converging to $\pi$ in a suitable metrics as $n$ grows.
+It is still possible to apply the control variates method in the similar manner to a simple Monte-Carlo case, yet the choice of the optimal control variate becomes much more difficult.
+Due to significant correlations between the elements of the Markov chain it might be not enough to minimize the marginal variances of \((X_p)_{p \geq 0}\) as it was in independent case.
+Instead one may choose the control variate by minimizing the corresponding asymptotic variance of the chain as it is suggested in~\cite{belomestny2019esvm}.
+At the same time it is possible to express the optimal control variate in terms of the solution of the Poisson equation for the corresponding Markov chain \((X_p)\).
+As it was observed in~\cite{henderson1997variance}, for a time-homogeneous Markov chain \(X_p\) with a stationary distribution \(\pi\), the function $U_{G}(x) := G(x)-\PE[G(X_{1})|X_0 = x]$ has a zero mean with respect to $\pi$ for arbitrary real-valued function \(G: \mathcal{X} \to \rset\), such that \(G  \in L^1(\pi)\).
+Hence, $U_{G}(x)$ is a valid control functional for a suitable choice of $G$, with the best $G$ given by a solution of the Poisson equation \(\PE [G(X_{1})|X_0 = x]-G(x)=-f(x)+\pi(f)\).
+For such $G$ we obtain $f(x) + U_G(x) = f(x) - f(x) + \pi(f) = \pi(f)$ leading to an ideal estimator with zero variance.
+Despite the fact that the Poisson equation involves the quantity of interest \(\pi(f)\)  and can not be  solved explicitly in most cases, this idea still can be used to construct some  approximations for the optimal zero-variance control variates.
 For example,  \cite{henderson1997variance} proposed to compute approximations for the solution of the Poisson equation for specific Markov chains with particular emphasis on models arising in stochastic network theory.
@@
 Then we estimate the cost-variance ratio
-\begin{eqnarray*}
-\frac{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]}{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}
-\end{eqnarray*}
+\begin{eqnarray}
+\label{eq:cvR}
+\mathcal{R}(K,N,n,n_{\mathrm{trunc}})=\frac{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n,n_{\mathrm{trunc}} }^N(f)|\TrainSet]} 
+\end{eqnarray}
 by its empirical counterpart, computed over $24$ independent trajectories with $n = 2\times 10^3$ and $N = 2 \times 10^3$.
-For different $n_{\mathrm{trunc}}$ and $K$ we estimate $\mathrm{cost}(\widehat \pi_{K,n}^N)$ as
- \begin{equation}
- \label{eq:cvR}
- %check this
- %add n_{truncation} here
- \mathrm{cost}(\widehat \pi_{K,n}^N) = \mathrm{cost}(\pi_{n}^N) \times n_{\mathrm{trunc}} \times d^{K}
- \end{equation}
- since the representations of $a_{r,\mathbf{k}} (x)$ might be pre-computed using formulas \eqref{eq:a_hat}.
+Since for fixed $K$ the cost of computing $\pi_{n}^N(f)$ is proportional to computing function $f$, we set for $K = 1$
+ \begin{equation*}
+ \mathrm{cost}(\widehat \pi_{K,n}^N) = \mathrm{cost}(\pi_{n}^N) \times n_{\mathrm{trunc}} \times 2
+ \end{equation*}
+ since there are only $2$ non-zero functions among $a_{r,\mathbf{k}}(x)$ for fixed $r$.
+Note that in this case each $a_{r,\mathbf{k}}(x)$  is a linear function, which can be computed at the same cost as $f$.
+Similarly, for $K = 2$ we set
+  \begin{equation*}
+ \mathrm{cost}(\widehat \pi_{K,n}^N) = \mathrm{cost}(\pi_{n}^N) \times n_{\mathrm{trunc}} \times 8
+ \end{equation*}
+since there are at most $8$ non-zero functions among $a_{r,\mathbf{k}}(x)$ for fixed $r$.
 Variance reduction costs for Gaussian potential and different $n_{\mathrm{trunc}}$  are summarized in Figure~\ref{fig:gaus_cost}, and for the Gaussian mixture - in Figure~\ref{fig:gmm_cost}.

### 2020-04-13 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 \begin{frontmatter}
-\title{Variance reduction for MCMC methods via martingale representations}
-\runtitle{Variance reduction for MCMC}
+\title{Variance reduction for Markov chains via martingale representations}
+\runtitle{Variance reduction for Markov chains}
 %\thankstext{T1}{This research was }
@@
 \begin{abstract}
-In this paper we propose an efficient variance reduction approach for  MCMC algorithms relying on a novel discrete time martingale representation for Markov chains.
-Our approach is fully non-asymptotic and does not require any type of ergodicity or special product structure of the underlying density.
-By rigorously analyzing the  convergence properties of the proposed algorithm, we show that it's complexity is indeed asymptotically smaller than one of the original MCMC algorithm.
-The numerical performance of the new method is illustrated  in the case of Gaussian mixtures and Bayesian regression models.
+In this paper we propose an efficient variance reduction approach for  additive functionals of Markov chains relying on a novel discrete time martingale representation.
+Our approach is fully non-asymptotic and does not require the knowledge of the stationary distribution (and even any type of ergodicity) or specific structure of the underlying density.
+By rigorously analyzing the  convergence properties of the proposed algorithm, we show that its complexity is indeed asymptotically smaller than one of the naive algorithm.
+The numerical performance of the new method is illustrated for Markov Chain Monte Carlo (MCMC) methods.
 \end{abstract}
@@
 
-Monte Carlo and Markov Chain Monte Carlo algorithms play crucial role in modern numerical analysis, finding various applications in such research areas as Bayesian statistics.
-Suppose that we aim at computing $\pi(f):= \PE\left[f(X)\right]$, where $X$ is a random vector in $\mathcal{X} \subseteq \rset^d$ admitting a density $\pi$ and $f$ is a square-integrable function $f:\mathcal{X} \to \rset$, $f \in L^2(\pi)$.
+Markov chains and Markov Chain Monte Carlo algorithms play crucial role in modern numerical analysis, finding various applications in such research areas as Bayesian inference, reinforcement learning, online learning.
+As an illustration, suppose that we aim at computing $\pi(f):= \PE\left[f(X)\right]$, where $X$ is a random vector in $\mathcal{X} \subseteq \rset^d$ admitting a density $\pi$ and $f$ is a square-integrable function $f:\mathcal{X} \to \rset$, $f \in L^2(\pi)$.
 Typically it is not possible to compute $\pi(f)$ analytically, leading to the Monte Carlo methods as a popular solution.
-Given an independent identically distributed observations $X_1,\ldots,X_n$ form $\pi$, we might estimate $\pi(f)$ by $\hat{\pi}(f):=1/n\sum_{k=1}^{n}f(X_k)$.
+Given an independent identically distributed observations $X_1,\ldots,X_n$ from $\pi$, we might estimate $\pi(f)$ by $\hat{\pi}(f):=1/n\sum_{k=1}^{n}f(X_k)$.
 The variance of such estimate equals $\sigma^2(f)/n$  with $\sigma^2(f)$ being the variance of the integrand.
@@
 Introduction to many of them can be found in \cite{rubinstein2016simulation}, \cite{GobetBook} and \cite{glasserman2013monte}.
-Recently one witnessed a revival of interest in efficient variance reduction methods  for MCMC algorithms, see for example \cite{dellaportas2012control}, \cite{mira2013zero}, \cite{brosse2018diffusion} and references therein.
+Recently one witnessed a revival of interest in efficient variance reduction methods  for Markov chains, mostly with applications to MCMC algorithms; see for example \cite{dellaportas2012control}, \cite{mira2013zero}, \cite{oates:girolami:chopin:2016}, \cite{south:mira:drovandi:2018}, \cite{brosse2018diffusion} and references therein.
 \par
-One of the popular approaches to variance reduction is the so called control variates method.
+One of the popular approaches to variance reduction is the control variates method.
 We aim at constructing cheaply computable random variable $\zeta$ (control variate) with $\PE[\zeta] = 0$ and $\PE[\zeta^2] < \infty$, such that the variance of the r.v.
@@
 The complexity of the problem of constructing classes $Z$ of control variates \(\zeta\) satisfying   $\PE[\zeta]=0$ essentially depends on the degree of our knowledge on \(\pi.\)
-For example, if \(\pi\) is analytically known and satisfies some regularity conditions, one can apply the well-known technique of  polynomial interpolation to construct control variates enjoying  some optimality properties, see, for example, Section~3.2 in \cite{dimov2008monte}.
+For example, if \(\pi\) is analytically known and satisfies some regularity conditions, one can apply the well-known technique of  polynomial interpolation to construct control variates enjoying  some optimality properties, see, for example \cite[Section~3.2]{dimov2008monte}.
 Alternatively, if an orthonormal system in \(L^2(\pi)\) is analytically available, one can build control variates \(\zeta\) as a linear combination of the corresponding basis functions, see \cite{GobetCV}.
-Furthermore, if \(\pi\) is known only up to a normalizing constant (which is often the case in Bayesian statistics), one can apply the recent approach of control variates  depending only on the gradient \(\nabla \log \pi\)  using Schr\(\ddot{\text{o}}\)dinger-type Hamiltonian operator in  \cite{mira2013zero} and Stein operator in \cite{brosse2018diffusion}.
+Furthermore, if \(\pi\) is known only up to a normalizing constant (which is often the case in Bayesian statistics), one can apply the recent approach of control variates  depending only on the gradient \(\nabla \log \pi\)  using Schr\(\ddot{\text{o}}\)dinger-type Hamiltonian operator in  \cite{assaraf1999zero}, \cite{mira2013zero}, and Stein operator in \cite{brosse2018diffusion}.
 In some situations \(\pi\) is not known analytically, but \(X\) can be represented as a function of  simple random variables with known distribution.
@@
 At the same time it is possible to express the optimal control variate in terms of the solution of the Poisson equation for the corresponding Markov chain \((X_p)\).
-As it was observed in~\cite{henderson1997variance}, for a time-homogeneous Markov chain \(X_p\) with a stationary distribution \(\pi\), the function $U_{G}(x) := G(x)-\PE[G(X_{1})|X_0 = x]$ has a zero mean with respect to $\pi$ for arbitrary real-valued function \(G: \mathcal{X} \to \rset\), such that \(G  \in L^1(\pi)\).
+As it was observed in~\cite{henderson1997variance,henderson2004}, for a time-homogeneous Markov chain \(X_p\) with a stationary distribution \(\pi\), the function $U_{G}(x) := G(x)-\PE[G(X_{1})|X_0 = x]$ has a zero mean with respect to $\pi$ for arbitrary real-valued function \(G: \mathcal{X} \to \rset\), such that \(G  \in L^1(\pi)\).
 Hence, $U_{G}(x)$ is a valid control functional for a suitable choice of $G$, with the best $G$ given by a solution of the Poisson equation \(\PE [G(X_{1})|X_0 = x]-G(x)=-f(x)+\pi(f)\).
@@
 We use the notations $\nset=\{1,2,\ldots\}$ and $\nset_0=\mathbb N\cup\{0\}$.
-We denote $\boldsymbol{\varphi}(z)=(2\pi)^{-d/2} \exp\{-|z|^2/2\}$, $z\in\mathbb R^d$ probability density function of the $d-$dimensional standard normal distribution.
+We denote $\boldsymbol{\varphi}(z)=(2\pi)^{-d/2} \exp(-|z|^2/2)$, $z\in\mathbb R^d$ probability density function of the $d-$dimensional standard normal distribution.
 For $x \in \rset^d$ and $r>0$ let $B_r(x) = \{y \in \rset^d | \|y - x\| < r\}$ where $\|\cdot\|$ is a standard Euclidean norm.
@@
 For a bounded Borel function $f: \rset^d \rightarrow \rset$ denote $\mathrm{osc}(f):=\sup_{x\in\mathbb R^d}f(x)-\inf_{x\in\mathbb R^d}f(x)$.
-Given a function $V: \rset^d \rightarrow \rset$, for a function $f: \rset^d \rightarrow \rset$ we define $\|f\|_{V} = \sup_{x \in \rset^d}\frac{|f(x)|}{V(x)}$ and the corresponding $V-$norm between probability measures $\mu$ and $\nu$ on $\mathcal{B}(\rset^d)$ as
-\[
-d_V(\mu, \nu) = \|\mu - \nu\|_{V} = \sup_{\|f\|_{V} \leq 1} \left[\int\limits_{\rset^d}f(x)\,d\mu(x) - \int\limits_{\rset^d}f(x)\,d\nu(x)\right]  \eqsp.
+Given a function $V: \rset^d \rightarrow \rset$, for a function $f: \rset^d \rightarrow \rset$ we define $\|f\|_{V} = \sup_{x \in \rset^d} |f(x)|/V(x)$ and the corresponding $V-$norm between probability measures $\mu$ and $\nu$ on $\mathcal{B}(\rset^d)$ as
+\[
+d_V(\mu, \nu) = \|\mu - \nu\|_{V} = \sup_{\|f\|_{V} \leq 1} \left[\int\nolimits_{\rset^d}f(x)\,d\mu(x) - \int\nolimits_{\rset^d}f(x)\,d\nu(x)\right]  \eqsp.
 \]
@@
 functions $\Phi_{p}\colon\mathcal{X}\times\mathbb{R}^{m}\to\mathcal{X}.$
-In fact, this is quite general class of Markov chains (see Theorem~1.3.6 in ~\cite{moulines2018})
+In fact, this is quite general class of Markov chains (see \cite[Theorem~1.3.6]{moulines2018})
 and many well-known MCMC algorithms can be represented in the form \eqref{eq:chain_gen}.
@@
 \begin{equation}\label{eq:stationary_distr}
-\pi(x)= \mathrm{const}\,\rme^{-\frac{U(x)}{2}}, \quad \mathrm{const}= \left. 1 \middle /{\int_{\mathbb{R}^{d}} \rme^{-\frac{U(x)}{2}}\, \rmd x,} \right.
-\end{equation}
-provided that \(\int_{\mathbb{R}^{d}} \rme^{-\frac{U(x)}{2}}\,\rmd x\) is finite. This method is usually referred to as
-Unadjusted Langevin Algorithm (ULA).
-In fact the Markov chain~\eqref{eq:chain}
+\pi(x)= Z^{-1} \rme^{-U(x)/2}, \quad Z= \int_{\mathbb{R}^{d}} \rme^{-U(x)/2}\, \rmd x,  
+\end{equation}
+provided that \(Z < \infty\). 
+This method is usually referred to as Unadjusted Langevin Algorithm (ULA).
+Denote by \(W\) the standard \(\mathbb{R}^m\)-valued Brownian motion.
+The Markov chain~\eqref{eq:chain}
 arises as the Euler-Maruyama discretization
@@
 \begin{eqnarray*}
-q_p(y|x)=(\gamma_{p+1})^{-d/2}\boldsymbol{\varphi}\Bigl([y-x+\gamma_{p+1}\mu(x)]/\sqrt{\gamma_{p+1}}\Bigr),
-\end{eqnarray*}
-where
-$\boldsymbol{\varphi}(z)=(2\pi)^{-d/2} \exp\{-|z|^2/2\}$,
-$z\in\mathbb R^d$, denotes the density of a $d$-dimensional
-standard Gaussian random vector.
-The MALA algorithms usually provide noticeable speed-ups in convergence for most problems.
-It is not difficult to see that the MALA chain can be compactly represented in the form
+q_p(y|x)=(\gamma_{p+1})^{-d/2}\boldsymbol{\varphi}\Bigl([y-x+\gamma_{p+1}\mu(x)]/\sqrt{\gamma_{p+1}}\Bigr) \,.
+\end{eqnarray*}
+The MALA algorithms usually provide noticeable speed-ups in convergence for most problems. It is not difficult to see that the MALA chain can be compactly represented in the form
 \begin{align*}
@@
 \end{eqnarray}
-where \(W\) is a standard \(\mathbb{R}^m\)-valued Brownian motion, \(b:\) \(\mathbb{R}^d\to \mathbb{R}^d\) and \(\sigma:\) \(\mathbb{R}^d\times \mathbb{R}^m \to \mathbb{R}^d\) are locally Lipschitz continuous functions with at most linear growth.
+where \(b:\) \(\mathbb{R}^d\to \mathbb{R}^d\) and \(\sigma:\) \(\mathbb{R}^d\times \mathbb{R}^m \to \mathbb{R}^d\) are locally Lipschitz continuous functions with at most linear growth.
 The process \((X_t)_{t\geq 0}\) is a Markov
@@
 \end{eqnarray*}
-where for all $y \in \rset^d$
-\begin{eqnarray*}
-\bar a_{r,k}(y)=\mathsf{E}\left[f(X^y_{r})\phi_k\left(\xi_{1}\right)\right], \quad r, k\in \mathbb{N}.
-\end{eqnarray*}
+where for all $y \in \rset^d$, 
+$\bar a_{r,k}(y)=\mathsf{E}\left[f(X^y_{r})\phi_k\left(\xi_{1}\right)\right]$, $r, k\in \mathbb{N}$.
 \end{cor}

### 2020-04-13 aap/Revision/vr-mcmc-revision-new-new.tex
update vr-mcmc-revision-new-new.tex

@@
 \begin{equation}\label{eq:29032018a2}
-\pi_{n}^{N}(f)=\frac{1}{n}\sum_{p=N+1}^{N+n} f(X^x_{p}),
+\pi_{n}^{N}(f)= \frac{1}{n} \sum_{p=N+1}^{N+n} f(X^x_{p}),
 \end{equation}
@@
 \\
-&=& \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\sum_{p=0}^{N+n-l}\bar{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l}).
+&=& \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}
+A_{N+n-1,k}(X_{l-1}^{x}) \phi_{k}(\xi_{l}),
 \label{eq:29032018a5}
 \end{eqnarray}
-Since \(X^x_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0\), we obtain
-\[
-\PE[g(X^x_{l-1})\phi_{k}\left(\xi_{l}\right)] = \mathsf{E}[g(X^x_{l-1})\CPE{\phi_{k}\left(\xi_{l}\right)}{\mathcal{G}_{l-1}]} = 0
-\]
-for any  function $g.$  Hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
-The representation \eqref{eq:coeff_mart} suggests also that the variance of  the   estimator
-\begin{equation}
-\label{eq:29032018a3}
-\pi_{K,n}^N(f)=\pi_n^N(f)-M_{K,n}^N(f)
-\end{equation}
-should be small for \(K\) large enough. Indeed, since $\PE[ \phi_k(\xi_l) \phi_{k'}(\xi_l)]=0$ if $k \ne k'$, we obtain
-\begin{equation}
-\label{eq:expression-variance}
-\PVar[\pi_{K,n}^N(f)]=\frac{1}{n^2}\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})],
-\end{equation}
 where
@@
 \end{eqnarray}
+Since \(X^x_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0\), we obtain
+\[
+\PE[g(X^x_{l-1})\phi_{k}\left(\xi_{l}\right)] = \mathsf{E}[g(X^x_{l-1})\CPE{\phi_{k}\left(\xi_{l}\right)}{\mathcal{G}_{l-1}]} = 0
+\]
+for any  function $g.$  Hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
+The representation \eqref{eq:coeff_mart} suggests also that the variance of  the   estimator
+\begin{equation}
+\label{eq:29032018a3}
+\pi_{K,n}^N(f)=\pi_n^N(f)-M_{K,n}^N(f)
+\end{equation}
+should be small for \(K\) large enough. Indeed, since $\PE[ \phi_k(\xi_l) \phi_{k'}(\xi_l)]=0$ if $k \ne k'$, we obtain
+\begin{equation}
+\label{eq:expression-variance}
+\PVar[\pi_{K,n}^N(f)]=\frac{1}{n^2}\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})],
+\end{equation}
 Hence $\mathsf{Var}[\pi_{K,n}^N(f)]$ is small provided that the coefficients $A_{s,k}$ decay fast enough as $k\to \infty.$
@@
 with $X^{(s)}_{0}=x,$ $s=1,\ldots,T.$
-Then we solve   the least squares optimization problems
+Then we solve  for $k=1,\ldots,K,$  the least squares optimization problems
 \begin{equation}\label{eq:06042018a1}
-\widetilde{A}_{q,k} \in \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \left| \sum_{r = 0}^{q}  f(X_{r+l_0}^{(s)})\phi_k(\xi^{(s)}_{l_0+1}) - \psi(X^{(s)}_{l_0})\right|^2
-\end{equation}
-for $k=1,\ldots,K,$ and some $l_0$ with $l_0+q\leq N+n,$
-where \(\Psi\) is a  class of  real-valued functions on \(\mathbb{R}^d\) and $\xi^{(s)}_{1},\ldots,\xi^{(s)}_{N+n}$ are iid random variables with distribution $P_\xi$ used to construct the $s$-th trajectory of $X.$   As usual in regression analysis (see Section~11 in \cite{gyorfi2006distribution}),  we  truncate  the least squares estimates $(\widetilde{A}_{q,k})$ and set
+\widetilde{A}_{q,k} \in \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \left| \sum_{r = 0}^{q}  f(X_{r+N}^{(s)})\phi_k(\xi^{(s)}_{N+1}) - \psi(X^{(s)}_{N})\right|^2
+\end{equation}
+where \(\Psi\) is a  class of  real-valued functions on \(\mathbb{R}^d\) and $\xi^{(s)}_{1},\ldots,\xi^{(s)}_{N+n}$ are i.i.d.
+random variables with distribution $P_\xi$ used to construct the $s$-th trajectory of $X.$   As usual in regression analysis (see \cite[Section~11]{gyorfi2006distribution}),  we  truncate  the least squares estimates $(\widetilde{A}_{q,k})$ and set
 \begin{eqnarray}
@@
 \begin{eqnarray*}
-\widehat M_{K,n}^N(f) := \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\widehat{A}_{N+n-l,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
+\widehat M_{K,n}^N(f) := \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n} \widehat{A}_{N+n-l,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\,.
 \end{eqnarray*}
@@
 \end{multline}
-In the next section we analyze the case of ULA and show that under some rather mild conditions (convexity of the potential $U$ outside a ball around zero)
-\begin{eqnarray*}
-\PE[A^2_{l,k}(X^x_{N+l-1})]\leq e^{\varkappa n\gamma^2}R^2_k ,\quad k=1,2,\ldots, \quad l=0,\ldots,n-1,
+In the next section we analyze the case of ULA algorithm and show that under some rather mild conditions (convexity of the potential $U$ outside a ball around zero)
+\begin{eqnarray*}
+\PE[A^2_{l,k}(X^x_{N+l-1})]\leq \rme^{\varkappa n\gamma^2}R^2_k ,\quad k \in \nset, \quad l=0,\ldots,n-1,
 \end{eqnarray*}
@@
 \[
-\bar A_{q,k}:=\argmin_{\psi\in\Psi}\mathsf{E}[|A_{q,k}(X_{l_0})-\psi(X_{l_0})|]^2
+\bar A_{q,k}:=\argmin_{\psi\in\Psi}\mathsf{E}[|A_{q,k}(X_{N})-\psi(X_{N})|^2]
 \]
@@
 \begin{eqnarray*}
-\PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]\leq \frac{2Ce^{\varkappa n\gamma^2}}{n}.
+\PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]\leq \frac{2C \rme^{\varkappa n\gamma^2}}{n}.
 \end{eqnarray*}
@@
 \label{eq:cost_theoretical}
-\frac{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]}{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}\leq C e^{\varkappa n\gamma^2} K \gamma \, T^2 \mathtt{D}_\Psi.
+\frac{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]}{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}\leq C \rme^{\varkappa n\gamma^2} K \gamma \, T^2 \mathtt{D}_\Psi.
 \end{equation}
@@
 \label{eq:Aqk2}
-A^2_{q,\mathbf{k}}(x)\leq C_Ke^{\varkappa n\gamma^2}(1+V(x)),\quad 1\leq |\mathbf{k}|\leq K,
+A^2_{q,\mathbf{k}}(x)\leq C_K \rme^{\varkappa n\gamma^2}(1+V(x)),\quad 1\leq |\mathbf{k}|\leq K,
 \end{eqnarray}
@@
 \label{eq:Aqk2sum}
-\sum_{\|\mathbf{k}\|\geq K+1} A^2_{q,\mathbf{k}}(x)\leq C_{K}e^{\varkappa n\gamma^2}\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
+\sum_{\|\mathbf{k}\|\geq K+1} A^2_{q,\mathbf{k}}(x)\leq C_{K} \rme^{\varkappa n\gamma^2}\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
 \left(\frac{\gamma}{2}\right)^{|I|K-1}

### 2020-04-13 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 &=& \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}
-A_{N+n-1,k}(X_{l-1}^{x}) \phi_{k}(\xi_{l}),
+A_{N+n-l,k}(X_{l-1}^{x}) \phi_{k}(\xi_{l}),
 \label{eq:29032018a5}
@@
 \end{eqnarray}
-For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
+For an estimator $\rho_n^N(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
 of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we shall be interested in the Mean Squared Error (MSE),
@@
 \label{eq:29032018a4}
-\mathrm{MSE}\left[\rho(f)\right] = \PE\left[\left\{\rho(f)-\pi(f)\right\}^2\right]
-=\left\{\PE[\rho(f)]-\pi(f)\right\}^2 +\PVar[\rho(f)].
+\mathrm{MSE}\left[\rho_n^N(f)\right] = \PE\left[\left\{\rho_n^N(f)-\pi(f)\right\}^2\right]
+=\left\{\PE[\rho_n^N(f)]-\pi(f)\right\}^2 +\PVar[\rho_n^N(f)].
 \end{equation}
@@
 Under assumptions {\bf (H1)} and {\bf (H2)}, the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$.
-From \cite[Theorem~10]{durmus:moulines:2017}
+From \cite[Theorem~10]{durmus:moulines:2017},
 %and Lemma~\ref{lem:v_ergodicity}
-it follows that
+it follows that there exist $\gamma_0 > 0$ and $C < \infty$ such that for all $\gamma \in (0,\gamma_0]$, we get
 \begin{equation}
@@
 \end{equation}
-for some constant $C>0.$ Let us now derive an upper bound for the variance of the classical estimator~\eqref{eq:29032018a2} for ULA-based chain.
-
+ Let us now derive an upper bound for the variance of the classical estimator~\eqref{eq:29032018a2} for ULA-based chain. Define
+ \begin{equation}
+ \label{eq:definition-V}
+ V(x) = 1+ \|x\|^2 \,.
+ \end{equation}
 \begin{lem}
 \label{lem:variance}
-Let $f$ be a bounded Borel function and $(X^x_{p})_{p \in \nset_0}$ be a Markov chain generated by ULA with constant step size $\gamma$, satisfying {\bf (H1)} and {\bf (H2)}.
-Assume also that $n\gamma > c_0$ for some fixed $c_0>0,$ then it holds
-
+Assume {\bf (H1)} and {\bf (H2)}.
+Let $f$ be a bounded Borel function and $(X^x_{p})_{p \in \nset_0}$ be a Markov chain generated by ULA with constant step size $\gamma \in (0,\gamma_0]$ where $\gamma_0= m/L^2$. 
+Let $c_0>0$. Then, for all $n \in \nset$, $\gamma \in (0,\gamma_0]$ satisfying $n\gamma > c_0$   
 \begin{equation}
@@
 \end{equation}
-with $V(x) = 1+ \|x\|^2$ where $\lesssim$ stands for inequality up to a constant not depending on $\gamma$ and $n$.
+where $\lesssim$ stands for inequality up to a constant not depending on $\gamma \in (0,\gamma_0]$ and $n \in \nset$.
 \end{lem}
@@
 \end{equation}
- Set $V(x) = 1 + \|x\|^2,$ then  it holds for $q=0,\ldots,n-1,$ and all $x$
+Then, it holds that for $q=0,\ldots,n-1,$ and all $x$
 \begin{eqnarray}

### 2020-04-13 aap/Revision/vr-mcmc-revision-new-new.tex
update vr-mcmc-revision-new-new.tex

@@
 \newtheorem{prop}[thm]{Proposition}
-\newtheorem{remark}{Remark}
-\newtheorem{cor}{Corollary}
-\newtheorem{example}{Example}
+\newtheorem{remark}[thm]{Remark}
+\newtheorem{cor}[thm]{Corollary}
+\newtheorem{example}[thm]{Example}
 \newenvironment{proof}{{\noindent \bf Proof }}
@@
 Assume {\bf (H1)} and {\bf (H2)}.
-Let $f$ be a bounded Borel function and $(X^x_{p})_{p \in \nset_0}$ be a Markov chain generated by ULA with constant step size $\gamma \in (0,\gamma_0]$ where $\gamma_0= m/L^2$. 
-Let $c_0>0$. Then, for all $n \in \nset$, $\gamma \in (0,\gamma_0]$ satisfying $n\gamma > c_0$   
+Let $f$ be a bounded Borel function and $(X^x_{p})_{p \in \nset_0}$ be a Markov chain generated by ULA with constant step size $\gamma$. 
+Let $c_0>0$. Then, there exist $C < \infty$ and $\gamma_0 > 0$, such that for all $n \in \nset$, $\gamma \in (0,\gamma_0]$ satisfying $n\gamma > c_0$, and $x \in \rset^d$, 
 \begin{equation}
 \label{eq:var-mc}
-\PVar[\pi_n^N(f)]\lesssim \frac{V(x)}{n\gamma} \eqsp,
-\end{equation}
-where $\lesssim$ stands for inequality up to a constant not depending on $\gamma \in (0,\gamma_0]$ and $n \in \nset$.
+\PVar\left[\pi_n^N(f)\right]\leq C (n \gamma)^{-1} V(x) \eqsp.
+\end{equation}
 \end{lem}
@@
 \label{rem:var-low}
-The bound in Lemma~\ref{lem:variance} is sharp and can not be improved even in the case of a normal distribution $\pi.$
+The bound in Lemma~\ref{lem:variance} is sharp and cannot be improved even in the case of a normal distribution $\pi.$
 Indeed,  ULA for the standard normal distribution takes form
@@
 \end{equation}
-Then, it holds that for $q=0,\ldots,n-1,$ and all $x$
+Then, there exist $\gamma_0 > 0$, $C_K < \infty$, and $\varkappa > 0$  such that, for all $\gamma \in (0,\gamma_0]$, $n \in \nset$, and $x \in \rset^d$,
 \begin{eqnarray}
 \label{eq:Aqk2}
-A^2_{q,\mathbf{k}}(x)\leq C_K \rme^{\varkappa n\gamma^2}(1+V(x)),\quad 1\leq |\mathbf{k}|\leq K,
+A^2_{q,\mathbf{k}}(x)\leq C_K \rme^{\varkappa n\gamma^2}\{1+V(x)\} \,,\quad 1\leq |\mathbf{k}|\leq K, \quad \text{for all  $q=0,\ldots,n-1,$}
 \end{eqnarray}
@@
 \sum_{\|\mathbf{k}\|\geq K+1} A^2_{q,\mathbf{k}}(x)\leq C_{K} \rme^{\varkappa n\gamma^2}\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
-\left(\frac{\gamma}{2}\right)^{|I|K-1}
-(1+V(x))
+\left(\frac{\gamma}{2}\right)^{|I|K-1} \, \{1+V(x)\} \eqsp, 
 \end{eqnarray}
-with some positive constants $\varkappa, C_K$ not depending on $n$ and  $\gamma.$  The sum in \eqref{eq:Aqk2} runs over all nonempty subsets $I$ of the set $\{1,\ldots,d\}$
+where the sum in \eqref{eq:Aqk2} runs over all nonempty subsets $I$ of the set $\{1,\ldots,d\}$.
 \end{thm}
 \begin{cor}
-Under assumptions of Theorem~\ref{th:mr},
+Under assumptions of Theorem~\ref{th:mr}, there exist $\gamma_0 > 0$, $\tilde{C} < \infty$ and $\varkappa > 0$ such that for all $n \in \nset$, $\gamma \in (0,\gamma_0]$, $n\in \nset$, and $x \in \rset^d$,
 \begin{eqnarray}
 \label{eq:var-bound}
-\PVar\left(\pi_{K,n}^{N}(f)\right)\lesssim n^{-1} \gamma^{K-1}  e^{\varkappa n\gamma^2}
-\eqsp,
+\PVar\left[\pi_{K,n}^{N}(f)\right] \leq \tilde{C}  n^{-1} \gamma^{K-1}  \rme^{\varkappa n\gamma^2}
+\eqsp.
 \end{eqnarray}
-where $\lesssim$ stands for inequality up to a constant not depending  on $n,$ $N$ and $\gamma$.
 \end{cor}
@@
 \PVar \left(\sum_{p=1}^{q}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X^x_{p}\right)
-\right)\leq C_{K}\gamma^{-1}e^{\varkappa n\gamma^2}(1+V(x)),\quad q=0,\ldots,n-1,
+\right)\leq C_{K}\gamma^{-1}\rme^{\varkappa n\gamma^2}(1+V(x)),\quad q=0,\ldots,n-1,
 \end{equation}
@@
 \PVar\left(\sum_{p=1}^{q}\partial_{1}^{\mathbf{K}_I}f\left(X^x_{p}\right)
-\right)\leq C\gamma^{-1}e^{\varkappa n\gamma^2}(1+V(x)),\quad q=1,\ldots,n,
+\right)\leq C\gamma^{-1} \rme^{\varkappa n\gamma^2}(1+V(x)),\quad q=1,\ldots,n,
 \end{eqnarray*}
@@
 $$
-\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right]\le  \gamma^{-1} C e^{\varkappa n\gamma^2}(1+V(x)), \quad q=1,\ldots,n,
+\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right]\le  \gamma^{-1} C \rme^{\varkappa n\gamma^2}(1+V(x)), \quad q=1,\ldots,n,
 $$
@@
 \[
-\zeta_{l,j}(u)=\mathsf{E}\left[e^{-u\gamma\sum_{k=l}^{j}\mu'(X^x_{k-1})}\right],\quad u>0.
+\zeta_{l,j}(u)=\mathsf{E}\left[\rme^{-u\gamma\sum_{k=l}^{j}\mu'(X^x_{k-1})}\right],\quad u>0.
 \]
@@
 \[
-\mathsf{E}\left[e^{-u\gamma\left(\sum_{k=l}^{j}[\mu'(X^x_{k-1})-\pi_\gamma(\mu')]\right)}\right]\leq C_1e^{\varkappa n\gamma^2}(1+V(x)),\quad u\gamma\leq \gamma_0,
+\mathsf{E}\left[\rme^{-u\gamma\left(\sum_{k=l}^{j}[\mu'(X^x_{k-1})-\pi_\gamma(\mu')]\right)}\right]\leq C_1 \rme^{\varkappa n\gamma^2}(1+V(x)),\quad u\gamma\leq \gamma_0,
 \]
@@
 \[
-\zeta_{l,j}(u)=\mathsf{E}\left[e^{-s\gamma\sum_{k=l}^{j}\mu'(X^x_{k-1})}\right]\leq  C_1 e^{\varkappa n\gamma^2}(1+V(x))e^{-s\gamma (j-l+1)\pi_\gamma(\mu')}.
+\zeta_{l,j}(u)=\mathsf{E}\left[\rme^{-s\gamma\sum_{k=l}^{j}\mu'(X^x_{k-1})}\right]\leq  C_1 \rme^{\varkappa n\gamma^2}(1+V(x)) \rme^{-s\gamma (j-l+1)\pi_\gamma(\mu')}.
 \]
@@
 \[
-\zeta_{l,j}(s) \leq  C_1e^{\varkappa n\gamma^2}(1+V(x))e^{-s\gamma(j-l+1)\alpha}
+\zeta_{l,j}(s) \leq  C_1 \rme^{\varkappa n\gamma^2}(1+V(x)) \rme^{-s\gamma(j-l+1)\alpha}
 \]
@@
 \begin{align*}
-\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4) &\leq (C_1e^{\varkappa n\gamma^2}(1+V(x)))^{1/4}\sum_{p=j+1}^{q+1}e^{-\gamma(p-j)\alpha}
-\\
-&\leq  (C_1e^{\varkappa n\gamma^2}(1+V(x)))^{1/4}\frac{1}{1-e^{- \gamma\alpha}}
+\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4) &\leq (C_1 \rme^{\varkappa n\gamma^2}(1+V(x)))^{1/4}\sum_{p=j+1}^{q+1} \rme^{-\gamma(p-j)\alpha}
+\\
+&\leq  (C_1 \rme^{\varkappa n\gamma^2}(1+V(x)))^{1/4}\frac{1}{1- \rme^{- \gamma\alpha}}
 \end{align*}
@@
 \begin{align*}
-\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \leq (C_1e^{\varkappa n\gamma^2}(1+V(x)))^{1/2}\sum_{j=0}^{q}e^{-\gamma j\alpha}
-\\
-\leq  (C_1e^{\varkappa n\gamma^2}(1+V(x)))^{1/2}\frac{ 1}{1-e^{- \gamma\alpha}}.
+\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \leq (C_1 \rme^{\varkappa n\gamma^2}(1+V(x)))^{1/2}\sum_{j=0}^{q} \rme^{-\gamma j\alpha}
+\\
+\leq  (C_1 \rme^{\varkappa n\gamma^2}(1+V(x)))^{1/2}\frac{ 1}{1- \rme^{- \gamma\alpha}}.
 \end{align*}
@@
 \begin{align*}
-\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X^x_{p}\right)\right] \leq \gamma^{-1} C_K e^{\varkappa n\gamma^2} (1+V(x))
+\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X^x_{p}\right)\right] \leq \gamma^{-1} C_K \rme^{\varkappa n\gamma^2} (1+V(x))
 \end{align*}
@@
 \label{lem:drift}
- Assume that the potential $U(x), x \in \mathbb{R}^d$ satisfies conditions {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$.
+ Assume that {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$. 
  Then the kernel $Q_\gamma$ from \eqref{eq:ula_kernel} satisfies drift condition \eqref{eq:lyapunov} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L}$ with drift function $V(x) = 1 + \|x\|^2$,   $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = \frac{25K_2^2}{8} + d + m$ with $K_2$ from \eqref{eq:H3}, $m$ from {\bf (H2)}.
@@
 \label{lem:v_ergodicity}
- Assume that the potential $U(x), x \in \mathbb{R}^d$ satisfies conditions {\bf (H1)} and {\bf (H2)}. Then for $0 < \gamma < \overline{\gamma}=\frac{m}{4L^2}$, for any $x \in X$ it holds
+ Assume  {\bf (H1)} and {\bf (H2)}. Then for $0 < \gamma < \overline{\gamma}=m / 4L^2$, for any $x \in \rset^d$ it holds
 $$

### 2020-04-13 aap/Revision/vr-mcmc-revision-new-new.tex
update vr-mcmc-revision-new-new.tex

@@
 where for all $y \in \rset^d$, 
-$\bar a_{r,k}(y)=\mathsf{E}\left[f(X^y_{r})\phi_k\left(\xi_{1}\right)\right]$, $r, k\in \mathbb{N}$.
+\begin{equation}
+\label{eq:definition-bar-a-0}
+\bar a_{r,k}(y)=\mathsf{E}\left[f(X^y_{r})\phi_k\left(\xi_{1}\right)\right] \,, \text{for $r, k\in \mathbb{N}$.}
+\end{equation}
 \end{cor}
@@
 \begin{prop}
-Let $q\geq l, k\in \nset$. Then the coefficients \(a_{q,l,k}(x)\) in \eqref{eq:coeff_mart}  can be alternatively represented as
+Let $q\geq l, k\in \nset$. Then the coefficients \(a_{q,l,k}\) in \eqref{eq:coeff_mart}  can be alternatively represented as
 \begin{eqnarray*}
@@
 with \(Q_{q,l}(x)=\mathsf{E}\left[f(X^x_{l,q})\right],\) \(q\geq l.\)
-In the case $\Phi_l=\Phi$ in \eqref{eq:ula_new} for all $l\geq 1,$ we have
+In the homogeneous case $\Phi_l=\Phi$, the coefficients $(\bar{a}_{r,k})$ in \eqref{eq:definition-bar-a-0} are given respectively  for all $l\geq 1,$ by
 \begin{equation}

### 2020-04-14 aap/Revision/vr-mcmc-revision-new-new.tex
update vr-mcmc-revision-new-new.tex

@@
 \section{Introduction}
-
 Markov chains and Markov Chain Monte Carlo algorithms play crucial role in modern numerical analysis, finding various applications in such research areas as Bayesian inference, reinforcement learning, online learning.
@@
 \end{cor}
+Another equivalent representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful to construct estimators:
+\begin{prop}
+Let $q\geq l, k\in \nset$. Then the coefficients \(a_{q,l,k}\) in \eqref{eq:coeff_mart}  can be alternatively represented as
+\begin{eqnarray*}
+a_{q,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{q,l}\left(\Phi_l(x,\xi)\right)\right]
+\end{eqnarray*}
+with \(Q_{q,l}(x)=\mathsf{E}\left[f(X^x_{l,q})\right],\) \(q\geq l.\)
+In the homogeneous case $\Phi_l=\Phi$, the coefficients $\bar{a}_{r,k}$ in \eqref{eq:definition-bar-a-0} are given respectively  for all $l\geq 1,$ by
+\begin{equation}
+\label{eq:definition-bar-a}
+\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right] \,
+\quad \text{with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$}
+\end{equation}
+\end{prop}
+
 \section{Variance reduction}
@@
 \]
-for any  function $g.$  Hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
+for any  function $g \in L^2(\pi).$  Hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
 The representation \eqref{eq:coeff_mart} suggests also that the variance of  the   estimator
@@
 \par
-The coefficients \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. One way to estimate them is to use nonparametric regression.
-We first present a generic regression algorithm and then in Section~\ref{sec:num} give further implementation details. Our algorithm starts with estimating the functions
-$A_{q,k}$ for $q=0,\ldots,n-1,$ $k=1,\ldots,K. $ We first generate \(\NtrainPath\)  paths  of the chain \(X\)
+The coefficients \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. 
+One way to estimate them is to use nonparametric regression.
+%%%%% Old Version
+%We first present a generic regression algorithm and then in Section~\ref{sec:num} give further implementation details. Our algorithm starts with estimating the functions
+%$A_{q,k}$ for $q=0,\ldots,n-1,$ $k=1,\ldots,K. $ We first generate \(\NtrainPath\)  paths  of the chain \(X\)
+%(the so-called ``training paths''):
+%\[
+%\TrainSet=\Bigl\{(X^{(s)}_{1},\ldots,X^{(s)}_{N+n}),\quad s=1,\ldots,\NtrainPath\Bigr\}
+%\]
+%with $X^{(s)}_{0}=x,$ $s=1,\ldots,T.$
+%Then we solve  for $k=1,\ldots,K,$  the least squares optimization problems
+%\begin{equation}\label{eq:06042018a1}
+%\widetilde{A}_{q,k} \in \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \left| \sum_{r = 0}^{q}  f(X_{r+N}^{(s)})\phi_k(\xi^{(s)}_{N+1}) - \psi(X^{(s)}_{N})\right|^2
+%\end{equation}
+%where \(\Psi\) is a  class of  real-valued functions on \(\mathbb{R}^d\) and $\xi^{(s)}_{1},\ldots,\xi^{(s)}_{N+n}$ are i.i.d.
+random variables with distribution $P_\xi$ used to construct the $s$-th trajectory of $X.$   As usual in regression analysis (see \cite[Section~11]{gyorfi2006distribution}),  we  truncate  the least squares estimates $(\widetilde{A}_{q,k})$ and set
+%\begin{eqnarray}
+%\label{eq:trunc}
+%\widehat{A}_{q,k}(y):=
+%\begin{cases}
+%\widetilde{A}_{q,k}(y), & |\widetilde{A}_{q,k}(y)|\leq W_k,
+%\\
+%W_k \,\mathrm{sign}(\widetilde{A}_{q,k}(y)), & \mathrm{otherwise}.
+%\end{cases}
+%\end{eqnarray}
+% for a sequence of positive real numbers $W_k,$ $k=1,\ldots,K.$
+%Upon estimating  the functions \((A_{q,k})\), one can construct the empirical estimate  of \(M_{K,n}^N(f)\) in the form
+%\begin{eqnarray*}
+%\widehat M_{K,n}^N(f) := \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n} \widehat{A}_{N+n-l,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\,.
+%\end{eqnarray*}
+%%%% Old Version
+We present a generic regression algorithm and then in Section~\ref{sec:num} give further implementation details. Our algorithm starts with estimating the functions
+$Q_r$ for $r=0,\ldots,n-1,$, defined in \eqref{eq:definition-bar-a}. 
+We first generate \(\NtrainPath\)  paths  of the chain \(X\)
 (the so-called ``training paths''):
@@
 with $X^{(s)}_{0}=x,$ $s=1,\ldots,T.$
-Then we solve  for $k=1,\ldots,K,$  the least squares optimization problems
+Then we solve  the least squares optimization problems
 \begin{equation}\label{eq:06042018a1}
-\widetilde{A}_{q,k} \in \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \left| \sum_{r = 0}^{q}  f(X_{r+N}^{(s)})\phi_k(\xi^{(s)}_{N+1}) - \psi(X^{(s)}_{N})\right|^2
-\end{equation}
-where \(\Psi\) is a  class of  real-valued functions on \(\mathbb{R}^d\) and $\xi^{(s)}_{1},\ldots,\xi^{(s)}_{N+n}$ are i.i.d.
-random variables with distribution $P_\xi$ used to construct the $s$-th trajectory of $X.$   As usual in regression analysis (see \cite[Section~11]{gyorfi2006distribution}),  we  truncate  the least squares estimates $(\widetilde{A}_{q,k})$ and set
-\begin{eqnarray}
-\label{eq:trunc}
-\widehat{A}_{q,k}(y):=
-\begin{cases}
-\widetilde{A}_{q,k}(y), & |\widetilde{A}_{q,k}(y)|\leq W_k,
-\\
-W_k \,\mathrm{sign}(\widetilde{A}_{q,k}(y)), & \mathrm{otherwise}.
-\end{cases}
-\end{eqnarray}
- for a sequence of positive real numbers $W_k,$ $k=1,\ldots,K.$
-Upon estimating  the functions \((A_{q,k})\), one can construct the empirical estimate  of \(M_{K,n}^N(f)\) in the form
+\hat{Q}_{r} \in \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \left|   f(X_{r+N}^{(s)}) - \psi(X^{(s)}_{N})\right|^2
+\end{equation}
+where \(\Psi\) is a  class of  real-valued functions on \(\mathbb{R}^d\).
+We then set 
+\begin{equation}
+\label{eq:definition-hat-a-A}
+\hat{a}_{r,k}(x)= \int \hat{Q}_r(\Phi(x,z)) \phi_k(z) P_\xi(\rmd z) \quad \text{and}
+\quad 
+\hat{A}_{q,k}(x) = \sum_{r=0}^q \hat{a}_{r,k}(x) \eqsp,
+\end{equation}
+and construct the empirical estimate  of \(M_{K,n}^N(f)\) in the form
 \begin{eqnarray*}
@@
 \end{eqnarray*}
+
 Obviously, \(\mathsf{E}[\widehat M_{K,n}^N(f)|\TrainSet]=0\) and \(\widehat M_{K,n}^N(f)\) is a valid control variate in that it
@@
 In this section we illustrate the performance of the proposed variance reduction method for ULA.
-First we note that from  numerical point of view another equivalent representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
-\begin{prop}
-Let $q\geq l, k\in \nset$. Then the coefficients \(a_{q,l,k}\) in \eqref{eq:coeff_mart}  can be alternatively represented as
-\begin{eqnarray*}
-a_{q,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{q,l}\left(\Phi_l(x,\xi)\right)\right]
-\end{eqnarray*}
-with \(Q_{q,l}(x)=\mathsf{E}\left[f(X^x_{l,q})\right],\) \(q\geq l.\)
-In the homogeneous case $\Phi_l=\Phi$, the coefficients $(\bar{a}_{r,k})$ in \eqref{eq:definition-bar-a-0} are given respectively  for all $l\geq 1,$ by
-\begin{equation}
-\label{eq:definition-bar-a}
-\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
-\end{equation}
-with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$
-\end{prop}
 First we construct a polynomial approximation  for each \(Q_r(x)\)  in the form:
@@
 The coefficients \(\widehat \beta_{\mathbf{s}}\in \mathbb{R}\) are obtained using a modified least-squares criteria based on $T$ independent training trajectories \(\bigl\{\bigl(X_1^{(s)},\ldots,X_{N+n}^{(s)}\bigr)\bigr\}_{s=1}^{T}\).
-More precisely  we define $\widehat{Q}_0(x) = f(x)$ and
+More precisely  we define $\widehat{Q}_0(x) = f(x)$ and for \(1 \leq r \leq n-1,\) we set
 \begin{eqnarray}
 \label{eq:Qregr}
-\widehat{Q}_r = \argmin_{\psi \in \Psi_m} \sum_{s = 1}^{T}  \left| f(X_{l_0+r}^{(s)}) - \psi(X_{l_0}^{(s)})\right|^2
+\widehat{Q}_r = \argmin_{\psi \in \Psi_m} \sum_{s = 1}^{T}  \left| f(X_{N+r}^{(s)}) - \psi(X_{N}^{(s)})\right|^2
 \end{eqnarray}
-for \(1 \leq l_0+r \leq n-1,\) where $\Psi_m$ is a class of polynomials
+ where $\Psi_m$ is a class of polynomials
 \[

### 2020-04-14 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 \end{eqnarray}
- where $\Psi_m$ is a class of polynomials
-\[
- \psi(x) = \sum_{\|\mathbf{s}\|\leq m} \alpha_{\mathbf{s}} x^{\mathbf{s}}, \quad \alpha_{\mathbf{s}}\in \mathbb{R}.
-\]
-Then using the identity
+ where $\Psi_m$ is a class of polynomials $ \psi(x) = \sum_{\|\mathbf{s}\|\leq m} \alpha_{\mathbf{s}} x^{\mathbf{s}}$ with $\alpha_{\mathbf{s}}\in \mathbb{R}$. Then using the identity
 \[
@@
 \]
-we obtain closed-form expression for the estimates $\widehat a_{r,k}(x)$ of functions $\bar a_{r,k}(x)$ in \eqref{eq:definition-bar-a}. Namely, for all $x \in \mathbb{R}^d$,
-\begin{eqnarray}
+we obtain closed-form expression for the estimates $\widehat a_{r,k}(x)$ of functions $\bar a_{r,k}(x)$ in \eqref{eq:definition-bar-a}. Namely, for all $x=(x^1,\dots,x^d) \in \mathbb{R}^d$,
+\begin{align}
 \nonumber
-\widehat a_{r,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(\xi) \widehat{Q}_{r}(x - \gamma \mu(x) + \sqrt{\gamma}\xi)}{\TrainSet}
-\\
-&=&
+\widehat a_{r,\mathbf{k}} (x) &= \int \mathbf{H}_\mathbf{k}(z) \widehat{Q}_{r}(x - \gamma \mu(x) + \sqrt{\gamma}z) P_\xi(\rmd z)
+\\
+&=
 \label{eq:a-pol}
-\sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{k_i,s_i}(x_i - \gamma \mu_i(x)),
-\end{eqnarray}
-where for any integer $k,s$ $P_{k,s}$
-is a one-dimensional polynomial (in \(x\)) of degree at most \(s\) with analytically known coefficients.
+\sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{k_i,s_i}(x^i - \gamma \mu_i(x)),
+\end{align}
+where for any integer $k,s$ $P_{k,s}$ is a one-dimensional polynomial of degree at most \(s\) with analytically known coefficients.
 We estimate $Q_{r}$ only for $r < n_{\mathrm{trunc}}$ where the truncation level \(n_{\mathrm{trunc}}\) may depend on \(d\) and \(\gamma.\) It allows us to use a smaller amount of training trajectories to approximate $Q_{r}(x)$.
@@
  \[
- \pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}}} \left( \rme^{-\frac{\|x-\mu\|^2}{2}} + \rme^{-\frac{\|x+\mu\|^2}{2}}\right)
+ \pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}}} \left( \rme^{-(1/2)\|x-\mu\|^2} + \rme^{-(1/2)\|x+\mu\|^2}\right)
  \]
@@
 \label{eq:gmm}
-\pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}|\Sigma|}} \left( \rme^{-\frac{(x-\mu)^T\Sigma^{-1}(x-\mu)}{2}} + \rme^{-\frac{(x+\mu)^T\Sigma^{-1}(x+\mu)}{2}}\right)
+\pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}|\Sigma|}} \left( \rme^{-(1/2)(x-\mu)^T\Sigma^{-1}(x-\mu)} + \rme^{-(1/2)(x+\mu)^T\Sigma^{-1}(x+\mu)}\right)
 \end{eqnarray}
-where $\rho \in (0,1), \mu \in \rset^d, \Sigma \in \rset^{2 \times 2}$ and $|\Sigma|$ is its determinant. The function $U(x)$ and its gradient are given by
+where $\rho \in (0,1)$, $\mu \in \rset^2$, $\Sigma$ is a symmetric positive definite $2 \times 2$ matrix and $|\Sigma|$ is its determinant. The function $U(x)$ and its gradient are given by
 \[

### 2020-04-14 aap/Revision/vr-mcmc-revision-new-new.tex
update vr-mcmc-revision-new-new.tex

@@
 \subsection{Gaussian mixtures}
- We consider  ULA with $\pi$  given by the mixture of two equally-weighted $2$-dimensional Gaussian distributions of the following form
+ We consider  ULA with $\pi$  given by the mixture of two equally-weighted $d$-dimensional Gaussian distributions of the following form
 \begin{eqnarray}
@@
 \end{eqnarray}
-where $\rho \in (0,1)$, $\mu \in \rset^2$, $\Sigma$ is a symmetric positive definite $2 \times 2$ matrix and $|\Sigma|$ is its determinant. The function $U(x)$ and its gradient are given by
+where  $\mu \in \rset^d$, $\Sigma$ is a positive-definite $d \times d$ matrix and $|\Sigma|$ is its determinant. The function $U(x)$ and its gradient are given by
 \[
@@
 \[
-\nabla U(x) = \Sigma^{-1}(x-\mu) +2\mu^\top\Sigma^{-1}\left(1 + \rme^{2 \mu^\top\Sigma^{-1} x}\right)^{-1},
+\nabla U(x) = \Sigma^{-1}(x-\mu) +2 \left(1 + \rme^{2 \mu^\top\Sigma^{-1} x}\right)^{-1} \Sigma^{-1}\mu ,
 \]
@@
 \subsection{Banana shape distribution}
-The “Banana-shape” distribution, proposed by \cite{haario1999adaptive}, can be obtained from a d-dimensional Gaussian vector with zero mean and covariance $\mathrm{diag}(p,1,\ldots,1)$ by applying transformation $\varphi_b(x) : \rset^d \rightarrow \rset^d$ of the form
-\[
-\varphi(x_1,\ldots,x_n) = (x_1, x_2 + bx^2_1 - pb,x_3,\ldots,x_d)
+The “Banana-shape” distribution, proposed by \cite{haario1999adaptive}, can be obtained from a $d$-dimensional Gaussian vector with zero mean and covariance $\mathrm{diag}(p,1,\ldots,1)$ by applying transformation $\varphi_b(x) : \rset^d \rightarrow \rset^d$ of the form
+\[
+\varphi_b(x_1,\ldots,x_n) = (x_1, x_2 + bx^2_1 - pb,x_3,\ldots,x_d)
 \]
@@
 \[
-U(x_1,\ldots, x_d) = \frac{x_1^2}{2} + \frac{(x_2 + bx_1^2 - pb)^2}{2} + \sum\limits_{k=3}^{d}\frac{x_k^2}{2}
+U(x_1,\ldots, x_d) = \frac{x_1^2}{2} + \frac{(x_2 + bx_1^2 - pb)^2}{2} + \frac{1}{2} \sum_{k=3}^{d} x_k^2 \eqsp.
 \]
@@
 We set the truncation level \(n_{\mathrm{trunc}} = 50\).
-To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 2 \times 10^3$ and plot boxplots of the ergodic averages in Figure~\ref{fig:banana_shape}.
+To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 2 \times 10^3$ and display boxplots of the ergodic averages in Figure~\ref{fig:banana_shape}.
 
@@
 \subsection{Binary Logistic Regression}
-Second experiment considers the problem of logistic regression.
+In this section, we consider  logistic regression.
 Let $\mathsf{Y} = \left(\mathsf{Y}_1,\ldots,\mathsf{Y}_n\right) \in \{0,1\}^m$ be binary response variables, $\mathsf{X} \in \rset^{m \times d}$ be a feature matrix and $\theta \in \rset^d$ - vector of regression parameters.
@@
 \[
-\ell(\mathsf{Y}_i | \theta, \mathsf{X}_i) = \mathsf{Y}_i \mathsf{X}_i^T\theta - \ln{\left(1 + \rme^{\mathsf{X}_i^T\theta}\right)}
-\]
-In order to estimate $\theta$ according to given data, the Bayesian approach introduces prior distribution $\pi_0(\theta)$ and inferences the posterior density $\pi(\theta| \mathsf{Y}, \mathsf{X})$ using Bayes' rule.
+\ell(\mathsf{Y}_i | \theta, \mathsf{X}_i) = \mathsf{Y}_i \mathsf{X}_i^T\theta - \ln{(1 + \rme^{\mathsf{X}_i^T\theta})}
+\]
+In order to estimate $\theta$ according to given data, the Bayesian approach introduces prior distribution $\pi_0(\theta)$ and consider the posterior density $\pi(\theta| \mathsf{Y}, \mathsf{X})$ using Bayes' rule.
 
@@
 \label{sec:proofs}
+
 \subsection{Proof of Theorem~\ref{thm:main-repr}}
@@
 \subsection{Proof of Lemma~\ref{lem:variance}}
+In the sequel, the notation $A(\gamma,n,x) \lesssim B(\gamma,n,x)$ means that there exist $\gamma_0 > 0$, and $C < \infty$ such that for all  $\gamma \in (0,\gamma_0]$, $x \in \rset^d$, $n \in \nset$, $A(\gamma,n,x) \leq C B(\gamma,n,x)$.
+
 Under assumptions {\bf (H1)} and {\bf (H2)} the Markov chain $(X^x_{p})_{p \in \nset_0}$ is $V-$geometrically ergodic with $V(x) = 1+\|x\|^2$ and $\rho = \rme^{-\kappa\gamma}$ with $\kappa$ specified in Lemma~\ref{lem:v_ergodicity}.
@@
 \begin{proof}
-Let $\boldsymbol{\varphi}(z)=(2\pi)^{-d/2} \exp(-|z|^2/2)$,
-$z\in\mathbb R^d$, denote the density of a $d$-dimensional
-standard Gaussian random vector.
-We first remark that, for the normalized Hermite polynomial $\mathbf H_{\mathbf k}$ on $\mathbb R^d$,
+Note that for the normalized Hermite polynomial $\mathbf H_{\mathbf k}$ on $\mathbb R^d$,
 $\mathbf k\in\mathbb N_0^d$, it holds
@@
 \]
-Note that $\mu^{\prime}(x)$ is bounded due  our assumptions and from Theorem~1 in \cite{delyon1999small} and Lemma~\ref{lem:v_ergodicity} in Appendix~\ref{sec:appendix} it follows that
+Note that $\mu^{\prime}(x)$ is bounded due  our assumptions and from \cite[Theorem~1]{delyon1999small} and Lemma~\ref{lem:v_ergodicity} in Appendix~\ref{sec:appendix} it follows that
 \[
@@
 \section{Bounds for moments of  ULA}\label{sec:appendix}
-First let us introduce some definitions
-\begin{definition}
-\label{def:drift}
-We say that the Markov kernel $Q_{\gamma}$ satisfies Foster-Lyapunov drift condition if there exist a measurable function $V: X \rightarrow [1; +\infty)$, real numbers $\lambda \in (0,1)$ and $C > 0$ such that for any $x \in X$,
-\begin{equation}
-\label{eq:lyapunov}
-Q_{\gamma}V(x) \leq \lambda^{\gamma} V(x) + \gamma C
-\end{equation}
-\end{definition}
 
@@
 
-Note that conditions {\bf (H1)} and {\bf (H2)} imply (\cite[Lemma~16]{brosse2018diffusion}) that for some $K_2 > 0$ it holds
-\begin{equation}
-\label{eq:H3}
-\ps{\nabla U(x)}{x} \geq \frac{m}{2}\|x\|^2, \quad \|x\| \geq K_2
-\end{equation}
-
-Now we shall show that conditions {\bf (H1)} and {\bf (H2)} are sufficient to show that ULA kernel $Q_\gamma$ satisfies the drift condition \eqref{eq:lyapunov}
-
+In all the sequel, without loss of generality, we assume that $\nabla U(0)=0$ and we set
+\begin{equation}
+\label{eq:definition-V}
+V(x) = 1 + \|x\|^2 \eqsp.
+\end{equation}
+\begin{lem}
+  \label{lem:quadratic_behaviour}
+  Assume {\bf (H1)} and {\bf (H2)}. Then there exists $K_2 \geq 0$ such that for any $ \| x \| \geq K_2$ $\ps{\nabla U(x)}{x} \geq (m/2) \| x \|^2$.
+\end{lem}
+\begin{proof}
+  Using {\bf (H1)} and {\bf (H2)}, we have for any $x \in \rset^d$, $\| x \|\geq K_1$,
+  \begin{align*}
+    \ps{\nabla U(x)}{x}
+    &= \int_{0}^{K_1/\| x \|} D^2 U(t x ) [x^{\otimes 2}] \rmd t + \int_{K_1/\| x \|} ^ 1 D^2 U(t x ) [x^{\otimes 2}] \rmd t\\
+    & \geq m \|x \|^2 \{1- K_1 (1 +L/m)   / \| x \| \}\eqsp,
+  \end{align*}
+which proves the first statement. The second statement is obvious.
+\end{proof}
 \begin{lem}
@@
  Assume that {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$.
- Then the kernel $Q_\gamma$ from \eqref{eq:ula_kernel} satisfies drift condition \eqref{eq:lyapunov} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L}$ with drift function $V(x) = 1 + \|x\|^2$,   $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = \frac{25K_2^2}{8} + d + m$ with $K_2$ from \eqref{eq:H3}, $m$ from {\bf (H2)}.
+ Then, for any $\gamma \in (0,\bar{\gamma}]$ with $\overline{\gamma} = m/4$  the kernel $Q_\gamma$ from \eqref{eq:ula_kernel} satisfies drift condition
+ \begin{equation}
+\label{eq:lyapunov}
+Q_{\gamma}V(x) \leq \lambda^{\gamma} V(x) + \gamma C
+\end{equation}
+with   $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = 25K_2^2/8 + d + m$ with $K_2$ from \eqref{eq:H3}, $m$ from {\bf (H2)}.
 \end{lem}
 
-\begin{proof} Let $V(x) = 1 + \|x\|^2$, then
+\begin{proof} 
 $$

### 2020-04-15 Code/zv_cv.py
update

@@ -1,7 +1,7 @@
 import numpy as np
-from baselines import set_function
+from baselines import set_function, Spectral_var
 
-def ZVpolyOne(traj, traj_grad, f_target, params):
+def ZVpolyOne(traj, traj_grad, f_target, params, W_spec):
     n, d = traj.shape
     if f_target == "sum":
         samples = traj.sum(axis = 1).reshape(-1,1)
@@ -24,9 +24,10 @@ def ZVpolyOne(traj, traj_grad, f_target, params):
     paramZV1 = -np.dot(A,covariance[:d, d:])
     ZV1 = samples - np.dot(traj_grad, paramZV1)
     mean_ZV1 = np.mean(ZV1, axis = 0)
-    return mean_ZV1
+    var_ZV1 = Spectral_var(ZV1[:,0],W_spec)
+    return mean_ZV1, var_ZV1
 
-def ZVpolyTwo(traj, traj_grad, f_target, params = None):
+def ZVpolyTwo(traj, traj_grad, f_target, params, W_spec):
     n, d = traj.shape
     if f_target == "sum":
         samples = traj.sum(axis = 1).reshape(-1,1)
@@ -60,9 +61,10 @@ def ZVpolyTwo(traj, traj_grad, f_target, params = None):
     paramZV2 = - np.dot(A,B)
     ZV2 = samples + np.dot(Lpoisson, paramZV2)
     mean_ZV2 = np.mean(ZV2, axis = 0)
-    return mean_ZV2
+    var_ZV2 = Spectral_var(ZV2[:,0],W_spec)
+    return mean_ZV2, var_ZV2
 
-def CVpolyOne(traj,traj_grad, f_target, params = None):
+def CVpolyOne(traj,traj_grad, f_target, params, W_spec):
     n, d = traj.shape
     if f_target == "sum":
         samples = traj.sum(axis = 1).reshape(-1,1)
@@ -83,9 +85,10 @@ def CVpolyOne(traj,traj_grad, f_target, params = None):
     paramCV1 = covariance[:d, d:]
     CV1 = samples - np.dot(traj_grad, paramCV1)
     mean_CV1 = np.mean(CV1, axis = 0)
-    return mean_CV1
+    var_CV1 = Spectral_var(CV1[:,0],W_spec)
+    return mean_CV1, var_CV1
 
-def CVpolyTwo(traj, traj_grad, f_target, params = None):
[…]

### 2020-04-15 Article/elsarticle-mcmc-var.tex

--- 2020-02-17 aap/vr-mcmc.tex
@@
-\documentclass[bj]{imsart}
+\documentclass[article]{elsarticle}
 \makeatletter
@@
 \usepackage{graphics}
-
 \usepackage{lineno,hyperref}
@@
 \usepackage{xargs}
-%\usepackage{showlabels}
 
@@
 
+\def\bs{\mathbf{s}}
+\def\TV{\operatorname{TV}}
 \def\rmd{\mathrm{d}}
@@
 \def\rset{\mathbb{R}}
+\def\nset{\mathbb{N}}
 \def\NtrainPath{T}
-\def\TrainSet{\mathcal{T}}
+\newcommandx{\TrainSet}[3][1=T,2=n,3=N]{\mathcal{T}_{#1,#2}^{#3}}
+\newcommandx{\TrainAlg}[3][1=T,2=n,3=N]{\mathcal{\widetilde{T}}_{#1,#2}^{#3}}
+
+\newcommandx{\EstFunc}[3][1=N]{\bar{Q}^{#1}_{#2,#3}}
+\newcommandx{\EstFuncSmp}[3][1=N]{\widehat{Q}^{#1}_{#2,#3}}
+\def\bk{\mathbf{k}}
+\def\bX{\mathbf{X}}
+\def\bY{\mathbf{Y}}
+\def\Id{\operatorname{I}}
+\newcommand{\jacob}[2]{\operatorname{J}_{#1}(#2)}
+\newcommandx{\acoef}[4][1=N]{\bar{a}_{#2,#3}^{#1,#4}}
+\newcommandx{\acoefest}[4][1=N]{\hat{a}_{#2,#3}^{#1,#4}}
+
 \modulolinenumbers[5]
-\newcommand{\proofendsign}{$\Box$}
+
+\journal{Journal of \LaTeX\ Templates}
+
+%%%%%%%%%%%%%%%%%%%%%%%
+%% Elsevier bibliography styles
+%%%%%%%%%%%%%%%%%%%%%%%
+%% To change the style, put a % in front of the second line of the current style and
+%% remove the % from the second line of the style you would like to use.
+%%%%%%%%%%%%%%%%%%%%%%%
+
+%% Numbered
+%\bibliographystyle{model1-num-names}
+
+%% Numbered without titles
+%\bibliographystyle{model1a-num-names}
+
+%% Harvard
+%\bibliographystyle{model2-names.bst}\biboptions{authoryear}
+
+%% Vancouver numbered
+%\usepackage{numcompress}\bibliographystyle{model3-num-names}
+
+%% Vancouver name/year
+%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}
+
+%% APA style
+%\bibliographystyle{model5-names}\biboptions{authoryear}
+
+%% AMA style
+%\usepackage{numcompress}\bibliographystyle{model6-num-names}
+
 \newtheorem{thm}{Theorem}
@@
 \newtheorem{prop}[thm]{Proposition}
-\newtheorem{remark}{Remark}
-\newtheorem{cor}{Corollary}
-\newtheorem{example}{Example}
-\newenvironment{proof}{{\noindent \bf Proof }}
- {{\hspace*{\fill}\proofendsign\par\bigskip}}
-\def\TV{\operatorname{TV}}
-\newcommand{\jac}[1]{\operatorname{J} _{#1}}
+\newdefinition{remark}{Remark}
+\newdefinition{cor}{Corollary}
+\newdefinition{example}{Example}
+\newproof{proof}{Proof}
+
+%% `Elsevier LaTeX' style
+\bibliographystyle{elsarticle-num}
+%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \begin{frontmatter}
-\title{Variance reduction for MCMC methods via martingale representations}
-\runtitle{Variance reduction for MCMC}
-%\thankstext{T1}{This research was }
-
-\begin{aug}
-\author{\snm{D. Belomestny}\thanksref{a,b}\ead[label=e1]{denis.belomestny@uni-due.de}}
-\and
-\author{\snm{E. Moulines}\thanksref{c, b}\ead[label=e2]{eric.moulines@polytechnique.edu}}
-\and
-\author{\snm{N. Shagadatov}\thanksref{b}\ead[label=e3]{nshagadatov@edu.hse.ru}}
-\and
-\author{\snm{M. Urusov}\thanksref{a}\ead[label=e4]{murusov@uni-due.de}}
-
-\runauthor{D. Belomestny et al}
-
-\affiliation{Duisburg-Essen University, National Research University Higher School of Economics,  Ecole Polytechnique  and Duisburg-Essen University}
-
-\address[a]{
-Duisburg-Essen University,
-Faculty of Mathematics,
-D-45127 Essen
-Germany }
-
-\address[c]{
-Centre de Math\'ematiques Appliqu\'ees, UMR 7641, Ecole Polytechnique,
-France
-}
-
-\address[b]{
-National University Higher School of Economics, Moscow, Russia
-}
-
-\end{aug}
+
+\title{Martingale representations  for Markov chains   with application to MCMC \tnoteref{mytitlenote}}
+\tnotetext[mytitlenote]{This work has been funded by the  Russian Academic Excellence Project '5-100'}
+
+%% or include affiliations in footnotes:
+\author[address1,address2]{D. Belomestny\corref{mycorrespondingauthor}}
+\ead[url]{www.uni-due.de/~hm0124}
+\ead{denis.belomestny@uni-due.de}
+
+\author[address3,address2]{E. Moulines}
+\author[address2]{N. Shagadatov}
+\author[address1]{M. Urusov}
+\cortext[mycorrespondingauthor]{Corresponding author}
+
+\address[address1]{Duisburg-Essen University, Essen}
+\address[address2]{National University Higher School of Economics, Moscow}
+\address[address3]{Centre de Math\'ematiques Appliqu\'ees, UMR 7641, Ecole Polytechnique, France}
 
 \begin{abstract}
-In this paper we propose an efficient variance reduction approach for  MCMC algorithms relying on a novel discrete time martingale representation for Markov chains.
+In this paper we propose an efficient variance reduction approach for  MCMC algorithms relying on a novel discrete time martingale representation.
 Our approach is fully non-asymptotic and does not require any type of ergodicity or special product structure of the underlying density.
-By rigorously analyzing the  convergence properties of the proposed algorithm, we show that it's complexity is indeed asymptotically smaller than one of the original MCMC algorithm.
-The numerical performance of the new method is illustrated  in the case of Gaussian mixtures and Bayesian regression models.
+We rigorously analyze   complexity of the proposed algorithm and prove its reduction.
+The numerical performance is illustrated  in the case of Gaussian mixtures and binary regression.
 \end{abstract}
 
-\begin{keyword}[class=MSC]
-\kwd[Primary ]{60G40}
-\kwd{60G40}
-\kwd[; secondary ]{91G80}
+\begin{keyword}
+Monte Carlo\sep optimal stopping \sep regression \sep boosting
+\MSC[2010] 65C05 \sep 60F05 \sep 62L10 \sep 65C40 \sep 60J05 \sep 93E35
+
 \end{keyword}
 
-\begin{keyword}
-\kwd{MCMC}
-\kwd{variance reduction}
-\kwd{martingale representation}
-\end{keyword}
-
 \end{frontmatter}
 
+\linenumbers
+
 \section{Introduction}
@@
 Monte Carlo integration typically has an error variance of the form
-$\sigma^{2}/n,$ where $n$ is a sample size and \(\sigma^2\) is the variance of  integrand. We can make the variance smaller
+$\sigma^{2}/n,$ where $n$ is a sample size and \(\sigma^2\) is the variance of the integrand. We can make the variance smaller
 by using a larger value of $n$.  Alternatively,
@@
 Monte Carlo and Markov Chain Monte Carlo methods.
-Introduction to many of the variance reduction techniques can be found in \cite{christian1999monte}, \cite{rubinstein2016simulation}, \cite{GobetBook} and \cite{glasserman2013monte}.
+Introduction to many of the variance reduction techniques can be found in \cite{christian1999monte}, \cite{rubinstein2016simulation} and \cite{glasserman2013monte}.
 Recently one witnessed a revival of interest in efficient variance reduction methods  for MCMC algorithms, see for example \cite{dellaportas2012control}, \cite{mira2013zero}, \cite{brosse2018diffusion} and references therein.
-\par
+
 Suppose that we wish to compute $\pi(f):=\mathsf{E}\left[f(X)\right]$, where $X$
@@
 $\mathcal{X}\to\mathbb{R}$ with $f\in L^2(\pi)$.
-The idea of the control variates variance reduction method
- is to find a cheaply computable random variable $\zeta$ with $\mathsf{E}[\zeta]=0$ and \(\mathsf{E}[\zeta^2]<\infty,\)
+The idea of the so-called control variates variance reduction method
+ is to find a cheaply computable random variable $\xi$ with $\mathsf{E}[\xi]=0$ and \(\mathsf{E}[\xi^2]<\infty,\)
 such that the variance of the r.v.
-$f(X)-\zeta$ is small.
-The complexity of the problem of constructing classes $Z$ of control variates \(\zeta\) satisfying   $\mathsf{E}[\zeta]=0$ essentially depends on the degree of our knowledge on \(\pi.\)
+$f(X)-\xi$ is small.
+The complexity of the problem of constructing classes $\Xi$ of control variates \(\xi\) satisfying   $\mathsf{E}[\xi]=0$ essentially depends on the degree of our knowledge on \(\pi.\)
 For example, if \(\pi\) is analytically known and satisfies some regularity conditions, one can apply the well-known technique of  polynomial interpolation to construct control variates enjoying  some optimality properties, see, for example, Section~3.2 in \cite{dimov2008monte}.
-Alternatively, if an orthonormal system in \(L^2(\pi)\) is analytically available, one can build control variates \(\zeta\) as a linear combination of the corresponding basis functions, see \cite{GobetCV}.
-Furthermore, if \(\pi\) is known only up to a normalizing constant (which is often the case in Bayesian statistics), one can apply the recent approach of control variates  depending only on the gradient \(\nabla \log \pi\)  using Schr\(\ddot{\text{o}}\)dinger-type Hamiltonian operator in  \cite{mira2013zero} and Stein operator in \cite{brosse2018diffusion}.
+Alternatively, if an orthonormal system in \(L^2(\pi)\) is analytically available, one can build control variates \(\xi\) as a linear combination of the corresponding basis functions.
+Furthermore, if \(\pi\) is known only up to a normalizing constant (which is often the case in Bayesian statistics), one can apply the recent approach of
+ \cite{mira2013zero} and further worked out in\cite{oates2017control}  suggesting  control variates  depending only on the gradient \(\nabla \log \pi.\)
+
 In some situations \(\pi\) is not known analytically, but \(X\) can be represented as a function of  simple random variables with known distribution.
@@
 Markov chain \(X_p,\) \(p=0,1,2,\ldots,\) whose marginal distribution  converges  to \(\pi\) as time grows.
-One important class of the variance reduction methods in this case  is based on the Poisson equation for the corresponding  Markov chain.
-It was observed in~\cite{henderson1997variance}  that if a time-homogeneous Markov chain \((X_p)\) is stationary with stationary distribution \(\pi,\) then for any real-valued function \(G \in L^1(\pi) \) defined on the state space of the Markov chain \((X_p),\)  the function \(U(x) := G(x)-\mathsf{E}[G(X_{1})|X_0 = x]\) has zero mean with respect to \(\pi\).
-The best choice for the function \(G\) corresponds to a solution of the Poisson equation  \(\mathsf{E}[G(X_{1})|X_0 = x]-G(x)=-f(x)+\pi(f)\).
-Moreover, it is also related to the minimal asymptotic variance in the corresponding central limit theorem, see \cite{duncan2016variance} and \cite{mira2013zero}.
+One important class of the variance reduction methods in this case  is based on the so-called Poisson equation for the corresponding  Markov chain.
+It was observed in Henderson~\cite{henderson1997variance}  that if a time-homogeneous Markov chain \((X_p)\) is stationary with stationary distribution \(\pi,\) then for any real-valued function \(G \in L^1(\pi) \) defined on the state space of the Markov chain \((X_p),\)  the function \(U(x) := G(x)-\mathsf{E}[G(X_{1})|X_0 = x]\) has zero mean with respect to \(\pi\), provided that \(\pi(|G|) < \infty\).
+The best choice for the function \(G\) corresponds to a solution of the so-called Poisson equation  \(\mathsf{E}[G(X_{1})|X_0 = x]-G(x)=-f(x)+\pi(f)\).
+In fact,  the Poisson equation leads to a zero-variance control variate for the empirical mean under \(\pi.\) Moreover, it is also related to the minimal asymptotic variance in the corresponding central limit theorem, see \cite{duncan2016variance} and \cite{mira2013zero}.
 Although the Poisson equation involves the quantity of interest \(\pi(f)\)  and can not be  solved explicitly in most cases, this idea still can be used to construct some  approximations for the optimal zero-variance control variates.
-For example,  \cite{henderson1997variance} proposed to compute approximations for the solution of the Poisson equation for specific Markov chains with particular emphasis on models arising in stochastic network theory.
+For example,  Henderson~\cite{henderson1997variance} proposed to compute approximations for the solution of the Poisson equation for specific Markov chains with particular emphasis on models arising in stochastic network theory.
 In \cite{dellaportas2012control} and \cite{brosse2018diffusion}  series-type control variates are introduced and studied for reversible Markov chains.
 It is assumed in \cite{dellaportas2012control}  that the one-step conditional expectations  can be computed explicitly  for a set of basis functions.
-The authors in \cite{brosse2018diffusion} proposed another approach tailored to diffusion setting which does not require the computation of integrals of basis functions and only involves  applications of the underlying generator.
+The authors in \cite{brosse2018diffusion} proposed another approach tailored to diffusion setting which doesn't require the computation of integrals of basis functions and only involves  applications of the underlying generator.
 \par
-In this paper we focus on the  Langevin type algorithms which got much attention recently, see \cite{dalalyan2017theoretical,durmus:moulines:2017, MR2353037, MR3861816, MR2977986} and references therein.
-We propose  a generic variance reduction method for these and other types algorithms, which is purely non-asymptotic and does not require that    the conditional expectations of the corresponding Markov chain can be computed  or that the generator is known analytically.
+In this paper we focus on the  Langevin type algorithms which got much attention recently, see \cite{dalalyan2017theoretical,durmus:moulines:2017} and references therein.
+We propose  a generic variance reduction method for these and other types algorithms, which is purely non-asymptotic and does not require that    the conditional expectations of the corresponding Markov chain can be computed explicitely or that the generator is known analytically.
 Moreover, we do not need to assume stationarity or/and sampling under the invariant distribution \(\pi.\) We rigorously analyse the convergence of the method and study its complexity.
@@
 Finally, numerical examples are presented in Section~\ref{sec:num}.
-
 \section{Setup}\label{sec:setup}
-Let  \(\mathcal{X}\) be a domain in \( \mathbb{R}^d.\)  Our aim is to numerically compute  expectations of the form
+Let  \(\mathcal{X}\) be a domain in \( \mathbb{R}^d.\)  Our aim is to numerically compute the expectations of the form
 \[
@@
 However, in many practical situations  direct sampling from \(\pi\) is impossible and this precludes the use of plain Monte Carlo methods in this case.
-One popular alternative to Monte Carlo  is Markov Chain Monte Carlo (MCMC), where one is looking for a discrete time  (possibly non-homogeneous) Markov chain   \((X_p)_{p\geq 0}\) such that \(\pi\) is its unique invariant measure.
+One popular alternative to Monte Carlo  is Markov Chain Monte Carlo, where one is looking for a discrete time  (possibly non-homogeneous) Markov chain   \((X_p)_{p\geq 0}\) such that \(\pi\) is its unique invariant measure.
 In this paper we study a class of MCMC algorithms with \((X_p)_{p\geq 0}\) satisfying the  the following recurrence relation:
@@
 standard Gaussian random vectors. If
-$\mu=(1/2) \nabla U$
+${\mu=(1/2)\nabla U}$
 for some continuously differentiable function $U,$ then Markov chain~\eqref{eq:chain} can be used to approximately sample from the density
@@
 \begin{enumerate}
-\item Generate \(Y_p\sim q_p(\cdot|x)\)\;
-\item Put
+\item Generate \(Y_p\sim q_p(y|x)\)\;
+\item Take
 \begin{eqnarray*}
@@
 \begin{eqnarray*}
-\alpha(x,y)=\min\left\{1,\frac{\pi(y)}{\pi(x)}\frac{q_p(x|y)}{q_p(y|x)}\right\}.
+\alpha(x,y)=\min\left\{\frac{\pi(y)}{\pi(x)}\frac{q_p(x|y)}{q_p(y|x)}\right\}.
 \end{eqnarray*}
@@
 \end{algorithm}
-This transition is reversible with respect to \(\pi\) and therefore preserves the stationary density \(\pi\); see \cite[Chapter~2]{moulines2018}. If \(q_p\) have a wide enough
+Then, as shown in Metropolis et al.~\cite{metropolis1953equation}, this transition is reversible with respect to \(\pi\) and therefore preserves the stationary density \(\pi\).
+If \(q\) has a wide enough
 support to eventually reach any region
@@
 \begin{eqnarray*}
-q_p(y|x)=(\gamma_{p+1})^{-d/2}\boldsymbol{\varphi}\Bigl([y-x+\gamma_{p+1}\mu(x)]/\sqrt{\gamma_{p+1}}\Bigr),
+q_p(y|x)=(\gamma_{p+1})^{-d/2}\boldsymbol{\varphi}\left([y-x+\gamma_{p+1}\mu(x)]/\sqrt{\gamma_{p+1}}\right),
 \end{eqnarray*}
@@
 standard Gaussian random vector.
-The MALA algorithms usually provide noticeable speed-ups in convergence for most problems.
+MALA algorithms usually provide noticeable speed-ups in convergence for most problems.
 It is not difficult to see that the MALA chain can be compactly represented in the form
 \begin{align*}
-X_{p+1} &=X_p+\mathbbm{1}\bigl(U_{p+1}\leq \alpha(X_{p},Y_{p})\bigr)(Y_{p}-X_p),  \,
+X_{p+1} &=X_p+\mathbbm{1}\bigl(U_{p+1}\leq \alpha(X_{p},Y_{p})\bigr)(Y_{p}-X_p),  \\
 Y_{p}&=X_p-\gamma_{p+1}\mu(X_p)+\sqrt{\gamma_{p+1}}Z_{p+1},
@@
 \end{example}
-\begin{example}
-  Let \((X_t)_{t\in [0,T]}\) be the unique strong solution to a  SDE of the form:
-\begin{eqnarray}
-\label{eq:sde-inv}
-dX_t=b(X_t)\, \rmd t+\sigma(X_t)\, \rmd W_t,\quad t\geq 0,	
-\end{eqnarray}
-where \(W\) is a standard \(\mathbb{R}^m\)-valued Brownian motion, \(b:\) \(\mathbb{R}^d\to \mathbb{R}^d\) and \(\sigma:\) \(\mathbb{R}^d\times \mathbb{R}^m \to \mathbb{R}^d\) are locally Lipschitz continuous functions with at most linear growth.
-The process \((X_t)_{t\geq 0}\) is a Markov
-process and let \(L\) denote its infinitesimal generator defined by
-\begin{eqnarray*}
-Lg=b^\top \nabla g+\frac{1}{2}\text{Tr}(\sigma^\top D^2g\sigma)
-\end{eqnarray*}
-for any \(g\in C^2(\mathbb{R}^d).\)
-If there exists a continuously twice differentiable Lyapunov function \(V:\) \(\mathbb{R}^d\to \mathbb{R}_{+}\) such that
-\begin{eqnarray*}
-\sup_{x\in \mathbb{R}^d} LV(x) <\infty,\quad \limsup_{|x|\to \infty} LV(x)<0,
-\end{eqnarray*}
-then there is an invariant probability measure \(\pi\)  for \(X,\) that is, \(X_t\sim \pi\) for all \(t> 0\) if \(X_0\sim \pi.\) Invariant measures are crucial in the study of the long term behaviour of stochastic differential
-systems \eqref{eq:sde-inv}.    Under some additional assumptions,  the invariant
-measure \(\pi\) is  ergodic and this property
-can be exploited  to compute  the integrals \(\pi(f)\) for \(f\in L^2(\pi)\) by means of ergodic averages.
-The idea is to replace the diffusion \(X\) by a (simulable) discretization scheme of the form (see e.g.
-\cite{MR3861816})
-\begin{eqnarray*}
-\bar X_{n+1}=\bar X_n+\gamma_{n+1} b(\bar X_n)+\sigma(\bar X_n)(W_{\Gamma_{n+1}}-W_{\Gamma_n}), \quad n\geq 0,\quad \bar X_0=X_0,
-\end{eqnarray*}
-where \(\Gamma_n=\gamma_1+\ldots+\gamma_n\) and \((\gamma_n)_{n\geq 1}\) is a non-increasing sequence of time steps. Then for a function \(f\in L^2(\pi)\) we can approximate \(\pi(f)\) via
-\begin{eqnarray*}
-\pi_n^\gamma(f)=\frac{1}{\Gamma_n}\sum_{i=1}^n \gamma_{i}f(\bar X_{i-1}).
-\end{eqnarray*}
-Due to typically high correlation between \(X_0,X_1,\ldots\)  variance reduction is of crucial importance here.
-As a matter of fact, in many cases there is no explicit formula for the invariant measure and this makes the use of gradient  based  variance reduction techniques (see e.g.
-\cite{mira2013zero})   impossible in this case.
-On the contrary, our method can be directly used to reduce the variance of the ergodic estimator \(\pi_n^\gamma\) without explicit knowledge of \(\pi.\)
-\end{example}
+
 \section{Martingale representation and variance reduction}
 \label{seq:mart_repr}
-In this section we give a general discrete-time martingale representation for the Markov chain \eqref{eq:chain_gen} which is  used below to construct an efficient variance reduction algorithm.
+In this section we give a general discrete-time martingale representation for the chain \eqref{eq:chain_gen} which later will be used to construct an efficient variance reduction algorithm.
 Let \((\phi_k)_{k\in \mathbb{Z}_+}\) be a complete orthonormal system in \(L^2(\mathbb{R}^m, P_{\xi})\) with \(\phi_0\equiv 1\).
@@
 \begin{eqnarray*}
-\mathsf{E}[\phi_i(\xi)\phi_j(\xi)]=\delta_{ij},\quad i,j\in  \mathbb{Z}_{+}
-\end{eqnarray*}
-with \(\xi \sim P_{\xi}.\)
+\mathsf{E}[\phi_i(\xi)\phi_j(\xi)]=\delta_{ij},\quad i,j\in  \mathbb{Z}_{+}.
+\end{eqnarray*}
 Notice that this implies that the random variables
 $\phi_k(\xi)$, $k\ge1$, are centered.
-As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of Shifted Legendre polynomials for "uniform part" and Hermite polynomials for "gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the Shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\) In the sequel, we denote by $(\mathcal{G}_p)_{p\in \mathbb{Z}_+}$  the filtration
-generated by generated by $(\xi_p)_{p \in \mathbb{N}^*}$ with the convention $\mathcal{G}_0=\mathrm{triv}$.
-
+As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and products of Hermite and Legendre polynomials for MALA, as the Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([-1,1].\)
 \begin{thm}\label{prop:29032018a1}
- Let $f$ be a Borel function $\mathbb{R}^{d}\to\mathbb R$ such that
-$\mathsf{E}\left[\left|f(X_{p})\right|^{2}\right]<\infty$, with a Markov chain $(X_{p})_{p\geq 0}$ defined in \eqref{eq:chain_gen}.
-Then,
+Denote by $(\mathcal{G}_p)_{p\in \mathbb{Z}_+}$  a filtration
+with $\mathcal{G}_0=\mathrm{triv}$
+generated by $(\xi_p)_{p=1,2,\ldots,}$. Let $f$ be a Borel function $\mathbb{R}^{d}\to\mathbb R$ such that
+$\mathsf{E}\left[\left|f(X_{p})\right|^{2}\right]<\infty$. Then,
 for $p>j$, the following representation holds in \(L^2(P)\)
@@
 %where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
-The functions \((Q_{p,l})_{l=0}^p\)  can be computed by the backward recurrence: $Q_{p,p}(x)=f(x)$ and for $l \in \{0,\cdots,p-1\}$
+The functions \(Q_{p,l}(x)\)   satisfy the recursion
 \begin{eqnarray}
 \label{eq:qpl}
-Q_{p,l}(x)=\mathsf{E}\left[\left.Q_{p,l+1}(X_{l+1})\right|X_{l}=x\right].
-%=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))]
+Q_{p,l}(x)=\mathsf{E}\left[\left.Q_{p,l+1}(X_{l+1})\right|X_{l}=x\right],\quad Q_{p,p}(x)=f(x),
+%=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))],
 \end{eqnarray}
+that is, they can be computed backward via  one-step expectations.
 \end{prop}
@@
 \label{eq:first-expression-bar-a-l-k}
-\bar{a}_{l,k}(x)
+\acoef{l}{n}{k}(x)
 =\sum_{p=l}^{N+n}\omega_{p,n}^{N}a_{p,l,k}(x)=\mathsf{E}\left[\left.\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right)\phi_k(\xi_{l})\right|X_{l-1}=x\right].
@@
 Since \(X_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\geq 1,\)  the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
+
+The coefficients \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. One way of estimating them can be based on nonparametric regression.
+\erici{to be discussed should we really present two algorithms}
+We present two regression algorithms. In both algorithms we first generate \(\NtrainPath\)  paths conditionally independent of the $\sigma$-algebra generated by the burn-in sequence $X_1,\dots,X_N$:
+\begin{equation}
+\label{eq:training-paths}
+\TrainSet=\left\{(X^{(s)}_{N+1},\ldots,X^{(s)}_{N+n}),\quad s=1,\ldots,\NtrainPath\right\}
+\end{equation}
+of the Markov chain \(X\)
+(the so-called ``training paths'').
+
+In \textbf{Algorithm~1}, we estimate by linear regression the family of functions $\EstFunc{l}{n}$ given for $l=N+1,\ldots,N+n$ by
+\begin{equation}
+\label{eq:definition-EstFunc}
+\EstFunc{l}{n}(x)=\sum_{p=l}^{N+n} \omega_{p,n}^{N} Q_{p,l}\left(x\right)=\CPE{\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})}{X_{l}=x} \, .
+\end{equation}
+More precisely, we solve for each \(l=N+1,\ldots, N+n-1.\) the least squares optimization problems
+\begin{equation}\label{eq:linear-regression-problem}
+\EstFuncSmp{l}{n}=\argmin_{\psi\in \Psi}\sum_{s=1}^{\NtrainPath} \left | \sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X^{(s)}_{p})-\psi(X^{(s)}_{l})\right|^2
+\end{equation}
+Next  we estimate the coefficients \( \acoef{l}{n}{k}\) using the formula
+\begin{equation}
+\label{eq:a_est_int}
+\acoefest{l}{n}{k}(x)=\CPE{\phi_k\left(\xi\right)\EstFuncSmp{l}{n} \left(\Phi_l(x,\xi)\right)}{\TrainSet} \,,
+\end{equation}
+where $\Phi_l$ is defined in \eqref{eq:chain_gen} and \(\Psi\) is a  class of  functions on \(\mathbb{R}^d.\)
+and $\xi$ is independent of $\TrainSet$ with distribution $P_\xi$.
+The integration \eqref{eq:a_est_int} can in many cases be done in closed analytical form.
+for more details,  see Section~\ref{sec:num}.
+
+%In \textbf{Algorithm~2}, we  estimate by linear regression the family of functions \(Q_{p,l}\) for $N+1 \leq l < p \leq N+n$ via solving the least-squares problem:
+%\begin{equation}\label{eq:est_a_direct}
+%\EstQ{l}{p}=\argmin_{\psi\in \Psi}\sum_{s=1}^{\NtrainPath} \left| \left(f(X^{(s)}_{p})\right)-\psi(X^{(s)}_{l-1})\right|^2,
+%\end{equation}
+
+Upon estimating  the coefficients \((\widehat a_{l,k})\) by one of the above approaches, one can construct the empirical version of \(M_{K,n}^N(f)\) in the form
+\begin{eqnarray*}
+\widehat M_{K,n}^N(f) =\sum_{k=1}^K \sum_{l=N+1}^{N+n}\widehat a_{l,k}(X_{l-1})\phi_k(\xi_{l}).
+\end{eqnarray*}
+Obviously \(\mathsf{E}[\widehat M_{K,n}^N(f)|\TrainSet]=0\) and \(\widehat M_{K,n}^N(f)\) is indeed a valid control variate in that it
+does not introduce any bias.
 
@@
 \end{equation}
-should be small for \(K\) large enough. In this section   we provide  an analysis of the variance-reduced ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
+should be small for \(K\) large enough.
+In this Section  we make this statement more precise for the case of ULA, here we provide  an analysis of the variance-reduced ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
 We shall use the notations $\mathbb N=\{1,2,\ldots\}$ and $\mathbb N_0=\mathbb N\cup\{0\}$.
@@
 $$
-\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
+\mathbf{H}_\mathbf{k}(x)=\prod_{i=1}^d H_{k_i}(x_i),\quad x=(x_i)\in\mathbb R^d.
 $$
@@
 $|\mathbf{k}|=\sum_{i=1}^d k_i$ for $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
-and we set $\mathcal G_p=\sigma(Z_1,\ldots,Z_p)$, $p\in\mathbb N$, and $\mathcal G_0=\mathrm{triv}$.
+and we set $\mathcal G_0=\mathrm{triv}$ and for $p \in \nset$,
+\begin{equation}
+\label{eq:definition-G-p}
+\mathcal G_p=\sigma(Z_1,\ldots,Z_p) \,.
+\end{equation}
 Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
@@
 \bar{a}_{l,\mathbf{k}}(x)
-& =\sum_{p=l}^{N+n}\omega_{p,n}^{N}a_{p,l,\mathbf{k}}(x) =\mathsf{E}\left[\left.\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right)\mathbf{H}_\mathbf{k}\left(Z_{l}\right)\right|X_{l-1}=x\right].
+& =\sum_{p=l}^{N+n}\omega_{p,n}^{N}a_{p,l,\mathbf{k}}(x)\\
+& =\mathsf{E}\left[\left.\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right)\mathbf{H}_\mathbf{k}\left(Z_{l}\right)\right|X_{l-1}=x\right].
 \end{align*}
 For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
-of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we are interested in its conditional
+of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a5}), we are interested in its conditional
 Mean Squared Error (MSE),
@@
 \omega_{p,n}^N\,
-\|Q_\gamma^{N+1,p}(X_N,\cdot)-\pi(\cdot)\|_{TV}^2,
+\|Q_\gamma^{N+1,p}(X_N,\cdot)-\pi(\cdot)\|_{\TV}^2,
 \notag
@@
 $$
-\|\mu-\nu\|_{TV}=\sup_{A\in\mathcal B(\mathbb R^d)}
+\|\mu-\nu\|_{\TV}=\sup_{A\in\mathcal B(\mathbb R^d)}
 |\mu(A)-\nu(A)|,
@@
 $$
-R_\gamma(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\}dy,
+R_\gamma(x,\cdot)=N\left(x-\gamma\mu(x),\gamma\right),
 $$
@@
 the kernel $Q_\gamma^{k,l}$ is
-$
+$$
 Q_\gamma^{k,l}=R_{\gamma_l}\cdots R_{\gamma_k},
-$
+$$
 which, finally, provides the (random) measure
@@
 used in the right-hand side of~\eqref{eq:06042018a2}.
+
 Different specific upper bounds for the squared bias
@@
 is provided in  in~\cite[Theorem~17]{durmus:moulines:2017}.
-As for the estimator~\eqref{eq:29032018a3}, we introduce lemma of conditional variance decomposition.
-It follows from~\eqref{eq:29032018a5}, Theorem~\ref{prop:29032018a1} and independence condition for the sequence $(Z_p)_{p \geq 1}$.
-\begin{lem}
-\label{cond_var_dec}
-	For variance-reduced estimator~\eqref{eq:29032018a3}, defined for the Markov Chain~\eqref{eq:chain_gen} the following representation holds
+As for estimator~\eqref{eq:29032018a3},
+it follows from~\eqref{eq:29032018a5}
+and Proposition~\ref{prop:29032018a1}
+applied for $j=N$ that
 \begin{equation}\label{eq:29032018a6}
@@
 \end{equation}
-\end{lem}
-Now we present an upper bound for the right-hand sides of~\eqref{eq:29032018a6} in case of ULA algorithm.
+Now we derive an upper bound for the right-hand sides of~\eqref{eq:29032018a6}.
 \begin{thm}\label{th:mr}
@@
 Suppose that $f$  and \(\mu\) are \(K+1\) times continuously differentiable
-and it holds
-\begin{eqnarray*}
-|\partial^{\mathbf{k}} f(x)|\le  B_f,\quad |\partial^{\mathbf{k}}\mu(x)|\le  B_\mu, \quad
+with bounded derivatives
+\begin{eqnarray*}
+|\partial^{\mathbf{k}} f(x)|&\le & B_f,\quad
+x\in\mathbb R^d,
+\\
+|\partial^{\mathbf{k}}\mu(x)|&\le & B_\mu,\quad
 x\in\mathbb R^d
@@
 \[
-\jac{\mu}(x)\ge b_\mu I ,\quad x\in\mathbb R^d,
-\]
-for some positive number $b_\mu\in(0,B_\mu]$  where  \(\jac{\mu}\) stands for Jacobian of \(\mu.\)
+\jacob{\mu}{x} \ge b_\mu \Id \, ,\quad x\in\mathbb R^d \, ,
+\]
+for some $b_\mu\in(0,B_\mu]$ and all \(i=1,\ldots, d.\)
 Let $(\gamma_k)_{k\in\mathbb N}$ be a sequence of positive
 and non-increasing step sizes with $\sum_{k=1}^\infty \gamma_k=\infty$.
-We  assume that $\gamma_1<\frac1{B_\mu}$ and that
+We also assume that $\gamma_1<\frac1{B_\mu}$ and that
 \begin{equation}\label{eq:17042018a1}
@@
 \begin{remark}
-Note that in  Theorem~\eqref{th:mr} we don't require that  the function \(\mu\) itself is bounded.
 Assumption~\eqref{eq:17042018a1} is not restrictive.
@@
 
-\section{Nonparametric regression}
-The functions \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. This amounts to solve a nonparametric regression problem.
-We present a generic regression algorithm. Algorithm starts with estimating the functions $\bar{Q}_{l}$ for $l=N+1,\ldots,N+n,$ where
-\begin{equation*}
-\bar{Q}_{l}(x)=\sum_{p=l}^{N+n} \omega_{p,n}^{N} Q_{p,l}\left(x\right)=\omega_{l,n}^{N}f(x)+\mathsf{E}\left[\left.\sum_{p=l+1}^{N+n}\omega_{p,n}^{N}f(X_{p})\right|X_{l}=x\right] .
-\end{equation*}
-We first generate \(\NtrainPath\)  paths conditionally independent of the $\sigma$-algebra generated by the burn-in sequence $X_1,\dots,X_N$:
-\[
-\TrainSet=\Bigl\{(X^{(s)}_{N+1},\ldots,X^{(s)}_{N+n}),\quad s=1,\ldots,\NtrainPath\Bigr\}
-\]
-of the chain \(X\) (referred to as \emph{training paths}).
-The regression algorithm proceeds with estimating the  functions \((\bar Q_{l})\) by solving  the least squares optimization problems
-\begin{equation}\label{eq:06042018a1}
-{\widehat  Q}_{l}=\argmin_{\psi\in \Psi}\sum_{s=1}^{\NtrainPath} \left | \sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X^{(s)}_{p})-\psi(X^{(s)}_{l})\right|^2
-\end{equation}
-for \(l=N+1,\ldots, N+n-1.\)
-where \(\Psi\) is a  class of  functions on \(\mathbb{R}^d.\)
-Next  we estimate the coefficients \( \bar a_{l,k}\) using
-the formula
-\begin{equation}
-\label{eq:a_est_int}
-\widehat a_{l,k}(x)=\mathsf{E}\left[\left.\phi_k\left(\xi\right)\widehat Q_{l}\left(\Phi_l(x,\xi)\right)\right | \TrainSet\right] \,,
-\end{equation}
-where $\xi$ is a random variable independent from $\mathcal{T}$ with law $P_\xi,$
-$\Phi_l$ is defined in \eqref{eq:chain_gen} and the expectation is taken according to known distribution \(P_\xi\).
-Conditioning on $\TrainSet$ underlines that estimation of $\widehat Q_{l}$ derived from training trajectories.
-In many cases integration can be done in closed analytical form.
-
- For example, in the case of ULA algorithm we have \(P_\xi$ is the multivariate normal distribution $ \mathcal{N}(0,I_d)\) and \(\Phi_l(x,\xi)=x-\gamma_{l}\mu(x)+\sqrt{\gamma_{l}}\xi.\) Therefore if we use polynomials to approximate \((\bar Q_{l}),\) then \((\widehat a_{l,k})\) can be even computed in closed form; implementations details are provided in  Section~\ref{sec:num}.
-
-Upon estimating  the coefficients \((\widehat a_{l,k})\), one can construct an empirical estimate  of \(M_{K,n}^N(f)\) in the form
-\begin{eqnarray*}
-\widehat M_{K,n}^N(f) =\sum_{1\leq  k \le K}\sum_{l=N+1}^{N+n}\widehat a_{l,k}(X_{l-1})\phi_k(\xi_{l}).
-\end{eqnarray*}
-Obviously \(\mathsf{E}[\widehat M_{K,n}^N(f)|\TrainSet]=0\) and \(\widehat M_{K,n}^N(f)\) is indeed a valid control variate in that it
-does not introduce any bias.
-
 \section{Complexity analysis for ULA}
 \label{sec:coeff}
-
-The following result quantifies the error of estimating the functions \((\bar Q_l)\)
-via the algorithm and its proof follows from Theorem~2.2 in \cite{audibert2011robust}.
+We now provide a complexity algorithm of our variance reduced estimator for the ULA algorithm.
+We focus on our  second algorithm where we estimate $\{ \EstFunc{l}{n} \}_{l=N}^{N+n}$ by linear regression over $K$ basis functions. We denote by
+\begin{equation}
+\label{eq:definition-train-alg}
+\TrainAlg= \sigma(\TrainSet) \vee \mathcal{G}_N \, ,
+\end{equation}
+where $\TrainSet$ is the set of training trajectories (see \eqref{eq:training-paths}) and $\mathcal{G}_N$ defined in \eqref{eq:definition-G-p}.
+
+The first step in the proof is obtained a high-probability bound for the quadratic risk $\CPE{|\EstFunc{l}{n}(X_l)-\EstFuncSmp{l}{n}(X_l)|^2}{\TrainAlg}$
 \begin{thm}\label{berm:theorem:regression_cv}
  Suppose that for any \(l\in \{N+1,\ldots,N+n+1\},\)
-\[
-\mathsf{E}\left[\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})-\bar Q_l(X_l)\right)^4\right]\leq \sigma_l^4,
-\]
-for some finite positive numbers \(\sigma_{N+1},\ldots, \sigma_{N+n}. \)
+\eric{I would put $\mathcal{G}_N$ there, no $X_j$}
+\[
+\mathsf{E}\left[\left.\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})-\bar Q_l(X_l)\right)^4\right | X_l\right]\leq \sigma_l^4,
+\]
+with probability \(1\) for some finite positive numbers \(\sigma_{N+1},\ldots, \sigma_{N+n}. \)
 Furthermore, assume that \(\Psi=\mathrm{span}\{\psi_1,\ldots,\psi_D\},\) where the functions  \(\psi_1,\ldots,\psi_D\) are uniformly bounded and satisfy
-\[
-\max_{l}\sup_{g\in \Psi\setminus \{0\}} \|g\|^2_\infty/\mathsf{E}[g^2(X_l) | \mathcal{G}_N]\leq B<\infty.
+\eric{I would put $\mathsf{E}[g^2(X_l) | \mathcal{G}_N]$ here}
+\[
+\max_{l}\sup_{g\in \Psi\setminus \{0\}} \|g\|^2_\infty/\mathsf{E}[g^2(X_l)]\leq B<\infty.
 \]
 Then for any values of \(\varepsilon\) and \(\NtrainPath\) such that \(2/\NtrainPath\leq \varepsilon\leq 1\) and
-\begin{eqnarray*}
+\[
 \NtrainPath\gtrsim B^2\left[BD+\log(2/\varepsilon)+\frac{B^2D^2}{\NtrainPath}\right]
-\end{eqnarray*}
-it holds with probability at least \(1-\varepsilon,\)
+\]
+it holds with probability at least \(1-\varepsilon,\) for any  \(l=N+1,\ldots,N+n, \)
 \begin{multline}
-\mathsf{E}\left[ \left.\Bigl|\bar Q_l(X_l)
--\widehat Q_l(X_l)\Bigr|^2\right | \TrainSet \vee \mathcal{G}_N\right]
+\CPE{\left|\EstFunc{l}{n}(X_l)-\EstFuncSmp{l}{n}(X_l)\right|^2}{\TrainAlg}
 \lesssim  \sigma^2_l B
@@
 \inf_{\psi\in \Psi}
-\mathsf{E}\left[\left.\left|\bar Q_l(X_{l})-\psi(X_{l})\right|^{2}\right | \mathcal{G}_N \right], \label{eq:regr_error}
+\CPE{\left|\EstFunc{l}{n}(X_{l})-\psi(X_{l})\right|^{2}}{\TrainAlg} \,, \label{eq:regr_error}
 \end{multline}
-for \(l=N+1,\ldots,N+n,
-\) with \(\lesssim\) standing for inequality up to a universal multiplicative constant.
+with \(\lesssim\) standing for inequality up to a universal multiplicative constant.
 \end{thm}
+\begin{proof}
+The proof follows from \cite[Theorem~2.2]{audibert2011robust}.
+\end{proof}
 Now we are able to give  a bound for the difference between \( M_{K,n}^N\) and \(\widehat M_{K,n}^N.\)
@@
 \label{eq:main_bound}
-&\mathsf{E}\left[\left.\left|\widehat{M}_{K,n}^{N}(f)-M_{K,n}^{N}(f)\right|^{2}\right | \TrainSet,\mathcal{G}_N\right] \\
+&\CPE{\left|\widehat{M}_{K,n}^{N}(f)-M_{K,n}^{N}(f)\right|^{2}}{\TrainAlg}\\
 \nonumber
@@
 & \phantom{\mathsf{E}|\widehat{M}_{K,n}^{N}} + K \sum_{l=n+1}^{N+n}\inf_{\psi\in \Psi}
-\mathsf{E}\left[\left.\left|\bar Q_l(X_{l})-\psi(X_{l})\right|^{2}\right | \mathcal{G}_N \right].
+\CPE{\left|\bar Q_l(X_{l})-\psi(X_{l})\right|^{2}}{\mathcal{G}_N}\, .
 \end{align}
@@
 \begin{eqnarray*}
-\mathsf{E}\left[\left.\left|\widehat{a}_{l,k}(X_{l-1})-\bar a_{l,k}(X_{l-1})\right|^{2}\right | \TrainSet \vee \mathcal{G}_N\right]\leq\mathsf{E}\left[\left.\left|\widehat{Q}_{l}\left(X_{l}\right)-\bar Q_{l}\left(X_{l}\right)\right|^{2}\right | \TrainSet \vee \mathcal{G}_N\right]
+\CPE{\left|\widehat{a}_{l,k}(X_{l-1})-\bar a_{l,k}(X_{l-1})\right|^{2}}{\TrainAlg}
+\leq
+\CPE{\left|\EstFuncSmp{l}{n} \left(X_{l}\right)-\EstFunc{l}{n} \left(X_{l}\right)\right|^{2}}{\TrainAlg}
 \end{eqnarray*}
@@
 \begin{multline*}
-\mathsf{E}\left[\left.\left|\widehat{M}_{K,n}^{N}(f)-M_{K,n}^{N}(f)\right|^{2}\right | \TrainSet \vee \mathcal{G}_N\right]
-\\
-\leq \sum_{1\leq k\le K}\sum_{l=N+1}^{N+n}\mathsf{E}\left[\left.\left|\widehat{a}_{l,k}(X_{l-1})-\bar a_{l,k}(X_{l-1})\right|^{2}\right | \TrainSet \vee \mathcal{G}_N\right].
+\CPE{\left|\widehat{M}_{K,n}^{N}(f)-M_{K,n}^{N}(f)\right|^{2}}{\TrainAlg}
+\\
+\leq \sum_{1\leq k\le K}\sum_{l=N+1}^{N+n}
+\CPE{\left|\widehat{a}_{l,k}(X_{l-1})-\bar a_{l,k}(X_{l-1})\right|^{2}}{\TrainAlg} \,.
 \end{multline*}
 \end{proof}
-\subsection{Complexity analysis of variance reduction for the ULA algorithm}
-In the case of ULA one can bound the quantities \(\sigma_l^2\) using  \(L^p\)-type Sobolev inequalities  (see \cite{Adamczak}), see Remark~\ref{rem:p-norm} in Section~\ref{sec:proofs}.
-In particular, we can derive that  under conditions of Theorem~\ref{th:mr} with \(K=1,\)
-\begin{eqnarray*}
-\sigma_l^2\lesssim \frac{1}{\Gamma^2_{N+2,N+n+1}},\quad l=N+1,\ldots,N+n+1,
-\end{eqnarray*}
-where \(\lesssim \) stands for the inequality up to a multiplicative constant not depending on \(n\) and \(N.\) Using this inequality and combining \eqref{eq:main_bound} with Theorem~\ref{th:mr}, we get for the variance of \(\widehat \pi_{K,n}^N(f)=\pi_n^N(f)-\widehat M_{K,n}^N(f),\) with probability at least \(1-\varepsilon,\)
+In the case of ULA one can bound the quantities \(\sigma_l^2\) using  Poincar\'e moment inequalities  (see \cite{aida1994moment}) similar to the proof of Lemma~\ref{lem:var_poincare}.
+In particular, we can derive that  under conditions of Theorem~\ref{th:mr} with \(K=0,\)
+\begin{eqnarray*}
+\sigma_l^2\lesssim \frac{1}{\Gamma^2_{N+2,N+n+1}},
+\end{eqnarray*}
+where \(\lesssim \) stands for the inequality up to a multiplicative constant not depending on \(n\) and \(N.\) Using this inequality and combining \eqref{eq:main_bound} with Theorem~\ref{th:mr}, we get for the variance of \(\widehat \pi_{K,n}^N(f)=\pi_n^N(f)-\widehat M_{K,n}^N(f),\) with probability at least \(1-\varepsilon\)
 \begin{eqnarray}
@@
 K \sum_{l=n+1}^{N+n}\inf_{\psi\in \Psi}
-\mathsf{E}\left[\left.\left|\bar Q_l(X_{l})-\psi(X_{l})\right|^{2}\right | \TrainSet \vee \mathcal{G}_N \right].
+\mathsf{E}\left[\left.\left|\bar Q_l(X_{l})-\psi(X_{l})\right|^{2}\right | \mathcal{G}_N \right].
 \end{eqnarray}
 In order to assess the complexity of the proposed algorithm,  we first prove that under some conditions the coefficients \(a_{p,l,\mathbf{k}}\) decay exponentially fast as \(|p-l|\to \infty.\)
+\erici{to be checked: formulation of Lemma 6 in the vector case}
 \begin{lem}
 \label{a_decay}
-Suppose that \(f,\mu\in C^1(\mathbb{R}^d),\) \(0<b_\mu I\leq\jac{\mu}(x)\leq B_\mu I\) and \(B_\mu\gamma_{l}\leq 1,\) then
-\begin{eqnarray*}
-\|a_{p,l,\mathbf{k}}\|_{\infty}\leq \sqrt{\gamma_l}\, \|\nabla f\|_\infty\exp\left(-b_{\mu}\sum_{r=l+1}^p \gamma_{r}  \right),\quad \mathbf{k}\in \mathbb{N}^d\setminus \{\mathbf{0}\}.
+Suppose that \(f,\mu\in C^1(\mathbb{R})\). Assume that there exist $b_\mu >0$, $B_\mu < \infty$ such that,
+for all $x \in \rset^d$, \(0<b_\mu \Id \leq \nabla \mu(x)$, $| \nabla \mu (x) | \leq \ B_\mu\), and \(B_\mu\gamma_{l}\leq 1\). Then
+\begin{eqnarray*}
+\|a_{p,l,\mathbf{k}}\|_{\infty}\leq \sqrt{\gamma_l}\, \| \nabla f\|_\infty\exp\left(-b_{\mu}\sum_{r=l+1}^p \gamma_{r}  \right),\quad \mathbf{k}\in \mathbb{N}^d\setminus \{\mathbf{0}\}
 \end{eqnarray*}
@@
 \begin{cor}
-Assume that
-\begin{eqnarray}
-\label{eq:gamk-decay}
-\gamma_k=\gamma_1 k^{-\alpha},\quad \alpha\in (0,1),
-\end{eqnarray}
- then
-\begin{eqnarray*}
-\|a_{p,l,\mathbf{k}}\|_{\infty}\leq \sqrt{\gamma_1} l^{-\alpha/2}\, \|\nabla f\|_\infty\exp\left(-c b_{\mu}\gamma_1 (p^{1-\alpha}-l^{1-\alpha}) \right)
+Assume that \(\gamma_k=\gamma_1 k^{-\alpha}\) for some \(\alpha\in (0,1),\) then
+\begin{eqnarray*}
+\|a_{p,l,\mathbf{k}}\|_{\infty}\leq \sqrt{\gamma_1} l^{-\alpha/2}\, \| \nabla f\|_\infty\exp\left(-c b_{\mu}\gamma_1 (p^{1-\alpha}-l^{1-\alpha}) \right)
 \end{eqnarray*}
@@
 \end{cor}
-Suppose now that the chain is close to stationarity, then \(Q_{p,l}\) are functions of \(p-l\) only.
+Suppose that the chain is close to stationarity, then \(Q_{p,l}\) are functions of \(p-l\) only.
 Furthermore,  according to Lemma~\ref{a_decay}, \(a_{p,l,\mathbf{k}}\) are exponentially small for \(p-l\) large.
 As a result we only need to compute a logarithmic (in \(n\)) number of functions \(Q_{p,l}.\)
-Hence the cost of computing the estimates  \(\widehat {a}_{p,l,\mathbf{k}}(x)\) for \(l=N+1,\ldots,N+n,\) and \(\|\mathbf{k}\|\leq K\) using regression on \(\Psi,\) is of order
+Hence the cost of computing the estimates  \(\widehat {a}_{p,l,\mathbf{k}}(x)\) for \(l=N+1,\ldots,N+n\) and \(\|\mathbf{k}\|\leq K\) using regression on \(\Psi,\) is of order
 \[
@@
 \]
-Suppose for simplicity that all functions \(\bar{Q}_l\) are in \(\Psi\) for some \(D>0,\) that is, the third term in \eqref{eq:ula_red_var} is zero.
-Then under \eqref{eq:gamk-decay} it is enough to take \(K=\lceil 1/\alpha \rceil+1\) to get
+Suppose for simplicity that all functions \(\EstFunc{l}{n}\) are in \(\Psi\) for some \(D>0,\) that is, the third term in \eqref{eq:ula_red_var} is zero.
+Then it is enough to take \(K=\lceil 1/\alpha \rceil+1\) to get
 \begin{eqnarray*}
@@
 \end{eqnarray*}
-for some constant \(C\) not depending on \(d.\) Then we have with high probability
-\begin{eqnarray*}
-\mathsf{Var}\left[\left.\widehat \pi_{K,n}^N(f)\right | \TrainSet \vee \mathcal{G}_N\right]\lesssim \frac{1}{n^{2(1-\alpha)}}\left[1+\frac{n}{\NtrainPath}\right],
-\end{eqnarray*}
-with corresponding cost proportional to \(\NtrainPath\log^{1/(1-\alpha)} (n) ,\) provided \(N/n=o(1).\) We should compare this to the standard weighted estimator \(\pi_{n}^N(f)\) with variance
+for some constant not depending on \(d.\) As a result we have with high probability
+\begin{eqnarray*}
+\mathsf{Var}\left[\left.\widehat \pi_{K,n}^N(f)\right | \TrainSet, \mathcal{G}_N\right]\lesssim \frac{1}{n^{2(1-\alpha)}}\left[1+\frac{n}{\NtrainPath}\right],
+\end{eqnarray*}
+with corresponding cost proportional to \(\log^{1/(1-\alpha)} (n) \NtrainPath,\) provided \(N/n=o(1).\) We should compare this to the standard weighted estimator \(\pi_{n}^N(f)\) with variance
 \begin{eqnarray*}
@@
 \begin{eqnarray*}
-\mathsf{Var}\left[\left. \pi_{n}^N(f)\right | \TrainSet \vee \mathcal{G}_N\right]\leq \varepsilon^2
+\mathsf{Var}\left[\left. \pi_{n}^N(f)\right | \TrainSet, \mathcal{G}_N\right]\leq \varepsilon^2
 \end{eqnarray*}
@@
 \]
-In the presence of approximation errors we have to increase \(D\) with \(n\) to balance  the statistical and approximation errors resulting in a smaller complexity reduction.
-Note that our approach does not require the computation of \(\nabla \log(\pi)\) for the construction of control variates.
-This leads to an additional complexity reduction especially in statistical application where the computation of \(\nabla \log(\pi)\) can be very costly.
+
 \section{Numerical results}
 \label{sec:num}
-If each function  \(Q_{p,l}(x) \) can be well approximated by polynomials, then we can use polynomial basis functions in   \eqref{eq:06042018a1} to approximate \(\bar Q_l\).
-Suppose that we constructed a polynomial approximation  for each function \(\bar Q_{l}\)  in the form:
+\subsection{Polynomial approximation}
+In this section we apply  polynomial regression to approximate functions \(\EstFunc{l}{n}(x) \) appearing in \eqref{eq:definition-EstFunc}.
+In this case one can derive explicit formula for the coefficients $ \acoef{l}{n}{k}(x)$  in the case of ULA with constant step size.
+Suppose we constructed a polynomial approximation  for \(\EstFuncSmp{l}{n}(x) \)  of the form:
 \begin{equation*}
-\widehat{Q}_{l}(x) = \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} x^{\mathbf{s}},\quad s=(s_1,\ldots,s_d)
+\EstFuncSmp{l}{n}(x) = \sum_{\|\mathbf{s}\|\leq m} \hat{\beta}^N_{l,n,\bs} x^{\mathbf{s}},\quad s=(s_1,\ldots,s_d)
 \end{equation*}
-with some coefficients \(\beta_{\mathbf{s}}\in \mathbb{R}.\)
-Then using the identity
-\[
-\xi^j = j! \sum_{r = 0}^{\lfloor j/2 \rfloor} \frac{1}{2^r  r! \sqrt{(j-2r)!}} H_{j-2r}(\xi),\quad \xi \in \mathbb{R},
+where the coefficients \(\hat{\beta}^N_{l,n,\bs}\) have been obtained by solving the least-square regression problem on a polynomial basis.  Then using the identity
+\[
+\xi^j = j! \sum_{r = 0}^{\lfloor j/2 \rfloor} \frac{1}{2^r  r! \sqrt{(j-2r)!}} H_{j-2r}(\xi),\quad \xi \in \rset^d,
 \]
 we derive for all $x \in \mathbb{R}^d$,
-\begin{eqnarray*}
-\widehat a_{l,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(x) \widehat{Q}_{l}(x - \gamma_l \mu(x) + \sqrt{\gamma_l}\xi)}{\TrainSet \vee \mathcal{G}_N}
-= \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{i,k_i,s_i}(x),
-\end{eqnarray*}
-where for all integers $i,k,s_i$ and $x \in \rset^d$,
+\eric{simplify this expression}
 \begin{align*}
-P_{l,k,s}(x) &= \mathsf{E} \left[ H_{k}(\xi_l) (x_l - \gamma_l \mu_l(x) + \sqrt{\gamma_l} \xi_l)^{s} \right]
+\acoefest{l}{n}{\bk}(x) =\CPE{\mathbf{H}_\mathbf{k}(x) \EstFuncSmp{l}{n}(x - \gamma \mu(x) + \sqrt{\gamma}\xi)}{\TrainAlg} = \sum_{\|\mathbf{s}\|\leq m} \hat{\beta}_{l,n,\mathbf{s}}^N \prod_{i=1}^d E_{i,k_i,s_i}(x)
 \end{align*}
-is a one-dimensional polynomial (in \(x\)) of degree at most \(s\) with analytically known coefficients.
-Alternatively, on practice we can work directly with functions \(Q_{p,l}.\) Due to (almost) time-homogenous Markov chain generated by ULA, each function $Q_{p,l}$ can be approximated by function, which depends only on the  $p-l$, that is,
-\[
- Q_{p,l}(x) = \mathsf{E} \left[f(X_p) | X_l = x\right] \approx Q^\circ_{p-l}(x)
- \]
-Thus we have
-\[
-  Q^\circ_{r}(x) = \mathsf{E} \left[f(X_{l+r}) | X_l = x \right].
- \]
-Consequently, the functions $Q^\circ_r$ can be estimated using a modified least squares criteria
-\begin{eqnarray}
-\label{eq:Qcircr}
-\widehat{Q}^\circ_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
-\end{eqnarray}
-for \(1 \leq r \leq n-1,\) where  $\widehat{Q}^\circ_0(x) = f(x)$ by definition.
-Due to Lemma~\ref{a_decay}, it is enough to estimate $Q^\circ_{r}$ for $|r| < n_{\mathrm{trunc}}$ for some truncation level \(n_{\mathrm{trunc}}\) depending on \(d\) and \(\gamma.\) It allows us to use a smaller amount of training trajectories to approximate conditional expectations $Q^\circ_{r}$.
-Finally  we construct a truncated version of our estimator:
-\[
-\pi_{K,n,n_{\mathrm{trunc}}}^N(f) = \pi_n^N(f) -  M_{K,n, n_{\mathrm{trunc}}}^{N}(f),
-\]
-where
-\begin{eqnarray*}
-{\widehat  M}_{K,n, n_{\mathrm{trunc}}}^N &=& \sum_{0<\left\| \mathbf{k}\right\| \leq K} \sum_{p=N+1}^{N+n} \omega_{p,n}^N
-\\
-&&\times
-\sum_{l=N+1}^p {\widehat  a}_{p-l,\mathbf{k}}(X_{l-1})\mathbf{H}_{\mathbf{k}}(Z_l) \mathbbm{1}\{ |p-l| < n_{\mathrm{trunc}} \}
-\end{eqnarray*}
+where for all integers $i,k,s$ and $x \in \rset^d$,
+\eric{simplify the expression}
+\begin{align*}
+E_{i,k,s}(x) &= \mathsf{E} \left[ H_{k}(\xi_i) (x_i - \gamma \mu_i(x) + \sqrt{\gamma} \xi_i)^{s} \right]
+\\
+%& = &  \int_{\mathbb{R}} H_{k}(\xi_i) (x_i - \gamma \mu(x)_i + \sqrt{\gamma} \xi_i)^{t_i} \varphi(\xi_i) d \xi_i =
+%\\
+& =  \sum_{j=0}^{s} \sum_{r= 0}^{ j/2} j!
+\frac{1}{2^r} \frac{1}{r!
+\sqrt{(t-2r)!}} \binom{s}{j} \left[ x_i - \gamma \mu_i(x) \right]^{s-j} \gamma^{j/2} \int_{\mathbb{R}} H_{k}(y) H_{j - 2r}(y) \varphi(y)\, d y
+\end{align*}
 and
 \begin{eqnarray*}
-\widehat a_{p-l,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(x) \widehat{Q}^\circ_{p-l}(x - \gamma_l \mu(x) + \sqrt{\gamma_l}\xi)}{\TrainSet \vee \mathcal{G}_N}.
-\end{eqnarray*}
+\int_{\mathbb{R}} H_{k_i}(y) H_{j - 2r}(y) \varphi(y)\, d y = \delta_{k_i, j-2r}.
+\end{eqnarray*}
+
+\begin{figure}[h]
+\centering
+\subfloat[Subfigure 1 list of figures text][d = 1]{
+\includegraphics[width=0.4\textwidth]{pictures/gm_1d_test_exp.png}
+\label{fig:subfig1}}
+\qquad
+\subfloat[Subfigure 2 list of figures text][d = 2]{
+\includegraphics[width=0.4\textwidth]{pictures/gm_2d_test.png}
+\label{fig:subfig2}}
+\caption{\label{fig:1}Histograms for Gaussian mixture.
+(a) 1-dimensional GM model: histograms of estimators for target function $f(x) = \rme^x$ on test sample (200 independent trajectories obtained by ULA algorithm).
+\textit{Red} bins correspondent to ordinary weighted estimators $\pi^N_n(f)$, \textit{green} - variance-reduced estimators $\pi^N_{1,n}(f)$.\\ (b) 2-dimensional GM model: histograms of estimators for target function $f(x) = x_1^2 + x_2^2 - \cos(x_1)$, \textit{red} bins: $\pi^N_n(f)$, \textit{green}: $\pi^N_{1,n}(f)$, \textit{yellow}: $\pi^N_{2,n}(f)$.}
+\end{figure}
+
 \subsection{Gaussian mixtures}
@@
 \[
-\pi (x) = \frac{1}{2(2\pi)^{d/2}} \left( \rme^{\frac{-\| x-a\|_2^2}{2}}  + \rme^{\frac{-\| x+a\|_2^2}{2}} \right), \quad x \in\mathbb R^d
+\pi (x) = \frac{1}{2(2\pi)^{d/2}} \left( \rme^{\frac{-| x-a|^2}{2}}  + \rme^{\frac{-| x+a|^2}{2}} \right), \quad x \in\mathbb R^d
 \]
 where $a \in \mathbb{R}^d$ is a given vector. The function $U(x)$ and its gradient are given by
-\[
-U(x) = \frac{1}{2} \|x - a\|_2^2 - \text{log}(1 + \rme^{-2x^\top a}), \,\,
+\eric{inconsistent notations withe the norm}
+\[
+U(x) = \frac{1}{2} \|x - a\|_2^2 - \text{log}(1 + \rme^{-2x^\top a})
+\]
+and
+\[
 \nabla U(x) = x-a +2a(1 + \rme^{2 x^\top a})^{-1},
@@
 respectively.
-In our experiments we considered dimensions $d = 2$ and $d = 8$ and take $a=( (2d)^{-1/2}, \dots, (2d)^{-1/2})^\top$.
-In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i,\)
- we have used constant step sizes $\gamma_i = \gamma=0.2$ and sampled $n = 10^4$ samples for $d = 2$ and $n = 2 \times 10^3$ for $d = 8$.
-We generated $\NtrainPath = 10$ independent "training" trajectories and solved the least squares  problems \eqref{eq:Qcircr} using the first order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section.
-The truncation level \(n_{\mathrm{trunc}}\) is chosen to be \(50.\) To test our variance reduction algorithm, we generated 100 independent trajectories.
- In Figure ~\ref{fig:1} we compare our approach to variance reduction methods of \cite{mira2013zero} and \cite{brosse2018diffusion}.
-
- \begin{figure}[tbh]
-\centering
-\subfloat[Subfigure 1 list of figures text][d = 2]{
-\includegraphics[width=0.45\textwidth]{pictures/GM_2d_f1.png}
-\label{fig:subfig1}}
-\qquad
-\subfloat[Subfigure 2 list of figures text][d = 8]{
-\includegraphics[width=0.45\textwidth]{pictures/GM_8d_f1.png}
-\label{fig:subfig2}}
-\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Gaussian mixture model.
-The compared estimators are the ordinary empirical average  (O), our estimator of control variates (CV-B), zero variance estimator (ZV) and the control variates using diffusion approximation (CV), along with the region obtained by the ordinary ULA of the length $10^2 \times n$ (green regions).
-\label{fig:1}}
-\end{figure}
-
-\subsection{Binary Logistic Regression}
-
-Second experiment considers the problem of logistic regression.
-Suppose we have i.i.d.
-sample $\left\{ (\mathbf{X}_i, Y_i)\right\}$ for $ i =1, \dots , m$ with features $\mathbf{X}_i \in \mathbb{R}^d $ and binary labels $Y_i \in \left\{0,1 \right\}$.
-The binary logistic regression model defines the conditional distribution of $Y$ given $X$ by a logistic function
-\begin{eqnarray*}
-r(\theta, x) = \frac{e^{\theta^T x}}{1 + e^{\theta^T x}},
-\end{eqnarray*}
-where $\theta$ is a parameter of model.
-In order to estimate $\theta$ according to given data, the Bayessian approach introduces prior distribution $\pi_0(\theta)$ and inferences the posterior density $\pi(\theta)$ using Bayes' rule.
+\eric{It is not clear which algorithm you are trying to illustrate; bowplots are better I guess}
+In our experiments we considered dimensions $d = 1$ and $d = 2$ and defined vector $a$ as $( (2d)^{-1/2}, \dots, (2d)^{-1/2})$.
+For ULA  we used constant step sizes $\gamma_i = 0.2$ and $n=1000$.
+In order to approximate coefficients $a_{p,l,\mathbf{k}}(x),$ we generated $\NtrainPath = 500$ independent "training" trajectories and solved the least squares  problems \eqref{eq:linear-regression-problem} with polynomial basis functions with maximum degree \(5\) and \(3\) for dimensions \(1\) and \(2\), respectively.
+More precisely we used the polynomials
+\begin{eqnarray*}
+\Psi^1 &=& \left\{1,x,x^2,x^3,x^4,x^5 \right\}, \text{ for } d = 1, \\
+\Psi^2 &=& \left\{1,x_1,x_2,x_1 x_2,x_1^2,x_2^2, x_1^2 x_2,x_1x_2^2,x_1^3,x_2^3 \right\}, \text{ for }  d = 2.
+\end{eqnarray*}
+We fixed \(K = 1\) for \(d=1\) and \(K=2\) for \(d = 2\).
+To test our variance reduction algorithm, we generated $N_{\mathrm{test}} = 200$ independent paths and computed empirical variance of the new variance reduced estimator $\pi_{K,n}^N(f)$ of target functions $f(x) = \rme^x$ for $d = 1$ and $f(x) = x_1^2 + x_2^2 - \cos(x_1)$ for $d = 2$.
+Figure ~\ref{fig:1} shows the histograms of weighted average estimator \(\pi_{n}^N(f)\) and variance reduced estimator \(\pi_{K,n}^N(f)\) computed on test sample.
+We have repeated the whole experiment \(5\) times \eric{far from being enough !
+make it 100 times at least}
+and presented the results in Table ~\ref{table:gm}. Eventually, we can see that new estimator has considerably reduced variance in comparison with ordinary estimator.
+
+\begin{table}[h]
+\begin{tabular}{|l|lllll|}	
+\hline
+ &  &  &\(d = 1\)  &  &  \\ \hline
+$\mathsf{Var}(\pi_n^N)$ &  0.28641 & 0.24557  &0.26398   &  0.27346 & 0.27845  \\
+$\mathsf{Var}(\pi_{1,n}^N)$ & 0.02046 &  0.02068 & 0.02789 & 0.02218 &  0.01957\\ \hline
+ &  &  &\(d = 2\)  &  &  \\ \hline
+ $\mathsf{Var}(\pi_n^N)$     & 0.09238 & 0.10268 & 0.09458 &  0.09024 &  0.09312\\
+ $\mathsf{Var}(\pi_{1,n}^N)$ & 0.01800 & 0.02156 & 0.01961 &  0.01288 &  0.01754\\
+ $\mathsf{Var}(\pi_{2,n}^N)$ & 0.01155 & 0.01341 & 0.01097 &  0.00842 &  0.01018 \\ \hline
+\end{tabular}
+\caption{Gaussian Mixtures: Empirical variances of ordinary weighted and variance-reduced estimators on test sample.}
+\label{table:gm}
+\end{table}
 
@@
 \begin{center}
-\includegraphics[width=0.6\textwidth]{pictures/BLR_banknotes_4d.png}
-\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Logistic Regression.
-The compared estimators are the ordinary empirical average  (O), our estimator of control variates (CV-B), zero variance estimator (ZV) and control variates using diffusion approximation (CV).
- \label{fig:2}}
+\includegraphics[height=0.5\textwidth]{pictures/comparison_gm.png}
+\caption{\label{fig:comparison}1-dimensional GM model.
+Vertical axis is empirical variance on test sample, horizontal axis is the length of test trajectories obtained by ULA algorithm.Red traceplot correspondents to ordinary weighted estimator, green traseplot illustrates empirical variances of variance-reduced estimator for K=1.}
 \end{center}
 \end{figure}
-In the case of  Gaussian prior $\pi_0$ with zero mean and covariance matrix $\sigma^2 I_d$, the posterior density takes the form:
-\begin{eqnarray*}
-\pi(\theta) \propto \exp \left\{ -  \mathbf{Y}^T \mathbf{X \theta} - \sum_{i=1}^m \text{log} (1 + e^{-\mathbf{\theta}^T \mathbf{X}_i}) - \frac{1}{2\sigma^2} \left\| \mathbf{\theta}\right\|_2^2 \right\},
-\end{eqnarray*}
-where $\mathbf{Y}$ is defined as $(Y_1, \dots, Y_m)^T \in \left\{ 0,1\right\}$ and $\sigma^2 $ is an additional parameter. We have
-\begin{eqnarray*}
-U( \mathbf{\theta}) = \mathbf{Y}^T \mathbf{X \theta} + \sum_{i=1}^m \text{log} (1 + e^{-\mathbf{\theta}^T \mathbf{X}_i}) + \frac{1}{2\sigma^2} \left\|\mathbf{\theta}\right\|_2^2, \,\,
-\nabla U(\mathbf{\theta}) = \mathbf{X}^T \mathbf{Y} - \sum_{i=1}^m \frac{\mathbf{X}_i}{1 + e^{\mathbf{\theta}^T \mathbf{X}_i}} + \frac{1}{\sigma^2} \mathbf{\theta}.
-\end{eqnarray*}
-To demonstrate the performance of the proposed control variates approach in the above Bayesian logistic regression model, we take a simple dataset from \cite{mira2013zero}, which contains the measurements of four variables on $m=200$ Swiss banknotes.
-Prior distribution of the regression parameter $\mathbf{\theta} = \left( \theta_1, \theta_2, \theta_3, \theta_4 \right)$ assumed to be normal with the covariance matrix $\sigma^2 I_4$, where $\sigma^2 = 100$.
-To construct trajectories of length $n = 5 \times 10^3, $ we take  constant step sizes $\gamma_i = 0.1$ for the ULA scheme with $N = 10^3$ burn-in steps.
-As in the previous experiment we use the first order polynomials approximations to analytically compute the coefficients $\hat{a}_{p-l,\mathbf{k}},$ where $T = 10$ and $n_{trunc} = 50$.
-The target function is taken to be $f(\theta) = \sum_{i=1}^{i=d}\theta_i$.
-In order to test our variance reduction algorithm, we generates \(100\) independent test trajectories.
-In Figure ~\ref{fig:2} we compare our approach to variance reduction methods of \cite{mira2013zero} and \cite{brosse2018diffusion}.
-
-\subsection{Binary Probit Regression}
-
-Finally we consider a Bayesian probit regression model in its ordinary form, following \cite{mira2013zero}. More precisely, the log-likelihood of the model looks as follows
-\begin{equation*}
-    \mathbf{L}(\mathbf{Y}| \theta, \mathbf{X}) = \sum_{i=1}^{m} \left[ \mathbf{Y}_i \text{log} (\Phi(\theta^T \mathbf{X}_i)) + (1-\mathbf{Y}_i) \text{log}(\Phi(-\theta^T \mathbf{X}_i))\right],
-\end{equation*}
-where $\theta \in \mathbb{R}^d$ is a parameter of the model and $\Phi$ is a cumulative distribution function of the standard normal distribution $\mathcal{N}(0,1)$.
-Then, if the prior $\pi(\mathbf{\theta}) = \mathcal{N}(0, \sigma^2 I_d)$, then the posterior distribution expresses as
-\begin{equation*}
-    \pi(\theta | \mathbf{Y}, \mathbf{X})  \propto \text{exp} \left\{\sum_{i=1}^{m} \left[\mathbf{Y}_i \text{log} (\Phi(\theta^T \mathbf{X}_i)) + (1-\mathbf{Y}_i) \text{log}(\Phi(-\theta^T \mathbf{X}_i))\right]- \frac{1}{2\sigma^2} \left\| \mathbf{\theta}\right\|_2^2  \right\}
-\end{equation*}
-Consequently,
-\begin{equation*}
-    U(\theta) =  - \sum_{i=1}^{m} \left[\mathbf{Y}_i \text{log} (\Phi(\theta^T \mathbf{X}_i)) + (1-\mathbf{Y}_i) \text{log}(\Phi(-\theta^T \mathbf{X}_i))\right] + \frac{1}{2\sigma^2} \left\| \mathbf{\theta}\right\|_2^2
-\end{equation*}
+In order to illustrate the dependence of variances of the proposed variance reduced estimator on  the  number of elements in trajectory,  we report in Figure ~\ref{fig:comparison}  the traceplots of the empirical variances versus $n$ for the case of the one-dimensional Gaussian mixture and $K =1$.
+One may observe that the sample size needed to achieve the "almost zero" variance  is much smaller for the variance reduced estimator \(\pi_{K,n}^N(f)\) than for the ordinary weighted average estimator \(\pi_{n}^N(f)\) \eric{what do you mean by that ?}.
 
@@
 \begin{center}
-\includegraphics[width=0.6\textwidth]{pictures/Probit_banknotes_4d.png}
-\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Probit Regression.
-The compared estimators are the ordinary empirical average  (O), our modified estimator (CV-B), zero variance estimator (ZV) and the estimator with control variates using asymptotic variance minimization (CV).
- \label{fig:3}}
+\includegraphics[height=0.5\textwidth]{pictures/blr_2d_test.png}
+\caption{\label{fig:blr} Binary Logistic Regression: Histograms of estimators for target function $f(\theta) = 2 \theta_1^2 + 7 \theta_2^2$ on test sample.
+\textit{Red} bins correspondent to ordinary weighted estimators $\pi^N_n(f)$, \textit{green} - variance-reduced estimators $\pi^N_{1,n}(f)$and \textit{yellow} - $\pi^N_{2,n}(f)$.}
 \end{center}
@@
 
-\begin{equation*}
-    \nabla U(\theta) = - \mathbf{X}^T s + \frac{1}{\sigma^2} \theta,
-\end{equation*}
-where the elements of the \(m\)-dimensional column vector $s$ are given by
-\begin{equation*}
-    s_i = \mathbf{Y}_i \Omega (\theta^T \mathbf{X}_i) - (1 - \mathbf{Y}_i) \Omega(- \theta^T \mathbf{X}_i), \quad i = 1, 2, \dots, m,
-\end{equation*}
-and $\Omega(\theta^T \mathbf{X}_i) : = \phi(\theta^T \mathbf{X}_i) / \Phi(\theta^T \mathbf{X}_i)$, where $\phi$ stands for the density of the standard normal distribution.
-As in the previous examples, to show the performance  of the proposed variance reduced estimator for the Bayesian probit regression model, banknotes dataset is used and the regularization parameter is taken to be $\sigma^2=100$.
-To generate trajectories of the length $n = 5 \times 10^3, $ we take a constant step size $\gamma_i = 0.1$ for the ULA scheme after $N = 10^3$ burn-in steps.
-We use again the first order polynomials approximations to  compute the coefficients $\hat{a}_{p-l,\mathbf{k}}$ and fixed $T = 10$, $n_{trunc} = 50$ and $K=1$.
-The target function is now $f(\theta) = \sum_{i=1}^{i=d}\theta_i$.
-In order to test our variance reduction algorithm, we generate \(100\) independent test trajectories.
-In Figure ~\ref{fig:3} we compare our approach to the variance reduction methods of \cite{mira2013zero} and \cite{brosse2018diffusion}.
+\subsection{Binary Logistic regression}
+Second experiment considers the problem of logistic regression, similar to that considered by Dalalyan \cite{dalalyan2017theoretical}.
+Suppose we have i.i.d.
+sample $\left\{ (\bX_i, Y_i)\right\}_{i=1}^m$ with features $\bX_i \in \mathbb{R}^p $ and binary labels $Y_i \in \left\{0,1 \right\}$.
+The binary logistic regression model defines the conditional distribution of $Y$ given $\bX$ as Bernoulli random variable with
+success probability specified by the logistic function
+\begin{equation}
+r(\theta, \bX) = \frac{\rme^{\theta^T \bX}}{1 + \rme^{\theta^T \bX}}
+\end{equation}
+where $\theta$ is parameter of model.
+We use a Bayesian setting and denote by $\pi_0(\theta)$ the prior distribution.
+We are willing to sample the posterior density $\pi_m(\theta)$ of the parameter $\theta$ given the observations  $\left\{ (\bX_i, Y_i)\right\}_{i=1}^m$.
+Taking a Gaussian prior $\pi_0$ with zero mean and covariance matrix proportional to the inverse of the Gram matrix $\Sigma_{\bX} = \frac{1}{n} \sum_{i=1}^m \bX_i \bX_i^T$ \eric{I think this is called a Zellner prior} the posterior density takes the form
+\begin{eqnarray*}
+\pi(\theta) \propto \exp \left\{ -  \bY^T \mathbf{X \theta} - \sum_{i=1}^m \text{log} (1 + \rme^{-\mathbf{\theta}^T \bX_i}) - \frac{\lambda}{2} \left\| \Sigma_{\bX}^{1/2} \mathbf{\theta}\right\|_2^2 \right\}
+\end{eqnarray*}
+where $\bY=(Y_1, \dots, Y_m)^T \in \left\{ 0,1\right\}$ and $\lambda > 0$ additional parameter specified by practitioner. Denote
+\begin{eqnarray*}
+U( \mathbf{\theta}) = \bY^T \mathbf{X \theta} + \sum_{i=1}^m \text{log} (1 + \rme^{-\mathbf{\theta}^T \bX_i}) + \frac{\lambda}{2} \left\| \Sigma_{\bX}^{1/2} \mathbf{\theta}\right\|_2^2
+\end{eqnarray*}
+\begin{eqnarray*}
+\nabla U(\mathbf{\theta}) = \bX^T \bY - \sum_{i=1}^m \frac{\bX_i}{1 + \rme^{\mathbf{\theta}^T \bX_i}} + \lambda \Sigma_{\bX} \theta \, .
+\end{eqnarray*}
+
+\begin{table}[h]
+\begin{tabular}{|l|lllll|}	
+\hline
+ &  &  &\(d = 2\)  &  &  \\ \hline
+$\operatorname{Var}(\pi_n^N)$    & 0.03727 &  0.02124 &  0.04769  & 0.01147  & 0.02974   \\
+$\operatorname{Var}(\pi_{1,n}^N)$& 0.00269 &  0.00224 &  0.00306  & 0.00179  & 0.00196  \\
+$\operatorname{Var}(\pi_{2,n}^N)$& 0.00182 &  0.00117 &  0.00213  & 0.00100  & 0.00107 \\ \hline
+\end{tabular}
+\caption{BLR: Empirical variance of ordinary weighted and variance-reduced estimators on test sample.}
+\label{table:blr}
+\end{table}
+
+In our second experiment, we randomly generated $m$ independent samples as in paper \cite{dalalyan2017theoretical}, more precisely features $\bX_i$ were generated from a Rademacher distribution and then normalized to have an norm equal to one.
+Each target variable $Y_i$ has been obtained from a Bernoulli distribution with parameter $r (\mathbf{\theta}_{\text{true}},\mathbf{x})$, where $\mathbf{\theta}_{\text{true}}$ is defined as $(1, \dots, 1)^T$.
+We fix $d = 2$ and generated $m = 50$ samples according Rademacher distribution.
+To construct trajectories of length $n = 500 $ we determined constant step size $\gamma_i = 0.02$ for ULA scheme.
+As in previous experiment we use polynomials approximations to explicitly compute $a_{p,l,k}$ and fixed $\NtrainPath = 300$, $N_{test} = 200$ and $K=2$.
+The target function defined as
+\eric{$\lambda$ is not specified}
+\begin{eqnarray*}
+f(\mathbf{\theta}) = 2 \theta_1^2 + 7 \theta_2^2
+\end{eqnarray*}
+Table ~\ref{table:blr} summarizes results of conventional weighted estimator and variance reduced estimator.
 
 \section{Proofs}
-\label{sec:proofs}
 \subsection{Proof of Theorem~\ref{th:mr}}
@@
 \\
-%& =
-%\gamma_l^{|\mathbf k'|/2}\frac{\sqrt{(\mathbf k-\mathbf k')!}}{\sqrt{\mathbf k!}}
-%\int_{\mathbb R^d}\ldots\int_{\mathbb R^d}
-%\partial_{y_l}^{\mathbf k'}[f\circ G_{p,l}](x,\sqrt{\gamma_{l}}z_{l},\ldots,\sqrt{\gamma_{p}}z_{p})
-%\\
+& =
+\gamma_l^{|\mathbf k'|/2}\frac{\sqrt{(\mathbf k-\mathbf k')!}}{\sqrt{\mathbf k!}}
+\int_{\mathbb R^d}\ldots\int_{\mathbb R^d}
+\partial_{y_l}^{\mathbf k'}[f\circ G_{p,l}](x,\sqrt{\gamma_{l}}z_{l},\ldots,\sqrt{\gamma_{p}}z_{p})
+\\
 &\hspace{1em}\times
@@
 &=\left(\gamma_{l}^{|\mathbf{k}'|}\frac{(\mathbf{k}-\mathbf{k}')!}{\mathbf{k}!}\right)^{1/2}
-\\
-&\times \sum_{p=l}^{N+n}\omega_{p,n}^N
+\,\sum_{p=l}^{N+n}\omega_{p,n}^N
 \mathsf E\left[\left.
@@
 where for any two multi-indices \(\mathbf{r},\) \(\mathbf{q}\) from \(\mathbb{N}_0^d\)
-\begin{multline*}
-Q(\mathbf{r},\mathbf{q})
-\\
-=\mathsf{E}\left\{
+\begin{eqnarray*}
+Q(\mathbf{r},\mathbf{q})=\mathsf{E}\left\{
 \left(\mathsf{E}\left[\sum_{p=l}^{N+n}\omega_{p,n}^{N}\left(\partial_{y_{l}}^{\mathbf r}f\left(X_{p}\right)-\mathsf{E}\left[\partial_{y_{l}}^{\mathbf r}f\left(X_{p}\right)|X_{l-1}\right]\right)\mathbf{H}_{\mathbf{q}}(Z_{l})\Big|X_{l-1}\right]\right)^{2}
 \Bigg|\mathcal G_N \right\}.
-\end{multline*}
+\end{eqnarray*}
 In \eqref{eq:sum_abar} the first sum runs over all nonempty subsets $I$ of the set $\{1,\ldots,d\}.$
@@
 \end{eqnarray*}
-Next we show that under the conditions  of Theorem~\ref{th:mr}, the quantity \(R_{l,n,N}^{I,K}\) is uniformly bounded in \(l,n,N,I.\) For the sake of simplicity we present the proof only in one-dimensional case.
-First we need to prove several auxiliary results.
+Next result we show that under the conditions  of Theorem~\ref{th:mr}, the quantity \(R_{l,n,N}^{I,K}\) is uniformly bounded in \(l,n,N,I.\) First we need to prove several auxiliary results.
 \begin{lem}\label{lem:06062018a1}
@@
 we obtain~\eqref{eq:08062018b3} for $r=1$.
+
 The induction hypothesis is now that the inequality
@@
 Furthermore,
-\begin{eqnarray*}
-\partial_{y_{j}}\left[\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X_{p-1}}{k!}\right)^{m_{k}}\right]&=&\sum_{s=1}^{r-1}\frac{m_{s}}{s!}\left(\frac{\partial_{y_{l}}^{s}X_{p-1}}{s!}\right)^{m_{s}-1}\partial_{y_{j}}\partial_{y_{l}}^{s}X_{p-1}
-\\
-&& \times \prod_{k\le r-1,\,k\neq s}\left(\frac{\partial_{y_{l}}^{k}X_{p-1}}{k!}\right)^{m_{k}}.
-\end{eqnarray*}
+\[
+\partial_{y_{j}}\left[\prod_{k=1}^{r-1}\left(\frac{\partial_{y_{l}}^{k}X_{p-1}}{k!}\right)^{m_{k}}\right]=\sum_{s=1}^{r-1}\frac{m_{s}}{s!}\left(\frac{\partial_{y_{l}}^{s}X_{p-1}}{s!}\right)^{m_{s}-1}\partial_{y_{j}}\partial_{y_{l}}^{s}X_{p-1}\prod_{k\le r-1,\,k\neq s}\left(\frac{\partial_{y_{l}}^{k}X_{p-1}}{k!}\right)^{m_{k}}.
+\]
 Using Lemma~\ref{lem:06062018a2},
@@
 \begin{align*}
-&\left|\epsilon_{l,j,p}\right|  \leq  B_{\mu}C_{r}\prod_{k=l+1}^{p-1}\alpha_{k}\prod_{k=j+1}^{p-1}\alpha_{k}+B_{\mu}\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\left[\prod_{k=j+1}^{p-1}\alpha_{k}\right]
-\\
-&\times \prod_{s=1}^{r-1}\left(\frac{C_{s}\prod_{k=l+1}^{p-1}\alpha_{k}}{s!}\right)^{m_{s}}\\
-&+B_{\mu}\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\sum_{t=1}^{r-1}\frac{m_t}{t!}\left(\frac{C_{t}\prod_{k=l+1}^{p-1}\alpha_{k}}{t!}\right)^{m_{t}-1}c_t\left[\prod_{k=l+1}^{p-1}\alpha_{k}\right]
-\\
-&\times \prod_{s\le r-1,\,s\neq t}\left(\frac{C_{s}\prod_{k=l+1}^{p-1}\alpha_{k}}{s!}\right)^{m_{s}}\leq \const\prod_{k=l+1}^{j}\alpha_{k}\prod_{k=j+1}^{p}\alpha_{k}^{2}
+&\left|\epsilon_{l,j,p}\right|  \leq  B_{\mu}C_{r}\prod_{k=l+1}^{p-1}\alpha_{k}\prod_{k=j+1}^{p-1}\alpha_{k}+B_{\mu}\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\left[\prod_{k=j+1}^{p-1}\alpha_{k}\right]\prod_{s=1}^{r-1}\left(\frac{C_{s}\prod_{k=l+1}^{p-1}\alpha_{k}}{s!}\right)^{m_{s}}\\
+&+B_{\mu}\sum\frac{r!}{m_{1}!\ldots m_{r-1}!\,}\sum_{t=1}^{r-1}\frac{m_t}{t!}\left(\frac{C_{t}\prod_{k=l+1}^{p-1}\alpha_{k}}{t!}\right)^{m_{t}-1}
+c_t\left[\prod_{k=l+1}^{p-1}\alpha_{k}\right]\prod_{s\le r-1,\,s\neq t}\left(\frac{C_{s}\prod_{k=l+1}^{p-1}\alpha_{k}}{s!}\right)^{m_{s}}\\
+& \leq \const\prod_{k=l+1}^{j}\alpha_{k}\prod_{k=j+1}^{p}\alpha_{k}^{2}
 \end{align*}
@@
 and $\|\cdot\|$ denotes the Euclidean norm.
-Notice that \(\partial_{Z_j} F=\sqrt{\gamma_j}\,\partial_{y_j} F\) and
-hence
+Notice that
+$$
+\partial_{Z_j} F=\sqrt{\gamma_j}\,\partial_{y_j} F,
+$$
+hence,
 $$
@@
 Then
-\begin{eqnarray*}
-\partial_{y_{j}}\partial_{y_{l}}^{K}f\left(X_{p}\right)
-&=&\sum\frac{K!}{m_{1}!\ldots m_{K}!}f^{(m_{1}+\ldots+m_{K}+1)}(X_{p})\left[\partial_{y_{j}}X_{p}\right]\prod_{k=1}^{K}\left(\frac{\partial_{y_{l}}^{k}X_{p}}{k!}\right)^{m_{k}} \\
-&&+\sum\frac{K!}{m_{1}!\ldots m_{K}!}f^{(m_{1}+\ldots+m_{K})}(X_{p})
+\begin{align*}
+&\partial_{y_{j}}\partial_{y_{l}}^{K}f\left(X_{p}\right)
+=\sum\frac{K!}{m_{1}!\ldots m_{K}!}f^{(m_{1}+\ldots+m_{K}+1)}(X_{p})\left[\partial_{y_{j}}X_{p}\right]\prod_{k=1}^{K}\left(\frac{\partial_{y_{l}}^{k}X_{p}}{k!}\right)^{m_{k}} \\
+&+\sum\frac{K!}{m_{1}!\ldots m_{K}!}f^{(m_{1}+\ldots+m_{K})}(X_{p})
 \sum_{s=1}^{K}\frac{m_{s}}{s!}
 \left(\frac{\partial_{y_{l}}^{s}X_{p}}{s!}\right)^{m_{s}-1}
-\\
-&& \times\left[\partial_{y_{j}}\partial_{y_{l}}^{s}X_{p}\right]
+\left[\partial_{y_{j}}\partial_{y_{l}}^{s}X_{p}\right]
 \prod_{k\le K,\,k\neq s}\left(\frac{\partial_{y_{l}}^{k}X_{p}}{k!}\right)^{m_{k}}.
-\end{eqnarray*}
+\end{align*}
 Using the bounds of
@@
 we obtain
-\begin{eqnarray}
-\label{eq:part-part}
+\[
 \left|\partial_{y_{j}}\partial_{y_{l}}^{K}f\left(X_{p}\right)\right|
 \leq A_{K}\prod_{k=l+1}^{p}\alpha_{k}
-\end{eqnarray}
+\]
 with a suitable constant $A_{K}$.
@@
 \\
-%&=
-%\frac{A_{K}^{2}}{(1-\gamma_1 b_\mu)^2}
-%\sum_{j=l}^{q} \gamma_{j}
-%\prod_{k=l+1}^{j} \alpha_{k}^2
-%\left(
-%\sum_{p=j+1}^{q+1} \gamma_{p} \prod_{k=j+1}^{p} \alpha_{k}
-%\right)^{2}
-%\\
+&=
+\frac{A_{K}^{2}}{(1-\gamma_1 b_\mu)^2}
+\sum_{j=l}^{q} \gamma_{j}
+\prod_{k=l+1}^{j} \alpha_{k}^2
+\left(
+\sum_{p=j+1}^{q+1} \gamma_{p} \prod_{k=j+1}^{p} \alpha_{k}
+\right)^{2}
+\\
 &\le
@@
 \end{proof}
-\begin{remark}\label{rem:p-norm}
-In fact in the proof of Lemma~\ref{lem:var_poincare} using \eqref{eq:part-part} we have shown that the expectations
-\begin{eqnarray*}
-\mathsf E\left[\left.
-\|\nabla_Z F(X_{l-1},Z_l,Z_{l+1},\ldots,Z_q)\|^r
-\right|X_{l-1}\right]
-\end{eqnarray*}
-are uniformly bounded for all natural \(l\leq q\) and any \(r\geq 2.\) Indeed it follows from \eqref{eq:17042018a1} that  the quantities
-\begin{eqnarray*}
-\left(\sum_{j=l}^{q} \gamma_{j}
-\prod_{k=l}^{j} \alpha_{k}\right)^{r/2}
-\left(
-\sum_{p=j+1}^{q+1} \gamma_{p} \prod_{k=j+1}^{p} \alpha_{k}
-\right)^{r}
-\end{eqnarray*}
-are bounded.
-This stronger result in combination with a \(L^r\)-Sobolev-type inequalities (see e.g.
-\cite{Adamczak}) implies that also higher order conditional (on \(X_{l-1}\)) central moments of the r.v.
-\begin{eqnarray*}
-\sum_{p=l}^{q}\gamma_{p+1}\partial_{y_{l}}^{K}f\left(X_{p}\right)
-\end{eqnarray*}
-are uniformly bounded in \(l\leq q\) and \(p.\)
-
-\end{remark}
 
 \subsection{Proof of Lemma~\ref{a_decay}}
-
-We have for any \(\mathbf{k}\neq \mathbf{0}\)
-\begin{eqnarray*}
+We have for any \(\mathbf{k}\neq \mathbf{0}\) and $x \in \rset^d$, we get
+\[
 a_{p,l,\mathbf{k}}(x)=\mathsf{E}\left[\mathbf{H}_\mathbf{k}\left(Z_l\right)[Q_{p,l}\left(\Phi_l(x,Z_l)\right)-Q_{p,l}\left(\Phi_l(x,0)\right)]\right].
-\end{eqnarray*}
-Hence
-\begin{eqnarray*}
-\|a_{p,l,\mathbf{k}}\|_{\infty}\leq \sqrt{\gamma_l}\|\nabla Q_{p,l}\|_{\infty}.
-\end{eqnarray*}
+\]
+showing that $\|a_{p,l,\mathbf{k}}\|_{\infty}\leq \sqrt{\gamma_l}\| \nabla Q_{p,l}\|_{\infty}$.
 Since \(f,\mu\in C^1(\mathbb{R}^d),\)  we have
-\begin{eqnarray*}
-\nabla Q_{p,l}(x)=\mathsf{E}\left[(I-\gamma_{l+1}\nabla\mu(x))\nabla Q_{p,l+1}(x-\gamma_{l+1}\mu(x)+\sqrt{\gamma_{l+1}}\xi)\right].
-\end{eqnarray*}
-Hence
-\begin{eqnarray*}
-\|\nabla Q_{p,l}\|_\infty\leq (1-b_{\mu}\gamma_{l+1})\|\nabla Q_{p,l+1}\|_\infty
-\end{eqnarray*}
-and
-\begin{eqnarray*}
-\|\nabla Q_{p,l}\|_\infty&\leq & \|\nabla f\|_\infty\prod_{r=l+1}^p (1-b_{\mu}\gamma_{r})
-\\
-&\leq & \|\nabla f\|_\infty\exp\left(-b_{\mu}\sum_{r=l+1}^p \gamma_{r}  \right).
-\end{eqnarray*}
-As a result
-\begin{eqnarray*}
-\|a_{p,l,\mathbf{k}}\|_{\infty}\leq \sqrt{\gamma_l}\, \|\nabla f\|_\infty\exp\left(-b_{\mu}\sum_{r=l+1}^p \gamma_{r}  \right).
-\end{eqnarray*}
-
-\bibliographystyle{plain}
+\[
+\nabla Q_{p,l}(x)=\mathsf{E}\left[(\Id-\gamma_{l+1} \jacob{\mu}{x}) \nabla Q_{p,l+1}(x-\gamma_{l+1}\mu(x)+\sqrt{\gamma_{l+1}}\xi)\right] \, .
+\]
+Hence we have $\| \nabla Q_{p,l}\|_\infty\leq (1-b_{\mu}\gamma_{l+1})\| \nabla Q_{p,l+1}\|_\infty$,
+which implies that
+\[
+\| \nabla Q_{p,l}\|_\infty \leq  \| \nabla f\|_\infty\prod_{r=l+1}^p (1-b_{\mu}\gamma_{r})
+\leq  \| \nabla f\|_\infty\exp\left(-b_{\mu}\sum_{r=l+1}^p \gamma_{r}  \right) \,
+\]
+which concludes the proof.
+
+\bibliographystyle{abbrv}
 \bibliography{refs-1}

### 2020-04-15 aap/Revision/vr-mcmc-revision-new-new.tex

@@
 \end{equation}
-with   $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = 25K_2^2/8 + d + m$ with $K_2$ from \eqref{eq:H3}, $m$ from {\bf (H2)}.
+with   $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = 25K_2^2/8 + d + m$ with $K_2$ from Lemma~\ref{lem:quadratic_behaviour}, and $m$ from {\bf (H2)}.
 \end{lem}
@@
 \begin{proof} 
-$$
-Q_{\gamma}V(x) = \int\limits_{\rset^d}V(y)Q_{\gamma}(x,dy) = \int\limits_{\rset^d}\frac{(1+\|y\|^2)}{(2\pi\gamma)^{\frac{d}{2}}}\exp{\left(-\frac{\|y-x+\gamma\nabla U(x)\|^2}{2\gamma}\right)}\,dy =
-$$
-
-$$
-=1 + \frac{1}{(2\pi\gamma)^{\frac{d}{2}}}\int\limits_{\mathbb{R}^d}\|z + x - \gamma \nabla U(x)\|^2 \exp{\left(-\frac{\|z\|^2}{2\gamma}\right)}\,dz
-$$
-Let us first consider the case $x \notin B(0,K_2)$. Note that
-$$
-\|z+x-\gamma \nabla U(x)\|^2 = \|z\|^2 + 2 \langle z, x - \gamma \nabla U(x) \rangle + \|x - \gamma \nabla U(x)\|^2
-$$
-and the linear term vanishes after integration, moreover, for $Z \sim \mathcal{N}(0,\gamma I_d)$, it holds $\mathbb{E}\|Z\|^2 = \gamma d$. It remains to notice that due to \eqref{eq:H3},
+Note first that
+\[
+Q_{\gamma}V(x) = 
+1+ \| x - \gamma \nabla U(x) \|^2 + \gamma d
+\]
+Let us first consider the case $x \notin B(0,K_2)$.  
+It remains to notice that, by Lemma~\ref{lem:quadratic_behaviour}, we get
 \[
@@
 $$
-Now let $x \in B(0,K_2)$. Then simply using $\|x-\gamma \nabla U(x)\|^2 \leq 2(1+L\gamma)^2\|x\|^2$, we obtain
+Consider now the case $x \in B(0,K_2)$. Then simply using $\|x-\gamma \nabla U(x)\|^2 \leq 2(1+L\gamma)^2\|x\|^2$, we obtain
 $$

### 2020-04-23 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 Hence \eqref{eq:vanilla_var} follows.
+\end{proof}
+
+\begin{lem}
+\label{lem:exp_strongly_convex}
+Let $(X_p^x)_{p \geq 0}$ be the ULA-generated Markov chain with ... . Then for any $\lambda \in \rset$, it holds
+\begin{equation}
+\label{eq:exp_strongly_convex}
+\PE_x \rme^{\lambda\left(\pi_n^N(f) - \PE_x\pi_n^N(f)\right)} \leq e^{\frac{4\lambda^2 \|f\|^2_{\text{Lip}}\left(1+\frac{N}{n}\right)d(m+L)^2}{\gamma n m^2L^2}}
+\end{equation}
+\end{lem}
+\begin{proof} Note that for ULA algorithm the distribution $\P_x$ is Gaussian for any $x$, thus, following \cite[Proposition 3]{dm2016HighDim}, we may bound the Wasserstein-1 distance between $\P_x$ and $\P_y$ by
+\[
+W_1(\P_x,\P_y) = W_1(\mathcal{N}(x-\gamma\mu(x),\gamma),\mathcal{N}(y-\gamma\mu(y),\gamma)) = \|x-y-\gamma(\mu(x)-\mu(y))\| \leq \sqrt{1-\kappa}\|x-y\|
+\]
+where $\kappa = \frac{2mL\gamma}{m+L}$. Hence, it holds
+\[
+W_1(\P_x,\P_y) \leq \left(1-\frac{\kappa}{2}\right)\|x-y\|
+\]
+Thus, we might apply the result of \cite[Theorem 4]{joulin2010}, yielding for any $\lambda \in \rset$  
+\[
+\PE_x \rme^{\lambda\left(\pi_n^N(f) - \PE_x\pi_n^N(f)\right)} \leq e^{4\lambda^2 \|f\|^2_{\text{Lip}} V^2}
+\]
+with the quantity $V^2 = \frac{4}{\kappa n}\left(1 + \frac{N}{n}\right)\sup_{x}\frac{\sigma^2(x)}{n_x \kappa}$, $\sigma^2(x) = \frac{1}{2}\iint \|y-z\|^2P_x(dy)P_x(dz) = \gamma d$ and $n_x \geq 1$.
+\end{proof}
+
+\begin{lem} Let $\mu(x) = \mu_{0}(x) + \Delta(x)$, where $\mu_0^{\prime}(x) <-m$ for some $m > 0$, $\mu_0(x) \in Lip(L)$ and suppose that $\Delta(x)$ is bounded. Then
+\begin{equation}
+\label{eq:exp_bound}
+\PE_x \rme^{u\gamma\sum\limits_{k=N+1}^{N+n}f(X_k)} \leq C\exp\left\{n\gamma\left(\frac{8u^2\|f\|^2_{\text{Lip}}\left(1+\frac{N}{n}\right)(m+L)^2}{m^2L^2} + M^2 + C_1u\sqrt{\gamma} +  u\pi(f)\right)\right\}
+\end{equation}
+with constants $C$ and $C_1$ depending only on $m,L$.
+\end{lem}
+\begin{proof} The Markov Chain $(X_p^x)$ has the form
+\[
+X_{k+1} = X_k + \gamma(\mu_0(X_k) + \Delta(X_k)) + \sqrt{\gamma}\xi_{k+1}, X_0 = x, \xi_{k+1} \sim \mathcal{N}(0,1)
+\]
+Consider the following auxiliary Markov Chain $(Y_p^x)$, corresponding to the diffusion with strongly log-concave potential
+\[
+Y_{k+1} = Y_k + \gamma\mu_0(Y_k) + \sqrt{\gamma}\xi_{k+1}, Y_0 = x, \xi_{k+1} \sim \mathcal{N}(0,1)
+\]
+The main idea of the proof is to change expectation w.r.t. $(X_p^x)$ to the expectation w.r.t. $(Y_p^x)$, which is easier to work with:
+\begin{eqnarray*}
+\PE_x \rme^{u\gamma\sum\limits_{k=N+1}^{N+n}f(X_k)} &=& \PE_x  \exp\left\{u\gamma\sum_{k=N+1}^{N+n}\left(f(Y_k) -\frac{(Y_k - Y_{k-1} - \gamma\mu(Y_{k-1}))^2 - (Y_k - Y_{k-1} - \gamma\mu_0(Y_{k-1}))^2}{2\gamma}\right)\right\} = \\
+& = & \PE_x \exp\left\{u\gamma\sum_{k=N+1}^{N+n}f(Y_k) + \sum_{k=1}^{n}\left[(Y_k-Y_{k-1})\Delta(Y_{k-1}) - \gamma\mu_0(Y_{k-1})\Delta(Y_{k-1}) - \frac{\gamma}{2}\Delta^2(Y_{k-1})\right]\right\} = \\
+ &=& \PE_x \exp\left\{u\gamma\sum_{k=N+1}^{N+n}f(Y_k) + \sum_{k=1}^{n}\left[\sqrt{\gamma}\Delta(Y_{k-1})\xi_{k} - \frac{\gamma}{2}\Delta^2(Y_{k-1})\right]\right\} \leq \\
+ & \leq & \left(\PE_x \exp\left\{2\sqrt{\gamma}\sum_{k=N+1}^{N+n}\Delta(Y_{k-1})\xi_{k}\right\}\right)^{1/2} \left(\PE \exp\left\{2u\gamma\sum_{k=1}^{n}f(Y_k)\right\}\right)^{1/2}
+\end{eqnarray*}
+with the last line obtained from Holder inequality. We bound the first expectation using Azuma-Hoeffding inequality, that is, setting $\mathcal{F}_k = \sigma(\xi_1,\ldots,\xi_k)$, 
+\[
+\PE_x \exp\left\{2\sqrt{\gamma}\sum_{k=N+1}^{N+n}\Delta(Y_{k-1})\xi_{k}\right\} = \PE_x\left(\exp\left\{2\sqrt{\gamma}\sum_{k=N+1}^{N+n-1}\Delta(Y_{k-1})\xi_{k}\right\} \PE\left[\exp\left\{2\sqrt{\gamma}\Delta(Y_{n-1})\xi_{n}\right\}| \mathcal{F}_{N+n-1}\right]\right) \leq
+\]
+\[
+\leq \PE_x\left(\exp\left\{2\sqrt{\gamma}\sum_{k=N+1}^{N+n-1}\Delta(Y_{k-1})\xi_{k}\right\}\exp\left\{2\gamma M^2\right\}\right) \leq \{\text{by induction}\} \leq \exp\left\{2n\gamma M^2\right\}
+\]
+To bound the second exponent, note that
+\[
+\PE \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}f(Y_k)\right\} =
+\PE \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}\left(f(Y_k) - \PE_xf(X_k)\right)\right\} \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}\PE_x f(Y_k)\right\} 
+\]
+Now we can apply Lemma~\ref{lem:exp_strongly_convex} to the first expectation with $\lambda = 2nu\gamma$, yielding the bound
+\begin{equation}
+\label{eq:1_bound}
+\PE \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}\left(f(Y_k) - \PE_xf(X_k)\right)\right\} \leq \exp\left\{\frac{16u^2 n\gamma \|f\|^2_{\text{Lip}}\left(1+\frac{N}{n}\right)(m+L)^2}{m^2L^2}\right\}
+\end{equation}
+Next we use \cite[Proposition 1]{joulin2010}, stating that
+\begin{equation}
+\label{eq:2_bound}
+\left|\sum_{k=N+1}^{N+n}\PE_x f(Y_k) - n\pi_{\gamma}(f)\right| \leq \frac{2\left(1-\frac{\kappa}{2}\right)^{N+1}}{\kappa}\|f\|_{\text{Lip}}\int_{\rset}|x-y|\,\pi_{\gamma}(dy) \leq \frac{m+L}{mL}\|f\|_{\text{Lip}}S(x)
+\end{equation}
+and \cite[Corollary 7]{dm2016HighDim}, that is
+\begin{equation}
+\label{eq:3_bound}
+|\pi_{\gamma}(f) - \pi(f)| \leq W_1(\pi_\gamma,\pi) \leq W_2(\pi_\gamma,\pi) \leq C_1\sqrt{\gamma}
+\end{equation}
+where $C_1 = 2L\sqrt{\frac{mL}{m+L}\left(\gamma + \frac{mL}{2(m+L)}\right)\left(2 + \frac{L^2\gamma}{m} + \frac{L^2\gamma^2}{6}\right)}$.
+Now we put together bounds \eqref{eq:1_bound}, \eqref{eq:2_bound} and \eqref{eq:3_bound}, leading to the final bound
+\[
+\PE \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}f(Y_k)\right\} \leq \exp\left\{\frac{16u^2 n\gamma \|f\|^2_{\text{Lip}}\left(1+\frac{N}{n}\right)(m+L)^2}{m^2L^2} + 2u\gamma \frac{m+L}{mL}\|f\|_{\text{Lip}}S(x) + 2n\gamma u(\pi(f) + C_1\sqrt{\gamma})\right\}
+\]
+Hence the statement of lemma follows.
 \end{proof}

### 2020-04-24 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 
-\begin{lem} Let $\mu(x) = \mu_{0}(x) + \Delta(x)$, where $\mu_0^{\prime}(x) <-m$ for some $m > 0$, $\mu_0(x) \in Lip(L)$ and suppose that $\Delta(x)$ is bounded. Then
+\begin{lem} Let $(X_p)_{p \geq 0}$ be the ULA-generated Markov chain of the form
+\[
+X_{p+1} = X_p + \gamma \mu(X_p) + \sqrt{\gamma}\xi_{p+1}, \xi_{p+1} \sim \mathcal{N}(0,1)
+\]
+ with $\mu(x) = \mu_{0}(x) + \Delta(x)$, where $\mu_0^{\prime}(x) <-m$ for some $m > 0$, $\mu_0(x) \in Lip(L)$ and suppose that $\exists M > 0, \left|\Delta(x)\right| < M$.
+Then for any Lipschitz function $f$,
 \begin{equation}

### 2020-04-25 aap/Revision/vr-mcmc-revision-new-new.tex
update

--- 2020-04-23 aap/Revision/vr-mcmc-revision-new-new.tex
@@
 \]
-Note that $\mu^{\prime}(x)$ is bounded due  our assumptions and from \cite[Theorem~1]{delyon1999small} and Lemma~\ref{lem:v_ergodicity} in Appendix~\ref{sec:appendix} it follows that
-\[
-\mathsf{E}\left[\rme^{-u\gamma\left(\sum_{k=l}^{j}[\mu'(X^x_{k-1})-\pi_\gamma(\mu')]\right)}\right]\leq C_1 \rme^{\varkappa n\gamma^2}(1+V(x)),\quad u\gamma\leq \gamma_0,
-\]
-for some constants $\gamma_0$ and $C_1,\varkappa>0$. Then we have
-\[
-\zeta_{l,j}(u)=\mathsf{E}\left[\rme^{-s\gamma\sum_{k=l}^{j}\mu'(X^x_{k-1})}\right]\leq  C_1 \rme^{\varkappa n\gamma^2}(1+V(x)) \rme^{-s\gamma (j-l+1)\pi_\gamma(\mu')}.
-\]
-Furthermore, since $\mu\pi',\mu'\pi\in L^{1}(\mathbb{R})$ and $\pi^{\prime}(x) = -\frac{1}{2}\pi(x)\mu(x)$, we have
+Note that $\mu^{\prime}(x)$ is bounded and Lipschitz due  our assumptions and from Lemma~\ref{lem:exp_bound} in Appendix~\ref{sec:appendix} it follows that
+\begin{align*}
+\zeta_{l,j}(u) \leq \begin{cases} C(x)\exp\left\{(j-l)\gamma\left(4u^2\kappa^{-2} \|f\|^2_{\text{Lip}} + M^2 + C_1u\sqrt{\gamma} -  u\pi(\mu')\right)\right\}, \quad \gamma(j-l) \geq C_0 := \left(\frac{4}{m+L} + \kappa^{-1}\right) \\
+\exp\left\{uC_0\max\limits_{y}|\mu'(y)|\right\}, \quad 0 \leq \gamma(j-l) \leq C_0	
+\end{cases}
+\end{align*}
+for constants $C(x)$ and $C_1$ which could be made precise. Furthermore, since $\mu\pi',\mu'\pi\in L^{1}(\mathbb{R})$ and $\pi^{\prime}(x) = -\frac{1}{2}\pi(x)\mu(x)$, we have
 \[
@@
 \]
-Note also that \(\pi_\gamma(\mu^{\prime}) \geq \pi_\gamma(\mu^{\prime}) - C_2\sqrt{\gamma}\)
-yielding the bound
-\[
-\zeta_{l,j}(s) \leq  C_1 \rme^{\varkappa n\gamma^2}(1+V(x)) \rme^{-s\gamma(j-l+1)\alpha}
-\]
-where $ \alpha = \frac{1}{2}\int \mu^2(x)\pi(x)\,dx - C_1\sqrt{\gamma} > 0$ for sufficiently small $\gamma$. Hence we obtain
+Hence the following bound follows
 \begin{align*}
-\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4) &\leq (C_1 \rme^{\varkappa n\gamma^2}(1+V(x)))^{1/4}\sum_{p=j+1}^{q+1} \rme^{-\gamma(p-j)\alpha}
-\\
-&\leq  (C_1 \rme^{\varkappa n\gamma^2}(1+V(x)))^{1/4}\frac{1}{1- \rme^{- \gamma\alpha}}
+\zeta_{l,j}(u) \leq \begin{cases} C(x)\exp\left\{-\gamma(j-l)\alpha\right\}, \quad \gamma(j-l) \geq C_0 \\
+\exp\left\{uC_0\max\limits_{y}|\mu'(y)|\right\}, \quad 0 \leq \gamma(j-l) \leq C_0	
+\end{cases}
+\end{align*}
+with the quantity $ \alpha = \min(...,...) > 0$. Hence we obtain
+\begin{align*}
+\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4) &\leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + C(x)\sum_{p=j+1}^{\infty} \rme^{-\gamma(p-j)\alpha}
+\leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + \frac{C(x)}{1-\rme^{- \gamma\alpha}}
 \end{align*}
@@
 \begin{align*}
-\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \leq (C_1 \rme^{\varkappa n\gamma^2}(1+V(x)))^{1/2}\sum_{j=0}^{q} \rme^{-\gamma j\alpha}
-\\
-\leq  (C_1 \rme^{\varkappa n\gamma^2}(1+V(x)))^{1/2}\frac{ 1}{1- \rme^{- \gamma\alpha}}.
+\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + \frac{C(x)}{1-\rme^{- \gamma\alpha}}
 \end{align*}
@@
 \begin{align*}
-\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X^x_{p}\right)\right] \leq \gamma^{-1} C_K \rme^{\varkappa n\gamma^2} (1+V(x))
+\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X^x_{p}\right)\right] \leq \gamma^{-1} C_K C(x)
 \end{align*}
@@
 
-\begin{lem} Let $\mu(x) = \mu_{0}(x) + \Delta(x)$, where $\mu_0^{\prime}(x) <-m$ for some $m > 0$, $\mu_0(x) \in Lip(L)$ and suppose that $\Delta(x)$ is bounded. Then
+\begin{lem} 
+\label{lem:exp_bound}
+Let $(X_p)_{p \geq 0}$ be the ULA-generated Markov chain of the form
+\[
+X_{p+1} = X_p + \gamma \mu(X_p) + \sqrt{\gamma}\xi_{p+1}, \xi_{p+1} \sim \mathcal{N}(0,1)
+\]
+ with $0 < \gamma < \frac{4}{m+L}, \quad \mu(x) = \mu_{0}(x) + \Delta(x)$, where $\mu_0^{\prime}(x) <-m$ for some $m > 0$, $\mu_0(x) \in Lip(L)$.
+Suppose that there exists $M > 0,$ such that $\left|\Delta(x)\right| \leq M$.
+ Then for any Lipschitz function $f$, $\forall N $
 \begin{equation}
 \label{eq:exp_bound}
-\PE_x \rme^{u\gamma\sum\limits_{k=N+1}^{N+n}f(X_k)} \leq C\exp\left\{n\gamma\left(\frac{8u^2\|f\|^2_{\text{Lip}}\left(1+\frac{N}{n}\right)(m+L)^2}{m^2L^2} + M^2 + C_1u\sqrt{\gamma} +  u\pi(f)\right)\right\}
-\end{equation}
-with constants $C$ and $C_1$ depending only on $m,L$.
+\PE_x \rme^{u\gamma\sum\limits_{k=N+1}^{N+n}f(X_k)} \leq C\exp\left\{n\gamma\left(2u^2\kappa^{-2} \|f\|^2_{\text{Lip}}\left(1+\frac{1}{n\gamma}\left(\frac{4}{m+L} + \kappa^{-1}\right)\right) + M^2 + C_1u\sqrt{\gamma} +  u\pi(f)\right)\right\}
+\end{equation}
+with positive constants $C < \exp\left\{\frac{4u\|f\|_{\text{Lip}}}{mL}\int_{\rset}|x-y|\,\pi_{\gamma}(dy)\right\}$ and $C_1$ depending only on $m,L$.
 \end{lem}
@@
 \PE_x \rme^{u\gamma\sum\limits_{k=N+1}^{N+n}f(X_k)} &=& \PE_x  \exp\left\{u\gamma\sum_{k=N+1}^{N+n}\left(f(Y_k) -\frac{(Y_k - Y_{k-1} - \gamma\mu(Y_{k-1}))^2 - (Y_k - Y_{k-1} - \gamma\mu_0(Y_{k-1}))^2}{2\gamma}\right)\right\} = \\
-& = & \PE_x \exp\left\{u\gamma\sum_{k=N+1}^{N+n}f(Y_k) + \sum_{k=1}^{n}\left[(Y_k-Y_{k-1})\Delta(Y_{k-1}) - \gamma\mu_0(Y_{k-1})\Delta(Y_{k-1}) - \frac{\gamma}{2}\Delta^2(Y_{k-1})\right]\right\} = \\
- &=& \PE_x \exp\left\{u\gamma\sum_{k=N+1}^{N+n}f(Y_k) + \sum_{k=1}^{n}\left[\sqrt{\gamma}\Delta(Y_{k-1})\xi_{k} - \frac{\gamma}{2}\Delta^2(Y_{k-1})\right]\right\} \leq \\
- & \leq & \left(\PE_x \exp\left\{2\sqrt{\gamma}\sum_{k=N+1}^{N+n}\Delta(Y_{k-1})\xi_{k}\right\}\right)^{1/2} \left(\PE \exp\left\{2u\gamma\sum_{k=1}^{n}f(Y_k)\right\}\right)^{1/2}
+& = & \PE_x \exp\left\{u\gamma\sum_{k=N+1}^{N+n}f(Y_k) + \sum_{k=N+1}^{N+n}\left[(Y_k-Y_{k-1})\Delta(Y_{k-1}) - \gamma\mu_0(Y_{k-1})\Delta(Y_{k-1}) - \frac{\gamma}{2}\Delta^2(Y_{k-1})\right]\right\} = \\
+ &=& \PE_x \exp\left\{u\gamma\sum_{k=N+1}^{N+n}f(Y_k) + \sum_{k=N+1}^{N+n}\left[\sqrt{\gamma}\Delta(Y_{k-1})\xi_{k} - \frac{\gamma}{2}\Delta^2(Y_{k-1})\right]\right\} \leq \\
+ & \leq & \left(\PE_x \exp\left\{2\sqrt{\gamma}\sum_{k=N+1}^{N+n}\Delta(Y_{k-1})\xi_{k}\right\}\right)^{1/2} \left(\PE \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}f(Y_k)\right\}\right)^{1/2}
 \end{eqnarray*}
@@
 \PE \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}f(Y_k)\right\} =
-\PE \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}\left(f(Y_k) - \PE_xf(X_k)\right)\right\} \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}\PE_x f(Y_k)\right\} 
-\]
-Now we can apply Lemma~\ref{lem:exp_strongly_convex} to the first expectation with $\lambda = 2nu\gamma$, yielding the bound
+\PE \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}\left(f(Y_k) - \PE_xf(Y_k)\right)\right\} \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}\PE_x f(Y_k)\right\} 
+\]
+Now we can apply \cite[Proposition 32]{dm2016HighDim} to the first expectation with $\lambda = 2nu\gamma$, yielding the bound
 \begin{equation}
 \label{eq:1_bound}
-\PE \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}\left(f(Y_k) - \PE_xf(X_k)\right)\right\} \leq \exp\left\{\frac{16u^2 n\gamma \|f\|^2_{\text{Lip}}\left(1+\frac{N}{n}\right)(m+L)^2}{m^2L^2}\right\}
+\PE \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}\left(f(Y_k) - \PE_xf(Y_k)\right)\right\} \leq \exp\left\{4u^2\kappa^{-2} n\gamma \|f\|^2_{\text{Lip}}\left(1+\frac{1}{n\gamma}\left(\frac{4}{m+L} + \kappa^{-1}\right)\right)\right\}
 \end{equation}
@@
 \[
-\PE \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}f(Y_k)\right\} \leq \exp\left\{\frac{16u^2 n\gamma \|f\|^2_{\text{Lip}}\left(1+\frac{N}{n}\right)(m+L)^2}{m^2L^2} + 2u\gamma \frac{m+L}{mL}\|f\|_{\text{Lip}}S(x) + 2n\gamma u(\pi(f) + C_1\sqrt{\gamma})\right\}
+\PE \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}f(Y_k)\right\} \leq \exp\left\{4u^2\kappa^{-2} n\gamma \|f\|^2_{\text{Lip}}\left(1+\frac{1}{n\gamma}\left(\frac{4}{m+L} + \kappa^{-1}\right)\right) + 2u\gamma \frac{m+L}{mL}\|f\|_{\text{Lip}}S(x) + 2n\gamma u(\pi(f) + C_1\sqrt{\gamma})\right\}
 \]

### 2020-04-30 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 \]
-Note that $\mu^{\prime}(x)$ is bounded and Lipschitz due  our assumptions and from Lemma~\ref{lem:exp_bound} in Appendix~\ref{sec:appendix} it follows that
+Note that due to our assumptios, $\mu(x) = \mu_0(x) + \Delta(x)$ with $\mu^{\prime}_0(x) < -m$. Thus by applying Holder inequality to $\zeta_{l,j}(u)$, we obtain
+\begin{equation}
+\label{eq:zeta}
+\zeta_{l,j}(u) \leq \left(\mathsf{E}\left[\rme^{-2u\gamma\sum_{k=l}^{j}\mu_0'(X^x_{k-1})}\right]\right)^{1/2}\left(\mathsf{E}\left[\rme^{-2u\gamma\sum_{k=l}^{j}\Delta'(X^x_{k-1})}\right]\right)^{1/2}
+\end{equation}
+Note that $\left|\Delta^{\prime}(x)\right| \leq M^{\prime}$, hence
+\[
+\left(\mathsf{E}\left[\rme^{-2u\gamma\sum_{k=l}^{j}\Delta'(X^x_{k-1})}\right]\right)^{1/2} \leq \mathsf{E}\rme^{2u(j-l)M^{\prime}}
+\]
+To bound the first exponential moment in \ref{eq:zeta}, we apply
+Lemma~\ref{lem:exp_bound} in Appendix~\ref{sec:appendix}. It implies that
 \begin{align*}
-\zeta_{l,j}(u) \leq \begin{cases} C(x)\exp\left\{(j-l)\gamma\left(4u^2\kappa^{-2} \|f\|^2_{\text{Lip}} + M^2 + C_1u\sqrt{\gamma} -  u\pi(\mu')\right)\right\}, \quad \gamma(j-l) \geq C_0 := \left(\frac{4}{m+L} + \kappa^{-1}\right) \\
+\zeta_{l,j}(u) \leq \begin{cases} C(x)\exp\left\{(j-l)\gamma\left(4u^2\kappa^{-2} \|\mu^{\prime}\|^2_{\text{Lip}} + M^2 + M^{\prime} + C_1u\sqrt{\gamma} -  u\pi(\mu')\right)\right\}, \quad \gamma(j-l) \geq C_0 \\
 \exp\left\{uC_0\max\limits_{y}|\mu'(y)|\right\}, \quad 0 \leq \gamma(j-l) \leq C_0	
@@
 \end{align*}
-for constants $C(x)$ and $C_1$ which could be made precise. Furthermore, since $\mu\pi',\mu'\pi\in L^{1}(\mathbb{R})$ and $\pi^{\prime}(x) = -\frac{1}{2}\pi(x)\mu(x)$, we have
+with $C_0 = \left(\frac{4}{m+L} + \kappa^{-1}\right)$, and constants $C(x)$ and $C_1$ given in Lemma~\ref{lem:exp_bound} .
+Furthermore, since $\mu\pi',\mu'\pi\in L^{1}(\mathbb{R})$ and $\pi^{\prime}(x) = -\frac{1}{2}\pi(x)\mu(x)$, we have
 \[
@@
 \appendix
-\section{Bounds for moments of  ULA}\label{sec:appendix}
+\section{Geometric ergodicity of the Unadjusted Langevin Algorithm}\label{sec:appendix}
 
@@
 \label{lem:v_ergodicity}
- Assume  {\bf (H1)} and {\bf (H2)}. Then for $0 < \gamma < \overline{\gamma}=m / 4L^2$, for any $x \in \rset^d$ it holds
+ Assume  {\bf (H1)} and {\bf (H2)}. Then for $0 < \gamma < \overline{\gamma}= \min(m / 4L^2,1/L)$, for $x \in \rset^d: \|x\| \leq K$  it holds
 $$
@@
 $$
-C = \left(1 + \exp{\left(-\frac{m\gamma}{2}\right)}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp{\left(-\frac{m\gamma}{2}\right)} - \frac{2b}{1+d})}\right);
-$$
-
-$$
-b = \gamma(\frac{25K_1^2}{8} + 2d + m); \quad \overline{b} = b\exp{\left(-\frac{m\gamma}{2}\right)} + d; \quad \varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right); \quad
-$$
-
-$$
-\rho = \exp{\left(-\gamma\left(\frac{m}{2} - 2\frac{\frac{25K_1^2}{8} + 2d + m}{d}\right)\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{m\gamma}{2}\right)} + \frac{2b}{d+1})}}\right)}
+C = \left(1 + \exp{\left(-\frac{m\gamma}{2}\right)}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp{\left(-\frac{m\gamma}{2}\right)} - \frac{2b}{2+K^2})}\right);
+$$
+
+$$
+b = \gamma(\frac{25K_1^2}{8} + 2d + m); \quad \overline{b} = b\exp{\left(-\frac{m\gamma}{2}\right)} + K^2 + 1; \quad \varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right); \quad
+$$
+
+$$
+\rho = \exp{\left(-\gamma\left(\frac{m}{2} - 2\frac{\frac{25K_1^2}{8} + 2d + m}{K^2 + 1}\right)\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{m\gamma}{2}\right)} + \frac{2b}{K^2+2})}}\right)}
 $$
@@
 
-\begin{proof} Note that the condition {\bf (H1)} implies that the Markov kernel $Q^{1,n}_{\boldsymbol{\gamma}}$ satisfies $(1,\varepsilon)$-Doeblin condition with $\varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right)$.
+\begin{proof} Note that the condition {\bf (H1)} implies that for any $x,y \in \rset^d$,
+\[
+\|x-y-\gamma(\nabla U(x) - \nabla U(y))\|^2 \leq (1 + 2L\gamma + \gamma^2)\|x-y\|^2
+\] 
+Hence, the result of [Corollary 5]\cite{DurmusBortoli} implies that
+\[
+\|\delta_xQ^{[1/\gamma]}_{\gamma} - \delta_yQ^{[1/\gamma]}_{\gamma}\|_{\text{TV}} \leq 2(1-\varepsilon)
+\]
+with $\varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right)$ for all $x,y$ such that $\|x\|,\|y\| \leq K$, which means that the kernel $Q_{\gamma}$ satisfies $([1/\gamma],\varepsilon)$-Doeblin codition on the set $\{x: V(x) < 1+K^2\}$.
 Together with drift condition \eqref{eq:lyapunov} it allows to apply \cite[Theorem~19.4.1]{moulines2018} with appropriate constants.
@@
 \]
- with $0 < \gamma < \frac{4}{m+L}, \quad \mu(x) = \mu_{0}(x) + \Delta(x)$, where $\mu_0^{\prime}(x) <-m$ for some $m > 0$, $\mu_0(x) \in Lip(L)$.
+ with $0 < \gamma < 4(m+L)^{-1}, \quad \mu(x) = \mu_{0}(x) + \Delta(x)$, where $\mu_0^{\prime}(x) <-m$ for some $m > 0$, $\mu_0(x) \in Lip(L)$. Denote by 
+ \[
+ \pi(y) = \frac{\exp\left\{\int\limits_{0}^{y}\mu_{0}(x)\,dx\right\}}{\int\limits_{-\infty}^{+\infty}\exp\left\{\int\limits_{0}^{y}\mu_{0}(x)\,dx\right\}\,dy}
+ \]
+ the stationary distribution of the corresponding continuous-time diffusion.
  Suppose that there exists $M > 0,$ such that $\left|\Delta(x)\right| \leq M$.
@@
 \label{eq:exp_bound}
-\PE_x \rme^{u\gamma\sum\limits_{k=N+1}^{N+n}f(X_k)} \leq C\exp\left\{n\gamma\left(2u^2\kappa^{-2} \|f\|^2_{\text{Lip}}\left(1+\frac{1}{n\gamma}\left(\frac{4}{m+L} + \kappa^{-1}\right)\right) + M^2 + C_1u\sqrt{\gamma} +  u\pi(f)\right)\right\}
-\end{equation}
-with positive constants $C < \exp\left\{\frac{4u\|f\|_{\text{Lip}}}{mL}\int_{\rset}|x-y|\,\pi_{\gamma}(dy)\right\}$ and $C_1$ depending only on $m,L$.
+\PE_x \rme^{u\gamma\sum\limits_{k=N+1}^{N+n}f(X_k)} \leq C(x)\exp\left\{n\gamma\left(2u^2\kappa^{-2} \|f\|^2_{\text{Lip}}\left(1+\frac{1}{n\gamma}\left(\frac{4}{m+L} + \kappa^{-1}\right)\right) + M^2 + C_1u\sqrt{\gamma} +  u\pi(f)\right)\right\}
+\end{equation}
+with $\kappa = 2mL(m+L)^{-1}$, positive constant $C_1$ depending only on $m,L$ and 
+\[
+C(x) = \exp\left\{\frac{4u\|f\|_{\text{Lip}}}{mL}\int_{\rset}|x-y|\,\pi_{\gamma}(dy)\right\}
+\]
 \end{lem}

### 2020-05-02 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 \label{lem:drift}
- Assume that {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$.
- Then, for any $\gamma \in (0,\bar{\gamma}]$ with $\overline{\gamma} = m/4$  the kernel $Q_\gamma$ from \eqref{eq:ula_kernel} satisfies drift condition
+ Assume {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$.
+ Then, for any $\gamma \in (0,\bar{\gamma}]$ with $\overline{\gamma} = m/2L^2$  the kernel $Q_\gamma$ from \eqref{eq:ula_kernel} satisfies drift condition
  \begin{equation}
 \label{eq:lyapunov}
-Q_{\gamma}V(x) \leq \lambda^{\gamma} V(x) + \gamma C
-\end{equation}
-with   $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = 25K_2^2/8 + d + m$ with $K_2$ from Lemma~\ref{lem:quadratic_behaviour}, and $m$ from {\bf (H2)}.
+Q_{\gamma}V(x) \leq \rho^{\gamma} V(x) + \gamma C
+\end{equation}
+with   $\rho = \exp{\left(-\frac{m}{2}\right)}$, $C = 2K_2^2(L+m) + d + m$ with $K_2$ from Lemma~\ref{lem:quadratic_behaviour}, and $m$ from {\bf (H2)}.
 \end{lem}
@@
 \[
-\|x - \gamma \nabla U(x) \|^2 = \|x\|^2 - 2\gamma \langle \nabla U(x), x\rangle + \gamma^2 \|\nabla U(x)\|^2 \leq \left(1 - \gamma m + 2\gamma^2 L^2\right)\|x\|^2
-\]
-Since $\gamma < \frac{m}{4L}$, we obtain
-$$
-Q_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma d + (\gamma m - 2\gamma^2 L^2) \leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma (d + m)
-$$
-Consider now the case $x \in B(0,K_2)$. Then simply using $\|x-\gamma \nabla U(x)\|^2 \leq 2(1+L\gamma)^2\|x\|^2$, we obtain
-$$
-Q_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma\left((m - 2\gamma L^2)(1+\|x\|^2) + d + 2(1+L\gamma)^2\|x\|^2\right) \leq
-$$
-
-$$
-\leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma(\frac{25K_2^2}{8} + d + m)
+\|x - \gamma \nabla U(x) \|^2 = \|x\|^2 - 2\gamma \langle \nabla U(x), x\rangle + \gamma^2 \|\nabla U(x)\|^2 \leq \left(1 - \gamma m + \gamma^2 L^2\right)\|x\|^2
+\]
+Since $\gamma < \frac{m}{2L^2}$, we obtain
+$$
+Q_{\gamma}V(x) \leq (1-\gamma m + \gamma^2 L^2)V(x) + \gamma (d + m - \gamma L^2) \leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma (d + m)
+$$
+Consider now the case $x \in B(0,K_2)$. Then simply using $\|x-\gamma \nabla U(x)\|^2 \leq (1+L\gamma)^2\|x\|^2$, we obtain
+$$
+Q_{\gamma}V(x) \leq (1-\gamma m + \gamma^2 L^2)V(x) + \gamma\left((m - \gamma L^2)(1+\|x\|^2) + d + L(2+L\gamma)\|x\|^2\right) \leq
+$$
+
+$$
+\leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma(2K_2^2(L+m) + d + m)
 $$
@@
 \label{lem:v_ergodicity}
- Assume  {\bf (H1)} and {\bf (H2)}. Then for $0 < \gamma < \overline{\gamma}= \min(m / 4L^2,1/L)$, for $x \in \rset^d: \|x\| \leq K$  it holds
-$$
-d_V(\delta_xQ_{\gamma}^{1,n},\pi_\gamma) \leq C\rho^n\left(V(x) + \pi_\gamma(V)\right)
+ Assume  {\bf (H1)} and {\bf (H2)}. Then for $0 < \gamma < \overline{\gamma}= \min(m / 4L^2,1/L)$, and for any $x \in \rset^d$  it holds
+$$
+d_V(\delta_xQ_{\gamma}^{n},\pi_\gamma) \leq C\rho^{\gamma n}\left(V(x) + \pi_\gamma(V)\right)
 $$
@@
 $$
-b = \gamma(\frac{25K_1^2}{8} + 2d + m); \quad \overline{b} = b\exp{\left(-\frac{m\gamma}{2}\right)} + K^2 + 1; \quad \varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right); \quad
-$$
-
-$$
-\rho = \exp{\left(-\gamma\left(\frac{m}{2} - 2\frac{\frac{25K_1^2}{8} + 2d + m}{K^2 + 1}\right)\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{m\gamma}{2}\right)} + \frac{2b}{K^2+2})}}\right)}
-$$
+b = \gamma(2K_2^2(L+m) + d + m); \quad \overline{b} = b\exp{\left(-\frac{m\gamma}{2}\right)} + K^2 + 1; \quad \varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right); \quad
+$$
+
+$$
+\rho = \exp\left\{\left(\frac{m}{2} - 2\frac{2K_2^2(L+m) + d + m}{K^2 + 1}\right)\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{m\gamma}{2}\right)} + \frac{2b}{K^2+2})}}\right\}
+$$
+where $K_2 > 0$ is from Lemma~\ref{lem:quadratic_behaviour} and $K > 0$ is arbitrary
 \end{lem}
@@
 \]
-with $\varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right)$ for all $x,y$ such that $\|x\|,\|y\| \leq K$, which means that the kernel $Q_{\gamma}$ satisfies $([1/\gamma],\varepsilon)$-Doeblin codition on the set $\{x: V(x) < 1+K^2\}$.
+with $\varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right)$ for all $x,y$ such that $\|x\|,\|y\| \leq K$, which means that the kernel $Q_{\gamma}$ satisfies $([1/\gamma],\varepsilon)$-Doeblin condition on the set $\{x: V(x) < 1+K^2\}$.
 Together with drift condition \eqref{eq:lyapunov} it allows to apply \cite[Theorem~19.4.1]{moulines2018} with appropriate constants.

### 2020-05-02 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 \begin{lem} 
-\label{lem:exp_strongly_convex}
-Let $(X_p^x)_{p \geq 0}$ be the ULA-generated Markov chain with ... . Then for any $\lambda \in \rset$, it holds
-\begin{equation}
-\label{eq:exp_strongly_convex}
-\PE_x \rme^{\lambda\left(\pi_n^N(f) - \PE_x\pi_n^N(f)\right)} \leq e^{\frac{4\lambda^2 \|f\|^2_{\text{Lip}}\left(1+\frac{N}{n}\right)d(m+L)^2}{\gamma n m^2L^2}}
-\end{equation}
-\end{lem}
-\begin{proof} Note that for ULA algorithm the distribution $\P_x$ is Gaussian for any $x$, thus, following \cite[Proposition 3]{dm2016HighDim}, we may bound the Wasserstein-1 distance between $\P_x$ and $\P_y$ by
-\[
-W_1(\P_x,\P_y) = W_1(\mathcal{N}(x-\gamma\mu(x),\gamma),\mathcal{N}(y-\gamma\mu(y),\gamma)) = \|x-y-\gamma(\mu(x)-\mu(y))\| \leq \sqrt{1-\kappa}\|x-y\|
-\]
-where $\kappa = \frac{2mL\gamma}{m+L}$. Hence, it holds
-\[
-W_1(\P_x,\P_y) \leq \left(1-\frac{\kappa}{2}\right)\|x-y\|
-\]
-Thus, we might apply the result of \cite[Theorem 4]{joulin2010}, yielding for any $\lambda \in \rset$  
-\[
-\PE_x \rme^{\lambda\left(\pi_n^N(f) - \PE_x\pi_n^N(f)\right)} \leq e^{4\lambda^2 \|f\|^2_{\text{Lip}} V^2}
-\]
-with the quantity $V^2 = \frac{4}{\kappa n}\left(1 + \frac{N}{n}\right)\sup_{x}\frac{\sigma^2(x)}{n_x \kappa}$, $\sigma^2(x) = \frac{1}{2}\iint \|y-z\|^2P_x(dy)P_x(dz) = \gamma d$ and $n_x \geq 1$.
-\end{proof}
-
-\begin{lem} 
 \label{lem:exp_bound}
@@
 \end{equation}
-Next we use \cite[Proposition 1]{joulin2010}, stating that
+Note that for ULA algorithm the distribution $\P_x$ is Gaussian for any $x \in \rset^d$, thus, following \cite[Proposition 3]{dm2016HighDim}, we may bound the Wasserstein-1 distance between $\P_x$ and $\P_y$ by
+\[
+W_1(\P_x,\P_y) = W_1(\mathcal{N}(x-\gamma\mu(x),\gamma),\mathcal{N}(y-\gamma\mu(y),\gamma)) = \|x-y-\gamma(\mu(x)-\mu(y))\| \leq \sqrt{1-\kappa\gamma}\|x-y\|
+\]
+where $\kappa = \frac{2mL}{m+L}$. Hence, it holds
+\[
+W_1(\P_x,\P_y) \leq \left(1-\frac{\kappa\gamma}{2}\right)\|x-y\|
+\]
+which means that the Markov kernel $\P$ is a strict contraction in Wasserstein metric. This allows to use \cite[Proposition 1]{joulin2010}, stating that
 \begin{equation}

### 2020-05-02 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 \]
-Note that due to our assumptios, $\mu(x) = \mu_0(x) + \Delta(x)$ with $\mu^{\prime}_0(x) < -m$. Thus by applying Holder inequality to $\zeta_{l,j}(u)$, we obtain
+Note that due to our assumptios, $\mu(x) = \mu_0(x) + \Delta(x)$ with $\mu^{\prime}_0(x) < -m$. Note that $\left|\Delta^{\prime}(x)\right| \leq M^{\prime}$, hence
 \begin{equation}
 \label{eq:zeta}
-\zeta_{l,j}(u) \leq \left(\mathsf{E}\left[\rme^{-2u\gamma\sum_{k=l}^{j}\mu_0'(X^x_{k-1})}\right]\right)^{1/2}\left(\mathsf{E}\left[\rme^{-2u\gamma\sum_{k=l}^{j}\Delta'(X^x_{k-1})}\right]\right)^{1/2}
-\end{equation}
-Note that $\left|\Delta^{\prime}(x)\right| \leq M^{\prime}$, hence
-\[
-\left(\mathsf{E}\left[\rme^{-2u\gamma\sum_{k=l}^{j}\Delta'(X^x_{k-1})}\right]\right)^{1/2} \leq \mathsf{E}\rme^{2u(j-l)M^{\prime}}
-\]
+\zeta_{l,j}(u) \leq \mathsf{E}\left[\rme^{-u\gamma\sum_{k=l}^{j}\mu_0'(X^x_{k-1})}\right]\rme^{u\gamma(j-l+1)M^{\prime}}
+\end{equation}
 To bound the first exponential moment in \ref{eq:zeta}, we apply
@@
 \begin{align*}
-\zeta_{l,j}(u) \leq \begin{cases} C(x)\exp\left\{(j-l)\gamma\left(4u^2\kappa^{-2} \|\mu^{\prime}\|^2_{\text{Lip}} + M^2 + M^{\prime} + C_1u\sqrt{\gamma} -  u\pi(\mu')\right)\right\}, \quad \gamma(j-l) \geq C_0 \\
-\exp\left\{uC_0\max\limits_{y}|\mu'(y)|\right\}, \quad 0 \leq \gamma(j-l) \leq C_0	
+\zeta_{l,j}(u) \leq \begin{cases} C(x)\exp\left\{(j-l+1)\gamma\left(4u^2\kappa^{-2} \|\mu_0^{\prime}\|^2_{\text{Lip}} + M^2 + M^{\prime} + C_1u\sqrt{\gamma} -  u\pi_{\mu_0}(\mu_0')\right)\right\}, \quad \gamma(j-l+1) \geq C_0 \\
+\exp\left\{uC_0(\max\limits_{y}|\mu_0'(y)| + M^{\prime})\right\}, \quad 0 \leq \gamma(j-l+1) \leq C_0	
 \end{cases}
 \end{align*}
-with $C_0 = \left(\frac{4}{m+L} + \kappa^{-1}\right)$, and constants $C(x)$ and $C_1$ given in Lemma~\ref{lem:exp_bound} .
+with $C_0 = \left(\frac{4}{m+L} + \kappa^{-1}\right)$, measure $\pi_{\mu_{0}}$ and constants $C(x)$ and $C_1$ given in Lemma~\ref{lem:exp_bound} .
 Furthermore, since $\mu\pi',\mu'\pi\in L^{1}(\mathbb{R})$ and $\pi^{\prime}(x) = -\frac{1}{2}\pi(x)\mu(x)$, we have
@@
   \end{align*}
-which proves the first statement. The second statement is obvious.
+Now the result follows given that $K_1 (1 +L/m)   / \| x \| \geq \frac{1}{2}$.
 \end{proof}
@@
 \]
-which means that the Markov kernel $\P$ is a strict contraction in Wasserstein metric. This allows to use \cite[Proposition 1]{joulin2010}, stating that
+which means that the Markov kernel $\P$ is a strict contraction in the Wasserstein metric. This allows to use \cite[Proposition 1]{joulin2010}, stating that
 \begin{equation}
 \label{eq:2_bound}
-\left|\sum_{k=N+1}^{N+n}\PE_x f(Y_k) - n\pi_{\gamma}(f)\right| \leq \frac{2\left(1-\frac{\kappa}{2}\right)^{N+1}}{\kappa}\|f\|_{\text{Lip}}\int_{\rset}|x-y|\,\pi_{\gamma}(dy) \leq \frac{m+L}{mL}\|f\|_{\text{Lip}}S(x)
-\end{equation}
-and \cite[Corollary 7]{dm2016HighDim}, that is
+\left|\sum_{k=N+1}^{N+n}\PE_x f(Y_k) - n\pi_{\gamma}(f)\right| \leq \frac{2\left(1-\frac{\kappa\gamma}{2}\right)^{N+1}}{\kappa\gamma}\|f\|_{\text{Lip}}\int_{\rset}|x-y|\,\pi_{\gamma}(dy) \leq \frac{m+L}{mL\gamma}\|f\|_{\text{Lip}}S(x)
+\end{equation}
+with $S(x) = \int_{\rset}|x-y|\,\pi_{\gamma}(dy)$. Then by \cite[Corollary 7]{dm2016HighDim}, 
 \begin{equation}

### 2020-05-02 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 To bound the first exponential moment in \ref{eq:zeta}, we apply
-Lemma~\ref{lem:exp_bound} in Appendix~\ref{sec:appendix}. It implies that
+Lemma~\ref{lem:exp_bound} in Appendix~\ref{sec:appendix} to $f(x) = -\mu^{\prime}_0(x)$. It implies that
 \begin{align*}
-\zeta_{l,j}(u) \leq \begin{cases} C(x)\exp\left\{(j-l+1)\gamma\left(4u^2\kappa^{-2} \|\mu_0^{\prime}\|^2_{\text{Lip}} + M^2 + M^{\prime} + C_1u\sqrt{\gamma} -  u\pi_{\mu_0}(\mu_0')\right)\right\}, \quad \gamma(j-l+1) \geq C_0 \\
+\zeta_{l,j}(u) \leq \begin{cases} C(x)\exp\left\{(j-l+1)\gamma\left(4u^2\kappa^{-2} \|\mu_0^{\prime}\|^2_{\text{Lip}} + M^2 + M^{\prime} + C_1u\sqrt{\gamma} +  u\pi_{\mu_0}(\mu_0')\right)\right\}, \quad \gamma(j-l+1) \geq C_0 \\
 \exp\left\{uC_0(\max\limits_{y}|\mu_0'(y)| + M^{\prime})\right\}, \quad 0 \leq \gamma(j-l+1) \leq C_0	
@@
 with $C_0 = \left(\frac{4}{m+L} + \kappa^{-1}\right)$, measure $\pi_{\mu_{0}}$ and constants $C(x)$ and $C_1$ given in Lemma~\ref{lem:exp_bound} .
-Furthermore, since $\mu\pi',\mu'\pi\in L^{1}(\mathbb{R})$ and $\pi^{\prime}(x) = -\frac{1}{2}\pi(x)\mu(x)$, we have
-\[
-\pi(\mu')=\int\mu'(x)\pi(x)\,dx=-\int\mu(x)\pi'(x)\,dx=\frac{1}{2}\int\mu^{2}(x)\pi(x)\,dx>0
+Furthermore, since $\mu_0\pi_{\mu_0}',\mu_0'\pi_{\mu_0}\in L^{1}(\rset)$ and $\pi_{\mu_0}^{\prime}(x) = \pi_{\mu_0}(x)\mu_0(x)$, we have
+\[
+\pi_{\mu_0}(\mu_0')=\int\mu_0'(x)\pi_{\mu_0}(x)\,dx=-\int\mu_{\pi_0}(x)\pi_{\mu_0}'(x)\,dx= -\int\mu_0^{2}(x)\pi_{\mu_0}(x)\,dx < 0
 \]
@@
 \begin{align*}
-\zeta_{l,j}(u) \leq \begin{cases} C(x)\exp\left\{-\gamma(j-l)\alpha\right\}, \quad \gamma(j-l) \geq C_0 \\
-\exp\left\{uC_0\max\limits_{y}|\mu'(y)|\right\}, \quad 0 \leq \gamma(j-l) \leq C_0	
+\zeta_{l,j}(u) \leq \begin{cases} C(x)\exp\left\{\gamma(j-l+1)\alpha\right\}, \quad \gamma(j-l+1) \geq C_0 \\
+\exp\left\{(\max\limits_{y}|\mu_0'(y)| + M^{\prime})\right\}, \quad 0 \leq \gamma(j-l+1) \leq C_0
 \end{cases}
 \end{align*}
-with the quantity $ \alpha = \min(...,...) > 0$. Hence we obtain
+with the quantity 
+\[
+\alpha = \max(64\kappa^{-2} \|\mu_0^{\prime}\|^2_{\text{Lip}} + M^2 + M^{\prime} + 4C_1\sqrt{\gamma} +  4\pi_{\mu_0}(\mu_0'), 16\kappa^{-2} \|\mu_0^{\prime}\|^2_{\text{Lip}} + M^2 + M^{\prime} + 2C_1\sqrt{\gamma} +  2\pi_{\mu_0}(\mu_0')) < 0
+\] 
+Hence we obtain
 \begin{align*}
-\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4) &\leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + C(x)\sum_{p=j+1}^{\infty} \rme^{-\gamma(p-j)\alpha}
-\leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + \frac{C(x)}{1-\rme^{- \gamma\alpha}}
+\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4) &\leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + C(x)\sum_{p=j+1}^{\infty} \rme^{\gamma(p-j)\alpha}
+\leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + \frac{C(x)}{1-\rme^{\gamma\alpha}}
 \end{align*}
@@
 \begin{align*}
-\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + \frac{C(x)}{1-\rme^{- \gamma\alpha}}
+\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + \frac{C(x)}{1-\rme^{\gamma\alpha}}
 \end{align*}
@@
 \[
-\PCov_x \left[f(X_k),f(X_{k+s})\right] = \PE_x\left[f(X_k) - \pi(f)\right]\left[f(X_{k+s})-\pi(f)\right] + \left[\pi(f) - \PE_xf(X_s)\right]\left[\PE_xf(X_k) - \pi(f)\right]
+\PCov_x \left[f(X_k),f(X_{k+s})\right] = \PE_x\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right] + \left[\pi(f) - \PE_xf(X_{k+s})\right]\left[\PE_xf(X_k) - \pi(f)\right]
 \]
@@
 \[
-\left|\left[\pi(f) - \PE_xf(X_s)\right]\left[\PE_xf(X_k) - \pi(f)\right]\right| \leq C^2V^2(x)\rho^{k+s}
+\left|\left[\pi(f) - \PE_xf(X_{k+s})\right]\left[\PE_xf(X_k) - \pi(f)\right]\right| \leq C^2V^2(x)\rho^{2k+s}
 \]
@@
 \begin{multline*}
-\left|\PE_x\left[f(X_k) - \pi(f)\right]\left[f(X_{k+s})-\pi(f)\right] - \PCov_{\pi}\left[f(X_k),f(X_{k+s})\right]\right|
+\left|\PE_x\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right] - \PCov_{\pi}\left[f(X_k),f(X_{k+s})\right]\right| = \left|\PE_x\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right] - \PE_\pi\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right]\right| \leq
 \\
@@
  \[
- \pi(y) = \frac{\exp\left\{\int\limits_{0}^{y}\mu_{0}(x)\,dx\right\}}{\int\limits_{-\infty}^{+\infty}\exp\left\{\int\limits_{0}^{y}\mu_{0}(x)\,dx\right\}\,dy}
+ \pi_{\mu_0}(y) = \frac{\exp\left\{\int\limits_{0}^{y}\mu_{0}(x)\,dx\right\}}{\int\limits_{-\infty}^{+\infty}\exp\left\{\int\limits_{0}^{y}\mu_{0}(x)\,dx\right\}\,dy}
  \]

### 2020-05-02 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 $$
-d_V(\delta_xQ_{\gamma}^{n},\pi_\gamma) \leq C\rho^{\gamma n}\left(V(x) + \pi_\gamma(V)\right)
+d_V(\delta_xQ_{\gamma}^{n},\pi_\gamma) \leq C\rho^{n}V(x)
 $$
@@
 $$
-C = \left(1 + \exp{\left(-\frac{m\gamma}{2}\right)}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp{\left(-\frac{m\gamma}{2}\right)} - \frac{2b}{2+K^2})}\right);
-$$
-
-$$
-b = \gamma(2K_2^2(L+m) + d + m); \quad \overline{b} = b\exp{\left(-\frac{m\gamma}{2}\right)} + K^2 + 1; \quad \varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right); \quad
-$$
-
-$$
-\rho = \exp\left\{\left(\frac{m}{2} - 2\frac{2K_2^2(L+m) + d + m}{K^2 + 1}\right)\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{m\gamma}{2}\right)} + \frac{2b}{K^2+2})}}\right\}
+C = \left(1+\pi_\gamma(V)\right)\left(1 + \exp{\left(-\frac{m\gamma}{2}\right)}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp{\left(-\frac{m\gamma}{2}\right)} - \frac{2b}{2+K^2})}\right);
+$$
+
+$$
+b = \gamma\left(2K_2^2(L+m) + d + m\right); \quad \overline{b} = b\exp{\left(-\frac{m\gamma}{2}\right)} + K^2 + 1; \quad \varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right); \quad
+$$
+
+$$
+\rho = \exp\left\{\gamma\left(\frac{m}{2} - 2\frac{2K_2^2(L+m) + d + m}{K^2 + 1}\right)\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{m\gamma}{2}\right)} + \frac{2b}{K^2+2})}}\right\}
 $$
@@
 Let $\tilde{f}(x) = f(x) - \pi(f)$.
-Then it holds for some constant $C > 0$
-\begin{equation}
-|\mathsf{E}_x[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq C\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x)
+Then it holds with constant $C > 0$ and $0 < \rho <1$ from \eqref{eq:v_ergodicity} that
+\begin{equation}
+\label{eq:fp_cov}
+|\mathsf{E}_x[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq 2C\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}V(x)
 \end{equation}
@@
 \begin{equation}
-|\mathsf{E}_{\pi}[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq C\pi(V)\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}
+\label{eq:fp_cov_stationary}
+|\mathsf{E}_{\pi}[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq 2C\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}\pi(V)
 \end{equation}
@@
 \[
-|\mathsf{E}_x[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq |\tilde{f}(x)| \mathsf{E}_x|\tilde{f}(X_s)| \leq \|\tilde{f}\|_{V^{\frac{1}{2}}}V^{\frac{1}{2}}(x)\int\limits_{\rset^d} V^{\frac{1}{2}}(y)|\mathsf{P}^{s}(x,\,dy)-\pi(dy)|
+|\mathsf{E}_x[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq |\tilde{f}(x)|^2 \mathsf{E}_x|\tilde{f}(X_s)| \leq \|\tilde{f}\|_{V^{\frac{1}{2}}}V^{\frac{1}{2}}(x)\int\limits_{\rset^d} V^{\frac{1}{2}}(y)|\mathsf{P}^{s}(x,\,dy)-\pi(dy)|
 \]
@@
 \label{eq:cov_v_ergodic}
-\PCov_x \left[f(X_k),f(X_{k+s})\right] \leq C^2V^2(x)\rho^{k+s} + C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x) + C\pi(V)\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}
-\end{equation}
+\PCov_x \left[f(X_k),f(X_{k+s})\right] \leq C^2V^2(x)\rho^{2k+s} + 2C^2\rho^{k+s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}V(x) + 2C\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}\pi(V)
+\end{equation}
+with constants $C > 0$ and $0 < \rho < 1$ from \eqref{eq:v_ergodicity}.
 \end{lem}
@@
 \]
-Due to $V-$ergodicity,
+Due to $V-$ergodicity \eqref{eq:v_ergodicity},
 \[
@@
 \leq \int\limits_{\rset^d}\left|\PE_y\left[\tilde{f}(X_0)\tilde{f}(X_{s})\right]\right| |\P^k(x,dy)-\pi(dy)|
-\leq C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x),
+\leq \{\eqref{eq:fp_cov}\}\leq 2C\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}\int\limits_{\rset^d} V(y) |\P^k(x,dy)-\pi(dy)| \leq 
+\\
+\leq 2C^2\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}\|\P^k(x,dy)-\pi(dy)\|_{V} \leq 2C^2\rho^{k+s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}V(x)
 \end{multline*}
-where the last inequality is due to lemma~\ref{lem:fp_covariance}, which implies \eqref{eq:cov_v_ergodic}.
+where $\left|\PCov_{\pi}\left[f(X_k),f(X_{k+s})\right]\right|$ can be bounded using \eqref{eq:fp_cov_stationary}. 
 \end{proof}
@@
 Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$.
-Assume also that $n(1-\rho^{1/2}) > 1$.
+Let $\tilde{f}(x) = f(x) - \pi(f)$ and assume that $n(1-\rho^{1/2}) > 1$.
 Then
@@
 \label{eq:vanilla_var}
-\PVar_x \left[ \pi_{N}^n(f) \right] \leq \frac{C\pi(V)\|\tilde{f}\|_{V^{1/2}}}{n(1-\rho^{1/2})} + \mathcal{O}(\frac{1}{n^2(1-\rho^{1/2})})
-\end{equation}
+\PVar_x \left[ \pi_{N}^n(f) \right] \leq \frac{C_1\pi(V)\|\tilde{f}\|^2_{V^{1/2}}}{n(1-\rho^{1/2})} + \mathcal{O}(\frac{1}{n^2(1-\rho^{1/2})})
+\end{equation}
+with constant $C_1 > 0$ and $0 < \rho <1$ from \eqref{eq:v_ergodicity}.
 \end{lem}
+
 \begin{proof}
@@
 \[
-S_1 \leq \frac{1}{n}C\pi(V)\|\tilde{f}\|_{V^{1/2}} + \frac{\rho^N C^2\left(V^2(x) + V(x)\|\tilde{f}\|_{V^{1/2}}\right)}{n^2(1-\rho)}
-\]
-\begin{eqnarray*}
-S_2 & \leq & \frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\left[C^2V^2(x)\rho^{k+1}\frac{1}{1-\rho} + C^2V(x)\rho^{k+1/2}\frac{1}{1-\rho^{1/2}} + C\pi(V)\frac{1}{1-\rho^{1/2}}\|\tilde{f}\|_{V^{1/2}}\right] \leq \\
-& \leq & \frac{2\rho^NC^2V^2(x)}{n^2(1-\rho)^2} + \frac{2\rho^NC^2V(x)}{n^2(1-\rho)(1-\rho^{1/2})} + \frac{C\pi(V)\|\tilde{f}\|_{V^{1/2}}}{n(1-\rho^{1/2})}
+S_1 \leq \frac{2}{n}C\pi(V)\|\tilde{f}\|^2_{V^{1/2}} + \frac{\rho^N C^2V(x)\left(\rho^NV(x) + 2\|\tilde{f}\|^2_{V^{1/2}}\right)}{n^2(1-\rho)}
+\]
+\begin{eqnarray*}
+S_2 & \leq & \frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\left[\frac{C^2V^2(x)\rho^{2k+1}}{1-\rho} + \frac{2C^2V(x)\|\tilde{f}\|^2_{V^{1/2}}\rho^{k+1/2}}{1-\rho^{1/2}} + \frac{2C\|\tilde{f}\|^2_{V^{1/2}}\pi(V)}{1-\rho^{1/2}}\right] \leq \\
+& \leq & \frac{2C^2\rho^{2N}V^2(x)}{n^2(1-\rho)^2} + \frac{4C^2\rho^{N+1/2}\|\tilde{f}\|^2_{V^{1/2}}V(x)}{n^2(1-\rho)(1-\rho^{1/2})} + \frac{4C\|\tilde{f}\|^2_{V^{1/2}}\pi(V)}{n(1-\rho^{1/2})}
 \end{eqnarray*}
@@
  Suppose that there exists $M > 0,$ such that $\left|\Delta(x)\right| \leq M$.
- Then for any Lipschitz function $f$, $\forall N $
+ Then for any Lipschitz function $f$ and for any $ N \geq 1$,
 \begin{equation}
 \label{eq:exp_bound}
-\PE_x \rme^{u\gamma\sum\limits_{k=N+1}^{N+n}f(X_k)} \leq C(x)\exp\left\{n\gamma\left(2u^2\kappa^{-2} \|f\|^2_{\text{Lip}}\left(1+\frac{1}{n\gamma}\left(\frac{4}{m+L} + \kappa^{-1}\right)\right) + M^2 + C_1u\sqrt{\gamma} +  u\pi(f)\right)\right\}
+\PE_x \rme^{u\gamma\sum\limits_{k=N+1}^{N+n}f(X_k)} \leq C(x)\exp\left\{n\gamma\left(2u^2\kappa^{-2} \|f\|^2_{\text{Lip}}\left(1+\frac{1}{n\gamma}\left(\frac{4}{m+L} + \kappa^{-1}\right)\right) + M^2 + C_1u\sqrt{\gamma} +  u\pi_{\mu_0}(f)\right)\right\}
 \end{equation}

### 2020-05-02 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 
-Under assumptions {\bf (H1)} and {\bf (H2)} the Markov chain $(X^x_{p})_{p \in \nset_0}$ is $V-$geometrically ergodic with $V(x) = 1+\|x\|^2$ and $\rho = \rme^{-\kappa\gamma}$ with $\kappa$ specified in Lemma~\ref{lem:v_ergodicity}.
-Due to lemma~\ref{lem:covariance},
+Under assumptions {\bf (H1)} and {\bf (H2)} the Markov chain $(X^x_{p})_{p \in \nset_0}$ is $V-$geometrically ergodic \eqref{eq:v_ergodicity} with $V(x) = 1+\|x\|^2$ and $\rho$ specified in Lemma~\ref{lem:v_ergodicity}.
+Due to lemma~\ref{lem:var_bound},
 \[
@@
 \begin{lem}
+\label{lem:var_bound}
 Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$.

### 2020-05-02 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 Let $f$ be a bounded Borel function and $(X^x_{p})_{p \in \nset_0}$ be a Markov chain generated by ULA with constant step size $\gamma$.
-Let $c_0>0$. Then, there exist $C < \infty$ and $\gamma_0 > 0$, such that for all $n \in \nset$, $\gamma \in (0,\gamma_0]$ satisfying $n\gamma > c_0$, and $x \in \rset^d$,
+Let $c_0>0$. Then, there exist $C < \infty$ and $\gamma_0 > 0$, such that for all $n \in \nset$, $\gamma \in (0,\gamma_0]$ and $x \in \rset^d$,
 \begin{equation}
 \label{eq:var-mc}
-\PVar\left[\pi_n^N(f)\right]\leq C (n \gamma)^{-1} V(x) \eqsp.
+\PVar\left[\pi_n^N(f)\right]\leq \frac{C}{n\gamma} + C\left(\frac{\rho^NV(x)+1}{n\gamma}\right)^2 \eqsp.
 \end{equation}
@@
 \[
-\PVar[\pi_n^N(f)]\lesssim \frac{1}{n(1-\rho^{1/2})}
-\]
-and the result of lemma follows from Taylor expansion of denominator.
+\PVar[\pi_n^N(f)]\lesssim \frac{1}{n(1-\rho^{1/2})} + \left(\frac{\rho^NV(x)+1}{n(1-\rho^{1/2})}\right)^2
+\]
+Now from Taylow expansion of denominator it follows that $1-\rho^{1/2} \sim C\gamma$ with constant $C$ not depending on $\gamma,N$ and $n$, thus yielding an estimate
+\[
+\PVar[\pi_n^N(f)] \lesssim \frac{1}{n\gamma} + \left(\frac{\rho^NV(x)+1}{n\gamma}\right)^2
+\]
 
@@
 Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$.
-Let $\tilde{f}(x) = f(x) - \pi(f)$ and assume that $n(1-\rho^{1/2}) > 1$.
+Let $\tilde{f}(x) = f(x) - \pi(f)$.
 Then
@@
 \label{eq:vanilla_var}
-\PVar_x \left[ \pi_{N}^n(f) \right] \leq \frac{C_1\pi(V)\|\tilde{f}\|^2_{V^{1/2}}}{n(1-\rho^{1/2})} + \mathcal{O}(\frac{1}{n^2(1-\rho^{1/2})})
-\end{equation}
-with constant $C_1 > 0$ and $0 < \rho <1$ from \eqref{eq:v_ergodicity}.
+\PVar_x \left[ \pi_{N}^n(f) \right] \leq \frac{C_1\pi(V)\|\tilde{f}\|^2_{V^{1/2}}}{n(1-\rho^{1/2})} + C_2\left(\frac{\rho^NV(x) + 1}{n(1-\rho^{1/2})}\right)^2
+\end{equation}
+with positive constants $C_1,C_2 > 0$ and $0 < \rho <1$ from \eqref{eq:v_ergodicity}.
 \end{lem}
@@
 \[
-S_1 \leq \frac{2}{n}C\pi(V)\|\tilde{f}\|^2_{V^{1/2}} + \frac{\rho^N C^2V(x)\left(\rho^NV(x) + 2\|\tilde{f}\|^2_{V^{1/2}}\right)}{n^2(1-\rho)}
+S_1 \leq \frac{2C\pi(V)\|\tilde{f}\|^2_{V^{1/2}}}{n} + \frac{\rho^N C^2V(x)\left(\rho^NV(x) + 2\|\tilde{f}\|^2_{V^{1/2}}\right)}{n^2(1-\rho)}
 \]

### 2020-05-03 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 $$
-C = \left(1+\pi_\gamma(V)\right)\left(1 + \exp{\left(-\frac{m\gamma}{2}\right)}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp{\left(-\frac{m\gamma}{2}\right)} - \frac{2b}{2+K^2})}\right);
-$$
-
-$$
-b = \gamma\left(2K_2^2(L+m) + d + m\right); \quad \overline{b} = b\exp{\left(-\frac{m\gamma}{2}\right)} + K^2 + 1; \quad \varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right); \quad
-$$
-
-$$
-\rho = \exp\left\{\gamma\left(\frac{m}{2} - 2\frac{2K_2^2(L+m) + d + m}{K^2 + 1}\right)\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{m\gamma}{2}\right)} + \frac{2b}{K^2+2})}}\right\}
-$$
-where $K_2 > 0$ is from Lemma~\ref{lem:quadratic_behaviour} and $K > 0$ is arbitrary
+C = \left(1+\pi_\gamma(V)\right)\left(1 + \exp{\left(-\frac{m\gamma}{2}\right)}\right)\left(1+\frac{b\exp{\left(-\frac{m\gamma}{2}\right)} + K^2 + 1}{(1-\varepsilon)(1-\exp{\left(-\frac{m\gamma}{2}\right)} - \frac{2b}{2+K^2})}\right);
+$$
+
+$$
+b = \gamma\left(2K_2^2(L+m) + d + m\right); \quad \varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right); \quad
+$$
+
+$$
+\rho = \exp\left\{\frac{\gamma m}{4}\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{m\gamma}{2}\right)} + \frac{2b}{K^2+2})}}\right\}
+$$
+where $K_2 > 0$ is from Lemma~\ref{lem:quadratic_behaviour} and $K>0$ is a positive parameter such that 
+$$
+K^2 > 4\frac{2K_2^2(L+m) + d + m}{m}
+$$
 \end{lem}

### 2020-05-04 aap/Revision/vr-mcmc-revision-new-new.tex
update

@@
 $$
-\rho = \exp\left\{\frac{\gamma m}{4}\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{m\gamma}{2}\right)} + \frac{2b}{K^2+2})}}\right\}
+\rho = \exp\left\{-\frac{\gamma m}{4}\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{m\gamma}{2}\right)} + \frac{2b}{K^2+2})}}\right\}
 $$

### 2020-05-04 aap/Revision/vr-mcmc-revision-final.tex
update

--- 2020-05-03 aap/Revision/vr-mcmc-revision-new-new.tex
@@
 We denote $\boldsymbol{\varphi}(z)=(2\pi)^{-d/2} \exp(-|z|^2/2)$, $z\in\mathbb R^d$ probability density function of the $d-$dimensional standard normal distribution.
-For $x \in \rset^d$ and $r>0$ let $B_r(x) = \{y \in \rset^d | \|y - x\| < r\}$ where $\|\cdot\|$ is a standard Euclidean norm.
+For $x \in \rset^d$ and $r>0$ let $B_r(x) = \{y \in \rset^d | |y - x| < r\}$ where $|\cdot|$ is a standard Euclidean norm.
 For the twice differentiable function $g: \rset^d \rightarrow \rset$ we denote by $D^2g(x)$ its Hessian at point $x$.
@@
 For a bounded Borel function $f: \rset^d \rightarrow \rset$ denote $\mathrm{osc}(f):=\sup_{x\in\mathbb R^d}f(x)-\inf_{x\in\mathbb R^d}f(x)$.
-Given a function $V: \rset^d \rightarrow \rset$, for a function $f: \rset^d \rightarrow \rset$ we define $\|f\|_{V} = \sup_{x \in \rset^d} |f(x)|/V(x)$ and the corresponding $V-$norm between probability measures $\mu$ and $\nu$ on $\mathcal{B}(\rset^d)$ as
+Given a function $V: \rset^d \rightarrow \rset$, for a function $f: \rset^d \rightarrow \rset$ we define $\|f\|_{V} = \sup_{x \in \rset^d} \{|f(x)|/V(x)\}$ and the corresponding $V-$norm between probability measures $\mu$ and $\nu$ on $\mathcal{B}(\rset^d)$ as
 \[
@@
 \begin{equation}\label{eq:06042018a1}
-\hat{Q}_{r} \in \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \left|   f(X_{r+N}^{(s)}) - \psi(X^{(s)}_{N})\right|^2
+\widehat{Q}_{r} \in \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \left|   f(X_{r+N}^{(s)}) - \psi(X^{(s)}_{N})\right|^2, \quad r=0,\ldots,n-1,
 \end{equation}
@@
 \label{eq:definition-hat-a-A}
-\hat{a}_{r,k}(x)= \int \hat{Q}_r(\Phi(x,z)) \phi_k(z) P_\xi(\rmd z) \quad \text{and}
+\widehat{a}_{r,k}(x)= \int \widehat{Q}_r(\Phi(x,z)) \phi_k(z) P_\xi(\rmd z),
 \quad
-\hat{A}_{q,k}(x) = \sum_{r=0}^q \hat{a}_{r,k}(x) \eqsp,
-\end{equation}
-and construct the empirical estimate  of \(M_{K,n}^N(f)\) in the form
+\widehat{A}_{q,k}(x) = \min\left\{W_k,\sum_{r=0}^q \widehat{a}_{r,k}(x)\right\},
+\end{equation}
+where $(W_k)$ is a sequence of truncation levels. Finally, we  construct the empirical estimate  of \(M_{K,n}^N(f)\) in the form
 \begin{eqnarray*}
@@
 \end{eqnarray*}
-
 Obviously, \(\mathsf{E}[\widehat M_{K,n}^N(f)|\TrainSet]=0\) and \(\widehat M_{K,n}^N(f)\) is a valid control variate in that it
@@
 \begin{eqnarray*}
-\widehat\pi_{K,n}^N(f):=\pi_n^N(f)-\widehat{M}_{K,n}^N(f)
+\widehat\pi_{K,n}^N(f):=\pi_n^N(f)-\widehat{M}_{K,n}^N(f).
 \end{eqnarray*}
@@
 \end{multline}
-In the next section we analyze the case of ULA algorithm and show that under some rather mild conditions (convexity of the potential $U$ outside a ball around zero)
-\begin{eqnarray*}
-\PE[A^2_{l,k}(X^x_{N+l-1})]\leq \rme^{\varkappa n\gamma^2}R^2_k ,\quad k \in \nset, \quad l=0,\ldots,n-1,
+In the next section we analyze the case of ULA algorithm and show that under some  mild conditions (Lipschitz continuity and convexity outside a ball of the potential $U$),
+\begin{eqnarray*}
+\PE[A^2_{l,k}(X^x_{N+l-1})]\leq R^2_k ,\quad k \in \nset, \quad l=0,\ldots,n-1,
 \end{eqnarray*}
 for a sequence  $(R_k)$ satisfying $\sum_{k=1}^\infty R^2_k\leq C$ with constant $C$  not depending  on $\gamma$ and $n$ (see \eqref{eq:Aqk2}).
-Hence if we take $W_k=R_k$ as truncation parameters in \eqref{eq:trunc}, then $\widehat A_{n-l,k}$ converges to
+Hence if we take $W_k=R_k$ as truncation parameters in \eqref{eq:definition-hat-a-A}, then $\widehat A_{n-l,k}$ converges to
 \[
@@
 \begin{eqnarray*}
-\PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]\leq \frac{2C \rme^{\varkappa n\gamma^2}}{n}.
-\end{eqnarray*}
- Note that the above estimate is rather rough, as we ignore here the fact that the difference between $\widehat A_{n-l,k}$ and $A_{n-l,k}$ is small.
-If the class of functions $\Psi$ is a linear one of dimension $\mathtt{D}_\Psi,$ then the cost of computing the coefficients $\widehat A_{l,k}$ for all $l=1,\ldots,n$ and $k=1,\ldots,K$ is of order  $ \mathtt{D}_\Psi K T^2 n$.
+\PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]\leq \frac{2C}{n}.
+\end{eqnarray*}
+% Note that the above estimate is rather rough, as we ignore here the fact that the difference between $\widehat A_{n-l,k}$ and $A_{n-l,k}$ is small.
+If the class of functions $\Psi$ is a linear one of dimension $\mathtt{D}_\Psi,$ then the cost of computing the coefficients $\widehat A_{l,k}$ for all $l=1,\ldots,n,$ and $k=1,\ldots,K,$ is of order  $ \mathtt{D}_\Psi K T^2 n$.
 Given $(\widehat A_{l,k})$ the cost of computing  $\widehat \pi_{K,n}^N$ is proportional to $\mathtt{D}_\Psi K n.$ At the same time the variance of the standard estimate $ \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved, see Lemma~\ref{lem:variance} and Remark~\ref{rem:var-low}.
@@
 \label{eq:cost_theoretical}
-\frac{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]}{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}\leq C \rme^{\varkappa n\gamma^2} K \gamma \, T^2 \mathtt{D}_\Psi.
+\frac{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]}{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}\leq C  K \gamma \, T^2 \mathtt{D}_\Psi.
 \end{equation}
@@
 \end{equation}
-Our analysis is carried out under the following two assumptions:
+Our analysis is carried out under the following two assumptions.
 \begin{description}
@@
 \begin{eqnarray*}
-\| \nabla U(x)-\nabla U(y) \| \leq L \|x-y\|, \quad x,y\in \rset^d.
-\end{eqnarray*}
-\item[(H2)][\textbf{Convexity outside a ball}] There exist $K>0$ and $m>0$ such that for any $x\not\in B_{K}(0)$  it holds
+| \nabla U(x)-\nabla U(y) | \leq L |x-y|, \quad x,y\in \rset^d.
+\end{eqnarray*}
+\item[(H2)][\textbf{Convexity}] The potential $U$ is of the form $U(x)=U_0(x)+\Delta(x),$ where $\Delta$ has  compact support and $U_0\in C^2(\mathbb{R}^d)$ with
 \begin{equation*}
-\ps{D^2U(x)}{x} \geq m\|x\|^2.
+\mathfrak{m} |x|^2 \leq \ps{D^2U_0(x)}{x} \leq \mathfrak{M} |x|^2,\quad   x\in \mathbb{R}
 \end{equation*}
+for some constants $\mathfrak{m},\mathfrak{M}>0.$
 \end{description}
@@
 \label{eq:ula_kernel}
-\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(2 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{2\gamma} \left\| y - x + \gamma \nabla U(x)\right\|^2\right\} \rmd y
+\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(2 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{2\gamma} | y - x + \gamma \nabla U(x)|^2\right\} \rmd y
 \end{equation}
@@
 $\pi_n^N(f)$ and $\pi_{K,n}^N(f)$ have the same  bias.
-Under assumptions {\bf (H1)} and {\bf (H2)}, the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$.
+Under assumptions {\bf (H1)} and {\bf (H2)}, the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$ which is different from $\pi$.
 From \cite[Theorem~10]{durmus:moulines:2017},
@@
  \label{eq:definition-V}
- V(x) = 1+ \|x\|^2 \,.
- \end{equation}
+ V(x) = 1+ |x|^2 \,.
+ \end{equation}
+ The following lemma follows from Lemma~\ref{eq:cov_v_ergodic} in Appendix.
 \begin{lem}
@@
 Let $f$ be a bounded Borel function and $(X^x_{p})_{p \in \nset_0}$ be a Markov chain generated by ULA with constant step size $\gamma$.
-Let $c_0>0$. Then, there exist $C < \infty$ and $\gamma_0 > 0$, such that for all $n \in \nset$, $\gamma \in (0,\gamma_0]$ and $x \in \rset^d$,
+Let $c_0>0$. Then, there exist $C < \infty$ and $\gamma_0 > 0$, such that for all $n \in \nset$, $\gamma \in (0,\gamma_0]$ satisfying $n\gamma > c_0$, and $x \in \rset^d$,
 \begin{equation}
 \label{eq:var-mc}
-\PVar\left[\pi_n^N(f)\right]\leq \frac{C}{n\gamma} + C\left(\frac{\rho^NV(x)+1}{n\gamma}\right)^2 \eqsp.
+\PVar\left[\pi_n^N(f)\right]\leq C (n \gamma)^{-1} V(x) \eqsp.
 \end{equation}
@@
 \begin{eqnarray*}
-X_{k+1} = (1-2\gamma)X_k + \sqrt{\gamma}\xi_{k+1},\quad \xi_{k+1} \sim \mathcal{N}(0,1)
+X_{k+1} = (1-2\gamma)X_k + \sqrt{\gamma}\xi_{k+1},\quad \xi_{k+1} \sim \mathcal{N}(0,1).
 \end{eqnarray*}
@@
 \begin{thm}\label{th:mr}
-Assume {\bf (H1)} and {\bf (H2)}.
-Fix some $K\geq 1$ and suppose additionally that a bounded function $f$ and $\mu=\nabla U$ are $K \times d $ times continuously differentiable
+Fix some $K\geq 1$ and suppose  that a bounded function $f$ and $\mu=\nabla U$ are $K \times d $ times continuously differentiable
 and for all $x\in\mathbb R^d$ and  \(\mathbf{k}\) satisfying \(0<\|\mathbf{k}\|\leq  K,\)
@@
 \end{equation}
-Then, there exist $\gamma_0 > 0$, $C_K < \infty$, and $\varkappa > 0$  such that, for all $\gamma \in (0,\gamma_0]$, $n \in \nset$, and $x \in \rset^d$,
+Moreover assume that 
+\begin{eqnarray}
+\label{eq:main-ass-prod}
+\max_{j\geq N}\sum_{p=j+1}^{\infty}\gamma\,\left\Vert\prod_{k=j+1}^{p} (I-\gamma\mathsf{D}^{2}U(X^x_{k-1}))\right \Vert_{r}\leq \Xi,\quad r=2,4,
+\end{eqnarray}
+for some $\Xi=\Xi(x)>0$ not depending on $\gamma,$ where for any random matrix $A$ we define $\|A\|_r=(\mathsf{E}|A|^r)^{1/r}.$ 
+Then, there exists a constant $C_K < \infty$  such that
 \begin{eqnarray}
 \label{eq:Aqk2}
-A^2_{q,\mathbf{k}}(x)\leq C_K \rme^{\varkappa n\gamma^2}\{1+V(x)\} \,,\quad 1\leq |\mathbf{k}|\leq K, \quad \text{for all  $q=0,\ldots,n-1,$}
+A^2_{q,\mathbf{k}}(x)\leq C_K \,\Xi(x) \,,\quad 1\leq |\mathbf{k}|\leq K, \quad q=0,\ldots,n-1,
 \end{eqnarray}
@@
 \label{eq:Aqk2sum}
-\sum_{\|\mathbf{k}\|\geq K+1} A^2_{q,\mathbf{k}}(x)\leq C_{K} \rme^{\varkappa n\gamma^2}\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
-\left(\frac{\gamma}{2}\right)^{|I|K-1} \, \{1+V(x)\} \eqsp,
+\sum_{\|\mathbf{k}\|\geq K+1} A^2_{q,\mathbf{k}}(x)\leq C_{K} \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
+\left(\frac{\gamma}{2}\right)^{|I|K-1} \, \Xi(x) \eqsp,
 \end{eqnarray}
@@
 \begin{cor}
-Under assumptions of Theorem~\ref{th:mr}, there exist $\gamma_0 > 0$, $\tilde{C} < \infty$ and $\varkappa > 0$ such that for all $n \in \nset$, $\gamma \in (0,\gamma_0]$, $n\in \nset$, and $x \in \rset^d$,
+Suppose that  {\bf (H2)} holds with  $\Delta \in C^2(\mathbb{R}^d)$ satisfying $|\mathsf{D}^{2}\Delta(x)|\leq D$ for all $x\in \mathbb{R}^d$ and some constant $D>0.$ Then for  $\gamma  <1/\mathfrak{M}$ we have
+ \begin{eqnarray*}
+\sum_{p=j+1}^{\infty}\gamma\left\Vert\prod_{k=j+1}^{p} (I-\gamma\mathsf{D}^{2}U(X^x_{k-1}))\right\Vert _{r}&\leq &  \frac{1}{\mathfrak{m}}+\frac{ D}{\mathfrak{m}^2(1-\gamma \mathfrak{M})}.
+\end{eqnarray*}
+Hence under assumptions of Theorem~\ref{th:mr}, there exists  a constant $\widetilde{C}_K < \infty$  such that for all $n \in \nset$ and $x \in \rset^d$,
 \begin{eqnarray}
 \label{eq:var-bound}
-\PVar\left[\pi_{K,n}^{N}(f)\right] \leq \tilde{C}  n^{-1} \gamma^{K-1}  \rme^{\varkappa n\gamma^2}
-\eqsp.
+\PVar\left[\pi_{K,n}^{N}(f)\right] \leq \widetilde{C}_K  n^{-1} \gamma^{K-1}  \eqsp.
 \end{eqnarray}
 \end{cor}
+\begin{proof}
+We have
+\begin{multline*}
+\prod_{k=j+1}^{p}\left(I-\gamma\mathsf{D}^{2}U(X_{k-1})\right)=\prod_{k=j+1}^{p}\left(I-\gamma\mathsf{D}^{2}U_0(X_{k-1})\right)+
+\\
+\gamma\sum_{l=j+1}^{p}\prod_{k=j+1}^{l-1}\left(I-\gamma\mathsf{D}^{2}U_{0}(X_{k-1})\right)\mathsf{D}^{2}\Delta(X_{l-1})\prod_{k=l+1}^{p}\left(I-\gamma\mathsf{D}^{2}U_0(X_{k-1})\right),
+\end{multline*}
+where empty products are equal $1$ by definition.
+Hence
+\begin{eqnarray*}
+\left|\prod_{k=j+1}^{p}\left(I-\gamma\mathsf{D}^{2}U(X_{k-1})\right)\right|&\leq &\left(1+\frac{\gamma (p-j) D}{1-\gamma\mathfrak{M}}\right)\prod_{k=j+1}^{p}\left |I-\gamma\mathsf{D}^{2}U_0(X_{k-1})\right|
+\\
+&\leq &  \left(1+\frac{\gamma (p-j) D}{1-\gamma\mathfrak{M}}\right) (1-\gamma \mathfrak{m})^{p-j}.
+\end{eqnarray*}
+
+\end{proof}
+
 Let us sketch the main steps of the proof. First using integration by parts, we prove that
@@
 \]
-Finally, using the Gaussian Poincare inequality, we show  that under our smoothness assumption \eqref{eq:smooth-mu} it holds
+Finally, using the Gaussian Poincare inequality, we show  that under  assumptions \eqref{eq:smooth-mu} and  \eqref{eq:main-ass-prod} it holds
 \begin{equation}
@@
 \PVar \left(\sum_{p=1}^{q}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X^x_{p}\right)
-\right)\leq C_{K}\gamma^{-1}\rme^{\varkappa n\gamma^2}(1+V(x)),\quad q=0,\ldots,n-1,
+\right)\leq C_{K}\gamma^{-1}\Xi(x),\quad q=0,\ldots,n-1,
 \end{equation}
@@
 \label{eq:Qregr}
-\widehat{Q}_r = \argmin_{\psi \in \Psi_m} \sum_{s = 1}^{T}  \left| f(X_{N+r}^{(s)}) - \psi(X_{N}^{(s)})\right|^2
+\widehat{Q}_r = \argmin_{\psi \in \Psi_m} \sum_{s = 1}^{T}  \left| f(X_{N+r}^{(s)}) - \psi(X_{N}^{(s)})\right|^2,
 \end{eqnarray}
@@
 
-Under assumptions {\bf (H1)} and {\bf (H2)} the Markov chain $(X^x_{p})_{p \in \nset_0}$ is $V-$geometrically ergodic \eqref{eq:v_ergodicity} with $V(x) = 1+\|x\|^2$ and $\rho$ specified in Lemma~\ref{lem:v_ergodicity}.
-Due to lemma~\ref{lem:var_bound},
+Under assumptions {\bf (H1)} and {\bf (H2)} the Markov chain $(X^x_{p})_{p \in \nset_0}$ is $V-$geometrically ergodic in the sense of \eqref{eq:v_ergodicity} with $V(x) = 1+|x|^2$ and $\rho\in (0,1) $ specified in Lemma~\ref{lem:v_ergodicity}.
+Due to Lemma~\ref{lem:var_bound},
 \[
@@
 \]
-Now from Taylow expansion of denominator it follows that $1-\rho^{1/2} \sim C\gamma$ with constant $C$ not depending on $\gamma,N$ and $n$, thus yielding an estimate
-\[
-\PVar[\pi_n^N(f)] \lesssim \frac{1}{n\gamma} + \left(\frac{\rho^NV(x)+1}{n\gamma}\right)^2
+It follows from \eqref{eq:rho-ula} that $1-\rho^{1/2} \geq C\gamma$ with a constant $C$ not depending on $\gamma,N$ and $n$, thus yielding an estimate
+\[
+\PVar[\pi_n^N(f)] \lesssim \frac{1}{n\gamma} + \left(\frac{\rho^NV(x)+1}{n\gamma}\right)^2.
 \]
@@
 \PVar\left(\sum_{p=1}^{q}\partial_{1}^{\mathbf{K}_I}f\left(X^x_{p}\right)
-\right)\leq C\gamma^{-1} \rme^{\varkappa n\gamma^2}(1+V(x)),\quad q=1,\ldots,n,
+\right)\leq C\gamma^{-1} \Xi(x),\quad q=1,\ldots,n,
 \end{eqnarray*}
@@
 Under assumptions of Theorem~\ref{th:mr}, it holds
-$$
-\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right]\le  \gamma^{-1} C \rme^{\varkappa n\gamma^2}(1+V(x)), \quad q=1,\ldots,n,
-$$
-where $C,\varkappa>0$ are constants that do not depend on $n$ and $\gamma.$
+\begin{eqnarray}
+\label{eq:lem-var}
+\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}^x\right)\right]\le  C_K\gamma^{-1}  \Xi(x), \quad q=1,\ldots,n,
+\end{eqnarray}
+where $C_K$ is a constant not depending on $x,$ $n$ and $\gamma.$
 \end{lem}
@@
 \end{align*}
-Now using the fact that $\prod\limits_{k=j+1}^{p}\alpha_{k} \leq \exp\left(-\sum\limits_{k=j+1}^{p}\gamma\mu'(X_{k-1}^x)\right),$
-we get
-\[
-\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=1}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]\leq\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \left(\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4)\right)^{2},
-\]
-where we denote
-\[
-\zeta_{l,j}(u)=\mathsf{E}\left[\rme^{-u\gamma\sum_{k=l}^{j}\mu'(X^x_{k-1})}\right],\quad u>0.
-\]
-Note that due to our assumptios, $\mu(x) = \mu_0(x) + \Delta(x)$ with $\mu^{\prime}_0(x) < -m$. Note that $\left|\Delta^{\prime}(x)\right| \leq M^{\prime}$, hence
-\begin{equation}
-\label{eq:zeta}
-\zeta_{l,j}(u) \leq \mathsf{E}\left[\rme^{-u\gamma\sum_{k=l}^{j}\mu_0'(X^x_{k-1})}\right]\rme^{u\gamma(j-l+1)M^{\prime}}
-\end{equation}
-To bound the first exponential moment in \ref{eq:zeta}, we apply
-Lemma~\ref{lem:exp_bound} in Appendix~\ref{sec:appendix} to $f(x) = -\mu^{\prime}_0(x)$. It implies that
-\begin{align*}
-\zeta_{l,j}(u) \leq \begin{cases} C(x)\exp\left\{(j-l+1)\gamma\left(4u^2\kappa^{-2} \|\mu_0^{\prime}\|^2_{\text{Lip}} + M^2 + M^{\prime} + C_1u\sqrt{\gamma} +  u\pi_{\mu_0}(\mu_0')\right)\right\}, \quad \gamma(j-l+1) \geq C_0 \\
-\exp\left\{uC_0(\max\limits_{y}|\mu_0'(y)| + M^{\prime})\right\}, \quad 0 \leq \gamma(j-l+1) \leq C_0	
-\end{cases}
-\end{align*}
-with $C_0 = \left(\frac{4}{m+L} + \kappa^{-1}\right)$, measure $\pi_{\mu_{0}}$ and constants $C(x)$ and $C_1$ given in Lemma~\ref{lem:exp_bound} .
-Furthermore, since $\mu_0\pi_{\mu_0}',\mu_0'\pi_{\mu_0}\in L^{1}(\rset)$ and $\pi_{\mu_0}^{\prime}(x) = \pi_{\mu_0}(x)\mu_0(x)$, we have
-\[
-\pi_{\mu_0}(\mu_0')=\int\mu_0'(x)\pi_{\mu_0}(x)\,dx=-\int\mu_{\pi_0}(x)\pi_{\mu_0}'(x)\,dx= -\int\mu_0^{2}(x)\pi_{\mu_0}(x)\,dx < 0
-\]
-Hence the following bound follows
-\begin{align*}
-\zeta_{l,j}(u) \leq \begin{cases} C(x)\exp\left\{\gamma(j-l+1)\alpha\right\}, \quad \gamma(j-l+1) \geq C_0 \\
-\exp\left\{(\max\limits_{y}|\mu_0'(y)| + M^{\prime})\right\}, \quad 0 \leq \gamma(j-l+1) \leq C_0
-\end{cases}
-\end{align*}
-with the quantity 
-\[
-\alpha = \max(64\kappa^{-2} \|\mu_0^{\prime}\|^2_{\text{Lip}} + M^2 + M^{\prime} + 4C_1\sqrt{\gamma} +  4\pi_{\mu_0}(\mu_0'), 16\kappa^{-2} \|\mu_0^{\prime}\|^2_{\text{Lip}} + M^2 + M^{\prime} + 2C_1\sqrt{\gamma} +  2\pi_{\mu_0}(\mu_0')) < 0
-\] 
-Hence we obtain
-\begin{align*}
-\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4) &\leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + C(x)\sum_{p=j+1}^{\infty} \rme^{\gamma(p-j)\alpha}
-\leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + \frac{C(x)}{1-\rme^{\gamma\alpha}}
-\end{align*}
-and
-\begin{align*}
-\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + \frac{C(x)}{1-\rme^{\gamma\alpha}}
-\end{align*}
-Thus the final bound follows:
-\begin{align*}
-\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X^x_{p}\right)\right] \leq \gamma^{-1} C_K C(x)
-\end{align*}
- with $C_K$  depending neither on $q$ nor on $\gamma$. The proof is completed.
+Now assumption \eqref{eq:main-ass-prod} implies \eqref{eq:lem-var}.
 \end{proof}
@@
 \section{Geometric ergodicity of the Unadjusted Langevin Algorithm}\label{sec:appendix}
+
+\begin{definition}
+\label{def:v_ergodicity}
+Let $(X_p)_{p \in \nset_0}$ be a Markov chain taking values in some space $X$ with the Markov kernel $Q$ and stationary distribution $\pi$.
+We say that $(X_p)_{p \in \nset_0}$ is $V-$geometrically ergodic for a given function $V: X \rightarrow [1; +\infty)$ if there exist real numbers $C > 0$ and $0 < \rho < 1$ such that for any $n \in \nset$,
+\begin{equation}
+\label{eq:v_ergodicity}
+d_V(\delta_xQ^{n},\pi) \leq C\rho^n V(x)
+\end{equation}
+\end{definition}
+In  the sequel, without loss of generality, we assume that $(X_p)_{p \in \nset_0}$ follows ULA dynamics \eqref{eq:chain} with $\nabla U(0)=0$ and  set
+\begin{equation}
+\label{eq:definition-V}
+V(x) = 1 + |x|^2 \eqsp.
+\end{equation}
+\begin{lem}
+  \label{lem:quadratic_behaviour}
+  Assume {\bf (H1)} and {\bf (H2)}. Then there exists $K_2 \geq 0$ such that for any $ | x | \geq K_2$ it holds $\ps{\nabla U(x)}{x} \geq (\mathfrak{m}/2) | x |^2$.
+\end{lem}
+\begin{proof}
+ Suppose that the function $\Delta$ in {\bf (H2)} is supported in the ball $\{x:|x|\leq K_1\}$, then we have 
+  \begin{align*}
+    \ps{\nabla U(x)}{x}
+    &= \int_{0}^{K_1/| x |} D^2 U(t x ) [x^{\otimes 2}] \rmd t + \int_{K_1/| x |} ^ 1 D^2 U(t x ) [x^{\otimes 2}] \rmd t\\
+    & \geq \mathfrak{m} |x |^2 \{1- K_1 (1 +L/\mathfrak{m})   / | x | \}\eqsp,\quad \mbox{for all} \quad  | x |\geq K_1,
+  \end{align*}
+which proves the  statement with $K_2=2K_{1}(1+L/\mathfrak{m}).$
+\end{proof}
+\begin{lem}
+\label{lem:drift}
+ Assume {\bf (H1)} and {\bf (H2)}.
+ Then, for any $\gamma \in (0,\bar{\gamma}]$ with $\overline{\gamma} = \mathfrak{m}/2L^2$  the kernel $Q_\gamma$ from \eqref{eq:ula_kernel} satisfies drift condition
+ \begin{equation}
+\label{eq:lyapunov}
+Q_{\gamma}V(x) \leq \rho^{\gamma} V(x) + \gamma C
+\end{equation}
+with   $\rho = \exp{\left(-\frac{\mathfrak{m}}{2}\right)}$, $C = 2K_2^2(L+\mathfrak{m}) + d + \mathfrak{m}$ with $K_2$ from Lemma~\ref{lem:quadratic_behaviour}, and $\mathfrak{m}$ from {\bf (H2)}.
+\end{lem}
+
+\begin{proof} 
+Note first that
+\[
+Q_{\gamma}V(x) = 
+1+ | x - \gamma \nabla U(x) |^2 + \gamma d
+\]
+Let us first consider the case $x \notin B(0,K_2)$.  
+It remains to notice that, by Lemma~\ref{lem:quadratic_behaviour}, we get
+\[
+|x - \gamma \nabla U(x) |^2 = |x|^2 - 2\gamma \langle \nabla U(x), x\rangle + \gamma^2 |\nabla U(x)|^2 \leq \left(1 - \gamma \mathfrak{m} + \gamma^2 L^2\right)|x|^2
+\]
+Since $\gamma < \frac{\mathfrak{m}}{2L^2}$, we obtain
+$$
+Q_{\gamma}V(x) \leq (1-\gamma \mathfrak{m} + \gamma^2 L^2)V(x) + \gamma (d + \mathfrak{m} - \gamma L^2) \leq \exp\Bigl(-\frac{\gamma \mathfrak{m}}{2}\Bigr)V(x) + \gamma (d + \mathfrak{m})
+$$
+Consider now the case $x \in B(0,K_2)$. Then simply using $|x-\gamma \nabla U(x)|^2 \leq (1+L\gamma)^2|x|^2$, we obtain
+\begin{eqnarray*}
+Q_{\gamma}V(x) &\leq & (1-\gamma \mathfrak{m} + \gamma^2 L^2)V(x) + \gamma\left((\mathfrak{m} - \gamma L^2)(1+|x|^2) + d + L(2+L\gamma)|x|^2\right) \leq
+\\
+&\leq & \exp\Bigl(-\frac{\gamma \mathfrak{m}}{2}\Bigr)V(x) + \gamma(2K_2^2(L+\mathfrak{m}) + d + \mathfrak{m})
+\end{eqnarray*}
+\end{proof}
+
+It is known that under assumption {\bf (H1)} the Markov chain generated by ULA with constant step size $\gamma$ would have unique stationary distribution $\pi_\gamma$, which is different from $\pi$.
+Yet this chain will be $V-$geometrically ergodic due to \cite[Theorem~19.4.1]{moulines2018}.
+Namely, the following lemma holds:
+
+\begin{lem}
+\label{lem:v_ergodicity}
+ Assume  {\bf (H1)} and {\bf (H2)}. Then for $0 < \gamma < \overline{\gamma}= \min(\mathfrak{m} / 4L^2,1/L)$, and for any $x \in \rset^d$  it holds
+$$
+d_V(\delta_xQ_{\gamma}^{n},\pi_\gamma) \leq C\rho^{ n}\left(V(x) + \pi_\gamma(V)\right)
+$$
+with $V(x) = 1 + |x|^2$ and constants
+$$
+C = \left(1 + \exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} - \frac{2b}{2+K^2})}\right),
+$$
+
+$$
+b = \gamma(2K_2^2(L+\mathfrak{m}) + d + \mathfrak{m}), \quad \overline{b} = b\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} + K^2 + 1, 
+$$
+
+\begin{eqnarray*}
+\varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right),
+\end{eqnarray*}
+
+\begin{eqnarray}
+\label{eq:rho-ula}
+\rho = \exp\left\{-\frac{\gamma m}{4}\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} + \frac{2b}{K^2+2})}}\right\},
+\end{eqnarray}
+where $K_2 > 0$ is from Lemma~\ref{lem:quadratic_behaviour} and $K>0$ is a positive parameter such that 
+$$
+K^2 > 4\frac{2K_2^2(L+m) + d + m}{m}.
+$$
+\end{lem}
+
+\begin{proof} Note that the condition {\bf (H1)} implies that for any $x,y \in \rset^d$,
+\[
+|x-y-\gamma(\nabla U(x) - \nabla U(y))|^2 \leq (1 + 2L\gamma + \gamma^2)|x-y|^2
+\] 
+Hence,   \cite[Corollary 5]{DurmusBortoli} implies that
+\[
+\|\delta_xQ^{[1/\gamma]}_{\gamma} - \delta_yQ^{[1/\gamma]}_{\gamma}\|_{\text{TV}} \leq 2(1-\varepsilon)
+\]
+with $\varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right)$ for all $x,y$ such that $|x|,|y| \leq K$, which means that the kernel $Q_{\gamma}$ satisfies $([1/\gamma],\varepsilon)$-Doeblin condition on the set $\{x: V(x) < 1+K^2\}$.
+Together with drift condition \eqref{eq:lyapunov} it allows to apply \cite[Theorem~19.4.1]{moulines2018} with appropriate constants.
+\end{proof}
+
+\section{Covariance estimation for $V-$geometrically ergodic Markov chains}\label{sec:appendix_moments}
+In this section we assume that $(X_p)_{p \in \nset_0}$ is a $V-$geometrically ergodic Markov chain and prove bounds on the variance of ergodic average $\pi_n^N(f)$ of the form \eqref{eq:29032018a2}.
+We use the same technique as in \cite{belomestny2019esvm} to control autocovariances for a given Markov chain.
+We start from some auxiliary lemmas:
+\begin{lem}
+\label{lem:fp_covariance}
+Let $(X_p)_{p\geq 0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and let $f$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$.
+Set $\tilde{f}(x) = f(x) - \pi(f),$ then it holds for some constant $C > 0,$
+\begin{equation}
+|\mathsf{E}_x[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq 2C\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}V(x),
+\end{equation}
+\begin{equation}
+|\mathsf{E}_{\pi}[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq 2C\pi(V)\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}.
+\end{equation}
+\end{lem}
+\begin{lem}
+\label{lem:covariance}
+Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and let $f$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$.
+Set $\tilde{f}(x) = f(x) - \pi(f),$ then
+\begin{equation}
+\label{eq:cov_v_ergodic}
+\PCov_x \left[f(X_k),f(X_{k+s})\right] \leq C^2V^2(x)\rho^{2k+s} + C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x) + C\pi(V)\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}
+\end{equation}
+\end{lem}
+\begin{proof}
+Note that
+\begin{eqnarray*}
+\PCov_x \left[f(X_k),f(X_{k+s})\right] &=& \PE_x\left[f(X_k) - \pi(f)\right]\left[f(X_{k+s})-\pi(f)\right] 
+\\
+&&+ \left[\pi(f) - \PE_xf(X_{k+s})\right]\left[\PE_xf(X_k) - \pi(f)\right].
+\end{eqnarray*}
+Due to $V-$ergodicity \eqref{eq:v_ergodicity},
+\[
+\left|\left[\pi(f) - \PE_xf(X_{k+s})\right]\left[\PE_xf(X_k) - \pi(f)\right]\right| \leq C^2V^2(x)\rho^{2k+s}.
+\]
+To bound the first term note that
+\begin{multline*}
+\left|\PE_x\left[f(X_k) - \pi(f)\right]\left[f(X_{k+s})-\pi(f)\right] - \PCov_{\pi}\left[f(X_k),f(X_{k+s})\right]\right|
+\\
+\leq \int\limits_{\rset^d}\left|\PE_y\left[\tilde{f}(X_0)\tilde{f}(X_{s})\right]\right| |\P^k(x,dy)-\pi(dy)|
+\\
+\leq 2C^2\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}\|\P^k(x,dy)-\pi(dy)\|_{V}
+\leq 2C^2\rho^{k+s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}V(x),
+\end{multline*}
+where we have used Lemma~\ref{lem:fp_covariance} and ergodicity of $(X_k)_{k\geq 0}.$ This implies \eqref{eq:cov_v_ergodic}.
+\end{proof}
+Now we state and prove the main result of this section on the variance bound for the estimator $\pi_N^n(f)$ in the case of a $V-$geometrically ergodic Markov chain.
+\begin{lem}
+\label{lem:var_bound}
+Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and let $f$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$.  Then
+\begin{equation}
+\label{eq:vanilla_var}
+\PVar_x \left[ \pi_{N}^n(f) \right] \leq \frac{C_1\pi(V)\|\tilde{f}\|^2_{V^{1/2}}}{n(1-\rho^{1/2})} + C_2\left(\frac{V(x) + 1}{n(1-\rho)}\right)^2
+\end{equation}
+with constants $C_1,C_2 > 0$ and $\rho\in (0,1)$ from \eqref{eq:v_ergodicity}.
+\end{lem}
+\begin{proof}
+Note that
+\begin{eqnarray*}
+\PVar_x \left[ \frac{1}{n}\sum\limits_{k=N+1}^{N+n}f(X_k) \right] = \underbrace{\frac{1}{n^2}\sum\limits_{k=N+1}^{N+n}\PVar_x \left[ f(X_k) \right]}_{S_1} + \underbrace{\frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\sum\limits_{s=1}^{n-k-1}\PCov_x\left[f(X_k),f(X_{k+s})\right]}_{S_2}.
+\end{eqnarray*}
+Now we bound each summand using Lemma~\ref{lem:covariance}:
+\[
+S_1 \leq \frac{1}{n}C\pi(V)\|\tilde{f}\|_{V^{1/2}} + \frac{\rho^N C^2\left(V^2(x) + V(x)\|\tilde{f}\|_{V^{1/2}}\right)}{n^2(1-\rho)},
+\]
+\begin{eqnarray*}
+S_2 & \leq & \frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\left[\frac{C^2V^2(x)\rho^{2k+1}}{1-\rho} + \frac{2C^2V(x)\|\tilde{f}\|^2_{V^{1/2}}\rho^{k+1/2}}{1-\rho^{1/2}} + \frac{2C\|\tilde{f}\|^2_{V^{1/2}}\pi(V)}{1-\rho^{1/2}}\right] \leq \\
+& \leq & \frac{2C^2\rho^{2N}V^2(x)}{n^2(1-\rho)^2} + \frac{4C^2\rho^{N+1/2}\|\tilde{f}\|^2_{V^{1/2}}V(x)}{n^2(1-\rho)(1-\rho^{1/2})} + \frac{4C\|\tilde{f}\|^2_{V^{1/2}}\pi(V)}{n(1-\rho^{1/2})}
+\end{eqnarray*}
+Hence \eqref{eq:vanilla_var} follows.
+\end{proof}
+
+\bibliographystyle{plain}
+\bibliography{refs-1}
+\end{document}
+
+\begin{cor}
+Suppose that {\bf (H1)} is satisfied and {\bf (H2)} holds with function $\Delta$ satisfying $\|\Delta\|_\infty\leq D.$   Moreover assume that $\mathsf{D}^{2}U$ is uniformly Lipschitz with a Lipschitz constant $L_2$ satisfying
+\begin{eqnarray*}
+L^2_2\leq c\min_{i=1,\ldots,d}\int (\partial_{x_i} U(x))^2\pi(x)\,dx 
+\end{eqnarray*}
+for some constant $c$ depending only on $L,m$ and $R.$ Then for $\gamma <4/(m+L)$
+\begin{eqnarray*}
+\max_{j\geq N}\sum_{p=j+1}^{\infty}\gamma\left\Vert\prod_{k=j+1}^{p} (I-\gamma\mathsf{D}^{2}U(X^x_{k-1}))\right\Vert _{r}\leq  C\exp\left( \frac{8L_2 |x|}{mL}\right)
+\end{eqnarray*}
+for some $C=C(L_2,m,L).$
+\end{cor}
+
+\appendix
+\section{Bounds for moments of  ULA}\label{sec:appendix}
 
@@
   \end{align*}
-Now the result follows given that $K_1 (1 +L/m)   / \| x \| \geq \frac{1}{2}$.
+which proves the first statement. The second statement is obvious.
 \end{proof}
@@
 \label{lem:drift}
- Assume {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$.
- Then, for any $\gamma \in (0,\bar{\gamma}]$ with $\overline{\gamma} = m/2L^2$  the kernel $Q_\gamma$ from \eqref{eq:ula_kernel} satisfies drift condition
+ Assume that {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$.
+ Then, for any $\gamma \in (0,\bar{\gamma}]$ with $\overline{\gamma} = m/4$  the kernel $Q_\gamma$ from \eqref{eq:ula_kernel} satisfies drift condition
  \begin{equation}
 \label{eq:lyapunov}
-Q_{\gamma}V(x) \leq \rho^{\gamma} V(x) + \gamma C
-\end{equation}
-with   $\rho = \exp{\left(-\frac{m}{2}\right)}$, $C = 2K_2^2(L+m) + d + m$ with $K_2$ from Lemma~\ref{lem:quadratic_behaviour}, and $m$ from {\bf (H2)}.
+Q_{\gamma}V(x) \leq \lambda^{\gamma} V(x) + \gamma C
+\end{equation}
+with   $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = 25K_2^2/8 + d + m$ with $K_2$ from Lemma~\ref{lem:quadratic_behaviour}, and $m$ from {\bf (H2)}.
 \end{lem}
@@
 \[
-\|x - \gamma \nabla U(x) \|^2 = \|x\|^2 - 2\gamma \langle \nabla U(x), x\rangle + \gamma^2 \|\nabla U(x)\|^2 \leq \left(1 - \gamma m + \gamma^2 L^2\right)\|x\|^2
-\]
-Since $\gamma < \frac{m}{2L^2}$, we obtain
-$$
-Q_{\gamma}V(x) \leq (1-\gamma m + \gamma^2 L^2)V(x) + \gamma (d + m - \gamma L^2) \leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma (d + m)
-$$
-Consider now the case $x \in B(0,K_2)$. Then simply using $\|x-\gamma \nabla U(x)\|^2 \leq (1+L\gamma)^2\|x\|^2$, we obtain
-$$
-Q_{\gamma}V(x) \leq (1-\gamma m + \gamma^2 L^2)V(x) + \gamma\left((m - \gamma L^2)(1+\|x\|^2) + d + L(2+L\gamma)\|x\|^2\right) \leq
-$$
-
-$$
-\leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma(2K_2^2(L+m) + d + m)
+\|x - \gamma \nabla U(x) \|^2 = \|x\|^2 - 2\gamma \langle \nabla U(x), x\rangle + \gamma^2 \|\nabla U(x)\|^2 \leq \left(1 - \gamma m + 2\gamma^2 L^2\right)\|x\|^2
+\]
+Since $\gamma < \frac{m}{4L}$, we obtain
+$$
+Q_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma d + (\gamma m - 2\gamma^2 L^2) \leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma (d + m)
+$$
+Consider now the case $x \in B(0,K_2)$. Then simply using $\|x-\gamma \nabla U(x)\|^2 \leq 2(1+L\gamma)^2\|x\|^2$, we obtain
+$$
+Q_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma\left((m - 2\gamma L^2)(1+\|x\|^2) + d + 2(1+L\gamma)^2\|x\|^2\right) \leq
+$$
+
+$$
+\leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma(\frac{25K_2^2}{8} + d + m)
 $$
@@
 \label{lem:v_ergodicity}
- Assume  {\bf (H1)} and {\bf (H2)}. Then for $0 < \gamma < \overline{\gamma}= \min(m / 4L^2,1/L)$, and for any $x \in \rset^d$  it holds
-$$
-d_V(\delta_xQ_{\gamma}^{n},\pi_\gamma) \leq C\rho^{n}V(x)
+ Assume  {\bf (H1)} and {\bf (H2)}. Then for $0 < \gamma < \overline{\gamma}=m / 4L^2$, for any $x \in \rset^d$ it holds
+$$
+d_V(\delta_xQ_{\gamma}^{1,n},\pi_\gamma) \leq C\rho^n\left(V(x) + \pi_\gamma(V)\right)
 $$
@@
 $$
-C = \left(1+\pi_\gamma(V)\right)\left(1 + \exp{\left(-\frac{m\gamma}{2}\right)}\right)\left(1+\frac{b\exp{\left(-\frac{m\gamma}{2}\right)} + K^2 + 1}{(1-\varepsilon)(1-\exp{\left(-\frac{m\gamma}{2}\right)} - \frac{2b}{2+K^2})}\right);
-$$
-
-$$
-b = \gamma\left(2K_2^2(L+m) + d + m\right); \quad \varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right); \quad
-$$
-
-$$
-\rho = \exp\left\{\frac{\gamma m}{4}\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{m\gamma}{2}\right)} + \frac{2b}{K^2+2})}}\right\}
-$$
-where $K_2 > 0$ is from Lemma~\ref{lem:quadratic_behaviour} and $K>0$ is a positive parameter such that 
-$$
-K^2 > 4\frac{2K_2^2(L+m) + d + m}{m}
+C = \left(1 + \exp{\left(-\frac{m\gamma}{2}\right)}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp{\left(-\frac{m\gamma}{2}\right)} - \frac{2b}{1+d})}\right);
+$$
+
+$$
+b = \gamma(\frac{25K_1^2}{8} + 2d + m); \quad \overline{b} = b\exp{\left(-\frac{m\gamma}{2}\right)} + d; \quad \varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right); \quad
+$$
+
+$$
+\rho = \exp{\left(-\gamma\left(\frac{m}{2} - 2\frac{\frac{25K_1^2}{8} + 2d + m}{d}\right)\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{m\gamma}{2}\right)} + \frac{2b}{d+1})}}\right)}
 $$
@@
 
-\begin{proof} Note that the condition {\bf (H1)} implies that for any $x,y \in \rset^d$,
-\[
-\|x-y-\gamma(\nabla U(x) - \nabla U(y))\|^2 \leq (1 + 2L\gamma + \gamma^2)\|x-y\|^2
-\] 
-Hence, the result of [Corollary 5]\cite{DurmusBortoli} implies that
-\[
-\|\delta_xQ^{[1/\gamma]}_{\gamma} - \delta_yQ^{[1/\gamma]}_{\gamma}\|_{\text{TV}} \leq 2(1-\varepsilon)
-\]
-with $\varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right)$ for all $x,y$ such that $\|x\|,\|y\| \leq K$, which means that the kernel $Q_{\gamma}$ satisfies $([1/\gamma],\varepsilon)$-Doeblin condition on the set $\{x: V(x) < 1+K^2\}$.
+\begin{proof} Note that the condition {\bf (H1)} implies that the Markov kernel $Q^{1,n}_{\boldsymbol{\gamma}}$ satisfies $(1,\varepsilon)$-Doeblin condition with $\varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right)$.
 Together with drift condition \eqref{eq:lyapunov} it allows to apply \cite[Theorem~19.4.1]{moulines2018} with appropriate constants.
@@
 Let $\tilde{f}(x) = f(x) - \pi(f)$.
-Then it holds with constant $C > 0$ and $0 < \rho <1$ from \eqref{eq:v_ergodicity} that
-\begin{equation}
-\label{eq:fp_cov}
-|\mathsf{E}_x[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq 2C\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}V(x)
+Then it holds for some constant $C > 0$
+\begin{equation}
+|\mathsf{E}_x[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq C\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x)
 \end{equation}
@@
 \begin{equation}
-\label{eq:fp_cov_stationary}
-|\mathsf{E}_{\pi}[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq 2C\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}\pi(V)
+|\mathsf{E}_{\pi}[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq C\pi(V)\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}
 \end{equation}
@@
 \[
-|\mathsf{E}_x[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq |\tilde{f}(x)|^2 \mathsf{E}_x|\tilde{f}(X_s)| \leq \|\tilde{f}\|_{V^{\frac{1}{2}}}V^{\frac{1}{2}}(x)\int\limits_{\rset^d} V^{\frac{1}{2}}(y)|\mathsf{P}^{s}(x,\,dy)-\pi(dy)|
+|\mathsf{E}_x[\tilde{f}(X_0)\tilde{f}(X_s)]| \leq |\tilde{f}(x)| \mathsf{E}_x|\tilde{f}(X_s)| \leq \|\tilde{f}\|_{V^{\frac{1}{2}}}V^{\frac{1}{2}}(x)\int\limits_{\rset^d} V^{\frac{1}{2}}(y)|\mathsf{P}^{s}(x,\,dy)-\pi(dy)|
 \]
@@
 \label{eq:cov_v_ergodic}
-\PCov_x \left[f(X_k),f(X_{k+s})\right] \leq C^2V^2(x)\rho^{2k+s} + 2C^2\rho^{k+s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}V(x) + 2C\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}\pi(V)
-\end{equation}
-with constants $C > 0$ and $0 < \rho < 1$ from \eqref{eq:v_ergodicity}.
+\PCov_x \left[f(X_k),f(X_{k+s})\right] \leq C^2V^2(x)\rho^{k+s} + C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x) + C\pi(V)\rho^{s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}
+\end{equation}
 \end{lem}
@@
 \[
-\PCov_x \left[f(X_k),f(X_{k+s})\right] = \PE_x\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right] + \left[\pi(f) - \PE_xf(X_{k+s})\right]\left[\PE_xf(X_k) - \pi(f)\right]
-\]
-Due to $V-$ergodicity \eqref{eq:v_ergodicity},
-\[
-\left|\left[\pi(f) - \PE_xf(X_{k+s})\right]\left[\PE_xf(X_k) - \pi(f)\right]\right| \leq C^2V^2(x)\rho^{2k+s}
+\PCov_x \left[f(X_k),f(X_{k+s})\right] = \PE_x\left[f(X_k) - \pi(f)\right]\left[f(X_{k+s})-\pi(f)\right] + \left[\pi(f) - \PE_xf(X_s)\right]\left[\PE_xf(X_k) - \pi(f)\right]
+\]
+Due to $V-$ergodicity,
+\[
+\left|\left[\pi(f) - \PE_xf(X_s)\right]\left[\PE_xf(X_k) - \pi(f)\right]\right| \leq C^2V^2(x)\rho^{k+s}
 \]
@@
 \begin{multline*}
-\left|\PE_x\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right] - \PCov_{\pi}\left[f(X_k),f(X_{k+s})\right]\right| = \left|\PE_x\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right] - \PE_\pi\left[\tilde{f}(X_{k})\tilde{f}(X_{k+s})\right]\right| \leq
+\left|\PE_x\left[f(X_k) - \pi(f)\right]\left[f(X_{k+s})-\pi(f)\right] - \PCov_{\pi}\left[f(X_k),f(X_{k+s})\right]\right|
 \\
 \leq \int\limits_{\rset^d}\left|\PE_y\left[\tilde{f}(X_0)\tilde{f}(X_{s})\right]\right| |\P^k(x,dy)-\pi(dy)|
-\leq \{\eqref{eq:fp_cov}\}\leq 2C\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}\int\limits_{\rset^d} V(y) |\P^k(x,dy)-\pi(dy)| \leq 
-\\
-\leq 2C^2\rho^{s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}\|\P^k(x,dy)-\pi(dy)\|_{V} \leq 2C^2\rho^{k+s/2}\|\tilde{f}\|^2_{V^{\frac{1}{2}}}V(x)
+\leq C^2\rho^{k+s/2}\|\tilde{f}\|_{V^{\frac{1}{2}}}V(x),
 \end{multline*}
-where $\left|\PCov_{\pi}\left[f(X_k),f(X_{k+s})\right]\right|$ can be bounded using \eqref{eq:fp_cov_stationary}. 
+where the last inequality is due to lemma~\ref{lem:fp_covariance}, which implies \eqref{eq:cov_v_ergodic}.
 \end{proof}
@@
 \begin{lem}
-\label{lem:var_bound}
 Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$.
-Let $\tilde{f}(x) = f(x) - \pi(f)$.
+Assume also that $n(1-\rho^{1/2}) > 1$.
 Then
@@
 \label{eq:vanilla_var}
-\PVar_x \left[ \pi_{N}^n(f) \right] \leq \frac{C_1\pi(V)\|\tilde{f}\|^2_{V^{1/2}}}{n(1-\rho^{1/2})} + C_2\left(\frac{\rho^NV(x) + 1}{n(1-\rho^{1/2})}\right)^2
-\end{equation}
-with positive constants $C_1,C_2 > 0$ and $0 < \rho <1$ from \eqref{eq:v_ergodicity}.
+\PVar_x \left[ \pi_{N}^n(f) \right] \leq \frac{C\pi(V)\|\tilde{f}\|_{V^{1/2}}}{n(1-\rho^{1/2})} + \mathcal{O}(\frac{1}{n^2(1-\rho^{1/2})})
+\end{equation}
 \end{lem}
-
 \begin{proof}
@@
 \[
-S_1 \leq \frac{2C\pi(V)\|\tilde{f}\|^2_{V^{1/2}}}{n} + \frac{\rho^N C^2V(x)\left(\rho^NV(x) + 2\|\tilde{f}\|^2_{V^{1/2}}\right)}{n^2(1-\rho)}
-\]
-\begin{eqnarray*}
-S_2 & \leq & \frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\left[\frac{C^2V^2(x)\rho^{2k+1}}{1-\rho} + \frac{2C^2V(x)\|\tilde{f}\|^2_{V^{1/2}}\rho^{k+1/2}}{1-\rho^{1/2}} + \frac{2C\|\tilde{f}\|^2_{V^{1/2}}\pi(V)}{1-\rho^{1/2}}\right] \leq \\
-& \leq & \frac{2C^2\rho^{2N}V^2(x)}{n^2(1-\rho)^2} + \frac{4C^2\rho^{N+1/2}\|\tilde{f}\|^2_{V^{1/2}}V(x)}{n^2(1-\rho)(1-\rho^{1/2})} + \frac{4C\|\tilde{f}\|^2_{V^{1/2}}\pi(V)}{n(1-\rho^{1/2})}
+S_1 \leq \frac{1}{n}C\pi(V)\|\tilde{f}\|_{V^{1/2}} + \frac{\rho^N C^2\left(V^2(x) + V(x)\|\tilde{f}\|_{V^{1/2}}\right)}{n^2(1-\rho)}
+\]
+\begin{eqnarray*}
+S_2 & \leq & \frac{2}{n^2}\sum\limits_{k=N+1}^{N+n-1}\left[C^2V^2(x)\rho^{k+1}\frac{1}{1-\rho} + C^2V(x)\rho^{k+1/2}\frac{1}{1-\rho^{1/2}} + C\pi(V)\frac{1}{1-\rho^{1/2}}\|\tilde{f}\|_{V^{1/2}}\right] \leq \\
+& \leq & \frac{2\rho^NC^2V^2(x)}{n^2(1-\rho)^2} + \frac{2\rho^NC^2V(x)}{n^2(1-\rho)(1-\rho^{1/2})} + \frac{C\pi(V)\|\tilde{f}\|_{V^{1/2}}}{n(1-\rho^{1/2})}
 \end{eqnarray*}
@@
 \label{lem:exp_bound}
-Let $(X_p)_{p \geq 0}$ be the ULA-generated Markov chain of the form
-\[
-X_{p+1} = X_p + \gamma \mu(X_p) + \sqrt{\gamma}\xi_{p+1}, \xi_{p+1} \sim \mathcal{N}(0,1)
-\]
- with $0 < \gamma < 4(m+L)^{-1}, \quad \mu(x) = \mu_{0}(x) + \Delta(x)$, where $\mu_0^{\prime}(x) <-m$ for some $m > 0$, $\mu_0(x) \in Lip(L)$. Denote by 
- \[
- \pi_{\mu_0}(y) = \frac{\exp\left\{\int\limits_{0}^{y}\mu_{0}(x)\,dx\right\}}{\int\limits_{-\infty}^{+\infty}\exp\left\{\int\limits_{0}^{y}\mu_{0}(x)\,dx\right\}\,dy}
- \]
- the stationary distribution of the corresponding continuous-time diffusion.
- Suppose that there exists $M > 0,$ such that $\left|\Delta(x)\right| \leq M$.
- Then for any Lipschitz function $f$ and for any $ N \geq 1$,
-\begin{equation}
+Let $(X_p)_{p \geq 0}$ be a Markov chain of the form
+\[
+X_{p+1} = X_p - \gamma \nabla U(X_p) + \sqrt{\gamma}\xi_{p+1}, \quad \xi_{p+1} \sim \mathcal{N}(0,1),
+\]
+where $U:$ $\mathbb{R}^d\to \mathbb{R}$ is a potential.
+ Assume that $U(x) = U_{0}(x) + \Delta(x),$ where $U_0 \in \mathsf{Lip}(L),$
+ \[
+ \ps{\alpha\mathsf{D}^2 U_0(x)}{\alpha}>m, \quad \left|\Delta(x)\right| \leq M,\quad x\in \mathbb{R}^d
+ \]
+ for all $\alpha\in \mathbb{R}^d$ with $|\alpha|=1$ and some $m,M > 0.$   Then for any natural $n,N,$  $0 < \gamma < \frac{4}{m+L}$ and any function $f\in \mathsf{Lip}(L_f)$ it holds
+\begin{multline}
 \label{eq:exp_bound}
-\PE_x \rme^{u\gamma\sum\limits_{k=N+1}^{N+n}f(X_k)} \leq C(x)\exp\left\{n\gamma\left(2u^2\kappa^{-2} \|f\|^2_{\text{Lip}}\left(1+\frac{1}{n\gamma}\left(\frac{4}{m+L} + \kappa^{-1}\right)\right) + M^2 + C_1u\sqrt{\gamma} +  u\pi_{\mu_0}(f)\right)\right\}
-\end{equation}
-with $\kappa = 2mL(m+L)^{-1}$, positive constant $C_1$ depending only on $m,L$ and 
-\[
-C(x) = \exp\left\{\frac{4u\|f\|_{\text{Lip}}}{mL}\int_{\rset}|x-y|\,\pi_{\gamma}(dy)\right\}
+\PE_x \exp \left(u\gamma\sum\limits_{k=N+1}^{N+n}f(X_k)\right)\leq
+ C_1(x)\exp\Bigl (n\gamma (2u^2\kappa^{-2} L_f^2+M^2 + C_2u\sqrt{\gamma} +   u\pi(f))\Bigr)
+\end{multline}
+with   $\kappa=\frac{2mL}{m+L},$
+\[ C_1(x) = \exp\left\{\frac{4u L_f}{mL}\int_{\rset}|x-y|\,\pi_{\gamma}(dy)+2u^2\kappa^{-2} L_f^2\left(\frac{4}{m+L} + \kappa^{-1}\right)\right\}
+\]
+ and 
+\[
+C_2=2L\sqrt{\frac{mL}{m+L}\left(\gamma + \frac{mL}{2(m+L)}\right)\left(2 + \frac{L^2\gamma}{m} + \frac{L^2\gamma^2}{6}\right)}.
 \] 
 \end{lem}
-\begin{proof} The Markov Chain $(X_p^x)$ has the form
-\[
-X_{k+1} = X_k + \gamma(\mu_0(X_k) + \Delta(X_k)) + \sqrt{\gamma}\xi_{k+1}, X_0 = x, \xi_{k+1} \sim \mathcal{N}(0,1)
-\]
-Consider the following auxiliary Markov Chain $(Y_p^x)$, corresponding to the diffusion with strongly log-concave potential
-\[
-Y_{k+1} = Y_k + \gamma\mu_0(Y_k) + \sqrt{\gamma}\xi_{k+1}, Y_0 = x, \xi_{k+1} \sim \mathcal{N}(0,1)
-\]
-The main idea of the proof is to change expectation w.r.t. $(X_p^x)$ to the expectation w.r.t. $(Y_p^x)$, which is easier to work with:
-\begin{eqnarray*}
-\PE_x \rme^{u\gamma\sum\limits_{k=N+1}^{N+n}f(X_k)} &=& \PE_x  \exp\left\{u\gamma\sum_{k=N+1}^{N+n}\left(f(Y_k) -\frac{(Y_k - Y_{k-1} - \gamma\mu(Y_{k-1}))^2 - (Y_k - Y_{k-1} - \gamma\mu_0(Y_{k-1}))^2}{2\gamma}\right)\right\} = \\
-& = & \PE_x \exp\left\{u\gamma\sum_{k=N+1}^{N+n}f(Y_k) + \sum_{k=N+1}^{N+n}\left[(Y_k-Y_{k-1})\Delta(Y_{k-1}) - \gamma\mu_0(Y_{k-1})\Delta(Y_{k-1}) - \frac{\gamma}{2}\Delta^2(Y_{k-1})\right]\right\} = \\
- &=& \PE_x \exp\left\{u\gamma\sum_{k=N+1}^{N+n}f(Y_k) + \sum_{k=N+1}^{N+n}\left[\sqrt{\gamma}\Delta(Y_{k-1})\xi_{k} - \frac{\gamma}{2}\Delta^2(Y_{k-1})\right]\right\} \leq \\
- & \leq & \left(\PE_x \exp\left\{2\sqrt{\gamma}\sum_{k=N+1}^{N+n}\Delta(Y_{k-1})\xi_{k}\right\}\right)^{1/2} \left(\PE \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}f(Y_k)\right\}\right)^{1/2}
-\end{eqnarray*}
-with the last line obtained from Holder inequality. We bound the first expectation using Azuma-Hoeffding inequality, that is, setting $\mathcal{F}_k = \sigma(\xi_1,\ldots,\xi_k)$, 
-\[
-\PE_x \exp\left\{2\sqrt{\gamma}\sum_{k=N+1}^{N+n}\Delta(Y_{k-1})\xi_{k}\right\} = \PE_x\left(\exp\left\{2\sqrt{\gamma}\sum_{k=N+1}^{N+n-1}\Delta(Y_{k-1})\xi_{k}\right\} \PE\left[\exp\left\{2\sqrt{\gamma}\Delta(Y_{n-1})\xi_{n}\right\}| \mathcal{F}_{N+n-1}\right]\right) \leq
-\]
-\[
-\leq \PE_x\left(\exp\left\{2\sqrt{\gamma}\sum_{k=N+1}^{N+n-1}\Delta(Y_{k-1})\xi_{k}\right\}\exp\left\{2\gamma M^2\right\}\right) \leq \{\text{by induction}\} \leq \exp\left\{2n\gamma M^2\right\}
-\]
-To bound the second exponent, note that
-\[
-\PE \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}f(Y_k)\right\} =
-\PE \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}\left(f(Y_k) - \PE_xf(Y_k)\right)\right\} \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}\PE_x f(Y_k)\right\} 
-\]
-Now we can apply \cite[Proposition 32]{dm2016HighDim} to the first expectation with $\lambda = 2nu\gamma$, yielding the bound
-\begin{equation}
-\label{eq:1_bound}
-\PE \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}\left(f(Y_k) - \PE_xf(Y_k)\right)\right\} \leq \exp\left\{4u^2\kappa^{-2} n\gamma \|f\|^2_{\text{Lip}}\left(1+\frac{1}{n\gamma}\left(\frac{4}{m+L} + \kappa^{-1}\right)\right)\right\}
-\end{equation}
-Note that for ULA algorithm the distribution $\P_x$ is Gaussian for any $x \in \rset^d$, thus, following \cite[Proposition 3]{dm2016HighDim}, we may bound the Wasserstein-1 distance between $\P_x$ and $\P_y$ by
-\[
-W_1(\P_x,\P_y) = W_1(\mathcal{N}(x-\gamma\mu(x),\gamma),\mathcal{N}(y-\gamma\mu(y),\gamma)) = \|x-y-\gamma(\mu(x)-\mu(y))\| \leq \sqrt{1-\kappa\gamma}\|x-y\|
-\]
-where $\kappa = \frac{2mL}{m+L}$. Hence, it holds
-\[
-W_1(\P_x,\P_y) \leq \left(1-\frac{\kappa\gamma}{2}\right)\|x-y\|
-\]
-which means that the Markov kernel $\P$ is a strict contraction in the Wasserstein metric. This allows to use \cite[Proposition 1]{joulin2010}, stating that
-\begin{equation}
-\label{eq:2_bound}
-\left|\sum_{k=N+1}^{N+n}\PE_x f(Y_k) - n\pi_{\gamma}(f)\right| \leq \frac{2\left(1-\frac{\kappa\gamma}{2}\right)^{N+1}}{\kappa\gamma}\|f\|_{\text{Lip}}\int_{\rset}|x-y|\,\pi_{\gamma}(dy) \leq \frac{m+L}{mL\gamma}\|f\|_{\text{Lip}}S(x)
-\end{equation}
-with $S(x) = \int_{\rset}|x-y|\,\pi_{\gamma}(dy)$. Then by \cite[Corollary 7]{dm2016HighDim}, 
-\begin{equation}
-\label{eq:3_bound}
-|\pi_{\gamma}(f) - \pi(f)| \leq W_1(\pi_\gamma,\pi) \leq W_2(\pi_\gamma,\pi) \leq C_1\sqrt{\gamma}
-\end{equation}
-where $C_1 = 2L\sqrt{\frac{mL}{m+L}\left(\gamma + \frac{mL}{2(m+L)}\right)\left(2 + \frac{L^2\gamma}{m} + \frac{L^2\gamma^2}{6}\right)}$.
-Now we put together bounds \eqref{eq:1_bound}, \eqref{eq:2_bound} and \eqref{eq:3_bound}, leading to the final bound
-\[
-\PE \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}f(Y_k)\right\} \leq \exp\left\{4u^2\kappa^{-2} n\gamma \|f\|^2_{\text{Lip}}\left(1+\frac{1}{n\gamma}\left(\frac{4}{m+L} + \kappa^{-1}\right)\right) + 2u\gamma \frac{m+L}{mL}\|f\|_{\text{Lip}}S(x) + 2n\gamma u(\pi(f) + C_1\sqrt{\gamma})\right\}
-\]
-Hence the statement of lemma follows.
-\end{proof}
-
-%Aforementioned lemmas allow us to bound the exponential moment of the additive functional of ULA. Namely, the following theorem holds
-%
-%\begin{thm}
-%\label{thm:ula_const_size}
-%Suppose that ULA kernel \ref{eq:kernel} satisfies assumptions {\bf (H1)} and {\bf (H2)}, and let $X_1,\ldots,X_n,\ldots$ be generated by ULA with constant step size $\gamma$.
-Let $X_0 = x$ be fixed.
-Then for any bounded function $g(x) : \mathbb{R}^d \rightarrow \mathbb{R}, |g(x)| \leq M$, it holds
-%\begin{eqnarray*}
-%\mathsf{E}\left[\exp\left(-\rho\gamma\sum_{k=l}^{p}\left(g(X_{k})-\pi_{\gamma}(g)\right)\right)\right] \leq C_1\exp{\left(s^2 \kappa \gamma^2 (p-l+1)\right)}
-%\end{eqnarray*}
-%for some absolute constant $C_1$ not depending on $l,p$.
-%\end{thm}
-%
-%\begin{proof} Conditions {\bf (H1)} and {\bf (H2)} imply that the Markov kernel $R_\gamma$ is $V-$ergodic (lemma \ref{lem:v_ergodicity}) and satisfies drift condition \ref{lem:drift}.
-Due to \cite[Theorem~3, Fact 3]{moulines:bound_dif:2019}, we obtain the following bound
-%\begin{align*}
-%\mathsf{E}_x\left[\exp\left(-s\gamma\sum_{k=l}^{p}\left(g(X_{k}) - \mathsf{E}_x g(X_k)\right)\right)\right] \leq \exp{\left(s^2 \kappa \gamma^2 (p-l+1)\right)}
-%\end{align*}
-%\sergey{I will put precise constant later}
-%with constant $\kappa = ...$. Note that
-%\begin{align*}
-%\mathsf{E}\left[\exp\left(-s\gamma\sum_{k=l}^{p}\left(g(X_{k})-\pi_{\gamma}(g)\right)\right)\right] =
-%\end{align*}
-%
-%\begin{align*}
-%=\mathsf{E}_x\left[\exp{\left(-s\gamma\sum_{k=l}^{p}\left(g(X_{k}) - \mathsf{E}_x g(X_k)\right)\right)}\right]
-%\exp{\left(s\gamma\sum\limits_{k=l}^{p}\left(\pi_\gamma (g) - \mathsf{E}_x g(X_k)\right)\right)}
-%\end{align*}
-%Using lemma~\ref{lem:v_ergodicity},
-%\begin{align*}
-%|\mathsf{E} g(X_k) - \pi_\gamma(g)| \leq Md_V(\pi_\gamma, \delta_xQ^{1,n}_{\boldsymbol{\gamma}}) \leq CM\rho^k\left(V(x) + \pi_\gamma(V)\right)
-%\end{align*}
-%Hence,
-%\begin{align*}
-%\exp{\left(s\gamma\sum\limits_{k=l}^{p}\left(\pi_\gamma (g) - \mathsf{E}_x g(X_k)\right)\right)} \leq \exp{\left(s\gamma CM(V(x) + \pi_\gamma(V))\frac{\rho^l - \rho^{p+1}}{1-\rho}\right)}
-%\end{align*}
-%and the statement follows.
-%\end{proof}
-%
-%Let us formulate and prove prove analogue of the previous theorem for inhomogeneous Markov chain with step size $\gamma_1,\ldots,\gamma_n,\ldots$.
-Note that under assumptions $\sum_{p=1}^{\infty} \gamma_p=\infty, \quad \sum_{p=1}^{\infty} \gamma^2_p<\infty$ and {\bf (H1)} the stationary distribution of ULA-based chain will be equal to $\pi$.
-We will need additional assumption
-%\begin{description}
-%\item[{\bf (H3)}] There exist such constants $\rho > 0, K_3 > 0$ that
-%\begin{eqnarray*}
-%d_V(\delta_xQ_{\boldsymbol{\gamma}}^{1,n},\pi) \leq K_3 V(x) \rho^n
-%\end{eqnarray*}
-%\end{description}
-%
-%\begin{thm}
-%\label{th:exp_mom_ula}
-%Suppose that
-%\begin{eqnarray*}
-%\sum_{p=1}^{\infty} \gamma_p=\infty, \quad \sum_{p=1}^{\infty} \gamma^2_p<\infty.
-%\end{eqnarray*}
-%Under assumptions {\bf (H1)}, {\bf (H2)} and {\bf (H3)}, for any bounded function $g$ on $\mathbb{R}^d$
-%we have
-%%\begin{eqnarray*}
-%%\mathsf{E}\left[\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}g(X_{p})-\bar Q_l(X_l)\right)^{4}\right] \lesssim \frac{1}{\Gamma^4_{N+2,N+n+1}}
-%%\end{eqnarray*}
-% %and
-%%\begin{eqnarray*}
-%%\bar Q_l(X_l)=\mathsf{E}\left[\left.\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right|X_{l}\right].
-%%\end{eqnarray*}
-%%Moreover,
-%\begin{eqnarray}
-%\label{eq:exp_decrease_step}
-%\mathsf{E}\left[\exp\left(-s\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k})-\left(\sum_{k=l}^{p}\gamma_{k+1}\right)\pi(g)\right)\right)\right]\leq C
-%\end{eqnarray}
-%for all natural $0<l\leq p<\infty$ with constant $C$ not depending on $l,p.$. Moreover,
-%\begin{eqnarray}
-%\label{eq:var_decrease_step}
-%\mathsf{Var}_x{\left(\sum_{k=l}^{l+n-1}\frac{\gamma_{k+1}}{\Gamma_{l+1,l+n}}g(X_k) - \mathsf{E}_x \sum_{k=l}^{l+n-1}\frac{\gamma_{k+1}}{\Gamma_{l+1,l+n}}g(X_k)\right)} \leq \frac{C_1}{\Gamma_{l+1,l+n}}
-%\end{eqnarray}
-%\end{thm}
-%\begin{proof}
-%
-% Note that under assumptions {\bf (H1)}, {\bf (H2)} and {\bf (H3)} it holds due to \cite[Theorem~4]{moulines:bound_dif:2019} that
-% \begin{align}
-% \label{eq:exp_bounded_dif}
-%\mathsf{E}_x\left[\exp\left(-s\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k}) - \mathsf{E}_x \left(\sum\limits_{k=l}^{p}\gamma_{k+1}g(X_k)\right)\right)\right)\right] \leq \exp{\left(s^2 \kappa \sum\limits_{k=l}^{p}\gamma_{k+1}^2\right)} \leq \exp{\left(s^2 \kappa \Gamma^*\right)}
-%\end{align}
-%\sergey{This theorem is not yet written in the second paper}
-%with $\Gamma^* = \sum\limits_{k=1}^{\infty}\gamma_k^2 < \infty$ and $\kappa = ...$. Using the same transformation as in theorem~\ref{thm:ula_const_size}, and using the bound
-%\begin{align*}
-%|\mathsf{E}_x f(X_k) - \pi(f)| \leq K_3 M \rho^{\Gamma_{1,k}}V(x)
-%\end{align*}
-%we obtain
-%\begin{align*}
-%\mathsf{E}\left[\exp\left(-s\left(\sum_{k=l}^{p}\gamma_{k+1}g(X_{k})-\left(\sum_{k=l}^{p}\gamma_{k+1}\right)\pi(g)\right)\right)\right] \leq \exp{\left(s^2 \kappa \Gamma^*\right)}\exp{\left(K_3 M V(x) \sum\limits_{k=l}^{p}\rho^{\Gamma_{1,k}}\gamma_{k+1}\right)} \leq
-%\end{align*}
-%
-%\begin{align*}
-%\leq \exp{\left(s^2 \kappa \Gamma^*\right)}\exp{\left(K_3 M V(x) \rho^{\Gamma_{1,k}} \int\limits_{0}^{+\infty}\rho^y\,dy\right)} = \exp{\left(s^2 \kappa \Gamma^*\right)}\exp{\left(K_3 M V(x) \frac{\rho^{\Gamma_{1,k}}}{\ln{\rho}}\right)}
-%\end{align*}
-%\end{proof}
-%which proves \ref{eq:exp_decrease_step}. Note that from \ref{eq:exp_bounded_dif} by Taylor expansion for small $s$ we obtain
-%\begin{align*}
-%1 + \frac{s^2}{2}\mathsf{Var}_x{\left(\sum_{k=l}^{l+n-1}\gamma_{k+1}g(X_k) - \mathsf{E}_x \sum_{k=l}^{l+n-1}\gamma_{k+1}g(X_k)\right)} + O(s^3) \leq 1 + s^2 \kappa \sum\limits_{k=l+1}^{l+n}\gamma_{k}^2 + O(s^4)
-%\end{align*}
-%Since it holds for arbitarily small $s$, the variance is bounded by
-%\begin{align*}
-%\mathsf{Var}_x{\left(\sum_{k=l}^{l+n-1}\gamma_{k+1}g(X_k) - \mathsf{E}_x \sum_{k=l}^{l+n-1}\gamma_{k+1}g(X_k)\right)} \leq 2\kappa \sum\limits_{k=l+1}^{l+n}\gamma_{k}^2 \leq 2\kappa \Gamma_{l+1,l+n}
-%\end{align*}
-%Now we need to divide both parts by $\Gamma^2_{l+1,l+n}$ to get \ref{eq:var_decrease_step}. The proof is completed.
-\bibliographystyle{plain}
-\bibliography{refs-1}
-\end{document}
-
-\par

### 2020-05-08 aap/Revision/vr-mcmc-revision-final.tex
update

@@
 Let $f$ be a bounded Borel function and $(X^x_{p})_{p \in \nset_0}$ be a Markov chain generated by ULA with constant step size $\gamma$.
-Let $c_0>0$. Then, there exist $C < \infty$ and $\gamma_0 > 0$, such that for all $n \in \nset$, $\gamma \in (0,\gamma_0]$ satisfying $n\gamma > c_0$, and $x \in \rset^d$,
+Then, there exist $C < \infty$ and $\gamma_0 > 0$, such that for all $n \in \nset$, $\gamma \in (0,\gamma_0]$ and $x \in \rset^d$,
 \begin{equation}
 \label{eq:var-mc}
-\PVar\left[\pi_n^N(f)\right]\leq C (n \gamma)^{-1} V(x) \eqsp.
+\PVar\left[\pi_n^N(f)\right]\leq C\left(\frac{1}{n\gamma} + \left(\frac{V(x)+1}{n\gamma}\right)^2\right) \eqsp.
 \end{equation}
@@
 
-Under assumptions {\bf (H1)} and {\bf (H2)} the Markov chain $(X^x_{p})_{p \in \nset_0}$ is $V-$geometrically ergodic in the sense of \eqref{eq:v_ergodicity} with $V(x) = 1+|x|^2$ and $\rho\in (0,1) $ specified in Lemma~\ref{lem:v_ergodicity}.
-Due to Lemma~\ref{lem:var_bound},
+Under assumptions {\bf (H1)} and {\bf (H2)} the Markov chain $(X^x_{p})_{p \in \nset_0}$ generated by ULA  is $V-$geometrically ergodic in the sense of \eqref{eq:v_ergodicity} with $V(x) = 1+|x|^2$ and $\rho\in (0,1) $ due to Lemma~\ref{lem:v_ergodicity}.
+It follows from Lemma~\ref{lem:var_bound}, that
 \[
@@
 \label{eq:lyapunov}
-Q_{\gamma}V(x) \leq \rho^{\gamma} V(x) + \gamma C
-\end{equation}
-with   $\rho = \exp{\left(-\frac{\mathfrak{m}}{2}\right)}$, $C = 2K_2^2(L+\mathfrak{m}) + d + \mathfrak{m}$ with $K_2$ from Lemma~\ref{lem:quadratic_behaviour}, and $\mathfrak{m}$ from {\bf (H2)}.
+Q_{\gamma}V(x) \leq \lambda^{\gamma} V(x) + \gamma b
+\end{equation}
+with   $\lambda = \exp{\left(-\frac{\mathfrak{m}}{2}\right)}$, $b = 2K_2^2(L+\mathfrak{m}) + d + \mathfrak{m}$ with $K_2$ from Lemma~\ref{lem:quadratic_behaviour}, and $\mathfrak{m}$ from {\bf (H2)}.
 \end{lem}
@@
 \label{lem:v_ergodicity}
- Assume  {\bf (H1)} and {\bf (H2)}. Then for $0 < \gamma < \overline{\gamma}= \min(\mathfrak{m} / 4L^2,1/L)$, and for any $x \in \rset^d$  it holds
-$$
-d_V(\delta_xQ_{\gamma}^{n},\pi_\gamma) \leq C\rho^{ n}\left(V(x) + \pi_\gamma(V)\right)
-$$
-with $V(x) = 1 + |x|^2$ and constants
-$$
-C = \left(1 + \exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} - \frac{2b}{2+K^2})}\right),
-$$
-
-$$
-b = \gamma(2K_2^2(L+\mathfrak{m}) + d + \mathfrak{m}), \quad \overline{b} = b\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} + K^2 + 1, 
-$$
-
-\begin{eqnarray*}
-\varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right),
-\end{eqnarray*}
-
-\begin{eqnarray}
-\label{eq:rho-ula}
-\rho = \exp\left\{-\frac{\gamma m}{4}\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} + \frac{2b}{K^2+2})}}\right\},
-\end{eqnarray}
-where $K_2 > 0$ is from Lemma~\ref{lem:quadratic_behaviour} and $K>0$ is a positive parameter such that 
-$$
-K^2 > 4\frac{2K_2^2(L+m) + d + m}{m}.
-$$
+ Assume  {\bf (H1)} and {\bf (H2)}. Then for $0 < \gamma < \overline{\gamma}= \min(\mathfrak{m} / 2L^2,1/L)$, and for any $x \in \rset^d$  it holds
+$$
+d_V(\delta_xQ_{\gamma}^{n},\pi_\gamma) \leq C\rho^{\gamma n} V(x)
+$$
+with $V(x) = 1 + |x|^2$ and some $0 < \rho < 1$ and $C > 0$.
+%$$
+%C = \left(1 + \exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} - \frac{2b}{2+K^2})}\right),
+%$$
+%
+%
+%
+%$$
+%b = \gamma(2K_2^2(L+\mathfrak{m}) + d + \mathfrak{m}), \quad \overline{b} = b\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} + K^2 + 1, 
+%$$
+%
+%\begin{eqnarray*}
+%\varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right),
+%\end{eqnarray*}
+%
+%\begin{eqnarray}
+%\label{eq:rho-ula}
+%\rho = \exp\left\{-\frac{\gamma m}{4}\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} + \frac{2b}{K^2+2})}}\right\},
+%\end{eqnarray}
+%where $K_2 > 0$ is from Lemma~\ref{lem:quadratic_behaviour} and $K>0$ is a positive parameter such that 
+%$$
+%K^2 > 4\frac{2K_2^2(L+m) + d + m}{m}.
+%$$
 \end{lem}
@@
 \] 
-Hence,   \cite[Corollary 5]{DurmusBortoli} implies that
-\[
-\|\delta_xQ^{[1/\gamma]}_{\gamma} - \delta_yQ^{[1/\gamma]}_{\gamma}\|_{\text{TV}} \leq 2(1-\varepsilon)
-\]
-with $\varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right)$ for all $x,y$ such that $|x|,|y| \leq K$, which means that the kernel $Q_{\gamma}$ satisfies $([1/\gamma],\varepsilon)$-Doeblin condition on the set $\{x: V(x) < 1+K^2\}$.
-Together with drift condition \eqref{eq:lyapunov} it allows to apply \cite[Theorem~19.4.1]{moulines2018} with appropriate constants.
+Hence,   \cite[Corollary 5]{DurmusBortoli} implies that for all $x,y$ such that $|x|,|y| \leq K$,
+\[
+\|\delta_xQ^{\lceil 1/\gamma \rceil}_{\gamma} - \delta_yQ^{ \lceil 1/\gamma \rceil}_{\gamma}\|_{\text{TV}} \leq 2(1-\varepsilon)
+\]
+where $\lceil \cdot \rceil$ denotes an upper integer part and $\varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right)$.
+Note also that the drift condition \eqref{eq:lyapunov} implies that
+\[
+Q^{[1/\gamma]}_{\gamma}V(x) \leq \lambda V(x) + \frac{\gamma b(1-\lambda^{\gamma(\lceil 1/\gamma\rceil + 1)})}{1-\lambda^{\gamma}} \leq \lambda V(x) + \frac{b}{\lambda \log{1 /\lambda}} 
+\]
+where the last inequality is due to \cite[Lemma 1]{durmus:moulines:2017}.
+Hence, the kernel $Q^{\lceil 1/\gamma \rceil}_{\gamma}$ satisfies $(1,\varepsilon)$-Doeblin condition on the set $\{x: V(x) < 1+K^2\}$ and drift condition \eqref{eq:lyapunov} with appropriate constants.
+Now the statement of the lemma follows from \cite[Theorem~19.4.1]{moulines2018}.
 \end{proof}

### 2020-05-08 aap/Revision/vr-mcmc-revision-final.tex
update

@@
 Now we state and prove the main result of this section on the variance bound for the estimator $\pi_N^n(f)$ in the case of a $V-$geometrically ergodic Markov chain.
+
 \begin{lem}
 \label{lem:var_bound}
-Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and let $f$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$.  Then
+Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and let $f$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$.
+Denote $\tilde{f}(x) = f(x) - \pi(f)$.
+Then
 \begin{equation}
@@
 Let $(X_p)_{p \in \nset_0}$ be a $V-$geometrically ergodic Markov chain with a stationary distribution $\pi$, and $f(x)$ be a function with $\|f\|_{V^{\frac{1}{2}}} < \infty$.
-Let $\tilde{f}(x) = f(x) - \pi(f)$.
+Denote $\tilde{f}(x) = f(x) - \pi(f)$.
 Then

### 2020-05-08 aap/Revision/vr-mcmc-revision-final.tex
update

@@
 
-Under assumptions {\bf (H1)} and {\bf (H2)} the Markov chain $(X^x_{p})_{p \in \nset_0}$ generated by ULA  is $V-$geometrically ergodic in the sense of \eqref{eq:v_ergodicity} with $V(x) = 1+|x|^2$ and $\rho\in (0,1) $ due to Lemma~\ref{lem:v_ergodicity}.
-It follows from Lemma~\ref{lem:var_bound}, that
-\[
-\PVar[\pi_n^N(f)]\lesssim \frac{1}{n(1-\rho^{1/2})} + \left(\frac{\rho^NV(x)+1}{n(1-\rho^{1/2})}\right)^2
-\]
-It follows from \eqref{eq:rho-ula} that $1-\rho^{1/2} \geq C\gamma$ with a constant $C$ not depending on $\gamma,N$ and $n$, thus yielding an estimate
+Under assumptions {\bf (H1)} and {\bf (H2)} the Markov chain $(X^x_{p})_{p \in \nset_0}$ generated by ULA  is $V-$geometrically ergodic in the sense of \eqref{eq:v_ergodicity} with $V(x) = 1+|x|^2$ due to Lemma~\ref{lem:v_ergodicity}.
+It follows from Lemma~\ref{lem:var_bound}, applied with $\rho^{\gamma}$ that
+\[ 
+\PVar[\pi_n^N(f)]\lesssim \frac{1}{n(1-\rho^{\gamma/2})} + \left(\frac{\rho^{\gamma N}V(x)+1}{n(1-\rho^{\gamma/2})}\right)^2
+\]
+Since $1-\rho^{\gamma/2} \geq C\gamma$ with a constant $C$ not depending on $\gamma,N$ and $n$, we end up with an estimate
 \[
@@
  Assume  {\bf (H1)} and {\bf (H2)}. Then for $0 < \gamma < \overline{\gamma}= \min(\mathfrak{m} / 2L^2,1/L)$, and for any $x \in \rset^d$  it holds
-$$
+\begin{equation}
+\label{eq:rho-ula}
 d_V(\delta_xQ_{\gamma}^{n},\pi_\gamma) \leq C\rho^{\gamma n} V(x)
-$$
+\end{equation}
 with $V(x) = 1 + |x|^2$ and some $0 < \rho < 1$ and $C > 0$.

### 2020-05-08 aap/Revision/vr-mcmc-revision-final.tex
update

--- 2020-05-08 aap/Revision/vr-mcmc-revision-final.tex
@@
 Our approach is fully non-asymptotic and does not require the knowledge of the stationary distribution (and even any type of ergodicity) or specific structure of the underlying density.
-By rigorously analyzing the  convergence properties of the proposed algorithm, we show that its complexity is indeed asymptotically smaller than one of the naive algorithm.
-The numerical performance of the new method is illustrated for Markov Chain Monte Carlo (MCMC) methods.
+By rigorously analyzing the  convergence properties of the proposed algorithm, we show that its cost-to-variance relation is indeed  smaller than one of the naive algorithm.
+The numerical performance of the new method is illustrated for Langevin-type Markov Chain Monte Carlo (MCMC) methods.
 \end{abstract}
@@
 $f(X) + \zeta$ is small.
-The complexity of the problem of constructing classes $Z$ of control variates \(\zeta\) satisfying   $\PE[\zeta]=0$ essentially depends on the degree of our knowledge on \(\pi.\)
+The complexity of the problem of constructing classes of control variates \(\zeta\) satisfying   $\PE[\zeta]=0$ essentially depends on the degree of our knowledge on \(\pi.\)
 For example, if \(\pi\) is analytically known and satisfies some regularity conditions, one can apply the well-known technique of  polynomial interpolation to construct control variates enjoying  some optimality properties, see, for example \cite[Section~3.2]{dimov2008monte}.
@@
 At the same time it is possible to express the optimal control variate in terms of the solution of the Poisson equation for the corresponding Markov chain \((X_p)\).
-As it was observed in~\cite{henderson1997variance,henderson2004}, for a time-homogeneous Markov chain \(X_p\) with a stationary distribution \(\pi\), the function $U_{G}(x) := G(x)-\PE[G(X_{1})|X_0 = x]$ has a zero mean with respect to $\pi$ for arbitrary real-valued function \(G: \mathcal{X} \to \rset\), such that \(G  \in L^1(\pi)\).
+As it was observed in~\cite{henderson1997variance,henderson2004}, for a time-homogeneous Markov chain \(X_p\) with a stationary distribution \(\pi\), the function $U_{G}(x) := G(x)-\PE[G(X_{1})|X_0 = x]$ has zero mean with respect to $\pi$ for an arbitrary real-valued function \(G: \mathcal{X} \to \rset\), such that \(G  \in L^1(\pi)\).
 Hence, $U_{G}(x)$ is a valid control functional for a suitable choice of $G$, with the best $G$ given by a solution of the Poisson equation \(\PE [G(X_{1})|X_0 = x]-G(x)=-f(x)+\pi(f)\).
@@
 Moreover, we do not need to assume stationarity or/and sampling under the invariant distribution \(\pi.\) We rigorously analyse the convergence of the method and study its complexity.
-It is shown that  our variance-reduced Langevin algorithm outperforms the standard Langevin algorithms in terms of complexity.
+It is shown that  our variance-reduced Langevin algorithm outperforms the standard Langevin algorithms in terms of its cost-to-variance relation.
 \par
@@
 However, in many practical situations  direct sampling from \(\pi\) is impossible and this precludes the use of plain Monte Carlo methods in this case.
-One popular alternative to Monte Carlo  is Markov Chain Monte Carlo (MCMC), where one is looking for a discrete time  (possibly non-homogeneous) Markov chain   \((X_p)_{p \in \nset_0}\) such that \(\pi\) is its unique invariant measure.
+One popular alternative to Monte Carlo  is Markov Chain Monte Carlo (MCMC) where one is looking for a discrete time  (possibly non-homogeneous) Markov chain   \((X_p)_{p \in \nset_0}\) such that \(\pi\) is its unique invariant measure.
 In this paper we study a class of MCMC algorithms with \((X_p)_{p \in \nset_0}\) satisfying the  the following recurrence relation:
@@
 \begin{example}
-  Let \((X_t)_{t\in [0,T]}\) be the unique strong solution to a  SDE of the form:
+  Let \((X_t)_{t\geq 0}\) be the unique strong solution to a  SDE of the form:
 \begin{eqnarray}
@@
 \label{seq:mart_repr}
-In this section we give a general discrete-time martingale representation for  Markov chains of the type  \eqref{eq:chain_gen} which is  used later to construct an efficient variance reduction algorithm.
+In this section we provide a general discrete-time martingale representation for  Markov chains of the type  \eqref{eq:chain_gen} which is  used later to construct an efficient variance reduction algorithm.
 Let \((\phi_k)_{k\in \mathbb{Z}_+}\) be a complete orthonormal system in \(L^2(\mathbb{R}^m, P_{\xi})\) with \(\phi_0\equiv 1\).
@@
 Let \((X^x_{p})_{p\geq 0}\) be a homogeneous Markov chain of the form \eqref{eq:ula_new} with $\Phi_l = \Phi,$ $l\geq 1$.
-Then for all $p \in \nset,$ $q \leq p$, $j < q \leq p$, $f \in \mathcal{F}_{2,\text{hom}}^p$ and $x \in \rset^d$ it holds in \(L^2(\mathsf{P})\)
+Then for all $p \in \nset,$ $q \leq p$, $j < q \leq p$, $f \in \mathcal{F}_{2,\text{hom}}^p$ and $x \in \rset^d$ it holds in \(L^2(\mathsf{P}),\)
 \begin{eqnarray*}
@@
 \label{eq:definition-bar-a-0}
-\bar a_{r,k}(y)=\mathsf{E}\left[f(X^y_{r})\phi_k\left(\xi_{1}\right)\right] \,, \text{for $r, k\in \mathbb{N}$.}
+\bar a_{r,k}(y)=\mathsf{E}\left[f(X^y_{r})\phi_k\left(\xi_{1}\right)\right] \quad  r, k\in \mathbb{N}
 \end{equation}
@@
 We present a generic regression algorithm and then in Section~\ref{sec:num} give further implementation details. Our algorithm starts with estimating the functions
-$Q_r$ for $r=0,\ldots,n-1,$, defined in \eqref{eq:definition-bar-a}.
+$Q_r$ for $r=0,\ldots,n-1,$ defined in \eqref{eq:definition-bar-a}.
 We first generate \(\NtrainPath\)  paths  of the chain \(X\)
@@
 Given $(\widehat A_{l,k})$ the cost of computing  $\widehat \pi_{K,n}^N$ is proportional to $\mathtt{D}_\Psi K n.$ At the same time the variance of the standard estimate $ \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved, see Lemma~\ref{lem:variance} and Remark~\ref{rem:var-low}.
-Thus, the ratio of the corresponding cost-variances is of order
+Thus, the ratio of the corresponding cost-to-variance characteristics  is of order
 \begin{equation}
@@
 
-Under assumptions {\bf (H1)} and {\bf (H2)} the Markov chain $(X^x_{p})_{p \in \nset_0}$ generated by ULA  is $V-$geometrically ergodic in the sense of \eqref{eq:v_ergodicity} with $V(x) = 1+|x|^2$ and $\rho\in (0,1) $ due to Lemma~\ref{lem:v_ergodicity}.
-It follows from Lemma~\ref{lem:var_bound}, that
+Under assumptions {\bf (H1)} and {\bf (H2)} the Markov chain $(X^x_{p})_{p \in \nset_0}$ is $V-$geometrically ergodic in the sense of \eqref{eq:v_ergodicity} with $V(x) = 1+|x|^2$ and $\rho\in (0,1) $ specified in Lemma~\ref{lem:v_ergodicity}.
+Due to Lemma~\ref{lem:var_bound},
 \[
@@
 \label{eq:lyapunov}
-Q_{\gamma}V(x) \leq \lambda^{\gamma} V(x) + \gamma b
-\end{equation}
-with   $\lambda = \exp{\left(-\frac{\mathfrak{m}}{2}\right)}$, $b = 2K_2^2(L+\mathfrak{m}) + d + \mathfrak{m}$ with $K_2$ from Lemma~\ref{lem:quadratic_behaviour}, and $\mathfrak{m}$ from {\bf (H2)}.
+Q_{\gamma}V(x) \leq \rho^{\gamma} V(x) + \gamma C
+\end{equation}
+with   $\rho = \exp{\left(-\frac{\mathfrak{m}}{2}\right)}$, $C = 2K_2^2(L+\mathfrak{m}) + d + \mathfrak{m}$ with $K_2$ from Lemma~\ref{lem:quadratic_behaviour}, and $\mathfrak{m}$ from {\bf (H2)}.
 \end{lem}
@@
 \label{lem:v_ergodicity}
- Assume  {\bf (H1)} and {\bf (H2)}. Then for $0 < \gamma < \overline{\gamma}= \min(\mathfrak{m} / 2L^2,1/L)$, and for any $x \in \rset^d$  it holds
-$$
-d_V(\delta_xQ_{\gamma}^{n},\pi_\gamma) \leq C\rho^{\gamma n} V(x)
-$$
-with $V(x) = 1 + |x|^2$ and some $0 < \rho < 1$ and $C > 0$.
-%$$
-%C = \left(1 + \exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} - \frac{2b}{2+K^2})}\right),
-%$$
-%
-%
-%
-%$$
-%b = \gamma(2K_2^2(L+\mathfrak{m}) + d + \mathfrak{m}), \quad \overline{b} = b\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} + K^2 + 1, 
-%$$
-%
-%\begin{eqnarray*}
-%\varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right),
-%\end{eqnarray*}
-%
-%\begin{eqnarray}
-%\label{eq:rho-ula}
-%\rho = \exp\left\{-\frac{\gamma m}{4}\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} + \frac{2b}{K^2+2})}}\right\},
-%\end{eqnarray}
-%where $K_2 > 0$ is from Lemma~\ref{lem:quadratic_behaviour} and $K>0$ is a positive parameter such that 
-%$$
-%K^2 > 4\frac{2K_2^2(L+m) + d + m}{m}.
-%$$
+ Assume  {\bf (H1)} and {\bf (H2)}. Then for $0 < \gamma < \overline{\gamma}= \min(\mathfrak{m} / 4L^2,1/L)$, and for any $x \in \rset^d$  it holds
+$$
+d_V(\delta_xQ_{\gamma}^{n},\pi_\gamma) \leq C\rho^{ n}\left(V(x) + \pi_\gamma(V)\right)
+$$
+with $V(x) = 1 + |x|^2$ and constants
+$$
+C = \left(1 + \exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} - \frac{2b}{2+K^2})}\right),
+$$
+
+$$
+b = \gamma(2K_2^2(L+\mathfrak{m}) + d + \mathfrak{m}), \quad \overline{b} = b\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} + K^2 + 1, 
+$$
+
+\begin{eqnarray*}
+\varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right),
+\end{eqnarray*}
+
+\begin{eqnarray}
+\label{eq:rho-ula}
+\rho = \exp\left\{-\frac{\gamma m}{4}\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp{\left(-\frac{\mathfrak{m}\gamma}{2}\right)} + \frac{2b}{K^2+2})}}\right\},
+\end{eqnarray}
+where $K_2 > 0$ is from Lemma~\ref{lem:quadratic_behaviour} and $K>0$ is a positive parameter such that 
+$$
+K^2 > 4\frac{2K_2^2(L+m) + d + m}{m}.
+$$
 \end{lem}
@@
 \] 
-Hence,   \cite[Corollary 5]{DurmusBortoli} implies that for all $x,y$ such that $|x|,|y| \leq K$,
-\[
-\|\delta_xQ^{\lceil 1/\gamma \rceil}_{\gamma} - \delta_yQ^{ \lceil 1/\gamma \rceil}_{\gamma}\|_{\text{TV}} \leq 2(1-\varepsilon)
-\]
-where $\lceil \cdot \rceil$ denotes an upper integer part and $\varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right)$.
-Note also that the drift condition \eqref{eq:lyapunov} implies that
-\[
-Q^{[1/\gamma]}_{\gamma}V(x) \leq \lambda V(x) + \frac{\gamma b(1-\lambda^{\gamma(\lceil 1/\gamma\rceil + 1)})}{1-\lambda^{\gamma}} \leq \lambda V(x) + \frac{b}{\lambda \log{1 /\lambda}} 
-\]
-where the last inequality is due to \cite[Lemma 1]{durmus:moulines:2017}.
-Hence, the kernel $Q^{\lceil 1/\gamma \rceil}_{\gamma}$ satisfies $(1,\varepsilon)$-Doeblin condition on the set $\{x: V(x) < 1+K^2\}$ and drift condition \eqref{eq:lyapunov} with appropriate constants.
-Now the statement of the lemma follows from \cite[Theorem~19.4.1]{moulines2018}.
+Hence,   \cite[Corollary 5]{DurmusBortoli} implies that
+\[
+\|\delta_xQ^{[1/\gamma]}_{\gamma} - \delta_yQ^{[1/\gamma]}_{\gamma}\|_{\text{TV}} \leq 2(1-\varepsilon)
+\]
+with $\varepsilon = 2\Phi\left(-K\sqrt{\left(1+\frac{1}{L}\right)\left(1 + 2L\right)}\right)$ for all $x,y$ such that $|x|,|y| \leq K$, which means that the kernel $Q_{\gamma}$ satisfies $([1/\gamma],\varepsilon)$-Doeblin condition on the set $\{x: V(x) < 1+K^2\}$.
+Together with drift condition \eqref{eq:lyapunov} it allows to apply \cite[Theorem~19.4.1]{moulines2018} with appropriate constants.
 \end{proof}

### 2020-05-09 aap/Revision/vr-mcmc-revision-final.tex
update

@@
 If the class of functions $\Psi$ is a linear one of dimension $\mathtt{D}_\Psi,$ then the cost of computing the coefficients $\widehat A_{l,k}$ for all $l=1,\ldots,n,$ and $k=1,\ldots,K,$ is of order  $ \mathtt{D}_\Psi K T^2 n$.
-Given $(\widehat A_{l,k})$ the cost of computing  $\widehat \pi_{K,n}^N$ is proportional to $\mathtt{D}_\Psi K n.$ At the same time the variance of the standard estimate $ \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved, see Lemma~\ref{lem:variance} and Remark~\ref{rem:var-low}.
+Given $(\widehat A_{l,k})$ the cost of evaluating  $\widehat \pi_{K,n}^N$ is proportional to $\mathtt{D}_\Psi K n.$ At the same time the variance of the standard estimate $ \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved, see Lemma~\ref{lem:variance} and Remark~\ref{rem:var-low}.
 Thus, the ratio of the corresponding cost-to-variance characteristics  is of order
@@
 \end{equation}
-Thus, for any fixed $K\geq 1,$  our variance reduction method is advantageous if $\gamma\ll\min\{1/(\mathtt{D}_\Psi T^2),1/\sqrt{n}\}.$  Note that in order to achieve convergence of the corresponding MSE to zero, we need to let $\gamma\to 0$ as the invariant measure of the ULA with a constant time step $\gamma$ is not equal to $\pi.$
+Thus, for any fixed $K\geq 1,$  our variance reduction method is advantageous if $\gamma\ll\min\{1/(\mathtt{D}_\Psi T^2),1/\sqrt{n}\}.$  Note that in order to achieve convergence of the corresponding MSE to zero, we need anyway that $\gamma\to 0$ as the invariant measure of the ULA with a constant time step $\gamma$ is not equal to $\pi.$
 \section{Analysis of variance reduced ULA}\label{sec:ula_analysis}
@@
 \end{eqnarray*}
-\item[(H2)][\textbf{Convexity}] The potential $U$ is of the form $U(x)=U_0(x)+\Delta(x),$ where $\Delta$ has  compact support and $U_0\in C^2(\mathbb{R}^d)$ with
+\item[(H2)][\textbf{Convexity}] The potential $U$ has the form $U(x)=U_0(x)+\Delta(x)$ where $\Delta$ has  a compact support and $U_0\in C^2(\mathbb{R}^d)$ with
 \begin{equation*}
@@
 Suppose that  {\bf (H2)} holds with  $\Delta \in C^2(\mathbb{R}^d)$ satisfying $|\mathsf{D}^{2}\Delta(x)|\leq D$ for all $x\in \mathbb{R}^d$ and some constant $D>0.$ Then for  $\gamma  <1/\mathfrak{M}$ we have
- \begin{eqnarray*}
+ \begin{eqnarray}
+ \label{eq:cor-bound}
 \sum_{p=j+1}^{\infty}\gamma\left\Vert\prod_{k=j+1}^{p} (I-\gamma\mathsf{D}^{2}U(X^x_{k-1}))\right\Vert _{r}&\leq &  \frac{1}{\mathfrak{m}}+\frac{ D}{\mathfrak{m}^2(1-\gamma \mathfrak{M})}.
-\end{eqnarray*}
+\end{eqnarray}
 Hence under assumptions of Theorem~\ref{th:mr}, there exists  a constant $\widetilde{C}_K < \infty$  such that for all $n \in \nset$ and $x \in \rset^d$,
@@
 \\
-\gamma\sum_{l=j+1}^{p}\prod_{k=j+1}^{l-1}\left(I-\gamma\mathsf{D}^{2}U_{0}(X_{k-1})\right)\mathsf{D}^{2}\Delta(X_{l-1})\prod_{k=l+1}^{p}\left(I-\gamma\mathsf{D}^{2}U_0(X_{k-1})\right),
+\gamma\sum_{l=j+1}^{p}\prod_{k=j+1}^{l-1}\left(I-\gamma\mathsf{D}^{2}U_{0}(X_{k-1})\right)\mathsf{D}^{2}\Delta(X_{l-1})\prod_{k=l+1}^{p}\left(I-\gamma\mathsf{D}^{2}U_0(X_{k-1})\right)
 \end{multline*}
@@
 \\
-&\leq &  \left(1+\frac{\gamma (p-j) D}{1-\gamma\mathfrak{M}}\right) (1-\gamma \mathfrak{m})^{p-j}.
-\end{eqnarray*}
-
+&\leq &  \left(1+\frac{\gamma (p-j) D}{1-\gamma\mathfrak{M}}\right) (1-\gamma \mathfrak{m})^{p-j}
+\end{eqnarray*}
+and  \eqref{eq:cor-bound} follows.
 \end{proof}
@@
 \begin{eqnarray*}
-A^2_{q,\mathbf{k}}(x)\leq \gamma^{|\mathbf{k}|} \PVar \left( \sum_{p=1}^{q}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X_{p}^x\right)
+A^2_{q,\mathbf{k}}(x)\leq \gamma^{|\mathbf{k}|} \PVar \left( \sum_{p=0}^{q}\partial_{Z_{1}}^{\mathbf{k}}f\left(X_{p}^x\right)
 \right),\quad q=0,\ldots,n-1.
@@
 \left(\frac{\gamma}{2}\right)^{|I|K}
-\PVar \left( \sum_{p=1}^{q}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X_{p}^x\right)
-\right),
+\PVar \left( \sum_{p=0}^{q}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X_{p}^x\right)
+\right)
 \end{eqnarray*}
@@
 \label{eq:varboundV}
-\PVar \left(\sum_{p=1}^{q}\partial_{Z_{1}}^{\mathbf{K}_I}f\left(X^x_{p}\right)
+\PVar \left(\sum_{p=1}^{q}\partial_{Z_{1}}^{\mathbf{k}}f\left(X^x_{p}\right)
 \right)\leq C_{K}\gamma^{-1}\Xi(x),\quad q=0,\ldots,n-1,
 \end{equation}
-for some positive constants $\varkappa, C_K$ not depending on $n$ and  $\gamma.$
+for any $\| \mathbf{k} \| \leq K$ and some positive constants $\varkappa, C_K$ not depending on $n$ and  $\gamma.$
 \section{Numerical analysis}

### 2020-05-09 aap/Revision/Response_1.tex
update

@@
 In this work we propose a novel variance reduction approach which is able to reduce the statistical error of MCMC algorithms.
-The main advantage of our method is provable increase of efficiency for Langevin-based MCMC algorithms as measured by  cost-times-variance criterion.
+The main advantage of our method is provable increase of efficiency for Langevin-based MCMC algorithms as measured by  cost-to-variance criterion.
 Our approach is based on a new discrete time martingale representation for Markov chains which is of its own interest.
 Therefore we first present this basic result and then discuss in Section 4 variance reduction method based on the derived martingale representation.
-In  Section~4 we  explain in accessible way why our approach leads to reduced cost-times-variance criterion.
+In  Section~4 we  explain in accessible way why our approach leads to reduced cost-to-variance criterion.
 These results are based on deep analysis of the variance reduced estimate postponed to Section 5.

### 2020-05-09 aap/Revision/vr-mcmc-revision-final.tex
update

@@
 We set $K=1$ or $K=2$ for $f(x)=\sum_{i=1}^d x_i$ or $f(x)=\sum_{i=1}^d x^2_i$ respectively.
-Then we estimate the cost-variance ratio
+Then we estimate the cost-to-variance ratio
 \begin{eqnarray}
@@
 by its empirical counterpart, computed over $24$ independent trajectories with $n = 2\times 10^3$ and $N = 2 \times 10^3$.
-Since for fixed $K$ the cost of computing $\pi_{n}^N(f)$ is proportional to computing function $f$, we set for $K = 1$
+Since for a fixed $K$ the cost of computing $\pi_{n}^N(f)$ is proportional to the cost of computing function $f$, we set for $K = 1$
  \begin{equation*}
@@
  \end{equation*}
- since there are only $2$ non-zero functions among $a_{r,\mathbf{k}}(x)$ for fixed $r$.
-Note that in this case each $a_{r,\mathbf{k}}(x)$  is a linear function, which can be computed at the same cost as $f$.
+ since there are only $2$ non-zero functions among $a_{r,\mathbf{k}}$ for a fixed $r$.
+Note that in this case each $a_{r,\mathbf{k}}$  is a linear function, which can be computed at the same cost as $f$.
 Similarly, for $K = 2$ we set
   \begin{equation*}
- \mathrm{cost}(\widehat \pi_{K,n}^N) = \mathrm{cost}(\pi_{n}^N) \times n_{\mathrm{trunc}} \times 8
+ \mathrm{cost}(\widehat \pi_{K,n}^N) = \mathrm{cost}(\pi_{n}^N) \times n_{\mathrm{trunc}} \times 8,
  \end{equation*}
-since there are at most $8$ non-zero functions among $a_{r,\mathbf{k}}(x)$ for fixed $r$.
+since there are at most $8$ non-zero functions among $a_{r,\mathbf{k}}(x)$ for a fixed $r$.
 Variance reduction costs for Gaussian potential and different $n_{\mathrm{trunc}}$  are summarized in Figure~\ref{fig:gaus_cost}, and for the Gaussian mixture - in Figure~\ref{fig:gmm_cost}.
@@
 \label{fig:subfig2}}
-\caption{Cost-variance ratios \eqref{eq:cvR} as  functions of the truncation level $n_{\mathrm{trunc}}$ for  two-dimensional standard Gaussian distribution and different test functions.
+\caption{Cost-to-variance ratios \eqref{eq:cvR} as  functions of the truncation level $n_{\mathrm{trunc}}$ for  a two-dimensional standard Gaussian distribution and different test functions.
 \label{fig:gaus_cost}}
@@
 \label{fig:subfig2}}
-\caption{Cost-variance ratios  \eqref{eq:cvR} as  functions of the truncation level $n_{\mathrm{trunc}}$ for a mixture of two-dimensional Gaussian distributions and different test functions.
+\caption{Cost-to-variance ratios  \eqref{eq:cvR} as  functions of the truncation level $n_{\mathrm{trunc}}$ for the mixture \eqref{eq:gmm} of two-dimensional Gaussian distributions and two  test functions.
 \label{fig:gmm_cost}}

### 2020-05-09 Code/ODE_samplers.py
update

@@ -122,7 +122,8 @@ def MALA_ODE(r_seed,Potential,step,params,N, n, d, typ, t = 1.0):
         #Lotki-Volterra system
         mu = params["mu"]
         sigma = params["sigma"]
-        step = step*np.array([10.0,0.1,10.0,0.1])
+        #step = step*np.array([10.0,0.1,10.0,0.1])
+        step = step*np.array([1.0,0.1,1.0,0.1])
         mu_init = Potential.theta_mle
         sigma_init = mu_init/20.0
         x = mu_init + sigma_init*np.random.randn(d)

### 2020-05-09 aap/Revision/Response_1.tex

@@
 \par
-Let us stress that  current literature on MCMC  mainly focuses on  convergence of MCMC algorithms in Total Variation and Wasserstein distances thus  ignoring a sample error of these algorithms.
+Let us first stress that  current literature on MCMC  mainly focuses on  convergence of MCMC algorithms in Total Variation and Wasserstein distances thus  ignoring a sample error of these algorithms.
 Only few references studied MSE error of the Langevin algorithm (see e.g.
@@
 In this work we propose a novel variance reduction approach which is able to reduce the statistical error of MCMC algorithms.
-The main advantage of our method is provable increase of efficiency for Langevin-based MCMC algorithms as measured by  cost-to-variance criterion.
+The main advantage of our method is provable increase of efficiency for Langevin-based MCMC algorithms as measured by  the cost-to-variance criterion.
 Our approach is based on a new discrete time martingale representation for Markov chains which is of its own interest.
 Therefore we first present this basic result and then discuss in Section 4 variance reduction method based on the derived martingale representation.
-In  Section~4 we  explain in accessible way why our approach leads to reduced cost-to-variance criterion.
+In  Section~4 we  explain in accessible way why our approach leads to a smaller cost-to-variance characteristic.
 These results are based on deep analysis of the variance reduced estimate postponed to Section 5.
-In the revised version we significantly weaken  assumptions and prove our main result (Theorem~3 and Corollary~2) for ULA under smoothness  and convexity of the underlying potential outside a ball (assumption  (H2)).
-Note that all assumptions needed are now collected in Theorem~3.
+In the revised version we significantly weaken  assumptions  and prove our main result (Theorem~10 and Corollary~11) for ULA under smoothness  and convexity of the underlying potential outside a ball (assumption  (H2)).
+Note that all assumptions needed are now collected in Theorem~10.
 For convenience of the reader we also sketch the main steps of the proof right after
-Theorem~3.
-We decided to present our main theoretical results as bounds for the functions $A_{q,k}$ since these functions play a crucial role in the variance bound (13).
+Theorem~10.
+We decided to present our main theoretical result in the form of bounds for the functions $A_{q,k}$ since these functions play a crucial role in the variance bound (13).
 Given these estimates it is easy to see how to get bounds for the variance.

### 2020-05-09 aap/Revision/Response_2.tex
update

@@
 \par
-In the revised version we significantly weaken  the assumptions and prove our main result (Theorem~3 and Corollary~2) for ULA under smoothness  and convexity of the underlying potential outside a ball (assumption  (H2)).
-Thus we avoid the strong convexity assumption.
-Note that all assumptions needed are now collected in Theorem~3.
-These  assumptions can now be validated for a large class of examples including example of Gaussian mixtures considered in Section~6.1.
+In the revised version we significantly weaken  these assumptions and prove our main result (Theorem~10 and Corollary~11) for ULA under smoothness  and convexity of the underlying potential outside a ball (assumption  (H2)) hence avoiding the strong convexity assumption.
+Note that all assumptions needed are now collected in Theorem~10 and Corollary~11.
+These  assumptions can now be validated for a large class of examples including the case of Gaussian mixtures considered in Section~6.2.
 \item \textit{...
@@
 \par
-Now we replaced the strong convexity assumption by a much weaker one (H2) (convexity outside a ball).
-Hence our method may gain additional efficiency as compared to other variance reduction approaches due to the fact that \(\nabla \log(\pi)\) doesn't need to be computed to construct control variates.
+As mentioned above we have replaced the strong convexity assumption by a much weaker one (H2) (convexity outside a ball).
+Hence our method may gain additional efficiency as compared to other variance reduction approaches due to the fact that \(\nabla \log(\pi)\) doesn't need to be computed for the construction of control variates.
 \end{itemize}

### 2020-05-09 aap/Revision/vr-mcmc-revision-final.tex
update

@@
 \begin{eqnarray*}
-X_{k+1} = (1-2\gamma)X_k + \sqrt{\gamma}\xi_{k+1},\quad \xi_{k+1} \sim \mathcal{N}(0,1).
+X_{k+1} = \left(1-\frac{\gamma}{2}\right)X_k + \sqrt{\gamma}\xi_{k+1},\quad \xi_{k+1} \sim \mathcal{N}(0,1).
 \end{eqnarray*}
@@
 \PVar_{0} [\pi_n^0(f)] &=& \PVar_{0}\left[\frac{1}{n}\sum\limits_{i=1}^{n}X_i\right]
-= \frac{1}{n^2}\PVar\left[\sum\limits_{i=1}^{n}\sum\limits_{j=1}^{i}\sqrt{\gamma}(1-2\gamma)^{i-j}\xi_{j}\right]
-\geq  \frac{1}{n\gamma} - \frac{2(1-2\gamma)}{n^2\gamma^2}.
+= \frac{1}{n^2}\PVar\left[\sum\limits_{i=1}^{n}\sum\limits_{j=1}^{i}\sqrt{\gamma}\left(1-\frac{\gamma}{2}\right)^{i-j}\xi_{j}\right]
+\geq  \frac{4}{n\gamma} - \frac{16(1-\frac{\gamma}{2})}{n^2\gamma^2}.
 \end{eqnarray*}
@@
 \label{eq:main-ass-prod}
-\max_{j\geq N}\sum_{p=j+1}^{\infty}\gamma\,\left\Vert\prod_{k=j+1}^{p} (I-\gamma\mathsf{D}^{2}U(X^x_{k-1}))\right \Vert_{r}\leq \Xi,\quad r=2,4,
+\max_{j\geq N}\sum_{p=j+1}^{\infty}\gamma\,\left\Vert\prod_{k=j+1}^{p} (I-\gamma\mathsf{D}^{2}U(X^x_{k-1}))\right \Vert_{r}\leq \Xi(x),\quad r=2,4,
 \end{eqnarray}
-for some $\Xi=\Xi(x)>0$ not depending on $\gamma,$ where for any random matrix $A$ we define $\|A\|_r=(\mathsf{E}|A|^r)^{1/r}.$ 
+for some $\Xi(x)>0$ not depending on $\gamma,$ where for any random matrix $A$ we define $\|A\|_r=(\mathsf{E}|A|^r)^{1/r}.$ 
 Then, there exists a constant $C_K < \infty$  such that
@@
 \end{figure}
-
 In the case of  Gaussian prior $\pi_0(\theta) \sim \mathcal{N}(0,\sigma^2 I_d)$, the unnormalized posterior density takes the form:
@@
 We put $n_{trunc} = 50$ and $K = 1$.
-The target function is taken to be $f(\theta) = \sum_{i=1}^{i=d}\theta_i$.
+The target function is taken to be $f(\theta) = \sum_{i=1}^{d}\theta_i$.
 In order to test our variance reduction algorithm, we generate \(100\) independent test trajectories of length $n = 10^3$.
@@
 \[
-Q^{[1/\gamma]}_{\gamma}V(x) \leq \lambda V(x) + \frac{\gamma b(1-\lambda^{\gamma(\lceil 1/\gamma\rceil + 1)})}{1-\lambda^{\gamma}} \leq \lambda V(x) + \frac{b}{\lambda \log{1 /\lambda}} 
+Q^{\lceil 1/\gamma \rceil}_{\gamma}V(x) \leq \lambda V(x) + \frac{\gamma b(1-\lambda^{\gamma(\lceil 1/\gamma\rceil + 1)})}{1-\lambda^{\gamma}} \leq \lambda V(x) + \frac{b}{\lambda \log{1 /\lambda}} 
 \]

### 2020-05-10 aap/Revision/vr-mcmc-revision-final.tex
update

@@
 \begin{frontmatter}
-\title{Variance reduction for Markov chains via martingale representations}
-\runtitle{Variance reduction for Markov chains}
+\title{Variance reduction for MCMC algorithms via martingale representations}
+\runtitle{Variance reduction for MCMC algorithms}
 %\thankstext{T1}{This research was }
@@
 Our approach is fully non-asymptotic and does not require the knowledge of the stationary distribution (and even any type of ergodicity) or specific structure of the underlying density.
-By rigorously analyzing the  convergence properties of the proposed algorithm, we show that its cost-to-variance relation is indeed  smaller than one of the naive algorithm.
-The numerical performance of the new method is illustrated for Langevin-type Markov Chain Monte Carlo (MCMC) methods.
+By rigorously analyzing the  convergence properties of the proposed algorithm, we show that its cost-to-variance product is indeed  smaller than one of the naive algorithm.
+The numerical performance of the new method is illustrated for the Langevin-type Markov Chain Monte Carlo (MCMC) methods.
 \end{abstract}
@@
 \label{eq:chain_gen}
-X_{p}=\Phi_{p}(X_{p-1},\xi_{p}),\quad p=1,2,\ldots ,\quad X_{0}=x,
+X_{p}=\Phi_{p}(X_{p-1},\xi_{p}),\quad p=1,2,\ldots ,\quad X_{0}=x
 \end{equation}
@@
 where \((U_{p})_{p\geq 1}\) is an i.i.d.
-sequence of uniformly distributed on \([0,1]\) random variables independent of \((Z_p)_{p\geq 1}.\) Thus we recover \eqref{eq:chain_gen} with  \(\xi_p=(U_p,Z_p)\in \mathbb{R}^{d+1}\) and
+sequence of uniformly distributed on \([0,1]\) random variables independent of \((Z_p)_{p\geq 1}.\) Thus, we recover \eqref{eq:chain_gen} with  \(\xi_p=(U_p,Z_p)\in \mathbb{R}^{d+1}\) and
 \begin{eqnarray*}
@@
 Given $(\widehat A_{l,k})$ the cost of evaluating  $\widehat \pi_{K,n}^N$ is proportional to $\mathtt{D}_\Psi K n.$ At the same time the variance of the standard estimate $ \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved, see Lemma~\ref{lem:variance} and Remark~\ref{rem:var-low}.
-Thus, the ratio of the corresponding cost-to-variance characteristics  is of order
+Thus, the ratio of the corresponding cost-to-variance characteristics (product of cost and variance)   can be bounded as
 \begin{equation}
@@
 \label{sec:num}
-In this section we illustrate the performance of the proposed variance reduction method for ULA.
-First we construct a polynomial approximation  for each \(Q_r(x)\)  in the form:
+In this section we illustrate the performance of the proposed variance reduced ULA algorithm for  various potential functions $U$. In our we implementation we follow the next steps.
+First we construct a polynomial approximation  for each \(Q_r\)  in the form:
 \begin{equation*}
@@
 \end{equation*}
-The coefficients \(\widehat \beta_{\mathbf{s}}\in \mathbb{R}\) are obtained using a modified least-squares criteria based on $T$ independent training trajectories \(\bigl\{\bigl(X_1^{(s)},\ldots,X_{N+n}^{(s)}\bigr)\bigr\}_{s=1}^{T}\).
+The coefficients \(\widehat \beta_{\mathbf{s}}\in \mathbb{R}\) are obtained via a modified least-squares criteria based on $T$ independent training trajectories \(\bigl\{\bigl(X_1^{(s)},\ldots,X_{N+n}^{(s)}\bigr)\bigr\}_{s=1}^{T}\).
 More precisely  we define $\widehat{Q}_0(x) = f(x)$ and for \(1 \leq r \leq n-1,\) we set
@@
 \]
-we obtain closed-form expression for the estimates $\widehat a_{r,k}(x)$ of functions $\bar a_{r,k}(x)$ in \eqref{eq:definition-bar-a}. Namely, for all $x=(x^1,\dots,x^d) \in \mathbb{R}^d$,
+we obtain a closed-form expression for the estimates $\widehat a_{r,\mathbf{k}}$ of functions $\bar a_{r,\mathbf{k}}$ in \eqref{eq:definition-bar-a}.
+Namely, for all $x=(x^1,\dots,x^d) \in \mathbb{R}^d$ and $\mu=\nabla U,$  we have
 \begin{align}
 \nonumber
-\widehat a_{r,\mathbf{k}} (x) &= \int \mathbf{H}_\mathbf{k}(z) \widehat{Q}_{r}(x - \gamma \mu(x) + \sqrt{\gamma}z) P_\xi(\rmd z)
+\widehat a_{r,\mathbf{k}} (x) &= \int \mathbf{H}_\mathbf{k}(z) \widehat{Q}_{r}(x - \gamma \mu(x) + \sqrt{\gamma}z) \boldsymbol{\varphi}(\rmd z)
 \\
@@
 We estimate $Q_{r}$ only for $r < n_{\mathrm{trunc}}$ where the truncation level \(n_{\mathrm{trunc}}\) may depend on \(d\) and \(\gamma.\) It allows us to use a smaller amount of training trajectories to approximate $Q_{r}(x)$.
-Finally  we construct a truncated version of the estimator \eqref{eq:29032018a3}:
-\[
-\pi_{K,n,n_{\mathrm{trunc}}}^N(f) = \pi_n^N(f) -  \widehat M_{K,n, n_{\mathrm{trunc}}}^{N}(f),
+Finally,  we construct a truncated version of the estimator \eqref{eq:29032018a3}:
+\[
+\pi_{K,n,n_{\mathrm{trunc}}}^N(f) = \pi_n^N(f) -  \widehat M_{K,n, n_{\mathrm{trunc}}}^{N}(f)
 \]
@@
 \subsection{Comparison with vanilla ULA}
-In this subsection we aim at comparing cost of variance reduction, achieved by the proposed algorithm, to the cost of vanilla ULA algorithm.
+In this subsection we aim at comparing the degree of variance reduction relative to costs, achieved by the proposed algorithm as compared to  the vanilla ULA algorithm.
 We consider samples, generated by ULA with $\pi$ being either the standard normal distribution in dimension $d$ or the mixture of two $d-$dimensional standard Gaussians of the form
@@
  where $d = 2$ and $\mu = (0.5,0.5)$.
-For both examples, we aimed at estimating $\pi(f)$ with $f(x)=\sum_{i=1}^d x_i$ and $f(x) = \sum_{i=1}^d x^2_i$.
-We used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 5 \times 10^4$ independent training trajectories, each one with the burn-in period $n_{\text{burn}} = 100$.
+For both examples, our goal is to estimate $\pi(f)$ with $f(x)=\sum_{i=1}^d x_i$ and $f(x) = \sum_{i=1}^d x^2_i$.
+We used a constant step size $\gamma=0.1$ and sampled $\NtrainPath = 5 \times 10^4$ independent training trajectories, each one with the burn-in period $n_{\text{burn}} = 100$.
 Then we solve the least squares  problems \eqref{eq:Qregr} using the first order polynomial approximations for $f(x)=\sum_{i=1}^d x_i$ and second order polynomial approximations for $f(x) = \sum_{i=1}^d x^2_i$ as described in the previous section.
-We set $K=1$ or $K=2$ for $f(x)=\sum_{i=1}^d x_i$ or $f(x)=\sum_{i=1}^d x^2_i$ respectively.
-Then we estimate the cost-to-variance ratio
+We set $K=1$ or $K=2$ for $f(x)=\sum_{i=1}^d x_i$ or $f(x)=\sum_{i=1}^d x^2_i,$ respectively.
+Then we estimate the cost-to-variance ratio (degree of variance reduction relative to costs)
 \begin{eqnarray}
@@
  \begin{equation*}
- \mathrm{cost}(\widehat \pi_{K,n}^N) = \mathrm{cost}(\pi_{n}^N) \times n_{\mathrm{trunc}} \times 2
+ \mathrm{cost}(\widehat \pi_{K,n}^N) = \mathrm{cost}(\pi_{n}^N) \times n_{\mathrm{trunc}} \times 2,
  \end{equation*}
@@
  \end{equation*}
-since there are at most $8$ non-zero functions among $a_{r,\mathbf{k}}(x)$ for a fixed $r$.
+since there are at most $8$ non-zero functions among $a_{r,\mathbf{k}}$ for a fixed $r$.
 Variance reduction costs for Gaussian potential and different $n_{\mathrm{trunc}}$  are summarized in Figure~\ref{fig:gaus_cost}, and for the Gaussian mixture - in Figure~\ref{fig:gmm_cost}.
-Note that for both examples the proposed algorithm allows us to obtain a sufficient gain in variance reduction cost.
+Note that for both examples the proposed algorithm allows us to obtain a significant gain in efficiency.
 
@@
 In our experiments we considered $d = 2$, $\mu = \left(0.5,0.5\right)$ and randomly generated positive-definite matrix $\Sigma$ (heterogeneous structure).
-In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i\) or \(f(x) = \sum_{i=1}^d x^2_i\) we used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 5 \times 10^4$ independent training trajectories, each one of size $n = 50$ with the burn-in period $n_{\text{burn}} = 100$.
+In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i\) or \(f(x) = \sum_{i=1}^d x^2_i,\) we used a constant step size $\gamma=0.1$ and sampled $\NtrainPath = 5 \times 10^4$ independent training trajectories, each one of size $n = 50$ with the burn-in period $n_{\text{burn}} = 100$.
 Then we solve the least squares  problems \eqref{eq:Qregr} using the first order polynomial approximations for $f(x)=\sum_{i=1}^d x_i$ and second order approximations for $f(x) = \sum_{i=1}^d x^2_i$ as described in the previous section.
-Hence we set $K = 1$ for $f(x)=\sum_{i=1}^d x_i$ or $K = 2$ for $f(x) = \sum_{i=1}^d x^2_i$.
+Also we set $K = 1$ for $f(x)=\sum_{i=1}^d x_i$ or $K = 2$ for $f(x) = \sum_{i=1}^d x^2_i$.
 We set the truncation level \(n_{\mathrm{trunc}} = 50\).
-To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 2 \times 10^3$ and plot boxplots of the corresponding ergodic averages in Figure ~\ref{fig:GMM}.
+To test our variance reduction algorithm, we generated \(100\) independent trajectories of length $n = 2 \times 10^3$ and plot boxplots of the corresponding ergodic averages in Figure ~\ref{fig:GMM}.
 Our approach (MDCV, Martingale decomposition control variates) is compared to other variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}.
-In the baselines we use first order polynomials in case of $K=1$ and second order polynomials for $K = 2$.
+In the baselines we used the first order polynomials in case of $K=1$ and second order polynomials for $K = 2$.
 \par
@@
 \label{fig:subfig2}}
-\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Gaussian mixture model.
-The compared estimators are the ordinary empirical average  (Vanilla), our estimator of control variates (MDCV), zero variance estimator (ZV) and control variates obtained with empirical spectral variance minimisation (ESVM)
+\caption{Boxplots of ergodic averages  obtained by the variance reduced ULA algorithm for the Gaussian mixture model.
+The compared estimators are the ordinary empirical average  (Vanilla), our method of martingale control variates (MDCV), zero variance estimator (ZV) and the control variates approach based on the empirical spectral variance minimisation (ESVM)
 \label{fig:GMM}}
@@
 We set $p = 100, b = 0.1$ and consider $d=8$.
-We solve the least squares  problems \eqref{eq:Qregr} using the second-order polynomial approximations for the coefficients \(\hat{a}_{p,\mathbf{k}}\) as described in the previous section, hence we set $K = 2$.
-We use $T= 5 \times 10^4$ independent training trajectories, each of size $n=50$ with the burn-in $n_{\text{burn}} = 100$.
+We solved the least squares  problems \eqref{eq:Qregr} using the second-order polynomial approximations for the coefficients \(\hat{a}_{p,\mathbf{k}}\) as described in the previous section, we also set $K = 2$.
+We used $T= 5 \times 10^4$ independent training trajectories, each of size $n=50$ with the burn-in $n_{\text{burn}} = 100$.
 We set the truncation level \(n_{\mathrm{trunc}} = 50\).
-To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 2 \times 10^3$ and display boxplots of the ergodic averages in Figure~\ref{fig:banana_shape}.
+To test our variance reduction algorithm, we generated \(100\) independent trajectories of length $n = 2 \times 10^3$ and display boxplots of the ergodic averages in Figure~\ref{fig:banana_shape}.
 
@@
 To construct trajectories of length $n = 1 \times 10^4, $ we take step size $\gamma = 0.1$ for the ULA scheme with $N = 10^3$ burn-in steps.
-As in the previous experiment we use the first order polynomials approximations to analytically compute the coefficients $\hat{a}_{p-l,\mathbf{k}},$ based on $T = 10$ training trajectories.
+As in the previous experiment we used the first order polynomials approximations to analytically compute the coefficients $\hat{a}_{p-l,\mathbf{k}},$ based on $T = 10$ training trajectories.
 We put $n_{trunc} = 50$ and $K = 1$.
@@
 In Figure ~\ref{fig:2} we compare our approach to the variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}.
-In the baselines we use first order polynomials for the same reasons as in the Gaussian mixtures example.
+In the baselines we use the first order polynomials for the same reasons as in the Gaussian mixtures example.
 
@@
 $$
-f\left(X_{p}\right) =f\circ G_{p}(X_{0},\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p}).
+f\left(X_{p}\right) =[f\circ G_{p}](X_{0},\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p}).
 $$
@@
 \le\mathsf E_x\left[
-\|\nabla_Z F(x,Z_1,Z_{2},\ldots,Z_q)\|^2
+|\nabla_Z F(x,Z_1,Z_{2},\ldots,Z_q)|^2
 \right],
 $$
-where $\nabla_Z F=(\partial_{Z_1} F,\ldots,\partial_{Z_q} F)$,
-and $\|\cdot\|$ denotes the Euclidean norm.
+where $\nabla_Z F=(\partial_{Z_1} F,\ldots,\partial_{Z_q} F)$.
 Notice that \(\partial_{Z_j} F=\sqrt{\gamma}\,\partial_{y_j} F\) and

### 2020-05-11 aap/Revision/arxiv/vr-mcmc-arxiv.tex
arxiv update

--- 2020-05-09 aap/Revision/vr-mcmc-revision-final.tex
@@
 \end{figure}
+
 In the case of  Gaussian prior $\pi_0(\theta) \sim \mathcal{N}(0,\sigma^2 I_d)$, the unnormalized posterior density takes the form:
@@
 We put $n_{trunc} = 50$ and $K = 1$.
-The target function is taken to be $f(\theta) = \sum_{i=1}^{d}\theta_i$.
+The target function is taken to be $f(\theta) = \sum_{i=1}^{i=d}\theta_i$.
 In order to test our variance reduction algorithm, we generate \(100\) independent test trajectories of length $n = 10^3$.

### 2020-05-11 Code/ODE_samplers.py

@@ -123,7 +123,7 @@ def MALA_ODE(r_seed,Potential,step,params,N, n, d, typ, t = 1.0):
         mu = params["mu"]
         sigma = params["sigma"]
         #step = step*np.array([10.0,0.1,10.0,0.1])
-        step = step*np.array([1.0,0.1,1.0,0.1])
+        step = step*np.array([10.0,0.1,10.0,0.1])
         mu_init = Potential.theta_mle
         sigma_init = mu_init/20.0
         x = mu_init + sigma_init*np.random.randn(d)

### 2020-05-11 aap/Revision/Response_1.tex

@@
 In this work we propose a novel variance reduction approach which is able to reduce the statistical error of MCMC algorithms.
-The main advantage of our method is provable increase of efficiency for Langevin-based MCMC algorithms as measured by  the cost-to-variance criterion.
+The main advantage of our method is provable increase of efficiency for Langevin-based MCMC algorithms as measured by  the cost-to-variance (product of cost and variance) criterion.
 Our approach is based on a new discrete time martingale representation for Markov chains which is of its own interest.
 Therefore we first present this basic result and then discuss in Section 4 variance reduction method based on the derived martingale representation.
-In  Section~4 we  explain in accessible way why our approach leads to a smaller cost-to-variance characteristic.
-These results are based on deep analysis of the variance reduced estimate postponed to Section 5.
+In  Section~4 we  also explain in an accessible way why our approach leads to  smaller cost-to-variance characteristics.
+These results are based on a thorough analysis of the variance reduced estimate postponed to Section 5.
 In the revised version we significantly weaken  assumptions  and prove our main result (Theorem~10 and Corollary~11) for ULA under smoothness  and convexity of the underlying potential outside a ball (assumption  (H2)).
-Note that all assumptions needed are now collected in Theorem~10.
+Note that all assumptions needed are now collected in Theorem~10 and Corollary~11.
 For convenience of the reader we also sketch the main steps of the proof right after
-Theorem~10.
-We decided to present our main theoretical result in the form of bounds for the functions $A_{q,k}$ since these functions play a crucial role in the variance bound (13).
-Given these estimates it is easy to see how to get bounds for the variance.
+Theorem~10. We decided to present our main theoretical result in the form of bounds for the functions $A_{q,k}$ as these functions play a crucial role for deriving  the variance bound (32). 
 \item \textit{...
@@
 The reader is left wondering whether or not the same variance reduction could have been achieved simply by drawing more MCMC samples ...}
+\par
+In the revised version (Section 6.1) we measure the gain of our VR approach by comparing its cost-to-variance characteristic (product of the obtained variance and its cost) to one of the plain MCMC method.
+This characteristic quantifies  the resulting variance reduction effect relative to computational effort spent.
 \item \textit{...

### 2020-05-11 aap/Revision/arxiv/vr-mcmc-arxiv.tex
arxiv update

--- 2020-05-10 aap/Revision/vr-mcmc-revision-final.tex
@@
-\documentclass[bj]{imsart}
+\documentclass[preprint]{imsart}
 \makeatletter
@@
 \subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
-\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_linear_gaussian.png}
+\includegraphics[width=0.45\textwidth]{vr_cost_linear_gaussian.png}
 \label{fig:subfig1}}
@@
 \subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
-\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_quadratic_gaussian.png}
+\includegraphics[width=0.45\textwidth]{vr_cost_quadratic_gaussian.png}
 \label{fig:subfig2}}
@@
 \subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
-\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_linear_gmm.png}
+\includegraphics[width=0.45\textwidth]{vr_cost_linear_gmm.png}
 \label{fig:subfig1}}
@@
 \subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
-\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_quadratic_gmm.png}
+\includegraphics[width=0.45\textwidth]{vr_cost_quadratic_gmm.png}
 \label{fig:subfig2}}
@@
 \subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
-\includegraphics[width=0.45\textwidth]{../pictures/GMM_2d_lin_sum.png}
+\includegraphics[width=0.45\textwidth]{GMM_2d_lin_sum.png}
 \label{fig:subfig1}}
@@
 \subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
-\includegraphics[width=0.45\textwidth]{../pictures/GMM_2d_lin_sum_squares.png}
+\includegraphics[width=0.45\textwidth]{GMM_2d_lin_sum_squares.png}
 \label{fig:subfig2}}
@@
 \begin{center}
-\includegraphics[width=0.6\textwidth]{../pictures/banana_shape.png}
+\includegraphics[width=0.6\textwidth]{banana_shape.png}
 \caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Banana shape density.
@@
 \begin{center}
-\includegraphics[width=0.6\textwidth]{../pictures/logreg_1st_order.png}
+\includegraphics[width=0.6\textwidth]{logreg_1st_order.png}
 \caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Logistic Regression.

### 2020-06-25 Code/samplers.py
update

@@ -44,7 +44,7 @@ def ULA(r_seed,Potential,step, N, n, d, burn_type = "SGLD",main_type = "SGLDFP")
     #select method for gradient updates during main loop   
     traj = np.zeros((n, d))
     traj_grad = np.zeros((n, d))
-    x = np.random.normal(scale=5.0, size=d) # initial value X_0
+    x = np.random.normal(scale=1.0, size=d) # initial value X_0
     for k in np.arange(N): # burn-in period
         grad_burn_val = grad_burn(x)
         x = x + step * grad_burn_val +\
@@ -110,7 +110,7 @@ def ULA_light(r_seed,Potential,step, N, n, d, return_noise = False, x0 = None, f
     #traj_grad = np.zeros((n, d))
     if return_noise:
         noise = np.zeros((n, d))
-    x = np.random.normal(scale=10.0, size=d) # initial value X_0
+    x = np.random.normal(scale=2.0, size=d) # initial value X_0
     for k in np.arange(N): # burn-in period
         grad_burn_val = grad_burn(x)
         x = x + step * grad_burn_val +\
@@ -179,7 +179,7 @@ def MALA(r_seed,Potential, step, N, n, d, burn_type = "SGLD",main_type = "SGLDFP
     grad_main = Potential.gradpotential
     traj = np.zeros((n, d))
     traj_grad = np.zeros((n, d))
-    x = np.random.normal(scale=1.0, size=d)
+    x = np.random.normal(scale=2.0, size=d)
     for k in np.arange(N):
         y = x + step * grad_burn(x) + np.sqrt(2*step)*np.random.normal(size=d)
         #if full gradient computed during burn-in, do acceptance-rejection step
@@ -221,7 +221,7 @@ def RWM(r_seed,Potential, step, N, n, d):
     grad_U = Potential.gradpotential # for control variates only
     traj = np.zeros((n, d))
     traj_grad = np.zeros((n, d))
-    x = np.random.normal(scale=5.0, size=d)
+    x = np.random.normal(scale=2.0, size=d)
     for k in np.arange(N):
         y = x + np.sqrt(2*step)*np.random.normal(size=d)
         logratio = U(y)-U(x)

### 2020-06-25 Code/zv_cv.py

@@ -22,6 +22,7 @@ def ZVpolyOne(traj, traj_grad, f_target, params, W_spec):
     A = np.linalg.inv(cov1)
     covariance = np.cov(np.concatenate((-traj_grad, samples), axis=1), rowvar=False)
     paramZV1 = -np.dot(A,covariance[:d, d:])
+    #print("ZV1: ",paramZV1)
     ZV1 = samples - np.dot(traj_grad, paramZV1)
     mean_ZV1 = np.mean(ZV1, axis = 0)
     var_ZV1 = Spectral_var(ZV1[:,0],W_spec)
@@ -81,8 +82,68 @@ def CVpolyOne(traj,traj_grad, f_target, params, W_spec):
         samples = set_function(f_target,traj,[0],params)
         traj = traj[0]
         samples = samples[0]
+    #covariance = np.cov(np.concatenate((traj, samples), axis=1), rowvar=False)
+    #paramCV1 = covariance[:d, d:]
+    paramCV1 = (np.transpose(traj)@(samples - np.mean(samples)))/traj.shape[0]
+    print("CV1: ",paramCV1)
+    CV1 = samples - np.dot(traj_grad, paramCV1)
+    mean_CV1 = np.mean(CV1, axis = 0)
+    var_CV1 = Spectral_var(CV1[:,0],W_spec)
+    return mean_CV1, var_CV1
+
+def CVpolyOneGaussian(traj,traj_grad, f_target, params, W_spec):
+    """
+    Version of CV's with family of $\psi$ given by 2-dimensional gaussians
+    """
+    n, d = traj.shape
+    if f_target == "sum":
+        samples = traj.sum(axis = 1).reshape(-1,1)
+    elif f_target == "sum_comps":
+        samples = traj[:,params["ind"]].reshape(-1,1)
+    elif f_target == "sum_squared":
+        samples = np.square(traj).sum(axis = 1).reshape(-1,1)
+    elif f_target == "sum_4th":
+        samples = ((traj)**4).sum(axis = 1).reshape(-1,1)
+    elif f_target == "exp_sum":
+        samples = np.exp(traj.sum(axis = 1)).reshape(-1,1)
+    else:
+        traj = np.expand_dims(traj, axis = 0)
+        samples = set_function(f_target,traj,[0],params)
+        traj = traj[0]
[…]

### 2020-06-26 Code/zv_cv.py
update

@@ -7,6 +7,8 @@ def ZVpolyOne(traj, traj_grad, f_target, params, W_spec):
         samples = traj.sum(axis = 1).reshape(-1,1)
     elif f_target == "sum_comps":
         samples = traj[:,params["ind"]].reshape(-1,1)
+    elif f_target == "sum_comps_squared":
+        samples = np.square(traj[:,params["ind"]]).reshape(-1,1)
     elif f_target == "sum_squared":
         samples = np.square(traj).sum(axis = 1).reshape(-1,1)
     elif f_target == "sum_4th":
@@ -19,7 +21,10 @@ def ZVpolyOne(traj, traj_grad, f_target, params, W_spec):
         traj = traj[0]
         samples = samples[0]
     cov1 = np.cov(traj_grad, rowvar=False)
-    A = np.linalg.inv(cov1)
+    if d==1 :
+        A = 1/cov1
+    else:
+        A = np.linalg.inv(cov1)
     covariance = np.cov(np.concatenate((-traj_grad, samples), axis=1), rowvar=False)
     paramZV1 = -np.dot(A,covariance[:d, d:])
     #print("ZV1: ",paramZV1)
@@ -36,6 +41,8 @@ def ZVpolyTwo(traj, traj_grad, f_target, params, W_spec):
         samples = traj[:,params["ind"]].reshape(-1,1)
     elif f_target == "sum_squared":
         samples = np.square(traj).sum(axis = 1).reshape(-1,1)
+    elif f_target == "sum_comps_squared":
+        samples = np.square(traj[:,params["ind"]]).reshape(-1,1)
     elif f_target == "sum_4th":
         samples = ((traj)**4).sum(axis = 1).reshape(-1,1)
     elif f_target == "exp_sum":
@@ -54,9 +61,11 @@ def ZVpolyTwo(traj, traj_grad, f_target, params, W_spec):
             Lpoisson[:,k] = -np.multiply(traj_grad[:,i], traj[:,j]) \
                     -np.multiply(traj_grad[:,j], traj[:,i])
             k=k+1
-    
     cov1 = np.cov(Lpoisson, rowvar=False)
-    A = np.linalg.inv(cov1)
+    if cov1.shape[0] == 1:
+        A = 1/cov1
+    else:
+        A = np.linalg.inv(cov1)
[…]

### 2020-06-28 Code/zv_cv.py
update

@@ -350,59 +350,5 @@ def Eval_ZVCV(traj,traj_grad, f_target, params, W_spec):
     #mean_CV1, var_CV1 = CVpolyOneUpdated(traj,traj_grad,f_target,params,W_spec)
     mean_CV1, var_CV1 = CVpolyOne(traj,traj_grad,f_target,params,W_spec)
     mean_CV2, var_CV2 = CVpolyTwo(traj,traj_grad,f_target,params,W_spec)
-    return (mean_vanilla,mean_ZV1, mean_ZV2, mean_CV1, mean_CV2), (vars_vanilla, var_ZV1, var_ZV2, var_CV1, var_CV2)
-
-def ZVpoly1d(traj,traj_grad,f_target,deg,lamda_reg,W_spec):
-    if f_target == "sum":
-        samples = traj[:,0].reshape(-1,1)
-    else:
-        raise "Unexpected function type in Eval_ZVCV_1d"
-    L_psi = np.zeros(deg,samples.shape[0])
-    for i in range(deg):
-        L_psi[i] = (i+1)*traj_grad*(traj**i) + i*(i+1)*(traj**(i-1))
-    #compute main matrix
-    H_zv = np.dot(L_psi,L_psi.T)/samples.shape[0]
-    #compute right side
-    b_zv = (L_psi.T @ (samples - samples.mean()))/samples.shape[0]
-    theta = np.linalg.inv(H_zv + lambda_reg*np.eye(deg)) @ b_zv
-    ZV_est = samples - L_psi.T @ theta
-    mean_ZV = np.mean(ZV_est, axis = 0)
-    var_ZV = Spectral_var(ZV_est[:,0],W_spec)
-    return mean_ZV, var_ZV
-
-def CVpoly1d(traj,traj_grad,f_target,deg,lamda_reg,W_spec):
-    if f_target == "sum":
-        samples = traj[:,0].reshape(-1,1)
-    else:
-        raise "Unexpected function type in Eval_ZVCV_1d"
-    Nabla_psi = np.zeros(deg,samples.shape[0])
-    Psi = np.zeros(deg,samples.shape[0])
-    L_psi = np.zeros(deg,samples.shape[0])
-    for i in range(deg):
-        Nabla_psi[i] = (i+1)*(traj**i)
-        Psi[i] = traj**(i+1)
-        L_psi[i] = (i+1)*traj_grad*(traj**i) + i*(i+1)*(traj**(i-1))
[…]
